### 2024-01-30 main.tex
First push

\documentclass{rapport}

\usepackage{myTools}

\usepackage{macros}

\author{Davide Fissore}
\title{Report - HO unification in Elpi} 
\date{January 29, 2024}

\xspaceaddexceptions{]\}}

\begin{document}

\maketitle

\def\elpi{\proglang{elpi}}
\def\coqelpi{\proglang{coq-elpi}}
\def\lambdaprolog{\proglang{$\lambda$-prolog}}
\def\coq{\proglang{coq}}
\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}

\section{Higher-order unification in \texorpdfstring{$L_\lambda$}{L\_lambda}}

Higher-order unification is a unification problem known to be undecidable. The
goal is to determine a most general unifier between two lambda terms, where at
least one of contain a sub-term $F$ applied to a non empty list of arguments
$a_1,\dots,a_n$ and $F$ is a variable. Some researches have been done in order to
identify some easy patterns where it is possible to solve this kind of
unification.

In \cite{miller1991}, D. Miller provides a unification algorithm (the function
$\eqtau$) for higher-order variable when they are in the so-called pattern
fragment.

\section{From the object to the target language}

Higher order unification becomes difficult when we try to unify terms
represented in the syntax of an object language embedded inside another
language. This kind of interaction between different languages is something that
happens quite often in \coqelpi, a \coq plugin embedding \elpi.

\elpi is a variant of \lambdaprolog equipped with constraint handling rules. On
the other hand, \coq is a proof assistant used to reason about mathematical
properties and certify them thanks to its typing system.

Both languages allows to quantify higher-order variables and are able to perform
higher-order unification when needed. However, when we want to reason with the
\coq unification inside \elpi, it is less evident to reproduce the unification
between two terms. With this paper, we want to provide a framework explaining
how the unification of an object language (\coq in our case) can be expressed in
the targeted language (\elpi in our case).

Without any loss of generality, in \cref{table:coq2elpi} we present a subset of
the terms in the object language and for each term we give its representation in
the targeted language.

\begin{table}
  \centering
  \begin{tabular}{c||c|c}
  Term        & Obj. lang.                    & Targ. lang.                                   \\
  \hline\hline
  Constant    & a                             & `a'                                     \\
  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
  Variable    & X                             & X                                             \\
  \end{tabular}
  \caption{Obj. lang. terms to targ. lang. terms representation}
  \label{table:coq2elpi}
\end{table}

\section{First-order unification}

Just as an introduction, we briefly show some small example of unification
between terms with only first-order unification variables. This way, we would
like the reader to become familiar between the communication of the two
languages.

Let's take as an example the following unification problem in the obj. lang.:

\begin{equation}
  f\ x\ 1 \eqtau f\ Y\ Z
  \label{eq:unif-fo-coq}
\end{equation}

\noindent where $f$, $x$ and $1$ are defined constant and $Y$ and $Z$ are both
unification variables. By convention we use upper case letter for quantified
variables. Moreover, for this first representation we do not really focus on the
type of the manipulated objects, since they do not condition the unification
algorithm.

It is quite evident that a valid substitution for \cref{eq:unif-fo-coq} is $\theta =
\{Y \mapsto `x', Z \mapsto `1'\}$. Now let's consider the same problem translated in
the targeted language.

\begin{equation}
  app[`f', `x', `1'] \eqtau app[`f', Y, Z]
  \label{eq:unif-fo-elpi}
\end{equation}

The unification of these terms is again quite simple since it is sufficient to
do a simple matching sub-term by sub-term so that variables can be instantiated.
We can therefore note that the same substitution $\theta$ will be produced.

\section{Higher-order unification}

The unification problem treated before was enough easy to be correctly
understood by both language representation. We want now to go a bit further and
reason with a more complex problem where a variable is a function of
higher-order.

We propose two different higher-order unification problem in the following
equations where, in the former we have rigid-flexible unification and in the
latter we have a flexible-flexible unification.

\begin{equation}
  f\ x\ 1 \eqtau F\ x
  \label{eq:unif-ho-coq1}
\end{equation}

\begin{equation}
  G\ x\ y \eqtau H\ y\ x
  \label{eq:unif-ho-coq2}
\end{equation}

The two substitutions for the previous examples are $\theta_1 = \{F \mapsto fun\
x \Rightarrow f\ x\ 1\}$ and $\theta_2 = \{H \mapsto fun\ y\ x \Rightarrow G\ x\
y\}$. We can note that to be in the pattern fragment, a functional variable
should be applied to distinct names.

If we translate the problem before in the target language, the unification
problems showed above become

\begin{equation}
  app[`f',\ `x',\ `1'] \eqtau app[F,\ `x']
  \label{eq:unif-ho-elpi1}
\end{equation}

\begin{equation}
  app[G,\ `x',\ `y'] \eqtau app[H,\ `y',\ `x']
  \label{eq:unif-ho-elpi2}
\end{equation}

Now, the new unification problems are no more expressed in the logic of the
targeted language and, therefore, in both cases, unification fails. The
procedure we can adopt in order to transform a higher-order unification problem
of the object language into the logic of the targeted language is to transform
the entry of the problem in a problem which can be understood by the targeted
language. The procedure is made of two steps:

\begin{enumerate}
  \item In the first place, we need to recognize the structure of the pattern
        fragment expressed in the term received in entry. This means that we
        need to find all the sub-terms of the form `app[X | L]', where `X' is
        a flexible variable and `L' is a list of distinct names.
  \item For any sub-term representing a higher-order unification in the object 
        language, we build a fresh variable `X\texttt{'}' such that the names 
        `L' are not in the scope of `X\texttt{'}', we call `X\texttt{'}' the 
        twin variable of `X'.
  \item We solve the new goal where each pattern fragment problem is replaced
        with a problem using twin variables and after each of these problems, we
        add a new premise linking these twin variables. The linking is done
        using the following criteria: for each abstraction in the resulting term
        `X\texttt{'}', unify recursively `X' to a lambda abstraction in the
        object language.
\end{enumerate}

The previous algorithm can be applied to \cref{eq:unif-ho-coq1,eq:unif-ho-coq2}
to provide the wanting solution. In particular, \cref{eq:unif-ho-coq1} is
transformed into the unification problem:

\begin{equation}
  f\ x\ 1 \eqtau F'\ x,\ ho-link\ F'\ F
  \label{eq:unif-ho-elpi1'}
\end{equation}

\begin{equation}
  G'\ x\ y \eqtau H'\ y\ x, ho-link\ G'\ G, ho-link\ H'\ H
  \label{eq:unif-ho-elpi2'}
\end{equation}

For instance, the former unification problem produce the substitution $\theta_1
= \{F' \mapsto (x\backslash\ f\ x\ 1)\}$. The ho-link function is then applied
to transform the substitution of $F'$ into the corresponding term of the object
langue: $F \mapsto fun \_\ \_\ (x\backslash\ f\ x\ 1)$ which correspond to the
term $fun\ x \Rightarrow f\ x\ 1$. The latter unification problem gives the
substitution $\theta_2 = \{H' \mapsto (y\ x \backslash G'\ x\ y)\}$ in the
targeted language. The first ho-link simply unify $G$ to $G'$ since $G'$ is
flexible, whereas $H$ is mapped to $fun\ y\ x \Rightarrow G'\ x\ y$.

The role of the ho-link is not only to instantiate the higher-order variable F
of the object language when F is flexible and the twin variable in the targeted
language is rigid. It may happen that F has already been partially instantiated.
The unification problem below gives such an example in the object language:

\begin{equation}
  G\ x\ y \eqtau H\ y\ x, H\ x\ y \eqtau x 
  \label{eq:unif-ho-coq4}
\end{equation}

\noindent producing the following substitution $\theta = \{G \mapsto (fun\ x\
y\Rightarrow x); F \mapsto (fun\ x\ y\Rightarrow y)\}$. This unification problem
is translated into:

\begin{equation}
  \begin{split}
    G'\ x\ y &\eqtau H'\ y\ x,\ ho-link\ G'\ G,\ ho-link\ H'\ H\\
    H''\ x\ y &\eqtau x,\ ho-link\ H''\ H
  \end{split}
  \label{eq:unif-ho-elpi4}
\end{equation}

\noindent The first line produces the same substitution as before: $\theta = \{H'
\mapsto (x\ y \backslash G'\ y\ x)\}$, where $G'$ is unified with $G$ and $H$ is
instantiated to <<$fun\ x\ y \Rightarrow G'\ y\ x$>>. While executing the second
line, we see $H''$ instantiated to the function $(x\ y\backslash x)$ in the
targeted language. The last ho-link is charged to link $H''$ with $H$ but this
time $H$ has already been partially instantiated. In particular the call to this
ho-link is as follows: $ho-link\ (x\ y\backslash x)\
(prod\ \_\ \_\ (x\backslash\ prod\ \_\ \_\ (y\backslash\ app [G', x, y])))$.

Since the two arguments have rigid heads, we start to traverse both terms
recursively by eating each lambda-abstraction. At the end of this procedure, the
remaining sub-terms are \todo{in our code this example though the eta mess
error} now $x$ and $app[G', x, y]$,

\section{HO unification in typed languages}

TODO: ho-link need the type of original term to produce a typed term in the
object language, example: f x 1 = F x => type of F = (A -> Prof) if type of f is 
(A -> nat -> Prop)

\section{Proof automation from \coq to \elpi}

TODO: representing a logic programming language into an other: compile rules
keeping higher order unification

\subsection{Dealing with FO \textit{non-syntactical} unification}

\subsection{Dealing with HO unification}

\end{document}

### 2024-01-30 elpi.py

@@ -1,0 +1,179 @@
+"""
+    pygments.lexers.elpi
+    ~~~~~~~~~~~~~~~~~~~~
+
+    Lexer for the `Elpi <http://github.com/LPCIC/elpi>`_ programming language.
+
+    :copyright: Copyright 2006-2023 by the Pygments team, see AUTHORS.
+    :license: BSD, see LICENSE for details.
+"""
+
+from pygments.lexer import RegexLexer, bygroups, include
+from pygments.token import Text, Comment, Operator, Keyword, Name, String, \
+    Number, Punctuation
+
+__all__ = ['ElpiLexer']
+
+
+class ElpiLexer(RegexLexer):
+    """
+    Lexer for the Elpi programming language.
+
+    .. versionadded:: 2.11
+    """
+
+    name = 'Elpi'
+    url = 'http://github.com/LPCIC/elpi'
+    aliases = ['elpi']
+    filenames = ['*.elpi']
+    mimetypes = ['text/x-elpi']
+
+    lcase_re = r"[a-z]"
+    ucase_re = r"[A-Z]"
+    digit_re = r"[0-9]"
+    schar2_re = r"([+*^?/<>`'@#~=&!])"
+    schar_re = r"({}|-|\$|_)".format(schar2_re)
+    idchar_re = r"({}|{}|{}|{})".format(lcase_re, ucase_re, digit_re, schar_re)
+    idcharstarns_re = r"({}*(\.({}|{}){}*)*)".format(idchar_re,
+                                                     lcase_re, ucase_re, idchar_re)
+    symbchar_re = r"({}|{}|{}|{}|:)".format(
+        lcase_re, ucase_re, digit_re, schar_re)
+    constant_re = r"({}{}*|{}{}|{}{}*|_{}+)".format(ucase_re, idchar_re,
+                                                    lcase_re, idcharstarns_re, schar2_re, symbchar_re, idchar_re)
+    symbol_re = r"(,|<=>|->|:-|;|\?-|->|&|=>|\bas\b|\buvar\b|<|=<|=|==|>=|>|\bi<|\bi=<|\bi>=|\bi>|\bis\b|\br<|\br=<|\br>=|\br>|\bs<|\bs=<|\bs>=|\bs>|@|::|\[\]|`->|`:|`:=|\^|-|\+|\bi-|\bi\+|r-|r\+|/|\*|\bdiv\b|\bi\*|\bmod\b|\br\*|~|\bi~|\br~)"
[…]

### 2024-01-30 pygmentize.py

@@ -1,0 +1,53 @@
+#! /usr/bin/env python
+import argparse
+import sys
+import pygments.cmdline as _cmdline
+import pygments.lexer as _lexer
+import pygments.token as _token
+
+def main(args):
+    parser = argparse.ArgumentParser()
+    parser.add_argument('-l', dest='lexer', type=str)
+    opts, rest = parser.parse_known_args(args[1:])
+    if opts.lexer == 'ssh_config':
+        args = [__file__, '-l', __file__ + ':SSHConfigLexer', '-x', *rest]
+    if args[2] == "elpi": 
+        args[2] = './elpi.py:ElpiLexer'
+    _cmdline.main(args + ['-x'])
+
+
+class SSHConfigLexer(_lexer.RegexLexer):
+    name = 'ssh_config'
+    tokens = {
+        'root': [
+            (r'(\s*Host)( .*\n)', _lexer.bygroups(
+                _token.Keyword, _token.String)),
+            (r'(\s*\w*)( ?.*\n)', _lexer.bygroups(
+                _token.Name.Attribute, _token.String)),
+        ],
+    }
+
+def isEmptyLine(x: str):
+    return x.lstrip(' ') == '\n'
+
+def getSpaceOfLine(x: str):
+    if isEmptyLine(x): 
+        return float('inf')
+    return len(x) - len(x.lstrip(' '))
+
+def getMinSpace(l):
+    mLen = min(map(getSpaceOfLine, l))
+    return [line[mLen:] if not(isEmptyLine(line)) else "\n" for line in l]
+
+import os
+
+if __name__ == '__main__':
+    f_name = ".aux/" + list(filter(lambda x: x.endswith("pyg"), os.listdir(".aux")))[0]
+    with open(f_name) as X:
+        lines = X.readlines()
+    with open(f_name, "w") as X:
+        X.write("".join(getMinSpace(lines)))
+    # with open("tt.txt", "w") as DUMP:
+    #     DUMP.write(str(os.listdir()))
+    #     DUMP.write("".join(getMinSpace(lines)))
+    main(sys.argv)

### 2024-02-05 main.tex
print bib
Mv file in elpi_code folder
Update tex

@@
 
-\author{Davide Fissore}
-\title{Report - HO unification in Elpi} 
-\date{January 29, 2024}
+\author{Enrico Tassi, Davide Fissore}
+\title{HO unification from object language to meta language} 
+\date{Month XX, 202Y}
 
@@
 
-\section{Higher-order unification in \texorpdfstring{$L_\lambda$}{L\_lambda}}
+\begin{abstract}
 
-Higher-order unification is a unification problem known to be undecidable. The
-goal is to determine a most general unifier between two lambda terms, where at
-least one of contain a sub-term $F$ applied to a non empty list of arguments
-$a_1,\dots,a_n$ and $F$ is a variable. Some researches have been done in order to
-identify some easy patterns where it is possible to solve this kind of
-unification.
+Unification is a fundamental process behind goal resolution in logic
+programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
+exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
+substitution is a mapping from unification variables to terms and it is the most
+general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
+t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
+Unification has been widely studied in the 90s [here the refs]. In this paper we
+want to give a framework to reason about unification in meta-programming.
 
-In \cite{miller1991}, D. Miller provides a unification algorithm (the function
-$\eqtau$) for higher-order variable when they are in the so-called pattern
-fragment.
+\end{abstract}
 
-\section{From the object to the target language}
+\section{Introduction}
 
-Higher order unification becomes difficult when we try to unify terms
-represented in the syntax of an object language embedded inside another
-language. This kind of interaction between different languages is something that
-happens quite often in \coqelpi, a \coq plugin embedding \elpi.
+Meta programming \cite{hill1994} is a programming technique in which a program
+can treat an other program as its data. This latter program is called object
+language, while the former is called meta language. At the heart of meta
+programming lays the necessity of representing terms of the object language in
+the meta language so that a wide set of program manipulation ranging from
+interpretation to compilation. Furthermore, since a logic program is a set of
+clauses, the behavior of the object language can be modified by simply adding
+new rules inside the database of the meta program.
 
-\elpi is a variant of \lambdaprolog equipped with constraint handling rules. On
-the other hand, \coq is a proof assistant used to reason about mathematical
-properties and certify them thanks to its typing system.
+The motivation pushing us to reason about unification of terms of the object
+language represented as terms of the meta language is to implement a type class
+solver for the \coq proof assistant into the \elpi logic programming language. A
+type class \cite{wadler89, sozeau08} is a typing structure allowing to introduce
+ad hoc polymorphism in functional languages and instances are implementations of
+type classes. From the point of view of logic programming, type class represent
+predicates parametrized by their argument and instances rules for those
+predicates. For instance, if \texttt{XXX} is the type class representing the
+AAA, then ZZZ and WWW are instances for XXX. In the code snippet below, we give
+such implementation in \coq.
 
-Both languages allows to quantify higher-order variables and are able to perform
-higher-order unification when needed. However, when we want to reason with the
-\coq unification inside \elpi, it is less evident to reproduce the unification
-between two terms. With this paper, we want to provide a framework explaining
-how the unification of an object language (\coq in our case) can be expressed in
-the targeted language (\elpi in our case).
+\begin{mintedTcolorbox}{coq}
+  Class XXX Args := {methods ...}.
+  Instance ZZZ : bla bla -> XXX Args := {...} Qed.
+  Instance WWW : bla bla -> XXX Args := {...} Qed.
+\end{mintedTcolorbox}
+
+This small set of instance thanks a first phase of class and instance\todo{explain compilation of pred and inst?}
+compilation is translated into the following \elpi program:
+
+\begin{mintedTcolorbox}{elpi}
+  type XXX Args .. -> term -> o.
+  XXX Args ... (ZZZ P1 ... PN) :-
+    premese...
+  WWW Args ... (ZZZ P1 ... PN) :-
+    premese...
+\end{mintedTcolorbox}
+
+Type-class resolution starts from a query, that is a class applied to some
+argument. This \coq term is translated into a term of the meta language and the
+search for a solution in the database is started. However, since unification is
+performed on the head of the clause, it may happen that the term representation
+in the meta language is not sufficiently powerful to understand that two terms
+unifying in the object language should also unify in the meta language.
+
+% \section{Higher-order unification in \texorpdfstring{$L_\lambda$}{L\_lambda}}
+
+% Higher-order unification is a unification problem known to be undecidable. The
+% goal is to determine a most general unifier between two lambda terms, where at
+% least one of contain a sub-term $F$ applied to a non empty list of arguments
+% $a_1,\dots,a_n$ and $F$ is a variable. Some researches have been done in order to
+% identify some easy patterns where it is possible to solve this kind of
+% unification.
+
+% In \cite{miller1991}, D. Miller provides a unification algorithm (the function
+% $\eqtau$) for higher-order variable when they are in the so-called pattern
+% fragment.
+
+\section{From the object to the meta language}
+
+In \coq, terms unify up to $\eta\beta\delta\iota\zeta$-equivalence and
+unification is also possible when dealing with higher order variables. Whereas
+\elpi allows higher order unification up to $\eta\beta$-equivalence, and the
+representation of \coq terms is more syntactical than one would expect, making 
+two terms that unify in \coq not unifiable in \elpi.
 
@@
 the terms in the object language and for each term we give its representation in
-the targeted language.
+the meta language.
 
@@
   \begin{tabular}{c||c|c}
-  Term        & Obj. lang.                    & Targ. lang.                                   \\
+  Term        & Obj. lang.                    & Meta lang.                                   \\
   \hline\hline
@@
   \end{tabular}
-  \caption{Obj. lang. terms to targ. lang. terms representation}
+  \caption{Obj. lang. terms to meta lang. terms representation}
   \label{table:coq2elpi}
@@
 
-\section{First-order unification}
+\subsection{First-order unification}
 
@@
 \{Y \mapsto `x', Z \mapsto `1'\}$. Now let's consider the same problem translated in
-the targeted language.
+the meta language.
 
@@
 
-\section{Higher-order unification}
+\subsection{Higher-order unification}
 
@@
 
-If we translate the problem before in the target language, the unification
+If we translate the problem before in the meta language, the unification
 problems showed above become
@@
 Now, the new unification problems are no more expressed in the logic of the
-targeted language and, therefore, in both cases, unification fails. The
+meta language and, therefore, in both cases, unification fails. The
 procedure we can adopt in order to transform a higher-order unification problem
-of the object language into the logic of the targeted language is to transform
-the entry of the problem in a problem which can be understood by the targeted
+of the object language into the logic of the meta language is to transform
+the entry of the problem in a problem which can be understood by the meta
 language. The procedure is made of two steps:
@@
 substitution $\theta_2 = \{H' \mapsto (y\ x \backslash G'\ x\ y)\}$ in the
-targeted language. The first ho-link simply unify $G$ to $G'$ since $G'$ is
+meta language. The first ho-link simply unify $G$ to $G'$ since $G'$ is
 flexible, whereas $H$ is mapped to $fun\ y\ x \Rightarrow G'\ x\ y$.
@@
 The role of the ho-link is not only to instantiate the higher-order variable F
-of the object language when F is flexible and the twin variable in the targeted
+of the object language when F is flexible and the twin variable in the meta
 language is rigid. It may happen that F has already been partially instantiated.
@@
 line, we see $H''$ instantiated to the function $(x\ y\backslash x)$ in the
-targeted language. The last ho-link is charged to link $H''$ with $H$ but this
+meta language. The last ho-link is charged to link $H''$ with $H$ but this
 time $H$ has already been partially instantiated. In particular the call to this
@@
 
+\printbibliography
+
 \end{document}

### 2024-02-06 main.tex
update tex

@@
 
-This small set of instance thanks a first phase of class and instance\todo{explain compilation of pred and inst?}
-compilation is translated into the following \elpi program:
+This small set of instances after a first phase of\todo{explain compilation of pred and inst?}
+compilation is translated into the following \elpi rules:
 
@@
   type XXX Args .. -> term -> o.
+  XXX Args ... (WWW P1 ... PN) :-
+    premese...
   XXX Args ... (ZZZ P1 ... PN) :-
-    premese...
-  WWW Args ... (ZZZ P1 ... PN) :-
     premese...
@@
 performed on the head of the clause, it may happen that the term representation
-in the meta language is not sufficiently powerful to understand that two terms
-unifying in the object language should also unify in the meta language.
+in the meta language may hide some unification properties that are true in the
+object language.
 
@@
 the terms in the object language and for each term we give its representation in
-the meta language.
+the meta language. This encoding is enough to fall into unification problems.
+For example, rules like
+
+\begin{center}
+  \elpiIn{p (lam x\ (app F x)) :- ... F ...}
+\end{center}
+
+cannot unify with goals like
+
+\begin{center}
+  \elpiIn{p (lam x\ app (app (const "f") x) x).}
+\end{center}
+
+Since \elpiIn{F} does not see \elpiIn{x} even though it is applied to \elpiIn{x}
+in the object language.
 

### 2024-02-06 main.tex
crazy

--- 2024-01-30 main.tex
@@
 \newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
+
+\section{intro}
+
+we are interested in using a meta language in LF style to write
+automation, proof search. In various works... they achieve that
+for a OL which is simpler than the LF, the equational theory is included
+in the one of the ML. This is exploited to piggy back on the unif
+of the ML.
+the peculiarity of our setting is that the OL has a richer equational
+theory of the ML, eg beta eta zeta bla bla. Moreover is HO logic, so
+quantifies over functions, so unif variable range on that too.
+We want to piggy back on the ML unif whenever the problem fits
+in its domain, eg pattern fragment.
+this is important for practical purposes.
+
+\subsection{in a nutshell}
+example, a rule for theorem
+$$
+forall f : A -> B, .. bla f .. -> prove (\forall x, f x)
+bla g
+$$
+\begin{minted}{elpi}
+type app ..
+prove (forall x\ app F x) :- ... bla F ...
+\end{minted}
+on goal \mint{elpi}|prove (forall x\ app (app g x) x))|
+would fail since \mint{elpi}|F != app g x|.
+
+of course one wants to avoid
+\begin{minted}{elpi}
+prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assignment "F" S' F, bla F S' ...
+\end{minted}
+
+Now, ML has HO variables
+\begin{minted}{elpi}
+type lam ..
+prove (forall x\ F x) :- ... bla F ...
+\end{minted}
+this time  \mint{elpi}|F x != app (app g x) x| has solution
+\mint{elpi}|F = a\ app (app g x) x| but F is not a term so
+bla needs to be adapted,
+\begin{minted}{elpi}
+prove (forall x\ F x) :- ... bla (lam F) ...
+bla g.
+\end{minted}
+this is too simplistic since \mint{elpi}|g != lam x\ app g x|
+
+\subsection{contribution}
+\begin{itemize}
+\item prover for HO OL in ML that uses unif
+\item eta beta
+\item test on stdpp and TLC
+\end{itemize}
 

### 2024-02-06 main.tex
tex add example of FO and HO unification problems from stdpp

--- 2024-02-06 main.tex
@@
 
-\begin{mintedTcolorbox}{coq}
-  Class XXX Args := {methods ...}.
-  Instance ZZZ : bla bla -> XXX Args := {...} Qed.
-  Instance WWW : bla bla -> XXX Args := {...} Qed.
+\begin{mintedTcolorbox}[escapeinside=&&]{coq}
+  Inductive sig (A : Type) (P : A -> Prop) : Type := ...
+
+  Class Decision (P : Prop) := decide : {P} + {not P}.
+  Class RelDecision {A B: Type} (R : A -> B -> Prop).
+  Class ProofIrrel (A : Type) : Prop := proof_irrel (x y : A) : x = y.
+
+  Instance decide_rel: forall (A B : Type) (R : A -> B -> Prop),
+    RelDecision R -> forall (x : A) (y : B), Decision (R x y). Admitted.
+  Instance True_pi : ProofIrrel True. Admitted.
+  Instance sig_eq_dec: forall (A : Type) (P : A -> Prop),
+    (forall x, ProofIrrel (P x)) -> RelDecision (@eq A) ->
+    RelDecision (@eq (sig A P)). Admitted.
 \end{mintedTcolorbox}
@@
 \begin{mintedTcolorbox}{elpi}
-  type XXX Args .. -> term -> o.
-  XXX Args ... (WWW P1 ... PN) :-
-    premese...
-  XXX Args ... (ZZZ P1 ... PN) :-
-    premese...
+  type tc-Decision term -> term -> o.
+  type tc-RelDecision term -> term -> term -> term -> o.
+  type tc-ProofIrrel term -> term -> o.
+
+  tc-ProofIrrel (`True`) (`True_pi`).
+  tc-Decision (app [R, X, Y]) 
+    (app [`decide_rel`, A, B, R, P, X, Y]) :-
+    tc-RelDecision A B R P.
+  tc-RelDecision (app [`sig`, A, P]) 
+    (app [`sig`, A, P]) 
+    (app [`eq`, app [`sig`, A, P]]) 
+    (app [`sig_eq_dec`, A, P, P1, P2]) :-
+    pi-decl c0 `x` A =>
+      tc-ProofIrrel (app [P, c0]) (app [P1, c0]), 
+    tc-RelDecision A A (app [`eq`, A]) P2.
 \end{mintedTcolorbox}
@@
 in the meta language may hide some unification properties that are true in the
-object language.
+object language. In the example below, the goal \texttt{Decision (@eq T a b)}
+for some \texttt{a} and \texttt{b} unifies with \texttt{Decision (R x y)} in the
+object language but not in its meta representation. Similarly, the goal
+\texttt{RelDecision (@eq (sig T ?P))} where \texttt{?P}, under the hypothesis
+\texttt{RelDecision (@eq nat)}, will trying to apply the rule for
+\texttt{sig\_eq\_dec}, we fall into an higher order unification problem, where
+\texttt{P} is applied to the local name \texttt{x}. However, the corresponding
+rule in the meta language exploit a first order variable \texttt{P}. Therefore,
+after the refinement of the goal to \texttt{sig\_eq\_dec}, the resolution
+immediately fail to solve the premise \texttt{tc-ProofIrrel (app [P, c0]) (app
+[P1, c0])}.
 

### 2024-02-11 main.tex
Update tex

@@
 \def\coq{\proglang{coq}}
+
+\newcommand*{\acronym}[1]{\texttt{#1}\xspace}
+
+\def\ol{\acronym{ol}}
+\def\ml{\acronym{ml}}
+
 \newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
@@
 can treat an other program as its data. This latter program is called object
-language, while the former is called meta language. At the heart of meta
-programming lays the necessity of representing terms of the object language in
-the meta language so that a wide set of program manipulation ranging from
-interpretation to compilation. Furthermore, since a logic program is a set of
-clauses, the behavior of the object language can be modified by simply adding
-new rules inside the database of the meta program.
+language (\ol for short), while the former is called meta language (\ml for
+short). At the heart of meta programming lays the necessity of representing
+terms of the \ol in the \ml so that a wide set of program
+manipulations ranging from interpretation to compilation. Furthermore, since a
+logic program is a set of clauses, the behavior of the \ol can be
+modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.
 
-The motivation pushing us to reason about unification of terms of the object
-language represented as terms of the meta language is to implement a type class
-solver for the \coq proof assistant into the \elpi logic programming language. A
-type class \cite{wadler89, sozeau08} is a typing structure allowing to introduce
-ad hoc polymorphism in functional languages and instances are implementations of
-type classes. From the point of view of logic programming, type class represent
-predicates parametrized by their argument and instances rules for those
-predicates. For instance, if \texttt{XXX} is the type class representing the
-AAA, then ZZZ and WWW are instances for XXX. In the code snippet below, we give
-such implementation in \coq.
+Meta programming has various application such as ... where thanks to meta
+programming it is possible to represent the logic of a language into a formal
+and formally verify the wanted properties. On the other hand, it is possible to
+embed a logic programming language into another so that some tasks can be
+delegated to \ml.
 
-\begin{mintedTcolorbox}[escapeinside=&&]{coq}
-  Inductive sig (A : Type) (P : A -> Prop) : Type := ...
-
-  Class Decision (P : Prop) := decide : {P} + {not P}.
-  Class RelDecision {A B: Type} (R : A -> B -> Prop).
-  Class ProofIrrel (A : Type) : Prop := proof_irrel (x y : A) : x = y.
-
-  Instance decide_rel: forall (A B : Type) (R : A -> B -> Prop),
-    RelDecision R -> forall (x : A) (y : B), Decision (R x y). Admitted.
-  Instance True_pi : ProofIrrel True. Admitted.
-  Instance sig_eq_dec: forall (A : Type) (P : A -> Prop),
-    (forall x, ProofIrrel (P x)) -> RelDecision (@eq A) ->
-    RelDecision (@eq (sig A P)). Admitted.
-\end{mintedTcolorbox}
-
-This small set of instances after a first phase of\todo{explain compilation of pred and inst?}
-compilation is translated into the following \elpi rules:
-
-\begin{mintedTcolorbox}{elpi}
-  type tc-Decision term -> term -> o.
-  type tc-RelDecision term -> term -> term -> term -> o.
-  type tc-ProofIrrel term -> term -> o.
-
-  tc-ProofIrrel (`True`) (`True_pi`).
-  tc-Decision (app [R, X, Y]) 
-    (app [`decide_rel`, A, B, R, P, X, Y]) :-
-    tc-RelDecision A B R P.
-  tc-RelDecision (app [`sig`, A, P]) 
-    (app [`sig`, A, P]) 
-    (app [`eq`, app [`sig`, A, P]]) 
-    (app [`sig_eq_dec`, A, P, P1, P2]) :-
-    pi-decl c0 `x` A =>
-      tc-ProofIrrel (app [P, c0]) (app [P1, c0]), 
-    tc-RelDecision A A (app [`eq`, A]) P2.
-\end{mintedTcolorbox}
-
-Type-class resolution starts from a query, that is a class applied to some
-argument. This \coq term is translated into a term of the meta language and the
-search for a solution in the database is started. However, since unification is
-performed on the head of the clause, it may happen that the term representation
-in the meta language may hide some unification properties that are true in the
-object language. In the example below, the goal \texttt{Decision (@eq T a b)}
-for some \texttt{a} and \texttt{b} unifies with \texttt{Decision (R x y)} in the
-object language but not in its meta representation. Similarly, the goal
-\texttt{RelDecision (@eq (sig T ?P))} where \texttt{?P}, under the hypothesis
-\texttt{RelDecision (@eq nat)}, will trying to apply the rule for
-\texttt{sig\_eq\_dec}, we fall into an higher order unification problem, where
-\texttt{P} is applied to the local name \texttt{x}. However, the corresponding
-rule in the meta language exploit a first order variable \texttt{P}. Therefore,
-after the refinement of the goal to \texttt{sig\_eq\_dec}, the resolution
-immediately fail to solve the premise \texttt{tc-ProofIrrel (app [P, c0]) (app
-[P1, c0])}.
+The latter situation motivates our works, since we are implementing a type-class
+solver for the \ol \coq in the \ml \elpi (a variant \lambdaprolog). A type class
+\cite{wadler89, sozeau08} is a typing structure allowing to introduce ad hoc
+polymorphism in functional languages. We call <<instance>> an implementation of
+a type class. The resolution of a type-class problem can be viewed as a logic
+program where type classes represent predicates parametrized by their arguments
+and where instances are rules for those predicates.
 
@@
 
-In \coq, terms unify up to $\eta\beta\delta\iota\zeta$-equivalence and
-unification is also possible when dealing with higher order variables. Whereas
-\elpi allows higher order unification up to $\eta\beta$-equivalence, and the
-representation of \coq terms is more syntactical than one would expect, making 
-two terms that unify in \coq not unifiable in \elpi.
+Even though type-class resolution is the motivating example of this paper, we
+provide a general framework allowing to solve reproduce the same unification
+properties of the \ol into the \ml. In other word, if two terms unify in the
+\ol, then they still unify in the \ml.
 
-Without any loss of generality, in \cref{table:coq2elpi} we present a subset of
-the terms in the object language and for each term we give its representation in
-the meta language. This encoding is enough to fall into unification problems.
-For example, rules like
+In the following, we consider the \ol being able to quantify over higher-order
+variables and accepts $\eta-\beta$-reductions. The same unification properties
+are considered valid for the \ml.
 
-\begin{center}
-  \elpiIn{p (lam x\ (app F x)) :- ... F ...}
-\end{center}
+There exist two different ways to encode the \ol in the \ml, we can either deep
+embed the \ol such that any term of the \ol is represented with a corresponding
+predicate. For example, if \texttt{f} is a function of type \coqIn{A -> B} in
+the \ol, then the \ml has the predicate \texttt{p} defined as \elpiIn{type p A'
+-> B' -> o}, where \elpiIn{A'} and \elpiIn{B'} are types corresponding
+respectively to \coqIn{A} and \coqIn{B} in the \ol. In a theorem prover like
+\coq, we can translate theorems like the following statement
 
-cannot unify with goals like
+\begin{equation}
+  \texttt{\coqIn{forall F X, p (f X) (fun x => g x (F x)).}}
+  \label{eq:coq1}
+\end{equation}
 
-\begin{center}
-  \elpiIn{p (lam x\ app (app (const "f") x) x).}
-\end{center}
+where \texttt{p, f} and \texttt{g} are defined constants of the language, into
 
-Since \elpiIn{F} does not see \elpiIn{x} even though it is applied to \elpiIn{x}
-in the object language.
+\begin{equation}
+  \texttt{\elpiIn{p (f X) (x\ g x (F x)).}}
+\end{equation}
 
-\begin{table}
-  \centering
-  \begin{tabular}{c||c|c}
-  Term        & Obj. lang.                    & Meta lang.                                   \\
-  \hline\hline
-  Constant    & a                             & `a'                                     \\
-  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
-  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
-  Variable    & X                             & X                                             \\
-  \end{tabular}
-  \caption{Obj. lang. terms to meta lang. terms representation}
-  \label{table:coq2elpi}
-\end{table}
+\noindent However, even if this encoding is quite appealing since it allows to
+mirror enough straightforwardly the terms of the \ol, we loose the possibility
+the manipulate the terms of the \ol into the \ml. In other words, we have no
+syntax allowing to know if the current term is a constant, an application, a
+lambda abstraction and so on. This is mainly due to the absence of a syntax in
+the encoding of the \ol terms. Moreover, another motivation for using syntax to
+represent terms of the \ol is that  the typing system of the \ol could
+potentially be more expressive than the typing system of the \ml\footnote{This
+is the case for \coq wrt \elpi, since in we have no immediate way to encode
+the dependent types of \coq into \elpi}.
 
+To simplify the understanding of our encoding, in the following code snippet
+we give the typing schema of the \ol terms represented into terms of the \ml.
+
+\begin{mintedTcolorbox}{elpi}
+  kind tm type.
+  type app list tm -> tm.
+  type lam (tm -> tm) -> tm.
+  type c string -> tm.
+  type uv  nat -> tm.
+\end{mintedTcolorbox}
+
+In particular, the type \elpiIn{tm} is the type of the terms of the \ol. The
+function applications of the \ol are represented as a list of \elpiIn{tm}
+prefixed by the constructor \elpiIn{app}. The \elpiIn{lam} constructor,
+represent lambda abstractions of the \ol binding a \elpiIn{tm} into an other
+\elpiIn{tm}. Constants as strings inside the constructor \elpiIn{c}. Lastly,
+unification variables are integers inside the constructor \elpiIn{uv}, where the
+integer is the index of the current variable wrt a list of optional \elpiIn{tm},
+standing for the substitution mapping of the \ol.
+
+This second encoding of the \ol into our \ml translate \cref{eq:coq1} into
+the term:
+
+\begin{equation}
+  \texttt{\elpiIn{app[c "p", app[c "f", uv 0], lam x\ app[c "g", x, app[uv 1, x]]].}}
+  \label{eq:elpi1}
+\end{equation}
+
+This second encoding of the \ol terms is now structured and as a drawback we are
+restricting the unification of the \ol, that is, terms that originally unify
+at the \ol level, do not unify in the \ml.
+
+For example, let \coqIn{a} and \coqIn{b} two defined constants and let's try to
+unify the \ol term
+
+\begin{equation}
+  \texttt{\coqIn{p (f a) (fun x => g x b)}}
+\end{equation}
+
+\noindent corresponding to
+
+\begin{equation}
+  \texttt{\elpiIn{app[c "p", app[c "f", c "a"], lam x\ app[c "g", x, c "b"]].}}
+\end{equation}
+
+\noindent with \cref{eq:coq1} (corresponding to \cref{eq:elpi1}). The
+unification of the \ml is able to instantiate \elpiIn{uv 0} (cf \coqIn{X}) to
+\elpiIn{c "a"}, but we are no longer capable to unify the sub-term
+\elpiIn{app[uv 1, x]} (cf \coqIn{F}) with \elpiIn{c "b"}.
+
+The result of this translation of terms inside the \ml causes a certain lack of
+powerfulness while symbolizing higher-order variables. Recall that we are
+considering a \ml capable to deal with higher-order variables, however, the
+sub-term \elpiIn{app[uv 1, x]} is not expressed into the canonical form where a
+higher-order variable of the \ml is in the pattern fragment \cite{miller1991},
+i.e. a variable applied to distinct names. Therefore, we need to preprocess the
+received unification problem $t_1 = t_2$ by 
+\begin{inlinelist}
+  \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
+  \ml
+  \item finding a valid substitution for $t_1'$ and $t_2'$ 
+  \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
+  is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
+\end{inlinelist}.
+
+% Let upper-case letter unification variables and lower case variables defined
+% constants of the object language. We take as example the following two
+% unification problems in our \ol:
+
+% \begin{center}
+%   \coqIn{(fun x => P x) = P}
+% \end{center}
+
+% The first problem, if
+
+% \begin{center}
+%   \elpiIn{p (lam x\ (app F x)) :- ... F ...}
+% \end{center}
+
+% cannot unify with goals like
+
+% \begin{center}
+%   \elpiIn{p (lam x\ app (app (const "f") x) x).}
+% \end{center}
+
+% Since \elpiIn{F} does not see \elpiIn{x} even though it is applied to \elpiIn{x}
+% in the object language.
+
+----------------------------- END -----------------------------
 \subsection{First-order unification}
@@
 
+\appendix
+
+\begin{table}
+  \centering
+  \begin{tabular}{c||c|c}
+  Term        & \ol                    & \ml                                   \\
+  \hline\hline
+  Constant    & a                             & `a'                                     \\
+  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
+  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
+  Variable    & X                             & X                                             \\
+  \end{tabular}
+  \caption{\ol terms to \ml terms representation}
+  \label{table:coq2elpi}
+\end{table}
+
+\section{TC in \coq}
+For instance, if
+\texttt{XXX} is the type class representing the AAA, then ZZZ and WWW are
+instances for XXX. In the code snippet below, we give such implementation in
+\coq.
+
+\begin{mintedTcolorbox}[escapeinside=&&]{coq}
+  Inductive sig (A : Type) (P : A -> Prop) : Type := ...
+
+  Class Decision (P : Prop) := decide : {P} + {not P}.
+  Class RelDecision {A B: Type} (R : A -> B -> Prop).
+  Class ProofIrrel (A : Type) : Prop := proof_irrel (x y : A) : x = y.
+
+  Instance decide_rel: forall (A B : Type) (R : A -> B -> Prop),
+    RelDecision R -> forall (x : A) (y : B), Decision (R x y). Admitted.
+  Instance True_pi : ProofIrrel True. Admitted.
+  Instance sig_eq_dec: forall (A : Type) (P : A -> Prop),
+    (forall x, ProofIrrel (P x)) -> RelDecision (@eq A) ->
+    RelDecision (@eq (sig A P)). Admitted.
+\end{mintedTcolorbox}
+
+This small set of instances after a first phase of\todo{explain compilation of
+pred and inst?} compilation is translated into the following \elpi rules:
+
+\begin{mintedTcolorbox}{elpi}
+  type tc-Decision term -> term -> o.
+  type tc-RelDecision term -> term -> term -> term -> o.
+  type tc-ProofIrrel term -> term -> o.
+
+  tc-ProofIrrel (`True`) (`True_pi`).
+  tc-Decision (app [R, X, Y]) 
+    (app [`decide_rel`, A, B, R, P, X, Y]) :-
+    tc-RelDecision A B R P.
+  tc-RelDecision (app [`sig`, A, P]) 
+    (app [`sig`, A, P]) 
+    (app [`eq`, app [`sig`, A, P]]) 
+    (app [`sig_eq_dec`, A, P, P1, P2]) :-
+    pi-decl c0 `x` A =>
+      tc-ProofIrrel (app [P, c0]) (app [P1, c0]), 
+    tc-RelDecision A A (app [`eq`, A]) P2.
+\end{mintedTcolorbox}
+
+In this paper we do not really want to explain how the translation of the
+class/instances is performed in our \ml, we prefer to focus our attention on
+unification of terms of the \ol in our \ml. Although, in \cref{table:coq2elpi},
+we provide a simple subset of the typing system used to represent the term of
+the \ol in the \ml.
+
+Type-class resolution starts from a query, that is a class applied to some
+arguments. This \coq term is translated into a term of the \ml and the search
+for a solution in the database is started. However, it may happen that the term
+representation in the \ml may hide some unification properties that are true in
+the \ol. In the example above, the goal \texttt{Decision (@eq T a b)} for some
+\texttt{a} and \texttt{b} unifies with \texttt{Decision (R x y)} in the \ol but
+not in its meta representation. Similarly, the goal \texttt{RelDecision (@eq
+(sig T ?P))} where \texttt{?P}, under the hypothesis \texttt{RelDecision (@eq
+nat)}, will trying to apply the rule for \texttt{sig\_eq\_dec}, we fall into an
+higher order unification problem, where \texttt{P} is applied to the local name
+\texttt{x}. However, the corresponding rule in the \ml exploit a first order
+variable \texttt{P}. Therefore, after the refinement of the goal to
+\texttt{sig\_eq\_dec}, the resolution immediately fail to solve the premise
+\texttt{tc-ProofIrrel (app [P, c0]) (app [P1, c0])}.
+
 \end{document}

### 2024-02-12 main.tex
Tex update

@@
 
-% Let upper-case letter unification variables and lower case variables defined
-% constants of the object language. We take as example the following two
-% unification problems in our \ol:
+\section{Term compilation}
 
-% \begin{center}
-%   \coqIn{(fun x => P x) = P}
-% \end{center}
+In order to present the the compilation of the \ol terms so that higher order
+unification can be performed, we need a second and mor powerful representation
+of the \ol terms so that variables have a scope. This specification is shown
+in the code snippet below.
 
-% The first problem, if
+\begin{mintedTcolorbox}{elpi}
+  kind ml.tm type.
+  type ml.app list ml.tm -> ml.tm.
+  type ml.lam (ml.tm -> ml.tm) -> ml.tm.
+  type ml.c string -> ml.tm.
+  type ml.uv  nat -> list ml.tm -> ml.tm.
+\end{mintedTcolorbox}
 
-% \begin{center}
-%   \elpiIn{p (lam x\ (app F x)) :- ... F ...}
-% \end{center}
+\noindent In particular a \elpiIn{ml.uv} term is meant as a unification variable
+of the meta-language. Therefore, the unification between \elpiIn{ml.lam x\
+ml.lam y\ ml.uv 1 [x, y]} and \elpiIn{ml.lam x\ ml.lam y\ ml.uv 0 [x]} is
+supposed to be unified by the internal unification procedure of the \ml (for
+this example, the substitution for \elpiIn{uv 0} is \elpiIn{ml.lam x\ ml.lam y\
+ml.uv 2 [x]} whereas the substitution for \elpiIn{uv 1} is \elpiIn{ml.lam x\
+ml.uv 2 [x]}).
 
-% cannot unify with goals like
+Moreover, if \elpiIn{ml.uv} stands for meta-variables, the \elpiIn{app} and the
+\elpiIn{lam} constructors are the nodes for the terms of the \ol. Therefore, we
+cannot claim that \elpiIn{ml.lam x\ ml.app [ml.c "f", x]} and \elpiIn{ml.c "f"}
+unify, since, even though the first is the $\eta$-expansion of the second, the
+\ml does not know how to $\eta\beta$-reduce the terms of the \ol.
 
-% \begin{center}
-%   \elpiIn{p (lam x\ app (app (const "f") x) x).}
-% \end{center}
+The compilation phase is quite straightforward, each constructor of type
+\elpiIn{tm} is mapped to its corresponding version of type \elpiIn{ml.tm}. A
+slight different approach is taken in the case of terms of the form \elpiIn{app
+[uv N | L]}, where the term is translated into \elpiIn{tm.un M L}, that is, a
+new meta-variable \elpiIn{M} with scope \elpiIn{L}.
 
-% Since \elpiIn{F} does not see \elpiIn{x} even though it is applied to \elpiIn{x}
-% in the object language.
+This latter term transformation is untying the original variable \elpiIn{N} of
+the \ol from the compiled term in the \ml. This means that when \elpiIn{M} is
+instantiated into the \ml, we need to transfer the substitution to the \ol. In
+order to bridge instantiation of meta-variables with the \ol variables, an ad
+hoc \texttt{link} is crafted between the two variables.
 

### 2024-02-12 main.tex
update tex

@@
 \noindent In particular a \elpiIn{ml.uv} term is meant as a unification variable
-of the meta-language. Therefore, the unification between \elpiIn{ml.lam x\
-ml.lam y\ ml.uv 1 [x, y]} and \elpiIn{ml.lam x\ ml.lam y\ ml.uv 0 [x]} is
-supposed to be unified by the internal unification procedure of the \ml (for
-this example, the substitution for \elpiIn{uv 0} is \elpiIn{ml.lam x\ ml.lam y\
-ml.uv 2 [x]} whereas the substitution for \elpiIn{uv 1} is \elpiIn{ml.lam x\
-ml.uv 2 [x]}).
+of the meta-language. Therefore, the unification between
+
+\begin{equation*}
+  \texttt{\elpiIn{ml.lam x\ ml.lam y\ ml.uv 1 [x, y]}}
+\end{equation*}
+
+and
+
+\begin{equation*}
+  \texttt{\elpiIn{ml.lam x\ ml.lam y\ ml.uv 0 [x]}}
+\end{equation*}
+
+is supposed to procedure of the substitution \elpiIn{ml.lam x\ ml.lam y\ ml.uv 2
+[x]} for \elpiIn{uv 0} and the substitution \elpiIn{ml.lam x\ ml.uv 2 [x]}) for
+\elpiIn{uv 1}.
 

### 2024-02-13 main.tex
update tex

@@
 
-\begin{equation}
-  \texttt{\elpiIn{app[c "p", app[c "f", uv 0], lam x\ app[c "g", x, app[uv 1, x]]].}}
-  \label{eq:elpi1}
-\end{equation}
+\elpiInEq[eq:elpi1]{app[c "p", app[c "f", uv 0], lam x\ app[c "g", x, app[uv 1, x]]].}
 
@@
   \texttt{\elpiIn{app[c "p", app[c "f", c "a"], lam x\ app[c "g", x, c "b"]].}}
+  \label{eq:elpi2}
 \end{equation}
@@
 
-In order to present the the compilation of the \ol terms so that higher order
-unification can be performed, we need a second and mor powerful representation
+In order to present the the compilation of the \ol terms, so that higher order
+unification can be performed, we need a second and more powerful representation
 of the \ol terms so that variables have a scope. This specification is shown
@@
 
-\begin{equation*}
-  \texttt{\elpiIn{ml.lam x\ ml.lam y\ ml.uv 1 [x, y]}}
-\end{equation*}
+\elpiInEqq{ml.lam x\ ml.lam y\ ml.uv 1 [x, y]}
 
@@
 
-\begin{equation*}
-  \texttt{\elpiIn{ml.lam x\ ml.lam y\ ml.uv 0 [x]}}
-\end{equation*}
+\elpiInEqq{ml.lam x\ ml.lam y\ ml.uv 0 [x]}
 
@@
 unify, since, even though the first is the $\eta$-expansion of the second, the
-\ml does not know how to $\eta\beta$-reduce the terms of the \ol.
+\ml does not know how to $\eta\beta$-reduce terms of the \ol.
+
+In our encoding, we explicitly encode the meta-variables with the \elpiIn{ml.uv}
+constructor. This is because we prefer to have the full control of the \ml,
+including the meta-variables instantiation. This way we are able to concretely
+touch the substitution performed by the \ml. In a further section, we show that
+there is no difference between our custom \ml language and any other \ml. Of
+course, a full control on the unification behind meta-variable assignment ask to
+drag the substitution mapping of the \ml and update it each time a variable is
+refined.
 
@@
 hoc \texttt{link} is crafted between the two variables.
+
+A link, \elpiIn{type link nat -> nat -> link}, takes two integers: the first 
+stands for the index of variables in the \ol and the second is the index of 
+the meta-variables.
+
+For example, if we take back the example in \cref{eq:elpi2}, and want to compile
+it, we obtain the new term:
+
+\elpiInEq{app[c "p", app[c "f", c "a"], lam x\ app[c "g", x, c "b"]].}
 

### 2024-02-13 main.tex

@@
 \newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
-
 \begin{abstract}
@@
   \end{abstract}
+
+\section{intro}
+
+we are interested in using a meta language in LF style to write
+automation, proof search. In various works... they achieve that
+for a OL which is simpler than the LF, the equational theory is included
+in the one of the ML. This is exploited to piggy back on the unif
+of the ML.
+the peculiarity of our setting is that the OL has a richer equational
+theory of the ML, eg beta eta zeta bla bla. Moreover is HO logic, so
+quantifies over functions, so unif variable range on that too.
+We want to piggy back on the ML unif whenever the problem fits
+in its domain, eg pattern fragment.
+this is important for practical purposes.
+
+\subsection{in a nutshell}
+example, a rule for theorem
+$$
+forall f : A -> B, .. bla f .. -> prove (\forall x, f x)
+bla g
+$$
+\begin{minted}{elpi}
+type app ..
+prove (forall x\ app F x) :- ... bla F ...
+\end{minted}
+on goal \mint{elpi}|prove (forall x\ app (app g x) x))|
+would fail since \mint{elpi}|F != app g x|.
+
+of course one wants to avoid
+\begin{minted}{elpi}
+prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assignment "F" S' F, bla F S' ...
+\end{minted}
+
+Now, ML has HO variables
+\begin{minted}{elpi}
+type lam ..
+prove (forall x\ F x) :- ... bla F ...
+\end{minted}
+this time  \mint{elpi}|F x != app (app g x) x| has solution
+\mint{elpi}|F = a\ app (app g x) x| but F is not a term so
+bla needs to be adapted,
+\begin{minted}{elpi}
+prove (forall x\ F x) :- ... bla (lam F) ...
+bla g.
+\end{minted}
+this is too simplistic since \mint{elpi}|g != lam x\ app g x|
+
+\subsection{contribution}
+\begin{itemize}
+\item prover for HO OL in ML that uses unif
+\item eta beta
+\item test on stdpp and TLC
+\end{itemize}
 

### 2024-02-13 main.tex
structure

@@
 
-\section{From the object to the meta language}
+esempietto in cui l'HO di elpi non risolve un problem HO del linguaggio
+oggetto FO. Equazione XX
+
+\begin{coqcode}
+(* HO unif *)
+Instance forall_dec A (P: A |$\to$| Prop) `{Finite A} `{|$\forall$|x, Decision (P x)} :
+  Decision (|$\forall$|x, P x).
+Instance and_dec P `{Decision P} Q `{Decision Q}:  Decision (P ∧ Q).
+
+(* FO approx *)
+Instance decide_rel A B (R: A |$\to$| B |$\to$| Prop) `{RelDecision R} :
+  forall x y, Decision (R x y).
+Instance decide_eq_nat : RelDecision (@eq nat).
+Check _ : Decision (@eq nat 2 3)
+
+Instance _ : Inj add.
+             Inj F -> Inj (fun x => F (G x))
+
+             (fun x => add x)
+
+  Decision (|$\forall$|x, P x).
+
+\end{coqcode}
+
+\subsection{Related work and alternative approaches}
+
+LF e PL with binders (sempre HOAS).
+
+citare FO logic et similia fatti in LP, twelf. qui l'unification dell'OL
+e' facile. cercare se in twelf hanno fatto un po' di ordine superiore.
+
+isabelle fa la stessa cosa, OL e HOL, che matcha il sistema di tipi.
+qui l'unificazione e' la stessa, e fanno un ecoding shallow (senza app).
+
+Noi abbiamo un OL piu complicato, i tipi non matchano, serve un nodo app
+per due ragioni: meta programmazione e arieta variabile
+
+(equazioni XX senza app e lam) XXX IDEA XXX
+
+\section{main idea: FO encoding - HO encoding}
+
+\begin{elpicode}
+  kind fo_tm type.
+  type fo_app list fo_tm -> fo_tm.
+  type fo_lam (tm -> fo_tm) -> fo_tm.
+  type fo_uv  nat -> fo_tm.
+  type fo_c string -> fo_tm.
+\end{elpicode}
+
+\begin{elpicode}
+  kind tm type.
+type app list tm -> tm.
+type lam (tm -> tm) -> tm.
+type uv  nat -> list tm -> tm.
+type c string -> tm.
+\end{elpicode}
+
+le due sintassi, XX tradotto, unif passa, bisogna riportare la soluzione
+nel mondo fo
+
+segnatura di decomp e unif, equalfo (fa beta, eta, deref)
+
+proprieta desiderate, serve equalho (fa deref)
+
+\subsection{implementation}
+
+l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
+in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
+elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
+ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.
+
+\begin{minted}{elpi}
+prove (app[c "decidable", all x\ P x]) Proof :- ho-link P P', prove (... P' ...).
+\end{minted}
+
+qui P nel paper diventa uv N [x] per un certo N.
+
+\section{recovering HO}
+
+\begin{elpicode}
+  p (all x\ app[F,x]) = p (all x\ app[f,x,x])
+\end{elpicode}
+
+fallisce perche le liste non hanno la stessa lunghezza.
+
+compile
+
+\begin{elpicode}
+  p (all x\ F' x) = p (all x\ app[f,x,x]) , link F F'.
+  F' = x\ app[f,x,x]
+  F = lam a\ app[f,a,a].
+\end{elpicode}
+
+\begin{elpicode}
+  type comp fo.tm -> ho.tm -> list link -> list link -> ho.subst -> ho.subst -> o.
+  comp (fo.c X)   (ho.c X) L L S S.
+  comp (fo.app [fo.uv N|Argsss]) TT L L3 S S3 :- %!,
+    % TODO: here split-pf to enter dist. names into N
+    % split-pf Argsss [] PF NPF,
+    split-pf Argsss [] PF [], NPF = [], % TODO: compile to (uv N L) + link-fo-app
+    print "In PF" PF NPF, 
+    if (NPF = []) (TT = ho.uv M PF1) (TT = ho.app [ho.uv M PF1 | NPF1]),
+    % pattern-fragment Args,
+    fold4 comp PF PF1 L L1 S S1,
+    fold4 comp NPF NPF1 L1 L2 S1 S2,
+    ho.new S2 M S3,
+    % TODO: maybe len can be given by split-pf
+    len PF Len,
+    L3 = [link N M Len| L2].
+  % TODO: if don't want to modify unif, we compile `fo.app [fo.c f, c0, c0, ho.c a]` into 
+  % `ho.app [ho.app[f, c0, c0], ho.c a]
+  comp (fo.app A) (ho.app A1) L L1 S S1 :- fold4 comp A A1 L L1 S S1.
+  comp (fo.lam F) (ho.lam F1) L L1 S S1 :- 
+    (pi x y\ (pi A S\ comp x y A A S S) => comp (F x) (F1 y) L L1 S S1).
+  comp (fo.uv N) (ho.uv M []) L [link N M z|L] S S1 :- ho.new S M S1.
+
+  decmop...
+\end{elpicode}
+
+\section{recovering eta}
+
+\begin{elpicode}
+q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
+F = fun a => app [f,a] ----> F = f
+\end{elpicode}
+
+l'utene da p su f, mentre l'istanza pe q forza F a fun ..
+
+\section{recovering beta}
+
+\begin{elpicode}
+  q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
+  F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
+\end{elpicode}
+
+qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
+p1 F1, e decomp beta F [a] F1.
+
+\section{recovering eta-beta within unification (non linear variables)}
+
+se i problemi di cui sopra avvengono nello stesso termine
+
+\begin{elpicode}
+  q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
+\end{elpicode}
+
+bisogna slegare le due F e poi unificare le soluzioni tra di loro
+
+\section{recoving binary app}
+
+fo approx / sub pattern fragment
+
+\begin{elpicode}
+  p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
+  p (all x\G x) F' = 
+  G = x\ f x x a
+  F = lam x\f x x
+  F' = (app[f,b,b])
+  link (F a) F'
+  link G F
+\end{elpicode}
+
+questo esempio recupera l'app binaria perche
+
+\begin{elpicode}
+  (app (app F x) a) = (app (f x x) a)
+\end{elpicode}
+
+\section{XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 

### 2024-02-14 main.tex
tex

@@
 
-\begin{coqcode}
+\begin{minted}[escapeinside=||,mathescape=true]{coq}
 (* HO unif *)
@@
   Decision (|$\forall$|x, P x).
-Instance and_dec P `{Decision P} Q `{Decision Q}:  Decision (P ∧ Q).
+Instance and_dec P `{Decision P} Q `{Decision Q}:  Decision (P /\ Q).
 
@@
 
-\end{coqcode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
   kind fo_tm type.
@@
   type fo_c string -> fo_tm.
-\end{elpicode}
+\end{minted}
 
-\begin{elpicode}
+\begin{minted}{elpi}
   kind tm type.
@@
 type c string -> tm.
-\end{elpicode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
   p (all x\ app[F,x]) = p (all x\ app[f,x,x])
-\end{elpicode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
   p (all x\ F' x) = p (all x\ app[f,x,x]) , link F F'.
@@
   F = lam a\ app[f,a,a].
-\end{elpicode}
+\end{minted}
 
-\begin{elpicode}
+\begin{minted}{elpi}
   type comp fo.tm -> ho.tm -> list link -> list link -> ho.subst -> ho.subst -> o.
@@
   decmop...
-\end{elpicode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
 q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
 F = fun a => app [f,a] ----> F = f
-\end{elpicode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
   q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
   F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
-\end{elpicode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
   q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
-\end{elpicode}
+\end{minted}
 
@@
 
-\begin{elpicode}
+\begin{minted}{elpi}
   p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
@@
   link G F
-\end{elpicode}
+\end{minted}
 
-questo esempio recupera l'app binaria perche
+% questo esempio recupera l'app binaria perche
 
-\begin{elpicode}
+\begin{minted}{elpi}
   (app (app F x) a) = (app (f x x) a)
-\end{elpicode}
+\end{minted}
 

### 2024-02-14 main.tex
coqcode and elpicode macro for minted

--- 2024-02-13 main.tex
@@
 $$
-\begin{minted}{elpi}
-type app ..
+\begin{elpicode}
+type app ...
 prove (forall x\ app F x) :- ... bla F ...
-\end{minted}
+\end{elpicode}
 on goal \mint{elpi}|prove (forall x\ app (app g x) x))|
@@
 of course one wants to avoid
-\begin{minted}{elpi}
+\begin{elpicode}
 prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assignment "F" S' F, bla F S' ...
-\end{minted}
+\end{elpicode}
 
 Now, ML has HO variables
-\begin{minted}{elpi}
+\begin{elpicode}
 type lam ..
 prove (forall x\ F x) :- ... bla F ...
-\end{minted}
+\end{elpicode}
 this time  \mint{elpi}|F x != app (app g x) x| has solution
@@
 bla needs to be adapted,
-\begin{minted}{elpi}
+\begin{elpicode}
 prove (forall x\ F x) :- ... bla (lam F) ...
 bla g.
-\end{minted}
+\end{elpicode}
 this is too simplistic since \mint{elpi}|g != lam x\ app g x|
@@
 (* HO unif *)
-Instance forall_dec A (P: A |$\to$| Prop) `{Finite A} `{|$\forall$|x, Decision (P x)} :
-  Decision (|$\forall$|x, P x).
-Instance and_dec P `{Decision P} Q `{Decision Q}:  Decision (P ∧ Q).
+Instance forall_dec A (P: A ~$\to$~ Prop) `{Finite A} `{~$\forall$~x, Decision (P x)} :
+  Decision (~$\forall$~x, P x).
+Instance and_dec P `{Decision P} Q `{Decision Q}:  Decision (P /\ Q).
 
 (* FO approx *)
-Instance decide_rel A B (R: A |$\to$| B |$\to$| Prop) `{RelDecision R} :
+Instance decide_rel A B (R: A ~$\to$~ B ~$\to$~ Prop) `{RelDecision R} :
   forall x y, Decision (R x y).
@@
 
-  Decision (|$\forall$|x, P x).
+  Decision (~$\forall$~x, P x).
 
@@
 
-\begin{minted}{elpi}
+\begin{elpicode}
 prove (app[c "decidable", all x\ P x]) Proof :- ho-link P P', prove (... P' ...).
-\end{minted}
+\end{elpicode}
 
@@
 
-questo esempio recupera l'app binaria perche
+% questo esempio recupera l'app binaria perche
 

### 2024-02-14 main.tex
wip

--- 2024-02-14 main.tex
@@
 \begin{minted}{elpi}
-prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assignment "F" S' F, bla F S' ...
+prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assmt "F" S' F, bla F S' ...
 \end{minted}
@@
 
+\begin{tabular}{c|c}
+\begin{minipage}{.5\textwidth}
 \begin{minted}{elpi}
@@
 type fo_c string -> fo_tm.
+
+typeabbrev subst list (option tm).
+type equal subst -> tm -> tm -> o.
 \end{minted}
-
+\end{minipage}
+&
+\begin{minipage}{.5\textwidth}
 \begin{minted}{elpi}
@@
 type c string -> tm.
+
+kind assmt type.
+type abs (tm -> assmt) -> assmt.
+type val tm -> assmt.
+
+typeabbrev subst list (option assmt).
+type equal subst -> tm -> tm -> o.
 \end{minted}
+\end{minipage}
+\end{tabular}
 
@@
 
+this should hold in both encoding
+%abella
+\begin{minted}{elpi}
+forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
+\end{minted}
+we have unif only in the HO one, where equal is
+weaker (but unif is stronger)
+
 segnatura di decomp e unif, equalfo (fa beta, eta, deref)
 
-proprieta desiderate, serve equalho (fa deref)
+\begin{tabular}{c}
+\begin{minipage}{.99\textwidth}
+\begin{minted}{elpi}
+% deref
+equal S (uv N) T1 :- assigned? N S T, equal S T T1.
+equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
+equal S (app [uv N|Args]) T1 :- assigned? N S F, beta F Args T, equal S T T1.
+equal S T1 (app [uv N|Args]) :- assigned? N S F, beta F Args T, equal S T1 T.
+% congruence
+equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
+equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
+equal _ (c X) (c X).
+equal _ (uv N) (uv N).
+% beta
+equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
+equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
+% eta
+equal S (lam F) T :- not (T = lam _),
+  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
+equal S T (lam F) :- not (T = lam _),
+  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
+\end{minted}
+\end{minipage}
+\\\hline
+\begin{minipage}{.99\textwidth}
+\begin{minted}{elpi}
+% deref
+equal S (uv N Args) T1 :- assigned? N S F, beta-abs F Args T, equal S T T1.
+equal S T1 (uv N Args) :- assigned? N S F, beta-abs F Args T, equal S T1 T.
+% congruence
+equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
+equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
+equal _ (c X) (c X).
+equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
+\end{minted}
+\end{minipage}
+
+\end{tabular}
+
+equal fo is congruence, deref, beta, eta (for the OL).
+equal ho is congruence, deref (for the OL).
+unif (fo or ho) should cover pattern fragment, i.e. well behaved unif.
+
+ho, and unif ho, is the LF/ML language/unif. here we describe fo and ho
+with a deeply embedded syntax, but the idea is that uv can be mapped to
+an actual unif variable.
 
@@
 there is no difference between our custom \ml language and any other \ml. Of
-course, a full control on the unification behind meta-variable assignment ask to
+course, a full control on the unification behind meta-variable assmt ask to
 drag the substitution mapping of the \ml and update it each time a variable is

### 2024-02-14 main.tex

--- 2024-02-14 main.tex
@@
 
-\begin{elpicode}
+\begin{tabular}{c|c}
+\begin{minipage}{.5\textwidth}
+\begin{minted}{elpi}
 kind fo_tm type.
@@
 type fo_c string -> fo_tm.
-\end{elpicode}
 
-\begin{elpicode}
+typeabbrev subst list (option tm).
+type equal subst -> tm -> tm -> o.
+\end{minted}
+\end{minipage}
+&
+\begin{minipage}{.5\textwidth}
+\begin{minted}{elpi}
 kind tm type.
@@
 type c string -> tm.
-\end{elpicode}
+
+kind assmt type.
+type abs (tm -> assmt) -> assmt.
+type val tm -> assmt.
+
+typeabbrev subst list (option assmt).
+type equal subst -> tm -> tm -> o.
+\end{minted}
+\end{minipage}
+\end{tabular}
 
@@
 
+this should hold in both encoding
+%abella
+\begin{minted}{elpi}
+forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
+\end{minted}
+we have unif only in the HO one, where equal is
+weaker (but unif is stronger)
+
 segnatura di decomp e unif, equalfo (fa beta, eta, deref)
 
-proprieta desiderate, serve equalho (fa deref)
+\begin{tabular}{c}
+\begin{minipage}{.9\textwidth}
+\begin{minted}{elpi}
+% deref
+equal S (uv N) T1 :- assigned? N S T, equal S T T1.
+equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
+equal S (app [uv N|Args]) T1 :- assigned? N S F, beta F Args T, equal S T T1.
+equal S T1 (app [uv N|Args]) :- assigned? N S F, beta F Args T, equal S T1 T.
+% congruence
+equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
+equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
+equal _ (c X) (c X).
+equal _ (uv N) (uv N).
+% beta
+equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
+equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
+% eta
+equal S (lam F) T :- not (T = lam _),
+  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
+equal S T (lam F) :- not (T = lam _),
+  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
+\end{minted}
+\end{minipage}
+\\\hline
+\begin{minipage}{.9\textwidth}
+\begin{minted}{elpi}
+% deref
+equal S (uv N Args) T1 :- assigned? N S F, beta-abs F Args T, equal S T T1.
+equal S T1 (uv N Args) :- assigned? N S F, beta-abs F Args T, equal S T1 T.
+% congruence
+equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
+equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
+equal _ (c X) (c X).
+equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
+\end{minted}
+\end{minipage}
+
+\end{tabular}
+
+equal fo is congruence, deref, beta, eta (for the OL).
+equal ho is congruence, deref (for the OL).
+unif (fo or ho) should cover pattern fragment, i.e. well behaved unif.
+
+ho, and unif ho, is the LF/ML language/unif. here we describe fo and ho
+with a deeply embedded syntax, but the idea is that uv can be mapped to
+an actual unif variable.
+
+of course we could write a unif for the ho that had beta and eta,
+but if we use the one of the LF it can't have eta beta of the OL baked in.
 

### 2024-02-15 main.tex
wip
Update tex

@@
 
+\newcommand{\library}[1]{\textit{#1}\xspace}
+\def\stdpp{\library{stdpp}}
+\def\iris{\library{iris}}
+\tableofcontents
+
 \newcommand*{\acronym}[1]{\texttt{#1}\xspace}
 
-\def\ol{\acronym{ol}}
-\def\ml{\acronym{ml}}
+\def\ol{\acronym{ol}} % object language
+\def\ml{\acronym{ml}} % meta language
+\def\lf{\acronym{lf}} % logical framework
 
@@
 
-  Unification is a fundamental process behind goal resolution in logic
-  programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
-  exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
-  substitution is a mapping from unification variables to terms and it is the most
-  general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
-  t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
-  Unification has been widely studied in the 90s [here the refs]. In this paper we
-  want to give a framework to reason about unification in meta-programming.
+  % Unification is a fundamental process behind goal resolution in logic
+  % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
+  % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
+  % substitution is a mapping from unification variables to terms and it is the most
+  % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
+  % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
+  % Unification has been widely studied in the 90s [here the refs]. In this paper we
+  % want to give a framework to reason about unification in meta-programming.
+  Meta-programming \cite{hill1994} is a programming technique in which a
+  program, referred to as meta language (\ml), can treat an other program, knwon
+  as the object language (\ol) as its data. This programming approach is used in
+  various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
+  capable of automating proof resolution. In many of the previously the
+  previously mentioned works, the equational theory expressed by the \ol is
+  weaker than the \ml one. Consequently, no effort is required to adjust the
+  unification algorithm of the \ml since it encompass the \ml one. In our
+  specific scenario, however, we work with an \ol with a richer equational
+  theory than our \ml. This paper introduces a framework that enable the \ml to
+  comprehend the unification of the \ol and provides some practical applications
+  of these findings. Additionally, we extend our work to unification in the case
+  of higher-order variable quantification.
 
 \end{abstract}
+
+\section{Introduction}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Context
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% meta programming
+Meta programming is a powerful programming style that enable the incorporation
+of an object language (\ol) into a meta language (\ml). This control over the
+\ol lies at the core of meta-programming, facilitating a spectrum of operations
+on the object language. Some of the principal program manipulations range
+from interpretation to compilation.
+
+% theorem prover and tc
+Thanks to meta programming, it is feasible to express the equational theory of a
+theorem prover, as noted in \cite{citeX, Y Z}, with the \ml serving as the
+foundational infrastructure for this ecosystem. Proving a theorem, that is
+solving a query using the jargon of logic programming consist in verifying if a
+proposition is a logical consequence of a given set of assumption. The benefit
+of of using a prologish verification style is that the proof resolution can be
+automatized thanks to different search strategies, such as a depth first search
+with backtracking. One of the main expression of this proof automation is
+type-class resolution\cite{wadler89, sozeau08}. Type classes are a typing
+structure to introduce ad-hoc polymorphism in functional programming languages.
+In \coq, the use of type class has become a more and more used programming
+style. Several are the libraries based on this machinery, we can site \stdpp and
+\iris.
+
+% embed tc in ml
+One of the
+
+\begin{itemize}
+  \item about meta programming (what it is, when to use it)
+  \item theorem provers (proof automation)
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Example
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{itemize}
+  \item TODO: context
+  \item TODO: example
+  \item TODO: riassunto delle next sections
+\end{itemize}
 
@@
 terms of the \ol in the \ml so that a wide set of program
-manipulations ranging from interpretation to compilation. Furthermore, since a
-logic program is a set of clauses, the behavior of the \ol can be
-modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.
+manipulations ranging from interpretation to compilation. 
+% Furthermore, since a
+% logic program is a set of clauses, the behavior of the \ol can be
+% modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.
 
@@
 
+\noindent
 \begin{tabular}{c}
-\begin{minipage}{.9\textwidth}
+\begin{minipage}{.95\textwidth}
 \begin{minted}{elpi}
@@
 equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
-equal S (app [uv N|Args]) T1 :- assigned? N S F, beta F Args T, equal S T T1.
-equal S T1 (app [uv N|Args]) :- assigned? N S F, beta F Args T, equal S T1 T.
+equal S (app [uv N|A]) T1 :- assigned? N S F, beta F A T, equal S T T1.
+equal S T1 (app [uv N|A]) :- assigned? N S F, beta F A T, equal S T1 T.
 % congruence
@@
 \\\hline
-\begin{minipage}{.9\textwidth}
+\begin{minipage}{.95\textwidth}
 \begin{minted}{elpi}
 % deref
-equal S (uv N Args) T1 :- assigned? N S F, beta-abs F Args T, equal S T T1.
-equal S T1 (uv N Args) :- assigned? N S F, beta-abs F Args T, equal S T1 T.
+equal S (uv N A) T1 :- assigned? N S F, beta-abs F A T, equal S T T1.
+equal S T1 (uv N A) :- assigned? N S F, beta-abs F A T, equal S T1 T.
 % congruence
@@
 
-\section{recoving binary app}
+\section{heuristic / binary app}
 

### 2024-02-15 main.tex
update tex

@@
 Thanks to meta programming, it is feasible to express the equational theory of a
-theorem prover, as noted in \cite{citeX, Y Z}, with the \ml serving as the
-foundational infrastructure for this ecosystem. Proving a theorem, that is
+theorem prover, as indicated in \cite{citeX, Y Z}, with the \ml serving as the
+foundational infrastructure for this ecosystem. The process of proving a theorem, that is
 solving a query using the jargon of logic programming consist in verifying if a
@@
 automatized thanks to different search strategies, such as a depth first search
-with backtracking. One of the main expression of this proof automation is
-type-class resolution\cite{wadler89, sozeau08}. Type classes are a typing
-structure to introduce ad-hoc polymorphism in functional programming languages.
-In \coq, the use of type class has become a more and more used programming
-style. Several are the libraries based on this machinery, we can site \stdpp and
-\iris.
+with backtracking. One prominent manifestation of this automated proof process is
+type-class resolution\cite{wadler89, sozeau08}. Type classes serve as a typing
+structure to introduce ad-hoc polymorphism in functional languages.
+In \coq, the use of type class has increasingly become a programming
+style. Numerous libraries, such as \stdpp and
+\iris are built upon this machinery.
 
 % embed tc in ml
-One of the
+The motivating example for this article is our investigation for an alternative
+type-class solver for \coq in \elpi\todo{citation to elpi} through the \coqelpi
+plugin. This plugin gives a deep embedding of \coq terms into \elpi syntax, but
+challenges arise whe attempting to solve type-class goals represented into their
+corresponding \elpi representation.
 

### 2024-02-15 main.tex
update tex

@@
 \def\iris{\library{iris}}
-\tableofcontents
+% \tableofcontents
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Context
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % meta programming
@@
 
-% embed tc in ml
+% tc motivates this paper
 The motivating example for this article is our investigation for an alternative
@@
 
-\begin{itemize}
-  \item about meta programming (what it is, when to use it)
-  \item theorem provers (proof automation)
-\end{itemize}
+% syntax of coq terms in elpi
+In particular, \coq terms are encoded following the following (very simplified)
+data type:
+
+\begin{elpicode}
+  kind term type.
+  type const gref -> term.
+  type app list term -> term.
+  type fun (term -> term) -> term.
+  type prod (term -> term) -> term.
+\end{elpicode}
+
+Where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
+identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
+is embedded in the \elpiIn{app} node, having the head and the arguments of the
+application expressed as a list of terms. Lambda abstractions are translated
+into the node \elpiIn{fun} binding a term to another term. Finally forall
+quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{forall x\ F}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Example
+% Examples
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Let's take the following example:
+\begin{tcolorboxSimple}
+  \begin{tabular}{c|c}
+    \begin{minipage}[c]{0.50\linewidth}
+      \begin{coqcode}
+        Instance forall_dec A (P: A ~$\to$~ Prop):
+          Finite A ~$\to$~ 
+          ~$\forall$~x, Decision (P x) ~$\to$~
+          Decision (~$\forall$~x, P x).
+      \end{coqcode}
+    \end{minipage}
+    &
+    \begin{minipage}[c]{0.46\linewidth}
+      \begin{elpicode}
+        tc-Decision (prod x\ app[P, x]) :- 
+          % premise for Finite A
+          % premise for $\forall$x, Decision (P x)
+      \end{elpicode}
+    \end{minipage}
+  \end{tabular}
+\end{tcolorboxSimple}
+
+The \coq instance on the left is compiled in the \elpi rule on the right. If 
+we try to solve the following query (already in the \elpi syntax)
+
+\elpiIn{tc-Decision (prod x\ app[f, y, x])}
+
+where \coqIn{f : A -> B -> Type} and \coqIn{y: A}, we can notice that the \elpi
+rule above cannot be applied; the terms \elpiIn{prod x\ app[P, x]} and
+\elpiIn{prod x\ app[f, y, x]} cannot unify since the lists under the
+\elpiIn{app} constructors are of different length.\todo{NOTE: the goal unify
+with forall\_dec in coq xké fa unif FO approx}
 

### 2024-02-15 main.tex
add example elpi->coq term

@@
 In particular, \coq terms are encoded following the following (very simplified)
-data type:
+data type
 
+% \begin{elpicode}
+%   kind term type.
+%   type const gref -> term.
+%   type app list term -> term.
+%   type fun (term -> term) -> term.
+%   type prod (term -> term) -> term.
+% \end{elpicode}
+
+\begin{tcolorboxSimple}
+  \begin{tabular}{c|c}
+    \begin{minipage}[c]{0.50\linewidth}
       \begin{elpicode}
@@
       \end{elpicode}
+    \end{minipage}
+    &
+    \begin{minipage}[c]{0.46\linewidth}
+      \begin{textcode}
+        Example:
+        a        ==> const 0
+        f a      ==> app[const f, const a]
+        ~$\lambda$~x => x  ==> fun x\ x
+        ~$\forall$x, F x~  ==> prod x\ app[F, x]
+      \end{textcode}
+    \end{minipage}
+  \end{tabular}
+\end{tcolorboxSimple}
 
-Where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
+\noindent where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
 identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
@@
 
-Let's take the following example:
+Let's take the following example\footnote{taken from the \stdpp library}:
 \begin{tcolorboxSimple}
@@
 with forall\_dec in coq xké fa unif FO approx}
+
+An other example of unification failure in the
 

### 2024-02-15 main.tex
update tex

@@
 \def\lf{\acronym{lf}} % logical framework
+\def\ho{\acronym{ho}} % higher order
 
@@
     &
-    \begin{minipage}[c]{0.46\linewidth}
+    \begin{minipage}[c]{0.48\linewidth}
       \begin{textcode}
@@
         a       ==> const 0
-        f a      ==> app[const f, const a]
+        f a     ==> app[const `f, const `a]
         ~$\lambda$~x => x ==> fun x\ x
@@
 
-The \coq instance on the left is compiled in the \elpi rule on the right. If 
-we try to solve the following query (already in the \elpi syntax)
+The \coq instance on the left is compiled in the \elpi rule on the right. If the 
+query to solve (already in the \elpi syntax) is
 
-\elpiIn{tc-Decision (prod x\ app[f, y, x])}
+\elpiInEq{tc-Decision (prod x\ const `nat)}
 
-where \coqIn{f : A -> B -> Type} and \coqIn{y: A}, we can notice that the \elpi
-rule above cannot be applied; the terms \elpiIn{prod x\ app[P, x]} and
-\elpiIn{prod x\ app[f, y, x]} cannot unify since the lists under the
-\elpiIn{app} constructors are of different length.\todo{NOTE: the goal unify
-with forall\_dec in coq xké fa unif FO approx}
+\noindent it is clear that the \elpi rule above will not unify with the query,
+Since the rigid node \elpiIn{app} is different from the rigid node
+\elpiIn{const}. The problem come form the lost of the fact that \coqIn{P} is an
+higher-order variable in \coq, and in \elpi it is translated into a first order
+variable (the variable \elpiIn{P} in \elpi does not see the binder \elpiIn{x}).
 
-An other example of unification failure in the
+Another unification problem we can da notare che il problema dopo, is the FO
+approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
+bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
+vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
+l'algo di unif di coq in questo caso specifico
 
-\begin{itemize}
-  \item TODO: context
-  \item TODO: example
-  \item TODO: riassunto delle next sections
-\end{itemize}
+\elpiInEq{tc-Decision (prod x\ app[const `f, const `y, x])}
+
+Specificare anche che non riusciamo a fare ne eta ne beta riduzione.
+
+Dire che elpi unifica in HO e con eta-beta
+
+% where \coqIn{f : A -> B -> Type} and \coqIn{y: A}, we can notice that the \elpi
+% rule above cannot be applied; the terms \elpiIn{prod x\ app[P, x]} and
+% \elpiIn{prod x\ app[f, y, x]} cannot unify since the lists under the
+% \elpiIn{app} constructors are of different length.\todo{NOTE: the goal unify
+% with forall\_dec in coq xké fa unif FO approx}
+
+% Another example of unification failure is given by the the following instance
 

### 2024-02-15 main.tex
update tex

@@
 \def\ho{\acronym{ho}} % higher order
+\def\Forall{$\forall$}
 
@@
 % syntax of coq terms in elpi
-In particular, \coq terms are encoded following the following (very simplified)
+In particular, \coq terms are encoded with the following (very simplified)
 data type
@@
 
-\begin{tcolorboxSimple}
+\noindent %
 \begin{tabular}{c|c}
@@
       Example:
-        a       ==> const 0
+      a       ==> const `a
       f a     ==> app[const `f, const `a]
@@
 \end{tabular}
-\end{tcolorboxSimple}
 
@@
 into the node \elpiIn{fun} binding a term to another term. Finally forall
-quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{forall x\ F}.
+quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{prod x\ F}.
 
@@
 
-Let's take the following example\footnote{taken from the \stdpp library}:
-\begin{tcolorboxSimple}
+Let's take the following example from the \stdpp library:
+
+\noindent %
 \begin{tabular}{c|c}
@@
 \end{tabular}
-\end{tcolorboxSimple}
 
-The \coq instance on the left is compiled in the \elpi rule on the right. If the 
-query to solve (already in the \elpi syntax) is
+\noindent In our type-class solver, we compile the \coqIn{forall_dec} instance
+of \coq (on the left) into the \elpi rule on the right. This compilation, which
+is blindly built from the type of \coqIn{forall_dec}, hides some unification
+properties accepted by the unification algorithm of the \ol but that are
+rejected by the unification algorithm of the \ml. The main problem of this
+compilation is how the subterm \coqIn{~$\forall$~x, P x} is translated in
+the \ml rule; \coqIn{P} is a higher order variable, in the former subterm, it
+does see the binder \coqIn{x}. However, the rule in the \ml, consider a first
+order variable \elpiIn{P}, not seeing \elpiIn{x}. Furthermore, this subterm is
+lodged within the \elpiIn{app} constructor making unification very structural.
 
-\elpiInEq{tc-Decision (prod x\ const `nat)}
-
-\noindent it is clear that the \elpi rule above will not unify with the query,
-Since the rigid node \elpiIn{app} is different from the rigid node
-\elpiIn{const}. The problem come form the lost of the fact that \coqIn{P} is an
-higher-order variable in \coq, and in \elpi it is translated into a first order
-variable (the variable \elpiIn{P} in \elpi does not see the binder \elpiIn{x}).
-
-Another unification problem we can da notare che il problema dopo, is the FO
-approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
-bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
-vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
-l'algo di unif di coq in questo caso specifico
-
+% Esempio HO in PF
+\elpiInEq{tc-Decision (prod x\ const `nat)}%
+% Esempio FO align
 \elpiInEq{tc-Decision (prod x\ app[const `f, const `y, x])}
 
-Specificare anche che non riusciamo a fare ne eta ne beta riduzione.
+\noindent The two example of queries above, already in the syntax of the \ml,
+represent two terms that would unify in the \ol, but fail in the \ml. In
+particular, the first fails since the \ml is not able to unify \elpiIn{const
+`nat} with the term \elpiIn{app[P, x]}: they have different rigid heads.
 
-Dire che elpi unifica in HO e con eta-beta
+The second goal refers to an approximation of the unification\todo{maybe there's
+something to cite} algorithm of the \ol. Spiegare l'approssimazione. E che
+attraverso il nostro framework è possibile "spiegare" l'algorithm di unification
+usato da ogni \ol nel ml desiderato. \todo{something to do here...}
+
+Following the definition give just before, the terms \elpiIn{app[const `f, const
+`y, x]} and \elpiIn{app[P, x]} should unify with substitution $\theta := \{P
+\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
+is not possible in the \ml since, this time we have two terms with the same head
+(the \elpiIn{app} node), but the corresponding lists have different lengths.
+
+An immediate solution to solve these unification problems\todo{mettere da qualche parte anche eta e beta}
+would be to be lazy and build very general rule for each instance in the database.
+This would build a rule with the following shape:
+
+\begin{center}
+  \begin{minipage}[c]{0.46\linewidth}
+    \begin{elpicode}
+      tc-Decision X :- 
+        ol.unify X (prod x\ app[P, x]),
+        % premise for Finite A
+        % premise for $\forall$x, Decision (P x)
+    \end{elpicode}
+  \end{minipage}
+\end{center}
+
+\noindent In this second implementation, any call to the \elpiIn{tc-Decision}
+will unify with the head of the rule for \coqIn{forall_dec}. The the first
+premise will ensure that the arguments unify with the what we expect from the
+instance definition. Notice that, this time, the very unification algorithm of
+the \ol is exploited, overcoming the unification problems listed just before.
+This approach is quite appealing but it has two main disadvantages: the first
+lays on the fact that the communication between the \ol and the \ml can
+potentially be slow, this is due to the transformation of terms from one language
+to the other. \todo{here i refer to constr->elpi e elpi->constr}
+The other drawback of this latter encoding is that clause discrimination cannot
+be done on the head of the clause, for instance, any indexing 
+algorithm allowing to filter rules become suddenly useless.
+
+% Contribution
+Finally
+
+% da notare che il problema dopo, is the FO
+% approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
+% bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
+% vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
+% l'algo di unif di coq in questo caso specifico
 

### 2024-02-15 main.tex
rephrase tex

@@
 
-\noindent In our type-class solver, we compile the \coqIn{forall_dec} instance
-of \coq (on the left) into the \elpi rule on the right. This compilation, which
-is blindly built from the type of \coqIn{forall_dec}, hides some unification
-properties accepted by the unification algorithm of the \ol but that are
-rejected by the unification algorithm of the \ml. The main problem of this
-compilation is how the subterm \coqIn{~$\forall$~x, P x} is translated in
-the \ml rule; \coqIn{P} is a higher order variable, in the former subterm, it
-does see the binder \coqIn{x}. However, the rule in the \ml, consider a first
-order variable \elpiIn{P}, not seeing \elpiIn{x}. Furthermore, this subterm is
-lodged within the \elpiIn{app} constructor making unification very structural.
+\noindent In our type-class solver, we translate the \coqIn{forall_dec} instance
+of \coq (on the left) into the \elpi rule (on the right). This translation, derived
+directly from the type of \coqIn{forall_dec}, conceals certain unification
+properties that are accepted by the unification algorithm of the \ol but 
+rejected by the unification algorithm of the \ml. The primary issue of this
+translation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
+the \ml rule; \coqIn{P} is a higher-order variable, and in the original subterm, it
+sees the binder \coqIn{x}. However, the rule in the \ml interprets \elpiIn{P} as a 
+first-order variable, not seeing \elpiIn{x}. Additionally, this subterm is
+nested within the \elpiIn{app} constructor, introducing a structural challenge 
+to the unification process.
 
@@
 
-\noindent The two example of queries above, already in the syntax of the \ml,
-represent two terms that would unify in the \ol, but fail in the \ml. In
-particular, the first fails since the \ml is not able to unify \elpiIn{const
-`nat} with the term \elpiIn{app[P, x]}: they have different rigid heads.
+\noindent The two examples above, already expressed in the syntax of the \ml,
+depict two terms that would successfully unify in the \ol, but encounter failure
+in the \ml. Specifically, the first query fails because the \ml is not able to unify \elpiIn{const
+`nat} with the term \elpiIn{app[P, x]} due to their distinct rigid heads.
 
@@
 
-Following the definition give just before, the terms \elpiIn{app[const `f, const
-`y, x]} and \elpiIn{app[P, x]} should unify with substitution $\theta := \{P
+Based on the definition provided earlier, the terms \elpiIn{app[const `f, const
+`y, x]} and \elpiIn{app[P, x]} are expected to unify with substitution $\theta := \{P
 \mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
-is not possible in the \ml since, this time we have two terms with the same head
-(the \elpiIn{app} node), but the corresponding lists have different lengths.
+is not possible in the \ml. This difficulty arises because, in this instance,
+we have two terms with the same head
+(the \elpiIn{app} node), but their corresponding lists have different lengths.
 
-An immediate solution to solve these unification problems\todo{mettere da qualche parte anche eta e beta}
-would be to be lazy and build very general rule for each instance in the database.
-This would build a rule with the following shape:
+% Implementation using ol unification on each argument
+An immediate approach to address these unification problems\todo{mettere da qualche parte anche eta e beta}
+would be to adopt a lazy strategy and construct highly general rules for each instance in the database.
+This approach would result in a rule with the following structure:
 
@@
 \noindent In this second implementation, any call to the \elpiIn{tc-Decision}
-will unify with the head of the rule for \coqIn{forall_dec}. The the first
-premise will ensure that the arguments unify with the what we expect from the
-instance definition. Notice that, this time, the very unification algorithm of
-the \ol is exploited, overcoming the unification problems listed just before.
-This approach is quite appealing but it has two main disadvantages: the first
-lays on the fact that the communication between the \ol and the \ml can
-potentially be slow, this is due to the transformation of terms from one language
-to the other. \todo{here i refer to constr->elpi e elpi->constr}
-The other drawback of this latter encoding is that clause discrimination cannot
-be done on the head of the clause, for instance, any indexing 
-algorithm allowing to filter rules become suddenly useless.
+will unify with the head of the rule for \coqIn{forall_dec}. The first
+premise ensures that the arguments unify with what we expect from the
+instance definition. Notably, in this case, the unification algorithm of
+the \ol is leveraged, , successfully addressing the previously mentioned unification problems.
+
+However, this approach has two main drawbacks. Firstly, the communication
+between the \ol and the \ml may potentially be slow due to the transformation of
+terms from one language to the other \todo{here I refer to constr->elpi and
+elpi->constr}. Secondly, the alternative encoding lacks the ability to
+discriminate clauses based on the head of the clause. For example, any indexing
+algorithm designed to filter rules becomes ineffective in this scenario.
 

### 2024-02-16 main.tex
add todo in tex
update tex

@@
   various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
-  capable of automating proof resolution. In many of the previously the
+  capable of automating proof resolution. In many of the
   previously mentioned works, the equational theory expressed by the \ol is
   weaker than the \ml one. Consequently, no effort is required to adjust the
-  unification algorithm of the \ml since it encompass the \ml one. In our
+  unification algorithm of the \ml to represent the unification process of 
+  the \ol: the former algorithm encompassing the second one. In our
   specific scenario, however, we work with an \ol with a richer equational
-  theory than our \ml. This paper introduces a framework that enable the \ml to
-  comprehend the unification of the \ol and provides some practical applications
-  of these findings. Additionally, we extend our work to unification in the case
-  of higher-order variable quantification.
+  theory than our \ml. This paper introduces a framework that enables the \ml to
+  comprehend the unification of the \ol in the context of higher-order 
+  unification. Practical applications of these findings will be provided.
 
@@
 Thanks to meta programming, it is feasible to express the equational theory of a
-theorem prover, as indicated in \cite{citeX, Y Z}, with the \ml serving as the
+theorem prover, as indicated in \cite{X, Y, Z}, with the \ml serving as the
 foundational infrastructure for this ecosystem. The process of proving a theorem, that is
-solving a query using the jargon of logic programming consist in verifying if a
+solving a query using the jargon of logic programming, consist in verifying if a
 proposition is a logical consequence of a given set of assumption. The benefit
-of of using a prologish verification style is that the proof resolution can be
+of using a prologish verification style is that the proof resolution can be
 automatized thanks to different search strategies, such as a depth first search
@@
 structure to introduce ad-hoc polymorphism in functional languages.
-In \coq, the use of type class has increasingly become a programming
+In \coq, the use of type classes has increasingly become a programming
 style. Numerous libraries, such as \stdpp and
@@
 application expressed as a list of terms. Lambda abstractions are translated
-into the node \elpiIn{fun} binding a term to another term. Finally forall
-quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{prod x\ F}.
+into the node \elpiIn{fun}, binding a term to another term. Finally forall
+quantification on the form \coqIn{~$\forall$~x,F} in the \ol are translated
+into \elpiIn{prod x\ F} within the \ml.
 
@@
 
-\noindent In our type-class solver, we translate the \coqIn{forall_dec} instance
-of \coq (on the left) into the \elpi rule (on the right). This translation, derived
+\noindent In our type-class solver, the \coqIn{forall_dec} instance 
+of \coq (on the left) is compiled into the \elpi rule (on the right). This 
+compilation, derived
 directly from the type of \coqIn{forall_dec}, conceals certain unification
@@
 rejected by the unification algorithm of the \ml. The primary issue of this
-translation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
-the \ml rule; \coqIn{P} is a higher-order variable, and in the original subterm, it
+compilation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
+the \ml rule: \coqIn{P} is a higher-order variable and it
 sees the binder \coqIn{x}. However, the rule in the \ml interprets \elpiIn{P} as a 
@@
 \mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
-is not possible in the \ml. This difficulty arises because, in this instance,
-we have two terms with the same head
-(the \elpiIn{app} node), but their corresponding lists have different lengths.
+is not possible in the \ml. This difficulty arises because, even though 
+the two terms share the same same head
+(the \elpiIn{app} node), their corresponding lists have different lengths.
 
@@
 
-\noindent In this second implementation, any call to the \elpiIn{tc-Decision}
-will unify with the head of the rule for \coqIn{forall_dec}. The first
+\noindent In this second implementation, any query to \elpiIn{tc-Decision}
+will unify with the head of the rule above. The first
 premise ensures that the arguments unify with what we expect from the
 instance definition. Notably, in this case, the unification algorithm of
-the \ol is leveraged, , successfully addressing the previously mentioned unification problems.
+the \ol is leveraged successfully addressing the previously mentioned unification problems.
 
@@
 terms from one language to the other \todo{here I refer to constr->elpi and
-elpi->constr}. Secondly, the alternative encoding lacks the ability to
-discriminate clauses based on the head of the clause. For example, any indexing
+elpi->constr}. Secondly, a lack of % the ability to
+clauses discrimination based on the head of the clause is no longer
+possible. For example, any indexing
 algorithm designed to filter rules becomes ineffective in this scenario.
@@
 
-% da notare che il problema dopo, is the FO
-% approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
-% bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
-% vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
-% l'algo di unif di coq in questo caso specifico
+% TODO: parlare non so ancora dove dell'encoding shallow
+% dei terming ol in ml. Invece di usare i nodi app, prod etc... riusiamo
+% la stessa sintassi del linguaggio oggetto. Questo potrebbe funzionare 
+% bene nel caso di un ol con logica equazionale piu debole del ml, ma nel nostro
+% caso non si può applicare. inoltre vogliamo tenere un encoding più strutturato
+% per meglio analizzare i termini che ci arrivano nel ml.
 
-% where \coqIn{f : A -> B -> Type} and \coqIn{y: A}, we can notice that the \elpi
-% rule above cannot be applied; the terms \elpiIn{prod x\ app[P, x]} and
-% \elpiIn{prod x\ app[f, y, x]} cannot unify since the lists under the
-% \elpiIn{app} constructors are of different length.\todo{NOTE: the goal unify
-% with forall\_dec in coq xké fa unif FO approx}
+% TODO: related work: chiedere a enrico
 
-% Another example of unification failure is given by the the following instance
+% nuova section dove si da la spec dei due linguaggi
+
+% TODO: spiegare le proprietà di unification
 

### 2024-02-16 main.tex
wip abstract

@@
 \newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
+
+\begin{comment}
 \begin{abstract}
@@
 \end{abstract}
+\end{comment}
+
+\begin{abstract}
+Specifying and implementing a logic from scratch requires a
+considerable effort. Logical frameworks and Higher Order Logic Programming
+Languages provide a reusable
+Higher Order Meta Languages (ML) to make that easier:
+the specification and implementation of binding and substitution comes for free 
+if one reuses the the ML binders for the Object Logic (OL);
+similarly proof construction (and even proof search) is greatly simplified
+by leveraging the unification of the ML.
+Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
+$\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
+that were used to implement various OLs such as
+First Order Logic~\cite{felty88cade},
+Set Theory~\cite{10.1007/BF00881873} and
+even Higher Order Logic~\cite{books/sp/NipkowPW02}.
+
+We are interested in using a ML to describe Dependent Type Theory (TT) and
+implement a proof search procedure based on the unification (U) of the ML.
+U is typically able to solve higher order problems in the pattern
+fragment~\cite{miller92jsc} and identify ML terms up to $\eta\beta$.
+The problem we face is that a natural encoding of the OL (FOOL) force on
+U a ``first order''.
+To overcome the problem we provide an alternative encoding of TT (HOOL) where
+higher order unification problems can be faithfully delegated to U.
+We show how to map FOOL to HOOL and how to map back the
+unifiers found by U, effectively implementing an higher order unification
+algorithm for FOOL.
+
+We apply this technique to the implementation of a type-class solver for Coq,
+which is a proof search procedure based on back-chaining designated lemmas.
+Type-class solvers provide essential automation to widely used 
+Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
+and TLC~\cite{10.1007/978-3-642-14052-5_15} that constitute our test bed.
+\end{abstract}
 
 \section{Introduction}
+
+related~\cite{felty93lics}
 

### 2024-02-16 main.tex
wip abstract

@@
 
-We apply this technique to the implementation of a type-class solver for Coq,
-which is a proof search procedure based on back-chaining designated lemmas.
-Type-class solvers provide essential automation to widely used 
+We apply this technique to the implementation of a type-class solver for Coq
+in Elpi~\cite{dunchev15lpar} (a dialect of $\lambda$Prolog).
+Type-class solvers
+are proof search procedures based on back-chaining designated lemmas and
+provide essential automation to widely used 
 Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}

### 2024-02-16 main.tex
wip abstract

--- 2024-02-16 main.tex
@@
 \end{abstract}
+\end{comment}
+
+\newcommand{\U}{\ensuremath{\mathcal{U}}}
+\newcommand{\Fo}{\ensuremath{\mathcal{F}}}
+\newcommand{\Ho}{\ensuremath{\mathcal{H}}}
+
+\begin{abstract}
+Specifying and implementing a logic from scratch requires a
+considerable effort. Logical frameworks and Higher Order Logic Programming
+Languages provide dedicated, high level, Meta Languages (ML) to ease the
+task in two ways: 1)
+variable binding and substitution comes for free when ML binders
+are use to represent the Object Logic (OL) ones; 2)
+proof construction, and even proof search, is greatly simplified
+by leveraging the unification procedure provided by the ML.
+Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
+$\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
+that were used to implement various OLs such as
+First Order Logic~\cite{felty88cade},
+Set Theory~\cite{10.1007/BF00881873} and
+even Higher Order Logic~\cite{books/sp/NipkowPW02}.
+
+We are interested in using a ML to describe Dependent Type Theory (DTT) and
+implement a proof search procedure based on unification.
+The unification \U{} of our ML of choice Elpi~\cite{dunchev15lpar},
+a dialect of $\lambda$Prolog, is able to solve higher order problems in
+the pattern fragment~\cite{miller92jsc} and identifies terms up to $\eta\beta$.
+Our objective is to lift \U{} to the OL retaining the same equational theory.
+The problem we face is that a natural encoding of DTT (\Fo{}) makes
+\U{} essentially first order. To overcome the problem we provide a better
+behaved encoding \Ho{}, we show how to map \Fo{} terms to \Ho{} terms and
+how to map back the unifiers found by \U, effectively implementing an higher
+order unification algorithm for \Fo{}.
+
+We apply this technique to the implementation of a type-class solver for Coq
+Type-class solvers are proof search procedures based on back-chaining designated
+lemmas and provide essential automation to widely used 
+Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
+and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two constitute our test bed.
+\end{abstract}
 
 \section{Introduction}
+
+related~\cite{felty93lics}
 

### 2024-02-16 main.tex
wip abstract

@@
 variable binding and substitution comes for free when ML binders
-are use to represent the Object Logic (OL) ones; 2)
+are use to represent the object logic ones; 2)
 proof construction, and even proof search, is greatly simplified
@@
 $\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-that were used to implement various OLs such as
+that were used to implement various object logics such as
 First Order Logic~\cite{felty88cade},
@@
 
-We are interested in using a ML to describe Dependent Type Theory (DTT) and
-implement a proof search procedure based on unification.
-The unification \U{} of our ML of choice Elpi~\cite{dunchev15lpar},
-a dialect of $\lambda$Prolog, is able to solve higher order problems in
-the pattern fragment~\cite{miller92jsc} and identifies terms up to $\eta\beta$.
-Our objective is to lift \U{} to the OL retaining the same equational theory.
-The problem we face is that a natural encoding of DTT (\Fo{}) makes
-\U{} essentially first order. To overcome the problem we provide a better
-behaved encoding \Ho{}, we show how to map \Fo{} terms to \Ho{} terms and
-how to map back the unifiers found by \U, effectively implementing an higher
-order unification algorithm for \Fo{}.
-
-We apply this technique to the implementation of a type-class solver for Coq
+The object logic we are interested in is Dependent Type Theory (DTT) for which
+want to implement a unification procedure using
+Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
+The unification procedure \U{} provided by Elpi solves problems in the pattern
+fragment~\cite{miller92jsc} and identifies terms up to $\eta\beta$:
+the equational theory that we want to lift to DTT.
+The problem we face is that the natural encoding of DTT, that we call \Fo{},
+makes \U{} essentially first order. To overcome the problem we provide a better
+behaved encoding \Ho{}, we show how to map unification problems in \Fo{}
+to related problems in \Ho{} and
+how to map back the unifiers found by \U{}, effectively implementing
+an higher order unification algorithm for \Fo{}.
+
+We apply this technique to the implementation of a type-class~\cite{wadler89}
+solver for Coq~\cite{Coq-refman}.
 Type-class solvers are proof search procedures based on back-chaining designated

### 2024-02-16 main.tex
wip abstract

@@
 
-\newcommand{\U}{\ensuremath{\mathcal{U}}}
-\newcommand{\Fo}{\ensuremath{\mathcal{F}}}
-\newcommand{\Ho}{\ensuremath{\mathcal{H}}}
+\newcommand{\U}{\ensuremath{=_o}}
+\newcommand{\Ue}{\ensuremath{=_\lambda}}
+\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}}}
+\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}}
 
@@
 Specifying and implementing a logic from scratch requires a
-considerable effort. Logical frameworks and Higher Order Logic Programming
+considerable effort. Logical Frameworks and Higher Order Logic Programming
 Languages provide dedicated, high level, Meta Languages (ML) to ease the
@@
 variable binding and substitution comes for free when ML binders
-are use to represent the object logic ones; 2)
+are used to represent object logic ones; 2)
 proof construction, and even proof search, is greatly simplified
@@
 $\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-that were used to implement various object logics such as
+that were used to implement various logics such as
 First Order Logic~\cite{felty88cade},
@@
 The object logic we are interested in is Dependent Type Theory (DTT) for which
-want to implement a unification procedure using
+want to implement a unification procedure \U{} using
 Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-The unification procedure \U{} provided by Elpi solves problems in the pattern
-fragment~\cite{miller92jsc} and identifies terms up to $\eta\beta$:
-the equational theory that we want to lift to DTT.
-The problem we face is that the natural encoding of DTT, that we call \Fo{},
-makes \U{} essentially first order. To overcome the problem we provide a better
-behaved encoding \Ho{}, we show how to map unification problems in \Fo{}
+Elpi comes equipped with and equational theory \Ue{} that comprises $\eta\beta$ equivalence
+and higher order unification restricted to the pattern
+fragment~\cite{miller92jsc}.
+We want \U{} to feature the same equational theory of \Ue{}, but on the
+object logic DTT. Unfortunately the natural encoding of DTT in Elpi, that we call \Fo{},
+``underuses'' \Ue{} by making it applicable to first order unification problems only.
+To overcome the problem we provide a better
+behaved encoding \Ho{}, we show how to map a unification problems in \Fo{}
 to related problems in \Ho{} and
-how to map back the unifiers found by \U{}, effectively implementing
-an higher order unification algorithm for \Fo{}.
+how to map back the unifiers found by \Ue{}, effectively implementing
+ \U{} on top of \Ue{} for the natural encoding \Fo{}.
 
@@
 Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two constitute our test bed.
+and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
+constitute our test bed.
 \end{abstract}

### 2024-02-16 main.tex
wip abstract

@@
 Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-Elpi comes equipped with and equational theory \Ue{} that comprises $\eta\beta$ equivalence
+Elpi comes equipped with the equational theory \Ue{} that comprises $\eta\beta$ equivalence
 and higher order unification restricted to the pattern

### 2024-02-16 main.tex
wip abstract

@@
 variable binding and substitution comes for free when ML binders
-are used to represent object logic ones; 2)
+are used to represent the object logic ones; 2)
 proof construction, and even proof search, is greatly simplified
@@
 $\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-that were used to implement various logics such as
+that were used to implement various formal systems such as
 First Order Logic~\cite{felty88cade},
@@
 The object logic we are interested in is Dependent Type Theory (DTT) for which
-want to implement a unification procedure \U{} using
+want to implement a unification procedure \U{} using the ML
 Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.

### 2024-02-19 main.tex
thanks IA

--- 2024-02-16 main.tex
@@
 \begin{abstract}
-Specifying and implementing a logic from scratch requires a
-considerable effort. Logical Frameworks and Higher Order Logic Programming
-Languages provide dedicated, high level, Meta Languages (ML) to ease the
-task in two ways: 1)
-variable binding and substitution comes for free when ML binders
-are used to represent object logic ones; 2)
-proof construction, and even proof search, is greatly simplified
-by leveraging the unification procedure provided by the ML.
+Specifying and implementing a logic from scratch requires significant effort.
+Logical Frameworks and Higher Order Logic Programming Languages provide
+dedicated, high-level Meta Languages (ML) to facilitate this task in two
+key ways: 1) variable binding and substitution are simplified when ML binders
+represent object logic ones; 2) proof construction, and even proof search, is
+greatly simplified by leveraging the unification procedure provided by the ML.
 Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
-$\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-that were used to implement various logics such as
+$\lambda$Prolog~\cite{miller_nadathur_2012} and
+Isabelle~\cite{10.1007/978-3-540-71067-7_7}
+which have been utilized to implement various formal systems such as
 First Order Logic~\cite{felty88cade},
@@
 
-The object logic we are interested in is Dependent Type Theory (DTT) for which
-want to implement a unification procedure \U{} using
+The object logic we are interested in is Dependent Type Theory (DTT),
+for which we aim to implement a unification procedure \U{} using the ML
 Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-Elpi comes equipped with the equational theory \Ue{} that comprises $\eta\beta$ equivalence
+Elpi comes equipped with the equational theory \Ue{}, comprising
+$\eta\beta$ equivalence
 and higher order unification restricted to the pattern
 fragment~\cite{miller92jsc}.
-We want \U{} to feature the same equational theory of \Ue{}, but on the
-object logic DTT. Unfortunately the natural encoding of DTT in Elpi, that we call \Fo{},
-``underuses'' \Ue{} by making it applicable to first order unification problems only.
-To overcome the problem we provide a better
-behaved encoding \Ho{}, we show how to map a unification problems in \Fo{}
-to related problems in \Ho{} and
+We want \U{} to feature the same equational theory as \Ue{} but on the
+object logic DTT.
+Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
+``underuses'' \Ue{} by restricting it to first-order unification problems only. 
+To address this issue, we propose a better-behaved encoding, \Ho{},
+demonstrate how to map unification problems in \Fo{}
+to related problems in \Ho{}, and illustrate
 how to map back the unifiers found by \Ue{}, effectively implementing
@@
 Type-class solvers are proof search procedures based on back-chaining designated
-lemmas and provide essential automation to widely used 
+lemmas, providing essential automation to widely used 
 Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}

### 2024-02-19 paper/main.tex
add paper conf template

--- 2024-02-19 main.tex
@@
-\documentclass{rapport}
+\documentclass[acmengage]{acmart}
 
 \usepackage{myTools}
-
+% \addbibresource{bib.bib}
 \usepackage{macros}
 
-\author{Enrico Tassi, Davide Fissore}
-\title{HO unification from object language to meta language} 
-\date{Month XX, 202Y}
-
-\xspaceaddexceptions{]\}}
-
-\begin{document}
-
-\maketitle
+% TODO: set this fields
+\setcopyright{cc}
+\setcctype{by}
+\copyrightyear{2024}
+\acmYear{XXXX 2024}
+\acmBooktitle{YYY}
+\acmDOI{ZZZZZZZZZZZZ}
+
+% \xspaceaddexceptions{]\}}
 
@@
 \def\iris{\library{iris}}
-% \tableofcontents
 
@@
 
-\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
-
-\begin{comment}
-\begin{abstract}
-
-  % Unification is a fundamental process behind goal resolution in logic
-  % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
-  % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
-  % substitution is a mapping from unification variables to terms and it is the most
-  % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
-  % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
-  % Unification has been widely studied in the 90s [here the refs]. In this paper we
-  % want to give a framework to reason about unification in meta-programming.
-  Meta-programming \cite{hill1994} is a programming technique in which a
-  program, referred to as meta language (\ml), can treat an other program, knwon
-  as the object language (\ol) as its data. This programming approach is used in
-  various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
-  capable of automating proof resolution. In many of the
-  previously mentioned works, the equational theory expressed by the \ol is
-  weaker than the \ml one. Consequently, no effort is required to adjust the
-  unification algorithm of the \ml to represent the unification process of 
-  the \ol: the former algorithm encompassing the second one. In our
-  specific scenario, however, we work with an \ol with a richer equational
-  theory than our \ml. This paper introduces a framework that enables the \ml to
-  comprehend the unification of the \ol in the context of higher-order 
-  unification. Practical applications of these findings will be provided.
-
-\end{abstract}
-\end{comment}
-
 \newcommand{\U}{\ensuremath{=_o}}
@@
 \newcommand{\Ho}{\ensuremath{\mathcal{H}_o}}
+
+\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
+
+\begin{document}
+
+\title{HO unification from object language to meta language}
+
+\author{Enrico Tassi}
+\email{enrico.tassi@inria.fr}
+\affiliation{%
+  \institution{Université Côte d'Azur, Inria}
+  % \city{Nice}
+  \country{France}}
+
+\author{Davide Fissore}
+\email{davide.fissore@inria.fr}
+\affiliation{%
+  \institution{Université Côte d'Azur, Inria}
+  % \city{Nice}
+  \country{France}}
 
@@
 
+\keywords{Logic Programming, Meta-Programming, Higher-Order Unification, Proof Automation}
+
+\maketitle
+
+% \begin{comment}
+% \begin{abstract}
+
+%   % Unification is a fundamental process behind goal resolution in logic
+%   % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
+%   % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
+%   % substitution is a mapping from unification variables to terms and it is the most
+%   % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
+%   % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
+%   % Unification has been widely studied in the 90s [here the refs]. In this paper we
+%   % want to give a framework to reason about unification in meta-programming.
+%   Meta-programming \cite{hill1994} is a programming technique in which a
+%   program, referred to as meta language (\ml), can treat an other program, knwon
+%   as the object language (\ol) as its data. This programming approach is used in
+%   various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
+%   capable of automating proof resolution. In many of the
+%   previously mentioned works, the equational theory expressed by the \ol is
+%   weaker than the \ml one. Consequently, no effort is required to adjust the
+%   unification algorithm of the \ml to represent the unification process of 
+%   the \ol: the former algorithm encompassing the second one. In our
+%   specific scenario, however, we work with an \ol with a richer equational
+%   theory than our \ml. This paper introduces a framework that enables the \ml to
+%   comprehend the unification of the \ol in the context of higher-order 
+%   unification. Practical applications of these findings will be provided.
+
+% \end{abstract}
+% \end{comment}
+
 \section{Introduction}
@@
 
-% \begin{elpicode}
-%   kind term type.
-%   type const gref -> term.
-%   type app list term -> term.
-%   type fun (term -> term) -> term.
-%   type prod (term -> term) -> term.
-% \end{elpicode}
+% % \begin{elpicode}
+% %   kind term type.
+% %   type const gref -> term.
+% %   type app list term -> term.
+% %   type fun (term -> term) -> term.
+% %   type prod (term -> term) -> term.
+% % \end{elpicode}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Examples
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % Examples
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{mintedTcolorbox}{elpi}
+\begin{minted}{elpi}
   kind tm type.
@@
   type uv  nat -> tm.
-\end{mintedTcolorbox}
+\end{minted}
 
@@
 received unification problem $t_1 = t_2$ by 
-\begin{inlinelist}
-  \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
-  \ml
-  \item finding a valid substitution for $t_1'$ and $t_2'$ 
-  \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
-  is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
-\end{inlinelist}.
+% \begin{inlinelist}
+%   \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
+%   \ml
+%   \item finding a valid substitution for $t_1'$ and $t_2'$ 
+%   \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
+%   is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
+% \end{inlinelist}.
 
@@
 
-\begin{mintedTcolorbox}{elpi}
+\begin{minted}{elpi}
   kind ml.tm type.
@@
   type ml.uv nat -> list ml.tm -> ml.tm.
-\end{mintedTcolorbox}
+\end{minted}
 
@@
 
-\printbibliography
+% \printbibliography
 
@@
 
-\begin{mintedTcolorbox}[escapeinside=&&]{coq}
+\begin{minted}[escapeinside=&&]{coq}
   Inductive sig (A : Type) (P : A -> Prop) : Type := ...
@@
     RelDecision (@eq (sig A P)). Admitted.
-\end{mintedTcolorbox}
+\end{minted}
 
@@
 
-\begin{mintedTcolorbox}{elpi}
+\begin{minted}{elpi}
   type tc-Decision term -> term -> o.
@@
     tc-RelDecision A A (app [`eq`, A]) P2.
-\end{mintedTcolorbox}
+\end{minted}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bib}
+
 \end{document}

### 2024-02-19 main.tex
wip intro

@@
 Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
+\marginpar{natural is not the right word}
 ``underuses'' \Ue{} by restricting it to first-order unification problems only. 
@@
 
-related~\cite{felty93lics}
+Specifying and implementing a logic from scratch requires significant effort.
+Logical Frameworks and Higher Order Logic Programming Languages provide
+dedicated, high-level Meta Languages (ML) to facilitate this task in two
+key ways: 1) variable binding and substitution are simplified when ML binders
+represent object logic ones; 2) proof construction, and even proof search, is
+greatly simplified by leveraging the unification procedure provided by the ML.
+Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
+$\lambda$Prolog~\cite{miller_nadathur_2012} and
+Isabelle~\cite{10.1007/978-3-540-71067-7_7}
+which have been utilized to implement various formal systems such as
+First Order Logic~\cite{felty88cade},
+Set Theory~\cite{10.1007/BF00881873} and
+even Higher Order Logic~\cite{books/sp/NipkowPW02}.
+
+The object logic we are interested in is Dependent Type Theory (DTT),
+and we want to code a type-class~\cite{wadler89}
+solver for Coq~\cite{Coq-refman}.
+Type-class solvers are proof search procedures based on back-chaining designated
+lemmas, providing essential automation to widely used 
+Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
+and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
+constitute our test bed.
+
+For example, these 3 designated lemmas (Instances in the TC slang)
+are supposed to be animated, backward chained, by the TC solver.
+fin n is the type of numbers smaller than n, nfact the number of prime
+factors and Finite/Decision properties designated as classes what are
+automatically proved by Coq using the designated instances.
+
+\begin{coqcode}
+Instance fin_fin n : Finite (fin n).
+Instance nfact_dec n nf : Decision (nfact n nf).
+Instance forall_dec A P : Finite A ~$\to$~ ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
+\end{coqcode}
+
+for example here
+
+\begin{coqcode}
+  Check _ : Decision (forall y: fin 7, nfact y 3).
+\end{coqcode}
+
+A natural encoding (others are discussed section xx)
+uses c for constants, all for forall, and then app and lam
+for the lambda calculus part.
+
+\begin{elpicode}
+finite (app[c"fin", N]).
+decision (app [c"nfact",N,NF]).
+decision (all A x\ app[P, x]) :- (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+unfortunately this does not work since the last rule cannt
+backchain on the encoded goal
+
+\begin{elpicode}
+decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"])
+\end{elpicode}
+
+indeed this unif problem has no solution
+
+\begin{elpicode}
+app[c"nfact", y, c"3"] = app[P, y]
+\end{elpicode}
+
+in this paper we study a more sophisticated encoding
+
+\begin{elpicode}
+decision (all A x\ Pm x) :- link Pm A P, (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+This time Pm is HO for elpi, an the extra link is there to explain how to
+get P out of it.
+
+\begin{elpicode}
+app[c"nfact", y, c"3"] = Pm y
+Pm = y\app[c"nfact", y, c"3"]
+P = lam (app[c"fin",c"7"]) a\ app[c"nfact", a, c"3"]
+\end{elpicode}
+
+We also see that this is not trivial, because in the rec call
+we have now a beta redex.
+
+\subsection{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
+
+our choice of encoding of DTT may look weird to the reader familiar with
+LF, since used a shallow encoding of classes and binders, but not of the
+``lambda calculus'' part of DTT. Here a more lightweight encoding
+that unfortunately does not fit our use case
+
+\begin{elpicode}
+finite (fin N).
+decision (nfact N NF).
+decision (all A x\ P x) :- (pi x\ decision (P x)), finite A.
+\end{elpicode}
+
+but in DTT this is not always possible and not handy in our use case,
+since the arity of constants is not fixed.
+
+\begin{coqcode}
+Fixpoint narr T n := if n is S m then T -> narr T m else T.
+Definition nsum n : narr nat (n+1).
+Check nsum 2   8 9 : nat.
+Check nsum 3 7 8 9 : nat.
+\end{coqcode}
+
+moreover we use the same encoding for meta programming, or even just to provide
+hand written rules. We want to access the syntax of OL, so our embedding cannot
+be that shallow. We want to keep it shallow for the binders, but we need
+a c, app and lam nodes.
+
+another alternative
+
+\begin{elpicode}
+decision X :- unif X (all A x\ app[P, x]), (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+gives up all half of what the ML gives us. moreover even if unif here embodies
+he eq theory of DTT which is much stringer than the on of ML, we don't need 
+it. according to our experience eta beta suffice, but HO is needed.
+
+Note that this~\cite{felty93lics} is related and make the
+discrepancy between the types of ML and DTT visible. In this case
+one needs 4 application nodes. Moreover the objective is an encoding
+of terms, proofs, not proof search. Also note the conv predicate,
+akin to the unif we rule out.
+
+This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
+
+\section{Introduction}
 

### 2024-02-19 paper/main.tex
pick the right style, integrate the intro sketch

@@
-\documentclass[acmengage]{acmart}
+\documentclass[sigconf]{acmart}
 
@@
 % TODO: set this fields
-\setcopyright{cc}
-\setcctype{by}
+%\setcopyright{cc}
+%\setcctype{by}
 \copyrightyear{2024}
@@
 % \end{comment}
+
+\section{Introduction}
+
+Specifying and implementing a logic from scratch requires significant effort.
+Logical Frameworks and Higher Order Logic Programming Languages provide
+dedicated, high-level Meta Languages (ML) to facilitate this task in two
+key ways: 1) variable binding and substitution are simplified when ML binders
+represent object logic ones; 2) proof construction, and even proof search, is
+greatly simplified by leveraging the unification procedure provided by the ML.
+Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
+$\lambda$Prolog~\cite{miller_nadathur_2012} and
+Isabelle~\cite{10.1007/978-3-540-71067-7_7}
+which have been utilized to implement various formal systems such as
+First Order Logic~\cite{felty88cade},
+Set Theory~\cite{10.1007/BF00881873} and
+even Higher Order Logic~\cite{books/sp/NipkowPW02}.
+
+The object logic we are interested in is Dependent Type Theory (DTT),
+and we want to code a type-class~\cite{wadler89}
+solver for Coq~\cite{Coq-refman}.
+Type-class solvers are proof search procedures based on back-chaining designated
+lemmas, providing essential automation to widely used 
+Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
+and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
+constitute our test bed.
+
+For example, these 3 designated lemmas (Instances in the TC slang)
+are supposed to be animated, backward chained, by the TC solver.
+fin n is the type of numbers smaller than n, nfact the number of prime
+factors and Finite/Decision properties designated as classes what are
+automatically proved by Coq using the designated instances.
+
+\begin{coqcode}
+Instance fin_fin n : Finite (fin n).
+Instance nfact_dec n nf : Decision (nfact n nf).
+Instance forall_dec A P : Finite A ~$\to$~ ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
+\end{coqcode}
+
+for example here
+
+\begin{coqcode}
+  Check _ : Decision (forall y: fin 7, nfact y 3).
+\end{coqcode}
+
+A natural encoding (others are discussed section xx)
+uses c for constants, all for forall, and then app and lam
+for the lambda calculus part.
+
+\begin{elpicode}
+finite (app[c"fin", N]).
+decision (app [c"nfact",N,NF]).
+decision (all A x\ app[P, x]) :- (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+unfortunately this does not work since the last rule cannt
+backchain on the encoded goal
+
+\begin{elpicode}
+decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"])
+\end{elpicode}
+
+indeed this unif problem has no solution
+
+\begin{elpicode}
+app[c"nfact", y, c"3"] = app[P, y]
+\end{elpicode}
+
+in this paper we study a more sophisticated encoding
+
+\begin{elpicode}
+decision (all A x\ Pm x) :- link Pm A P, (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+This time Pm is HO for elpi, an the extra link is there to explain how to
+get P out of it.
+
+\begin{elpicode}
+app[c"nfact", y, c"3"] = Pm y
+Pm = y\app[c"nfact", y, c"3"]
+P = lam (app[c"fin",c"7"]) a\ app[c"nfact", a, c"3"]
+\end{elpicode}
+
+We also see that this is not trivial, because in the rec call
+we have now a beta redex.
+
+\subsection{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
+
+our choice of encoding of DTT may look weird to the reader familiar with
+LF, since used a shallow encoding of classes and binders, but not of the
+``lambda calculus'' part of DTT. Here a more lightweight encoding
+that unfortunately does not fit our use case
+
+\begin{elpicode}
+finite (fin N).
+decision (nfact N NF).
+decision (all A x\ P x) :- (pi x\ decision (P x)), finite A.
+\end{elpicode}
+
+but in DTT this is not always possible and not handy in our use case,
+since the arity of constants is not fixed.
+
+\begin{coqcode}
+Fixpoint narr T n := if n is S m then T -> narr T m else T.
+Definition nsum n : narr nat (n+1).
+Check nsum 2   8 9 : nat.
+Check nsum 3 7 8 9 : nat.
+\end{coqcode}
+
+moreover we use the same encoding for meta programming, or even just to provide
+hand written rules. We want to access the syntax of OL, so our embedding cannot
+be that shallow. We want to keep it shallow for the binders, but we need
+a c, app and lam nodes.
+
+another alternative
+
+\begin{elpicode}
+decision X :- unif X (all A x\ app[P, x]), (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+gives up all half of what the ML gives us. moreover even if unif here embodies
+he eq theory of DTT which is much stringer than the on of ML, we don't need 
+it. according to our experience eta beta suffice, but HO is needed.
+
+Note that this~\cite{felty93lics} is related and make the
+discrepancy between the types of ML and DTT visible. In this case
+one needs 4 application nodes. Moreover the objective is an encoding
+of terms, proofs, not proof search. Also note the conv predicate,
+akin to the unif we rule out.
+
+This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
 

### 2024-02-19 elpi.py
update main

@@ -1,179 +1,0 @@
-"""
-    pygments.lexers.elpi
-    ~~~~~~~~~~~~~~~~~~~~
-
-    Lexer for the `Elpi <http://github.com/LPCIC/elpi>`_ programming language.
-
-    :copyright: Copyright 2006-2023 by the Pygments team, see AUTHORS.
-    :license: BSD, see LICENSE for details.
-"""
-
-from pygments.lexer import RegexLexer, bygroups, include
-from pygments.token import Text, Comment, Operator, Keyword, Name, String, \
-    Number, Punctuation
-
-__all__ = ['ElpiLexer']
-
-
-class ElpiLexer(RegexLexer):
-    """
-    Lexer for the Elpi programming language.
-
-    .. versionadded:: 2.11
-    """
-
-    name = 'Elpi'
-    url = 'http://github.com/LPCIC/elpi'
-    aliases = ['elpi']
-    filenames = ['*.elpi']
-    mimetypes = ['text/x-elpi']
-
-    lcase_re = r"[a-z]"
-    ucase_re = r"[A-Z]"
-    digit_re = r"[0-9]"
-    schar2_re = r"([+*^?/<>`'@#~=&!])"
-    schar_re = r"({}|-|\$|_)".format(schar2_re)
-    idchar_re = r"({}|{}|{}|{})".format(lcase_re, ucase_re, digit_re, schar_re)
-    idcharstarns_re = r"({}*(\.({}|{}){}*)*)".format(idchar_re,
-                                                     lcase_re, ucase_re, idchar_re)
-    symbchar_re = r"({}|{}|{}|{}|:)".format(
-        lcase_re, ucase_re, digit_re, schar_re)
-    constant_re = r"({}{}*|{}{}|{}{}*|_{}+)".format(ucase_re, idchar_re,
-                                                    lcase_re, idcharstarns_re, schar2_re, symbchar_re, idchar_re)
-    symbol_re = r"(,|<=>|->|:-|;|\?-|->|&|=>|\bas\b|\buvar\b|<|=<|=|==|>=|>|\bi<|\bi=<|\bi>=|\bi>|\bis\b|\br<|\br=<|\br>=|\br>|\bs<|\bs=<|\bs>=|\bs>|@|::|\[\]|`->|`:|`:=|\^|-|\+|\bi-|\bi\+|r-|r\+|/|\*|\bdiv\b|\bi\*|\bmod\b|\br\*|~|\bi~|\br~)"
[…]

### 2024-02-19 pygmentize.py

@@ -1,53 +1,0 @@
-#! /usr/bin/env python
-import argparse
-import sys
-import pygments.cmdline as _cmdline
-import pygments.lexer as _lexer
-import pygments.token as _token
-
-def main(args):
-    parser = argparse.ArgumentParser()
-    parser.add_argument('-l', dest='lexer', type=str)
-    opts, rest = parser.parse_known_args(args[1:])
-    if opts.lexer == 'ssh_config':
-        args = [__file__, '-l', __file__ + ':SSHConfigLexer', '-x', *rest]
-    if args[2] == "elpi": 
-        args[2] = './elpi.py:ElpiLexer'
-    _cmdline.main(args + ['-x'])
-
-
-class SSHConfigLexer(_lexer.RegexLexer):
-    name = 'ssh_config'
-    tokens = {
-        'root': [
-            (r'(\s*Host)( .*\n)', _lexer.bygroups(
-                _token.Keyword, _token.String)),
-            (r'(\s*\w*)( ?.*\n)', _lexer.bygroups(
-                _token.Name.Attribute, _token.String)),
-        ],
-    }
-
-def isEmptyLine(x: str):
-    return x.lstrip(' ') == '\n'
-
-def getSpaceOfLine(x: str):
-    if isEmptyLine(x): 
-        return float('inf')
-    return len(x) - len(x.lstrip(' '))
-
-def getMinSpace(l):
-    mLen = min(map(getSpaceOfLine, l))
-    return [line[mLen:] if not(isEmptyLine(line)) else "\n" for line in l]
-
-import os
-
-if __name__ == '__main__':
-    f_name = ".aux/" + list(filter(lambda x: x.endswith("pyg"), os.listdir(".aux")))[0]
-    with open(f_name) as X:
-        lines = X.readlines()
-    with open(f_name, "w") as X:
-        X.write("".join(getMinSpace(lines)))
-    # with open("tt.txt", "w") as DUMP:
-    #     DUMP.write(str(os.listdir()))
-    #     DUMP.write("".join(getMinSpace(lines)))
-    main(sys.argv)

### 2024-02-19 paper/main.tex
updates compile
Some typos

@@
-\documentclass[sigconf]{acmart}
+\documentclass[sigconf,natbib=false]{acmart}
+\usepackage[]{biblatex}
+
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    Bib\TeX}}}
+\addbibresource{bib.bib}
 
 \usepackage{myTools}
-% \addbibresource{bib.bib}
 \usepackage{macros}
@@
 Instance nfact_dec n nf : Decision (nfact n nf).
-Instance forall_dec A P : Finite A ~$\to$~ ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
+Instance forall_dec A P : Finite A ~$\to$~ 
+  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
 \end{coqcode}
 
-for example here
+\noindent For example here
 
@@
 
-A natural encoding (others are discussed section xx)
+A natural encoding (others are discussed in section xx)
 uses c for constants, all for forall, and then app and lam
@@
 decision (app [c"nfact",N,NF]).
-decision (all A x\ app[P, x]) :- (pi x\ decision (app[P, x])), finite A.
-\end{elpicode}
-
-unfortunately this does not work since the last rule cannt
+decision (all A x\ app[P, x]) :- 
+  (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+Unfortunately this does not work since the last rule cannt
 backchain on the encoded goal
@@
 \begin{elpicode}
-decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"])
+decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"]).
 \end{elpicode}
@@
 \begin{elpicode}
-app[c"nfact", y, c"3"] = app[P, y]
-\end{elpicode}
-
-in this paper we study a more sophisticated encoding
-
-\begin{elpicode}
-decision (all A x\ Pm x) :- link Pm A P, (pi x\ decision (app[P, x])), finite A.
+app[c"nfact", y, c"3"] = app[P, y].
+\end{elpicode}
+
+In this paper we study a more sophisticated encoding
+
+\begin{elpicode}
+decision (all A x\ Pm x) :- link Pm A P, 
+  (pi x\ decision (app[P, x])), finite A.
 \end{elpicode}
@@
 
-our choice of encoding of DTT may look weird to the reader familiar with
+Our choice of encoding of DTT may look weird to the reader familiar with
 LF, since used a shallow encoding of classes and binders, but not of the
@@
 decision (nfact N NF).
-decision (all A x\ P x) :- (pi x\ decision (P x)), finite A.
+decision (all A x\ P x) :- 
+  (pi x\ decision (P x)), finite A.
 \end{elpicode}
@@
 \begin{coqcode}
-Fixpoint narr T n := if n is S m then T -> narr T m else T.
+Fixpoint narr T n := 
+  if n is S m then T -> narr T m else T.
 Definition nsum n : narr nat (n+1).
@@
 be that shallow. We want to keep it shallow for the binders, but we need
-a c, app and lam nodes.
-
-another alternative
-
-\begin{elpicode}
-decision X :- unif X (all A x\ app[P, x]), (pi x\ decision (app[P, x])), finite A.
-\end{elpicode}
-
-gives up all half of what the ML gives us. moreover even if unif here embodies
-he eq theory of DTT which is much stringer than the on of ML, we don't need 
-it. according to our experience eta beta suffice, but HO is needed.
+the c, app and lam nodes. % all ?
+
+Another alternative
+
+\begin{elpicode}
+decision X :- unif X (all A x\ app[P, x]), 
+  (pi x\ decision (app[P, x])), finite A.
+\end{elpicode}
+
+gives up all half of what the ML gives us. Moreover even if unif here embodies
+the eq theory of DTT which is much stronger than the one of the ML, we don't need 
+it. According to our experience eta beta suffice, but HO is needed.
 
@@
 \noindent %
-\begin{tabular}{c|c}
-  \begin{minipage}[c]{0.50\linewidth}
+% \begin{tabular}{c|c}
+  \begin{center}
+    \begin{minipage}[c]{0.8\linewidth}
       \begin{elpicode}
@@
     \end{minipage}
-  &
-  \begin{minipage}[c]{0.48\linewidth}
-    \begin{textcode}
-      Example:
-      a       ==> const `a
-      f a     ==> app[const `f, const `a]
-      ~$\lambda$~x => x ==> fun x\ x
-      ~$\forall$x, F x~ ==> prod x\ app[F, x]
-    \end{textcode}
-  \end{minipage}
-\end{tabular}
+  \end{center}
+
+  % &
+  % \begin{minipage}[c]{0.48\linewidth}
+  %   \begin{textcode}
+  %     Example:
+  %     a       ==> const `a
+  %     f a     ==> app[const `f, const `a]
+  %     ~$\lambda$~x => x ==> fun x\ x
+  %     ~$\forall$x, F x~ ==> prod x\ app[F, x]
+  %   \end{textcode}
+  % \end{minipage}
+% \end{tabular}
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bib}
+\printbibliography
 

### 2024-02-20 paper/main.tex
more work on intro

@@
   First Order Logic~\cite{felty88cade},
-  Set Theory~\cite{10.1007/BF00881873} and
-  even Higher Order Logic~\cite{books/sp/NipkowPW02}.
-
-  The object logic we are interested in is Dependent Type Theory (DTT),
+  Set Theory~\cite{10.1007/BF00881873},
+  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
+  Constuctions~\cite{felty93lics}.
+
+  The object logic we are interested in is Coq's~\cite{Coq-refman}
+  Dependent Type Theory (DTT),
   for which we aim to implement a unification procedure \U{} using the ML
@@
   We want \U{} to feature the same equational theory as \Ue{} but on the
-  object logic DTT.
-  Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
+  object logic DTT. Elpi also comes with an encoding for DTT that works well
+  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
+  Unfortunately this encoding, which we refer to as \Fo{},
   ``underuses'' \Ue{} by restricting it to first-order unification problems only. 
@@
   how to map back the unifiers found by \Ue{}, effectively implementing
-   \U{} on top of \Ue{} for the natural encoding \Fo{}.
+   \U{} on top of \Ue{} for the encoding \Fo{}.
 
@@
   solver for Coq~\cite{Coq-refman}.
-  Type-class solvers are proof search procedures based on back-chaining designated
-  lemmas, providing essential automation to widely used 
+  Type-class solvers are proof search procedures based on
+  unification that back-chain designated lemmas, providing essential
+  automation to widely used 
   Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
@@
 \maketitle
-
-% \begin{comment}
-% \begin{abstract}
-
-%   % Unification is a fundamental process behind goal resolution in logic
-%   % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
-%   % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
-%   % substitution is a mapping from unification variables to terms and it is the most
-%   % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
-%   % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
-%   % Unification has been widely studied in the 90s [here the refs]. In this paper we
-%   % want to give a framework to reason about unification in meta-programming.
-%   Meta-programming \cite{hill1994} is a programming technique in which a
-%   program, referred to as meta language (\ml), can treat an other program, knwon
-%   as the object language (\ol) as its data. This programming approach is used in
-%   various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
-%   capable of automating proof resolution. In many of the
-%   previously mentioned works, the equational theory expressed by the \ol is
-%   weaker than the \ml one. Consequently, no effort is required to adjust the
-%   unification algorithm of the \ml to represent the unification process of 
-%   the \ol: the former algorithm encompassing the second one. In our
-%   specific scenario, however, we work with an \ol with a richer equational
-%   theory than our \ml. This paper introduces a framework that enables the \ml to
-%   comprehend the unification of the \ol in the context of higher-order 
-%   unification. Practical applications of these findings will be provided.
-
-% \end{abstract}
-% \end{comment}
 
@@
 First Order Logic~\cite{felty88cade},
-Set Theory~\cite{10.1007/BF00881873} and
-even Higher Order Logic~\cite{books/sp/NipkowPW02}.
-
-The object logic we are interested in is Dependent Type Theory (DTT),
-and we want to code a type-class~\cite{wadler89}
-solver for Coq~\cite{Coq-refman}.
-Type-class solvers are proof search procedures based on back-chaining designated
-lemmas, providing essential automation to widely used 
-Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
-constitute our test bed.
-
-For example, these 3 designated lemmas (Instances in the TC slang)
-are supposed to be animated, backward chained, by the TC solver.
-fin n is the type of numbers smaller than n, nfact the number of prime
-factors and Finite/Decision properties designated as classes what are
-automatically proved by Coq using the designated instances.
+Set Theory~\cite{10.1007/BF00881873},
+Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
+Constuctions~\cite{felty93lics}.
+
+The object logic we are interested in is Coq's~\cite{Coq-refman}
+Dependent Type Theory (DTT), and we want to code a type-class~\cite{wadler89}
+solver for Coq~\cite{Coq-refman} using the Coq-Elpi~\cite{tassi:hal-01637063}
+meta programming framework.
+Type-class solvers are unification based proof search procedures
+that combine a set of designated lemmas in order to providing essential
+automation to widely used Coq libraries.
+
+As the running example we take the \coqIn{Decide} type class,
+from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
+library. The class identifies predicates equipped with a decision procedure.
+The following three designated lemmas (called \coqIn{Instances} in the
+type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
+smaller than \coqIn{n} is finite; 2) the predicate \coqIn{nfact n nf},
+linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
+3) the universal closure of a predicate has a decision procedure if the
+predicate has and if its domain is finite.
 
@@
 
-\noindent For example here
+\noindent Under this context the type-class solver is able to prove
+the the following statement automatically by back-chaining
+the three instances.
 
@@
 
-A natural encoding (others are discussed in section xx)
-uses c for constants, all for forall, and then app and lam
-for the lambda calculus part.
+\noindent
+The encoding of DTT provided by Elpi, that we will discuss at length later in
+section~\ref{sec:encodings} and ~\ref{} features the following term constructors:
+
+\begin{elpicode}
+kind tm type.
+type lam tm -> (tm -> tm) -> tm. % lambda abstraction
+type app list tm -> tm.          % n-ary application
+type all tm -> (tm -> tm) -> tm. % forall quantifier
+type c string -> tm.             % constants
+\end{elpicode}
+
+\noindent
+\marginpar{TODO: explain HOAS}
+Following this term encoding the three instances are represented by the
+following rules:
 
@@
 finite (app[c"fin", N]).
-decision (app [c"nfact",N,NF]).
-decision (all A x\ app[P, x]) :- 
-  (pi x\ decision (app[P, x])), finite A.
-\end{elpicode}
-
-Unfortunately this does not work since the last rule cannt
-backchain on the encoded goal
-
-\begin{elpicode}
-decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"]).
-\end{elpicode}
-
-indeed this unif problem has no solution
-
-\begin{elpicode}
-app[c"nfact", y, c"3"] = app[P, y].
-\end{elpicode}
-
-In this paper we study a more sophisticated encoding
-
-\begin{elpicode}
-decision (all A x\ Pm x) :- link Pm A P, 
-  (pi x\ decision (app[P, x])), finite A.
-\end{elpicode}
-
-This time Pm is HO for elpi, an the extra link is there to explain how to
-get P out of it.
+decision (app [c"nfact", N, NF]).
+decision (all A x\ app[P, x]) :- finite A,
+  pi x\ decision (app[P, x]).
+\end{elpicode}
+
+\noindent
+\marginpar{TODO: explain pi, cons}
+Unfortunately this direct translation of the instances considers the
+predicate \coqIn{P} as a first order term. If we try to backchain the
+third rule on the encoding of the goal above:
+
+\begin{elpicode}
+decision (all (app[c"fin", c"7"]) y\
+  app[c"nfact", y, c"3"]).
+\end{elpicode}
+
+\noindent
+we fail because of this ``higher order'' unification problem (in DTT)
+is phrased as a first order unification problem in the meta language.
+
+\begin{elpicode}
+app[c"nfact", y, c"3"] = app[P, y]
+\end{elpicode}
+
+\noindent
+In this paper we study a more sophisticated encoding of Coq terms allowing
+us to rephrase the problematic rule as follows:
+
+\begin{elpicode}
+decision (all A x\ Pm x) :- link Pm A P, finite A,
+  pi x\ decision (app[P, x]).
+\end{elpicode}
+
+\noindent
+This time \elpiIn{Pm} is an higher order unification variable (of type
+\elpiIn{tm -> tm}). The resulting unification problem is now:
 
@@
 app[c"nfact", y, c"3"] = Pm y
-Pm = y\app[c"nfact", y, c"3"]
-P = lam (app[c"fin",c"7"]) a\ app[c"nfact", a, c"3"]
-\end{elpicode}
-
-We also see that this is not trivial, because in the rec call
-we have now a beta redex.
-
-\subsection{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
+\end{elpicode}
+
+\noindent
+That admits one solution:
+
+\begin{elpicode}
+Pm = y\ app[c"nfact", y, c"3"]
+A = app[c"fin",c"7"]
+\end{elpicode}
+
+\noindent
+Elpi succeeds in the application of the new rule and then runs
+the premise \elpiIn{link Pm A P} that is in charge of bringing the
+assignment back to the domain of Coq terms (the type \elpiIn{tm}):
+
+\begin{elpicode}
+P = lam A a\ app[c"nfact", a, c"3"]
+\end{elpicode}
+
+\noindent
+This simple example is sufficient to show that the encoding we seek
+is not trivial. Indeed the solution for \elpiIn{P} generates a
+(Coq) $\beta$-redex in the second premise (under the \elpiIn{pi x}):
+
+\begin{elpicode}
+decision (app[lam A (a\ app[c"nfact", a, c"3"]), x])
+\end{elpicode}
+
+\noindent
+In turn the redex prevents the second rule to backchain properly since
+the following unification problem has no solution:
+
+\begin{elpicode}
+app[lam A (a\ app[c"nfact", a, c"3"]), x] =
+app[c"nfact", N, NF]
+\end{elpicode}
+
+\noindent
+This time the root cause is that the unification procedure of \Ue{} of the
+meta language is not aware of the equational theory of the object logic \U{},
+even if both theories include $\eta\beta$-conversion and admit most general
+unifiers for problems in the pattern fragment~\cite{miller92jsc}.
+
+In this paper we discuss alternative encodings of Coq in
+Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo{}
+in which the problems sketched here can be fully described.
+We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho{} (the language of
+the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
+bla bla..
+
+\section{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
+\label{sec:encodings}
 

### 2024-02-20 paper/elpi.py
add elpi lexer

@@ -1,0 +1,174 @@
+"""
+    pygments.lexers.elpi
+    ~~~~~~~~~~~~~~~~~~~~
+
+    Lexer for the `Elpi <http://github.com/LPCIC/elpi>`_ programming language.
+
+    :copyright: Copyright 2006-2024 by the Pygments team, see AUTHORS.
+    :license: BSD, see LICENSE for details.
+"""
+
+from pygments.lexer import RegexLexer, bygroups, include
+from pygments.token import Text, Comment, Operator, Keyword, Name, String, \
+    Number, Punctuation
+
+__all__ = ['ElpiLexer']
+
+
+class ElpiLexer(RegexLexer):
+    """
+    Lexer for the Elpi programming language.
+    """
+
+    name = 'Elpi'
+    url = 'http://github.com/LPCIC/elpi'
+    aliases = ['elpi']
+    filenames = ['*.elpi']
+    mimetypes = ['text/x-elpi']
+    version_added = '2.11'
+
+    lcase_re = r"[a-z]"
+    ucase_re = r"[A-Z]"
+    digit_re = r"[0-9]"
+    schar2_re = r"([+*^?/<>`'@#~=&!])"
+    schar_re = r"({}|-|\$|_)".format(schar2_re)
+    idchar_re = r"({}|{}|{}|{})".format(lcase_re,ucase_re,digit_re,schar_re)
+    idcharstarns_re = r"({}*(\.({}|{}){}*)*)".format(idchar_re, lcase_re, ucase_re, idchar_re)
+    symbchar_re = r"({}|{}|{}|{}|:)".format(lcase_re, ucase_re, digit_re, schar_re)
+    constant_re = r"({}{}*|{}{}|{}{}*|_{}+)".format(ucase_re, idchar_re, lcase_re, idcharstarns_re, schar2_re, symbchar_re, idchar_re)
+    symbol_re = r"(,|<=>|->|:-|;|\?-|->|&|=>|\bas\b|\buvar\b|<|=<|=|==|>=|>|\bi<|\bi=<|\bi>=|\bi>|\bis\b|\br<|\br=<|\br>=|\br>|\bs<|\bs=<|\bs>=|\bs>|@|::|\[\]|`->|`:|`:=|\^|-|\+|\bi-|\bi\+|r-|r\+|/|\*|\bdiv\b|\bi\*|\bmod\b|\br\*|~|\bi~|\br~)"
+    escape_re = r"\(({}|{})\)".format(constant_re,symbol_re)
[…]

### 2024-02-20 paper/main_old.tex
update main

--- 2024-02-20 paper/main.tex
@@
 
-\section{main idea: FO encoding - HO encoding}
-
-\begin{tabular}{c|c}
-\begin{minipage}{.5\textwidth}
-\begin{minted}{elpi}
+% \section{main idea: FO encoding - HO encoding}
+
+\section{Lanuages description}
+
+% Description of languages (OL and ML)
+
+In order to reason about unification of the terms of an objet language within a
+meta lanauge, we start by formally describing the two languages. We use
+meta-programming to accomplish this result. In \cref{code:lang-descr}, we
+provide on the left the term representation of the object language. %
+% Variables in the OL and ML are represented with integers
+In this encoding, the fo\_tm type stands for first order representation of the
+object language terms. We model unification variables as integers corresponding
+to memory adresses. This memory is represented witht the subst type which is 
+list of optional terms. If the cell $i$ is none, then the variable $i$ is not
+assigned.
+
+% Why the fo prefix for fo_tm && Uvar in OL have no scope
+This encoding is called first order, since variables have no scope. If we take
+back the subterm \coqIn{~$\forall$~x,P x} from the forall\_dec instance, we have
+\coqIn{P: A ~$\to$~ Prop} and in that subterm, \coqIn{P} has in scope the
+variable \coqIn{x}. The translation of this term in the ML is 
+\elpiIn{all x\app[uv 1, x]}; \elpiIn{x} is not in the scope of \elpiIn{uv 1}.
+
+% Uvar in ML have scope
+To solve this problem, we give more expressiveness to the OL variables, by
+giving an alternative term encoding, where unification variables are the
+variables of the meta language. These variables have a scope and therefore
+can be of higher order.
+
+% The subst mappings
+In order to represent variable assignation, we declare the two types subst\_fo
+and subst coressponding to the substitution mapping of unification variable
+for both the OL and the ML.
+
+\def\eqfo{equal\_fo\xspace}
+\def\eqho{equal\_ho\xspace}
+
+% We have two different equal functions
+Unification over terms is possible thanks to the equal predicate. There exist
+two version of this predicate: \eqfo and \eqho. Their implementation is given in
+\cref{code:term-equal}. The first tests if two terms are equal in the object
+langauge, that is, given a substitution $\theta$, is it true that the two terms
+are equal? Note that no variable instanciation is done: unification is not
+performed by the \elpiIn{equal} predicate. Therefore, a unification variabile
+$i$ equal to another term $t$ if $t$ is a variable with same index $i$, or if
+$i$ is assigend to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover
+\eqfo, is able to verify if two terms are equal up to $\eta\beta$.
+
+The \eqho predicate to test if two terms of the
+
+% Compile FO to HO
+
+% Assignement in HO should be decompiled
+
+-------------------------------------------------------------------
+
+\begin{figure*}
+  \begin{tabular}{c}
+    \begin{minipage}{.35\textwidth}
+
+      \begin{elpicode}
         kind fo_tm type.
         type fo_app list fo_tm -> fo_tm.
-type fo_lam (tm -> fo_tm) -> fo_tm.
+        type fo_lam (fo_tm -> fo_tm) -> fo_tm.
         type fo_uv  nat -> fo_tm.
@@
 
-typeabbrev subst list (option tm).
-type equal subst -> tm -> tm -> o.
-\end{minted}
+        typeabbrev subst list (option fo_tm).
+        type eq subst -> fo_tm -> fo_tm -> o.
+      \end{elpicode}
     \end{minipage}
-&
-\begin{minipage}{.5\textwidth}
-\begin{minted}{elpi}
+    \begin{minipage}{.35\textwidth}
+
+      \begin{elpicode}
         kind tm type.
@@
         typeabbrev subst list (option assmt).
-type equal subst -> tm -> tm -> o.
-\end{minted}
+        type eq subst -> tm -> tm -> o.
+      \end{elpicode}
     \end{minipage}
@@
 
-le due sintassi, XX tradotto, unif passa, bisogna riportare la soluzione
-nel mondo fo
-
-this should hold in both encoding
-%abella
-\begin{minted}{elpi}
-forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
-\end{minted}
-we have unif only in the HO one, where equal is
-weaker (but unif is stronger)
-
-segnatura di decomp e unif, equalfo (fa beta, eta, deref)
-
-\noindent
+  \caption{Language description}
+  \label{code:lang-descr}
+  \Description[lang-spec]{Language description}
+\end{figure*}
+
+% The two equals: equal of OL and ML
+% In ML we don't have eta beta -> only unification is allowed on variables
+
+\begin{figure*}
   \begin{tabular}{c}
@@
     \end{minipage}
-\\\hline
+    \\ \hline
     \begin{minipage}{.95\textwidth}
@@
     \end{minipage}
-
   \end{tabular}
+  \caption{Term equality}
+  \label{code:term-equal}
+  \Description[term-equal]{Term equality}
+\end{figure*}
+
+le due sintassi, XX tradotto, unif passa, bisogna riportare la soluzione
+nel mondo fo
+
+this should hold in both encoding
+%abella
+\begin{minted}{elpi}
+forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
+\end{minted}
+we have unif only in the HO one, where equal is
+weaker (but unif is stronger)
+
+segnatura di decomp e unif, equalfo (fa beta, eta, deref)
+
+\noindent
 

### 2024-02-20 paper/main.tex
clear main

@@
 
-\section{Introduction}
+\section{Lanuages description}
 
-related~\cite{felty93lics}
+\input{lang_descr.tex}
 
-% meta programming
-Meta programming is a powerful programming style that enable the incorporation
-of an object language (\ol) into a meta language (\ml). This control over the
-\ol lies at the core of meta-programming, facilitating a spectrum of operations
-on the object language. Some of the principal program manipulations range
-from interpretation to compilation.
+% Description of languages (OL and ML)
 
-% theorem prover and tc
-Thanks to meta programming, it is feasible to express the equational theory of a
-theorem prover, as indicated in \cite{X, Y, Z}, with the \ml serving as the
-foundational infrastructure for this ecosystem. The process of proving a theorem, that is
-solving a query using the jargon of logic programming, consist in verifying if a
-proposition is a logical consequence of a given set of assumption. The benefit
-of using a prologish verification style is that the proof resolution can be
-automatized thanks to different search strategies, such as a depth first search
-with backtracking. One prominent manifestation of this automated proof process is
-type-class resolution\cite{wadler89, sozeau08}. Type classes serve as a typing
-structure to introduce ad-hoc polymorphism in functional languages.
-In \coq, the use of type classes has increasingly become a programming
-style. Numerous libraries, such as \stdpp and
-\iris are built upon this machinery.
+In order to reason about unification of the terms of an objet language within a
+meta lanauge, we start by formally describing the two languages. We use
+meta-programming to accomplish this result. In \cref{code:lang-descr}, we
+provide on the left the term representation of the object language. %
+% Variables in the OL and ML are represented with integers
+In this encoding, the fo\_tm type stands for first order representation of the
+object language terms. We model unification variables as integers corresponding
+to memory adresses. This memory is represented witht the subst type which is 
+list of optional terms. If the cell $i$ is none, then the variable $i$ is not
+assigned.
 
-% tc motivates this paper
-The motivating example for this article is our investigation for an alternative
-type-class solver for \coq in \elpi\todo{citation to elpi} through the \coqelpi
-plugin. This plugin gives a deep embedding of \coq terms into \elpi syntax, but
-challenges arise whe attempting to solve type-class goals represented into their
-corresponding \elpi representation.
+% Why the fo prefix for fo_tm && Uvar in OL have no scope
+This encoding is called first order, since variables have no scope. To better
+understand this definition, let's take the subterm \coqIn{P x} from the instance
+forall\_dec. \coqIn{P} has type \coqIn{A ~$\to$~ Prop} and \coqIn{x} is a name
+bound to \coqIn{P}. The application of \coqIn{P} to \coqIn{x} at the OL level,
+put this name in the scope of the variable \coqIn{P}. However, at the meta
+level, the translation of \coqIn{P x} become \elpiIn{app[uv 1, x]}; \elpiIn{x}
+is not in the scope of \elpiIn{uv 1}.
 
-% syntax of coq terms in elpi
-In particular, \coq terms are encoded with the following (very simplified)
-data type
+% Uvar in ML have scope
+To solve this problem, we give more expressiveness to the OL variables, by
+giving an alternative term encoding, where unification variables are the
+variables of the meta language. These variables have a scope and therefore
+can be of higher order.
+per la equal, 
+% The subst mappings
+In order to represent variable assignation, we declare the two types subst\_fo
+and subst coressponding to the substitution mapping of unification variable
+for both the OL and the ML.
 
-% % \begin{elpicode}
-% %   kind term type.
-% %   type const gref -> term.
-% %   type app list term -> term.
-% %   type fun (term -> term) -> term.
-% %   type prod (term -> term) -> term.
-% % \end{elpicode}
+\def\eqfo{equal\_fo\xspace}
+\def\eqho{equal\_ho\xspace}
+\def\etabeta{$\eta\beta$\xspace}
 
-\noindent %
-% \begin{tabular}{c|c}
-  \begin{center}
-    \begin{minipage}[c]{0.8\linewidth}
-      \begin{elpicode}
-        kind term type.
-        type const gref -> term.
-        type app list term -> term.
-        type fun (term -> term) -> term.
-        type prod (term -> term) -> term.
-      \end{elpicode}
-    \end{minipage}
-  \end{center}
+% We have two different equal functions
+Unification over terms is possible thanks to the equal predicate. There exist
+two version of this predicate: \eqfo and \eqho. Their implementation is given in
+\cref{code:term-equal}. The first tests if two terms are equal in the object
+langauge, that is, given a substitution $\theta$, is it true that the two terms
+are equal? Note that no variable instanciation is done: unification is not
+performed by the \elpiIn{equal} predicate. Therefore, a unification variabile
+$i$ equal to another term $t$ if $t$ is a variable with same index $i$, or if
+$i$ is assigend to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover
+\eqfo, is able to verify if two terms are equal up to \etabeta. This equality
+relation is in fact the equality relation of the object langauge and therefore,
+since it accepts \etabeta equality, then \eqfo should take into account these
+reductions.
 
-  % &
-  % \begin{minipage}[c]{0.48\linewidth}
-  %   \begin{textcode}
-  %     Example:
-  %     a       ==> const `a
-  %     f a     ==> app[const `f, const `a]
-  %     ~$\lambda$~x => x ==> fun x\ x
-  %     ~$\forall$x, F x~ ==> prod x\ app[F, x]
-  %   \end{textcode}
-  % \end{minipage}
-% \end{tabular}
+The \eqho predicate test if two terms are equal in the world of the meta
+languages. This eaquality function is not capable to understand the \etabeta
+reductions of the OL. This is mainly due to the fact that the meta language 
+understand \etabeta, but, since the nodes \elpiIn{app} and \elpiIn{lam} do
+not respectively represent the internal application and lambda abstraction
+of the language, the \etabeta reducation is not possible.
 
-\noindent where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
-identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
-is embedded in the \elpiIn{app} node, having the head and the arguments of the
-application expressed as a list of terms. Lambda abstractions are translated
-into the node \elpiIn{fun}, binding a term to another term. Finally forall
-quantification on the form \coqIn{~$\forall$~x,F} in the \ol are translated
-into \elpiIn{prod x\ F} within the \ml.
+As we can
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % Examples
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Compile FO to HO
 
-Let's take the following example from the \stdpp library:
+% Assignement in HO should be decompiled
 
-\noindent %
-\begin{tabular}{c|c}
-  \begin{minipage}[c]{0.50\linewidth}
-    \begin{coqcode}
-      Instance forall_dec A (P: A ~$\to$~ Prop):
-        Finite A ~$\to$~ 
-        ~$\forall$~x, Decision (P x) ~$\to$~
-        Decision (~$\forall$~x, P x).
-    \end{coqcode}
-  \end{minipage}
-  &
-  \begin{minipage}[c]{0.46\linewidth}
-    \begin{elpicode}
-      tc-Decision (prod x\ app[P, x]) :- 
-        % premise for Finite A
-        % premise for $\forall$x, Decision (P x)
-    \end{elpicode}
-  \end{minipage}
-\end{tabular}
-
-\noindent In our type-class solver, the \coqIn{forall_dec} instance 
-of \coq (on the left) is compiled into the \elpi rule (on the right). This 
-compilation, derived
-directly from the type of \coqIn{forall_dec}, conceals certain unification
-properties that are accepted by the unification algorithm of the \ol but 
-rejected by the unification algorithm of the \ml. The primary issue of this
-compilation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
-the \ml rule: \coqIn{P} is a higher-order variable and it
-sees the binder \coqIn{x}. However, the rule in the \ml interprets \elpiIn{P} as a 
-first-order variable, not seeing \elpiIn{x}. Additionally, this subterm is
-nested within the \elpiIn{app} constructor, introducing a structural challenge 
-to the unification process.
-
-% Esempio HO in PF
-\elpiInEq{tc-Decision (prod x\ const `nat)}%
-% Esempio FO align
-\elpiInEq{tc-Decision (prod x\ app[const `f, const `y, x])}
-
-\noindent The two examples above, already expressed in the syntax of the \ml,
-depict two terms that would successfully unify in the \ol, but encounter failure
-in the \ml. Specifically, the first query fails because the \ml is not able to unify \elpiIn{const
-`nat} with the term \elpiIn{app[P, x]} due to their distinct rigid heads.
-
-The second goal refers to an approximation of the unification\todo{maybe there's
-something to cite} algorithm of the \ol. Spiegare l'approssimazione. E che
-attraverso il nostro framework è possibile "spiegare" l'algorithm di unification
-usato da ogni \ol nel ml desiderato. \todo{something to do here...}
-
-Based on the definition provided earlier, the terms \elpiIn{app[const `f, const
-`y, x]} and \elpiIn{app[P, x]} are expected to unify with substitution $\theta := \{P
-\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
-is not possible in the \ml. This difficulty arises because, even though 
-the two terms share the same same head
-(the \elpiIn{app} node), their corresponding lists have different lengths.
-
-% Implementation using ol unification on each argument
-An immediate approach to address these unification problems\todo{mettere da qualche parte anche eta e beta}
-would be to adopt a lazy strategy and construct highly general rules for each instance in the database.
-This approach would result in a rule with the following structure:
-
-\begin{center}
-  \begin{minipage}[c]{0.46\linewidth}
-    \begin{elpicode}
-      tc-Decision X :- 
-        ol.unify X (prod x\ app[P, x]),
-        % premise for Finite A
-        % premise for $\forall$x, Decision (P x)
-    \end{elpicode}
-  \end{minipage}
-\end{center}
-
-\noindent In this second implementation, any query to \elpiIn{tc-Decision}
-will unify with the head of the rule above. The first
-premise ensures that the arguments unify with what we expect from the
-instance definition. Notably, in this case, the unification algorithm of
-the \ol is leveraged successfully addressing the previously mentioned unification problems.
-
-However, this approach has two main drawbacks. Firstly, the communication
-between the \ol and the \ml may potentially be slow due to the transformation of
-terms from one language to the other \todo{here I refer to constr->elpi and
-elpi->constr}. Secondly, a lack of % the ability to
-clauses discrimination based on the head of the clause is no longer
-possible. For example, any indexing
-algorithm designed to filter rules becomes ineffective in this scenario.
-
-% Contribution
-Finally
-
-% TODO: parlare non so ancora dove dell'encoding shallow
-% dei terming ol in ml. Invece di usare i nodi app, prod etc... riusiamo
-% la stessa sintassi del linguaggio oggetto. Questo potrebbe funzionare 
-% bene nel caso di un ol con logica equazionale piu debole del ml, ma nel nostro
-% caso non si può applicare. inoltre vogliamo tenere un encoding più strutturato
-% per meglio analizzare i termini che ci arrivano nel ml.
-
-% TODO: related work: chiedere a enrico
-
-% nuova section dove si da la spec dei due linguaggi
-
-% TODO: spiegare le proprietà di unification
-
-\section{intro}
-
-we are interested in using a meta language in LF style to write
-automation, proof search. In various works... they achieve that
-for a OL which is simpler than the LF, the equational theory is included
-in the one of the ML. This is exploited to piggy back on the unif
-of the ML.
-the peculiarity of our setting is that the OL has a richer equational
-theory of the ML, eg beta eta zeta bla bla. Moreover is HO logic, so
-quantifies over functions, so unif variable range on that too.
-We want to piggy back on the ML unif whenever the problem fits
-in its domain, eg pattern fragment.
-this is important for practical purposes.
-
-\subsection{in a nutshell}
-example, a rule for theorem
-$$
-forall f : A -> B, .. bla f .. -> prove (\forall x, f x)
-bla g
-$$
-\begin{elpicode}
-type app ...
-prove (forall x\ app F x) :- ... bla F ...
-\end{elpicode}
-on goal \mint{elpi}|prove (forall x\ app (app g x) x))|
-would fail since \mint{elpi}|F != app g x|.
-
-of course one wants to avoid
-\begin{elpicode}
-prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assignment "F" S' F, bla F S' ...
-\end{elpicode}
-
-Now, ML has HO variables
-\begin{elpicode}
-type lam ..
-prove (forall x\ F x) :- ... bla F ...
-\end{elpicode}
-this time  \mint{elpi}|F x != app (app g x) x| has solution
-\mint{elpi}|F = a\ app (app g x) x| but F is not a term so
-bla needs to be adapted,
-\begin{elpicode}
-prove (forall x\ F x) :- ... bla (lam F) ...
-bla g.
-\end{elpicode}
-this is too simplistic since \mint{elpi}|g != lam x\ app g x|
-
-\subsection{contribution}
-\begin{itemize}
-\item prover for HO OL in ML that uses unif
-\item eta beta
-\item test on stdpp and TLC
-\end{itemize}
-
-\section{Introduction}
-
-Meta programming \cite{hill1994} is a programming technique in which a program
-can treat an other program as its data. This latter program is called object
-language (\ol for short), while the former is called meta language (\ml for
-short). At the heart of meta programming lays the necessity of representing
-terms of the \ol in the \ml so that a wide set of program
-manipulations ranging from interpretation to compilation. 
-% Furthermore, since a
-% logic program is a set of clauses, the behavior of the \ol can be
-% modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.
-
-Meta programming has various application such as ... where thanks to meta
-programming it is possible to represent the logic of a language into a formal
-and formally verify the wanted properties. On the other hand, it is possible to
-embed a logic programming language into another so that some tasks can be
-delegated to \ml.
-
-The latter situation motivates our works, since we are implementing a type-class
-solver for the \ol \coq in the \ml \elpi (a variant \lambdaprolog). A type class
-\cite{wadler89, sozeau08} is a typing structure allowing to introduce ad hoc
-polymorphism in functional languages. We call <<instance>> an implementation of
-a type class. The resolution of a type-class problem can be viewed as a logic
-program where type classes represent predicates parametrized by their arguments
-and where instances are rules for those predicates.
-
-% \section{Higher-order unification in \texorpdfstring{$L_\lambda$}{L\_lambda}}
-
-% Higher-order unification is a unification problem known to be undecidable. The
-% goal is to determine a most general unifier between two lambda terms, where at
-% least one of contain a sub-term $F$ applied to a non empty list of arguments
-% $a_1,\dots,a_n$ and $F$ is a variable. Some researches have been done in order to
-% identify some easy patterns where it is possible to solve this kind of
-% unification.
-
-% In \cite{miller1991}, D. Miller provides a unification algorithm (the function
-% $\eqtau$) for higher-order variable when they are in the so-called pattern
-% fragment.
-
-esempietto in cui l'HO di elpi non risolve un problem HO del linguaggio
-oggetto FO. Equazione XX
-
-\begin{coqcode}
-(* HO unif *)
-Instance forall_dec A (P: A ~$\to$~ Prop) `{Finite A} `{~$\forall$~x, Decision (P x)} :
-  Decision (~$\forall$~x, P x).
-Instance and_dec P `{Decision P} Q `{Decision Q}:  Decision (P /\ Q).
-
-(* FO approx *)
-Instance decide_rel A B (R: A ~$\to$~ B ~$\to$~ Prop) `{RelDecision R} :
-  forall x y, Decision (R x y).
-Instance decide_eq_nat : RelDecision (@eq nat).
-Check _ : Decision (@eq nat 2 3)
-
-Instance _ : Inj add.
-             Inj F -> Inj (fun x => F (G x))
-
-             (fun x => add x)
-
-  Decision (~$\forall$~x, P x).
-
-\end{coqcode}
-
-\subsection{Related work and alternative approaches}
-
-LF e PL with binders (sempre HOAS).
-
-citare FO logic et similia fatti in LP, twelf. qui l'unification dell'OL
-e' facile. cercare se in twelf hanno fatto un po' di ordine superiore.
-
-isabelle fa la stessa cosa, OL e HOL, che matcha il sistema di tipi.
-qui l'unificazione e' la stessa, e fanno un ecoding shallow (senza app).
-
-Noi abbiamo un OL piu complicato, i tipi non matchano, serve un nodo app
-per due ragioni: meta programmazione e arieta variabile
-
-(equazioni XX senza app e lam) XXX IDEA XXX
-
-\section{main idea: FO encoding - HO encoding}
-
-\begin{tabular}{c|c}
-\begin{minipage}{.5\textwidth}
-\begin{minted}{elpi}
-kind fo_tm type.
-type fo_app list fo_tm -> fo_tm.
-type fo_lam (tm -> fo_tm) -> fo_tm.
-type fo_uv  nat -> fo_tm.
-type fo_c string -> fo_tm.
-
-typeabbrev subst list (option tm).
-type equal subst -> tm -> tm -> o.
-\end{minted}
-\end{minipage}
-&
-\begin{minipage}{.5\textwidth}
-\begin{minted}{elpi}
-kind tm type.
-type app list tm -> tm.
-type lam (tm -> tm) -> tm.
-type uv  nat -> list tm -> tm.
-type c string -> tm.
-
-kind assmt type.
-type abs (tm -> assmt) -> assmt.
-type val tm -> assmt.
-
-typeabbrev subst list (option assmt).
-type equal subst -> tm -> tm -> o.
-\end{minted}
-\end{minipage}
-\end{tabular}
+-------------------------------------------------------------------
 
@@
 %abella
-\begin{minted}{elpi}
+\begin{elpicode}
 forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
-\end{minted}
+\end{elpicode}
 we have unif only in the HO one, where equal is
@@
 \noindent
-\begin{tabular}{c}
-\begin{minipage}{.95\textwidth}
-\begin{minted}{elpi}
-% deref
-equal S (uv N) T1 :- assigned? N S T, equal S T T1.
-equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
-equal S (app [uv N|A]) T1 :- assigned? N S F, beta F A T, equal S T T1.
-equal S T1 (app [uv N|A]) :- assigned? N S F, beta F A T, equal S T1 T.
-% congruence
-equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
-equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
-equal _ (c X) (c X).
-equal _ (uv N) (uv N).
-% beta
-equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
-equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
-% eta
-equal S (lam F) T :- not (T = lam _),
-  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
-equal S T (lam F) :- not (T = lam _),
-  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
-\end{minted}
-\end{minipage}
-\\\hline
-\begin{minipage}{.95\textwidth}
-\begin{minted}{elpi}
-% deref
-equal S (uv N A) T1 :- assigned? N S F, beta-abs F A T, equal S T T1.
-equal S T1 (uv N A) :- assigned? N S F, beta-abs F A T, equal S T1 T.
-% congruence
-equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
-equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
-equal _ (c X) (c X).
-equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
-\end{minted}
-\end{minipage}
-
-\end{tabular}
 
@@
 \begin{elpicode}
-  p (all x\ app[F,x]) = p (all x\ app[f,x,x])
+  :index (1)
+  p (all x\ app[F,x]).
 \end{elpicode}
@@
 
-\section{XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-Even though type-class resolution is the motivating example of this paper, we
-provide a general framework allowing to solve reproduce the same unification
-properties of the \ol into the \ml. In other word, if two terms unify in the
-\ol, then they still unify in the \ml.
-
-In the following, we consider the \ol being able to quantify over higher-order
-variables and accepts $\eta-\beta$-reductions. The same unification properties
-are considered valid for the \ml.
-
-There exist two different ways to encode the \ol in the \ml, we can either deep
-embed the \ol such that any term of the \ol is represented with a corresponding
-predicate. For example, if \texttt{f} is a function of type \coqIn{A -> B} in
-the \ol, then the \ml has the predicate \texttt{p} defined as \elpiIn{type p A'
--> B' -> o}, where \elpiIn{A'} and \elpiIn{B'} are types corresponding
-respectively to \coqIn{A} and \coqIn{B} in the \ol. In a theorem prover like
-\coq, we can translate theorems like the following statement
-
-\begin{equation}
-  \texttt{\coqIn{forall F X, p (f X) (fun x => g x (F x)).}}
-  \label{eq:coq1}
-\end{equation}
-
-where \texttt{p, f} and \texttt{g} are defined constants of the language, into
-
-\begin{equation}
-  \texttt{\elpiIn{p (f X) (x\ g x (F x)).}}
-\end{equation}
-
-\noindent However, even if this encoding is quite appealing since it allows to
-mirror enough straightforwardly the terms of the \ol, we loose the possibility
-the manipulate the terms of the \ol into the \ml. In other words, we have no
-syntax allowing to know if the current term is a constant, an application, a
-lambda abstraction and so on. This is mainly due to the absence of a syntax in
-the encoding of the \ol terms. Moreover, another motivation for using syntax to
-represent terms of the \ol is that  the typing system of the \ol could
-potentially be more expressive than the typing system of the \ml\footnote{This
-is the case for \coq wrt \elpi, since in we have no immediate way to encode
-the dependent types of \coq into \elpi}.
-
-To simplify the understanding of our encoding, in the following code snippet
-we give the typing schema of the \ol terms represented into terms of the \ml.
-
-\begin{minted}{elpi}
-  kind tm type.
-  type app list tm -> tm.
-  type lam (tm -> tm) -> tm.
-  type c string -> tm.
-  type uv  nat -> tm.
-\end{minted}
-
-In particular, the type \elpiIn{tm} is the type of the terms of the \ol. The
-function applications of the \ol are represented as a list of \elpiIn{tm}
-prefixed by the constructor \elpiIn{app}. The \elpiIn{lam} constructor,
-represent lambda abstractions of the \ol binding a \elpiIn{tm} into an other
-\elpiIn{tm}. Constants as strings inside the constructor \elpiIn{c}. Lastly,
-unification variables are integers inside the constructor \elpiIn{uv}, where the
-integer is the index of the current variable wrt a list of optional \elpiIn{tm},
-standing for the substitution mapping of the \ol.
-
-This second encoding of the \ol into our \ml translate \cref{eq:coq1} into
-the term:
-
-\elpiInEq[eq:elpi1]{app[c "p", app[c "f", uv 0], lam x\ app[c "g", x, app[uv 1, x]]].}
-
-This second encoding of the \ol terms is now structured and as a drawback we are
-restricting the unification of the \ol, that is, terms that originally unify
-at the \ol level, do not unify in the \ml.
-
-For example, let \coqIn{a} and \coqIn{b} two defined constants and let's try to
-unify the \ol term
-
-\begin{equation}
-  \texttt{\coqIn{p (f a) (fun x => g x b)}}
-\end{equation}
-
-\noindent corresponding to
-
-\begin{equation}
-  \texttt{\elpiIn{app[c "p", app[c "f", c "a"], lam x\ app[c "g", x, c "b"]].}}
-  \label{eq:elpi2}
-\end{equation}
-
-\noindent with \cref{eq:coq1} (corresponding to \cref{eq:elpi1}). The
-unification of the \ml is able to instantiate \elpiIn{uv 0} (cf \coqIn{X}) to
-\elpiIn{c "a"}, but we are no longer capable to unify the sub-term
-\elpiIn{app[uv 1, x]} (cf \coqIn{F}) with \elpiIn{c "b"}.
-
-The result of this translation of terms inside the \ml causes a certain lack of
-powerfulness while symbolizing higher-order variables. Recall that we are
-considering a \ml capable to deal with higher-order variables, however, the
-sub-term \elpiIn{app[uv 1, x]} is not expressed into the canonical form where a
-higher-order variable of the \ml is in the pattern fragment \cite{miller1991},
-i.e. a variable applied to distinct names. Therefore, we need to preprocess the
-received unification problem $t_1 = t_2$ by 
-% \begin{inlinelist}
-%   \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
-%   \ml
-%   \item finding a valid substitution for $t_1'$ and $t_2'$ 
-%   \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
-%   is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
-% \end{inlinelist}.
-
-\section{Term compilation}
-
-In order to present the the compilation of the \ol terms, so that higher order
-unification can be performed, we need a second and more powerful representation
-of the \ol terms so that variables have a scope. This specification is shown
-in the code snippet below.
-
-\begin{minted}{elpi}
-  kind ml.tm type.
-  type ml.app list ml.tm -> ml.tm.
-  type ml.lam (ml.tm -> ml.tm) -> ml.tm.
-  type ml.c string -> ml.tm.
-  type ml.uv nat -> list ml.tm -> ml.tm.
-\end{minted}
-
-\noindent In particular a \elpiIn{ml.uv} term is meant as a unification variable
-of the meta-language. Therefore, the unification between
-
-\elpiInEqq{ml.lam x\ ml.lam y\ ml.uv 1 [x, y]}
-
-and
-
-\elpiInEqq{ml.lam x\ ml.lam y\ ml.uv 0 [x]}
-
-is supposed to procedure of the substitution \elpiIn{ml.lam x\ ml.lam y\ ml.uv 2
-[x]} for \elpiIn{uv 0} and the substitution \elpiIn{ml.lam x\ ml.uv 2 [x]}) for
-\elpiIn{uv 1}.
-
-Moreover, if \elpiIn{ml.uv} stands for meta-variables, the \elpiIn{app} and the
-\elpiIn{lam} constructors are the nodes for the terms of the \ol. Therefore, we
-cannot claim that \elpiIn{ml.lam x\ ml.app [ml.c "f", x]} and \elpiIn{ml.c "f"}
-unify, since, even though the first is the $\eta$-expansion of the second, the
-\ml does not know how to $\eta\beta$-reduce terms of the \ol.
-
-In our encoding, we explicitly encode the meta-variables with the \elpiIn{ml.uv}
-constructor. This is because we prefer to have the full control of the \ml,
-including the meta-variables instantiation. This way we are able to concretely
-touch the substitution performed by the \ml. In a further section, we show that
-there is no difference between our custom \ml language and any other \ml. Of
-course, a full control on the unification behind meta-variable assmt ask to
-drag the substitution mapping of the \ml and update it each time a variable is
-refined.
-
-The compilation phase is quite straightforward, each constructor of type
-\elpiIn{tm} is mapped to its corresponding version of type \elpiIn{ml.tm}. A
-slight different approach is taken in the case of terms of the form \elpiIn{app
-[uv N | L]}, where the term is translated into \elpiIn{tm.un M L}, that is, a
-new meta-variable \elpiIn{M} with scope \elpiIn{L}.
-
-This latter term transformation is untying the original variable \elpiIn{N} of
-the \ol from the compiled term in the \ml. This means that when \elpiIn{M} is
-instantiated into the \ml, we need to transfer the substitution to the \ol. In
-order to bridge instantiation of meta-variables with the \ol variables, an ad
-hoc \texttt{link} is crafted between the two variables.
-
-A link, \elpiIn{type link nat -> nat -> link}, takes two integers: the first 
-stands for the index of variables in the \ol and the second is the index of 
-the meta-variables.
-
-For example, if we take back the example in \cref{eq:elpi2}, and want to compile
-it, we obtain the new term:
-
-\elpiInEq{app[c "p", app[c "f", c "a"], lam x\ app[c "g", x, c "b"]].}
-
------------------------------ END -----------------------------
-\subsection{First-order unification}
-
-Just as an introduction, we briefly show some small example of unification
-between terms with only first-order unification variables. This way, we would
-like the reader to become familiar between the communication of the two
-languages.
-
-Let's take as an example the following unification problem in the obj. lang.:
-
-\begin{equation}
-  f\ x\ 1 \eqtau f\ Y\ Z
-  \label{eq:unif-fo-coq}
-\end{equation}
-
-\noindent where $f$, $x$ and $1$ are defined constant and $Y$ and $Z$ are both
-unification variables. By convention we use upper case letter for quantified
-variables. Moreover, for this first representation we do not really focus on the
-type of the manipulated objects, since they do not condition the unification
-algorithm.
-
-It is quite evident that a valid substitution for \cref{eq:unif-fo-coq} is $\theta =
-\{Y \mapsto `x', Z \mapsto `1'\}$. Now let's consider the same problem translated in
-the meta language.
-
-\begin{equation}
-  app[`f', `x', `1'] \eqtau app[`f', Y, Z]
-  \label{eq:unif-fo-elpi}
-\end{equation}
-
-The unification of these terms is again quite simple since it is sufficient to
-do a simple matching sub-term by sub-term so that variables can be instantiated.
-We can therefore note that the same substitution $\theta$ will be produced.
-
-\subsection{Higher-order unification}
-
-The unification problem treated before was enough easy to be correctly
-understood by both language representation. We want now to go a bit further and
-reason with a more complex problem where a variable is a function of
-higher-order.
-
-We propose two different higher-order unification problem in the following
-equations where, in the former we have rigid-flexible unification and in the
-latter we have a flexible-flexible unification.
-
-\begin{equation}
-  f\ x\ 1 \eqtau F\ x
-  \label{eq:unif-ho-coq1}
-\end{equation}
-
-\begin{equation}
-  G\ x\ y \eqtau H\ y\ x
-  \label{eq:unif-ho-coq2}
-\end{equation}
-
-The two substitutions for the previous examples are $\theta_1 = \{F \mapsto fun\
-x \Rightarrow f\ x\ 1\}$ and $\theta_2 = \{H \mapsto fun\ y\ x \Rightarrow G\ x\
-y\}$. We can note that to be in the pattern fragment, a functional variable
-should be applied to distinct names.
-
-If we translate the problem before in the meta language, the unification
-problems showed above become
-
-\begin{equation}
-  app[`f',\ `x',\ `1'] \eqtau app[F,\ `x']
-  \label{eq:unif-ho-elpi1}
-\end{equation}
-
-\begin{equation}
-  app[G,\ `x',\ `y'] \eqtau app[H,\ `y',\ `x']
-  \label{eq:unif-ho-elpi2}
-\end{equation}
-
-Now, the new unification problems are no more expressed in the logic of the
-meta language and, therefore, in both cases, unification fails. The
-procedure we can adopt in order to transform a higher-order unification problem
-of the object language into the logic of the meta language is to transform
-the entry of the problem in a problem which can be understood by the meta
-language. The procedure is made of two steps:
-
-\begin{enumerate}
-  \item In the first place, we need to recognize the structure of the pattern
-        fragment expressed in the term received in entry. This means that we
-        need to find all the sub-terms of the form `app[X | L]', where `X' is
-        a flexible variable and `L' is a list of distinct names.
-  \item For any sub-term representing a higher-order unification in the object 
-        language, we build a fresh variable `X\texttt{'}' such that the names 
-        `L' are not in the scope of `X\texttt{'}', we call `X\texttt{'}' the 
-        twin variable of `X'.
-  \item We solve the new goal where each pattern fragment problem is replaced
-        with a problem using twin variables and after each of these problems, we
-        add a new premise linking these twin variables. The linking is done
-        using the following criteria: for each abstraction in the resulting term
-        `X\texttt{'}', unify recursively `X' to a lambda abstraction in the
-        object language.
-\end{enumerate}
-
-The previous algorithm can be applied to \cref{eq:unif-ho-coq1,eq:unif-ho-coq2}
-to provide the wanting solution. In particular, \cref{eq:unif-ho-coq1} is
-transformed into the unification problem:
-
-\begin{equation}
-  f\ x\ 1 \eqtau F'\ x,\ ho-link\ F'\ F
-  \label{eq:unif-ho-elpi1'}
-\end{equation}
-
-\begin{equation}
-  G'\ x\ y \eqtau H'\ y\ x, ho-link\ G'\ G, ho-link\ H'\ H
-  \label{eq:unif-ho-elpi2'}
-\end{equation}
-
-For instance, the former unification problem produce the substitution $\theta_1
-= \{F' \mapsto (x\backslash\ f\ x\ 1)\}$. The ho-link function is then applied
-to transform the substitution of $F'$ into the corresponding term of the object
-langue: $F \mapsto fun \_\ \_\ (x\backslash\ f\ x\ 1)$ which correspond to the
-term $fun\ x \Rightarrow f\ x\ 1$. The latter unification problem gives the
-substitution $\theta_2 = \{H' \mapsto (y\ x \backslash G'\ x\ y)\}$ in the
-meta language. The first ho-link simply unify $G$ to $G'$ since $G'$ is
-flexible, whereas $H$ is mapped to $fun\ y\ x \Rightarrow G'\ x\ y$.
-
-The role of the ho-link is not only to instantiate the higher-order variable F
-of the object language when F is flexible and the twin variable in the meta
-language is rigid. It may happen that F has already been partially instantiated.
-The unification problem below gives such an example in the object language:
-
-\begin{equation}
-  G\ x\ y \eqtau H\ y\ x, H\ x\ y \eqtau x 
-  \label{eq:unif-ho-coq4}
-\end{equation}
-
-\noindent producing the following substitution $\theta = \{G \mapsto (fun\ x\
-y\Rightarrow x); F \mapsto (fun\ x\ y\Rightarrow y)\}$. This unification problem
-is translated into:
-
-\begin{equation}
-  \begin{split}
-    G'\ x\ y &\eqtau H'\ y\ x,\ ho-link\ G'\ G,\ ho-link\ H'\ H\\
-    H''\ x\ y &\eqtau x,\ ho-link\ H''\ H
-  \end{split}
-  \label{eq:unif-ho-elpi4}
-\end{equation}
-
-\noindent The first line produces the same substitution as before: $\theta = \{H'
-\mapsto (x\ y \backslash G'\ y\ x)\}$, where $G'$ is unified with $G$ and $H$ is
-instantiated to <<$fun\ x\ y \Rightarrow G'\ y\ x$>>. While executing the second
-line, we see $H''$ instantiated to the function $(x\ y\backslash x)$ in the
-meta language. The last ho-link is charged to link $H''$ with $H$ but this
-time $H$ has already been partially instantiated. In particular the call to this
-ho-link is as follows: $ho-link\ (x\ y\backslash x)\
-(prod\ \_\ \_\ (x\backslash\ prod\ \_\ \_\ (y\backslash\ app [G', x, y])))$.
-
-Since the two arguments have rigid heads, we start to traverse both terms
-recursively by eating each lambda-abstraction. At the end of this procedure, the
-remaining sub-terms are \todo{in our code this example though the eta mess
-error} now $x$ and $app[G', x, y]$,
-
-\section{HO unification in typed languages}
-
-TODO: ho-link need the type of original term to produce a typed term in the
-object language, example: f x 1 = F x => type of F = (A -> Prof) if type of f is 
-(A -> nat -> Prop)
-
-\section{Proof automation from \coq to \elpi}
-
-TODO: representing a logic programming language into an other: compile rules
-keeping higher order unification
-
-\subsection{Dealing with FO \textit{non-syntactical} unification}
-
-\subsection{Dealing with HO unification}
-
-% \printbibliography
-
-\appendix
-
-\begin{table}
-  \centering
-  \begin{tabular}{c||c|c}
-  Term        & \ol                    & \ml                                   \\
-  \hline\hline
-  Constant    & a                             & `a'                                     \\
-  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
-  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
-  Variable    & X                             & X                                             \\
-  \end{tabular}
-  \caption{\ol terms to \ml terms representation}
-  \label{table:coq2elpi}
-\end{table}
-
-\section{TC in \coq}
-For instance, if
-\texttt{XXX} is the type class representing the AAA, then ZZZ and WWW are
-instances for XXX. In the code snippet below, we give such implementation in
-\coq.
-
-\begin{minted}[escapeinside=&&]{coq}
-  Inductive sig (A : Type) (P : A -> Prop) : Type := ...
-
-  Class Decision (P : Prop) := decide : {P} + {not P}.
-  Class RelDecision {A B: Type} (R : A -> B -> Prop).
-  Class ProofIrrel (A : Type) : Prop := proof_irrel (x y : A) : x = y.
-
-  Instance decide_rel: forall (A B : Type) (R : A -> B -> Prop),
-    RelDecision R -> forall (x : A) (y : B), Decision (R x y). Admitted.
-  Instance True_pi : ProofIrrel True. Admitted.
-  Instance sig_eq_dec: forall (A : Type) (P : A -> Prop),
-    (forall x, ProofIrrel (P x)) -> RelDecision (@eq A) ->
-    RelDecision (@eq (sig A P)). Admitted.
-\end{minted}
-
-This small set of instances after a first phase of\todo{explain compilation of
-pred and inst?} compilation is translated into the following \elpi rules:
-
-\begin{minted}{elpi}
-  type tc-Decision term -> term -> o.
-  type tc-RelDecision term -> term -> term -> term -> o.
-  type tc-ProofIrrel term -> term -> o.
-
-  tc-ProofIrrel (`True`) (`True_pi`).
-  tc-Decision (app [R, X, Y]) 
-    (app [`decide_rel`, A, B, R, P, X, Y]) :-
-    tc-RelDecision A B R P.
-  tc-RelDecision (app [`sig`, A, P]) 
-    (app [`sig`, A, P]) 
-    (app [`eq`, app [`sig`, A, P]]) 
-    (app [`sig_eq_dec`, A, P, P1, P2]) :-
-    pi-decl c0 `x` A =>
-      tc-ProofIrrel (app [P, c0]) (app [P1, c0]), 
-    tc-RelDecision A A (app [`eq`, A]) P2.
-\end{minted}
-
-In this paper we do not really want to explain how the translation of the
-class/instances is performed in our \ml, we prefer to focus our attention on
-unification of terms of the \ol in our \ml. Although, in \cref{table:coq2elpi},
-we provide a simple subset of the typing system used to represent the term of
-the \ol in the \ml.
-
-Type-class resolution starts from a query, that is a class applied to some
-arguments. This \coq term is translated into a term of the \ml and the search
-for a solution in the database is started. However, it may happen that the term
-representation in the \ml may hide some unification properties that are true in
-the \ol. In the example above, the goal \texttt{Decision (@eq T a b)} for some
-\texttt{a} and \texttt{b} unifies with \texttt{Decision (R x y)} in the \ol but
-not in its meta representation. Similarly, the goal \texttt{RelDecision (@eq
-(sig T ?P))} where \texttt{?P}, under the hypothesis \texttt{RelDecision (@eq
-nat)}, will trying to apply the rule for \texttt{sig\_eq\_dec}, we fall into an
-higher order unification problem, where \texttt{P} is applied to the local name
-\texttt{x}. However, the corresponding rule in the \ml exploit a first order
-variable \texttt{P}. Therefore, after the refinement of the goal to
-\texttt{sig\_eq\_dec}, the resolution immediately fail to solve the premise
-\texttt{tc-ProofIrrel (app [P, c0]) (app [P1, c0])}.
-
 \printbibliography

### 2024-02-20 paper/main.tex
tex update

@@
 
+\def\eqfo{eq\_fo\xspace}
+
 In order to reason about unification of the terms of an objet language within a
-meta lanauge, we start by formally describing the two languages. We use
+meta language, we start by formally describing the two languages. We use
 meta-programming to accomplish this result. In \cref{code:lang-descr}, we
-provide on the left the term representation of the object language. %
-% Variables in the OL and ML are represented with integers
-In this encoding, the fo\_tm type stands for first order representation of the
-object language terms. We model unification variables as integers corresponding
-to memory adresses. This memory is represented witht the subst type which is 
-list of optional terms. If the cell $i$ is none, then the variable $i$ is not
-assigned.
+provide on the left the typing structure of the terms of the object language
+together with the signature of the \eqfo function.
+
+In this encoding, the fo\_tm type stands for the first order representation of
+the object language terms. We model unification variables as integers
+corresponding to memory addresses. This memory is represented with the subst
+type abbreviation which is list of optional terms. If the cell $i$ is none, then
+the variable $i$ is not assigned.
 
 % Why the fo prefix for fo_tm && Uvar in OL have no scope
-This encoding is called first order, since variables have no scope. To better
-understand this definition, let's take the subterm \coqIn{P x} from the instance
-forall\_dec. \coqIn{P} has type \coqIn{A ~$\to$~ Prop} and \coqIn{x} is a name
-bound to \coqIn{P}. The application of \coqIn{P} to \coqIn{x} at the OL level,
-put this name in the scope of the variable \coqIn{P}. However, at the meta
-level, the translation of \coqIn{P x} become \elpiIn{app[uv 1, x]}; \elpiIn{x}
-is not in the scope of \elpiIn{uv 1}.
+Because of our translation of the terms of the object language into the meta
+language we have sketched in section one, a object language term has no scope.
+An example of this claim can be given by analyzing deeply the subterm \coqIn{P x} from
+the instance forall\_dec. \coqIn{P} is an higher order variables
+with type \coqIn{A ~$\to$~ Prop} and
+\coqIn{x} is a name bound to \coqIn{P}.
+However, at the meta level, the translation of \coqIn{P x} become \elpiIn{app[uv
+1, x]}, that is, \elpiIn{uv 1} cannot see \elpiIn{x}.
 
-% Uvar in ML have scope
-To solve this problem, we give more expressiveness to the OL variables, by
-giving an alternative term encoding, where unification variables are the
-variables of the meta language. These variables have a scope and therefore
-can be of higher order.
-per la equal, 
-% The subst mappings
-In order to represent variable assignation, we declare the two types subst\_fo
-and subst coressponding to the substitution mapping of unification variable
-for both the OL and the ML.
+Since our goal is to work with higher order variables, we need to compile the
+term received in entry into a more expressive version. This is possible by using
+a second type structure for the terms of the OL. This term definition is given
+in the second column of the \cref{code:lang-descr}. The main different here is
+that a unification variable (always identified with by an integer) has now a
+list of terms representing the scope of the variable. We symbolize the
+abstractions of the ML using the constructor \elpiIn{abs} of the type
+\elpiIn{assmt}. The constructor \elpiIn{val} is is used to contain terms of the
+object language. In this context we do not need to have an application node for
+the ML.
 
@@
 % We have two different equal functions
-Unification over terms is possible thanks to the equal predicate. There exist
+The equality relation over terms under a given substitution mapping is possible 
+thanks to the equal predicate. There exist
 two version of this predicate: \eqfo and \eqho. Their implementation is given in
-\cref{code:term-equal}. The first tests if two terms are equal in the object
-langauge, that is, given a substitution $\theta$, is it true that the two terms
-are equal? Note that no variable instanciation is done: unification is not
-performed by the \elpiIn{equal} predicate. Therefore, a unification variabile
-$i$ equal to another term $t$ if $t$ is a variable with same index $i$, or if
-$i$ is assigend to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover
-\eqfo, is able to verify if two terms are equal up to \etabeta. This equality
-relation is in fact the equality relation of the object langauge and therefore,
-since it accepts \etabeta equality, then \eqfo should take into account these
-reductions.
+\cref{code:term-equal}.
 
-The \eqho predicate test if two terms are equal in the world of the meta
-languages. This eaquality function is not capable to understand the \etabeta
-reductions of the OL. This is mainly due to the fact that the meta language 
-understand \etabeta, but, since the nodes \elpiIn{app} and \elpiIn{lam} do
-not respectively represent the internal application and lambda abstraction
-of the language, the \etabeta reducation is not possible.
+The first tests if two terms are equal in the object language, that is, given a
+substitution $\theta$, is it true that the two terms are equal? Note that no
+variable instantiation is done: unification is not performed by the
+\elpiIn{equal} predicate. Therefore, a unification variable $i$ equals to
+another term $t$ if $t$ is a variable with same index $i$, or if $i$ is assigned
+to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover, since \eqfo 
+represents the equality of the OL, and since the OL accept term equality up to
+\etabeta, then \eqfo also quotient over these reductions.
 
-As we can
+The \eqho predicate tests if two terms are equal in the world of the meta
+languages. This equality function is not capable to understand the \etabeta
+reductions over terms of the object language. The ML sees a term of the type
+\elpiIn{tm} as atom on which only structural equality can be performed. It is
+only possible to dereference a variable if assigned.
 
+\subsection[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}
+
+- how we transform an fo\_tm in tm
+- the role of links
+
+\subsection{Unification in ML}
+
+\subsection{Term de-compilation}
 % Compile FO to HO

### 2024-02-20 paper/main.tex
tex clean

@@
 
-\newcommand{\U}{\ensuremath{=_o}}
-\newcommand{\Ue}{\ensuremath{=_\lambda}}
-\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}}}
-\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}}
+\newcommand{\U}{\ensuremath{=_o}\xspace}
+\newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
+\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
+\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}
 
@@
   Dependent Type Theory (DTT),
-  for which we aim to implement a unification procedure \U{} using the ML
+  for which we aim to implement a unification procedure \U using the ML
   Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-  Elpi comes equipped with the equational theory \Ue{}, comprising
+  Elpi comes equipped with the equational theory \Ue, comprising
   $\eta\beta$ equivalence
@@
   fragment~\cite{miller92jsc}.
-  We want \U{} to feature the same equational theory as \Ue{} but on the
+  We want \U to feature the same equational theory as \Ue but on the
   object logic DTT. Elpi also comes with an encoding for DTT that works well
   for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
-  Unfortunately this encoding, which we refer to as \Fo{},
-  ``underuses'' \Ue{} by restricting it to first-order unification problems only. 
-  To address this issue, we propose a better-behaved encoding, \Ho{},
-  demonstrate how to map unification problems in \Fo{}
-  to related problems in \Ho{}, and illustrate
-  how to map back the unifiers found by \Ue{}, effectively implementing
-   \U{} on top of \Ue{} for the encoding \Fo{}.
+  Unfortunately this encoding, which we refer to as \Fo,
+  ``underuses'' \Ue by restricting it to first-order unification problems only. 
+  To address this issue, we propose a better-behaved encoding, \Ho,
+  demonstrate how to map unification problems in \Fo
+  to related problems in \Ho, and illustrate
+  how to map back the unifiers found by \Ue, effectively implementing
+   \U on top of \Ue for the encoding \Fo.
 
@@
 The encoding of DTT provided by Elpi, that we will discuss at length later in
-section~\ref{sec:encodings} and ~\ref{} features the following term constructors:
+\cref{sec:encodings, ...} features the following term constructors:
 
@@
 \noindent
-This time the root cause is that the unification procedure of \Ue{} of the
-meta language is not aware of the equational theory of the object logic \U{},
+This time the root cause is that the unification procedure of \Ue of the
+meta language is not aware of the equational theory of the object logic \U,
 even if both theories include $\eta\beta$-conversion and admit most general
@@
 In this paper we discuss alternative encodings of Coq in
-Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo{}
+Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
 in which the problems sketched here can be fully described.
-We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho{} (the language of
+We then detail an encoding \elpiIn{comp} from \Fo to \Ho (the language of
 the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
@@
 Because of our translation of the terms of the object language into the meta
-language we have sketched in section one, a object language term has no scope.
+language we have sketched in section 1, a object language term has no scope.
 An example of this claim can be given by analyzing deeply the subterm \coqIn{P x} from
@@
 only possible to dereference a variable if assigned.
+\marginpar{TODO: say HO is same as ML}
 
@@
 
+- we accept HO unif with PF
+
 \subsection{Term de-compilation}
@@
 
-le due sintassi, XX tradotto, unif passa, bisogna riportare la soluzione
-nel mondo fo
-
-this should hold in both encoding
-%abella
-\begin{elpicode}
-forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
-\end{elpicode}
-we have unif only in the HO one, where equal is
-weaker (but unif is stronger)
-
-segnatura di decomp e unif, equalfo (fa beta, eta, deref)
-
 \noindent
-
-equal fo is congruence, deref, beta, eta (for the OL).
-equal ho is congruence, deref (for the OL).
-unif (fo or ho) should cover pattern fragment, i.e. well behaved unif.
-
-ho, and unif ho, is the LF/ML language/unif. here we describe fo and ho
-with a deeply embedded syntax, but the idea is that uv can be mapped to
-an actual unif variable.
-
-of course we could write a unif for the ho that had beta and eta,
-but if we use the one of the LF it can't have eta beta of the OL baked in.
 
@@
 ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.
-
-\begin{elpicode}
-prove (app[c "decidable", all x\ P x]) Proof :- ho-link P P', prove (... P' ...).
-\end{elpicode}
-
-qui P nel paper diventa uv N [x] per un certo N.
-
-\section{recovering HO}
-
-\begin{elpicode}
-  :index (1)
-  p (all x\ app[F,x]).
-\end{elpicode}
-
-fallisce perche le liste non hanno la stessa lunghezza.
-
-compile
-
-\begin{elpicode}
-  p (all x\ F' x) = p (all x\ app[f,x,x]) , link F F'.
-  F' = x\ app[f,x,x]
-  F = lam a\ app[f,a,a].
-\end{elpicode}
-
-\begin{elpicode}
-  type comp fo.tm -> ho.tm -> list link -> list link -> ho.subst -> ho.subst -> o.
-  comp (fo.c X)   (ho.c X) L L S S.
-  comp (fo.app [fo.uv N|Argsss]) TT L L3 S S3 :- %!,
-    % TODO: here split-pf to enter dist. names into N
-    % split-pf Argsss [] PF NPF,
-    split-pf Argsss [] PF [], NPF = [], % TODO: compile to (uv N L) + link-fo-app
-    print "In PF" PF NPF, 
-    if (NPF = []) (TT = ho.uv M PF1) (TT = ho.app [ho.uv M PF1 | NPF1]),
-    % pattern-fragment Args,
-    fold4 comp PF PF1 L L1 S S1,
-    fold4 comp NPF NPF1 L1 L2 S1 S2,
-    ho.new S2 M S3,
-    % TODO: maybe len can be given by split-pf
-    len PF Len,
-    L3 = [link N M Len| L2].
-  % TODO: if don't want to modify unif, we compile `fo.app [fo.c f, c0, c0, ho.c a]` into 
-  % `ho.app [ho.app[f, c0, c0], ho.c a]
-  comp (fo.app A) (ho.app A1) L L1 S S1 :- fold4 comp A A1 L L1 S S1.
-  comp (fo.lam F) (ho.lam F1) L L1 S S1 :- 
-    (pi x y\ (pi A S\ comp x y A A S S) => comp (F x) (F1 y) L L1 S S1).
-  comp (fo.uv N) (ho.uv M []) L [link N M z|L] S S1 :- ho.new S M S1.
-
-  decmop...
-\end{elpicode}
 

### 2024-02-21 paper/main.tex
minor changes
update tex

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 The encoding of DTT provided by Elpi, that we will discuss at length later in
-\cref{sec:encodings, ...} features the following term constructors:
+\cref{sec:encodings,sec:lang-spec}, features the following term constructors:
 
@@
 
-\section{Lanuages description}
+\section{Languages description}
+\label{sec:lang-spec}
 
@@
 In order to reason about unification of the terms of an objet language within a
-meta language, we start by formally describing the two languages. We use
-meta-programming to accomplish this result. In \cref{code:lang-descr}, we
-provide on the left the typing structure of the terms of the object language
-together with the signature of the \eqfo function.
+meta language, we start by formally describing the two languages. Employing
+meta-programming for this purpose, \cref{code:lang-descr} presents, on the left,
+the typing structure of object language terms along with the signature of the
+\eqfo function.
 
@@
 the object language terms. We model unification variables as integers
-corresponding to memory addresses. This memory is represented with the subst
-type abbreviation which is list of optional terms. If the cell $i$ is none, then
-the variable $i$ is not assigned.
+corresponding to memory addresses. The memory is represented as list of optional
+terms (the subst\_fo type abbreviation). If the cell $i$ is none, then the
+variable $i$ is not instantiated, otherwise it has already been assigned to some
+value.
 
@@
 Because of our translation of the terms of the object language into the meta
-language we have sketched in section 1, a object language term has no scope.
-An example of this claim can be given by analyzing deeply the subterm \coqIn{P x} from
-the instance forall\_dec. \coqIn{P} is an higher order variables
+language outlined in \cref{sec:intro}, the variables of the object language have 
+no scope.
+An illustration of this observation can be found by analyzing deeply the subterm \coqIn{P x} from
+the instance forall\_dec. Here, \coqIn{P} is an higher order variables
 with type \coqIn{A ~$\to$~ Prop} and
 \coqIn{x} is a name bound to \coqIn{P}.
-However, at the meta level, the translation of \coqIn{P x} become \elpiIn{app[uv
-1, x]}, that is, \elpiIn{uv 1} cannot see \elpiIn{x}.
+However, at the meta level, the translation of \coqIn{P x} becomes \elpiIn{app[uv
+1, x]}, that is, \elpiIn{uv 1} cannot reference \elpiIn{x}.
 
-Since our goal is to work with higher order variables, we need to compile the
-term received in entry into a more expressive version. This is possible by using
-a second type structure for the terms of the OL. This term definition is given
-in the second column of the \cref{code:lang-descr}. The main different here is
-that a unification variable (always identified with by an integer) has now a
-list of terms representing the scope of the variable. We symbolize the
-abstractions of the ML using the constructor \elpiIn{abs} of the type
-\elpiIn{assmt}. The constructor \elpiIn{val} is is used to contain terms of the
-object language. In this context we do not need to have an application node for
-the ML.
+As our objective is to handle higher-order variables, a viable solution involves
+addressing this issue through the compilation of the term received as input into
+a more expressive version. This can be achieved by employing a second type
+structure for the terms of the OL, as outlined in the second column of
+\cref{code:lang-descr}. The primary difference lies in the fact that a
+unification variable (always identified by an integer) now takes a list of terms
+representing its scope. The abstractions of the ML are
+symbolized using the constructor \elpiIn{abs} of the type \elpiIn{assmt}. The
+constructor \elpiIn{val} is utilized to contain terms of the object language,
+and in this context, there is no need for an application node in the ML.
 

### 2024-02-21 paper/main.tex
update tex

@@
 
-In order to reason about unification of the terms of an objet language within a
+In order to reason about unification of the terms of the objet language within the
 meta language, we start by formally describing the two languages. Employing
-meta-programming for this purpose, \cref{code:lang-descr} presents, on the left,
-the typing structure of object language terms along with the signature of the
-\eqfo function.
+meta-programming for this purpose, \cref{code:lang-descr} presents the type
+\elpiIn{tm} containing the constructors for term application, lambda abstraction
+and constants. Moreover, in order to represent unification variables, we need 
+to give two different constructors.
+
+In the case of the OL, variables have no scope. For example, the subterm
+\coqIn{P x} from the instance forall\_dec. Here, \coqIn{P} is an higher order
+variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} is a name bound to
+\coqIn{P}. However, at the meta level, the translation of \coqIn{P x} becomes
+\elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
+encoding, unification variables are encoded as integer corresponding to memory
+addresses, and the constructor for unification variables in the OL is
+\elpiIn{fo_uv}. This constructor is prefixed with \elpiIn{fo} since the
+variable, having no scope, is a first-order variable.
+
+On the other hand, in the case of the ML, we want to use its unification
+algorithm to make variable assignment. Since the ML is an Higher Order
+Programming Language, we represent unification variables with the \elpiIn{uv}
+constructor, which, this time, can see a list of terms. Of course some attention
+should be payed when dealing with this constructor, since we have to certify
+each time that an \elpiIn{uv i} remain in the pattern fragment, that is, the
+list of term in the scope of \elpiIn{i} is a list of distinct names. Finally,
+the following code
+
+\begin{elpicode}
+  kind assmt type.
+  type abs (tm -> assmt) -> assmt.
+  type val tm -> assmt.
+\end{elpicode}
+
+\noindent
+illustrates the \elpiIn{assmt} representing variable assignment in the ML.
+
+Variable assignment
+
+\vspace{30pt}
 
@@
 
-\subsection[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}
+% \section[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}
 
-- how we transform an fo\_tm in tm
-- the role of links
+% - how we transform an fo\_tm in tm \\
+% - the role of links \\
+% - decomp
+% - esempio che va in questa semplice rappresentazione (from intro)
+% - esempio che non va, multi-var, eta, beta
+% % \section{Term de-compilation}
 
-\subsection{Unification in ML}
+% \section{Unification in ML}
 
-- we accept HO unif with PF
+% - we accept HO unif with PF \\
+% - need of multiple vars for a single OL var
 
-\subsection{Term de-compilation}
-% Compile FO to HO
+% % Compile FO to HO
 
-% Assignement in HO should be decompiled
+% % Assignement in HO should be decompiled
 
--------------------------------------------------------------------
+% -------------------------------------------------------------------
 
-\noindent
+% \noindent
 
-\subsection{implementation}
+% \subsection{implementation}
 
-l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
-in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
-elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
-ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.
+% l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
+% in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
+% elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
+% ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.
 
-\section{recovering eta}
+% \section{recovering eta}
 
-\begin{elpicode}
-q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
-F = fun a => app [f,a] ----> F = f
-\end{elpicode}
+% \begin{elpicode}
+% q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
+% F = fun a => app [f,a] ----> F = f
+% \end{elpicode}
 
-l'utene da p su f, mentre l'istanza pe q forza F a fun ..
+% l'utene da p su f, mentre l'istanza pe q forza F a fun ..
 
-\section{recovering beta}
+% \section{recovering beta}
 
-\begin{elpicode}
-  q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
-  F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
-\end{elpicode}
+% \begin{elpicode}
+%   q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
+%   F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
+% \end{elpicode}
 
-qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
-p1 F1, e decomp beta F [a] F1.
+% qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
+% p1 F1, e decomp beta F [a] F1.
 
-\section{recovering eta-beta within unification (non linear variables)}
+% \section{recovering eta-beta within unification (non linear variables)}
 
-se i problemi di cui sopra avvengono nello stesso termine
+% se i problemi di cui sopra avvengono nello stesso termine
 
-\begin{elpicode}
-  q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
-\end{elpicode}
+% \begin{elpicode}
+%   q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
+% \end{elpicode}
 
-bisogna slegare le due F e poi unificare le soluzioni tra di loro
+% bisogna slegare le due F e poi unificare le soluzioni tra di loro
 
-\section{heuristic / binary app}
+% \section{heuristic / binary app}
 
-fo approx / sub pattern fragment
+% fo approx / sub pattern fragment
 
-\begin{elpicode}
-  p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
-  p (all x\G x) F' = 
-  G = x\ f x x a
-  F = lam x\f x x
-  F' = (app[f,b,b])
-  link (F a) F'
-  link G F
-\end{elpicode}
+% \begin{elpicode}
+%   p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
+%   p (all x\G x) F' = 
+%   G = x\ f x x a
+%   F = lam x\f x x
+%   F' = (app[f,b,b])
+%   link (F a) F'
+%   link G F
+% \end{elpicode}
 
-% questo esempio recupera l'app binaria perche
+% % questo esempio recupera l'app binaria perche
 
-\begin{elpicode}
-  (app (app F x) a) = (app (f x x) a)
-\end{elpicode}
+% \begin{elpicode}
+%   (app (app F x) a) = (app (f x x) a)
+% \end{elpicode}
 
-\printbibliography
+% \section{Stats}
+
+% Concretamente:
+
+% - quante volte l'ordine superiore è necessario in stdpp/tlc
+
+% \section{Conclusion}
+
+% \printbibliography
 

### 2024-02-21 paper/main.tex
update tex

@@
 \noindent
-illustrates the \elpiIn{assmt} representing variable assignment in the ML.
+illustrates the \elpiIn{assmt} representing variable assignment in the ML. \marginpar{todo: explain better}
 
-Variable assignment
+The memory of the two languages are represented with lists of substitutions. In
+particular, the substitution of the OL, called \elpiIn{fo_subst}, is made by
+optional terms such that, if the substitution is \elpiIn{none}, then the
+variable is not instantiated. Note that the variables in the \elpiIn{fo_subst}
+have always the \elpiIn{fo_uv} constructor. On the other hand, the ML
+substitution is an optional assignment and in that assignment, variables are
+considered to have the \elpiIn{uv} constructor.
 
-\vspace{30pt}
+A key property needed in unification is being able to verify if two terms are
+equal. This is kind of a structural equality verification between two terms,
+where variable dereferencing is performed when the variable is assigned. A
+sketch of the equality function is given for the OL language in
+\cref{code:term-equal}. Though, this equality relation over terms of a language
+can be powered by other reduction rules depending equational theory being
+considered. In our case, the OL terms are equal under $\eta\beta$ redex. This mean
+that new rules for those two redexes are added in the implementation of
+\elpiIn{fo_equal}.
 
-In this encoding, the fo\_tm type stands for the first order representation of
-the object language terms. We model unification variables as integers
-corresponding to memory addresses. The memory is represented as list of optional
-terms (the subst\_fo type abbreviation). If the cell $i$ is none, then the
-variable $i$ is not instantiated, otherwise it has already been assigned to some
-value.
+If \elpiIn{fo_equal} is conceived to manage equality between terms of the OL,
+the same equality predicate in the ML behave slightly different. By the given
+definition of the ML, the ML allows $\eta\beta$ congruence of terms, but, since
+the node \elpiIn{app} and \elpiIn{abs} are constructor representing the
+applications and the abstractions of the OL, these two reduction rules cannot
+applied on them. We build therefore a predicate \elpiIn{equal} working for terms
+in the ML which is implemented merely with the rules for the \elpiIn{fo_equal}
+predicate.\todo{not true for uv} For example, if \elpiIn{fo_equal []
+(abs x\ [c"f", c]) (c"f")} is true in the object language, \elpiIn{equal [] (abs
+x\ [c"f", c]) (c"f")} produces a failure.
 
-% Why the fo prefix for fo_tm && Uvar in OL have no scope
-Because of our translation of the terms of the object language into the meta
-language outlined in \cref{sec:intro}, the variables of the object language have 
-no scope.
-An illustration of this observation can be found by analyzing deeply the subterm \coqIn{P x} from
-the instance forall\_dec. Here, \coqIn{P} is an higher order variables
-with type \coqIn{A ~$\to$~ Prop} and
-\coqIn{x} is a name bound to \coqIn{P}.
-However, at the meta level, the translation of \coqIn{P x} becomes \elpiIn{app[uv
-1, x]}, that is, \elpiIn{uv 1} cannot reference \elpiIn{x}.
+The solution we are proposing aim to overcome these unification issues by 1)
+compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
+in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
+variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
+the variables of the OL with the decompiled version of their corresponding meta
+variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
+and that the substitution in the object language is the same as the one returned
+by the ML. \todo{same or $\supseteq$ or $\subseteq$}
 
-As our objective is to handle higher-order variables, a viable solution involves
-addressing this issue through the compilation of the term received as input into
-a more expressive version. This can be achieved by employing a second type
-structure for the terms of the OL, as outlined in the second column of
-\cref{code:lang-descr}. The primary difference lies in the fact that a
-unification variable (always identified by an integer) now takes a list of terms
-representing its scope. The abstractions of the ML are
-symbolized using the constructor \elpiIn{abs} of the type \elpiIn{assmt}. The
-constructor \elpiIn{val} is utilized to contain terms of the object language,
-and in this context, there is no need for an application node in the ML.
+Mathematically, we what to prove the following property:
 
-\def\eqfo{equal\_fo\xspace}
-\def\eqho{equal\_ho\xspace}
-\def\etabeta{$\eta\beta$\xspace}
+\begin{coqcode}
+  forall (t  u : term_ol) (s : subst_ol) 
+         (t' u': term_ml) (s': subst_ml), 
+    equal_ol t u s <->
+      comp t t' /\ comp u u' /\ 
+      unif_ml t' u' s' /\ decomp s' s.
+\end{coqcode}
 
-% We have two different equal functions
-The equality relation over terms under a given substitution mapping is possible 
-thanks to the equal predicate. There exist
-two version of this predicate: \eqfo and \eqho. Their implementation is given in
-\cref{code:term-equal}.
+$$Math\ formula$$
 
-The first tests if two terms are equal in the object language, that is, given a
-substitution $\theta$, is it true that the two terms are equal? Note that no
-variable instantiation is done: unification is not performed by the
-\elpiIn{equal} predicate. Therefore, a unification variable $i$ equals to
-another term $t$ if $t$ is a variable with same index $i$, or if $i$ is assigned
-to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover, since \eqfo 
-represents the equality of the OL, and since the OL accept term equality up to
-\etabeta, then \eqfo also quotient over these reductions.
+In the following section we explain how we deal with term (de)compilation and
+unification variable linking.
 
-The \eqho predicate tests if two terms are equal in the world of the meta
-languages. This equality function is not capable to understand the \etabeta
-reductions over terms of the object language. The ML sees a term of the type
-\elpiIn{tm} as atom on which only structural equality can be performed. It is
-only possible to dereference a variable if assigned.
-\marginpar{TODO: say HO is same as ML}
-
-% \section[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}
+\section[Compilation: fo\_tm to tm]{Compilation: \texttt{tm} $\to$ \texttt{tm}}
 

### 2024-02-21 paper/main.tex
wip intro

@@
 \newcommand{\U}{\ensuremath{=_o}\xspace}
+\newcommand{\nU}{\ensuremath{\neq_o}\xspace}
 \newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
+\newcommand{\nUe}{\ensuremath{\neq_\lambda}\xspace}
 \newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
@@
 \begin{coqcode}
-Instance fin_fin n : Finite (fin n).
-Instance nfact_dec n nf : Decision (nfact n nf).
-Instance forall_dec A P : Finite A ~$\to$~ 
-  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
+Instance fin_fin n : Finite (fin n).             (* r1 *)
+Instance nfact_dec n nf : Decision (nfact n nf). (* r2 *)
+Instance forall_dec A P : Finite A ~$\to$~            (* r3 *)
+  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
 \end{coqcode}
 
-\noindent Under this context the type-class solver is able to prove
-the the following statement automatically by back-chaining
-the three instances.
+\noindent Under this context of instances a type-class solver is able to prove
+the the following statement automatically by back-chaining.
 
 \begin{coqcode}
-  Check _ : Decision (forall y: fin 7, nfact y 3).
+  Check _ : Decision (forall y: fin 7, nfact y 3).  (* g *)
 \end{coqcode}
@@
 The encoding of DTT provided by Elpi, that we will discuss at length later in
-\cref{sec:encodings,sec:lang-spec}, features the following term constructors:
+section~\ref{sec:encodings,sec:lang-spec} and ~\ref{}, is an Higher Order Abstract
+Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
+constructors:
 
 \begin{elpicode}
-kind tm type.
 type lam  tm -> (tm -> tm) -> tm.     % lambda abstraction
@@
 \noindent
-\marginpar{TODO: explain HOAS}
-Following this term encoding the three instances are represented by the
-following rules:
+Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
+the concrete syntax to abstract, at the meta level, an expression
+\elpiIn{e} over a variable \elpiIn{x}
+is \elpiIn{x\ e}, and square brackets denote a list of
+terms separated by comma. As an example we show the encoding of the Coq term
+``$\forall y:\mathrm{t}.\mathrm{nfact}~ y~ 3$'':
 
 \begin{elpicode}
-finite (app[c"fin", N]).
-decision (app [c"nfact", N, NF]).
-decision (all A x\ app[P, x]) :- finite A,
+all (c"t") y\ app[c"nfact", y, c"3"]
+\end{elpicode}
+
+\noindent
+We now illustrate the encoding of the three instances above as higher order
+logic programming rules: capital letters denote rule
+parameters; \elpiIn{:-} separates the rule's head from the premises;
+\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
+for the premise \elpiIn{p}.
+
+\begin{elpicode}
+finite (app[c"fin", N]).                             % r1
+decision (app [c"nfact", N, NF]).                    % r2
+decision (all A x\ app[P, x]) :- finite A,           % r3
+  pi w\ decision (app[P, w]).
+\end{elpicode}
+
+\noindent
+Unfortunately this direct translation of rule (r3) uses the
+predicate \coqIn{P} essentially as a first order term for the meta
+language (its type is \elpiIn{tm}).
+If we try to backchain the rule (r3) on the encoding of the goal (g) above:
+
+\begin{elpicode}
+decision (all (app[c"fin", c"7"]) y\                 % g
+  app[c"nfact", y, c"3"]).
+\end{elpicode}
+
+\noindent
+we fail because of this ``higher order'' unification problem (for DTT)
+is phrased as a first order unification problem in the meta language:
+the two lists of terms have different lengths!
+
+\begin{elpicode}
+app[c"nfact", y, c"3"] = app[P, y]                   % p
+\end{elpicode}
+
+\noindent
+In this paper we study a more sophisticated encoding of Coq terms allowing
+us to rephrase the problematic rule (r3) as follows:
+
+\begin{elpicode}
+decision (all A x\ Pm x) :- link Pm A P, finite A,   % r3a
   pi x\ decision (app[P, x]).
@@
 \noindent
-\marginpar{TODO: explain pi, cons}
-Unfortunately this direct translation of the instances considers the
-predicate \coqIn{P} as a first order term. If we try to backchain the
-third rule on the encoding of the goal above:
+Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
+in its scope, the unification problem (pa) admits one solution:
 
 \begin{elpicode}
-decision (all (app[c"fin", c"7"]) y\
-  app[c"nfact", y, c"3"]).
+app[c"nfact", y, c"3"] = Pm y                       % pa
+Pm = x\ app[c"nfact", x, c"3"]       % assignment for Pm
+A = app[c"fin", c"7"]                % assignment for A
 \end{elpicode}
@@
 \noindent
-we fail because of this ``higher order'' unification problem (in DTT)
-is phrased as a first order unification problem in the meta language.
-
-\begin{elpicode}
-app[c"nfact", y, c"3"] = app[P, y]
-\end{elpicode}
-
-\noindent
-In this paper we study a more sophisticated encoding of Coq terms allowing
-us to rephrase the problematic rule as follows:
-
-\begin{elpicode}
-decision (all A x\ Pm x) :- link Pm A P, finite A,
-  pi x\ decision (app[P, x]).
-\end{elpicode}
-
-\noindent
-This time \elpiIn{Pm} is an higher order unification variable (of type
-\elpiIn{tm -> tm}). The resulting unification problem is now:
-
-\begin{elpicode}
-app[c"nfact", y, c"3"] = Pm y
-\end{elpicode}
-
-\noindent
-That admits one solution:
-
-\begin{elpicode}
-Pm = y\ app[c"nfact", y, c"3"]
-A = app[c"fin",c"7"]
-\end{elpicode}
-
-\noindent
-Elpi succeeds in the application of the new rule and then runs
+After unifying the head of rule (r3a) with the goal Elpi runs
 the premise \elpiIn{link Pm A P} that is in charge of bringing the
-assignment back to the domain of Coq terms (the type \elpiIn{tm}):
+assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
+back to the domain of Coq terms (the type \elpiIn{tm}):
 
@@
 This simple example is sufficient to show that the encoding we seek
-is not trivial. Indeed the solution for \elpiIn{P} generates a
-(Coq) $\beta$-redex in the second premise (under the \elpiIn{pi x}):
+is not trivial, since the solution for \elpiIn{P} above generates a
+(Coq) $\beta$-redex in the second premise (the predicate
+under the \elpiIn{pi w\ }\hspace{-0.8em}):
 
 \begin{elpicode}
-decision (app[lam A (a\ app[c"nfact", a, c"3"]), x])
+decision (app[lam A (a\ app[c"nfact", a, c"3"]), w])
 \end{elpicode}
@@
 \noindent
-In turn the redex prevents the second rule to backchain properly since
+In turn this redex prevents the rule (r2) to backchain properly since
 the following unification problem has no solution:
@@
 \noindent
-This time the root cause is that the unification procedure of \Ue of the
-meta language is not aware of the equational theory of the object logic \U,
-even if both theories include $\eta\beta$-conversion and admit most general
+The root cause of the problems we face is that the unification procedure
+\Ue of the meta language is not aware of the equational theory of the object
+logic \U, even if both theories include $\eta\beta$-conversion and admit most general
 unifiers for problems in the pattern fragment~\cite{miller92jsc}.
 
+\paragraph{Contributions}
 In this paper we discuss alternative encodings of Coq in
@@
 the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
-bla bla..
+bla bla.. TODO citare Teyjus.
 
-\section{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
+\section{Problem statement and alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
 \label{sec:encodings}
+
+The equational theory of Coq's Dependent Type Theory is very rich. In
+addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
+are compared up to definition unfolding and fixpoint unrolling. Still,
+for efficiency and predictability reasons, most form of automatic proof search
+employ a unification procedure that captures a simpler one,
+just $\eta\beta$, and that solves higher order problems
+restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
+We call this unification procedure \U{}.
+
+The equational theory of the meta language Elpi that we want to use to
+implement a form of proof automation is strikingly similar, since it
+it comprises $\eta\beta$ (for the meta language functions), and the
+unification procedure \Ue{} solves higher order problems in
+$\mathcal{L}_\lambda$.
+
+In spite of the similarity the link between \Ue{} and \U{} is not trivial,
+since the abstraction and application term constructors
+the two unification procedure deal with are different. For example
+
+\begin{tabular}{lcl}
+\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
+\elpiIn{lam A x\ app[c"f", x]} & \U{} & \elpiIn{c"f"}\\
+\elpiIn{lam A x\ app[c"f", x]} & \nUe{} & \elpiIn{c"f"} \\
+\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
+\elpiIn{app[P, x]} & \U{} & \elpiIn{x}\\
+\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
+\end{tabular}
+
+\noindent
+One could ignore this similarity, and ``just'' describe the object language
+unification procedure in the meta language, that is crafting a \elpiIn{unif}
+predicate to be used as follows in rule (r3):
+
+\begin{elpicode}
+decision X :- unif X (all A x\ app[P, x]), finite A,
+  pi x\ decision (app[P, x]).
+\end{elpicode}
+
+\noindent
+This choice would underuse the logic programming engine provided by
+the metalanguage since by removing any datum from the head of rules
+indexing degenerates. Moreover the unification procedure built in the
+meta language is likely to be faster than one implemented in it,
+especially if the meta language is interpreted as Elpi is.
+
+To state precisely the problem we solve we need a \Fo{} representation
+of DTT terms, a \Ho{} one. Then for any problem $s_1 \U{} s_2$ for
+$s_i$ in \Fo{}, we get a  $t_1 \Ue{} t_2$ for $s_i$ in \Fo{} and
+a $L = \bigwedge_i \mathrm{link}~ s_i~ t_i $ such that:
+
+\begin{align*}
+s_1 \U{} s_2 & \Leftrightarrow  t_1 \Ue{} t_2 \land L \\
+s_1 \nU{} s_2 & \Rightarrow t_1 \nUe{} t_2 \lor \lnot L
+\end{align*}
+
+where $\mathrm{comp}~t_i = (s_i, \mathrm{link}~ s_i~ t_i)$ and 
+link is a predicate allowed to use \U{} on ground terms (hence equal, not unif)
+
+These properties allow us to simulate a unification based
+backward search on DTT by using \Ue{}. In particular
+any unification that would work in DTT will work,
+and whenever a unification in DTT would have failed
+either the corresponding \Ue{} or the link fails.
+
+This grants that the trace of the logic program
+performing the search not only gives the same result, but
+also takes the same paths, that is it fails as early as possible.
+
+\subsection{Alternative encodings and related work}
 
@@
 the c, app and lam nodes. % all ?
-
-Another alternative
-
-\begin{elpicode}
-decision X :- unif X (all A x\ app[P, x]), 
-  (pi x\ decision (app[P, x])), finite A.
-\end{elpicode}
-
-gives up all half of what the ML gives us. Moreover even if unif here embodies
-the eq theory of DTT which is much stronger than the one of the ML, we don't need 
-it. According to our experience eta beta suffice, but HO is needed.
 

### 2024-02-21 paper/main.tex
update tex

@@
 
-\section[Compilation: fo\_tm to tm]{Compilation: \texttt{tm} $\to$ \texttt{tm}}
+\section[Compilation: fo\_tm to tm]{Compilation}
 
-% - how we transform an fo\_tm in tm \\
-% - the role of links \\
-% - decomp
-% - esempio che va in questa semplice rappresentazione (from intro)
-% - esempio che non va, multi-var, eta, beta
-% % \section{Term de-compilation}
+The compilation step is meant to recover the higher-order variables of the OL,
+expressed in a first order way, by replacing them with higher-order variables in
+the ML. In particular, every time a variable of the OL is encountered in the
+original term, it is replaced with a meta variable, and if the OL variable is
+applied to a list $L$ of terms, then this list becomes the scope of the variable
+if $L$ is a list of distinct names. For all the other constructors of
+\elpiIn{tm}, the same term constructor is returned and its arguments are
+recursively compiled. The predicate in charge for term compilation is:\\
+\elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
+Where, we take the term of the OL, produce the term of the ML, take a list
+of link and produce a list of new links, take a substitution and return a 
+new substitution.
 
-% \section{Unification in ML}
+In particular, due to programming constraints, we need to drag the old subst and
+return a new one extended, if needed, with the new declared meta-variables.
 
-% - we accept HO unif with PF \\
-% - need of multiple vars for a single OL var
+The following code
+\begin{elpicode}
+  kind link type.
+  type link nat -> nat -> nat -> subst.
+\end{elpicode}
 
-% % Compile FO to HO
+\noindent
+defines a link, which is a relation between to variables indexes, the first
+being the index of a OL variable and the second being the index of a ML
+variable. The third integer\todo{integer or nat?} is the number of term in the
+scope of the two variables, or equivalently, in a typed language, their arity.
 
-% % Assignement in HO should be decompiled
+\noindent
+- how we transform an fo\_tm in tm \\
+- the role of links \\
+- decomp
+- esempio che va in questa semplice rappresentazione (from intro)
+- esempio che non va, multi-var, eta, beta
+% \section{Term de-compilation}
+
+\section{Unification in ML}
+
+- we accept HO unif with PF \\
+- need of multiple vars for a single OL var
+
+% Compile FO to HO
+
+% Assignement in HO should be decompiled
 

### 2024-02-22 paper/main.tex
reduce space before equation
wip

@@
 
-\newcommand{\U}{\ensuremath{=_o}\xspace}
-\newcommand{\nU}{\ensuremath{\neq_o}\xspace}
-\newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
-\newcommand{\nUe}{\ensuremath{\neq_\lambda}\xspace}
-\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
+\newcommand{\EqualRel}{\ensuremath{=}}
+\newcommand{\nEqualRel}{\ensuremath{\new}}
+\newcommand{\UnifRel}{\ensuremath{\simeq}}
+\newcommand{\nUnifRel}{\ensuremath{\not\simeq}}
+
+\newcommand{\Uo}{\ensuremath{\UnifRel_o}\xspace}
+\newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
+\newcommand{\Eo}{\ensuremath{\EqualRel_o}\xspace}
+\newcommand{\nEo}{\ensuremath{\nEqualRel_o}\xspace}
+
+\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
+\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
+\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
+\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
+
+\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}} % space non va
 \newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}
@@
   Dependent Type Theory (DTT),
-  for which we aim to implement a unification procedure \U using the ML
+  for which we aim to implement a unification procedure \Uo using the ML
   Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-  Elpi comes equipped with the equational theory \Ue, comprising
-  $\eta\beta$ equivalence
-  and higher order unification restricted to the pattern
+  Elpi's equational theory comprises
+  $\eta\beta$ equivalence and comes equipped with a
+  higher order unification procedure \Ue restricted to the pattern
   fragment~\cite{miller92jsc}.
-  We want \U to feature the same equational theory as \Ue but on the
-  object logic DTT. Elpi also comes with an encoding for DTT that works well
+  We want \Uo to be as powerful as \Ue but on the object logic DTT.
+  Elpi also comes with an encoding for DTT that works well
   for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
@@
   how to map back the unifiers found by \Ue, effectively implementing
-   \U on top of \Ue for the encoding \Fo.
+   \Uo on top of \Ue for the encoding \Fo.
 
@@
 \Ue of the meta language is not aware of the equational theory of the object
-logic \U, even if both theories include $\eta\beta$-conversion and admit most general
+logic \Uo, even if both theories include $\eta\beta$-conversion and admit most general
 unifiers for problems in the pattern fragment~\cite{miller92jsc}.
@@
 restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
-We call this unification procedure \U{}.
+We call this unification procedure \Uo{}.
 
@@
 
-In spite of the similarity the link between \Ue{} and \U{} is not trivial,
+In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
 since the abstraction and application term constructors
@@
 \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
-\elpiIn{lam A x\ app[c"f", x]} & \U{} & \elpiIn{c"f"}\\
+\elpiIn{lam A x\ app[c"f", x]} & \Uo{} & \elpiIn{c"f"}\\
 \elpiIn{lam A x\ app[c"f", x]} & \nUe{} & \elpiIn{c"f"} \\
 \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \U{} & \elpiIn{x}\\
+\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
 \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
@@
 
-To state precisely the problem we solve we need a \Fo{} representation
-of DTT terms, a \Ho{} one. Then for any problem $s_1 \U{} s_2$ for
-$s_i$ in \Fo{}, we get a  $t_1 \Ue{} t_2$ for $s_i$ in \Fo{} and
-a $L = \bigwedge_i \mathrm{link}~ s_i~ t_i $ such that:
+To state precisely the problem we solve we need a \Fo representation
+of DTT terms and a \Ho one.
+We call \Eo the equality over ground terms in \Fo,
+\Ee the equality over ground terms in \Ho,
+\Uo the unification procedure we want to implement and
+\Ue the one provided by the meta language.
+TODO extend \Eo and \Ee with reflexivity on uvars.
 
-\begin{align*}
-s_1 \U{} s_2 & \Leftrightarrow  t_1 \Ue{} t_2 \land L \\
-s_1 \nU{} s_2 & \Rightarrow t_1 \nUe{} t_2 \lor \lnot L
-\end{align*}
+We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
+$\sigma$, we write $\sigma t$ for the application of the substitution to
+$t$ and we assume that the unification of our meta language is correct:
 
-where $\mathrm{comp}~t_i = (s_i, \mathrm{link}~ s_i~ t_i)$ and 
-link is a predicate allowed to use \U{} on ground terms (hence equal, not unif)
+$$
+t_1 \Ue t_2 \mapsto \sigma \Rightarrow \sigma t_1 \Ee \sigma t_2
+$$
+
+\newcommand{\C}[3]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3)}
+\newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}
+
+We illustrate a compilation $\C{s}{t}{l}$ that
+maps a term $s$ in \Fo to a term $t$ in \Ho and a list of links $l$.
+The links connect unification variables in \Ho with variables
+in \Fo and are used to decompile the assignment,
+$\D{\sigma}{l}{\rho}$.
+
+Given 
+$$
+\C{s_1}{t_1}{l_1} \land \C{s_2}{t_2}{l_2}
+$$
+we define
+$$
+s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}
+    t_1 \Ue t_2 \mapsto \sigma \land
+    \D{\sigma}{l_1 + l_2}{\rho}
+$$
+
+We write $s \in \mathcal{L}_\lambda$ if all unif variables in
+$s$ are applied to distinct bound variables.
+
+\begin{align}
+t_1 \Uo t_2 \mapsto \rho &\Rightarrow \rho t_1 \Eo \rho t_2\\
+s_i \in \mathcal{L}_\lambda \land \exists \rho, \rho s_1 \Eo \rho s_2 &\Leftrightarrow s_1 \Uo s_2 \mapsto \rho' \subseteq \rho \\
+\forall \rho, \rho s_1 \Eo \rho s_2 &\Rightarrow \rho t_1 \Uo \rho t_2
+\end{align}
+
+the first one is that we are correct
+
+the second one is that we are complete and mgu iff the problems are nice
+
+the third one is still broken since we did not define how
+to apply a fo subst to a ho term but it morally means
+that any valid substitution for s1 s2 cannot break
+the unification of terms t1 t2 obtained by compiling
+s1 and s2 before knowing the subst.
 
 These properties allow us to simulate a unification based
-backward search on DTT by using \Ue{}. In particular
-any unification that would work in DTT will work,
-and whenever a unification in DTT would have failed
-either the corresponding \Ue{} or the link fails.
-
-This grants that the trace of the logic program
+backward search on DTT by using \Ue, in a faithful way:
+the trace of the logic program
 performing the search not only gives the same result, but

### 2024-02-22 paper/main.tex
update tex

@@
   To address this issue, we propose a better-behaved encoding, \Ho,
-  demonstrate how to map unification problems in \Fo
+  demonstrate how to map unification problems in \Fo{}
   to related problems in \Ho, and illustrate
@@
 language (its type is \elpiIn{tm}).
-If we try to backchain the rule (r3) on the encoding of the goal (g) above:
+If we try to backchain the rule (r3) on the encoding of the goal (g) below:
 
@@
 in which the problems sketched here can be fully described.
-We then detail an encoding \elpiIn{comp} from \Fo to \Ho (the language of
+We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
 the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
@@
 
-To state precisely the problem we solve we need a \Fo representation
+To state precisely the problem we solve we need a \Fo{} representation
 of DTT terms and a \Ho one.
@@
 We illustrate a compilation $\C{s}{t}{l}$ that
-maps a term $s$ in \Fo to a term $t$ in \Ho and a list of links $l$.
+maps a term $s$ in \Fo{} to a term $t$ in \Ho and a list of links $l$.
 The links connect unification variables in \Ho with variables
-in \Fo and are used to decompile the assignment,
+in \Fo{} and are used to decompile the assignment,
 $\D{\sigma}{l}{\rho}$.
@@
 
+\
+
 In the case of the OL, variables have no scope. For example, the subterm
@@
       comp t t' /\ comp u u' /\ 
-      unif_ml t' u' s' /\ decomp s' s.
+      unif_ml t' u' s' /\ links ... /\ decomp s' s.
 \end{coqcode}
@@
 original term, it is replaced with a meta variable, and if the OL variable is
-applied to a list $L$ of terms, then this list becomes the scope of the variable
-if $L$ is a list of distinct names. For all the other constructors of
+applied to a list of distinct names $L$, then this list becomes the scope of the variable. 
+For all the other constructors of
 \elpiIn{tm}, the same term constructor is returned and its arguments are
@@
 The following code
+%
 \begin{elpicode}
@@
 \end{elpicode}
-
+%
 \noindent
@@
 scope of the two variables, or equivalently, in a typed language, their arity.
+
+As an example, let's study the following unification problem (a slightly 
+modified version from \cref{sec:intro}):
+
+\begin{elpicode}
+  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
+    lam x\ app [c"decision", app[uv 0, x]]
+\end{elpicode}
+
+\noindent
+we have the main unification problem where the nested \elpiIn{app} nodes have
+lists of different lengths making the unification to fail. The compilation of 
+these terms produces a new unification problem with the following shape:
+
+\begin{elpicode}
+  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~ 
+    lam x\ app [c"decision", uv 1 [x]]
+\end{elpicode}
+
+\noindent
+The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
+the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
+ML, that is, the index \elpiIn{0} for that unification variable of the OL term
+has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
+different substitution mapping, one for the OL and one for the ML and the indexes
+of variable point to the respective substitution.
 

### 2024-02-22 paper/main.tex
update tex

@@
 meta language, we start by formally describing the two languages. Employing
-meta-programming for this purpose, \cref{code:lang-descr} presents the type
-\elpiIn{tm} containing the constructors for term application, lambda abstraction
-and constants. Moreover, in order to represent unification variables, we need 
-to give two different constructors.
+meta-programming for this purpose, we present in \cref{code:common-terms} a new type for the terms of
+the OL.
+%
+\setlength{\abovecaptionskip}{0pt}
+\setlength{\belowcaptionskip}{-13pt}
 
-\
+\begin{figure}[H]
+  \begin{elpicode}
+    kind tm type.
+    type app list tm -> tm.
+    type lam (tm -> tm) -> tm.
+    type c   string -> tm.
+  \end{elpicode}
+  \caption{Common terms}
+  \label{code:common-terms}
+  \Description[code:common-terms]{code:common-terms}
+\end{figure}
+%
+\noindent
+This encoding is very similar to the one introduced in \cref{sec:intro},\todo{say we do not care about types} except
+for the \elpiIn{all} constructor. We explicitly neglect it since blabla.
+Moreover, since we are working with unification variables we need to introduce a
+new constructor.
 
 In the case of the OL, variables have no scope. For example, the subterm
-\coqIn{P x} from the instance forall\_dec. Here, \coqIn{P} is an higher order
-variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} is a name bound to
-\coqIn{P}. However, at the meta level, the translation of \coqIn{P x} becomes
+\coqIn{P x} from the instance forall\_dec has \coqIn{P} an higher order
+variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} as bound variable. 
+However, at the meta level, the translation of \coqIn{P x} becomes
 \elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
-encoding, unification variables are encoded as integer corresponding to memory
-addresses, and the constructor for unification variables in the OL is
-\elpiIn{fo_uv}. This constructor is prefixed with \elpiIn{fo} since the
-variable, having no scope, is a first-order variable.
+encoding, we represent the variables of the OL in the following way:
 
-On the other hand, in the case of the ML, we want to use its unification
-algorithm to make variable assignment. Since the ML is an Higher Order
-Programming Language, we represent unification variables with the \elpiIn{uv}
-constructor, which, this time, can see a list of terms. Of course some attention
-should be payed when dealing with this constructor, since we have to certify
-each time that an \elpiIn{uv i} remain in the pattern fragment, that is, the
-list of term in the scope of \elpiIn{i} is a list of distinct names. Finally,
-the following code
+\begin{elpicode}
+  type fo_uv nat -> tm.
+\end{elpicode}
+
+\noindent 
+In particular, a variable of the OL is identified by a integer
+refereeing a memory address. This constructor is prefixed with \elpiIn{fo} since
+the variable, having no scope, is a first-order variable. Finally, the memory of
+the OL is depicted by the following type abbreviation:
+
+\begin{elpicode}
+  typeabbrev fo_subst list (option tm).
+\end{elpicode}
+
+\noindent 
+which is a list of optional terms. If a cell is \elpiIn{none}, then the variable
+corresponding to this cell is not instantiated.
+
+On the other hand, the terms of the ML have exactly the same shape
+has the terms of \cref{code:common-terms}. However, since the ML is an Higher
+Order Programming Language, unification unification variables must have a scope.
+We represent them in the following way:
+
+\begin{elpicode}
+  type ho_uv nat -> list tm -> tm.
+\end{elpicode}
+
+Of course some attention
+should be payed when dealing with \elpiIn{ho_uv}, since we have to certify
+that an \elpiIn{uv i} remain in the pattern fragment, that is, the
+list of term in the scope of \elpiIn{i} is always a list of distinct names.
+
+Another important implementation should be provided in order to represent the
+lambda abstractions and the applications of the ML. As outlined in
+\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not the same semantic as 
+<<\elpiIn{lam x\ app[f, x]}>>, since, even though, the two terms represent the same
+concept, the former is the lambda abstraction of \elpiIn{f} applied to the 
+binder in ML, whereas the latter is the same term representation but wrt the OL.
+We already have defined the nodes \elpiIn{app} and \elpiIn{lam} for the OL,
+therefore, in the following code snippet, we give their representation in the
+ML.
 
@@
 \noindent
-illustrates the \elpiIn{assmt} representing variable assignment in the ML. \marginpar{todo: explain better}
+In particular, the node \elpiIn{abs} stands for the lambda abstraction. The
+node \elpiIn{val} contains terms of the object language. The node for variable
+application is not really necessary, since, using the syntax of the ML, it is
+defined by putting two terms of the ML side by side. For example, the terms 
+<<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.
 
-The memory of the two languages are represented with lists of substitutions. In
-particular, the substitution of the OL, called \elpiIn{fo_subst}, is made by
-optional terms such that, if the substitution is \elpiIn{none}, then the
-variable is not instantiated. Note that the variables in the \elpiIn{fo_subst}
-have always the \elpiIn{fo_uv} constructor. On the other hand, the ML
-substitution is an optional assignment and in that assignment, variables are
-considered to have the \elpiIn{uv} constructor.
+The memory of the ML, is defined with the following type abbreviation:
+
+\begin{elpicode}
+  typeabbrev ho_subst list (option tm).
+\end{elpicode}
+
+\noindent
+with the invariant that the \elpiIn{tm} inside the cell never contains the node
+\elpiIn{fo_uv}. Note that, reciprocally, the \elpiIn{fo_subst} does not contain
+any node of the form \elpiIn{ho_uv}.
+
+\subsection{Term equality}
+
+\old
 

### 2024-02-23 paper/main.tex
some typos

@@
 \noindent Under this context of instances a type-class solver is able to prove
-the the following statement automatically by back-chaining.
+the following statement automatically by back-chaining.
 
@@
 \elpiIn{e} over a variable \elpiIn{x}
-is \elpiIn{x\ e}, and square brackets denote a list of
+is <<\elpiIn{x\ e}>>, and square brackets denote a list of
 terms separated by comma. As an example we show the encoding of the Coq term
-``$\forall y:\mathrm{t}.\mathrm{nfact}~ y~ 3$'':
+<<\coqIn{~$\forall$~y:t, nfact y 3}>>:
 
@@
 \noindent
-We now illustrate the encoding of the three instances above as higher order
-logic programming rules: capital letters denote rule
+We now illustrate the encoding of the three instances above as higher-order
+logic-programming rules: capital letters denote rule
 parameters; \elpiIn{:-} separates the rule's head from the premises;
@@
 \noindent
-we fail because of this ``higher order'' unification problem (for DTT)
-is phrased as a first order unification problem in the meta language:
+we fail because this ``higher order'' unification problem (for DTT)
+is phrased as a first-order unification problem in the meta language:
 the two lists of terms have different lengths!
@@
 \noindent
-Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
+Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
 in its scope, the unification problem (pa) admits one solution:
@@
 employ a unification procedure that captures a simpler one,
-just $\eta\beta$, and that solves higher order problems
+just $\eta\beta$, and that solves higher-order problems
 restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
@@
 it comprises $\eta\beta$ (for the meta language functions), and the
-unification procedure \Ue{} solves higher order problems in
+unification procedure \Ue{} solves higher-order problems in
 $\mathcal{L}_\lambda$.
@@
 since the abstraction and application term constructors
-the two unification procedure deal with are different. For example
+the two unification procedures deal with are different. For example
 
@@
 $t$ and we assume that the unification of our meta language is correct:
-
+%
 $$
@@
 
-% \printbibliography
+\printbibliography
 

### 2024-02-23 paper/main.tex
update tex

@@
 \newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
+\newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}
 
@@
 
+\newcommand{\specunif}[3]{
+  #3_i \in \mathcal{L}_\lambda \Rightarrow %
+    \exists \rho, %
+      \rho #3_1 #1 \rho #3_2  %
+        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho 
+}
+
+\newcommand{\unifcorrect}[3]{
+  % \forall \rho #3_1 #3_2, %
+    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    #3_i \in \llambda \Rightarrow
+      #3_1 #2 #3_2 \mapsto \rho
+        \Rightarrow 
+          \rho #3_1 #1 \rho #3_2  %
+}
+
+\newcommand{\unifcomplete}[3]{
+  % \forall #3_1 #3_2, %
+    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    #3_i \in \llambda \Rightarrow
+      % \forall \rho, %
+        \rho #3_1 #1 \rho #3_2  %
+          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho 
+}
+
 We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
@@
 $t$ and we assume that the unification of our meta language is correct:
+
 %
-$$
-t_1 \Ue t_2 \mapsto \sigma \Rightarrow \sigma t_1 \Ee \sigma t_2
-$$
+\begin{gather}
+  \unifcorrect{\Ee}{\Ue}{t}\\
+  \unifcomplete{\Ee}{\Ue}{t}
+\end{gather}
 
@@
 $s$ are applied to distinct bound variables.
-
-\begin{align}
-t_1 \Uo t_2 \mapsto \rho &\Rightarrow \rho t_1 \Eo \rho t_2\\
-s_i \in \mathcal{L}_\lambda \land \exists \rho, \rho s_1 \Eo \rho s_2 &\Leftrightarrow s_1 \Uo s_2 \mapsto \rho' \subseteq \rho \\
-\forall \rho, \rho s_1 \Eo \rho s_2 &\Rightarrow \rho t_1 \Uo \rho t_2
-\end{align}
+%
+\begin{gather}
+\unifcorrect{\Eo}{\Uo}{s}\\
+\unifcomplete{\Eo}{\Uo}{s}\\
+% \forall \rho'\rho, 
+  \rho s_1 \Eo \rho s_2 \Rightarrow
+  \rho' \subseteq \rho \Rightarrow
+  \rho's_i \in \mathcal{L}_\lambda \Rightarrow
+  \rho' s_1 \Uo \rho' s_2
+\end{gather}
 
@@
 
-\input{lang_descr.tex}
+% \input{lang_descr.tex}
 
@@
 A key property needed in unification is being able to verify if two terms are
-equal. This is kind of a structural equality verification between two terms,
-where variable dereferencing is performed when the variable is assigned. A
-sketch of the equality function is given for the OL language in
-\cref{code:term-equal}. Though, this equality relation over terms of a language
-can be powered by other reduction rules depending equational theory being
-considered. In our case, the OL terms are equal under $\eta\beta$ redex. This mean
+equal wrt a given equational theory. This relation allow to compare terms under
+a certain substitution mapping, so that any time a variable $v$ is assigned in a
+subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
+test for equality is continued on the new-created subterm.
+
+The base equality function over terms can be defined as follows:
+
+\begin{figure}
+  \begin{elpicode}
+    % congruence
+    type ~(\Eo)~ tm -> tm -> o.
+    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
+    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
+    con X ~\Eo~con X.
+    fov N ~\Eo~fov N.
+    % eta
+    lam F ~\Eo~T :- not (T = lam _), 
+      pi x\ beta T [x] (T' x), lam F ~\Eo~lam T'.
+    T ~\Eo~(lam F) :- not (T = lam _), 
+      pi x\ beta T [x] (T' x), lam T' ~\Eo~lam F.
+    % beta
+    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T.
+    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'.
+  \end{elpicode}
+  \caption{Equal predicate}
+  \label{code:eq-pred}
+  \Description[code:equal-pred]{code:equal-pred}
+\end{figure}
+
+Though, this equality relation is to be extended with $\eta\beta$ reduction
+rules to mirror the theory of the OL. This mean
 that new rules for those two redexes are added in the implementation of

### 2024-02-23 paper/main.tex
labels

@@
 \begin{coqcode}
-  Check _ : Decision (forall y: fin 7, nfact y 3).  (* g *)
+  Check _ : Decision (forall y: fin 7, nfact y 3).   ~\customlabel{goal:g}{(g)}~
 \end{coqcode}
@@
 \begin{elpicode}
-finite (app[c"fin", N]).                             % r1
-decision (app [c"nfact", N, NF]).                    % r2
-decision (all A x\ app[P, x]) :- finite A,           % r3
+finite (app[c"fin", N]).                              ~\customlabel{clause:r1}{(r1)}~
+decision (app [c"nfact", N, NF]).                     ~\customlabel{clause:r2}{(r2)}~
+decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
   pi w\ decision (app[P, w]).
@@
 \noindent
-Unfortunately this direct translation of rule (r3) uses the
+Unfortunately this direct translation of rule \ref{clause:r3} uses the
 predicate \coqIn{P} essentially as a first order term for the meta
 language (its type is \elpiIn{tm}).
-If we try to backchain the rule (r3) on the encoding of the goal (g) below:
+If we try to backchain the rule \ref{clause:r3} on the encoding of the goal \ref{goal:g} below:
 
 \begin{elpicode}
-decision (all (app[c"fin", c"7"]) y\                 % g
+decision (all (app[c"fin", c"7"]) y\                 
   app[c"nfact", y, c"3"]).
@@
 \noindent
-we fail because this ``higher order'' unification problem (for DTT)
-is phrased as a first-order unification problem in the meta language:
-the two lists of terms have different lengths!
+we obtain an unsolvable unification problem \ref{problem:p}:
+the two lists of terms have different lengths! The root cause is that
+\ref{problem:p} is an higher order in DTT, but becomes
+first order due to the encoding in the meta language.
 
 \begin{elpicode}
-app[c"nfact", y, c"3"] = app[P, y]                   % p
+app[c"nfact", y, c"3"] = app[P, y]                     ~\customlabel{problem:p}{(p)}~
 \end{elpicode}
@@
 In this paper we study a more sophisticated encoding of Coq terms allowing
-us to rephrase the problematic rule (r3) as follows:
+us to rephrase the problematic rule \ref{clause:r3} as follows:
 
 \begin{elpicode}
-decision (all A x\ Pm x) :- link Pm A P, finite A,   % r3a
+decision (all A x\ Pm x) :- link Pm A P, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
   pi x\ decision (app[P, x]).
@@
 \noindent
-Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
-in its scope, the unification problem (pa) admits one solution:
+Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
+in its scope, the unification problem \ref{problem:pa} admits one solution:
 
 \begin{elpicode}
-app[c"nfact", y, c"3"] = Pm y                       % pa
+app[c"nfact", y, c"3"] = Pm y                         ~\customlabel{problem:pa}{(pa)}~
 Pm = x\ app[c"nfact", x, c"3"]       % assignment for Pm
@@
 \noindent
-After unifying the head of rule (r3a) with the goal Elpi runs
+After unifying the head of rule \ref{clause:r3a} with the goal Elpi runs
 the premise \elpiIn{link Pm A P} that is in charge of bringing the
@@
 \noindent
-In turn this redex prevents the rule (r2) to backchain properly since
+In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
 the following unification problem has no solution:
@@
 unification procedure in the meta language, that is crafting a \elpiIn{unif}
-predicate to be used as follows in rule (r3):
+predicate to be used as follows in rule \ref{clause:r3}:
 
@@
 
-Our choice of encoding of DTT may look weird to the reader familiar with
-LF, since used a shallow encoding of classes and binders, but not of the
-``lambda calculus'' part of DTT. Here a more lightweight encoding
-that unfortunately does not fit our use case
+Our encoding of DTT may look ``semi shallow'' since we we use the meta language
+lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
+A fully shallow encoding unfortunately does not fit our use case, although
+it would make the running example work:
 
@@
 decision (nfact N NF).
-decision (all A x\ P x) :- 
-  (pi x\ decision (P x)), finite A.
+decision (all A x\ P x) :- finite A, pi x\ decision (P x).
 \end{elpicode}
 
-but in DTT this is not always possible and not handy in our use case,
-since the arity of constants is not fixed.
+\noindent
+There are two reasons for dismissing this encoding. The first one is that
+in DTT it is not always possible to adopt it since the type system
+of the meta language is too weak to accommodate terms with a variable arity,
+like the following example:
 
 \begin{coqcode}
-Fixpoint narr T n := 
-  if n is S m then T -> narr T m else T.
-Definition nsum n : narr nat (n+1).
-Check nsum 2   8 9 : nat.
-Check nsum 3 7 8 9 : nat.
+Fixpoint arr T n := if n is S m then T -> arr T m else T.
+Definition sum n : arr nat n := ...
+Check sum 2   7 8   : nat.
+Check sum 3   7 8 9 : nat.
 \end{coqcode}
 
-moreover we use the same encoding for meta programming, or even just to provide
-hand written rules. We want to access the syntax of OL, so our embedding cannot
-be that shallow. We want to keep it shallow for the binders, but we need
-the c, app and lam nodes. % all ?
+\noindent
+The second reason is the encoding for Coq is used for meta programming the
+system, hence it must accommodate the manipulation of terms that are now
+know in advance (not even defined in Coq) without using introspection
+primitives such as Prologs's \texttt{functor} and \texttt{arg}.
 
-Note that this~\cite{felty93lics} is related and make the
+In the literature we could find a few related encoding of DTT.
+TODO In~\cite{felty93lics} is related and make the
 discrepancy between the types of ML and DTT visible. In this case
@@
 
-This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
+TODO This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
+
+None of the encodings above provide a solution to our problem.
 

### 2024-02-23 paper/main.tex
refactor eta expand rule in elpi and tex
update constructor name
tex update

@@
     type lam (tm -> tm) -> tm.
-    type c   string -> tm.
+    type con string -> tm.
   \end{elpicode}
@@
 
-\subsection{Term equality}
+\subsection{Auxiliary relations}
+
+Together with the description of the terms of the language, we need some
+auxiliary functions to perform operations like term equality, unification,
+dereferencing. There predicates are supposed to be implemented in the OL and the
+ML following respectively their specification. In the following few paragraphs
+we give a brief definition of them.
+
+\paragraph{Term equality: \Eo vs. \Ee}
+\paragraph{Term unification: \Uo vs. \Ue}
+\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}
+TODO
 
@@
     % eta
-    lam F ~\Eo~T :- not (T = lam _), 
-      pi x\ beta T [x] (T' x), lam F ~\Eo~lam T'.
-    T ~\Eo~(lam F) :- not (T = lam _), 
-      pi x\ beta T [x] (T' x), lam T' ~\Eo~lam F.
+    lam F ==o T :- pi x\ beta T [x] (T' x), 
+      x ==o x => F x ==o T' x.
+    T ==o lam F :- pi x\ beta T [x] (T' x), 
+      x ==o x => T' x ==o F x.
     % beta

### 2024-02-23 paper/main.tex
c -> con

@@
 type all  tm -> (tm -> tm) -> tm.     % forall quantifier
-type c    string -> tm.               % constants
+type con  string -> tm.               % constants
 \end{elpicode}
@@
 \begin{elpicode}
-all (c"t") y\ app[c"nfact", y, c"3"]
+all (con"t") y\ app[con"nfact", y, con"3"]
 \end{elpicode}
@@
 \begin{elpicode}
-finite (app[c"fin", N]).                              ~\customlabel{clause:r1}{(r1)}~
-decision (app [c"nfact", N, NF]).                     ~\customlabel{clause:r2}{(r2)}~
+finite (app[con"fin", N]).                            ~\customlabel{clause:r1}{(r1)}~
+decision (app [con"nfact", N, NF]).                   ~\customlabel{clause:r2}{(r2)}~
 decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
@@
 \begin{elpicode}
-decision (all (app[c"fin", c"7"]) y\                 
-  app[c"nfact", y, c"3"]).
+decision (all (app[con"fin", con"7"]) y\                 
+  app[con"nfact", y, con"3"]).
 \end{elpicode}
@@
 \begin{elpicode}
-app[c"nfact", y, c"3"] = app[P, y]                     ~\customlabel{problem:p}{(p)}~
+app[con"nfact", y, con"3"] = app[P, y]                 ~\customlabel{problem:p}{(p)}~
 \end{elpicode}
@@
 \begin{elpicode}
-app[c"nfact", y, c"3"] = Pm y                         ~\customlabel{problem:pa}{(pa)}~
-Pm = x\ app[c"nfact", x, c"3"]       % assignment for Pm
-A = app[c"fin", c"7"]                % assignment for A
+app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(pa)}~
+Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
+A = app[con"fin", con"7"]              % assignment for A
 \end{elpicode}
@@
 \begin{elpicode}
-P = lam A a\ app[c"nfact", a, c"3"]
+P = lam A a\ app[con"nfact", a, con"3"]
 \end{elpicode}
@@
 \begin{elpicode}
-decision (app[lam A (a\ app[c"nfact", a, c"3"]), w])
+decision (app[lam A (a\ app[con"nfact", a, con"3"]), w])
 \end{elpicode}
@@
 \begin{elpicode}
-app[lam A (a\ app[c"nfact", a, c"3"]), x] =
-  app[c"nfact", N, NF]
+app[lam A (a\ app[con"nfact", a, con"3"]), x] =
+  app[con"nfact", N, NF]
 \end{elpicode}
@@
 \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
-\elpiIn{lam A x\ app[c"f", x]} & \Uo{} & \elpiIn{c"f"}\\
-\elpiIn{lam A x\ app[c"f", x]} & \nUe{} & \elpiIn{c"f"} \\
+\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"}\\
+\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"} \\
 \elpiIn{P x} & \Ue{} & \elpiIn{x}\\

### 2024-02-24 paper/main.tex
wip

@@
 \noindent
-Unfortunately this direct translation of rule \ref{clause:r3} uses the
-predicate \coqIn{P} essentially as a first order term for the meta
+Unfortunately this translation of rule \ref{clause:r3} uses the
+predicate \coqIn{P} as a first order term for the meta
 language (its type is \elpiIn{tm}).
-If we try to backchain the rule \ref{clause:r3} on the encoding of the goal \ref{goal:g} below:
+If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
+\ref{goal:g}:
 
@@
 we obtain an unsolvable unification problem \ref{problem:p}:
-the two lists of terms have different lengths! The root cause is that
-\ref{problem:p} is an higher order in DTT, but becomes
-first order due to the encoding in the meta language.
+the two lists of terms have different lengths!
+%The root cause is that
+%\ref{problem:p} is an higher order in DTT, but becomes
+%first order in the meta language due to the ``naive'' encoding.
 
@@
 \begin{elpicode}
-app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(pa)}~
+app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(p')}~
 Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
@@
 assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
-back to the domain of Coq terms (the type \elpiIn{tm}):
+back to the domain of Coq terms:
 
@@
 This simple example is sufficient to show that the encoding we seek
-is not trivial, since the solution for \elpiIn{P} above generates a
+is not trivial and does not only concern the head of rules, but the entire sequence
+of unification problems that constitute the execution of a logic program.
+In fact
+the solution for \elpiIn{P} above generates a
 (Coq) $\beta$-redex in the second premise (the predicate
-under the \elpiIn{pi w\ }\hspace{-0.8em}):
+under the \elpiIn{pi w\ }\hspace{-0.4em}):
 
@@
 \end{elpicode}
-
 \noindent
-The root cause of the problems we face is that the unification procedure
-\Ue of the meta language is not aware of the equational theory of the object
-logic \Uo, even if both theories include $\eta\beta$-conversion and admit most general
-unifiers for problems in the pattern fragment~\cite{miller92jsc}.
+~\\
+The root cause of the problems we sketched in the running example
+is that the unification procedure \Ue of the meta language is not aware
+of the equational theory of the object logic, even if both theories
+include $\eta\beta$-conversion and admit most general
+unifiers for unification problems in the pattern fragment \llambda~\cite{miller92jsc}.
 
@@
 
-\section{Problem statement and alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
+\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
 \label{sec:encodings}
@@
 $\sigma$, we write $\sigma t$ for the application of the substitution to
-$t$ and we assume that the unification of our meta language is correct:
+$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
+and we assume that the unification of our meta language is correct:
 
@@
 $$
+Where $l_1 + l_2$ is the list concatenation of links.
 
@@
 \begin{gather}
-\unifcorrect{\Eo}{\Uo}{s}\\
-\unifcomplete{\Eo}{\Uo}{s}\\
+\unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
+\unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}\\
 % \forall \rho'\rho, 
@@
   \rho's_i \in \mathcal{L}_\lambda \Rightarrow
-  \rho' s_1 \Uo \rho' s_2
+  \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
 \end{gather}
 
-the first one is that we are correct
+Properties \ref{prop:correct} and \ref{prop:complete} state, respectively, that
+in \llambda the implementation of \Uo is correct, complete and returns the most
+general unifier.
 
-the second one is that we are complete and mgu iff the problems are nice
+Property \ref{prop:simulation} states that \Uo is resilient to problems outside
+\llambda if a third party provides a (partial) solution for its problem. Since
+we are interested in using \Uo in a proof search procedure, made of a sequence
+of unification problems, not necessarily in \llambda. In practice it is often
+the case the order in which these problems are stated matters. A Typical example
+is the following problem
 
-the third one is still broken since we did not define how
-to apply a fo subst to a ho term but it morally means
-that any valid substitution for s1 s2 cannot break
-the unification of terms t1 t2 obtained by compiling
-s1 and s2 before knowing the subst.
+\begin{elpicode}
+app [F, con"a"] = app[F, con"a", con"a"]
+\end{elpicode}
 
-These properties allow us to simulate a unification based
-backward search on DTT by using \Ue, in a faithful way:
-the trace of the logic program
-performing the search not only gives the same result, but
-also takes the same paths, that is it fails as early as possible.
+\noindent
+preceded by
 
-\subsection{Alternative encodings and related work}
+\begin{elpicode}
+F = lam x\app[const f,x,x]
+\end{elpicode}
+
+\noindent
+becomes solvable in DTT trivially, since the term is ground (hence in \llambda),
+but is one substitutes F in the LHS does not find, structurally, the RHS
+hence \Ue would fail (since \Ue does not know about the $\beta$ rule of DTT).
+Our compiler takes care of making property \ref{prop:simulation} hold, see
+section XXX.
+
+Property \ref{prop:complete} is also relevant to use \Uo for logic programming.
+In particular we want failures to occur as early as possible, so want
+the decomp phase to take place immediately after \Ue, and fail if need be.
+This becomes particularly important since compile may introduce two
+ho variables for the same fo one, leaving the task of unifying the solutions
+to decomp.
+
+\section{Alternative encodings and related work}
 

### 2024-02-25 paper/main.tex
Update aux relation

@@
 \noindent
-Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
+Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
 in its scope, the unification problem \ref{problem:pa} admits one solution:
@@
 \noindent
-After unifying the head of rule \ref{clause:r3a} with the goal Elpi runs
-the premise \elpiIn{link Pm A P} that is in charge of bringing the
+After unifying the head of rule \ref{clause:r3a} with the goal, Elpi runs
+the premise <<\elpiIn{link Pm A P}>> that is in charge of bringing the
 assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
@@
 \begin{gather}
-  \unifcorrect{\Ee}{\Ue}{t}\\
+  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
   \unifcomplete{\Ee}{\Ue}{t}
@@
 
-\paragraph{Term equality: \Eo vs. \Ee}
+\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}
+
+Since in our encoding we explicitly drag substitution mappings, we need to
+dereference variables of the two languages. This is particularly useful to check
+if two terms in the OL (resp. in the ML) are equal. The constructors
+representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
+make the dereferencing procedure to recurse over the their subterms. The code
+below display this behavior for the dereferencing performed by the OL:
+
+\begin{elpicode}
+  type fo_deref fo_subst -> tm -> tm -> o.
+  fo_deref S (app L1) (app L2) :- forall2 (fo_deref S) L1 L2.
+  fo_deref S (lam F1) (lam F2) :- 
+    pi x\ fo_deref S x x => fo_deref S (F1 x) (F2 x).
+  fo_deref _ (con X) (con X).
+\end{elpicode}
+
+A very similar implementation for the dereferencing performed by the ML, given 
+in the appendix, can be encoded respecting the signature:
+
+\begin{elpicode}
+  type ho_deref ho_subst -> tm -> tm -> o.
+\end{elpicode}
+
+More attention should be payed when the dereferencing encounters a subterm that
+is unification variable. The following new rules should be added to the previous
+code snipper to complete the dereferencing of variables of the OL:
+
+\begin{elpicode}
+  fo_deref S (fov N) R :- assigned? N S T, fo_deref S T R.
+  fo_deref S (app [fov N|L]) R :- 
+    assigned? N S T, beta T L R', fo_deref S R' R.
+  fo_deref _ (fov N) (fov N).
+\end{elpicode}
+
+In particular, 1) if the variable \elpiIn{fov N} is assigned to a term
+\elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
+dereferencing \elpiIn{T}; 2) if the current term is the application of a
+variable \elpiIn{fov N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
+is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
+reduction of \elpiIn{T} to \elpiIn{L}. This is due to the semantics of the
+unification variables within the OL; 3) finally, if \elpiIn{uv N} is not
+assigned, then the same variable is returned.
+
+A different reasoning is to be addressed to the variables of the ML. Firstly, a
+meta variable cannot appear in the \elpiIn{app} node as the first element of the
+list, we will explain why in \cref{sec:compilation}; secondly, variables in the
+ML have a scope which should be taken into account. The following few lines of
+code outline the rules of dereferencing for the ML:
+
+\begin{elpicode}
+  ho_deref H (hov N L) R :- assigned? N H T, 
+    ho_beta T L R', ho_deref H R' R.
+  ho_deref H (hov X L) (hov X L1) :- 
+    forall2 (ho_deref H) L L1.
+\end{elpicode}
+
+\noindent
+where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
+return the dereferencing of the beta reduction \elpiIn{T} to the scope
+\elpiIn{L}.
+
+\begin{figure}
+  \begin{elpicode}
+    type ~(\Eo)~ tm -> tm -> o.
+    % congruence
+    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
+    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
+    con X ~\Eo~con X.
+    fov N ~\Eo~fov N.
+    % eta
+    lam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
+      x ~\Eo~x => F x ~\Eo~T' x.
+    T ~\Eo~lam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
+      x ~\Eo~x => T' x ~\Eo~F x.
+    % beta
+    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
+    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
+  \end{elpicode}
+  \caption{Equal predicate}
+  \label{code:eq-pred}
+  \Description[code:equal-pred]{code:equal-pred}
+\end{figure}
+
+\paragraph{Term equality: \Eo vs. \Ee} We can test if two terms are equal
+following the equational theory of the language being considered. In
+\cref{code:eq-pred} we provide an implementation of the \Eo predicate. The first
+four rules check if the two terms are equal regarding the structure of the
+current node, that is, two terms are equal if they have same head and if
+recursively each subterm is two by two equal. Moreover, since the theory of the
+OL accepts $\eta$- and $\beta$-equivalence, then we implement the corresponding
+rules.
+
+\noindent
+The equality relation for the ML, accepts $\eta\beta$-equivalence between terms
+of the ML. Recall that \elpiIn{abs x\ f x}, is a valid $\eta$ expansion of the
+function \elpiIn{f} and that \elpiIn{lam x\ app[f, x]} is not that equivalent to
+\elpiIn{f} at meta level. However, since we are interested in using the
+unification procure of the ML, by \cref{prop:correct-ml}, we can use the
+\Ue relation to test, when needed if two terms are equal in the ML.
+
 \paragraph{Term unification: \Uo vs. \Ue}
-\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}
 TODO
@@
 
-\begin{figure}
-  \begin{elpicode}
-    % congruence
-    type ~(\Eo)~ tm -> tm -> o.
-    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
-    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
-    con X ~\Eo~con X.
-    fov N ~\Eo~fov N.
-    % eta
-    lam F ==o T :- pi x\ beta T [x] (T' x), 
-      x ==o x => F x ==o T' x.
-    T ==o lam F :- pi x\ beta T [x] (T' x), 
-      x ==o x => T' x ==o F x.
-    % beta
-    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T.
-    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'.
-  \end{elpicode}
-  \caption{Equal predicate}
-  \label{code:eq-pred}
-  \Description[code:equal-pred]{code:equal-pred}
-\end{figure}
-
-Though, this equality relation is to be extended with $\eta\beta$ reduction
-rules to mirror the theory of the OL. This mean
-that new rules for those two redexes are added in the implementation of
-\elpiIn{fo_equal}.
-
-If \elpiIn{fo_equal} is conceived to manage equality between terms of the OL,
-the same equality predicate in the ML behave slightly different. By the given
-definition of the ML, the ML allows $\eta\beta$ congruence of terms, but, since
-the node \elpiIn{app} and \elpiIn{abs} are constructor representing the
-applications and the abstractions of the OL, these two reduction rules cannot
-applied on them. We build therefore a predicate \elpiIn{equal} working for terms
-in the ML which is implemented merely with the rules for the \elpiIn{fo_equal}
-predicate.\todo{not true for uv} For example, if \elpiIn{fo_equal []
-(abs x\ [c"f", c]) (c"f")} is true in the object language, \elpiIn{equal [] (abs
-x\ [c"f", c]) (c"f")} produces a failure.
-
 The solution we are proposing aim to overcome these unification issues by 1)
@@
 
-Mathematically, we what to prove the following property:
-
-\begin{coqcode}
-  forall (t  u : term_ol) (s : subst_ol) 
-         (t' u': term_ml) (s': subst_ml), 
-    equal_ol t u s <->
-      comp t t' /\ comp u u' /\ 
-      unif_ml t' u' s' /\ links ... /\ decomp s' s.
-\end{coqcode}
-
-$$Math\ formula$$
-
 In the following section we explain how we deal with term (de)compilation and
-unification variable linking.
+links between unification variables.
 
 \section[Compilation: fo\_tm to tm]{Compilation}
+\label{sec:compilation}
 

### 2024-02-25 paper/main.tex
update tex

@@
 \paragraph{Term unification: \Uo vs. \Ue} 
-TODO
+The last but not least important relation we should take care of before
+presenting our full algorithm aiming to unify terms of the OL in the ML and
+provide the substitution produced in the ML to the OL, is term unification. This
+procedure is a more powerful version of the equal predicate, since unification
+checks if two terms can be equal be assigning unification variables. In our
+representation, variable assignment (or refinement) is performed by modifying
+the corresponding substitution mapping. We will not give an implementation of
+\Uo in the world of the OL, since we are giving an implementation of it using
+our algorithm, see \cref{XX}.
+
+On the other hand, we can
 

### 2024-02-25 paper/main.tex
update main.tex

@@
 \elpiIn{tm}, the same term constructor is returned and its arguments are
-recursively compiled. The predicate in charge for term compilation is:\\
+recursively compiled. The predicate in charge for term compilation is:
+
 \elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
-Where, we take the term of the OL, produce the term of the ML, take a list
+
+\noindent
+where, we take the term of the OL, produce the term of the ML, take a list
 of link and produce a list of new links, take a substitution and return a 

### 2024-02-26 paper/main.tex
tex update

@@
 \begin{elpicode}
-decision (all A x\ Pm x) :- link Pm A P, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
+decision (all A x\ Pm x) :- link Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
   pi x\ decision (app[P, x]).
@@
 the corresponding substitution mapping. We will not give an implementation of
-\Uo in the world of the OL, since we are giving an implementation of it using
+\Uo, since we are giving an implementation of it using
 our algorithm, see \cref{XX}.
 
-On the other hand, we can
+\begin{figure}
+  \begin{elpicode}
+    type unif tm -> tm -> ho_subst -> ho_subst ->  o.
+    % Base cases
+    unif (app L1) (app L2) S S1 :- fold2 unif L1 L2 S S1.
+    unif (lam F1) (lam F2) S S1 :- 
+      pi x\ (pi S\ unif x x S S) => unif (F1 x) (F2 x) S S1.
+    unif (con X) (con X) S S.
+    % Assigned variables
+    unif (hov N Args) T S S1 :- 
+      assigned-deref N S F, beta-abs F Args T1, unif T1 T S S1.
+    unif T (hov N Args) S S1 :- 
+      assigned-deref N S F, beta-abs F Args T1, unif T T1 S S1.
+    % Non-assigned variabless
+    unif (hov N Args1) (hov M Args2) S S3 :-  
+      pattern-fragment Args1, pattern-fragment Args2, 
+      new S W S1, prune W Args1 Args2 Ass, 
+      assign N S1 Ass S2, assign M S2 Ass S3.
+    unif (hov N Args) T S S1 :- pattern-fragment Args,
+      not (T = hov _ _), not_occ N S T,  
+      bind Args T T1, assign N S T1 S1.
+    unif T (hov N Args) S S1 :- pattern-fragment Args,
+      not (T = hov _ _), not_occ N S T, 
+       bind Args T T1, assign N S T1 S1.
+  \end{elpicode}
+  \caption{The \Ue predicate}
+  \label{code:ue-pred}
+  \Description[code:unif-pred]{code:unif-pred}
+\end{figure}
+
+On the other hand, unification in the ML needs to be defined. In
+\cref{code:ue-pred}, we give an implementation of \Ue. This predicate has four
+arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
+$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
+\subseteq \rho_2$. The first three rules unify terms with same rigid heads, and
+call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
+assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
+unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
+terms are unification variables, we test that their arguments are in the pattern
+fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
+pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
+$w$ and return the new mapping $\rho_2$ containing all the new variable
+assignment. Finally, if only one of the two terms is an unification variable
+$v$, after having verified that $v$ does not occur in the other term $t$, we
+bind $v$ to $t$ and return the new substitution mapping.
 

### 2024-02-26 paper/main.tex
update tex

@@
 
+\otext{Goal: $s_1 \Uo s_2$ is compiled into $t_1 \Ue t_2$}
+\otext{What is done: uvars \elpiIn{fo_uv} of OL are replaced into uvars \elpiIn{ho_uv} of the ML}
+\otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
+\otext{Example: \elpiIn{lam x\ app[uv 0, x] ~\Ue~lam x\ c"f"}}
+\otext{Links used to instantiate vars of elpi}
+\otext{After all links, the solution in links are compacted and given to coq}
+\otext{It is not so simple, see next sections (multi-vars, eta, beta)}
+
 The compilation step is meant to recover the higher-order variables of the OL,
@@
 
-\noindent
-- how we transform an fo\_tm in tm \\
-- the role of links \\
-- decomp
-- esempio che va in questa semplice rappresentazione (from intro)
-- esempio che non va, multi-var, eta, beta
-% \section{Term de-compilation}
+\section{Use of multivars}
 
-\section{Unification in ML}
+\subsection{Problems with $\eta$}
+\otext{The following goal: \elpiIn{X = lam x\ lam y\ Y x y, X = lam x\ f}}
+\otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
+\otext{It is not doable, with the same elpi var}
 
-- we accept HO unif with PF \\
-- need of multiple vars for a single OL var
+\subsection{Problems with $\beta$}
 
-% Compile FO to HO
+\otext{The following goal: \elpiIn{X = lam x\ x, app[X, 3] = 3}}
+\otext{We use links-beta}
 
-% Assignement in HO should be decompiled
+\section{First order approximation}
 
-% -------------------------------------------------------------------
+\otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
+\otext{We can re-use part of the algo for $\beta$ given before}
 
-% \noindent
+\section{Unif encoding in real life}
+\otext{Il ML presentato qui è esattamente elpi}
+\otext{Il OL presentato qui è esattamente coq}
+\otext{Come implementatiamo tutto ciò nel solver}
+
+\section{Results: stdpp and tlc}
+\otext{How may rule are we solving?}
+\otext{Can we do some perf test}
+
+\section{Conclusion}
 

### 2024-02-26 paper/main.tex
update tex
wip

@@
 
-Our encoding of DTT may look ``semi shallow'' since we we use the meta language
+Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
+
+Our encoding of DTT may look ``semi shallow'' since we use the meta-language
 lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
@@
 \begin{figure}[H]
-  \begin{elpicode}
+  \begin{tabular}{ll}
+  \begin{minipage}{0.21\textwidth}
+  \begin{elpicodetab}
+kind fm type.
+type fapp list fm -> fm.
+type flam (fm -> fm) -> fm.
+type fcon string -> fm.
+type fuva nat -> fm.
+\end{elpicodetab}
+  \end{minipage}
+  &
+  \begin{minipage}{0.24\textwidth}
+  \begin{elpicodetab}
 kind tm type.
@@
 type con string -> tm.
-  \end{elpicode}
-  \caption{Common terms}
+type uva nat -> list tm -> tm.
+  \end{elpicodetab}
+  \end{minipage}
+  \end{tabular}
+  \caption{Source and Target language}
   \label{code:common-terms}
@@
 \end{figure}
-%
+
 \noindent
@@
 
-In the case of the OL, variables have no scope. For example, the subterm
-\coqIn{P x} from the instance forall\_dec has \coqIn{P} an higher order
-variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} as bound variable. 
-However, at the meta level, the translation of \coqIn{P x} becomes
-\elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
-encoding, we represent the variables of the OL in the following way:
+In the case of the OL unification variables \elpiIn{fuv} have no scope.
+For example the statement of the instance \coqIn{forall_dec} contains the
+term \coqIn{P x} where \coqIn{P} stands for a function of type
+\coqIn{A ~$\to$~ Prop}. However our representation of \coqIn{P x} in the meta
+language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
+In the terms of the ML unification variables come with a scope.
 
 \begin{elpicode}
-  type fo_uv nat -> tm.
+  type uva nat -> list tm -> tm.
 \end{elpicode}
@@
 \noindent
-In particular, a variable of the OL is identified by a integer
-refereeing a memory address. This constructor is prefixed with \elpiIn{fo} since
-the variable, having no scope, is a first-order variable. Finally, the memory of
+Of course some attention
+should be payed when dealing with \elpiIn{uva}, since we have to certify
+that an \elpiIn{uva i} remain in the pattern fragment, that is, the
+list of term in the scope of \elpiIn{i} is always a list of distinct names.
+
+Unification variables of the OL are identified by a natural number
+referring to memory address. 
+Finally, the memory of
 the OL is depicted by the following type abbreviation:
@@
 \begin{elpicode}
-  typeabbrev fo_subst list (option tm).
+  typeabbrev fsubst list (option fm).
 \end{elpicode}
@@
 
-On the other hand, the terms of the ML have exactly the same shape
-has the terms of \cref{code:common-terms}. However, since the ML is an Higher
-Order Programming Language, unification unification variables must have a scope.
-We represent them in the following way:
-
-\begin{elpicode}
-  type ho_uv nat -> list tm -> tm.
-\end{elpicode}
-
-Of course some attention
-should be payed when dealing with \elpiIn{ho_uv}, since we have to certify
-that an \elpiIn{uv i} remain in the pattern fragment, that is, the
-list of term in the scope of \elpiIn{i} is always a list of distinct names.
-
 Another important implementation should be provided in order to represent the
 lambda abstractions and the applications of the ML. As outlined in
-\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not the same semantic as 
-<<\elpiIn{lam x\ app[f, x]}>>, since, even though, the two terms represent the same
-concept, the former is the lambda abstraction of \elpiIn{f} applied to the 
-binder in ML, whereas the latter is the same term representation but wrt the OL.
-We already have defined the nodes \elpiIn{app} and \elpiIn{lam} for the OL,
-therefore, in the following code snippet, we give their representation in the
-ML.
+\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
+<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
+concept, the latter uses the abstraction and application of the OL.
 
@@
   type val tm -> assmt.
+  typeabbrev subst list (option tm).
 \end{elpicode}
@@
 \noindent
-In particular, the node \elpiIn{abs} stands for the lambda abstraction. The
+In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
+The
 node \elpiIn{val} contains terms of the object language. The node for variable
@@
 
-The memory of the ML, is defined with the following type abbreviation:
-
-\begin{elpicode}
-  typeabbrev ho_subst list (option tm).
-\end{elpicode}
+The memory of the ML is represented by a \elpiIn{subst}.
 
@@
 
-\begin{elpicode}
-  type fo_deref fo_subst -> tm -> tm -> o.
-  fo_deref S (app L1) (app L2) :- forall2 (fo_deref S) L1 L2.
-  fo_deref S (lam F1) (lam F2) :- 
-    pi x\ fo_deref S x x => fo_deref S (F1 x) (F2 x).
-  fo_deref _ (con X) (con X).
-\end{elpicode}
+\begin{elpicodetab}
+  type fderef fsubst -> fm -> fm -> o.
+  fderef S (fuva N) R :- assigned? N S T, fderef S T R.
+  fderef S (app [fuva N|L]) R :- 
+    assigned? N S (app NL), append NL L AL, fderef S (app AL) R.
+  fderef _ (fuva N) (fuva N).
+  fderef S (fapp L1) (fapp L2) :- forall2 (fderef S) L1 L2.
+  fderef S (flam F1) (flam F2) :- 
+    pi x\ fderef S x x => fderef S (F1 x) (F2 x).
+  fderef _ (fcon X) (fcon X).
+\end{elpicodetab}
+
+More attention should be payed when the dereferencing encounters a subterm that
+is unification variable. The following new rules should be added to the previous
+code snipper to complete the dereferencing of variables of the OL:
 
@@
 \begin{elpicode}
-  type ho_deref ho_subst -> tm -> tm -> o.
-\end{elpicode}
-
-More attention should be payed when the dereferencing encounters a subterm that
-is unification variable. The following new rules should be added to the previous
-code snipper to complete the dereferencing of variables of the OL:
-
-\begin{elpicode}
-  fo_deref S (fov N) R :- assigned? N S T, fo_deref S T R.
-  fo_deref S (app [fov N|L]) R :- 
-    assigned? N S T, beta T L R', fo_deref S R' R.
-  fo_deref _ (fov N) (fov N).
+  type deref subst -> tm -> tm -> o.
+  deref S (uva N L) R :- assigned? N S T, beta-abs T L T', deref S T' R.
 \end{elpicode}
@@
 
+\noindent
+\otext{An other example: \\
+  \elpiIn{@lam x\ @app[@f, @app[@X, x]] = @Y, (@lam x\ x) = @X.}}
+
 \subsection{Problems with $\beta$}
@@
 \otext{We use links-beta}
+
+\subsection{Tricky examples}
+
+\begin{elpicode}
+  triple ok (@lam x\ @app[@f, @app[@X, x]]) @Y,
+  triple ok @X (@lam x\ x),
+  triple ok @Y @f
+\end{elpicode}
 
@@
 
-% \subsection{implementation}
-
-% l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
-% in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
-% elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
-% ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.
-
-% \section{recovering eta}
-
-% \begin{elpicode}
-% q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
-% F = fun a => app [f,a] ----> F = f
-% \end{elpicode}
-
-% l'utene da p su f, mentre l'istanza pe q forza F a fun ..
-
-% \section{recovering beta}
-
-% \begin{elpicode}
-%   q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
-%   F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
-% \end{elpicode}
-
-% qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
-% p1 F1, e decomp beta F [a] F1.
-
-% \section{recovering eta-beta within unification (non linear variables)}
-
-% se i problemi di cui sopra avvengono nello stesso termine
-
-% \begin{elpicode}
-%   q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
-% \end{elpicode}
-
-% bisogna slegare le due F e poi unificare le soluzioni tra di loro
-
-% \section{heuristic / binary app}
-
-% fo approx / sub pattern fragment
-
-% \begin{elpicode}
-%   p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
-%   p (all x\G x) F' = 
-%   G = x\ f x x a
-%   F = lam x\f x x
-%   F' = (app[f,b,b])
-%   link (F a) F'
-%   link G F
-% \end{elpicode}
-
-% % questo esempio recupera l'app binaria perche
-
-% \begin{elpicode}
-%   (app (app F x) a) = (app (f x x) a)
-% \end{elpicode}
-
-% \section{Stats}
-
-% Concretamente:
-
-% - quante volte l'ordine superiore è necessario in stdpp/tlc
-
-% \section{Conclusion}
-
 \printbibliography

### 2024-02-26 paper/main.tex
typos in tex

--- 2024-02-26 paper/main.tex
@@
 
-Our encoding of DTT may look ``semi shallow'' since we we use the meta language
+Our encoding of DTT may look ``semi shallow'' since we use the meta-language
 lambda abstraction but not its application (for the terms of type \elpiIn{tm}).

### 2024-02-26 paper/main.tex
wip

--- 2024-02-26 paper/main.tex
@@
 language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
-In the terms of the ML unification variables come with a scope.
-
-\begin{elpicode}
-  type uva nat -> list tm -> tm.
-\end{elpicode}
-
-\noindent
+In the terms of the ML unification variables come with a scope (the second
+argument of the \elpiIn{uva} constructor).
 Of course some attention
@@
 
+\marginpar{maybe move code to the appendix}
 \begin{elpicode}
-  typeabbrev fsubst list (option fm).
+typeabbrev memory A (list (option A)).
+type set nat -> memory A -> A -> o.
+set N S T :- nth N S (some T).
+type unset nat -> memory A -> o.
+unset N S :- nth N S none.
+type assign nat -> memory A -> A -> memory A -> o.
+assign z [none|M] T [some T|M].
+assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
 \end{elpicode}
@@
 \begin{elpicode}
+  typeabbrev fsubst (memory ftm).
+
   kind assmt type.
@@
   type val tm -> assmt.
-  typeabbrev subst list (option tm).
+  typeabbrev subst (memory assmt).
 \end{elpicode}
@@
 
-\subsection{Auxiliary relations}
+\subsection{equational theory and unification}
+
+here we give the functions/signatures to express the properties 3-5 in the problem
+statement
 
@@
 
-\begin{elpicodetab}
+\begin{elpicode}
 type fderef fsubst -> fm -> fm -> o.
-  fderef S (fuva N) R :- assigned? N S T, fderef S T R.
-  fderef S (app [fuva N|L]) R :- 
-    assigned? N S (app NL), append NL L AL, fderef S (app AL) R.
-  fderef _ (fuva N) (fuva N).
-  fderef S (fapp L1) (fapp L2) :- forall2 (fderef S) L1 L2.
-  fderef S (flam F1) (flam F2) :- 
-    pi x\ fderef S x x => fderef S (F1 x) (F2 x).
-  fderef _ (fcon X) (fcon X).
-\end{elpicodetab}
+fderef S (fapp [fuva N|A]) R :- set N S (fapp B),
+  append A B C, fderef S (fapp C) R.
+fderef S (fapp [fuva N|A]) (fapp [fuva N|B]) :- unset N S,
+  forall2 (fderef S) A B.
+fderef S (fapp A) (fapp B) :- forall2 (fderef S) A B.
+fderef S (flam F) (flam G) :- 
+  pi x\ fderef S x x => fderef S (F x) (G x).
+fderef _ (fcon C) (fcon C).
+fderef S (fuva N) R :- set N S T, fderef S T R.
+fderef _ (fuva N) (fuva N) :- unset N S.
+\end{elpicode}
 
@@
 
+\marginpar{what about app uva?}
 \begin{elpicode}
   type deref subst -> tm -> tm -> o.
-  deref S (uva N L) R :- assigned? N S T, beta-abs T L T', deref S T' R.
+  % ... the same
+  deref S (uva N A) R :- set N S T,
+    instantiate T A T', deref S T' R.
+  deref S (uva N A) (uva N B) :-unset N S,
+    forall2 (deref S) A B.
 \end{elpicode}
 
-In particular, 1) if the variable \elpiIn{fov N} is assigned to a term
+In particular, 1) if the variable \elpiIn{fuva N} is assigned to a term
 \elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
 dereferencing \elpiIn{T}; 2) if the current term is the application of a
-variable \elpiIn{fov N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
+variable \elpiIn{fuva N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
 is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
@@
 
-\begin{elpicode}
-  ho_deref H (hov N L) R :- assigned? N H T, 
-    ho_beta T L R', ho_deref H R' R.
-  ho_deref H (hov X L) (hov X L1) :- 
-    forall2 (ho_deref H) L L1.
-\end{elpicode}
-
-\noindent
 where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
@@
   \begin{elpicode}
-    type ~(\Eo)~ tm -> tm -> o.
+    type ~(\Eo)~ ftm -> ftm -> o.
     % congruence
-    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
-    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
-    con X ~\Eo~con X.
-    fov N ~\Eo~fov N.
+    fapp A ~\Eo~fapp B :- forall2 ~(\Eo)~ A B.
+    flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
+    fcon C ~\Eo~fcon C.
+    fuva N ~\Eo~fuva N.
     % eta
-    lam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
+    flam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
       x ~\Eo~x => F x ~\Eo~T' x.
-    T ~\Eo~lam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
+    T ~\Eo~flam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
       x ~\Eo~x => T' x ~\Eo~F x.
     % beta
-    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
-    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
+    fapp [flam X | TL] ~\Eo~T :- beta (flam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
+    T ~\Eo~fapp [flam X | TL] :- beta (flam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
   \end{elpicode}
-  \caption{Equal predicate}
+  \caption{Equal predicate OL}
   \label{code:eq-pred}
   \Description[code:equal-pred]{code:equal-pred}
+\end{figure}
+
+\begin{figure}
+  \begin{elpicode}
+    type ~(\Ee)~ tm -> tm -> o.
+    app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
+    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
+    con C ~\Ee~fcon C.
+    uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
+  \end{elpicode}
+  \caption{Equal predicate ML}
+  \label{code:eq-pred-meta}
+  \Description[code:equal-pred-meta]{code:equal-pred-meta}
 \end{figure}
@@
 \begin{elpicode}
-    type unif tm -> tm -> ho_subst -> ho_subst ->  o.
-    % Base cases
-    unif (app L1) (app L2) S S1 :- fold2 unif L1 L2 S S1.
-    unif (lam F1) (lam F2) S S1 :- 
-      pi x\ (pi S\ unif x x S S) => unif (F1 x) (F2 x) S S1.
-    unif (con X) (con X) S S.
-    % Assigned variables
-    unif (hov N Args) T S S1 :- 
-      assigned-deref N S F, beta-abs F Args T1, unif T1 T S S1.
-    unif T (hov N Args) S S1 :- 
-      assigned-deref N S F, beta-abs F Args T1, unif T T1 S S1.
-    % Non-assigned variabless
-    unif (hov N Args1) (hov M Args2) S S3 :-  
-      pattern-fragment Args1, pattern-fragment Args2, 
-      new S W S1, prune W Args1 Args2 Ass, 
+type ~(\Ue)~ tm -> tm -> subst -> subst ->  o.
+% Congruence
+(app A ~\Ue~app B) S S1 :- fold2 ~\Ue~A B S S1.
+(lam F ~\Ue~lam G) S S1 :- 
+  pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
+~\Ue~(con C) (con C) S S.
+% deref
+(uva N A ~\Ue~T) S S1 :- set N S F,
+  beta-abs F A T1, (T1 ~\Ue~T) S S1.
+(T ~\Ue~uva N A) S S1 :- set N S F,
+  beta-abs F A T1, (T ~\Ue~T1) S S1.
+% flex-flex
+(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
+  pattern-fragment A, pattern-fragment B, 
+  new S W S1, prune W Args1 B Ass, 
   assign N S1 Ass S2, assign M S2 Ass S3.
-    unif (hov N Args) T S S1 :- pattern-fragment Args,
-      not (T = hov _ _), not_occ N S T,  
-      bind Args T T1, assign N S T1 S1.
-    unif T (hov N Args) S S1 :- pattern-fragment Args,
-      not (T = hov _ _), not_occ N S T, 
-       bind Args T T1, assign N S T1 S1.
+% assignment
+(uva N A ~\Ue~T) S S1 :- pattern-fragment A,
+  not (T = uva _ _), not_occ N S T,  
+  bind A T T1, assign N S T1 S1.
+(T ~\Ue~uva N A) S S1 :- pattern-fragment A,
+  not (T = uva _ _), not_occ N S T, 
+    bind A T T1, assign N S T1 S1.
+%  Note: (a infix b) c d === (infix) a b c d
+
 \end{elpicode}
-  \caption{The \Ue predicate}
+\caption{The \Ue predicate // TODO move to the appendix}
 \label{code:ue-pred}

### 2024-02-27 paper/main.tex
tex add eta example
tex add tricky unif pb
tex

@@
   \begin{elpicode}
-    type ~(\Eo)~ ftm -> ftm -> o.
+    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.
     % congruence
@@
   \begin{elpicode}
-    type ~(\Ee)~ tm -> tm -> o.
+    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
     app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
@@
 \begin{elpicode}
-type ~(\Ue)~ tm -> tm -> subst -> subst ->  o.
+~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst o}~.
 % Congruence

### 2024-02-27 paper/main.tex
4

@@
 
-% \input{lang_descr.tex}
-
-% Description of languages (OL and ML)
-
-\def\eqfo{eq\_fo\xspace}
-
-In order to reason about unification of the terms of the objet language within the
-meta language, we start by formally describing the two languages. Employing
-meta-programming for this purpose, we present in \cref{code:common-terms} a new type for the terms of
-the OL.
+In order to reason about unification we provide a description of the
+\Fo{} and \Ho languages where unification variables
+are first class terms, i.e. they have a concrete syntax. We keep these languages
+minimal, for example we omit the \elpiIn{all} quantifier of DTT we used
+in the example in Section~\ref{sec:intro}. 
 %
@@
   \end{tabular}
-  \caption{Source and Target language}
+  \caption{\Fo{} and \Ho language}
   \label{code:common-terms}
@@
 \noindent
-This encoding is very similar to the one introduced in \cref{sec:intro},\todo{say we do not care about types} except
-for the \elpiIn{all} constructor. We explicitly neglect it since blabla.
-Moreover, since we are working with unification variables we need to introduce a
-new constructor.
+In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
+Unification variables standing for functions are applied to variables
+via the \elpiIn{fapp} constructor.
+For example in the statement of the instance \coqIn{forall_dec} the
+term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N}
+is of type \elpiIn{nat} and \elpiIn{x} is a bound variable.
 
-In the case of the OL unification variables \elpiIn{fuv} have no scope.
-For example the statement of the instance \coqIn{forall_dec} contains the
-term \coqIn{P x} where \coqIn{P} stands for a function of type
-\coqIn{A ~$\to$~ Prop}. However our representation of \coqIn{P x} in the meta
-language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
-In the terms of the ML unification variables come with a scope (the second
-argument of the \elpiIn{uva} constructor).
-Of course some attention
-should be payed when dealing with \elpiIn{uva}, since we have to certify
-that an \elpiIn{uva i} remain in the pattern fragment, that is, the
-list of term in the scope of \elpiIn{i} is always a list of distinct names.
+In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]}.
+We say that the unification variable \elpiIn{uva N L} is in \llambda iff
+\elpiIn{distinct-names L} holds.
 
-Unification variables of the OL are identified by a natural number
-referring to memory address. 
-Finally, the memory of
-the OL is depicted by the following type abbreviation:
+\begin{elpicode}
+type distinct-names list tm -> o.
+distinct-names [].
+distinct-names [X|XS] :- name X, not(mem X XS),
+  distinct-names XS.
+\end{elpicode}
 
-\marginpar{maybe move code to the appendix}
+\noindent
+The \elpiIn{name} builtin predicate tests if a term is a bound variable.
+
+In both languages unification variables are identified by a natural number,
+which can be seen as a memory address. The memory and its associated
+operations are described below:
+
+\marginpar{maybe move code to the appendix and keep only signatures}
 \begin{elpicode}
@@
 \noindent 
-which is a list of optional terms. If a cell is \elpiIn{none}, then the variable
-corresponding to this cell is not instantiated.
+If a memory cell is \elpiIn{none}, then the corresponding unification variable
+is not set.
 
-Another important implementation should be provided in order to represent the
-lambda abstractions and the applications of the ML. As outlined in
-\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
-<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
-concept, the latter uses the abstraction and application of the OL.
+Since \Ho unification variables have a scope, their solution needs to be
+abstracted over these in order to enables the instantiation of a single
+solution to different scopes. On the contrary solutions to \Fo variables
+are just terms.
 
@@
 \noindent
-In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
-The
-node \elpiIn{val} contains terms of the object language. The node for variable
-application is not really necessary, since, using the syntax of the ML, it is
-defined by putting two terms of the ML side by side. For example, the terms 
-<<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.
+One can see \elpiIn{abs} as a syntax for the lambda abstraction of the meta
+language.
+%Luckily  we need to reason on these abstractions only when we
+%manipulate the solution given by the unification of the meta language; when
+%we manipulate the terms we don't 
+%
+%Another important implementation should be provided in order to represent the
+%lambda abstractions and the applications of the ML. As outlined in
+%\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
+%<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
+%concept, the latter uses the abstraction and application of the OL.
+%
+%In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
+% The
+% node \elpiIn{val} contains terms of the object language. The node for variable
+% application is not really necessary, since, using the syntax of the ML, it is
+% defined by putting two terms of the ML side by side. For example, the terms 
+% <<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.
 
-The memory of the ML is represented by a \elpiIn{subst}.
-
-\noindent
-with the invariant that the \elpiIn{tm} inside the cell never contains the node
-\elpiIn{fo_uv}. Note that, reciprocally, the \elpiIn{fo_subst} does not contain
-any node of the form \elpiIn{ho_uv}.
+The memory \Fo{} is called \elpiIn{fsubst} while the one of \Ho is
+called \elpiIn{subst}.
+Both have the invariant that they are not cyclic, TODO explain.
 

### 2024-02-27 paper/main.tex
4

@@
 
-\section{Languages description}
+\section{Preliminaries: \Fo{} and \Ho}
 \label{sec:lang-spec}
@@
   \end{tabular}
-  \caption{\Fo{} and \Ho language}
+  \caption{\Fo{} and \Ho language}\vspace{0.3em}
   \label{code:common-terms}
@@
 In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
-Unification variables standing for functions are applied to variables
+Unification variables standing for functions are applied their arguments
 via the \elpiIn{fapp} constructor.
@@
 \noindent
-The \elpiIn{name} builtin predicate tests if a term is a bound variable.
+The \elpiIn{name} builtin predicate tests if a term is a bound variable. The
+compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
+reasons, so we don't assume all out terms are in \llambda but rather test.
 
@@
 
-\marginpar{maybe move code to the appendix and keep only signatures}
 \begin{elpicode}
 typeabbrev memory A (list (option A)).
-type set nat -> memory A -> A -> o.
-set N S T :- nth N S (some T).
+type set?   nat -> memory A -> A -> o.
 type unset  nat -> memory A -> o.
-unset N S :- nth N S none.
 type assign nat -> memory A -> A -> memory A -> o.
-assign z [none|M] T [some T|M].
-assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
 \end{elpicode}
@@
 If a memory cell is \elpiIn{none}, then the corresponding unification variable
-is not set.
+is not set. \elpiIn{assign} sets an unset cell to the given value.
 
-Since \Ho unification variables have a scope, their solution needs to be
-abstracted over these in order to enables the instantiation of a single
+Since in \Ho unification variables have a scope, their solution needs to be
+abstracted over it in order to enable the instantiation of a single
 solution to different scopes. On the contrary solutions to \Fo variables
-are just terms.
+are plain terms.
 
@@
   typeabbrev fsubst (memory ftm).
-
   kind assmt type.
@@
 \noindent
-One can see \elpiIn{abs} as a syntax for the lambda abstraction of the meta
-language.
-%Luckily  we need to reason on these abstractions only when we
-%manipulate the solution given by the unification of the meta language; when
-%we manipulate the terms we don't 
-%
-%Another important implementation should be provided in order to represent the
-%lambda abstractions and the applications of the ML. As outlined in
-%\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
-%<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
-%concept, the latter uses the abstraction and application of the OL.
-%
-%In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
-% The
-% node \elpiIn{val} contains terms of the object language. The node for variable
-% application is not really necessary, since, using the syntax of the ML, it is
-% defined by putting two terms of the ML side by side. For example, the terms 
-% <<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.
-
-The memory \Fo{} is called \elpiIn{fsubst} while the one of \Ho is
-called \elpiIn{subst}.
+We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
+the one of \Ho.
 Both have the invariant that they are not cyclic, TODO explain.
 
-\subsection{equational theory and unification}
+\subsection{Equational theory and Unification}
 
@@
 
-\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}
+\paragraph{Term dereferencing: $\rho s$ and $\sigma t$}
 
-Since in our encoding we explicitly drag substitution mappings, we need to
+Since in our encoding we explicitly carry a substitution we need to
+define the operation that applies it to a term. Its definition
+takes care to normalize (flatten) applications, for example
+it turns
+\elpiIn{(app [app [con"f",con"a"],con"b"])}
+into
+\elpiIn{(app [con"f",con"a",con"b"])}.
+
 dereference variables of the two languages. This is particularly useful to check
@@
 type fderef fsubst -> fm -> fm -> o.
-fderef S (fapp [fuva N|A]) R :- set N S (fapp B),
+% flattening
+fderef S (fapp [fuva N|A]) R :- set? N S (fapp B), !,
   append A B C, fderef S (fapp C) R.
-fderef S (fapp [fuva N|A]) (fapp [fuva N|B]) :- unset N S,
-  forall2 (fderef S) A B.
-fderef S (fapp A) (fapp B) :- forall2 (fderef S) A B.
+% traversal
+fderef S (fapp A) (fapp B) :- map (fderef S) A B.
 fderef S (flam F) (flam G) :- 
@@
 fderef _ (fcon C) (fcon C).
-fderef S (fuva N) R :- set N S T, fderef S T R.
+% dereferencing
+fderef S (fuva N) R :- set? N S T, fderef S T R.
 fderef _ (fuva N) (fuva N) :- unset N S.
@@
 
-More attention should be payed when the dereferencing encounters a subterm that
-is unification variable. The following new rules should be added to the previous
-code snipper to complete the dereferencing of variables of the OL:
+\noindent
+We use the cut operator to keep the code compact. It is possible to rewrite
+the rule for application traversal so that it is mutually exclusive with the
+first one, but that requires a rather verbose analysis of the head of \elpiIn{A}.
 
-A very similar implementation for the dereferencing performed by the ML, given 
-in the appendix, can be encoded respecting the signature:
+The corresponding code for \Ho is similar, we only show the last two rules that
+differ in a substantial way:
 
-\marginpar{what about app uva?}
 \begin{elpicode}
 type deref subst -> tm -> tm -> o.
-  % ... the same
-  deref S (uva N A) R :- set N S T,
-    instantiate T A T', deref S T' R.
-  deref S (uva N A) (uva N B) :-unset N S,
-    forall2 (deref S) A B.
+% ... similar to above ...
+deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
+deref S (uva N A) (uva N B) :- unset N S, map (deref S) A B.
+type move assignment -> list tm -> tm -> o.
+move (abs F) [X | L] R :- move (F X) L R.
+move (val A)         [] A :- !.
+move (val (app A))   L (app X) :- append A L X.
+move (val (uva N A)) L (uva N X) :- append A L X.
+move (val (con H))   L (app [con H | L]).
 \end{elpicode}
 
-In particular, 1) if the variable \elpiIn{fuva N} is assigned to a term
-\elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
-dereferencing \elpiIn{T}; 2) if the current term is the application of a
-variable \elpiIn{fuva N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
-is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
-reduction of \elpiIn{T} to \elpiIn{L}. This is due to the semantics of the
-unification variables within the OL; 3) finally, if \elpiIn{uv N} is not
-assigned, then the same variable is returned.
+\noindent
+Note that when the substitution \elpiIn{S} maps a unification variable
+\elpiIn{N} to an assignment \elpiIn{F} we ....
 
+Important!!!
 A different reasoning is to be addressed to the variables of the ML. Firstly, a
 meta variable cannot appear in the \elpiIn{app} node as the first element of the
-list, we will explain why in \cref{sec:compilation}; secondly, variables in the
-ML have a scope which should be taken into account. The following few lines of
-code outline the rules of dereferencing for the ML:
-
-where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
-return the dereferencing of the beta reduction \elpiIn{T} to the scope
-\elpiIn{L}.
+list, we will explain why in \cref{sec:compilation}
 
@@
     % congruence
-    fapp A ~\Eo~fapp B :- forall2 ~(\Eo)~ A B.
+    fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
     flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
@@
     ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-    app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
+    app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
     lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
     con C ~\Ee~fcon C.
-    uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
+    uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
   \end{elpicode}
@@
 
-\begin{figure}
 \begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst o}~.
-% Congruence
-(app A ~\Ue~app B) S S1 :- fold2 ~\Ue~A B S S1.
-(lam F ~\Ue~lam G) S S1 :- 
-  pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
-~\Ue~(con C) (con C) S S.
-% deref
-(uva N A ~\Ue~T) S S1 :- set N S F,
-  beta-abs F A T1, (T1 ~\Ue~T) S S1.
-(T ~\Ue~uva N A) S S1 :- set N S F,
-  beta-abs F A T1, (T ~\Ue~T1) S S1.
-% flex-flex
-(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
-  pattern-fragment A, pattern-fragment B, 
-  new S W S1, prune W Args1 B Ass, 
-  assign N S1 Ass S2, assign M S2 Ass S3.
-% assignment
-(uva N A ~\Ue~T) S S1 :- pattern-fragment A,
-  not (T = uva _ _), not_occ N S T,  
-  bind A T T1, assign N S T1 S1.
-(T ~\Ue~uva N A) S S1 :- pattern-fragment A,
-  not (T = uva _ _), not_occ N S T, 
-    bind A T T1, assign N S T1 S1.
-%  Note: (a infix b) c d === (infix) a b c d
-
+~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
 \end{elpicode}
-\caption{The \Ue predicate // TODO move to the appendix}
-\label{code:ue-pred}
-\Description[code:unif-pred]{code:unif-pred}
-\end{figure}
 
 On the other hand, unification in the ML needs to be defined. In
-\cref{code:ue-pred}, we give an implementation of \Ue. This predicate has four
+\cref{code:ue-pred}, we give an implementation of \Ue but
+that is actually what our meta language provides as a builtin.
+
+This predicate has four
 arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
@@
 
+\clearpage
+\section*{Appendix}
+
+Note that \elpiIn{(a infix b) c d} de-sugars to \elpiIn{(infix) a b c d}.
+
+\begin{figure*}
+\begin{elpicode}
+~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
+% Congruence
+(app A ~\Ue~app B) S S1 :- fold2 ~(\Ue)~A B S S1.
+(lam F ~\Ue~lam G) S S1 :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
+~\Ue~(con C) (con C) S S.
+% deref
+(uva N A ~\Ue~T) S S1 :- set? N S F, move F A T1, (T1 ~\Ue~T) S S1.
+(T ~\Ue~uva N A) S S1 :- set? N S F, move F A T1, (T ~\Ue~T1) S S1.
+% flex-flex
+(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
+  distinct-names A, distinct-names B, 
+  new S W S1, prune W Args1 B Ass, 
+  assign N S1 Ass S2, assign M S2 Ass S3.
+% assignment
+(uva N A ~\Ue~T) S S1 :- distinct-names A, not (T = uva _ _), not_occ N S T,  
+  bind A T T1, assign N S T1 S1.
+(T ~\Ue~uva N A) S S1 :- distinct-names A, not (T = uva _ _), not_occ N S T, 
+  bind A T T1, assign N S T1 S1.
+
+type distinct-names list tm -> o.
+distinct-names [].
+distinct-names [X|XS] :- name X, not(mem X XS),
+distinct-names XS.
+
+typeabbrev memory A (list (option A)).
+type set? nat -> memory A -> A -> o.
+set? N S T :- nth N S (some T).
+type unset nat -> memory A -> o.
+unset N S :- nth N S none.
+type assign nat -> memory A -> A -> memory A -> o.
+assign z [none|M] T [some T|M].
+assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
+kind nat type.
+type z nat.
+type s nat -> nat.
+type nth nat -> list A -> A -> o.
+nth z [X|_] X.
+nth (s N) [_|L] X :- nth N L X.
+
+type new memory A -> nat -> memory A -> o.
+type prune .
+type move .
+type beta.
+TODO
+
+\end{elpicode}
+\caption{Implementation of the \Ue predicate for \Ho}
+\label{code:ue-pred}
+\Description[code:unif-pred]{code:unif-pred}
+\end{figure*}
+
 \end{document}

### 2024-02-27 paper/main.tex
wip

@@
 \noindent
-The \elpiIn{name} builtin predicate tests if a term is a bound variable. The
+The \elpiIn{name} builtin predicate tests if a term is a bound variable.
+\footnote{one could always load name x for every x under a pi and get rid of the name builtin}
+The
 compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
@@
 type set?   nat -> memory A -> A -> o.
-type unset  nat -> memory A -> o.
+type unset? nat -> memory A -> o.
 type assign nat -> memory A -> A -> memory A -> o.
@@
 fderef S (fuva N) R :- set? N S T, fderef S T R.
-fderef _ (fuva N) (fuva N) :- unset N S.
+fderef S (fuva N) (fuva N) :- unset? N S.
 \end{elpicode}
@@
 deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
-deref S (uva N A) (uva N B) :- unset N S, map (deref S) A B.
+deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
 type move assignment -> list tm -> tm -> o.
@@
 % flex-flex
-(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
+(uva N A ~\Ue~uva M B) S S3 :-  unset? M, unset? N,
   distinct-names A, distinct-names B, 
@@
 set? N S T :- nth N S (some T).
-type unset nat -> memory A -> o.
-unset N S :- nth N S none.
+type unset? nat -> memory A -> o.
+unset? N S :- nth N S none.
 type assign nat -> memory A -> A -> memory A -> o.
@@
 type beta.
+type not_occ.
 TODO

### 2024-02-27 paper/main.tex
wip

@@
 \begin{elpicode}
-type fderef fsubst -> fm -> fm -> o.
+type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fderef}{\rho s}$)~
 % flattening
@@
 \begin{elpicode}
-type deref subst -> tm -> tm -> o.
+type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
 % ... similar to above ...
@@
 
-\begin{figure}
 \begin{elpicode}
-    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.
-    % congruence
+  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
   fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
@@
   fuva N ~\Eo~fuva N.
-    % eta
-    flam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
-      x ~\Eo~x => F x ~\Eo~T' x.
-    T ~\Eo~flam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
-      x ~\Eo~x => T' x ~\Eo~F x.
-    % beta
-    fapp [flam X | TL] ~\Eo~T :- beta (flam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
-    T ~\Eo~fapp [flam X | TL] :- beta (flam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
+  flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
+    pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
+  T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
+    pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
+  fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
+  T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
 \end{elpicode}
-  \caption{Equal predicate OL}
-  \label{code:eq-pred}
-  \Description[code:equal-pred]{code:equal-pred}
-\end{figure}
 
@@
 \otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
-\otext{Example: \elpiIn{lam x\ app[uv 0, x] ~\Ue~lam x\ c"f"}}
+\otext{Example needing the compiler v0 (tra l'altro lo scope è ignorato):\\ \elpiIn{lam x\ app[con"g",app[uv 0, x]] ~\Uo~lam x\ app[con"g", c"a"]}}
 \otext{Links used to instantiate vars of elpi}
@@
 
+decomp che mappa abs verso lam
+\noindent
+\otext{An other example: \\
+  \elpiIn{lam x\ app[f, app[X, x]] = Y, (lam x\ x) = X.}}
+
 \section{Use of multivars}
 
+Se il termine initziale è della forma
+
+\begin{elpicode}
+  app[con"xxx", (lam x\ lam y\ Y y x), (lam x\ f)]
+  =
+  app[con"xxx",X,X]
+\end{elpicode}
+
+allora se non uso due X diverse non ho modo di recuperare il quoziente che mi manca.
+
+a sto punto consideriamo liste di problemi e così da eliminare sta xxx senza
+perdità di generalità (e facciamo problemi più corti, e modellizziamo anche la
+sequenza)
+
 \subsection{Problems with $\eta$}
-\otext{The following goal: \elpiIn{X = lam x\ lam y\ Y x y, X = lam x\ f}}
+\otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
 \otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
 \otext{It is not doable, with the same elpi var}
-
-\noindent
-\otext{An other example: \\
-  \elpiIn{@lam x\ @app[@f, @app[@X, x]] = @Y, (@lam x\ x) = @X.}}
 
@@
   triple ok @Y @f
+\end{elpicode}
+
+\begin{elpicode}
+% @okl 22 [
+%   triple ok (@lam x\ @lam y\ @app[@Y, y, x]) @X,
+%   triple ok (@lam x\ @f) @X,
+% ].
 \end{elpicode}

### 2024-02-28 paper/main.tex
wip

@@
 Unfortunately this translation of rule \ref{clause:r3} uses the
-predicate \coqIn{P} as a first order term for the meta
-language (its type is \elpiIn{tm}).
+predicate \coqIn{P} as a first order term: for the meta
+language its type is \elpiIn{tm}.
 If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
-\ref{goal:g}:
+\ref{goal:g} given below
 
@@
 \noindent
-Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
-in its scope, the unification problem \ref{problem:pa} admits one solution:
+Since \elpiIn{Pm} is an higher-order unification variable,
+of type \elpiIn{tm -> tm},
+with \elpiIn{x}
+in its scope, the unification problem \ref{problem:pa}
+admits one solution:
 
@@
 the premise <<\elpiIn{link Pm A P}>> that is in charge of bringing the
-assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
-back to the domain of Coq terms:
+assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:
 
@@
 In this paper we discuss alternative encodings of Coq in
-Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
+Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
 in which the problems sketched here can be fully described.
@@
 \begin{elpicode}
-app [F, con"a"] = app[F, con"a", con"a"]
+app [F, con"a"] = app[con"f", con"a", con"a"]
 \end{elpicode}
@@
 \begin{elpicode}
-F = lam x\app[const f,x,x]
+F = lam x\ app[con"f",x,x]
 \end{elpicode}
@@
 becomes solvable in DTT trivially, since the term is ground (hence in \llambda),
-but is one substitutes F in the LHS does not find, structurally, the RHS
+but if one substitutes \elpiIn{F} in the LHS does not find,
+structurally, the RHS
 hence \Ue would fail (since \Ue does not know about the $\beta$ rule of DTT).

### 2024-03-13 paper/main.tex
manual rebase of main.tex from main

@@
 addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
-are compared up to definition unfolding and fixpoint unrolling. Still,
+are compared up to proposition unfolding and fixpoint unrolling. Still,
 for efficiency and predictability reasons, most form of automatic proof search
@@
 just $\eta\beta$, and that solves higher-order problems
-restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
+restricted to the pattern fragment $\llambda$~\cite{miller92jsc}.
 We call this unification procedure \Uo{}.
@@
 unification procedure \Ue{} solves higher-order problems in
-$\mathcal{L}_\lambda$.
+$\llambda$.
 
@@
 \newcommand{\specunif}[3]{
-  #3_i \in \mathcal{L}_\lambda \Rightarrow %
+  #3_i \in \llambda \Rightarrow %
     \exists \rho, %
@@
 
-Given 
+We represent a logic program run for a goal g as
+a list steps $p$ of length $\mathcal{N}$ each made of a unification problem
+between terms $\mathcal{S}_{p_l}$ and $\mathcal{S}_{p_r}$
+taken from the set of all terms $\mathcal{S}$.
+If a rule is used
+twice in a run we make two copies of the terms (the predicates)
+occurring in the head or premises. So a run is the conjunction of all
+unification problems, where $\rho_0$ is the empty substitution
+%
 $$
-\C{s_1}{t_1}{l_1} \land \C{s_2}{t_2}{l_2}
+\bigwedge_{p = 1 \ldots \mathcal{N}}
+\underbrace{
+  \rho_{(p-1)}\mathcal{S}_{p_l} \Uo \rho_{(p-1)}\mathcal{S}_{p_l} \mapsto \rho_p
+}_\mathrm{step}
 $$
-we define
+
+We simulate this run as where $\sigma_0$ is the empty substitution
+%
+\newcommand{\solve}{\ensuremath{\mathrm{check}}}
+\begin{gather*}
+\mathcal{T} \times \mathbb{L} = \{ (t_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{l_j} \}\\
+\bigwedge_{p = 1 \ldots \mathcal{N}}
+ \underbrace{
+  \sigma_{(p-1)}\mathcal{T}_{p_l} \Ue \sigma_{(p-1)}\mathcal{T}_{p_r} \mapsto \sigma_p' \land
+  \solve~(\mathbb{L},\sigma_p')\mapsto \sigma_p
+ }_\mathrm{step'}\\
+\D{\sigma_{\mathcal{N}}}{\mathbb{L}}{\rho_{\mathcal{N}}}
+\end{gather*}
+
+\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
+  $\forall p, \mathrm{step}(p,\rho) \mapsto \rho' \Leftrightarrow
+              \mathrm{step'}(p,\sigma) \mapsto \sigma'$
+\end{proposition}
+
+\newcommand{\maybebeta}{\ensuremath{\stackrel{?}{\beta}}\xspace}
+\begin{definition}[\maybebeta]
+  $t$ is in \maybebeta if it contains a subterm matching the
+  pattern \elpiIn{app[uva _ _|_]}
+\end{definition}
+
+\newcommand{\maybeeta}{\ensuremath{\stackrel{?}{\eta}}\xspace}
+\begin{definition}[\maybeeta]
+  $t$ is in \maybeeta if it contains a subterm matching the
+  pattern \elpiIn{lam x\ .. lam y\ app[con _,x, ...,uva _ [y]]}\\
+   or
+  \elpiIn{lam x\lam y\uva _ [y,x]}
+\end{definition}
+
+\begin{definition}[Normal form]
+  We write $norm(\mathcal{T})$ when $\mathcal{T} \cap (\maybebeta \cup \maybeeta) = \emptyset$
+\end{definition}
+
+We write $\sigma\mathcal{T} = \{ \sigma t_i | t_i \in \mathcal{T}\}$.
+
+\begin{proposition}[Normal form preservation]
+  $norm(\sigma\mathcal{T}) \land \mathrm{step'}(p,\sigma) \mapsto \sigma'
+  \Rightarrow norm(\sigma\mathcal{T})$   
+\end{proposition}
+
+If we focus on a single unification problem, that is
+$\mathcal{S} = \{ s_1, s_2 \}$ and 
+$\C{s_1}{t_1}{l_1}$ and $\C{s_2}{t_2}{l_2}$ we write
 $$
-s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}
-    t_1 \Ue t_2 \mapsto \sigma \land
-    \D{\sigma}{l_1 + l_2}{\rho}
+\begin{array}{l}
+s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}\\
+\quad    t_1 \Ue t_2 \mapsto \sigma' \land
+    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land
+    \D{\sigma''}{\{l_1,l_2\}}{\rho}
+\end{array}
 $$
-Where $l_1 + l_2$ is the list concatenation of links.
-
-We write $s \in \mathcal{L}_\lambda$ if all unif variables in
+\noindent
+We write $s \in \llambda$ if all unif variables in
 $s$ are applied to distinct bound variables.
-%
+
+\begin{proposition}[Properties of \Uo{}]
 \begin{gather}
@@
   \rho' \subseteq \rho \Rightarrow
-  \rho's_i \in \mathcal{L}_\lambda \Rightarrow
+  \rho's_i \in \llambda \Rightarrow
   \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
 \end{gather}
+\end{proposition}
 
@@
 
-Property \ref{prop:complete} is also relevant to use \Uo for logic programming.
+Property \ref{prop:fidelity} is also relevant to use \Uo for logic programming.
 In particular we want failures to occur as early as possible, so want
-the decomp phase to take place immediately after \Ue, and fail if need be.
-This becomes particularly important since compile may introduce two
-ho variables for the same fo one, leaving the task of unifying the solutions
-to decomp.
+the check phase to take place immediately after \Ue, and fail if need be.
+
+\paragraph{The intuition} is that a term s is compiled in a term t where any
+``problematic'' sp is replaced by a hole hp and an accessory
+(suspended) unification problem tp \Ue hp (called link) where
+tp is the compiled of sp. 
+On t \Ue is well behaved (captures \Eo), and check solves links when it can by
+using \Ue or keeps them suspended until the decompilation phase.
+check also detects occur checks trough the suspended unif problems
+in order to grant \cref{prop:fidelity}.
 
@@
 We say that the unification variable \elpiIn{uva N L} is in \llambda iff
-\elpiIn{distinct-names L} holds.
-
-\begin{elpicode}
-type distinct-names list tm -> o.
-distinct-names [].
-distinct-names [X|XS] :- name X, not(mem X XS),
-  distinct-names XS.
+\elpiIn{distinct L} holds.
+
+\begin{elpicode}
+type distinct list A -> o.
+distinct [].
+distinct [X|XS] :- name X, not(mem X XS),
+  distinct XS.
 \end{elpicode}
@@
 ML following respectively their specification. In the following few paragraphs
-we give a brief definition of them.
+we give a brief proposition of them.
 
@@
 Since in our encoding we explicitly carry a substitution we need to
-define the operation that applies it to a term. Its definition
+define the operation that applies it to a term. Its proposition
 takes care to normalize (flatten) applications, for example
@@
 \begin{elpicode}
-type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fderef}{\rho s}$)~
-% flattening
-fderef S (fapp [fuva N|A]) R :- set? N S (fapp B), !,
-  append A B C, fderef S (fapp C) R.
-% traversal
-fderef S (fapp A) (fapp B) :- map (fderef S) A B.
-fderef S (flam F) (flam G) :- 
-  pi x\ fderef S x x => fderef S (F x) (G x).
-fderef _ (fcon C) (fcon C).
-% dereferencing
-fderef S (fuva N) R :- set? N S T, fderef S T R.
-fderef S (fuva N) (fuva N) :- unset? N S.
-\end{elpicode}
-
-\noindent
+type fder fsubst -> fm -> fm -> o.
+fder S (fapp A) (fapp B) :- map (fder S) A B.
+fder S (flam F) (flam G) :- 
+  pi x\ fder S x x => fder S (F x) (G x).
+fder _ (fcon C) (fcon C).
+fder S (fuva N) R :- set? N S T, fder S T R.
+fder S (fuva N) (fuva N) :- unset? N S.
+
+type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fder}{\rho s}$)~
+fderef S T R :- fder S T T', napp T' R.
+
+type napp fm -> fm -> o.
+\end{elpicode}
+
+\noindent
+TODO explain napp.
 We use the cut operator to keep the code compact. It is possible to rewrite
@@
 type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
-% ... similar to above ...
+deref S (app A) (app B) :- map (deref S) A B.
+deref S (lam F) (lam G) :- 
+  pi x\ deref S x x => deref S (F x) (G x).
+deref _ (con C) (con C).
 deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
@@
 \noindent
+TODO: no need to napp, see the beta section.
 Note that when the substitution \elpiIn{S} maps a unification variable
@@
 
-\section[Compilation: fo\_tm to tm]{Compilation}
+\section[Compilation: fo\_tm to tm]{Basic compilation \Fo{} to \Ho{}}
 \label{sec:compilation}
+
+In this section we describe a basic compilation scheme that we refine
+later, in the following sections.
+
+The objective of the compilation is to recognize the higher-order variables
+available in \Ho{} when expressed in a first order way in \Fo{}. The compiler
+also generates a list of links that are used to bring back the substitution from
+\Ho{} to \Fo{} and allocates in the memory a cell for each variable.
+
+\begin{elpicode}
+kind link type.
+type link nat -> nat -> nat -> subst. % link Fo Ho Arity
+typeabbrev links list link.
+type comp fm -> tm -> links -> links -> subst -> subst -> o.
+comp (fcon X) (con X) L L S S.
+comp (flam F) (lam G) K L R S :- pi x y\
+  (pi A S\ comp x y L L S S) => comp (F x) (G y) K L R S.
+comp (fuva M) (uva N []) K [link M N z|K] R S :- new R N S.
+comp (fapp[fuva M|A]) (uva N B) K L R S :- distinct A, !,
+  fold4 comp A B K K R R,
+  new R N S, len A Arity,
+  L = [link N M Arity | K].
+comp (fapp A) (app B) K L R S :- fold4 comp A B K L R S.
+\end{elpicode}
+
+\noindent
+Note that \elpiIn{link} carries the arity (number of expected arguments) of
+the variable. \marginpar{say when this is needed}
+
+\begin{elpicode}
+type solve-links links -> links -> subst -> subst -> o.
+solve-links L L S S.
+\end{elpicode}
+
+Then decomp
+
+\begin{elpicode}
+type decompile links -> subst -> fsubst -> o.
+decompile L S O :-
+  map (_\r\r = none) S O1, % allocate empty fsubst
+  (pi N X\ knil N X :- mem L (link X N _) ; N = X) =>
+    decompl S L O1 O.
+type knil nat -> nat -> o.
+
+type decompl links -> subst -> fsubst -> o.
+decompl S [] [].
+decompl S [link _ N _|L] O P :- unset? N S X,
+  decompl S L O P.
+decompl S [link M N _|L] O P :- set? N S X,
+  decomp-assignment S X T, assign M O (some T) O1,
+  decompl S L O1 P.
+
+type decomp-assignment subst -> assignment -> fm -> o.
+decomp-assignment S (abs F) (flam G) :-
+  pi x y\ decomp-tm S x y => decomp-assignment S (F x) (G y).
+decomp-assignment S (val T) T1 :- decomp S T T1.
+
+type decomp subst -> tm -> fm.
+decomp _ (con C) (fcon C).
+decomp S (app A) (app B) :- map (decomp S) A B.
+decomp S (lam F) (flam G) :-
+  pi x y\ decomp S x y => decomp S (F x) (G y).
+decomp S (uva N A) R :- set? N S F,
+  move F A T, decomp S T R.
+decomp S (uva N A) R :- unset? N S,
+  map (decomp S) A B, knil N M, napp (fapp[fuva M|B]) R.
+\end{elpicode}
+
+\marginpar{TODO link}
+\marginpar{TODO nuove subst}
+
+Now unif\marginpar{TODO: code unif}
+
+\begin{elpicode}
+~\PYG{k+kd}{type} \PYG{n+nf}{(\Uo)} \PYG{k+kt}{fm -> fm -> subst -> subst -> o}~.
+(X ~\Uo~Y) S S1 :-
+  fderef S X X0, fderef S Y Y0,                    ~($\customlabel{step:norm}{norm}$)~
+  comp X0 X1 [] S0 [] L0,                        ~($\customlabel{step:compile}{compile}$)~
+  comp Y0 Y1 S0 S1 L0 L1,
+  (X1 ~\Ue~Y1) [] HS0,                              ~($\customlabel{step:unify}{unify}$)~
+  solve-links L1 L2 HS0 HS1,                        ~($\customlabel{step:link}{link}$)~
+  decompile L2 HS1  S1.                        ~($\customlabel{step:decompile}{decompile}$)~
+\end{elpicode}
+
+\subsection{Prolog simulation}
+Allows us to express the properties.
+we take all terms involved in a search (if a rule is used twice we simply
+take a copy of it), we compile all of them, and then we pick the unification
+prblems among these terms and step trough them.
+
+\begin{elpicode}
+type pick list A -> (pair nat nat) -> (pair A A) -> o.
+pick L (pr X Y) (pr TX TY) :- nth X L TX, nth Y L TY.
+
+type prolog-fo list fm -> list (pair nat nat) -> subst -> o.
+prolog-fo Terms Problems S :-
+  map (pick Terms) Problems FoProblems,
+  fold4 ~(\Uo)~ FoProblems [] S.
+
+type step-ho (pair tm tm) -> links -> links -> subst -> subst -> o.
+step-ho (pr X Y) L0 L1 S0 S2 :-
+  (X1 ~\Ue~Y1) S0 S1,
+  solve-links L0 L1 S1 S2.
+
+type prolog-ho list fm -> list (pair nat nat) -> subst -> o.
+prolog-ho Terms Problems S :-
+  fold4 comp Terms HoTerms [] L0 [] HS0,
+  map (pick HoTerms) Problems HoProblems,
+  fold4 step-ho HoProblems L0 L HS0 HS,
+  decompile L HS S.
+\end{elpicode}
+
+\noindent
+the proprty is that if a step for Fo succeds then the Ho one does, and
+if Fo fails then the Ho fails ()
+
+\subsection{Example}
+
+OK
+
+\begin{elpicode}
+Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]] 
+      , flam x\ fapp[fcon"g", fcon"a"] ]
+Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$ 
+lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
+link z z (s z)
+HS = [some (abs x\con"a")]
+S = [some (flam x\fcon a)]
+\end{elpicode}
+
+KO
+
+\begin{elpicode}
+  Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]] 
+  , flam x\ fapp[fcon"g", fcon"a"] ]
+Problems = [ pr 0 1   % $A = \lambda x.x$
+           , pr 2 3 ] % $A a = a$ 
+lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
+link z z (s z)
+HS = [some (abs x\con"a")]
+S = [some (flam x\fcon a)]
+lam x\ app[f, app[X, x]] = Y,
+  lam x\ x) = X.
+\end{elpicode}
 
@@
 \begin{elpicode}
+~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
+fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
+flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
+fcon C ~\Eo~fcon C.
+fuva N ~\Eo~fuva N.
+flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
+  pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
+T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
+  pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
+fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
+T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
+
+type beta fm -> list fm -> fm -> o.
+beta A [] A.
+beta (flam F) [H | L] R :- subst F H B,
+  beta B L R. % since F could be x\app[x|_] and H be lam _ 
+beta (fapp A) L (fapp X) :- append A L X.
+beta (fuva N) L (fapp [fuva N | L]).
+beta (fcon H) L (fapp [fcon H | L]).
+
+type subst (fm -> fm) -> fm -> fm -> o.
+subst F H B :- napp (F H) B. % since (F H) may generate (app[app _|_])
+
+type napp fm -> fm -> o.
+napp (fcon C) (fcon C).
+napp (flam F) (flam G) :- pi x\ napp x x => napp (F x) (G x).
+napp (fapp[fapp L|M]) R :- !, append L M N, napp (fapp N) R.
+napp (fapp[X]) R :- !, napp X R.
+napp (fapp A) (fapp B) :- map napp A B.
+napp (fuva N) (fuva N).
+\end{elpicode}
+\caption{Full implementation of the \Eo predicate for \Fo}
+\label{code:eo-pred}
+\Description[code:eo-pred]{code:eo-pred}
+\end{figure*}
+
+\begin{figure*}
+\begin{elpicode}
 ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
 % Congruence
-(app A ~\Ue~app B) S S1 :- fold2 ~(\Ue)~A B S S1.
-(lam F ~\Ue~lam G) S S1 :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
+(app A ~\Ue~app B) R S :- fold2 ~(\Ue)~A B R S.
+(lam F ~\Ue~lam G) R S :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) R S.
 ~\Ue~(con C) (con C) S S.
 % deref
-(uva N A ~\Ue~T) S S1 :- set? N S F, move F A T1, (T1 ~\Ue~T) S S1.
-(T ~\Ue~uva N A) S S1 :- set? N S F, move F A T1, (T ~\Ue~T1) S S1.
+(uva N A ~\Ue~T) R S :- set? N S F, move F A T1, (T1 ~\Ue~T) R S.
+(T ~\Ue~uva N A) R S :- set? N S F, move F A T1, (T ~\Ue~T1) R S.
 % flex-flex
 (uva N A ~\Ue~uva M B) S S3 :-  unset? M, unset? N,
-  distinct-names A, distinct-names B, 
+  distinct A, distinct B, 
   new S W S1, prune W Args1 B Ass, 
@@
 % assignment
-(uva N A ~\Ue~T) S S1 :- distinct-names A, not (T = uva _ _), not_occ N S T,  
+(uva N A ~\Ue~T) R S :- distinct A, not (T = uva _ _), not_occ N S T,  
   bind A T T1, assign N S T1 S1.
-(T ~\Ue~uva N A) S S1 :- distinct-names A, not (T = uva _ _), not_occ N S T, 
+(T ~\Ue~uva N A) R S :- distinct A, not (T = uva _ _), not_occ N S T, 
   bind A T T1, assign N S T1 S1.
 
-type distinct-names list tm -> o.
-distinct-names [].
-distinct-names [X|XS] :- name X, not(mem X XS),
-distinct-names XS.
+type distinct list A -> o.
+distinct [].
+distinct [X|XS] :- name X, not(mem X XS),
+distinct XS.
 
@@
 type new memory A -> nat -> memory A -> o.
+new [] z [none].
+new [X|XS] (s N) [X|YS] :- new XS N YS.
+
 type prune .
@@
 type beta.
+type bind.
 type not_occ.
@@
 
+type fold2 (A -> A1 -> B -> B -> o) -> list A -> list A1 -> B -> B -> o.
+fold2 _ [] [] A A.
+fold2 F [X|XS] [Y|YS] A A1 :- F X Y A A0, fold2 F XS YS A0 A1.
+
 \end{elpicode}
@@
 \label{code:ue-pred}
-\Description[code:unif-pred]{code:unif-pred}
+\Description[code:ue-pred]{code:ue-pred}
 \end{figure*}
 
+\begin{figure*}
+\begin{elpicode}
+type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
+fold4 _ [] [] A A B B.
+fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.
+
+type len list A -> nat -> o.
+len [] z.
+len [_|L] (s X) :- len L X.
+
+\end{elpicode}
+\caption{Implementation of the compiler}
+\label{code:ue-pred}
+\Description[code:ue-pred]{code:ue-pred}
+\end{figure*}
+
 \end{document}

### 2024-03-13 paper/elpi.py
update lexer elpi

@@ -79,7 +79,7 @@ class ElpiLexer(RegexLexer):
              'elpi-chr-rule-start'),
 
             (r"(?=[A-Z_]){}".format(constant_re), Name.Variable),
-            (r"(?=[a-z_]){}\\".format(constant_re), Name.Variable),
+            (r"(?=[a-z_])({}|_)\\".format(constant_re), Name.Variable),
             (r"_", Name.Variable),
             (r"({}|!|=>|;)".format(symbol_re), Keyword.Declaration),
             (constant_re, Text),

### 2024-03-15 paper/main.tex
w

@@
 \noindent
-Since \elpiIn{Pm} is an higher-order unification variable,
+Since \elpiIn{Pm} is an higher-order unification variable
 of type \elpiIn{tm -> tm},
@@
 and we assume that the unification of our meta language is correct:
-
 %
@@
 
-\newcommand{\C}[3]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3)}
+\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
 \newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}
 
-We illustrate a compilation $\C{s}{t}{l}$ that
-maps a term $s$ in \Fo{} to a term $t$ in \Ho and a list of links $l$.
-The links connect unification variables in \Ho with variables
-in \Fo{} and are used to decompile the assignment,
-$\D{\sigma}{l}{\rho}$.
+We illustrate a compilation $\C{s}{t}{m}{l}$ that
+maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
+list of links $l$.
+The variable map connects unification variables in \Ho with variables
+in \Fo{} and is used to ``decompile'' the assignment,
+$\D{\sigma}{m}{\rho}$. Links represent problematic sub-terms which
+are linked to the unification variable that stands in their place in the
+compiled term. These links are checked for or progress XXX improve....
 
@@
 %
+\newcommand{\solve}{\ensuremath{\mathrm{check}}}
+\newcommand{\step}[4]{\ensuremath{\mathrm{step}(#1,#2,#3) \mapsto #4}}
+\newcommand{\stepD}[4]{\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
+\newcommand{\runD}[2]{\ensuremath{\bigwedge_{p = 1 \ldots #2} \step{#1}{p}{\rho_{(p-1)}}{\rho_{p}}}}
+\newcommand{\run}[2]{\ensuremath{\mathrm{run}(#1,#2)}}
+$$
+\step{\mathcal{S}}{p}{\rho}{\rho'} := \stepD{\mathcal{S}}{p}{\rho}{\rho'}
+$$
+$$
+\run{\mathcal{S}}{\mathcal{N}} := \runD{\mathcal{S}}{\mathcal{N}}
+$$
 $$
@@
 %
-\newcommand{\solve}{\ensuremath{\mathrm{check}}}
 \begin{gather*}
-\mathcal{T} \times \mathbb{L} = \{ (t_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{l_j} \}\\
+\mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
 \bigwedge_{p = 1 \ldots \mathcal{N}}
@@
 $\mathcal{S} = \{ s_1, s_2 \}$ and 
-$\C{s_1}{t_1}{l_1}$ and $\C{s_2}{t_2}{l_2}$ we write
+$\C{s_1}{t_1}{m_1}{l_1}$ and $\C{s_2}{t_2}{m_2}{l_2}$ we write
 $$

### 2024-03-15 paper/main.tex
latex

@@
 \newcommand{\solve}{\ensuremath{\mathrm{check}}}
-\newcommand{\step}[4]{\ensuremath{\mathrm{step}(#1,#2,#3) \mapsto #4}}
-\newcommand{\stepD}[4]{\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
-\newcommand{\runD}[2]{\ensuremath{\bigwedge_{p = 1 \ldots #2} \step{#1}{p}{\rho_{(p-1)}}{\rho_{p}}}}
-\newcommand{\run}[2]{\ensuremath{\mathrm{run}(#1,#2)}}
+\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
+\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
+\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
+\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
+\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
+\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
+\newcommand{\stepFD}[4]{\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
+\newcommand{\stepHD}[6]{\ensuremath{%
+#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#4) \mapsto #5}}
+\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
+\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
+\newcommand{\runFD}[2]{\ensuremath{\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
+\newcommand{\runHD}[3]{\ensuremath{\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{\sigma'}{\sigma_{p}}{#3}}}
+\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
 $$
-\step{\mathcal{S}}{p}{\rho}{\rho'} := \stepD{\mathcal{S}}{p}{\rho}{\rho'}
+\begin{array}{l}
+\stepF{\mathcal{S}}{p}{\rho}{\rho'}
+\deff
+\stepFD{\mathcal{S}}{p}{\rho}{\rho'}\\
+\runF{\mathcal{S}}{\mathcal{N}}{\rho}
+\deff
+\runFD{\mathcal{S}}{\mathcal{N}}
+\end{array}
 $$
+
+We simulate this run as where $\sigma_0$ is the empty substitution
 $$
-\run{\mathcal{S}}{\mathcal{N}} := \runD{\mathcal{S}}{\mathcal{N}}
+\begin{array}{l}
+\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\\
+  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\\
+\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \\
+  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
+  \qquad \runFD{\mathcal{S}}{\mathcal{N}}{\mathbb{L}}\\
+  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\rho_{\mathcal{N}}}
+\end{array}
 $$
+
+\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
+In the context of \hrun, we have that
+$\forall p \in 1 \ldots \mathcal{N}$
 $$
-\bigwedge_{p = 1 \ldots \mathcal{N}}
-\underbrace{
-  \rho_{(p-1)}\mathcal{S}_{p_l} \Uo \rho_{(p-1)}\mathcal{S}_{p_l} \mapsto \rho_p
-}_\mathrm{step}
+\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
+\Leftrightarrow
+\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
 $$
-
-We simulate this run as where $\sigma_0$ is the empty substitution
-%
-\begin{gather*}
-\mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
-\bigwedge_{p = 1 \ldots \mathcal{N}}
- \underbrace{
-  \sigma_{(p-1)}\mathcal{T}_{p_l} \Ue \sigma_{(p-1)}\mathcal{T}_{p_r} \mapsto \sigma_p' \land
-  \solve~(\mathbb{L},\sigma_p')\mapsto \sigma_p
- }_\mathrm{step'}\\
-\D{\sigma_{\mathcal{N}}}{\mathbb{L}}{\rho_{\mathcal{N}}}
-\end{gather*}
-
-\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-  $\forall p, \mathrm{step}(p,\rho) \mapsto \rho' \Leftrightarrow
-              \mathrm{step'}(p,\sigma) \mapsto \sigma'$
 \end{proposition}

### 2024-03-15 paper/main.tex
latex

@@
 \newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
-\newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}
+\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
 
@@
 in \Fo{} and is used to ``decompile'' the assignment,
-$\D{\sigma}{m}{\rho}$. Links represent problematic sub-terms which
+$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
 are linked to the unification variable that stands in their place in the
@@
 
-We represent a logic program run for a goal g as
-a list steps $p$ of length $\mathcal{N}$ each made of a unification problem
-between terms $\mathcal{S}_{p_l}$ and $\mathcal{S}_{p_r}$
-taken from the set of all terms $\mathcal{S}$.
-If a rule is used
-twice in a run we make two copies of the terms (the predicates)
-occurring in the head or premises. So a run is the conjunction of all
-unification problems, where $\rho_0$ is the empty substitution
+We represent a logic program \emph{run} in \Fo as
+a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
+unification problem between terms $\mathcal{S}_{p_l}$ and
+$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
+The composition of these steps starting from the
+empty substitution $\rho_0$ produces the final
+substitution $\rho_\mathcal{N}$.
+\footnote{If the same rule is used multiple time in a run we
+just consider as many copies as needed of the terms composing the
+rules, with fresh unification variables each time}
+The initial here $\rho_0$ is the empty substitution
 %
@@
 \newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
+\newcommand{\stepFD}[5]{%
+\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
 \newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
-\newcommand{\stepFD}[4]{\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
 \newcommand{\stepHD}[6]{\ensuremath{%
-#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#4) \mapsto #5}}
+#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
 \newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
+\newcommand{\runFD}[2]{\ensuremath{%
+\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
 \newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
-\newcommand{\runFD}[2]{\ensuremath{\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
-\newcommand{\runHD}[3]{\ensuremath{\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{\sigma'}{\sigma_{p}}{#3}}}
+\newcommand{\runHD}[3]{\ensuremath{%
+\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
@@
 \begin{array}{l}
-\stepF{\mathcal{S}}{p}{\rho}{\rho'}
+\stepF{\mathcal{S}}{p}{\rho}{\rho''}
 \deff
-\stepFD{\mathcal{S}}{p}{\rho}{\rho'}\\
+\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho'}\\
 \runF{\mathcal{S}}{\mathcal{N}}{\rho}
@@
 
-We simulate this run as where $\sigma_0$ is the empty substitution
+We simulate each run in \Fo with a run in \Ho as follows.
+Note that $\sigma_0$ is the empty substitution.
 $$
@@
   \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
-  \qquad \runFD{\mathcal{S}}{\mathcal{N}}{\mathbb{L}}\\
-  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\rho_{\mathcal{N}}}
+  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\\
+  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
 \end{array}
@@
 
+\noindent
+Here \hstep is made of two sub-steps: a call to \Ue (on the compiled
+terms) and a call to \solve on the set of links. We claim the following:
+
+\begin{proposition}[Simulation]\label{prop:sumul}
+$\forall \mathcal{S}, \forall \mathcal{N}$
+$$
+  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
+  \Leftrightarrow
+  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
+$$
+\end{proposition}
+
+\noindent
+That is, the two executions give the same result. Moreover:
+
 \begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of \hrun, we have that
+In the context of~ \hrun, we have that
 $\forall p \in 1 \ldots \mathcal{N}$
@@
 \end{proposition}
-
-\newcommand{\maybebeta}{\ensuremath{\stackrel{?}{\beta}}\xspace}
+\noindent
+Note that this property guarantees that a failure in the \Fo run
+is matched by a failure
+in \Ho
+\emph{at the same step}.
+
+\newcommand{\maybebeta}{\ensuremath{\beta}\xspace}
 \begin{definition}[\maybebeta]
@@
 
-\newcommand{\maybeeta}{\ensuremath{\stackrel{?}{\eta}}\xspace}
+\newcommand{\maybeeta}{\ensuremath{\eta}\xspace}
 \begin{definition}[\maybeeta]
@@
 
-\begin{proposition}[Normal form preservation]
+\begin{proposition}[Normal form preservation]\label{prop:nf}
   $norm(\sigma\mathcal{T}) \land \mathrm{step'}(p,\sigma) \mapsto \sigma'
@@
 
+TODO: check if this is also true for $\rho$ after decomp.
+In particular this guaarantees that is we start from normal terms
+we never introduce eta-long or non-beta-normal terms. This is required in
+order to show the fidelity, since there terms would not be considered
+as equal by \Ue.
+
 If we focus on a single unification problem, that is
-$\mathcal{S} = \{ s_1, s_2 \}$ and 
-$\C{s_1}{t_1}{m_1}{l_1}$ and $\C{s_2}{t_2}{m_2}{l_2}$ we write
+$\mathcal{S} = \{ s_1, s_2 \}$ we can define \Uo as follows:
 $$
 \begin{array}{l}
-s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}\\
+s_1 \Uo s_2 \mapsto \rho \deff\\
+\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\\
 \quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land
-    \D{\sigma''}{\{l_1,l_2\}}{\rho}
+    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\\
+\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
 \end{array}
@@
 \noindent
-We write $s \in \llambda$ if all unif variables in
-$s$ are applied to distinct bound variables.
+Note that proposition \ref{prop:nf} does not hold in this case since
+decompilation can introduce terms in \maybebeta or \maybebeta. Nevertheless
+\Uo has the following properties (where we
+write $s \in \llambda$ if all unif variables in
+$s$ are applied to distinct bound variables):
 
@@
 
-\paragraph{The intuition} is that a term s is compiled in a term t where any
-``problematic'' sp is replaced by a hole hp and an accessory
-(suspended) unification problem tp \Ue hp (called link) where
-tp is the compiled of sp. 
-On t \Ue is well behaved (captures \Eo), and check solves links when it can by
-using \Ue or keeps them suspended until the decompilation phase.
-check also detects occur checks trough the suspended unif problems
+\subsection{The intuition in a nutshell}
+A term $s$ is compiled in a term $t$ where any
+``problematic'' subterm $tp$ is replaced by a unification variable
+$hp$ and an accessory
+(suspended) unification problem $(hp \Ue tp) \in \mathcal{L}$.
+Since $t \in \llambda$, \Ue is well behaved (captures \Eo).
+
+Problematic terms are terms where the head constructor may
+change if instantiated, that is, they are in \maybebeta or \maybeeta.
+On these terms \Ue is weak/incorrect.
+\begin{itemize}
+\item \elpiIn{app[F,con "a"]} may be \Eo to \elpiIn{con "a"}
+\item \elpiIn{lam x\lam y\app[F,y,x]} may be \Eo to \elpiIn{con "f"} 
+\end{itemize}
+
+\noindent
+\solve{} solves links when they are no more ``problematic'', for example
+in the examples above if \elpiIn{F} materializes, or if the
+other side of the link forces a commit. EXPLAIN.
+In these cases the problems are ``resumed'' by using \Ue,
+otherwise they are kept suspended until the
+decompilation phase.
+\solve also detects occur checks trough the suspended unif problems
 in order to grant \cref{prop:fidelity}.
+
+Note tha the first case is outside $\llambda$ but is frequent in practice
+so we handle it. It is also what prop 5 is about.
 

### 2024-03-18 paper/main.tex
2

@@
 \deff
-\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho'}\\
+\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
 \runF{\mathcal{S}}{\mathcal{N}}{\rho}
@@
 
-We simulate each run in \Fo with a run in \Ho as follows.
+We simulate each run in \Fo{} with a run in \Ho as follows.
 Note that $\sigma_0$ is the empty substitution.
@@
 \begin{array}{l}
-\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\\
-  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\\
-\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \\
-  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
-  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\\
+\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\vspace{2pt}\\
+  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\vspace{2pt}\\
+\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
+  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\vspace{2pt}\\
   \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
@@
 \noindent
-Here \hstep is made of two sub-steps: a call to \Ue (on the compiled
-terms) and a call to \solve on the set of links. We claim the following:
+Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
+terms) and a call to \solve{} on the set of links. We claim the following:
 
@@
 \noindent
-Note that this property guarantees that a failure in the \Fo run
-is matched by a failure
-in \Ho
-\emph{at the same step}.
-
-\newcommand{\maybebeta}{\ensuremath{\beta}\xspace}
-\begin{definition}[\maybebeta]
-  $t$ is in \maybebeta if it contains a subterm matching the
-  pattern \elpiIn{app[uva _ _|_]}
-\end{definition}
-
-\newcommand{\maybeeta}{\ensuremath{\eta}\xspace}
-\begin{definition}[\maybeeta]
-  $t$ is in \maybeeta if it contains a subterm matching the
-  pattern \elpiIn{lam x\ .. lam y\ app[con _,x, ...,uva _ [y]]}\\
-   or
-  \elpiIn{lam x\lam y\uva _ [y,x]}
-\end{definition}
-
-\begin{definition}[Normal form]
-  We write $norm(\mathcal{T})$ when $\mathcal{T} \cap (\maybebeta \cup \maybeeta) = \emptyset$
-\end{definition}
-
-We write $\sigma\mathcal{T} = \{ \sigma t_i | t_i \in \mathcal{T}\}$.
-
-\begin{proposition}[Normal form preservation]\label{prop:nf}
-  $norm(\sigma\mathcal{T}) \land \mathrm{step'}(p,\sigma) \mapsto \sigma'
-  \Rightarrow norm(\sigma\mathcal{T})$   
-\end{proposition}
-
-TODO: check if this is also true for $\rho$ after decomp.
-In particular this guaarantees that is we start from normal terms
-we never introduce eta-long or non-beta-normal terms. This is required in
-order to show the fidelity, since there terms would not be considered
-as equal by \Ue.
-
-If we focus on a single unification problem, that is
-$\mathcal{S} = \{ s_1, s_2 \}$ we can define \Uo as follows:
+In particular this property guarantees that a \emph{failure} in the \Fo{} run
+is matched by a failure in \Ho{} \emph{at the same step}. We consider this
+property very important from a practical point of view since it guarantees
+that the execution traces are strongly related and in turn this enables a user
+to debug a logic program in \Fo{} by looking at its execution trace in
+\Ho{}.
+
+We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
+$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
+%
 $$
 \begin{array}{l}
-s_1 \Uo s_2 \mapsto \rho \deff\\
-\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\\
+s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
+\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
 \quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\\
+    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
 \quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
@@
 $$
-\noindent
-Note that proposition \ref{prop:nf} does not hold in this case since
-decompilation can introduce terms in \maybebeta or \maybebeta. Nevertheless
-\Uo has the following properties (where we
-write $s \in \llambda$ if all unif variables in
-$s$ are applied to distinct bound variables):
 
@@
 
-Property \ref{prop:simulation} states that \Uo is resilient to problems outside
-\llambda if a third party provides a (partial) solution for its problem. Since
-we are interested in using \Uo in a proof search procedure, made of a sequence
-of unification problems, not necessarily in \llambda. In practice it is often
-the case the order in which these problems are stated matters. A Typical example
-is the following problem
-
-\begin{elpicode}
-app [F, con"a"] = app[con"f", con"a", con"a"]
-\end{elpicode}
-
-\noindent
-preceded by
-
-\begin{elpicode}
-F = lam x\ app[con"f",x,x]
-\end{elpicode}
-
-\noindent
-becomes solvable in DTT trivially, since the term is ground (hence in \llambda),
-but if one substitutes \elpiIn{F} in the LHS does not find,
-structurally, the RHS
-hence \Ue would fail (since \Ue does not know about the $\beta$ rule of DTT).
-Our compiler takes care of making property \ref{prop:simulation} hold, see
-section XXX.
-
-Property \ref{prop:fidelity} is also relevant to use \Uo for logic programming.
-In particular we want failures to occur as early as possible, so want
-the check phase to take place immediately after \Ue, and fail if need be.
+Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
+is resilient to unification problems outside \llambda solved by
+a third party. We believe this property is of practical interest since we
+want the user to be able to add heuristics via hand written rules
+to the ones obtained by our compilation scheme. A Typical example
+is the following problem \ref{problem:q} that is outside \llambda:
+
+\begin{elpicode}
+app [F, con"a"] = app[con"f", con"a", con"a"]   ~\customlabel{problem:q}{(q)}~
+F = lam x\ app[con"f",x,x]                      ~\customlabel{heuristic:h}{(h)}~
+\end{elpicode}
+
+\noindent
+Instead of rejecting it our scheme accepts it and guarantees that if
+\ref{heuristic:h} is given (after the compilation part of the scheme, as
+a run time hint) then ...
 
@@
 Since $t \in \llambda$, \Ue is well behaved (captures \Eo).
+
+\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
+\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
+\begin{definition}[\maybeeta]
+  $\maybeeta = \{ t | \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
+\end{definition}
+
+\noindent
+An example of term $t$ in \maybeeta{} is
+$\lambda x.\lambda y.F~y~x$
+since the substitution
+$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
+makes $\rho t = \lambda x.\lambda y.f x y$
+that is the eta long form of $f$.
+
+\begin{definition}[\maybebeta]
+  $t$ is in \maybebeta if it contains a subterm matching the
+  pattern \elpiIn{app[uva _ _|_]}
+\end{definition}
+
+\newcommand{\normal}{\ensuremath{\mathrm{normal}}}
+\begin{definition}[Normal form]
+Given a set of terms $X \subseteq \Ho{}$,
+$$
+\normal(X) =
+X \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
+$$
+\end{definition}
+
+\noindent
+We write $\sigma X = \{ \sigma t_i | t_i \in X\}$.
+
+\begin{proposition}[Normal form preservation]\label{prop:nf}
+$\forall \mathcal{T}, \forall \mathbb{L}, \forall p,$
+$$
+\normal(\sigma\mathcal{T}) \land
+\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
+\Rightarrow \normal(\sigma'\mathcal{T})
+$$
+\end{proposition}
+
+TODO: check if this is also true for $\rho$ after decomp.
+In particular this guaarantees that is we start from normal terms
+we never introduce eta-long or non-beta-normal terms. This is required in
+order to show the fidelity, since there terms would not be considered
+as equal by \Ue.
 
@@
 
+Note that proposition \ref{prop:nf} does not hold in this case since
+decompilation can introduce terms in \maybebeta or \maybebeta.
 \section{Alternative encodings and related work}
@@
 distinct [].
-distinct [X|XS] :- name X, not(mem X XS),
-  distinct XS.
+distinct [X|XS] :- name X, not(mem X XS), distinct XS.
 \end{elpicode}
@@
 type move assignment -> list tm -> tm -> o.
-move (abs F) [X | L] R :- move (F X) L R.
+move (abs Bo)        [H|L] R :- move (Bo H) L R.
 move (val A)         []    A :- !.
-move (val (app A))   L (app X) :- append A L X.
-move (val (uva N A)) L (uva N X) :- append A L X.
-move (val (con H))   L (app [con H | L]).
+move (val (uva N A)) L     (uva N X) :- std.append A L X.
 \end{elpicode}

### 2024-03-18 paper/main.tex
more tex

@@
 A term $s$ is compiled in a term $t$ where any
-``problematic'' subterm $tp$ is replaced by a unification variable
-$hp$ and an accessory
-(suspended) unification problem $(hp \Ue tp) \in \mathcal{L}$.
-Since $t \in \llambda$, \Ue is well behaved (captures \Eo).
+``problematic'' subterm $tp$ is replaced by a unification variable $h$
+and an accessory link, that represent a suspended unification problem
+$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
+We now define ``problematic'' formally.
 
@@
 \begin{definition}[\maybebeta]
-  $t$ is in \maybebeta if it contains a subterm matching the
-  pattern \elpiIn{app[uva _ _|_]}
+  $\maybebeta = \{ X t_1 \ldots t_n | t \not\in \llambda \}$.
 \end{definition}
 
+\noindent
+An example of $t$ in \maybebeta{} is $F a$ for a constant $a$. Note however tha
+an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
+that makes the resulting term fall outside of \maybebeta.
+
+\begin{definition}[Subterm $\mathcal{P}(t)$]
+$$
+\begin{array}{l}
+t \in \mathcal{P}(t)\\
+t = f t_1\ldots t_n \Rightarrow \mathcal{P}(t_i) \subseteq \mathcal{P}(t) \land f \in \mathcal{P}(t)\\
+t = \lambda x.t' \Rightarrow \mathcal{P}(t') \subseteq \mathcal{P}(t)\\
+\end{array}
+$$
+\end{definition}
+
+\noindent
+We write $\mathcal{P}(X) = \bigcup_{t\in X} \mathcal{P}(t)$.
+
+\marginpar{TODO: find better word}
 \newcommand{\normal}{\ensuremath{\mathrm{normal}}}
@@
 \normal(X) =
-X \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
+\mathcal{P}(X) \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
 $$
@@
 \noindent
-We write $\sigma X = \{ \sigma t_i | t_i \in X\}$.
+We write $\sigma X = \{ \sigma t | t \in X\}$.
 
 \begin{proposition}[Normal form preservation]\label{prop:nf}
-$\forall \mathcal{T}, \forall \mathbb{L}, \forall p,$
+$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
 $$
@@
 \stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
-\Rightarrow \normal(\sigma'\mathcal{T})
+\Rightarrow \normal((\sigma \cup \sigma')\mathcal{T})
 $$
@@
 
-TODO: check if this is also true for $\rho$ after decomp.
-In particular this guaarantees that is we start from normal terms
-we never introduce eta-long or non-beta-normal terms. This is required in
-order to show the fidelity, since there terms would not be considered
-as equal by \Ue.
-
-Problematic terms are terms where the head constructor may
-change if instantiated, that is, they are in \maybebeta or \maybeeta.
-On these terms \Ue is weak/incorrect.
-\begin{itemize}
-\item \elpiIn{app[F,con "a"]} may be \Eo to \elpiIn{con "a"}
-\item \elpiIn{lam x\lam y\app[F,y,x]} may be \Eo to \elpiIn{con "f"} 
-\end{itemize}
-
-\noindent
-\solve{} solves links when they are no more ``problematic'', for example
-in the examples above if \elpiIn{F} materializes, or if the
-other side of the link forces a commit. EXPLAIN.
-In these cases the problems are ``resumed'' by using \Ue,
-otherwise they are kept suspended until the
-decompilation phase.
-\solve also detects occur checks trough the suspended unif problems
-in order to grant \cref{prop:fidelity}.
-
-Note tha the first case is outside $\llambda$ but is frequent in practice
-so we handle it. It is also what prop 5 is about.
-
-Note that proposition \ref{prop:nf} does not hold in this case since
-decompilation can introduce terms in \maybebeta or \maybebeta.
+\noindent
+In particular this guarantees that is we start from normal terms
+we never introduce eta-long or non-beta-normal terms in $\sigma'$.
+
+Note that proposition \ref{prop:nf} does not hold for \Uo{} since
+decompilation can introduce (actually restore) terms in
+\maybeeta or \maybebeta.
+
 \section{Alternative encodings and related work}

### 2024-03-18 paper/main.tex
review mode
latex

@@
-\documentclass[sigconf,natbib=false]{acmart}
+\documentclass[sigconf,natbib=false,review]{acmart}
 \usepackage[]{biblatex}
@@
 are first class terms, i.e. they have a concrete syntax. We keep these languages
-minimal, for example we omit the \elpiIn{all} quantifier of DTT we used
-in the example in Section~\ref{sec:intro}. 
+minimal, for example, we omit the \elpiIn{all} quantifier of DTT we used
+in the example in Section~\ref{sec:intro} together with the type notation of
+terms carried by the \elpiIn{lam} constructor. 
 %
@@
 In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
-Unification variables standing for functions are applied their arguments
-via the \elpiIn{fapp} constructor.
-For example in the statement of the instance \coqIn{forall_dec} the
-term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N}
-is of type \elpiIn{nat} and \elpiIn{x} is a bound variable.
+the arguments of an higher order unification variable are via the \elpiIn{fapp}
+constructor. For example in the statement of the instance \coqIn{forall_dec} the
+term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is
+a memory address and \elpiIn{x} is a bound variable.
 
@@
 compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
-reasons, so we don't assume all out terms are in \llambda but rather test.
+reasons, so we don't assume all out terms are in \llambda but rather test. \textbf{\textcolor{red}{what?? }}
 
@@
 \elpiIn{N} to an assignment \elpiIn{F} we ....
+
+TODO: invariant: variables in subst are always fully applied, and length of scope is the
+arity of the HO variable.
 

### 2024-03-19 paper/appendix.tex
appendix in separate file

--- 2024-03-18 paper/main.tex
@@
-\documentclass[sigconf,natbib=false]{acmart}
-\usepackage[]{biblatex}
-
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-\addbibresource{bib.bib}
-
-\usepackage{myTools}
-\usepackage{macros}
-
-% TODO: set this fields
-%\setcopyright{cc}
-%\setcctype{by}
-\copyrightyear{2024}
-\acmYear{XXXX 2024}
-\acmBooktitle{YYY}
-\acmDOI{ZZZZZZZZZZZZ}
-
-% \xspaceaddexceptions{]\}}
-
-\def\elpi{\proglang{elpi}}
-\def\coqelpi{\proglang{coq-elpi}}
-\def\lambdaprolog{\proglang{$\lambda$-prolog}}
-\def\coq{\proglang{coq}}
-
-\newcommand{\library}[1]{\textit{#1}\xspace}
-\def\stdpp{\library{stdpp}}
-\def\iris{\library{iris}}
-
-\newcommand*{\acronym}[1]{\texttt{#1}\xspace}
-
-\def\ol{\acronym{ol}} % object language
-\def\ml{\acronym{ml}} % meta language
-\def\lf{\acronym{lf}} % logical framework
-\def\ho{\acronym{ho}} % higher order
-\def\Forall{$\forall$}
-
-\newcommand{\EqualRel}{\ensuremath{=}}
-\newcommand{\nEqualRel}{\ensuremath{\new}}
-\newcommand{\UnifRel}{\ensuremath{\simeq}}
-\newcommand{\nUnifRel}{\ensuremath{\not\simeq}}
-
-\newcommand{\Uo}{\ensuremath{\UnifRel_o}\xspace}
-\newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
-\newcommand{\Eo}{\ensuremath{\EqualRel_o}\xspace}
-\newcommand{\nEo}{\ensuremath{\nEqualRel_o}\xspace}
-
-\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
-\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
-\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
-\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
-\newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}
-
-\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}} % space non va
-\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}
-
-\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
-
-\begin{document}
-
-\title{HO unification from object language to meta language}
-
-\author{Enrico Tassi}
-\email{enrico.tassi@inria.fr}
-\affiliation{%
-  \institution{Université Côte d'Azur, Inria}
-  % \city{Nice}
-  \country{France}}
-
-\author{Davide Fissore}
-\email{davide.fissore@inria.fr}
-\affiliation{%
-  \institution{Université Côte d'Azur, Inria}
-  % \city{Nice}
-  \country{France}}
-
-\begin{abstract}
-  Specifying and implementing a logic from scratch requires significant effort.
-  Logical Frameworks and Higher Order Logic Programming Languages provide
-  dedicated, high-level Meta Languages (ML) to facilitate this task in two
-  key ways: 1) variable binding and substitution are simplified when ML binders
-  represent object logic ones; 2) proof construction, and even proof search, is
-  greatly simplified by leveraging the unification procedure provided by the ML.
-  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
-  $\lambda$Prolog~\cite{miller_nadathur_2012} and
-  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-  which have been utilized to implement various formal systems such as
-  First Order Logic~\cite{felty88cade},
-  Set Theory~\cite{10.1007/BF00881873},
-  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
-  Constuctions~\cite{felty93lics}.
-
-  The object logic we are interested in is Coq's~\cite{Coq-refman}
-  Dependent Type Theory (DTT),
-  for which we aim to implement a unification procedure \Uo using the ML
-  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-  Elpi's equational theory comprises
-  $\eta\beta$ equivalence and comes equipped with a
-  higher order unification procedure \Ue restricted to the pattern
-  fragment~\cite{miller92jsc}.
-  We want \Uo to be as powerful as \Ue but on the object logic DTT.
-  Elpi also comes with an encoding for DTT that works well
-  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
-  Unfortunately this encoding, which we refer to as \Fo,
-  ``underuses'' \Ue by restricting it to first-order unification problems only. 
-  To address this issue, we propose a better-behaved encoding, \Ho,
-  demonstrate how to map unification problems in \Fo{}
-  to related problems in \Ho, and illustrate
-  how to map back the unifiers found by \Ue, effectively implementing
-   \Uo on top of \Ue for the encoding \Fo.
-
-  We apply this technique to the implementation of a type-class~\cite{wadler89}
-  solver for Coq~\cite{Coq-refman}.
-  Type-class solvers are proof search procedures based on
-  unification that back-chain designated lemmas, providing essential
-  automation to widely used 
-  Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-  and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
-  constitute our test bed.
-\end{abstract}
-
-\keywords{Logic Programming, Meta-Programming, Higher-Order Unification, Proof Automation}
-
-\maketitle
-
-\section{Introduction}
-\label{sec:intro}
-
-Specifying and implementing a logic from scratch requires significant effort.
-Logical Frameworks and Higher Order Logic Programming Languages provide
-dedicated, high-level Meta Languages (ML) to facilitate this task in two
-key ways: 1) variable binding and substitution are simplified when ML binders
-represent object logic ones; 2) proof construction, and even proof search, is
-greatly simplified by leveraging the unification procedure provided by the ML.
-Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
-$\lambda$Prolog~\cite{miller_nadathur_2012} and
-Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-which have been utilized to implement various formal systems such as
-First Order Logic~\cite{felty88cade},
-Set Theory~\cite{10.1007/BF00881873},
-Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
-Constuctions~\cite{felty93lics}.
-
-The object logic we are interested in is Coq's~\cite{Coq-refman}
-Dependent Type Theory (DTT), and we want to code a type-class~\cite{wadler89}
-solver for Coq~\cite{Coq-refman} using the Coq-Elpi~\cite{tassi:hal-01637063}
-meta programming framework.
-Type-class solvers are unification based proof search procedures
-that combine a set of designated lemmas in order to providing essential
-automation to widely used Coq libraries.
-
-As the running example we take the \coqIn{Decide} type class,
-from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-library. The class identifies predicates equipped with a decision procedure.
-The following three designated lemmas (called \coqIn{Instances} in the
-type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
-smaller than \coqIn{n} is finite; 2) the predicate \coqIn{nfact n nf},
-linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
-3) the universal closure of a predicate has a decision procedure if the
-predicate has and if its domain is finite.
-
-\begin{coqcode}
-Instance fin_fin n : Finite (fin n).              (* r1 *)
-Instance nfact_dec n nf : Decision (nfact n nf).  (* r2 *)
-Instance forall_dec A P : Finite A ~$\to$~             (* r3 *)
-  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
-\end{coqcode}
-
-\noindent Under this context of instances a type-class solver is able to prove
-the following statement automatically by back-chaining.
-
-\begin{coqcode}
-  Check _ : Decision (forall y: fin 7, nfact y 3).       ~\customlabel{goal:g}{(g)}~
-\end{coqcode}
-
-\noindent
-The encoding of DTT provided by Elpi, that we will discuss at length later in
-section~\ref{sec:encodings,sec:lang-spec} and ~\ref{}, is an Higher Order Abstract
-Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
-constructors:
-
-\begin{elpicode}
-type lam  tm -> (tm -> tm) -> tm.     % lambda abstraction
-type app  list tm -> tm.              % n-ary application
-type all  tm -> (tm -> tm) -> tm.     % forall quantifier
-type con  string -> tm.               % constants
-\end{elpicode}
-
-\noindent
-Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
-the concrete syntax to abstract, at the meta level, an expression
-\elpiIn{e} over a variable \elpiIn{x}
-is <<\elpiIn{x\ e}>>, and square brackets denote a list of
-terms separated by comma. As an example we show the encoding of the Coq term
-<<\coqIn{~$\forall$~y:t, nfact y 3}>>:
-
-\begin{elpicode}
-all (con"t") y\ app[con"nfact", y, con"3"]
-\end{elpicode}
-
-\noindent
-We now illustrate the encoding of the three instances above as higher-order
-logic-programming rules: capital letters denote rule
-parameters; \elpiIn{:-} separates the rule's head from the premises;
-\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
-for the premise \elpiIn{p}.
-
-\begin{elpicode}
-finite (app[con"fin", N]).                            ~\customlabel{clause:r1}{(r1)}~
-decision (app [con"nfact", N, NF]).                   ~\customlabel{clause:r2}{(r2)}~
-decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
-  pi w\ decision (app[P, w]).
-\end{elpicode}
-
-\noindent
-Unfortunately this translation of rule \ref{clause:r3} uses the
-predicate \coqIn{P} as a first order term: for the meta
-language its type is \elpiIn{tm}.
-If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
-\ref{goal:g} given below
-
-\begin{elpicode}
-decision (all (app[con"fin", con"7"]) y\                 
-  app[con"nfact", y, con"3"]).
-\end{elpicode}
-
-\noindent
-we obtain an unsolvable unification problem \ref{problem:p}:
-the two lists of terms have different lengths!
-%The root cause is that
-%\ref{problem:p} is an higher order in DTT, but becomes
-%first order in the meta language due to the ``naive'' encoding.
-
-\begin{elpicode}
-app[con"nfact", y, con"3"] = app[P, y]                 ~\customlabel{problem:p}{(p)}~
-\end{elpicode}
-
-\noindent
-In this paper we study a more sophisticated encoding of Coq terms allowing
-us to rephrase the problematic rule \ref{clause:r3} as follows:
-
-\begin{elpicode}
-decision (all A x\ Pm x) :- link Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
-  pi x\ decision (app[P, x]).
-\end{elpicode}
-
-\noindent
-Since \elpiIn{Pm} is an higher-order unification variable
-of type \elpiIn{tm -> tm},
-with \elpiIn{x}
-in its scope, the unification problem \ref{problem:pa}
-admits one solution:
-
-\begin{elpicode}
-app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(p')}~
-Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
-A = app[con"fin", con"7"]              % assignment for A
-\end{elpicode}
-
-\noindent
-After unifying the head of rule \ref{clause:r3a} with the goal, Elpi runs
-the premise <<\elpiIn{link Pm A P}>> that is in charge of bringing the
-assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:
-
-\begin{elpicode}
-P = lam A a\ app[con"nfact", a, con"3"]
-\end{elpicode}
-
-\noindent
-This simple example is sufficient to show that the encoding we seek
-is not trivial and does not only concern the head of rules, but the entire sequence
-of unification problems that constitute the execution of a logic program.
-In fact
-the solution for \elpiIn{P} above generates a
-(Coq) $\beta$-redex in the second premise (the predicate
-under the \elpiIn{pi w\ }\hspace{-0.4em}):
-
-\begin{elpicode}
-decision (app[lam A (a\ app[con"nfact", a, con"3"]), w])
-\end{elpicode}
-
-\noindent
-In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
-the following unification problem has no solution:
-
-\begin{elpicode}
-app[lam A (a\ app[con"nfact", a, con"3"]), x] =
-app[con"nfact", N, NF]
-\end{elpicode}
-\noindent
-~\\
-The root cause of the problems we sketched in the running example
-is that the unification procedure \Ue of the meta language is not aware
-of the equational theory of the object logic, even if both theories
-include $\eta\beta$-conversion and admit most general
-unifiers for unification problems in the pattern fragment \llambda~\cite{miller92jsc}.
-
-\paragraph{Contributions}
-In this paper we discuss alternative encodings of Coq in
-Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
-in which the problems sketched here can be fully described.
-We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
-the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
-bla bla.. TODO citare Teyjus.
-
-\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
-\label{sec:encodings}
-
-The equational theory of Coq's Dependent Type Theory is very rich. In
-addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
-are compared up to proposition unfolding and fixpoint unrolling. Still,
-for efficiency and predictability reasons, most form of automatic proof search
-employ a unification procedure that captures a simpler one,
-just $\eta\beta$, and that solves higher-order problems
-restricted to the pattern fragment $\llambda$~\cite{miller92jsc}.
-We call this unification procedure \Uo{}.
-
-The equational theory of the meta language Elpi that we want to use to
-implement a form of proof automation is strikingly similar, since it
-it comprises $\eta\beta$ (for the meta language functions), and the
-unification procedure \Ue{} solves higher-order problems in
-$\llambda$.
-
-In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
-since the abstraction and application term constructors
-the two unification procedures deal with are different. For example
-
-\begin{tabular}{lcl}
-\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
-\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"}\\
-\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"} \\
-\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
-\end{tabular}
-
-\noindent
-One could ignore this similarity, and ``just'' describe the object language
-unification procedure in the meta language, that is crafting a \elpiIn{unif}
-predicate to be used as follows in rule \ref{clause:r3}:
-
-\begin{elpicode}
-decision X :- unif X (all A x\ app[P, x]), finite A,
-  pi x\ decision (app[P, x]).
-\end{elpicode}
-
-\noindent
-This choice would underuse the logic programming engine provided by
-the metalanguage since by removing any datum from the head of rules
-indexing degenerates. Moreover the unification procedure built in the
-meta language is likely to be faster than one implemented in it,
-especially if the meta language is interpreted as Elpi is.
-
-To state precisely the problem we solve we need a \Fo{} representation
-of DTT terms and a \Ho one.
-We call \Eo the equality over ground terms in \Fo,
-\Ee the equality over ground terms in \Ho,
-\Uo the unification procedure we want to implement and
-\Ue the one provided by the meta language.
-TODO extend \Eo and \Ee with reflexivity on uvars.
-
-\newcommand{\specunif}[3]{
-  #3_i \in \llambda \Rightarrow %
-    \exists \rho, %
-      \rho #3_1 #1 \rho #3_2  %
-        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho 
-}
-
-\newcommand{\unifcorrect}[3]{
-  % \forall \rho #3_1 #3_2, %
-    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
-    #3_i \in \llambda \Rightarrow
-      #3_1 #2 #3_2 \mapsto \rho
-        \Rightarrow 
-          \rho #3_1 #1 \rho #3_2  %
-}
-
-\newcommand{\unifcomplete}[3]{
-  % \forall #3_1 #3_2, %
-    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
-    #3_i \in \llambda \Rightarrow
-      % \forall \rho, %
-        \rho #3_1 #1 \rho #3_2  %
-          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho 
-}
-
-We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
-$\sigma$, we write $\sigma t$ for the application of the substitution to
-$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
-and we assume that the unification of our meta language is correct:
-%
-\begin{gather}
-  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
-  \unifcomplete{\Ee}{\Ue}{t}
-\end{gather}
-
-\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
-\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
-
-We illustrate a compilation $\C{s}{t}{m}{l}$ that
-maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
-list of links $l$.
-The variable map connects unification variables in \Ho with variables
-in \Fo{} and is used to ``decompile'' the assignment,
-$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
-are linked to the unification variable that stands in their place in the
-compiled term. These links are checked for or progress XXX improve....
-
-We represent a logic program \emph{run} in \Fo as
-a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
-unification problem between terms $\mathcal{S}_{p_l}$ and
-$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
-The composition of these steps starting from the
-empty substitution $\rho_0$ produces the final
-substitution $\rho_\mathcal{N}$.
-\footnote{If the same rule is used multiple time in a run we
-just consider as many copies as needed of the terms composing the
-rules, with fresh unification variables each time}
-The initial here $\rho_0$ is the empty substitution
-%
-\newcommand{\solve}{\ensuremath{\mathrm{check}}}
-\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
-\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
-\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
-\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
-\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
-\newcommand{\stepFD}[5]{%
-\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
-\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
-\newcommand{\stepHD}[6]{\ensuremath{%
-#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
-\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
-\newcommand{\runFD}[2]{\ensuremath{%
-\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
-\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
-\newcommand{\runHD}[3]{\ensuremath{%
-\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
-\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
-$$
-\begin{array}{l}
-\stepF{\mathcal{S}}{p}{\rho}{\rho''}
-\deff
-\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
-\runF{\mathcal{S}}{\mathcal{N}}{\rho}
-\deff
-\runFD{\mathcal{S}}{\mathcal{N}}
-\end{array}
-$$
-
-We simulate each run in \Fo{} with a run in \Ho as follows.
-Note that $\sigma_0$ is the empty substitution.
-$$
-\begin{array}{l}
-\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\vspace{2pt}\\
-  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\vspace{2pt}\\
-\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
-  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\vspace{2pt}\\
-  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
-\end{array}
-$$
-
-\noindent
-Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
-terms) and a call to \solve{} on the set of links. We claim the following:
-
-\begin{proposition}[Simulation]\label{prop:sumul}
-$\forall \mathcal{S}, \forall \mathcal{N}$
-$$
-  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
-  \Leftrightarrow
-  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
-$$
-\end{proposition}
-
-\noindent
-That is, the two executions give the same result. Moreover:
-
-\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of~ \hrun, we have that
-$\forall p \in 1 \ldots \mathcal{N}$
-$$
-\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
-\Leftrightarrow
-\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
-$$
-\end{proposition}
-\noindent
-In particular this property guarantees that a \emph{failure} in the \Fo{} run
-is matched by a failure in \Ho{} \emph{at the same step}. We consider this
-property very important from a practical point of view since it guarantees
-that the execution traces are strongly related and in turn this enables a user
-to debug a logic program in \Fo{} by looking at its execution trace in
-\Ho{}.
-
-We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
-$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
-%
-$$
-\begin{array}{l}
-s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
-\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
-\quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
-\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
-\end{array}
-$$
-
-\begin{proposition}[Properties of \Uo{}]
-\begin{gather}
-  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
-\unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}\\
-% \forall \rho'\rho, 
-  \rho s_1 \Eo \rho s_2 \Rightarrow
-  \rho' \subseteq \rho \Rightarrow
-  \rho's_i \in \llambda \Rightarrow
-  \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
-\end{gather}
-\end{proposition}
-
-Properties \ref{prop:correct} and \ref{prop:complete} state, respectively, that
-in \llambda the implementation of \Uo is correct, complete and returns the most
-general unifier.
-
-Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
-is resilient to unification problems outside \llambda solved by
-a third party. We believe this property is of practical interest since we
-want the user to be able to add heuristics via hand written rules
-to the ones obtained by our compilation scheme. A Typical example
-is the following problem \ref{problem:q} that is outside \llambda:
-
-\begin{elpicode}
-app [F, con"a"] = app[con"f", con"a", con"a"]   ~\customlabel{problem:q}{(q)}~
-F = lam x\ app[con"f",x,x]                      ~\customlabel{heuristic:h}{(h)}~
-\end{elpicode}
-
-\noindent
-Instead of rejecting it our scheme accepts it and guarantees that if
-\ref{heuristic:h} is given (after the compilation part of the scheme, as
-a run time hint) then ...
-
-\subsection{The intuition in a nutshell}
-A term $s$ is compiled in a term $t$ where any
-``problematic'' subterm $tp$ is replaced by a unification variable $h$
-and an accessory link, that represent a suspended unification problem
-$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
-We now define ``problematic'' formally.
-
-\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
-\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
-\begin{definition}[\maybeeta]
-  $\maybeeta = \{ t | \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
-\end{definition}
-
-\noindent
-An example of term $t$ in \maybeeta{} is
-$\lambda x.\lambda y.F~y~x$
-since the substitution
-$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
-makes $\rho t = \lambda x.\lambda y.f x y$
-that is the eta long form of $f$.
-
-\begin{definition}[\maybebeta]
-  $\maybebeta = \{ X t_1 \ldots t_n | t \not\in \llambda \}$.
-\end{definition}
-
-\noindent
-An example of $t$ in \maybebeta{} is $F a$ for a constant $a$. Note however tha
-an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
-that makes the resulting term fall outside of \maybebeta.
-
-\begin{definition}[Subterm $\mathcal{P}(t)$]
-$$
-\begin{array}{l}
-t \in \mathcal{P}(t)\\
-t = f t_1\ldots t_n \Rightarrow \mathcal{P}(t_i) \subseteq \mathcal{P}(t) \land f \in \mathcal{P}(t)\\
-t = \lambda x.t' \Rightarrow \mathcal{P}(t') \subseteq \mathcal{P}(t)\\
-\end{array}
-$$
-\end{definition}
-
-\noindent
-We write $\mathcal{P}(X) = \bigcup_{t\in X} \mathcal{P}(t)$.
-
-\marginpar{TODO: find better word}
-\newcommand{\normal}{\ensuremath{\mathrm{normal}}}
-\begin{definition}[Normal form]
-Given a set of terms $X \subseteq \Ho{}$,
-$$
-\normal(X) =
-\mathcal{P}(X) \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
-$$
-\end{definition}
-
-\noindent
-We write $\sigma X = \{ \sigma t | t \in X\}$.
-
-\begin{proposition}[Normal form preservation]\label{prop:nf}
-$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
-$$
-\normal(\sigma\mathcal{T}) \land
-\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
-\Rightarrow \normal((\sigma \cup \sigma')\mathcal{T})
-$$
-\end{proposition}
-
-\noindent
-In particular this guarantees that is we start from normal terms
-we never introduce eta-long or non-beta-normal terms in $\sigma'$.
-
-Note that proposition \ref{prop:nf} does not hold for \Uo{} since
-decompilation can introduce (actually restore) terms in
-\maybeeta or \maybebeta.
-
-\section{Alternative encodings and related work}
-
-Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
-
-Our encoding of DTT may look ``semi shallow'' since we use the meta-language
-lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
-A fully shallow encoding unfortunately does not fit our use case, although
-it would make the running example work:
-
-\begin{elpicode}
-finite (fin N).
-decision (nfact N NF).
-decision (all A x\ P x) :- finite A, pi x\ decision (P x).
-\end{elpicode}
-
-\noindent
-There are two reasons for dismissing this encoding. The first one is that
-in DTT it is not always possible to adopt it since the type system
-of the meta language is too weak to accommodate terms with a variable arity,
-like the following example:
-
-\begin{coqcode}
-Fixpoint arr T n := if n is S m then T -> arr T m else T.
-Definition sum n : arr nat n := ...
-Check sum 2   7 8   : nat.
-Check sum 3   7 8 9 : nat.
-\end{coqcode}
-
-\noindent
-The second reason is the encoding for Coq is used for meta programming the
-system, hence it must accommodate the manipulation of terms that are now
-know in advance (not even defined in Coq) without using introspection
-primitives such as Prologs's \texttt{functor} and \texttt{arg}.
-
-In the literature we could find a few related encoding of DTT.
-TODO In~\cite{felty93lics} is related and make the
-discrepancy between the types of ML and DTT visible. In this case
-one needs 4 application nodes. Moreover the objective is an encoding
-of terms, proofs, not proof search. Also note the conv predicate,
-akin to the unif we rule out.
-
-TODO This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
-
-None of the encodings above provide a solution to our problem.
-
-\section{Preliminaries: \Fo{} and \Ho}
-\label{sec:lang-spec}
-
-In order to reason about unification we provide a description of the
-\Fo{} and \Ho languages where unification variables
-are first class terms, i.e. they have a concrete syntax. We keep these languages
-minimal, for example we omit the \elpiIn{all} quantifier of DTT we used
-in the example in Section~\ref{sec:intro}. 
-%
-\setlength{\abovecaptionskip}{0pt}
-\setlength{\belowcaptionskip}{-13pt}
-
-\begin{figure}[H]
-  \begin{tabular}{ll}
-  \begin{minipage}{0.21\textwidth}
-  \begin{elpicodetab}
-kind fm type.
-type fapp list fm -> fm.
-type flam (fm -> fm) -> fm.
-type fcon string -> fm.
-type fuva nat -> fm.
-\end{elpicodetab}
-  \end{minipage}
-  &
-  \begin{minipage}{0.24\textwidth}
-  \begin{elpicodetab}
-kind tm type.
-type app list tm -> tm.
-type lam (tm -> tm) -> tm.
-type con string -> tm.
-type uva nat -> list tm -> tm.
-  \end{elpicodetab}
-  \end{minipage}
-  \end{tabular}
-  \caption{\Fo{} and \Ho language}\vspace{0.3em}
-  \label{code:common-terms}
-  \Description[code:common-terms]{code:common-terms}
-\end{figure}
-
-\noindent
-In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
-Unification variables standing for functions are applied their arguments
-via the \elpiIn{fapp} constructor.
-For example in the statement of the instance \coqIn{forall_dec} the
-term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N}
-is of type \elpiIn{nat} and \elpiIn{x} is a bound variable.
-
-In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]}.
-We say that the unification variable \elpiIn{uva N L} is in \llambda iff
-\elpiIn{distinct L} holds.
-
-\begin{elpicode}
-type distinct list A -> o.
-distinct [].
-distinct [X|XS] :- name X, not(mem X XS), distinct XS.
-\end{elpicode}
-
-\noindent
-The \elpiIn{name} builtin predicate tests if a term is a bound variable.
-\footnote{one could always load name x for every x under a pi and get rid of the name builtin}
-The
-compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
-reasons, so we don't assume all out terms are in \llambda but rather test.
-
-In both languages unification variables are identified by a natural number,
-which can be seen as a memory address. The memory and its associated
-operations are described below:
-
-\begin{elpicode}
-typeabbrev memory A (list (option A)).
-type set?   nat -> memory A -> A -> o.
-type unset? nat -> memory A -> o.
-type assign nat -> memory A -> A -> memory A -> o.
-\end{elpicode}
-
-\noindent 
-If a memory cell is \elpiIn{none}, then the corresponding unification variable
-is not set. \elpiIn{assign} sets an unset cell to the given value.
-
-Since in \Ho unification variables have a scope, their solution needs to be
-abstracted over it in order to enable the instantiation of a single
-solution to different scopes. On the contrary solutions to \Fo variables
-are plain terms.
-
-\begin{elpicode}
-  typeabbrev fsubst (memory ftm).
-  kind assmt type.
-  type abs (tm -> assmt) -> assmt.
-  type val tm -> assmt.
-  typeabbrev subst (memory assmt).
-\end{elpicode}
-
-\noindent
-We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
-the one of \Ho.
-Both have the invariant that they are not cyclic, TODO explain.
-
-\subsection{Equational theory and Unification}
-
-here we give the functions/signatures to express the properties 3-5 in the problem
-statement
-
-Together with the description of the terms of the language, we need some
-auxiliary functions to perform operations like term equality, unification,
-dereferencing. There predicates are supposed to be implemented in the OL and the
-ML following respectively their specification. In the following few paragraphs
-we give a brief proposition of them.
-
-\paragraph{Term dereferencing: $\rho s$ and $\sigma t$}
-
-Since in our encoding we explicitly carry a substitution we need to
-define the operation that applies it to a term. Its proposition
-takes care to normalize (flatten) applications, for example
-it turns
-\elpiIn{(app [app [con"f",con"a"],con"b"])}
-into
-\elpiIn{(app [con"f",con"a",con"b"])}.
-
-dereference variables of the two languages. This is particularly useful to check
-if two terms in the OL (resp. in the ML) are equal. The constructors
-representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
-make the dereferencing procedure to recurse over the their subterms. The code
-below display this behavior for the dereferencing performed by the OL:
-
-\begin{elpicode}
-type fder fsubst -> fm -> fm -> o.
-fder S (fapp A) (fapp B) :- map (fder S) A B.
-fder S (flam F) (flam G) :- 
-  pi x\ fder S x x => fder S (F x) (G x).
-fder _ (fcon C) (fcon C).
-fder S (fuva N) R :- set? N S T, fder S T R.
-fder S (fuva N) (fuva N) :- unset? N S.
-
-type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fder}{\rho s}$)~
-fderef S T R :- fder S T T', napp T' R.
-
-type napp fm -> fm -> o.
-\end{elpicode}
-
-\noindent
-TODO explain napp.
-We use the cut operator to keep the code compact. It is possible to rewrite
-the rule for application traversal so that it is mutually exclusive with the
-first one, but that requires a rather verbose analysis of the head of \elpiIn{A}.
-
-The corresponding code for \Ho is similar, we only show the last two rules that
-differ in a substantial way:
-
-\begin{elpicode}
-type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
-deref S (app A) (app B) :- map (deref S) A B.
-deref S (lam F) (lam G) :- 
-  pi x\ deref S x x => deref S (F x) (G x).
-deref _ (con C) (con C).
-deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
-deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
-type move assignment -> list tm -> tm -> o.
-move (abs Bo)        [H|L] R :- move (Bo H) L R.
-move (val A)         []    A :- !.
-move (val (uva N A)) L     (uva N X) :- std.append A L X.
-\end{elpicode}
-
-\noindent
-TODO: no need to napp, see the beta section.
-Note that when the substitution \elpiIn{S} maps a unification variable
-\elpiIn{N} to an assignment \elpiIn{F} we ....
-
-Important!!!
-A different reasoning is to be addressed to the variables of the ML. Firstly, a
-meta variable cannot appear in the \elpiIn{app} node as the first element of the
-list, we will explain why in \cref{sec:compilation}
-
-\begin{elpicode}
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
-  fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
-  flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
-  fcon C ~\Eo~fcon C.
-  fuva N ~\Eo~fuva N.
-  flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
-    pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
-  T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
-    pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
-  fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
-  T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
-\end{elpicode}
-
-\begin{figure}
-  \begin{elpicode}
-    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-    app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
-    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
-    con C ~\Ee~fcon C.
-    uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
-  \end{elpicode}
-  \caption{Equal predicate ML}
-  \label{code:eq-pred-meta}
-  \Description[code:equal-pred-meta]{code:equal-pred-meta}
-\end{figure}
-
-\paragraph{Term equality: \Eo vs. \Ee} We can test if two terms are equal
-following the equational theory of the language being considered. In
-\cref{code:eq-pred} we provide an implementation of the \Eo predicate. The first
-four rules check if the two terms are equal regarding the structure of the
-current node, that is, two terms are equal if they have same head and if
-recursively each subterm is two by two equal. Moreover, since the theory of the
-OL accepts $\eta$- and $\beta$-equivalence, then we implement the corresponding
-rules.
-
-\noindent
-The equality relation for the ML, accepts $\eta\beta$-equivalence between terms
-of the ML. Recall that \elpiIn{abs x\ f x}, is a valid $\eta$ expansion of the
-function \elpiIn{f} and that \elpiIn{lam x\ app[f, x]} is not that equivalent to
-\elpiIn{f} at meta level. However, since we are interested in using the
-unification procure of the ML, by \cref{prop:correct-ml}, we can use the
-\Ue relation to test, when needed if two terms are equal in the ML.
-
-\paragraph{Term unification: \Uo vs. \Ue} 
-The last but not least important relation we should take care of before
-presenting our full algorithm aiming to unify terms of the OL in the ML and
-provide the substitution produced in the ML to the OL, is term unification. This
-procedure is a more powerful version of the equal predicate, since unification
-checks if two terms can be equal be assigning unification variables. In our
-representation, variable assignment (or refinement) is performed by modifying
-the corresponding substitution mapping. We will not give an implementation of
-\Uo, since we are giving an implementation of it using
-our algorithm, see \cref{XX}.
-
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
-\end{elpicode}
-
-On the other hand, unification in the ML needs to be defined. In
-\cref{code:ue-pred}, we give an implementation of \Ue but
-that is actually what our meta language provides as a builtin.
-
-This predicate has four
-arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
-$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
-\subseteq \rho_2$. The first three rules unify terms with same rigid heads, and
-call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
-assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
-unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
-terms are unification variables, we test that their arguments are in the pattern
-fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
-pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
-$w$ and return the new mapping $\rho_2$ containing all the new variable
-assignment. Finally, if only one of the two terms is an unification variable
-$v$, after having verified that $v$ does not occur in the other term $t$, we
-bind $v$ to $t$ and return the new substitution mapping.
-
-\old
-
-A key property needed in unification is being able to verify if two terms are
-equal wrt a given equational theory. This relation allow to compare terms under
-a certain substitution mapping, so that any time a variable $v$ is assigned in a
-subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
-test for equality is continued on the new-created subterm.
-
-The base equality function over terms can be defined as follows:
-
-The solution we are proposing aim to overcome these unification issues by 1)
-compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
-in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
-variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
-the variables of the OL with the decompiled version of their corresponding meta
-variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
-and that the substitution in the object language is the same as the one returned
-by the ML. \todo{same or $\supseteq$ or $\subseteq$}
-
-In the following section we explain how we deal with term (de)compilation and
-links between unification variables.
-
-\section[Compilation: fo\_tm to tm]{Basic compilation \Fo{} to \Ho{}}
-\label{sec:compilation}
-
-In this section we describe a basic compilation scheme that we refine
-later, in the following sections.
-
-The objective of the compilation is to recognize the higher-order variables
-available in \Ho{} when expressed in a first order way in \Fo{}. The compiler
-also generates a list of links that are used to bring back the substitution from
-\Ho{} to \Fo{} and allocates in the memory a cell for each variable.
-
-\begin{elpicode}
-kind link type.
-type link nat -> nat -> nat -> subst. % link Fo Ho Arity
-typeabbrev links list link.
-type comp fm -> tm -> links -> links -> subst -> subst -> o.
-comp (fcon X) (con X) L L S S.
-comp (flam F) (lam G) K L R S :- pi x y\
-  (pi A S\ comp x y L L S S) => comp (F x) (G y) K L R S.
-comp (fuva M) (uva N []) K [link M N z|K] R S :- new R N S.
-comp (fapp[fuva M|A]) (uva N B) K L R S :- distinct A, !,
-  fold4 comp A B K K R R,
-  new R N S, len A Arity,
-  L = [link N M Arity | K].
-comp (fapp A) (app B) K L R S :- fold4 comp A B K L R S.
-\end{elpicode}
-
-\noindent
-Note that \elpiIn{link} carries the arity (number of expected arguments) of
-the variable. \marginpar{say when this is needed}
-
-\begin{elpicode}
-type solve-links links -> links -> subst -> subst -> o.
-solve-links L L S S.
-\end{elpicode}
-
-Then decomp
-
-\begin{elpicode}
-type decompile links -> subst -> fsubst -> o.
-decompile L S O :-
-  map (_\r\r = none) S O1, % allocate empty fsubst
-  (pi N X\ knil N X :- mem L (link X N _) ; N = X) =>
-    decompl S L O1 O.
-type knil nat -> nat -> o.
-
-type decompl links -> subst -> fsubst -> o.
-decompl S [] [].
-decompl S [link _ N _|L] O P :- unset? N S X,
-  decompl S L O P.
-decompl S [link M N _|L] O P :- set? N S X,
-  decomp-assignment S X T, assign M O (some T) O1,
-  decompl S L O1 P.
-
-type decomp-assignment subst -> assignment -> fm -> o.
-decomp-assignment S (abs F) (flam G) :-
-  pi x y\ decomp-tm S x y => decomp-assignment S (F x) (G y).
-decomp-assignment S (val T) T1 :- decomp S T T1.
-
-type decomp subst -> tm -> fm.
-decomp _ (con C) (fcon C).
-decomp S (app A) (app B) :- map (decomp S) A B.
-decomp S (lam F) (flam G) :-
-  pi x y\ decomp S x y => decomp S (F x) (G y).
-decomp S (uva N A) R :- set? N S F,
-  move F A T, decomp S T R.
-decomp S (uva N A) R :- unset? N S,
-  map (decomp S) A B, knil N M, napp (fapp[fuva M|B]) R.
-\end{elpicode}
-
-\marginpar{TODO link}
-\marginpar{TODO nuove subst}
-
-Now unif\marginpar{TODO: code unif}
-
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Uo)} \PYG{k+kt}{fm -> fm -> subst -> subst -> o}~.
-(X ~\Uo~Y) S S1 :-
-  fderef S X X0, fderef S Y Y0,                    ~($\customlabel{step:norm}{norm}$)~
-  comp X0 X1 [] S0 [] L0,                        ~($\customlabel{step:compile}{compile}$)~
-  comp Y0 Y1 S0 S1 L0 L1,
-  (X1 ~\Ue~Y1) [] HS0,                              ~($\customlabel{step:unify}{unify}$)~
-  solve-links L1 L2 HS0 HS1,                        ~($\customlabel{step:link}{link}$)~
-  decompile L2 HS1  S1.                        ~($\customlabel{step:decompile}{decompile}$)~
-\end{elpicode}
-
-\subsection{Prolog simulation}
-Allows us to express the properties.
-we take all terms involved in a search (if a rule is used twice we simply
-take a copy of it), we compile all of them, and then we pick the unification
-prblems among these terms and step trough them.
-
-\begin{elpicode}
-type pick list A -> (pair nat nat) -> (pair A A) -> o.
-pick L (pr X Y) (pr TX TY) :- nth X L TX, nth Y L TY.
-
-type prolog-fo list fm -> list (pair nat nat) -> subst -> o.
-prolog-fo Terms Problems S :-
-  map (pick Terms) Problems FoProblems,
-  fold4 ~(\Uo)~ FoProblems [] S.
-
-type step-ho (pair tm tm) -> links -> links -> subst -> subst -> o.
-step-ho (pr X Y) L0 L1 S0 S2 :-
-  (X1 ~\Ue~Y1) S0 S1,
-  solve-links L0 L1 S1 S2.
-
-type prolog-ho list fm -> list (pair nat nat) -> subst -> o.
-prolog-ho Terms Problems S :-
-  fold4 comp Terms HoTerms [] L0 [] HS0,
-  map (pick HoTerms) Problems HoProblems,
-  fold4 step-ho HoProblems L0 L HS0 HS,
-  decompile L HS S.
-\end{elpicode}
-
-\noindent
-the proprty is that if a step for Fo succeds then the Ho one does, and
-if Fo fails then the Ho fails ()
-
-\subsection{Example}
-
-OK
-
-\begin{elpicode}
-Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]] 
-      , flam x\ fapp[fcon"g", fcon"a"] ]
-Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$ 
-lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
-link z z (s z)
-HS = [some (abs x\con"a")]
-S = [some (flam x\fcon a)]
-\end{elpicode}
-
-KO
-
-\begin{elpicode}
-  Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]] 
-  , flam x\ fapp[fcon"g", fcon"a"] ]
-Problems = [ pr 0 1   % $A = \lambda x.x$
-           , pr 2 3 ] % $A a = a$ 
-lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
-link z z (s z)
-HS = [some (abs x\con"a")]
-S = [some (flam x\fcon a)]
-lam x\ app[f, app[X, x]] = Y,
-  lam x\ x) = X.
-\end{elpicode}
-
-\otext{Goal: $s_1 \Uo s_2$ is compiled into $t_1 \Ue t_2$}
-\otext{What is done: uvars \elpiIn{fo_uv} of OL are replaced into uvars \elpiIn{ho_uv} of the ML}
-\otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
-\otext{Example needing the compiler v0 (tra l'altro lo scope è ignorato):\\ \elpiIn{lam x\ app[con"g",app[uv 0, x]] ~\Uo~lam x\ app[con"g", c"a"]}}
-\otext{Links used to instantiate vars of elpi}
-\otext{After all links, the solution in links are compacted and given to coq}
-\otext{It is not so simple, see next sections (multi-vars, eta, beta)}
-
-The compilation step is meant to recover the higher-order variables of the OL,
-expressed in a first order way, by replacing them with higher-order variables in
-the ML. In particular, every time a variable of the OL is encountered in the
-original term, it is replaced with a meta variable, and if the OL variable is
-applied to a list of distinct names $L$, then this list becomes the scope of the variable. 
-For all the other constructors of
-\elpiIn{tm}, the same term constructor is returned and its arguments are
-recursively compiled. The predicate in charge for term compilation is:
-
-\elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
-
-\noindent
-where, we take the term of the OL, produce the term of the ML, take a list
-of link and produce a list of new links, take a substitution and return a 
-new substitution.
-
-In particular, due to programming constraints, we need to drag the old subst and
-return a new one extended, if needed, with the new declared meta-variables.
-
-The following code
-%
-\begin{elpicode}
-  kind link type.
-  type link nat -> nat -> nat -> subst.
-\end{elpicode}
-%
-\noindent
-defines a link, which is a relation between to variables indexes, the first
-being the index of a OL variable and the second being the index of a ML
-variable. The third integer\todo{integer or nat?} is the number of term in the
-scope of the two variables, or equivalently, in a typed language, their arity.
-
-As an example, let's study the following unification problem (a slightly 
-modified version from \cref{sec:intro}):
-
-\begin{elpicode}
-  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
-    lam x\ app [c"decision", app[uv 0, x]]
-\end{elpicode}
-
-\noindent
-we have the main unification problem where the nested \elpiIn{app} nodes have
-lists of different lengths making the unification to fail. The compilation of 
-these terms produces a new unification problem with the following shape:
-
-\begin{elpicode}
-  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~
-    lam x\ app [c"decision", uv 1 [x]]
-\end{elpicode}
-
-\noindent
-The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
-the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
-ML, that is, the index \elpiIn{0} for that unification variable of the OL term
-has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
-different substitution mapping, one for the OL and one for the ML and the indexes
-of variable point to the respective substitution.
-
-decomp che mappa abs verso lam
-\noindent
-\otext{An other example: \\
-  \elpiIn{lam x\ app[f, app[X, x]] = Y, (lam x\ x) = X.}}
-
-\section{Use of multivars}
-
-Se il termine initziale è della forma
-
-\begin{elpicode}
-  app[con"xxx", (lam x\ lam y\ Y y x), (lam x\ f)]
-  =
-  app[con"xxx",X,X]
-\end{elpicode}
-
-allora se non uso due X diverse non ho modo di recuperare il quoziente che mi manca.
-
-a sto punto consideriamo liste di problemi e così da eliminare sta xxx senza
-perdità di generalità (e facciamo problemi più corti, e modellizziamo anche la
-sequenza)
-
-\subsection{Problems with $\eta$}
-\otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
-\otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
-\otext{It is not doable, with the same elpi var}
-
-\subsection{Problems with $\beta$}
-
-\otext{The following goal: \elpiIn{X = lam x\ x, app[X, 3] = 3}}
-\otext{We use links-beta}
-
-\subsection{Tricky examples}
-
-\begin{elpicode}
-  triple ok (@lam x\ @app[@f, @app[@X, x]]) @Y,
-  triple ok @X (@lam x\ x),
-  triple ok @Y @f
-\end{elpicode}
-
-\begin{elpicode}
-% @okl 22 [
-%   triple ok (@lam x\ @lam y\ @app[@Y, y, x]) @X,
-%   triple ok (@lam x\ @f) @X,
-% ].
-\end{elpicode}
-
-\section{First order approximation}
-
-\otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
-\otext{We can re-use part of the algo for $\beta$ given before}
-
-\section{Unif encoding in real life}
-\otext{Il ML presentato qui è esattamente elpi}
-\otext{Il OL presentato qui è esattamente coq}
-\otext{Come implementatiamo tutto ciò nel solver}
-
-\section{Results: stdpp and tlc}
-\otext{How may rule are we solving?}
-\otext{Can we do some perf test}
-
-\section{Conclusion}
-
-\printbibliography
-
-\clearpage
 \section*{Appendix}
@@
 \end{figure*}
-
-\end{document}

### 2024-03-19 paper/main.tex
appendix in separate file

@@
 %
-\newcommand{\solve}{\ensuremath{\mathrm{check}}}
-\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
-\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
-\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
-\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
+\newcommand{\solve}{\ensuremath{\mathrm{check}}\xspace}
+\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
+\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
+\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
+\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
 \newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
@@
 \subsection{The intuition in a nutshell}
+\label{sec:nutshell}
 A term $s$ is compiled in a term $t$ where any
@@
 Both have the invariant that they are not cyclic, TODO explain.
+Other invariant: the terms in ho\_subst never contains eta and beta expansion
 
@@
 
-\otext{The following goal: \elpiIn{X = lam x\ x, app[X, 3] = 3}}
+$\beta$-reduction problems (\maybebeta) appears any time we deal with a subterm $t
+= X t_1 \dots t_n$, where $X$ is flexible and the list $[t_1 \dots t_n]$ in not
+in \llambda. This unification problem is not solvable without loss of
+generality, since there is not a most general unifier. If we take back the
+example given in \cref{sec:nutshell}, the unification $F a = a$ admits two solutions for $F$:
+$\rho_1 = \{F \mapsto \lambda x.x\}$ and $\rho_2 = \{F \mapsto \lambda \_.a\}$.
+Despite this, it is possible to work with $\maybebeta$ if an oracle provides a
+substitution $\rho$ such that $\rho t$ falls again in the \llambda.
+
+On the other hand, the \Ue is not designed to understand how the $\beta$-redexes
+work in the onject language. Therefore, even if we know that $F$ is assigned
+to $\lambda x.x$, \Ue is not able to unify $F a$ with $a$. On the other hand,
+the problem $F a = G$ is solvable by \Ue, but the final result is that $G$ is
+assigned to $(\lambda x.x) a$ which breaks the invariant saying that the 
+substitution of the meta language always contain only terms in normal form.
+
+\otext{The following goal: \elpiIn{X = lam x\ x, app[X, a] = a}}
 \otext{We use links-beta}
@@
 \clearpage
-\section*{Appendix}
-
-Note that \elpiIn{(a infix b) c d} de-sugars to \elpiIn{(infix) a b c d}.
-
-\begin{figure*}
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
-fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
-flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
-fcon C ~\Eo~fcon C.
-fuva N ~\Eo~fuva N.
-flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
-  pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
-T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
-  pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
-fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
-T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
-
-type beta fm -> list fm -> fm -> o.
-beta A [] A.
-beta (flam F) [H | L] R :- subst F H B,
-  beta B L R. % since F could be x\app[x|_] and H be lam _ 
-beta (fapp A) L (fapp X) :- append A L X.
-beta (fuva N) L (fapp [fuva N | L]).
-beta (fcon H) L (fapp [fcon H | L]).
-
-type subst (fm -> fm) -> fm -> fm -> o.
-subst F H B :- napp (F H) B. % since (F H) may generate (app[app _|_])
-
-type napp fm -> fm -> o.
-napp (fcon C) (fcon C).
-napp (flam F) (flam G) :- pi x\ napp x x => napp (F x) (G x).
-napp (fapp[fapp L|M]) R :- !, append L M N, napp (fapp N) R.
-napp (fapp[X]) R :- !, napp X R.
-napp (fapp A) (fapp B) :- map napp A B.
-napp (fuva N) (fuva N).
-\end{elpicode}
-\caption{Full implementation of the \Eo predicate for \Fo}
-\label{code:eo-pred}
-\Description[code:eo-pred]{code:eo-pred}
-\end{figure*}
-
-\begin{figure*}
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
-% Congruence
-(app A ~\Ue~app B) R S :- fold2 ~(\Ue)~A B R S.
-(lam F ~\Ue~lam G) R S :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) R S.
-~\Ue~(con C) (con C) S S.
-% deref
-(uva N A ~\Ue~T) R S :- set? N S F, move F A T1, (T1 ~\Ue~T) R S.
-(T ~\Ue~uva N A) R S :- set? N S F, move F A T1, (T ~\Ue~T1) R S.
-% flex-flex
-(uva N A ~\Ue~uva M B) S S3 :-  unset? M, unset? N,
-  distinct A, distinct B, 
-  new S W S1, prune W Args1 B Ass, 
-  assign N S1 Ass S2, assign M S2 Ass S3.
-% assignment
-(uva N A ~\Ue~T) R S :- distinct A, not (T = uva _ _), not_occ N S T,  
-  bind A T T1, assign N S T1 S1.
-(T ~\Ue~uva N A) R S :- distinct A, not (T = uva _ _), not_occ N S T, 
-  bind A T T1, assign N S T1 S1.
-
-type distinct list A -> o.
-distinct [].
-distinct [X|XS] :- name X, not(mem X XS),
-distinct XS.
-
-typeabbrev memory A (list (option A)).
-type set? nat -> memory A -> A -> o.
-set? N S T :- nth N S (some T).
-type unset? nat -> memory A -> o.
-unset? N S :- nth N S none.
-type assign nat -> memory A -> A -> memory A -> o.
-assign z [none|M] T [some T|M].
-assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
-kind nat type.
-type z nat.
-type s nat -> nat.
-type nth nat -> list A -> A -> o.
-nth z [X|_] X.
-nth (s N) [_|L] X :- nth N L X.
-
-type new memory A -> nat -> memory A -> o.
-new [] z [none].
-new [X|XS] (s N) [X|YS] :- new XS N YS.
-
-type prune .
-type move .
-type beta.
-type bind.
-type not_occ.
-TODO
-
-type fold2 (A -> A1 -> B -> B -> o) -> list A -> list A1 -> B -> B -> o.
-fold2 _ [] [] A A.
-fold2 F [X|XS] [Y|YS] A A1 :- F X Y A A0, fold2 F XS YS A0 A1.
-
-\end{elpicode}
-\caption{Implementation of the \Ue predicate for \Ho}
-\label{code:ue-pred}
-\Description[code:ue-pred]{code:ue-pred}
-\end{figure*}
-
-\begin{figure*}
-\begin{elpicode}
-type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
-fold4 _ [] [] A A B B.
-fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.
-
-type len list A -> nat -> o.
-len [] z.
-len [_|L] (s X) :- len L X.
-
-\end{elpicode}
-\caption{Implementation of the compiler}
-\label{code:ue-pred}
-\Description[code:ue-pred]{code:ue-pred}
-\end{figure*}
+
+\input{appendix.tex}
 

### 2024-03-19 paper/main.tex
update latex

@@
 \newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}
+
+\newcommand{\linkbeta}{\texttt{link-}\ensuremath{\beta}\xspace}
+\newcommand{\linketa}{\texttt{link-}\ensuremath{\eta}\xspace}
 
@@
 
+Invarianti:
+A destra della eta abbiamo sempre un termine che comincia per $\lambda x. bla$
+
+La deduplicate eta:
+  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2] (a destra non c'è mai un termine con testa rigida)
+  - i due termini a dx vengono unificati con la unif e uno dei due link viene buttato
+    -> si mettono più lambda 
+  - dopo l'unificazione rimane un link [variabile] -> [etaX] e quando
+    risolviamo un eta-link, allora dobbiamo eta-espandere qb il termine di sinistra per
+    poterlo  unifcare con quello di dx.
+
 \subsection{Problems with $\beta$}
@@
 
+The solution to this problem is to modify the compiler such that any sub-term $t$
+considered as a potential $\beta$-redex is replaced with a hole $h$ and a new dedicated
+link, called \linkbeta.
+
+\begin{elpicode}
+  type link-beta tm -> tm -> link.
+\end{elpicode}
+
+This link carries two terms, the former representing the variable $h$ for the
+new created hole and the latter containing the subterm $t$. As for the \linketa,
+we will call $h$ and $t$ respectively the left $lhs$ and the right $rhs$ hand
+side of the \linkbeta.
+
+At compile time, a subterm is \maybebeta, if it has the shape \elpiIn{fapp[fuva
+N | L]} and \elpiIn{distinct L} does not hold. In that case, \elpiIn{L} is split
+in two sublist \elpiIn{PF} and \elpiIn{NPF} such that former is the longest
+prefix of \elpiIn{L} such that \elpiIn{distinct PF} holds. \elpiIn{NPF} is the
+list such that \elpiIn{append PF NPF L}. The $lhs$ is set to a new variable
+named $M$ with \elpiIn{PF} in scope whereas the $rhs$ is given by the term
+\elpiIn{app[uva N' PF | NPF]} where the \Ho variable identified by \elpiIn{N'}
+is mapped to the \Fo variable named \elpiIn{N}.
+
+After its creation, a \linkbeta remain suspended until the head of $rhs$ is
+instantiated by the oracle (see \cref{prop:simulation}). In this case, $rhs$ is
+$\beta$-reduced to a new term, say $t'$. $t'$ is either a term in \llambda, in
+which case $t'$ is unified with the $lhs$, otherwise, the link remain suspended
+and no progress is performed. Another way to wake a \linkbeta up is when the
+$lhs$ is a term \elpiIn{T} and $rhs$ has the shape \elpiIn{app[uva N PF | NPF]}
+and some of the arguments in the \elpiIn{NPF} list become names. This is
+possible after the resolution of other links. In this case, the list \elpiIn{L}
+obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
+again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
+same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
+new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
+create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.
+
+An example justifying this last link manipulation is given by the following
+unification problem:
+
+\begin{elpicode}
+  f = flam x\ flam y\ fapp[F, (X x), y] % $f = \lambda x.\lambda y.F (X x) y$
+\end{elpicode}
+
+\noindent
+from the subsitution $\rho = \{X \mapsto \lambda x.x\}$.
+
+\noindent
 \otext{The following goal: \elpiIn{X = lam x\ x, app[X, a] = a}}

### 2024-03-20 paper/main.tex
Update latex

@@
 \newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}
+
+\newcommand{\linketaM}[3]{\texttt{#1 |- #2 =$\eta$= #3}}
+\newcommand{\linkbetaM}[3]{\texttt{#1 |- #2 =$\beta$= #3}}
 
@@
 
+\begin{textcode}
   La deduplicate eta:
-  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2] (a destra non c'è mai un termine con testa rigida)
+  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2] 
+    (a destra non c'è mai un termine con testa rigida)
   - i due termini a dx vengono unificati con la unif e uno dei due link viene buttato
-    -> si mettono più lambda 
-  - dopo l'unificazione rimane un link [variabile] -> [etaX] e quando
-    risolviamo un eta-link, allora dobbiamo eta-espandere qb il termine di sinistra per
-    poterlo  unifcare con quello di dx.
+    NOTA!! A dx abbiamo sempre un termine della forma lam x.VAR x!!!
+    Altrimenti il link sarebbe stato risolto!!
+  - dopo l'unificazione rimane un link [variabile] -> [etaX] 
+  - nella progress-eta, se a sx abbiamo una constante o un'app, allora eta-espandiamo
+    di uno per poter unificare con il termine di dx.
+\end{textcode}
 
@@
 \begin{elpicode}
-  f = flam x\ flam y\ fapp[F, (X x), y] % $f = \lambda x.\lambda y.F (X x) y$
-\end{elpicode}
-
-\noindent
-from the subsitution $\rho = \{X \mapsto \lambda x.x\}$.
-
-\noindent
-\otext{The following goal: \elpiIn{X = lam x\ x, app[X, a] = a}}
-\otext{We use links-beta}
+  f = flam x\ fapp[F, (X x), a] % $f = \lambda x.F (X x) a$
+\end{elpicode}
+
+\noindent
+under the subsitution $\rho = \{X \mapsto \lambda x.x\}$.
+
+\noindent
+The links generated from this unification problem are:
+
+\begin{textcode}
+X ~$\mapsto$~ X1; F ~$\mapsto$~ X2 % The mappings
+  ~$\vdash$~ X0   =~$\eta$~= x\ `X3 x'
+x ~$\vdash$~ X3 x =~$\beta$~= X2 `X1 x' a
+\end{textcode}
+
+\noindent
+where the first link is a \linketa between the variable \texttt{X0}, representing
+the right side of the unification problem (it is a \maybeeta) and
+\texttt{X3}; and a \linkbeta between the variable \texttt{X3} and the subterm
+\texttt{c0\textbackslash\ X2 `X1 c0' a} (it is a \maybebeta).
+The substitution tells that \textIn{x ~$\vdash$~ X1 x = x}.
+
+We can now represent the \hrun execution from this configuration which will, at
+first, dereference all the links, and then try to solve them. The only link
+being modified is the second one, which is set to \textIn{X3 x =~$\beta$~= X2
+x a}. The $rhs$ of the link has now a variable which is partially in the PF,
+we can therefore remove the original \linkbeta and replace it with the following 
+couple on links:
+
+\begin{textcode}
+  ~$\vdash$~ X1   =~$\eta$~= x\ `X4 x'
+x ~$\vdash$~ X3 x =~$\beta$~= x\ `X4 x' a
+\end{textcode}
+
+By these links we say that \texttt{X1} is now $\eta$-linked to a fresh variable
+\texttt{X4} with arity one. This new variable is used in the new \linkbeta where
+the name \texttt{x} is in its scope. This allows
 

### 2024-03-20 paper/main.tex
rework 2.1

@@
 
-We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
-$\sigma$, we write $\sigma t$ for the application of the substitution to
-$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
-and we assume that the unification of our meta language is correct:
+We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
+with substitution $\sigma$; we write $\sigma t$ for the application of
+the substitution to $t$, and $\sigma X = \{ \sigma t | t \in X\}$ when
+$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
+general than $\sigma'$. We assume that the unification of our meta
+language is correct:
 %
@@
 %
-\newcommand{\solve}{\ensuremath{\mathrm{check}}\xspace}
+\newcommand{\progress}{\ensuremath{\mathrm{check}}\xspace}
 \newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
@@
 \newcommand{\stepHD}[6]{\ensuremath{%
-#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
+#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
 \newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
@@
 Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
-terms) and a call to \solve{} on the set of links. We claim the following:
+terms) and a call to \progress{} on the set of links. We claim the following:
 
@@
 \quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
+    \progress~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
 \quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
@@
 A term $s$ is compiled in a term $t$ where any
-``problematic'' subterm $tp$ is replaced by a unification variable $h$
-and an accessory link, that represent a suspended unification problem
-$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
-We now define ``problematic'' formally.
+``problematic'' subterm $tp$ is replaced by a fresh unification variable $h$
+and an accessory link that represent a suspended unification problem
+$h \Ue tp$. As a result \Ue is ``well behaved'' on $t$, meaning it does not
+contradict \Eo (as it would do on ``problematic'' terms).
+We now define ``problematic'' and ``well behaved'' more formally.
 
@@
 \begin{definition}[\maybeeta]
-  $\maybeeta = \{ t | \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
+  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
 \end{definition}
@@
 \begin{definition}[\maybebeta]
-  $\maybebeta = \{ X t_1 \ldots t_n | t \not\in \llambda \}$.
+  $\maybebeta = \{ X t_1 \ldots t_n ~|~ X t_1 \ldots t_n \not\in \llambda \}$.
 \end{definition}
@@
 
-\marginpar{TODO: find better word}
-\newcommand{\normal}{\ensuremath{\mathrm{normal}}}
-\begin{definition}[Normal form]
+\newcommand{\wellb}{\ensuremath{\mathcal{W}}}
+\begin{definition}[Well behaved set]
 Given a set of terms $X \subseteq \Ho{}$,
 $$
-\normal(X) =
-\mathcal{P}(X) \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
+\wellb(X) \Leftrightarrow \forall t \in \mathcal{P}(X), t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
 $$
@@
 \noindent
-We write $\sigma X = \{ \sigma t | t \in X\}$.
-
-\begin{proposition}[Normal form preservation]\label{prop:nf}
+
+\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
 $\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
 $$
-\normal(\sigma\mathcal{T}) \land
+\wellb(\sigma\mathcal{T}) \land
 \stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
-\Rightarrow \normal((\sigma \cup \sigma')\mathcal{T})
+\Rightarrow \wellb((\sigma \cup \sigma')\mathcal{T})
 $$
@@
 \noindent
-In particular this guarantees that is we start from normal terms
-we never introduce eta-long or non-beta-normal terms in $\sigma'$.
+A less formal way to state \ref{prop:nf} is that \hstep{} never
+``commits'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
+that could be erased by an $\eta$-contraction),
+nor puts in $\sigma$ a flexible application outside \llambda{} (an application
+node that could be erased by a $\beta$-reduction).
 
@@
 decompilation can introduce (actually restore) terms in
-\maybeeta or \maybebeta.
+\maybeeta or \maybebeta that were move out of the way (put in $\mathbb{L}$)
+by compilation.
 

### 2024-03-20 paper/main.tex
fix

@@
 A term $s$ is compiled in a term $t$ where any
-``problematic'' subterm $tp$ is replaced by a fresh unification variable $h$
+``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
 and an accessory link that represent a suspended unification problem
-$h \Ue tp$. As a result \Ue is ``well behaved'' on $t$, meaning it does not
+$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, meaning it does not
 contradict \Eo (as it would do on ``problematic'' terms).
@@
 \stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
-\Rightarrow \wellb((\sigma \cup \sigma')\mathcal{T})
+\Rightarrow \wellb(\sigma' \mathcal{T})
 $$

### 2024-03-20 paper/appendix.tex
update appendix
appendix meta language

@@
 
-\begin{figure*}
+\section{The object language}
+
 \begin{elpicode}
@@
 \end{elpicode}
-\caption{Full implementation of the \Eo predicate for \Fo}
-\label{code:eo-pred}
-\Description[code:eo-pred]{code:eo-pred}
-\end{figure*}
 
-\begin{figure*}
+\section{The meta language}
+% \begin{figure*}
 \begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
-% Congruence
-(app A ~\Ue~app B) R S :- fold2 ~(\Ue)~A B R S.
-(lam F ~\Ue~lam G) R S :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) R S.
-~\Ue~(con C) (con C) S S.
-% deref
-(uva N A ~\Ue~T) R S :- set? N S F, move F A T1, (T1 ~\Ue~T) R S.
-(T ~\Ue~uva N A) R S :- set? N S F, move F A T1, (T ~\Ue~T1) R S.
+  typeabbrev subst list (option assignment).
+
+  kind inctx type -> type.
+  type abs (tm -> inctx A) -> inctx A.
+  type val A -> inctx A.
+
+  typeabbrev assignment (inctx tm).
+
+  kind tm  type.
+  type app list tm -> tm.
+  type lam (tm -> tm) -> tm.
+  type con string -> tm.
+  type uva  address -> list tm -> tm.
+
+  type (==l) tm -> tm -> subst -> subst ->  o.
+  % congruence
+  ((app L1) ==l (app L2)) S S1 :- fold2 (==l) L1 L2 S S1.
+  ((lam F1) ==l (lam F2)) S S1 :- 
+    pi x\ copy x x => ((F1 x) ==l (F2 x)) S S1.
+  ((con X) ==l (con X)) S S.
+  % set variables
+  ((uva N Args) ==l T) S S1 :- 
+    mem.set? N S F,!, move F Args T1, (T1 ==l T) S S1.
+  (T ==l (uva N Args)) S S1 :- 
+    mem.set? N S F,!, move F Args T1, (T ==l T1) S S1.
   % flex-flex
-(uva N A ~\Ue~uva M B) S S3 :-  unset? M, unset? N,
-  distinct A, distinct B, 
-  new S W S1, prune W Args1 B Ass, 
-  assign N S1 Ass S2, assign M S2 Ass S3.
-% assignment
-(uva N A ~\Ue~T) R S :- distinct A, not (T = uva _ _), not_occ N S T,  
-  bind A T T1, assign N S T1 S1.
-(T ~\Ue~uva N A) R S :- distinct A, not (T = uva _ _), not_occ N S T, 
-  bind A T T1, assign N S T1 S1.
+  ((uva M A1) ==l (uva N A2)) S1 S2 :- !, 
+    pattern-fragment A1, pattern-fragment A2, 
+    prune! M A1 N A2 S1 S2.
+  ((uva N Args) ==l T) S S1 :- not_occ N S T, pattern-fragment Args, 
+    bind T Args T1, mem.assign N S T1 S1.
+  % variable assigment
+  (T ==l (uva N Args)) S S1 :- not_occ N S T, pattern-fragment Args, 
+    bind T Args T1, mem.assign N S T1 S1.
+  (N ==l N) S S :- name N.
 
-type distinct list A -> o.
-distinct [].
-distinct [X|XS] :- name X, not(mem X XS),
-distinct XS.
+  % Note: We suppose the scopes to always be in PF
+  type prune! address -> list ho.tm -> address -> 
+              list ho.tm -> subst -> subst -> o.
+  prune! N A  N A  S S :- !.
+  prune! M A  N A  S1 S2 :- !, bind (uva M A) A Ass, 
+    mem.assign N S1 Ass S2.
+  prune! N A1 N A2 S1 S3 :- !,
+    std.assert!(len A1 {len A2}) "Not typechecking", !,
+    mem.new S1 W S2, prune-same-variable W A1 A2 [] Ass,
+    mem.assign N S2 Ass S3.
+  prune! N A1 M A2 S1 S4 :- !,
+    mem.new S1 W S2, prune-diff-variables W A1 A2 Ass1 Ass2, 
+    mem.assign N S2 Ass1 S3,
+    mem.assign M S3 Ass2 S4.
 
-typeabbrev memory A (list (option A)).
-type set? nat -> memory A -> A -> o.
-set? N S T :- nth N S (some T).
-type unset? nat -> memory A -> o.
-unset? N S :- nth N S none.
-type assign nat -> memory A -> A -> memory A -> o.
-assign z [none|M] T [some T|M].
-assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
-kind nat type.
-type z nat.
-type s nat -> nat.
-type nth nat -> list A -> A -> o.
-nth z [X|_] X.
-nth (s N) [_|L] X :- nth N L X.
+  type prune-same-variable address -> list tm -> list tm -> 
+                            list tm -> assignment -> o.
+  prune-same-variable N [] [] ACC (val (uva N Args)) :- 
+    std.rev ACC Args.
+  prune-same-variable N [X|XS] [X|YS] ACC (abs F) :- 
+    pi x\ prune-same-variable N XS YS [x|ACC] (F x).
+  prune-same-variable N [_|XS] [_|YS] ACC (abs F) :- 
+    pi x\ prune-same-variable N XS YS ACC (F x).
 
-type new memory A -> nat -> memory A -> o.
-new [] z [none].
-new [X|XS] (s N) [X|YS] :- new XS N YS.
+  type prune-build-ass1 address -> list tm -> 
+                        list bool -> assignment -> o.
+  prune-build-ass1 N Acc [] (val (uva N Args)) :- 
+    std.rev Acc Args.
+  prune-build-ass1 N Acc [tt|L] (abs T) :- 
+    pi x\ prune-build-ass1 N [x|Acc] L (T x).
+  prune-build-ass1 N Acc [ff|L] (abs T) :- 
+    pi x\ prune-build-ass1 N Acc L (T x).
 
-type prune .
-type move .
-type beta.
-type bind.
-type not_occ.
-TODO
+  type build-order list nat -> list tm -> list tm -> o.
+  build-order L T R :-
+    len L Len, list-init Len z 
+      (p\r\ sigma Index Elt\ index L p Index, nth Index T r) R.
 
-type fold2 (A -> A1 -> B -> B -> o) -> list A -> list A1 -> B -> B -> o.
-fold2 _ [] [] A A.
-fold2 F [X|XS] [Y|YS] A A1 :- F X Y A A0, fold2 F XS YS A0 A1.
+  type prune-build-ass2 address -> list tm -> list bool -> 
+                      list nat -> assignment -> o.
+  prune-build-ass2 N Acc [] Pos (val (uva N Args)) :- 
+    std.rev Acc Acc', build-order Pos Acc' Args.
+  prune-build-ass2 N Acc [tt|L] Pos (abs T) :- 
+    pi x\ prune-build-ass2 N [x|Acc] L Pos (T x).
+  prune-build-ass2 N Acc [ff|L] Pos (abs T) :- 
+    pi x\ prune-build-ass2 N Acc L Pos (T x).
 
+  type keep list A -> A -> bool -> o.
+  keep L A tt :- mem L A, !.
+  keep _ _ ff.
+
+  type prune-diff-variables address -> list tm -> list tm -> 
+                            assignment -> assignment -> o.
+  prune-diff-variables N Args1 Args2 Ass1 Ass2 :-
+    std.map Args1 (keep Args2) Bits1, 
+    prune-build-ass1 N [] Bits1 Ass1,
+    std.map Args2 (keep Args1) Bits2, 
+    std.filter Args1 (mem Args2) ToKeep1,
+    std.filter Args2 (mem Args1) ToKeep2,
+    std.map ToKeep2 (index ToKeep1) Pos,
+    prune-build-ass2 N [] Bits2 Pos Ass2.
+
+  type move assignment -> list tm -> tm -> o.
+  move (abs Bo)        [H|L] R :- move (Bo H) L R.
+  move (val A)         []    A :- !.
+  move (val (uva N A)) L     (uva N X) :- std.append A L X.
+
+  type beta tm -> list tm -> tm -> o.
+  beta A [] A.
+  beta (lam Bo) [H | L] R :- beta (Bo H) L R.
+  beta (app A) L (app X) :- std.append A L X.
+  beta (uva N A) L (uva N A') :- std.append A L A'.
+  beta (con H) L (app [con H | L]).
+
+  type not_occ_aux address -> subst -> tm -> o.
+  not_occ_aux N S (uva M _) :- mem.unset? M S, not (N = M).
+  not_occ_aux N S (uva M Args) :- mem.set? M S F, 
+    move F Args T, not_occ_aux N S T.
+  not_occ_aux N S (app L) :- forall1 (not_occ_aux N S) L.
+  not_occ_aux N S (lam F) :- pi x\ not_occ_aux N S (F x).
+  not_occ_aux _ _ (con _).
+  not_occ_aux _ _ X :- name X.
+
+  type not_occ  address -> subst -> tm -> o.
+  not_occ N _ (uva N _).
+  not_occ N S (uva M Args) :- mem.set? M S F, 
+    move F Args T, not_occ N S T.
+  not_occ N S (uva M Args) :- mem.unset? M S, 
+    std.forall Args (not_occ_aux N S).
+  not_occ _ _ (con _).
+  not_occ N S (app L) :- not_occ_aux N S (app L).
+  not_occ N S (lam L) :- pi x\ not_occ N S (L x).
+  not_occ _ _ X :- name X.
+
+  type copy tm -> tm -> o.
+  copy (app L)   (app L') :- forall2 copy L L'.
+  copy (lam T)   (lam T') :- pi x\ copy x x => copy (T x) (T' x).
+  copy (uva N L) (uva N L') :- forall2 copy L L'.
+  copy (con C)   (con C).
+  copy N N :- name N.
+
+  type bind tm -> list tm -> assignment -> o.
+  bind T [] (val T') :- copy T T'.
+  bind T [X | TL] (abs T') :- pi x\ copy X x => bind T TL (T' x).
+
+  type deref subst -> tm -> tm -> o.
+  deref S X _ :- (var S; var X), halt "flex deref".
+  deref H (uva N L)  X         :- mem.set? N H T, 
+    move T L X', !, deref H X' X.
+  deref H (app L)   (app L1)   :- forall2 (deref H) L L1.
+  deref _ (con X)   (con X).
+  deref H (uva X L) (uva X L1) :- mem.unset? X H, 
+    forall2 (deref H) L L1.
+  deref H (lam F)   (lam G)    :- pi x\ deref H (F x) (G x).
+  deref _ N         N          :- name N.
+
+  type deref-assmt subst -> assignment -> assignment -> o.
+  deref-assmt S (abs T) (abs R) :- pi x\ deref-assmt S (T x) (R x).
+  deref-assmt S (val T) (val R) :- deref S T R.
 \end{elpicode}
-\caption{Implementation of the \Ue predicate for \Ho}
-\label{code:ue-pred}
-\Description[code:ue-pred]{code:ue-pred}
-\end{figure*}
+% \caption{Implementation of the \Ue predicate for \Ho}
+% \label{code:ue-pred}
+% \Description[code:ue-pred]{code:ue-pred}
+% \end{figure*}
 
-\begin{figure*}
+\section{The compiler}
+% \begin{figure*}
+% \caption{Implementation of the compiler}
+% \label{code:ue-pred}
+% \Description[code:ue-pred]{code:ue-pred}
+% \end{figure*}
+
+\section{The progress function}
+
+\section{The decompiler}
+
+\section{Auxiliary functions}
+
 \begin{elpicode}
@@
   \end{elpicode}
-\caption{Implementation of the compiler}
-\label{code:ue-pred}
-\Description[code:ue-pred]{code:ue-pred}
-\end{figure*}

### 2024-03-20 paper/appendix.tex
code refactor
refactor maybe-eta-of

@@
 Note that \elpiIn{(a infix b) c d} de-sugars to \elpiIn{(infix) a b c d}.
+
+\section{The memory}
+
+\begin{elpicode}
+  kind address type.
+  type addr nat -> address.
+
+  typeabbrev (mem A) (list (option A)).
+
+  type get nat -> mem A -> A -> o.
+  get z (some Y :: _) Y.
+  get (s N) (_ :: L) X :- get N L X.
+
+  type alloc-aux nat -> mem A -> mem A -> o.
+  alloc-aux z [] [none] :- !.
+  alloc-aux z L L.
+  alloc-aux (s N) [] [none | M] :- alloc-aux N [] M.
+  alloc-aux (s N) [X | L] [X | M] :- alloc-aux N L M.
+
+  type alloc address -> mem A -> mem A -> o.
+  alloc (addr A as Ad) Mem1 Mem2 :- unset? Ad Mem1, 
+    alloc-aux A Mem1 Mem2.
+
+  type new-aux mem A -> nat -> mem A -> o.
+  new-aux [] z [none].
+  new-aux [A | As] (s N) [A | Bs] :- new-aux As N Bs.
+
+  type new mem A -> address -> mem A -> o.
+  new Mem1 (addr Ad) Mem2 :- new-aux Mem1 Ad Mem2.
+
+  type set? address -> mem A -> A -> o.
+  set? (addr A) Mem Val :- get A Mem Val.
+
+  type unset? address -> mem A -> o.
+  unset? Addr Mem :- not (set? Addr Mem _).
+
+  type assign-aux nat -> mem A -> A -> mem A -> o.
+  assign-aux z (none :: L) Y (some Y :: L).
+  assign-aux (s N) (X :: L) Y (X :: L1) :- assign-aux N L Y L1.
+
+  type assign address -> mem A -> A -> mem A -> o.
+  assign (addr A) Mem1 Val Mem2 :- assign-aux A Mem1 Val Mem2.
+\end{elpicode}
 
@@
 \section{The compiler}
+
+\begin{elpicode}
+
+\end{elpicode}
 % \begin{figure*}

### 2024-03-20 paper/appendix.tex
update appendix

@@
   deref-assmt S (val T) (val R) :- deref S T R.
-\end{elpicode}
-% \caption{Implementation of the \Ue predicate for \Ho}
-% \label{code:ue-pred}
-% \Description[code:ue-pred]{code:ue-pred}
-% \end{figure*}
 
@@
 \begin{elpicode}
+  type use-binder fo.fm -> fo.fm -> o.
+  use-binder N N.
+  use-binder N (fo.fapp L) :- exists (use-binder N) L.
+  use-binder N (fo.flam B) :- pi x\ use-binder N (B x).
 
+  type maybe-eta fo.fm -> fo.fm -> list fo.fm -> o.
+  maybe-eta N (fo.fapp[fo.fuva _|Args]) _ :- !,
+    exists (x\ maybe-eta-of [] N x) Args, !. 
+  maybe-eta N (fo.flam B) L :- !, pi x\ maybe-eta N (B x) [x | L].
+  maybe-eta _ (fo.fapp [fo.fcon _|Args]) L :-
+    split-last-n {len L} Args First Last,
+    forall1 (x\ forall1 (y\ not (use-binder x y)) First) L,
+    forall2 (maybe-eta-of []) {std.rev L} Last.
+
+  type maybe-eta-of list fo.fm -> fo.fm -> fo.fm -> o.
+  maybe-eta-of _ N N :- !.
+  maybe-eta-of L N (fo.fapp[fo.fuva _|Args]) :- !, 
+    % Head is flex -> if N is in Args, we can have eta
+    forall1 (x\ exists (maybe-eta-of [] x) Args) [N|L]. 
+  maybe-eta-of L N (fo.flam B) :- !, 
+    pi x\ maybe-eta-of [x | L] N (B x).
+  maybe-eta-of L N (fo.fapp [N|Args]) :-
+    % Head is rigid -> to be an eta we consider also L
+    last-n {len L} Args R,
+    forall2 (maybe-eta-of []) R {std.rev L}.
+
+  type locale ho.tm -> o.
+  type get-scope-aux ho.tm -> list ho.tm -> o.
+  get-scope-aux (ho.con _) [].
+  get-scope-aux (ho.uva _ L) L1 :- 
+    forall2 get-scope-aux L R,
+    flatten R L1.
+  get-scope-aux (ho.lam B) L1 :- 
+    pi x\ locale x => get-scope-aux (B x) L1.
+  get-scope-aux (ho.app L) L1 :- 
+    forall2 get-scope-aux L R,
+    flatten R L1.
+  get-scope-aux X [X] :- name X, not (locale X).
+  get-scope-aux X [] :- name X, (locale X).
+
+  type get-scope ho.tm -> list ho.tm -> o.
+  get-scope T Scope :- names N,
+    get-scope-aux T ScopeDuplicata,
+    std.filter N (mem ScopeDuplicata) Scope.
+
+  type close-links (ho.tm -> links) -> links -> o.
+  close-links (_\[]) [].
+  close-links (v\[L|XS v]) [L|YS] :- !, close-links XS YS.
+  close-links (v\[(L v)|XS v]) [ho.abs L|YS] :- !, 
+    close-links XS YS.
+
+  type comp-lam (fo.fm -> fo.fm) -> (ho.tm -> ho.tm) -> 
+    mappings -> mappings -> links -> links -> ho.subst -> 
+      ho.subst -> o.
+  comp-lam F F1 Mappings1 Mappings2 L L2 S S1 :-
+    pi x y\ (pi M L S\ comp x y M M L L S S) => 
+      comp (F x) (F1 y) Mappings1 Mappings2 L (L1 y) S S1, 
+        close-links L1 L2.
+
+  type comp fo.fm -> ho.tm -> mappings -> mappings -> links -> 
+    links -> ho.subst -> ho.subst -> o.
+  comp (fo.fcon X) (ho.con X)        Map1 Map1 L1 L1 S1 S1.
+  comp (fo.flam F) (ho.uva E0 Scope) Map1 Map2 L1 L3 S1 S3 :-
+    (pi x\ maybe-eta x (F x) [x]), !,
+    mem.new S1 E0 S2,
+    comp-lam F F1 Map1 Map2 L1 L2 S2 S3,
+    (pi x\ locale x => get-scope (F1 x) Scope), 
+    NewLink = ho.val(link-eta (ho.uva E0 Scope) (ho.lam F1)), 
+    L3 = [NewLink| L2].
+  comp (fo.flam F) (ho.lam F1)       Map1 Map2 L1 L2 S1 S2 :- 
+    comp-lam F F1 Map1 Map2 L1 L2 S1 S2.
+  comp (fo.fuva N) (ho.uva M []) Map1 Map2 L L S S1 :- 
+    alloc Map1 Map2 (fv N) (hv M (arity z)) S S1.
+  comp (fo.fapp [fo.fuva N|Ag]) (ho.uva M' Scope) Map1 Map3 L1 L3 S1 S4 :- !,
+    split-pf Ag PF NPF,
+    forall2 (x\y\ comp x y _ _ _ _ _ _) PF PF1,
+    fold6 comp NPF NPF1 Map1 Map2 L1 L2 S1 S2,
+    len PF Arity, 
+    alloc Map2 Map3 (fv N) (hv M (arity Arity)) S2 S3,
+    if (NPF = []) 
+      (Scope = PF1, M' = M, S3 = S4, L2 = L3)
+      (get-scope (ho.app {std.append PF1 NPF1}) Scope, 
+        mem.new S3 M' S4,
+        L3 = [@val-link-beta (ho.uva M' Scope) (ho.app[ho.uva M PF1 | NPF1]) | L2]).
+  comp (fo.fapp A) (ho.app A1) Map1 Map2 L1 L2 S1 S2 :- 
+    fold6 comp A A1 Map1 Map2 L1 L2 S1 S2.
+
+  type make-eta-link-aux nat -> address -> address -> 
+    list ho.tm -> links -> ho.subst -> ho.subst -> o.
+  make-eta-link-aux z Ad1 Ad2 Scope1 L H1 H1  :-
+    std.rev Scope1 Scope, eta-expand (ho.uva Ad2 Scope) @one T1,
+    L = [@val-link-eta (ho.uva Ad1 Scope) T1].
+  make-eta-link-aux (s N) Ad1 Ad2 Scope1 L H1 H3 :-
+    std.rev Scope1 Scope, mem.new H1 Ad H2, 
+    eta-expand (ho.uva Ad Scope) @one T2,
+    (pi x\ make-eta-link-aux N Ad Ad2 [x|Scope1] (L1 x) H2 H3),
+    close-links L1 L2,
+    L = [@val-link-eta (ho.uva Ad1 Scope) T2 | L2].
+
+  type make-eta-link nat -> nat -> address -> address -> 
+          list ho.tm -> links -> ho.subst -> ho.subst -> o.
+  make-eta-link (s N) z Ad1 Ad2 Vars L H H1 :- 
+    make-eta-link-aux N Ad2 Ad1 Vars L H H1.
+  make-eta-link z (s N) Ad1 Ad2 Vars L H H1 :- 
+    make-eta-link-aux N Ad1 Ad2 Vars L H H1.
+  make-eta-link (s N) (s M) Ad1 Ad2 Vars Links H H1 :-
+    (pi x\ make-eta-link N M Ad1 Ad2 [x|Vars] (L x) H H1),
+    close-links L Links.
+
+  type deduplicate-mappings mappings -> mappings -> 
+          ho.subst -> ho.subst -> links -> links -> o.
+  deduplicate-mappings [] [] H H L L.
+  deduplicate-mappings [(mapping (fv O) (hv M (arity LenM)) as X1) | Map1] Map2 H1 H3 L1 L3 :-
+    take-list Map1 (mapping (fv O) (hv M' (arity LenM'))) _, !,
+    make-eta-link LenM LenM' M M' [] New H1 H2,
+    std.append New L1 L2,
+    deduplicate-mappings Map1 Map2 H2 H3 L2 L3.
+  deduplicate-mappings [A|As] [A|Bs] H1 H2 L1 L2 :- 
+    deduplicate-mappings As Bs H1 H2 L1 L2, !.
 \end{elpicode}
-% \begin{figure*}
-% \caption{Implementation of the compiler}
-% \label{code:ue-pred}
-% \Description[code:ue-pred]{code:ue-pred}
-% \end{figure*}
 

### 2024-03-20 paper/appendix.tex
appendix add progress

@@
 
+\begin{elpicode}
+
+  macro @one :- s z.
+
+  type contract-rigid list ho.tm -> ho.tm -> ho.tm -> o.
+  contract-rigid L (ho.lam F) T :- 
+    pi x\ contract-rigid [x|L] (F x) T. % also checks H Prefix does not see x
+  contract-rigid L (ho.app [H|Args]) T :- 
+    std.rev L LRev, std.appendR Prefix LRev Args,
+    if (Prefix = []) (T = H) (T = ho.app [H|Prefix]).
+
+  type progress-eta-link ho.tm -> ho.tm -> ho.subst -> ho.subst -> links -> o.
+  progress-eta-link (ho.app _ as T) (ho.lam x\ _ as T1) H H1 [] :- !, 
+    ({eta-expand T @one} ==l T1) H H1.
+  progress-eta-link (ho.con _ as T) (ho.lam x\ _ as T1) H H1 [] :- !, 
+    ({eta-expand T @one} ==l T1) H H1.
+  progress-eta-link (ho.lam _ as T) T1 H H1 [] :- !, 
+    (T ==l T1) H H1.
+  progress-eta-link (ho.uva _ _ as X) T H H1 [] :- 
+    contract-rigid [] T T1, !, (X ==l T1) H H1.
+  progress-eta-link (ho.uva Ad _ as T1) T2 H H [@val-link-eta T1 T2] :- !, 
+    if (ho.not_occ Ad H T2) true fail.
+
+  type is-in-pf ho.tm -> o.
+  is-in-pf (ho.app [ho.uva _ _ | _]) :- !, fail.
+  is-in-pf (ho.lam B) :- !, pi x\ is-in-pf (B x).
+  is-in-pf (ho.con _).
+  is-in-pf (ho.app L) :- forall1 is-in-pf L.
+  is-in-pf N :- name N.
+  is-in-pf (ho.uva _ L) :- pattern-fragment L.
+
+  type arity ho.tm -> nat -> o.
+  arity (ho.con _) z.
+  arity (ho.app L) A :- len L A.
+
+  type occur-check-err ho.tm -> ho.tm -> ho.subst -> o.
+  occur-check-err (ho.con _) _ _ :- !.
+  occur-check-err (ho.app _) _ _ :- !.
+  occur-check-err (ho.lam _) _ _ :- !.
+  occur-check-err (ho.uva Ad _) T S :-
+    not (ho.not_occ Ad S T).
+
+  type progress-beta-link-aux ho.tm -> ho.tm -> 
+          ho.subst -> ho.subst -> links -> o.
+  progress-beta-link-aux T1 T2 S1 S2 [] :-  is-in-pf T2, !,
+    (T1 ==l T2) S1 S2.
+  progress-beta-link-aux T1 T2 S S [@val-link-beta T1 T2] :- !.
+
+  type progress-beta-link ho.tm -> ho.tm -> ho.subst -> 
+        ho.subst -> links -> o.
+  progress-beta-link T (ho.app[ho.uva V Scope | L] as T2) S S2 [@val-link-beta T T2] :- 
+    arity T Arity, len L ArgsNb, ArgsNb >n Arity, !,
+    minus ArgsNb Arity Diff, mem.new S V1 S1,
+    eta-expand (ho.uva V1 Scope) Diff T1,
+    ((ho.uva V Scope) ==l T1) S1 S2.
+
+  progress-beta-link (ho.uva _ _ as T) (ho.app[ho.uva Ad1 Scope1 | L1] as T1) S1 S3 NewLinks :-
+    std.append Scope1 L1 Scope1L,
+    split-pf Scope1L Scope2 L2,
+    not (Scope1 = Scope2), !,
+    mem.new S1 Ad2 S2,
+    len Scope1 Scope1Len,
+    len Scope2 Scope2Len,
+    make-eta-link Scope1Len Scope2Len Ad1 Ad2 [] LinkEta S2 S3,
+    if (L2 = []) (NewLinks = LinkEta, T2 = ho.uva Ad2 Scope2) 
+      (T2 = ho.app [ho.uva Ad2 Scope2 | L2], 
+      NewLinks = [@val-link-beta T T2 | LinkEta]).
+
+  progress-beta-link T1 (ho.app[ho.uva _ _ | _] as T2) _ _ _ :- 
+    not (T1 = ho.uva _ _), !, fail.
+
+  progress-beta-link (ho.uva _ _ as T) (ho.app[ho.uva _ _ | _] as T2) S1 _ _ :- 
+    occur-check-err T T2 S1, !, fail.
+
+  progress-beta-link T1 (ho.app[ho.uva _ _ | _] as T2) H H [@val-link-beta T1 T2] :- !.
+
+  progress-beta-link T1 (ho.app [Hd | Tl]) S1 S2 B :-
+    ho.beta Hd Tl T3, 
+    progress-beta-link-aux T1 T3 S1 S2 B.
+
+  type solve-link-abs link -> links -> ho.subst -> ho.subst -> o.
+  solve-link-abs (ho.abs X) R H H1 :- 
+    pi x\ ho.copy x x => (pi S\ ho.deref S x x) => 
+      solve-link-abs (X x) (R' x) H H1,
+    close-links R' R.
+
+  solve-link-abs (@val-link-eta A B) NewLinks S S1 :- !,
+    progress-eta-link A B S S1 NewLinks.
+
+  solve-link-abs (@val-link-beta A B) NewLinks S S1 :- !,
+    progress-beta-link A B S S1 NewLinks.
+
+  type take-link link -> links -> link -> links -> o.
+  take-link A [B|XS] B XS :- link-abs-same-lhs A B, !.
+  take-link A [L|XS] B [L|YS] :- take-link A XS B YS.
+
+  type link-abs-same-lhs link -> link -> o.
+  link-abs-same-lhs (ho.abs F) B :- 
+    pi x\ link-abs-same-lhs (F x) B.
+  link-abs-same-lhs A (ho.abs G) :- 
+    pi x\ link-abs-same-lhs A (G x).
+  link-abs-same-lhs (@val-link-eta (ho.uva N _) _) (@val-link-eta (ho.uva N _) _).
+
+  type solve-links links -> links -> ho.subst -> ho.subst -> o.
+  solve-links [] [] X X.
+  solve-links [A|L1] [A|L3] S S2 :- take-link A L1 B L2, !,
+    same-link-eta A B S S1, 
+    solve-links L2 L3 S1 S2.
+  solve-links [L0|L1] L3 S S2 :- deref-link S L0 L,
+    solve-link-abs L R S S1, !,
+    solve-links L1 L2 S1 S2, std.append R L2 L3.
+\end{elpicode}
+
 \section{The decompiler}

### 2024-03-20 paper/appendix.tex
appendix: update progress

@@
 
+  type same-link-eta link -> link -> ho.subst -> ho.subst -> o.
+  same-link-eta (ho.abs F) B H H1 :- !, pi x\ same-link-eta (F x) B H H1.
+  same-link-eta A (ho.abs G) H H1 :- !, pi x\ same-link-eta A (G x) H H1.
+  same-link-eta (@val-link-eta (ho.uva N S1) A)
+                (@val-link-eta (ho.uva N S2) B) H H1 :-
+    std.map2 S1 S2 (x\y\r\ r = ho.copy x y) Perm,
+    Perm => ho.copy A A',
+    (A' ==l B) H H1.
+
   type solve-links links -> links -> ho.subst -> ho.subst -> o.

### 2024-03-20 paper/appendix.tex
appendix: update decompiler section

@@
   deref-assmt S (val T) (val R) :- deref S T R.
-
+\end{elpicode}
 \section{The compiler}
@@
 
+\begin{elpicode}
+
+  type abs->lam ho.assignment -> ho.tm -> o.
+  abs->lam (ho.abs T) (ho.lam R)  :- !, pi x\ abs->lam (T x) (R x).
+  abs->lam (ho.val A) A.
+
+  type commit-links-aux link -> ho.subst -> ho.subst -> o.
+  commit-links-aux (@val-link-eta T1 T2) H1 H2 :- 
+    ho.deref H1 T1 T1', ho.deref H1 T2 T2',
+    (T1' ==l T2') H1 H2.
+  commit-links-aux (@val-link-beta T1 T2) H1 H2 :- 
+    ho.deref H1 T1 T1', ho.deref H1 T2 T2',
+    (T1' ==l T2') H1 H2.
+  commit-links-aux (ho.abs B) H H1 :- 
+    pi x\ commit-links-aux (B x) H H1.
+
+  type commit-links links -> links -> ho.subst -> ho.subst -> o.
+  commit-links [] [] H H.
+  commit-links [Abs | Links] L H H2 :- 
+    commit-links-aux Abs H H1, !, commit-links Links L H1 H2.
+
+  type decompl-subst mappings -> mappings -> ho.subst -> 
+    fo.subst -> fo.subst -> o.
+  decompl-subst _ [A|_] _ _ _ :- fail.
+  decompl-subst _ [] _ F F.
+  decompl-subst Map [mapping (fv VO) (hv VM _)|Tl] H F F2 :- 
+    mem.set? VM H T, !, 
+    ho.deref-assmt H T TTT,
+    abs->lam TTT T', tm->fm Map T' T1, 
+    fo.eta-contract T1 T2, mem.assign VO F T2 F1,
+    decompl-subst Map Tl H F1 F2. 
+  decompl-subst Map [mapping _ (hv VM _)|Tl] H F F2 :- 
+    mem.unset? VM H, decompl-subst Map Tl H F F2.
+
+  type tm->fm mappings -> ho.tm -> fo.fm -> o.
+  tm->fm _ (ho.con C)  (fo.fcon C).
+  tm->fm L (ho.lam B1) (fo.flam B2) :- 
+    pi x y\ tm->fm _ x y => tm->fm L (B1 x) (B2 y).
+  tm->fm L (ho.app L1) T :- forall2 (tm->fm L) L1 [Hd|Tl], 
+    fo.mk-app Hd Tl T.
+  tm->fm L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
+    forall2 (tm->fm L) TL T1, fo.mk-app (fo.fuva VO) T1 T.
+
+  type add-new-mappings-aux ho.subst -> list ho.tm -> mappings -> 
+        mappings ->  fo.subst -> fo.subst -> o.
+  add-new-mappings-aux _ [] _ [] S S.
+  add-new-mappings-aux H [T|Ts] L L2 S S2 :- 
+    add-new-mappings H T L L1 S S1, 
+    add-new-mappings-aux H Ts L1 L2 S1 S2.
+
+  type add-new-mappings ho.subst -> ho.tm -> mappings -> 
+      mappings ->  fo.subst -> fo.subst -> o.
+  add-new-mappings _ (ho.uva N _) Map [] F1 F1 :- 
+    mem Map (mapping _ (hv N _)), !.
+  add-new-mappings H (ho.uva N L) Map [Map1 | MapL] F1 F3 :-
+    mem.new F1 M F2,
+    len L Arity, Map1 = mapping (fv M) (hv N (arity Arity)),
+    add-new-mappings H (ho.app L) [Map1 | Map] MapL F2 F3.
+  add-new-mappings H (ho.lam B) Map NewMap F1 F2 :- 
+    pi x\ add-new-mappings H (B x) Map NewMap F1 F2.
+  add-new-mappings H (ho.app L) Map NewMap F1 F3 :- 
+    add-new-mappings-aux H L Map NewMap F1 F3.
+  add-new-mappings _ (ho.con _) _ [] F F :- !.
+  add-new-mappings _ N _ [] F F :- name N.
+
+  type complete-mapping-under-ass ho.subst -> ho.assignment -> 
+    mappings -> mappings ->  fo.subst -> fo.subst -> o.
+  complete-mapping-under-ass H (ho.val Val) Map1 Map2 F1 F2 :- 
+    add-new-mappings H Val Map1 Map2 F1 F2.
+  complete-mapping-under-ass H (ho.abs Abs) Map1 Map2 F1 F2 :- 
+    pi x\ complete-mapping-under-ass H (Abs x) Map1 Map2 F1 F2.
+
+  type complete-mapping ho.subst -> ho.subst ->   
+    mappings -> mappings -> fo.subst -> fo.subst -> o.
+  complete-mapping _ [] L L F F.
+  complete-mapping H [none | Tl] L1 L2 F1 F2 :-   
+    complete-mapping H Tl L1 L2 F1 F2.
+  complete-mapping H [some T0 | Tl] L1 L3 F1 F3 :-
+    ho.deref-assmt H T0 T,
+    complete-mapping-under-ass H T L1 L2 F1 F2, 
+    std.append L1 L2 LAll,
+    complete-mapping H Tl LAll L3 F2 F3.
+
+  type decompile mappings -> links -> ho.subst -> 
+    fo.subst -> fo.subst -> o.
+  decompile Map1 L HO FO FO2 :- 
+    commit-links L L1_ HO HO1, !,
+    complete-mapping HO1 HO1 Map1 Map2 FO FO1,
+    decompl-subst Map2 Map2 HO1 FO1 FO2.
+\end{elpicode}
+
 \section{Auxiliary functions}
@@
 \begin{elpicode}
-  type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
+  type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> 
+    list A1 -> B -> B -> C -> C -> o.
   fold4 _ [] [] A A B B.
-  fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.
+  fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, 
+    fold4 F XS YS A0 A1 B0 B1.
 

### 2024-03-20 paper/main.tex
some comments on 6

@@
 On the other hand, the \Ue is not designed to understand how the $\beta$-redexes
-work in the onject language. Therefore, even if we know that $F$ is assigned
+work in the object language. Therefore, even if we know that $F$ is assigned
 to $\lambda x.x$, \Ue is not able to unify $F a$ with $a$. On the other hand,
@@
 assigned to $(\lambda x.x) a$ which breaks the invariant saying that the 
-substitution of the meta language always contain only terms in normal form.
+substitution of the meta language does not generate terms outide \wellb{} (Property \ref{prop:nf}).
 
@@
 new created hole and the latter containing the subterm $t$. As for the \linketa,
-we will call $h$ and $t$ respectively the left $lhs$ and the right $rhs$ hand
-side of the \linkbeta.
-
-At compile time, a subterm is \maybebeta, if it has the shape \elpiIn{fapp[fuva
+we will call $h$ and $t$ respectively the left hand side (LHS)
+and the right hand side (RHS) of the \linkbeta.
+
+A subterm is in \maybebeta{} if it has the shape \elpiIn{fapp[fuva
 N | L]} and \elpiIn{distinct L} does not hold. In that case, \elpiIn{L} is split
@@
 prefix of \elpiIn{L} such that \elpiIn{distinct PF} holds. \elpiIn{NPF} is the
-list such that \elpiIn{append PF NPF L}. The $lhs$ is set to a new variable
-named $M$ with \elpiIn{PF} in scope whereas the $rhs$ is given by the term
+list such that \elpiIn{append PF NPF L}. The LHS is set to a new variable
+named $M$ with \elpiIn{PF} in scope whereas the RHS is given by the term
 \elpiIn{app[uva N' PF | NPF]} where the \Ho variable identified by \elpiIn{N'}
-is mapped to the \Fo variable named \elpiIn{N}.
-
-After its creation, a \linkbeta remain suspended until the head of $rhs$ is
-instantiated by the oracle (see \cref{prop:simulation}). In this case, $rhs$ is
-$\beta$-reduced to a new term, say $t'$. $t'$ is either a term in \llambda, in
-which case $t'$ is unified with the $lhs$, otherwise, the link remain suspended
+is mapped to the \Fo{} variable named \elpiIn{N}.
+
+\marginpar{signposting: a link is reconsidered in two cases. then make 2 paragraphs}
+After its creation, a \linkbeta remain suspended until the head of the RHS is
+instantiated by the oracle (see \cref{prop:simulation}). In this case the RHS is
+$\beta$-reduced to a new term $t$. $t$ is either a term in \llambda, in
+which case $t$ is unified with the LHS, otherwise, the link remain suspended
 and no progress is performed. Another way to wake a \linkbeta up is when the
-$lhs$ is a term \elpiIn{T} and $rhs$ has the shape \elpiIn{app[uva N PF | NPF]}
+LHS is a term \elpiIn{T} and RHS has the shape \elpiIn{app[uva N PF | NPF]}
 and some of the arguments in the \elpiIn{NPF} list become names. This is
@@
 
+\marginpar{A bit too fast, we first compile, then unify, then the oracle, then the manip...}
 An example justifying this last link manipulation is given by the following
@@
 being modified is the second one, which is set to \textIn{X3 x =~$\beta$~= X2
-x a}. The $rhs$ of the link has now a variable which is partially in the PF,
+x a}. The RHS of the link has now a variable which is partially in the PF,
 we can therefore remove the original \linkbeta and replace it with the following 

### 2024-03-21 paper/appendix.tex
small cleanup in comp

@@
     alloc Map1 Map2 (fv N) (hv M (arity z)) S S1.
+  comp (fo.fapp [fo.fuva N|Ag]) (ho.uva M Scope) Map1 Map2 L1 L1 S1 S2 :-
+    pattern-fragment Ag, !,
+      fold6 comp Ag Scope Map1 Map1 L1 L1 S1 S1,
+      len Ag Arity, 
+      alloc Map1 Map2 (fv N) (hv M (arity Arity)) S1 S2.
   comp (fo.fapp [fo.fuva N|Ag]) (ho.uva M' Scope) Map1 Map3 L1 L3 S1 S4 :- !,
-    split-pf Ag PF NPF,
-    forall2 (x\y\ comp x y _ _ _ _ _ _) PF PF1,
+    pattern-fragment-prefix Ag PF NPF,
+      fold6 comp PF PF1 Map1 Map1 L1 L1 S1 S1,
       fold6 comp NPF NPF1 Map1 Map2 L1 L2 S1 S2,
@@
       alloc Map2 Map3 (fv N) (hv M (arity Arity)) S2 S3,
-    if (NPF = []) 
-      (Scope = PF1, M' = M, S3 = S4, L2 = L3)
-      (get-scope (ho.app {std.append PF1 NPF1}) Scope, 
+      get-scope (ho.app {std.append PF1 NPF1}) Scope, 
       mem.new S3 M' S4,
-        L3 = [@val-link-beta (ho.uva M' Scope) (ho.app[ho.uva M PF1 | NPF1]) | L2]).
+      L3 = [@val-link-beta (ho.uva M' Scope) (ho.app[ho.uva M PF1 | NPF1]) | L2].
   comp (fo.fapp A) (ho.app A1) Map1 Map2 L1 L2 S1 S2 :- 
@@
     std.append Scope1 L1 Scope1L,
-    split-pf Scope1L Scope2 L2,
+    pattern-fragment-prefix Scope1L Scope2 L2,
     not (Scope1 = Scope2), !,

### 2024-03-21 paper/appendix.tex
renaming

@@
   move (val A)         []    A :- !.
-  move (val (uva N A)) L     (uva N X) :- std.append A L X.
+  move (val (uva N A)) L     (uva N X) :- append A L X.
 
@@
   beta (lam Bo) [H | L] R :- beta (Bo H) L R.
-  beta (app A) L (app X) :- std.append A L X.
-  beta (uva N A) L (uva N A') :- std.append A L A'.
+  beta (app A) L (app X) :- append A L X.
+  beta (uva N A) L (uva N A') :- append A L A'.
   beta (con H) L (app [con H | L]).
@@
 
-  type locale ho.tm -> o.
+  type locally-bound ho.tm -> o.
   type get-scope-aux ho.tm -> list ho.tm -> o.
@@
   get-scope-aux (ho.lam B) L1 :- 
-    pi x\ locale x => get-scope-aux (B x) L1.
+    pi x\ locally-bound x => get-scope-aux (B x) L1.
   get-scope-aux (ho.app L) L1 :- 
@@
     flatten R L1.
-  get-scope-aux X [X] :- name X, not (locale X).
-  get-scope-aux X [] :- name X, (locale X).
+  get-scope-aux X [X] :- name X, not (locally-bound X).
+  get-scope-aux X [] :- name X, (locally-bound X).
 
@@
       comp-lam F F1 Map1 Map2 L1 L2 S2 S3,
-      (pi x\ locale x => get-scope (F1 x) Scope), 
+      (pi x\ locally-bound x => get-scope (F1 x) Scope), 
       NewLink = ho.val(link-eta (ho.uva E0 Scope) (ho.lam F1)), 
@@
       alloc Map2 Map3 (fv N) (hv M (arity Arity)) S2 S3,
-      get-scope (ho.app {std.append PF1 NPF1}) Scope, 
+      get-scope (ho.app {append PF1 NPF1}) Scope, 
       mem.new S3 M' S4,
@@
     make-eta-link LenM LenM' M M' [] New H1 H2,
-    std.append New L1 L2,
+    append New L1 L2,
     deduplicate-mappings Map1 Map2 H2 H3 L2 L3.
@@
   contract-rigid L (ho.app [H|Args]) T :- 
-    std.rev L LRev, std.appendR Prefix LRev Args,
+    std.rev L LRev, append Prefix LRev Args,
     if (Prefix = []) (T = H) (T = ho.app [H|Prefix]).
@@
   progress-beta-link (ho.uva _ _ as T) (ho.app[ho.uva Ad1 Scope1 | L1] as T1) S1 S3 NewLinks :-
-    std.append Scope1 L1 Scope1L,
+    append Scope1 L1 Scope1L,
     pattern-fragment-prefix Scope1L Scope2 L2,
@@
     solve-link-abs L R S S1, !,
-    solve-links L1 L2 S1 S2, std.append R L2 L3.
+    solve-links L1 L2 S1 S2, append R L2 L3.
 \end{elpicode}
@@
     complete-mapping-under-ass H T L1 L2 F1 F2, 
-    std.append L1 L2 LAll,
+    append L1 L2 LAll,
     complete-mapping H Tl LAll L3 F2 F3.

### 2024-03-21 paper/appendix.tex
gentex

@@
 
-\begin{elpicode}
-  kind address type.
-  type addr nat -> address.
-
-  typeabbrev (mem A) (list (option A)).
-
-  type get nat -> mem A -> A -> o.
-  get z (some Y :: _) Y.
-  get (s N) (_ :: L) X :- get N L X.
-
-  type alloc-aux nat -> mem A -> mem A -> o.
-  alloc-aux z [] [none] :- !.
-  alloc-aux z L L.
-  alloc-aux (s N) [] [none | M] :- alloc-aux N [] M.
-  alloc-aux (s N) [X | L] [X | M] :- alloc-aux N L M.
-
-  type alloc address -> mem A -> mem A -> o.
-  alloc (addr A as Ad) Mem1 Mem2 :- unset? Ad Mem1, 
-    alloc-aux A Mem1 Mem2.
-
-  type new-aux mem A -> nat -> mem A -> o.
-  new-aux [] z [none].
-  new-aux [A | As] (s N) [A | Bs] :- new-aux As N Bs.
-
-  type new mem A -> address -> mem A -> o.
-  new Mem1 (addr Ad) Mem2 :- new-aux Mem1 Ad Mem2.
-
-  type set? address -> mem A -> A -> o.
-  set? (addr A) Mem Val :- get A Mem Val.
-
-  type unset? address -> mem A -> o.
-  unset? Addr Mem :- not (set? Addr Mem _).
-
-  type assign-aux nat -> mem A -> A -> mem A -> o.
-  assign-aux z (none :: L) Y (some Y :: L).
-  assign-aux (s N) (X :: L) Y (X :: L1) :- assign-aux N L Y L1.
-
-  type assign address -> mem A -> A -> mem A -> o.
-  assign (addr A) Mem1 Val Mem2 :- assign-aux A Mem1 Val Mem2.
-\end{elpicode}
-
-\section{The object language}
-
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
-fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
-flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
-fcon C ~\Eo~fcon C.
-fuva N ~\Eo~fuva N.
-flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
-  pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
-T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
-  pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
-fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
-T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
-
-type beta fm -> list fm -> fm -> o.
-beta A [] A.
-beta (flam F) [H | L] R :- subst F H B,
-  beta B L R. % since F could be x\app[x|_] and H be lam _ 
-beta (fapp A) L (fapp X) :- append A L X.
-beta (fuva N) L (fapp [fuva N | L]).
-beta (fcon H) L (fapp [fcon H | L]).
-
-type subst (fm -> fm) -> fm -> fm -> o.
-subst F H B :- napp (F H) B. % since (F H) may generate (app[app _|_])
-
-type napp fm -> fm -> o.
-napp (fcon C) (fcon C).
-napp (flam F) (flam G) :- pi x\ napp x x => napp (F x) (G x).
-napp (fapp[fapp L|M]) R :- !, append L M N, napp (fapp N) R.
-napp (fapp[X]) R :- !, napp X R.
-napp (fapp A) (fapp B) :- map napp A B.
-napp (fuva N) (fuva N).
-\end{elpicode}
+\input{code/mem}
 

### 2024-03-21 paper/code/generate_tex.py

@@ -1,0 +1,25 @@
+import os, sys
+
+def get_file_cnt(lines):
+    indexBegin = lines.index("%BEGIN\n")
+    indexEnd = lines.index("%END\n")
+    return lines[indexBegin+1:indexEnd]
+
+def print_tex(lines, fout):
+    with open(fout, "w") as f:
+        f.write("\\begin{elpicode}\n")
+        f.write("".join(lines))
+        f.write("\\end{elpicode}\n")
+
+def mk_fname(fname):
+    return fname[:-4] + "tex"
+
+def read_file(fname):
+    with open("../../code/deep/" + fname) as f:
+        lines = f.readlines()
+        print_tex(get_file_cnt(lines), mk_fname(fname))
+        
+if __name__ == "__main__":
+    fname = sys.argv[1]
+    # print(sys.argv)
+    read_file(fname)

### 2024-03-21 paper/appendix.tex
build

@@
 
-\begin{elpicode}
-  type use-binder fo.fm -> fo.fm -> o.
-  use-binder N N.
-  use-binder N (fo.fapp L) :- exists (use-binder N) L.
-  use-binder N (fo.flam B) :- pi x\ use-binder N (B x).
-
-  type maybe-eta fo.fm -> fo.fm -> list fo.fm -> o.
-  maybe-eta N (fo.fapp[fo.fuva _|Args]) _ :- !,
-    exists (x\ maybe-eta-of [] N x) Args, !. 
-  maybe-eta N (fo.flam B) L :- !, pi x\ maybe-eta N (B x) [x | L].
-  maybe-eta _ (fo.fapp [fo.fcon _|Args]) L :-
-    split-last-n {len L} Args First Last,
-    forall1 (x\ forall1 (y\ not (use-binder x y)) First) L,
-    forall2 (maybe-eta-of []) {std.rev L} Last.
-
-  type maybe-eta-of list fo.fm -> fo.fm -> fo.fm -> o.
-  maybe-eta-of _ N N :- !.
-  maybe-eta-of L N (fo.fapp[fo.fuva _|Args]) :- !, 
-    % Head is flex -> if N is in Args, we can have eta
-    forall1 (x\ exists (maybe-eta-of [] x) Args) [N|L]. 
-  maybe-eta-of L N (fo.flam B) :- !, 
-    pi x\ maybe-eta-of [x | L] N (B x).
-  maybe-eta-of L N (fo.fapp [N|Args]) :-
-    % Head is rigid -> to be an eta we consider also L
-    last-n {len L} Args R,
-    forall2 (maybe-eta-of []) R {std.rev L}.
-
-  type locally-bound ho.tm -> o.
-  type get-scope-aux ho.tm -> list ho.tm -> o.
-  get-scope-aux (ho.con _) [].
-  get-scope-aux (ho.uva _ L) L1 :- 
-    forall2 get-scope-aux L R,
-    flatten R L1.
-  get-scope-aux (ho.lam B) L1 :- 
-    pi x\ locally-bound x => get-scope-aux (B x) L1.
-  get-scope-aux (ho.app L) L1 :- 
-    forall2 get-scope-aux L R,
-    flatten R L1.
-  get-scope-aux X [X] :- name X, not (locally-bound X).
-  get-scope-aux X [] :- name X, (locally-bound X).
-
-  type get-scope ho.tm -> list ho.tm -> o.
-  get-scope T Scope :- names N,
-    get-scope-aux T ScopeDuplicata,
-    std.filter N (mem ScopeDuplicata) Scope.
-
-  type close-links (ho.tm -> links) -> links -> o.
-  close-links (_\[]) [].
-  close-links (v\[L|XS v]) [L|YS] :- !, close-links XS YS.
-  close-links (v\[(L v)|XS v]) [ho.abs L|YS] :- !, 
-    close-links XS YS.
-
-  type comp-lam (fo.fm -> fo.fm) -> (ho.tm -> ho.tm) -> 
-    mappings -> mappings -> links -> links -> ho.subst -> 
-      ho.subst -> o.
-  comp-lam F F1 Mappings1 Mappings2 L L2 S S1 :-
-    pi x y\ (pi M L S\ comp x y M M L L S S) => 
-      comp (F x) (F1 y) Mappings1 Mappings2 L (L1 y) S S1, 
-        close-links L1 L2.
-
-  type comp fo.fm -> ho.tm -> mappings -> mappings -> links -> 
-    links -> ho.subst -> ho.subst -> o.
-  comp (fo.fcon X) (ho.con X)        Map1 Map1 L1 L1 S1 S1.
-  comp (fo.flam F) (ho.uva E0 Scope) Map1 Map2 L1 L3 S1 S3 :-
-    (pi x\ maybe-eta x (F x) [x]), !,
-      mem.new S1 E0 S2,
-      comp-lam F F1 Map1 Map2 L1 L2 S2 S3,
-      (pi x\ locally-bound x => get-scope (F1 x) Scope), 
-      NewLink = ho.val(link-eta (ho.uva E0 Scope) (ho.lam F1)), 
-      L3 = [NewLink| L2].
-  comp (fo.flam F) (ho.lam F1)       Map1 Map2 L1 L2 S1 S2 :- 
-    comp-lam F F1 Map1 Map2 L1 L2 S1 S2.
-  comp (fo.fuva N) (ho.uva M []) Map1 Map2 L L S S1 :- 
-    alloc Map1 Map2 (fv N) (hv M (arity z)) S S1.
-  comp (fo.fapp [fo.fuva N|Ag]) (ho.uva M Scope) Map1 Map2 L1 L1 S1 S2 :-
-    pattern-fragment Ag, !,
-      fold6 comp Ag Scope Map1 Map1 L1 L1 S1 S1,
-      len Ag Arity, 
-      alloc Map1 Map2 (fv N) (hv M (arity Arity)) S1 S2.
-  comp (fo.fapp [fo.fuva N|Ag]) (ho.uva M' Scope) Map1 Map3 L1 L3 S1 S4 :- !,
-    pattern-fragment-prefix Ag PF NPF,
-      fold6 comp PF PF1 Map1 Map1 L1 L1 S1 S1,
-      fold6 comp NPF NPF1 Map1 Map2 L1 L2 S1 S2,
-      len PF Arity, 
-      alloc Map2 Map3 (fv N) (hv M (arity Arity)) S2 S3,
-      get-scope (ho.app {append PF1 NPF1}) Scope, 
-      mem.new S3 M' S4,
-      L3 = [@val-link-beta (ho.uva M' Scope) (ho.app[ho.uva M PF1 | NPF1]) | L2].
-  comp (fo.fapp A) (ho.app A1) Map1 Map2 L1 L2 S1 S2 :- 
-    fold6 comp A A1 Map1 Map2 L1 L2 S1 S2.
-
-  type make-eta-link-aux nat -> address -> address -> 
-    list ho.tm -> links -> ho.subst -> ho.subst -> o.
-  make-eta-link-aux z Ad1 Ad2 Scope1 L H1 H1  :-
-    std.rev Scope1 Scope, eta-expand (ho.uva Ad2 Scope) @one T1,
-    L = [@val-link-eta (ho.uva Ad1 Scope) T1].
-  make-eta-link-aux (s N) Ad1 Ad2 Scope1 L H1 H3 :-
-    std.rev Scope1 Scope, mem.new H1 Ad H2, 
-    eta-expand (ho.uva Ad Scope) @one T2,
-    (pi x\ make-eta-link-aux N Ad Ad2 [x|Scope1] (L1 x) H2 H3),
-    close-links L1 L2,
-    L = [@val-link-eta (ho.uva Ad1 Scope) T2 | L2].
-
-  type make-eta-link nat -> nat -> address -> address -> 
-          list ho.tm -> links -> ho.subst -> ho.subst -> o.
-  make-eta-link (s N) z Ad1 Ad2 Vars L H H1 :- 
-    make-eta-link-aux N Ad2 Ad1 Vars L H H1.
-  make-eta-link z (s N) Ad1 Ad2 Vars L H H1 :- 
-    make-eta-link-aux N Ad1 Ad2 Vars L H H1.
-  make-eta-link (s N) (s M) Ad1 Ad2 Vars Links H H1 :-
-    (pi x\ make-eta-link N M Ad1 Ad2 [x|Vars] (L x) H H1),
-    close-links L Links.
-
-  type deduplicate-mappings mappings -> mappings -> 
-          ho.subst -> ho.subst -> links -> links -> o.
-  deduplicate-mappings [] [] H H L L.
-  deduplicate-mappings [(mapping (fv O) (hv M (arity LenM)) as X1) | Map1] Map2 H1 H3 L1 L3 :-
-    take-list Map1 (mapping (fv O) (hv M' (arity LenM'))) _, !,
-    make-eta-link LenM LenM' M M' [] New H1 H2,
-    append New L1 L2,
-    deduplicate-mappings Map1 Map2 H2 H3 L2 L3.
-  deduplicate-mappings [A|As] [A|Bs] H1 H2 L1 L2 :- 
-    deduplicate-mappings As Bs H1 H2 L1 L2, !.
-\end{elpicode}
+\input{code/comp}
 

### 2024-03-21 paper/code/generate_tex.py

@@ -1,4 +1,4 @@
-import os, sys
+import os, sys, re
 
 def get_file_cnt(lines):
     indexBegin = lines.index("%BEGIN\n")
@@ -8,14 +8,16 @@ def get_file_cnt(lines):
 def print_tex(lines, fout):
     with open(fout, "w") as f:
         f.write("\\begin{elpicode}\n")
-        f.write("".join(lines))
+        for l in lines:
+            l = re.sub("^ *% .*\n","",l)
+            f.write(l)
         f.write("\\end{elpicode}\n")
 
 def mk_fname(fname):
-    return fname[:-4] + "tex"
+    return fname.split("/")[-1][:-4] + "tex"
 
 def read_file(fname):
-    with open("../../code/deep/" + fname) as f:
+    with open(fname) as f:
         lines = f.readlines()
         print_tex(get_file_cnt(lines), mk_fname(fname))
         

### 2024-03-21 paper/appendix.tex
usine a gaz

@@
 \section{The meta language}
-% \begin{figure*}
-\begin{elpicode}
-  typeabbrev subst list (option assignment).
 
-  kind inctx type -> type.
-  type abs (tm -> inctx A) -> inctx A.
-  type val A -> inctx A.
+\input{code/meta\_language}
 
-  typeabbrev assignment (inctx tm).
-
-  kind tm  type.
-  type app list tm -> tm.
-  type lam (tm -> tm) -> tm.
-  type con string -> tm.
-  type uva  address -> list tm -> tm.
-
-  type (==l) tm -> tm -> subst -> subst ->  o.
-  % congruence
-  ((app L1) ==l (app L2)) S S1 :- fold2 (==l) L1 L2 S S1.
-  ((lam F1) ==l (lam F2)) S S1 :- 
-    pi x\ copy x x => ((F1 x) ==l (F2 x)) S S1.
-  ((con X) ==l (con X)) S S.
-  % set variables
-  ((uva N Args) ==l T) S S1 :- 
-    mem.set? N S F,!, move F Args T1, (T1 ==l T) S S1.
-  (T ==l (uva N Args)) S S1 :- 
-    mem.set? N S F,!, move F Args T1, (T ==l T1) S S1.
-  % flex-flex
-  ((uva M A1) ==l (uva N A2)) S1 S2 :- !, 
-    pattern-fragment A1, pattern-fragment A2, 
-    prune! M A1 N A2 S1 S2.
-  ((uva N Args) ==l T) S S1 :- not_occ N S T, pattern-fragment Args, 
-    bind T Args T1, mem.assign N S T1 S1.
-  % variable assigment
-  (T ==l (uva N Args)) S S1 :- not_occ N S T, pattern-fragment Args, 
-    bind T Args T1, mem.assign N S T1 S1.
-  (N ==l N) S S :- name N.
-
-  % Note: We suppose the scopes to always be in PF
-  type prune! address -> list ho.tm -> address -> 
-              list ho.tm -> subst -> subst -> o.
-  prune! N A  N A  S S :- !.
-  prune! M A  N A  S1 S2 :- !, bind (uva M A) A Ass, 
-    mem.assign N S1 Ass S2.
-  prune! N A1 N A2 S1 S3 :- !,
-    std.assert!(len A1 {len A2}) "Not typechecking", !,
-    mem.new S1 W S2, prune-same-variable W A1 A2 [] Ass,
-    mem.assign N S2 Ass S3.
-  prune! N A1 M A2 S1 S4 :- !,
-    mem.new S1 W S2, prune-diff-variables W A1 A2 Ass1 Ass2, 
-    mem.assign N S2 Ass1 S3,
-    mem.assign M S3 Ass2 S4.
-
-  type prune-same-variable address -> list tm -> list tm -> 
-                            list tm -> assignment -> o.
-  prune-same-variable N [] [] ACC (val (uva N Args)) :- 
-    std.rev ACC Args.
-  prune-same-variable N [X|XS] [X|YS] ACC (abs F) :- 
-    pi x\ prune-same-variable N XS YS [x|ACC] (F x).
-  prune-same-variable N [_|XS] [_|YS] ACC (abs F) :- 
-    pi x\ prune-same-variable N XS YS ACC (F x).
-
-  type prune-build-ass1 address -> list tm -> 
-                        list bool -> assignment -> o.
-  prune-build-ass1 N Acc [] (val (uva N Args)) :- 
-    std.rev Acc Args.
-  prune-build-ass1 N Acc [tt|L] (abs T) :- 
-    pi x\ prune-build-ass1 N [x|Acc] L (T x).
-  prune-build-ass1 N Acc [ff|L] (abs T) :- 
-    pi x\ prune-build-ass1 N Acc L (T x).
-
-  type build-order list nat -> list tm -> list tm -> o.
-  build-order L T R :-
-    len L Len, list-init Len z 
-      (p\r\ sigma Index Elt\ index L p Index, nth Index T r) R.
-
-  type prune-build-ass2 address -> list tm -> list bool -> 
-                      list nat -> assignment -> o.
-  prune-build-ass2 N Acc [] Pos (val (uva N Args)) :- 
-    std.rev Acc Acc', build-order Pos Acc' Args.
-  prune-build-ass2 N Acc [tt|L] Pos (abs T) :- 
-    pi x\ prune-build-ass2 N [x|Acc] L Pos (T x).
-  prune-build-ass2 N Acc [ff|L] Pos (abs T) :- 
-    pi x\ prune-build-ass2 N Acc L Pos (T x).
-
-  type keep list A -> A -> bool -> o.
-  keep L A tt :- mem L A, !.
-  keep _ _ ff.
-
-  type prune-diff-variables address -> list tm -> list tm -> 
-                            assignment -> assignment -> o.
-  prune-diff-variables N Args1 Args2 Ass1 Ass2 :-
-    std.map Args1 (keep Args2) Bits1, 
-    prune-build-ass1 N [] Bits1 Ass1,
-    std.map Args2 (keep Args1) Bits2, 
-    std.filter Args1 (mem Args2) ToKeep1,
-    std.filter Args2 (mem Args1) ToKeep2,
-    std.map ToKeep2 (index ToKeep1) Pos,
-    prune-build-ass2 N [] Bits2 Pos Ass2.
-
-  type move assignment -> list tm -> tm -> o.
-  move (abs Bo)        [H|L] R :- move (Bo H) L R.
-  move (val A)         []    A :- !.
-  move (val (uva N A)) L     (uva N X) :- append A L X.
-
-  type beta tm -> list tm -> tm -> o.
-  beta A [] A.
-  beta (lam Bo) [H | L] R :- beta (Bo H) L R.
-  beta (app A) L (app X) :- append A L X.
-  beta (uva N A) L (uva N A') :- append A L A'.
-  beta (con H) L (app [con H | L]).
-
-  type not_occ_aux address -> subst -> tm -> o.
-  not_occ_aux N S (uva M _) :- mem.unset? M S, not (N = M).
-  not_occ_aux N S (uva M Args) :- mem.set? M S F, 
-    move F Args T, not_occ_aux N S T.
-  not_occ_aux N S (app L) :- forall1 (not_occ_aux N S) L.
-  not_occ_aux N S (lam F) :- pi x\ not_occ_aux N S (F x).
-  not_occ_aux _ _ (con _).
-  not_occ_aux _ _ X :- name X.
-
-  type not_occ  address -> subst -> tm -> o.
-  not_occ N _ (uva N _).
-  not_occ N S (uva M Args) :- mem.set? M S F, 
-    move F Args T, not_occ N S T.
-  not_occ N S (uva M Args) :- mem.unset? M S, 
-    std.forall Args (not_occ_aux N S).
-  not_occ _ _ (con _).
-  not_occ N S (app L) :- not_occ_aux N S (app L).
-  not_occ N S (lam L) :- pi x\ not_occ N S (L x).
-  not_occ _ _ X :- name X.
-
-  type copy tm -> tm -> o.
-  copy (app L)   (app L') :- forall2 copy L L'.
-  copy (lam T)   (lam T') :- pi x\ copy x x => copy (T x) (T' x).
-  copy (uva N L) (uva N L') :- forall2 copy L L'.
-  copy (con C)   (con C).
-  copy N N :- name N.
-
-  type bind tm -> list tm -> assignment -> o.
-  bind T [] (val T') :- copy T T'.
-  bind T [X | TL] (abs T') :- pi x\ copy X x => bind T TL (T' x).
-
-  type deref subst -> tm -> tm -> o.
-  deref S X _ :- (var S; var X), halt "flex deref".
-  deref H (uva N L)  X         :- mem.set? N H T, 
-    move T L X', !, deref H X' X.
-  deref H (app L)   (app L1)   :- forall2 (deref H) L L1.
-  deref _ (con X)   (con X).
-  deref H (uva X L) (uva X L1) :- mem.unset? X H, 
-    forall2 (deref H) L L1.
-  deref H (lam F)   (lam G)    :- pi x\ deref H (F x) (G x).
-  deref _ N         N          :- name N.
-
-  type deref-assmt subst -> assignment -> assignment -> o.
-  deref-assmt S (abs T) (abs R) :- pi x\ deref-assmt S (T x) (R x).
-  deref-assmt S (val T) (val R) :- deref S T R.
-\end{elpicode}
 \section{The compiler}

### 2024-03-21 paper/code/generate_tex.py

@@ -10,6 +10,10 @@ def print_tex(lines, fout):
         f.write("\\begin{elpicode}\n")
         for l in lines:
             l = re.sub("^ *% .*\n","",l)
+            l = re.sub("==l",r"~$\\Ue$~",l)
+            l = re.sub(".*%HIDE\n","",l)
+            l = re.sub("type \(~\$([^ ]+)\$~\) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~.",l)
+            l = re.sub("type (\([^ ]+\)) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~.",l)
             f.write(l)
         f.write("\\end{elpicode}\n")
 

### 2024-03-21 paper/code/generate_tex.py
w

@@ -9,9 +9,10 @@ def print_tex(lines, fout):
     with open(fout, "w") as f:
         f.write("\\begin{elpicode}\n")
         for l in lines:
-            l = re.sub("^ *% .*\n","",l)
+            l = re.sub("^ *% *.*\n","",l)
             l = re.sub("==l",r"~$\\Ue$~",l)
-            l = re.sub(".*%HIDE\n","",l)
+            l = re.sub("==o",r"~$\\Eo$~",l)
+            l = re.sub(".*% *HIDE *\n","",l)
             l = re.sub("type \(~\$([^ ]+)\$~\) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~.",l)
             l = re.sub("type (\([^ ]+\)) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~.",l)
             f.write(l)

### 2024-03-21 paper/elpi.py

@@ -65,6 +65,9 @@ class ElpiLexer(RegexLexer):
             (r"\b(typeabbrev)(\s+)({})".format(const_sym_re),
              bygroups(Keyword.Declaration, Text.Whitespace, Name.Function),
              'elpi-type'),
+            (r"\b(typeabbrev)(\s+)(\([^)]+\))",
+             bygroups(Keyword.Declaration, Text.Whitespace, Name.Function),
+             'elpi-type'),
             (r"\b(accumulate)(\s+)(\")",
              bygroups(Keyword.Declaration, Text.Whitespace, String.Double),
              'elpi-string'),

### 2024-03-21 paper/code/generate_tex.py
clean

@@ -12,7 +12,7 @@ def print_tex(lines, fout):
             l = re.sub("^ *% *.*\n","",l)
             l = re.sub("==l",r"~$\\Ue$~",l)
             l = re.sub("==o",r"~$\\Eo$~",l)
-            l = re.sub(".*% *HIDE *\n","",l)
+            l = re.sub(".*% *HIDE.*\n","",l)
             l = re.sub("type \(~\$([^ ]+)\$~\) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~.",l)
             l = re.sub("type (\([^ ]+\)) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~.",l)
             f.write(l)

### 2024-03-21 paper/elpi.py
c

@@ -104,7 +104,8 @@ class ElpiLexer(RegexLexer):
         ],
         'elpi-multiline-comment': [
             (r'\*/', Comment, '#pop'),
-            (r'.', Comment)
+            (r'.', Comment),
+            (r'\n', Comment)
         ],
         'elpi-indexing-expr':[
             (r'\(', Punctuation, '#push'),

### 2024-03-21 paper/main.tex
renaming
w
wip

@@
 \begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of~ \hrun, we have that
+In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
 $\forall p \in 1 \ldots \mathcal{N}$
@@
 \Ho{}.
+
+XXX permuting hrun does not change the final result if check dooes not fail eagerly
+
+XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to
+a non MGU too early
 
@@
 
-\begin{definition}[Subterm $\mathcal{P}(t)$]
+\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
+\begin{definition}[Subterm \subterm{t}]
 $$
 \begin{array}{l}
-t \in \mathcal{P}(t)\\
-t = f t_1\ldots t_n \Rightarrow \mathcal{P}(t_i) \subseteq \mathcal{P}(t) \land f \in \mathcal{P}(t)\\
-t = \lambda x.t' \Rightarrow \mathcal{P}(t') \subseteq \mathcal{P}(t)\\
+t \in \subterm{t}\\
+t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
+t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
 \end{array}
@@
 \noindent
-We write $\mathcal{P}(X) = \bigcup_{t\in X} \mathcal{P}(t)$.
+We write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$.
 
@@
 $$
-\wellb(X) \Leftrightarrow \forall t \in \mathcal{P}(X), t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
+\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
 $$
@@
 
+\subsection{Notations}
+
+we use math mode for ho.
+
 \subsection{Equational theory and Unification}
@@
 move (val A)         []    A :- !.
-move (val (uva N A)) L     (uva N X) :- std.append A L X.
+move (val (uva N A)) L     (uva N X) :- append A L X.
 \end{elpicode}

### 2024-03-21 paper/code/generate_tex.py

@@ -9,7 +9,8 @@ def print_tex(lines, fout):
     with open(fout, "w") as f:
         f.write("\\begin{elpicode}\n")
         for l in lines:
-            l = re.sub("^ *% *.*\n","",l)
+            l = re.sub("^ *% +.*\n","",l)
+            l = re.sub("^ *%%%.*\n","",l)
             l = re.sub("==l",r"~$\\Ue$~",l)
             l = re.sub("==o",r"~$\\Eo$~",l)
             l = re.sub(".*% *HIDE.*\n","",l)

### 2024-03-21 paper/main.tex
notazione preliminare
python DSL %SNIP: file, %ENDSNIP

@@
 
-we use math mode for ho.
+We use math mode for \Ho.
+
+\begin{tabular}{ll}
+  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
+  $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
+  $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
+  $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
+\end{tabular}
 
@@
 
-\begin{elpicode}
-kind link type.
-type link nat -> nat -> nat -> subst. % link Fo Ho Arity
-typeabbrev links list link.
-type comp fm -> tm -> links -> links -> subst -> subst -> o.
-comp (fcon X) (con X) L L S S.
-comp (flam F) (lam G) K L R S :- pi x y\
-  (pi A S\ comp x y L L S S) => comp (F x) (G y) K L R S.
-comp (fuva M) (uva N []) K [link M N z|K] R S :- new R N S.
-comp (fapp[fuva M|A]) (uva N B) K L R S :- distinct A, !,
-  fold4 comp A B K K R R,
-  new R N S, len A Arity,
-  L = [link N M Arity | K].
-comp (fapp A) (app B) K L R S :- fold4 comp A B K L R S.
-\end{elpicode}
+\input{code/comp_base_types}
+\input{code/comp_base}
 
@@
 \subsection{Problems with $\eta$}
+
+\input{code/comp_eta}
+
 \otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
@@
 \subsection{Problems with $\beta$}
+
+\input{code/comp_beta}
 

### 2024-03-21 paper/code/generate_tex.py

@@ -1,9 +1,13 @@
 import os, sys, re
 
 def get_file_cnt(lines):
-    indexBegin = lines.index("%BEGIN\n")
-    indexEnd = lines.index("%END\n")
-    return lines[indexBegin+1:indexEnd]
+    try:
+        indexBegin = lines.index("%BEGIN\n")
+        indexEnd = lines.index("%END\n")
+        return lines[indexBegin+1:indexEnd]
+    finally:
+        return []
+
 
 def print_tex(lines, fout):
     with open(fout, "w") as f:
@@ -22,10 +26,38 @@ def print_tex(lines, fout):
 def mk_fname(fname):
     return fname.split("/")[-1][:-4] + "tex"
 
+def get_snippets(lines):
+    snips = {}
+    ingrp = False
+    name = ""
+    curgrp = []
+    for l in lines:
+        m = re.match(r"^%ENDSNIP",l)
+        if not (m is None):
+            snips[name] = curgrp
+            ingrp = False
+            curgrp = []
+        if ingrp is True:
+            curgrp = curgrp + [l]
+        m = re.match(r"^%SNIP: *(.*) *$",l)
+        if not (m is None):
+            ingrp = True
+            name = m.group(1)
+            if name in snips:
+                curgrp = snips[name]
+            else:
+                curgrp = []
+    return snips
+
 def read_file(fname):
     with open(fname) as f:
         lines = f.readlines()
         print_tex(get_file_cnt(lines), mk_fname(fname))
+        snippets = get_snippets(lines)
+        for fname in snippets:
+            lines = snippets[fname]
+            print_tex(lines, fname + ".tex")
+
         
 if __name__ == "__main__":
     fname = sys.argv[1]

### 2024-03-21 paper/appendix.tex
w
cleanup
rename

@@
 
+Explain builtin name (can be implemented by loading name after each pi)
+
 \section{The memory}
@@
 \input{code/mem}
+
+\section{The object language}
+
+\input{code/object\_language}
 
@@
   contract-rigid L (ho.app [H|Args]) T :- 
-    std.rev L LRev, append Prefix LRev Args,
+    rev L LRev, append Prefix LRev Args,
     if (Prefix = []) (T = H) (T = ho.app [H|Prefix]).
@@
 
-  type decompl-subst mappings -> mappings -> ho.subst -> 
+  type decompl-subst map -> map -> ho.subst -> 
     fo.subst -> fo.subst -> o.
@@
 
-  type tm->fm mappings -> ho.tm -> fo.fm -> o.
+  type tm->fm map -> ho.tm -> fo.fm -> o.
   tm->fm _ (ho.con C)  (fo.fcon C).
@@
 
-  type add-new-mappings-aux ho.subst -> list ho.tm -> mappings -> 
-        mappings ->  fo.subst -> fo.subst -> o.
-  add-new-mappings-aux _ [] _ [] S S.
-  add-new-mappings-aux H [T|Ts] L L2 S S2 :- 
-    add-new-mappings H T L L1 S S1, 
-    add-new-mappings-aux H Ts L1 L2 S1 S2.
+  type add-new-map-aux ho.subst -> list ho.tm -> map -> 
+        map ->  fo.subst -> fo.subst -> o.
+  add-new-map-aux _ [] _ [] S S.
+  add-new-map-aux H [T|Ts] L L2 S S2 :- 
+    add-new-map H T L L1 S S1, 
+    add-new-map-aux H Ts L1 L2 S1 S2.
 
-  type add-new-mappings ho.subst -> ho.tm -> mappings -> 
-      mappings ->  fo.subst -> fo.subst -> o.
-  add-new-mappings _ (ho.uva N _) Map [] F1 F1 :- 
+  type add-new-map ho.subst -> ho.tm -> map -> 
+      map ->  fo.subst -> fo.subst -> o.
+  add-new-map _ (ho.uva N _) Map [] F1 F1 :- 
     mem Map (mapping _ (hv N _)), !.
-  add-new-mappings H (ho.uva N L) Map [Map1 | MapL] F1 F3 :-
+  add-new-map H (ho.uva N L) Map [Map1 | MapL] F1 F3 :-
     mem.new F1 M F2,
     len L Arity, Map1 = mapping (fv M) (hv N (arity Arity)),
-    add-new-mappings H (ho.app L) [Map1 | Map] MapL F2 F3.
-  add-new-mappings H (ho.lam B) Map NewMap F1 F2 :- 
-    pi x\ add-new-mappings H (B x) Map NewMap F1 F2.
-  add-new-mappings H (ho.app L) Map NewMap F1 F3 :- 
-    add-new-mappings-aux H L Map NewMap F1 F3.
-  add-new-mappings _ (ho.con _) _ [] F F :- !.
-  add-new-mappings _ N _ [] F F :- name N.
+    add-new-map H (ho.app L) [Map1 | Map] MapL F2 F3.
+  add-new-map H (ho.lam B) Map NewMap F1 F2 :- 
+    pi x\ add-new-map H (B x) Map NewMap F1 F2.
+  add-new-map H (ho.app L) Map NewMap F1 F3 :- 
+    add-new-map-aux H L Map NewMap F1 F3.
+  add-new-map _ (ho.con _) _ [] F F :- !.
+  add-new-map _ N _ [] F F :- name N.
 
   type complete-mapping-under-ass ho.subst -> ho.assignment -> 
-    mappings -> mappings ->  fo.subst -> fo.subst -> o.
+    map -> map ->  fo.subst -> fo.subst -> o.
   complete-mapping-under-ass H (ho.val Val) Map1 Map2 F1 F2 :- 
-    add-new-mappings H Val Map1 Map2 F1 F2.
+    add-new-map H Val Map1 Map2 F1 F2.
   complete-mapping-under-ass H (ho.abs Abs) Map1 Map2 F1 F2 :- 
@@
   type complete-mapping ho.subst -> ho.subst ->   
-    mappings -> mappings -> fo.subst -> fo.subst -> o.
+    map -> map -> fo.subst -> fo.subst -> o.
   complete-mapping _ [] L L F F.
@@
 
-  type decompile mappings -> links -> ho.subst -> 
+  type decompile map -> links -> ho.subst -> 
     fo.subst -> fo.subst -> o.

### 2024-03-22 paper/code/generate_tex.py
update python

@@ -1,21 +1,24 @@
 import os, sys, re
 
 def get_file_cnt(lines):
+    res = []
     try:
         indexBegin = lines.index("%BEGIN\n")
         indexEnd = lines.index("%END\n")
-        return lines[indexBegin+1:indexEnd]
+        res = lines[indexBegin+1:indexEnd]
     finally:
-        return []
+        return res
 
 
 def print_tex(lines, fout):
     with open(fout, "w") as f:
         f.write("\\begin{elpicode}\n")
         for l in lines:
-            l = re.sub("^ *% +.*\n","",l)
-            l = re.sub("^ *%%%.*\n","",l)
-            l = re.sub("==l",r"~$\\Ue$~",l)
+            l = re.sub("^ *% +.*\n","",l)   
+            l = re.sub("^ *%SNIP.*\n","",l)   
+            l = re.sub("^ *%ENDSNIP.*\n","",l)   
+            l = re.sub("^ *%%%.*\n","",l)   
+            l = re.sub("==l",r"~$\\Ue$~",l) 
             l = re.sub("==o",r"~$\\Eo$~",l)
             l = re.sub(".*% *HIDE.*\n","",l)
             l = re.sub("type \(~\$([^ ]+)\$~\) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~.",l)

### 2024-03-22 paper/main.tex
rename
cleanup
import all declared snippets
update beta

@@
 
-\newcommand{\linketaM}[3]{\texttt{#1 |- #2 =$\eta$= #3}}
-\newcommand{\linkbetaM}[3]{\texttt{#1 |- #2 =$\beta$= #3}}
+\newcommand{\linketaM}[3]{\ensuremath{#1 \vdash #2 =_\eta #3}}
+\newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_\beta #3}}
+\newcommand{\substCell}[3]{\ensuremath{#1 \vdash #2 = #3}}
+\newcommand{\mapping}[3]{\ensuremath{#1 \mapsto #2^#3}}
 
@@
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
+\def\linkStore{\ensuremath{\mathbb{L}}}
+
 $$
@@
 \begin{array}{l}
-\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\vspace{2pt}\\
-  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\vspace{2pt}\\
+\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''} \deff\vspace{2pt}\\
+  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
 \runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
-  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\vspace{2pt}\\
-  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
+  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
+  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore}{\rho_{\mathcal{N}}}
 \end{array}
@@
 \Leftrightarrow
-\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
+\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}
 $$
@@
 \begin{proposition}[\wellb{}-preservation]\label{prop:nf}
-$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
+$\forall \mathcal{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
 $$
 \wellb(\sigma\mathcal{T}) \land
-\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
+\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma'}
 \Rightarrow \wellb(\sigma' \mathcal{T})
@@
 decompilation can introduce (actually restore) terms in
-\maybeeta or \maybebeta that were move out of the way (put in $\mathbb{L}$)
+\maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
 by compilation.
@@
 
+types
 \input{code/comp_base_types}
+
+core
+
 \input{code/comp_base}
+
+aux
+
+\input{code/comp_base2}
 
@@
 \input{code/comp_eta}
+
+and aux
+
+\input{code/maybe_eta}
 
@@
 
-\input{code/comp_beta}
-
 $\beta$-reduction problems (\maybebeta) appears any time we deal with a subterm $t
@@
 
+\def\rhs{\ensuremath{rhs}\xspace}
+\def\lhs{\ensuremath{lhs}\xspace}
+
 This link carries two terms, the former representing the variable $h$ for the
 new created hole and the latter containing the subterm $t$. As for the \linketa,
-we will call $h$ and $t$ respectively the left hand side (LHS)
-and the right hand side (RHS) of the \linkbeta.
-
-A subterm is in \maybebeta{} if it has the shape \elpiIn{fapp[fuva
-N | L]} and \elpiIn{distinct L} does not hold. In that case, \elpiIn{L} is split
-in two sublist \elpiIn{PF} and \elpiIn{NPF} such that former is the longest
-prefix of \elpiIn{L} such that \elpiIn{distinct PF} holds. \elpiIn{NPF} is the
-list such that \elpiIn{append PF NPF L}. The LHS is set to a new variable
-named $M$ with \elpiIn{PF} in scope whereas the RHS is given by the term
-\elpiIn{app[uva N' PF | NPF]} where the \Ho variable identified by \elpiIn{N'}
-is mapped to the \Fo{} variable named \elpiIn{N}.
-
-\marginpar{signposting: a link is reconsidered in two cases. then make 2 paragraphs}
-After its creation, a \linkbeta remain suspended until the head of the RHS is
-instantiated by the oracle (see \cref{prop:simulation}). In this case the RHS is
-$\beta$-reduced to a new term $t$. $t$ is either a term in \llambda, in
-which case $t$ is unified with the LHS, otherwise, the link remain suspended
-and no progress is performed. Another way to wake a \linkbeta up is when the
-LHS is a term \elpiIn{T} and RHS has the shape \elpiIn{app[uva N PF | NPF]}
+we will call $h$ and $t$ respectively the left hand side (\lhs)
+and the right hand side (\rhs) of the \linkbeta.
+
+In order to build a \linkbeta, we need to adapt the compiler so that it can
+recognize these ``problematic'' subterms. The following code snippet illustrate
+such behavior, we suppose the rule to be added just after \cref{rule:comp-pf}.
+
+\input{code/comp_beta}
+
+A term is \maybebeta if it has the shape \elpiIn{fapp[fuva A|Ag]} and
+\elpiIn{distinct Ag} does not hold. In that case, \elpiIn{Ag} is split in two
+sublist \elpiIn{Pf} and \elpiIn{Extra} such that former is the longest prefix of
+\elpiIn{Ag} such that \elpiIn{distinct Pf} holds. \elpiIn{Extra} is the list
+such that \elpiIn{append Pf Extra Ag}. Next important step is to compile
+recursively the terms of these lists and allocate a memory adress \elpiIn{B}
+from the substitution in order to map the \Fo variable \elpiIn{fuva A} to
+the \Ho variable \elpiIn{uva B}. The \linkbeta to return in the end is given 
+by the term \elpiIn{Beta = app[uva B Scope1 | Extra1]} constituting the \rhs,
+and a fresh variable \elpiIn{C} having in scope all the free variables occurring
+in \elpiIn{Beta} (this is \lhs). We point out that the \rhs is intentionally
+built as an \elpiIn{uva} where \elpiIn{Extra1} are not in scope, since by 
+invariant, we want all the variables appearing in \Ho to be in \llambda.
+
+One created, there exist two main situations waking up a suspended \linkbeta.
+The former is strictly connected to the definition of $\beta$-redex and occurs
+when the head of \rhs is materialized by the oracle (see
+\cref{prop:simulation}). In this case \rhs is safely\marginpar{explain why}
+$\beta$-reduced to a new term $t'$ and the result can be unified with \lhs. In
+this scenario the \linkbeta has accomplished its goal and can be removed from
+\linkStore.
+
+The second circumstance making the \linkbeta to progress is the instantiation 
+of the variables in the \elpiIn{Extra}.
+
 and some of the arguments in the \elpiIn{NPF} list become names. This is
@@
 
-\begin{textcode}
-X ~$\mapsto$~ X1; F ~$\mapsto$~ X2 % The mappings
-  ~$\vdash$~ X0   =~$\eta$~= x\ `X3 x'
-x ~$\vdash$~ X3 x =~$\beta$~= X2 `X1 x' a
-\end{textcode}
+\begin{gather}
+  \mapping{X}{X1}{1}; \mapping{F}{X2}{0} (\mathrm{The map})\\
+  \linketaM{}{X0}{\lambda x.X3_x}\\
+  \linkbetaM{x}{X3_x}{X2 ~ X1_x ~ a}
+\end{gather}
 
@@
 \texttt{X3}; and a \linkbeta between the variable \texttt{X3} and the subterm
-\texttt{c0\textbackslash\ X2 `X1 c0' a} (it is a \maybebeta).
-The substitution tells that \textIn{x ~$\vdash$~ X1 x = x}.
+$\lambda x.X1_x ~ a$ (it is a \maybebeta).
+The substitution tells that \substCell{x}{X1_x}{x}.
 
@@
 first, dereference all the links, and then try to solve them. The only link
-being modified is the second one, which is set to \textIn{X3 x =~$\beta$~= X2
-x a}. The RHS of the link has now a variable which is partially in the PF,
-we can therefore remove the original \linkbeta and replace it with the following 
-couple on links:
+being modified is the second one, which is set to \linkbetaM{x}{X3}{X2 x a}. The
+\rhs of the link has now a variable which is partially in the PF, we can
+therefore remove the original \linkbeta and replace it with the following couple
+on links:
 

### 2024-03-22 paper/main.tex
update beta

@@
 
-The second circumstance making the \linkbeta to progress is the instantiation 
-of the variables in the \elpiIn{Extra}.
-
-and some of the arguments in the \elpiIn{NPF} list become names. This is
-possible after the resolution of other links. In this case, the list \elpiIn{L}
-obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
-again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
-same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
-new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
-create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.
+The second circumstance making the \linkbeta to progress is the instantiation of
+the variables in the \elpiIn{Extra1} making the corresponding arguments to
+reduce to names. In this case, we want to take the list \elpiIn{Scope1} and
+append to it the largest prefix of \elpiIn{Extra1} in a new variable
+\elpiIn{Scope2} such that \elpiIn{Scope2} remains in \llambda; we call
+\elpiIn{Extra2} the suffix of \elpiIn{Extra1} such that the concatenation of
+\elpiIn{Scope1} and \elpiIn{Extra1} is the same as the concatenation of
+\elpiIn{Scope2} and \elpiIn{Extra2}.
+
+% and some of the arguments in the \elpiIn{NPF} list become names. This is
+% possible after the resolution of other links. In this case, the list \elpiIn{L}
+% obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
+% again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
+% same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
+% new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
+% create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.
 

### 2024-03-22 paper/main.tex
wip

@@
 
-\section[Compilation: fo\_tm to tm]{Basic compilation \Fo{} to \Ho{}}
+\section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
 \label{sec:compilation}
@@
 In this section we describe a basic compilation scheme that we refine
-later, in the following sections.
+later, in the following sections. This scheme is sufficient to implement
+an \Uo{} that respects $\beta$-conversion for terms in \llambda. The extension to
+$\eta\beta$-conversion is described in Section \ref{sec:eta} and the support
+for terms outside \llambda in Section \ref{sec:beta}.
 
@@
 available in \Ho{} when expressed in a first order way in \Fo{}. The compiler
-also generates a list of links that are used to bring back the substitution from
-\Ho{} to \Fo{} and allocates in the memory a cell for each variable.
-
-types
+also generates a map to bring back the substitution from
+\Ho{} to \Fo{}.
+
+\newtheorem{invariant}{Invariant}
+
 \input{code/comp_base_types}
 
-core
+\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
+in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
+\elpiIn{(uvar A L)} is such that \elpiIn{(len L N)} holds
+\end{invariant}
+
+\noindent The arity of a variable in \Ho (a \elpiIn{hvariable} is stored in
+the mapping. In particular \elpiIn{m-alloc}
+bla bla explain. Multiple mappings for the same \elpiIn{fvariable}
+are handled in section \ref{sec:eta}.
+
+\input{code/alloc_mapping}
+
+The signature of the \elpiIn{comp} predicate below allows for the generation of
+links (suspended unification problems) that play no role in this section
+but play a major role in Sections \ref{sec:eta} and \ref{sec:beta}.
 
@@
 
-aux
+This preliminary version of \elpiIn{comp} simply recognizes \Fo variables
+applied to a (possibly empty) duplicate free list of names (i.e.
+\elpiIn{pattern-fragment} detects variables in \llambda).
+
+The auxiliary function \elpiIn{close-liks}
 
@@
 
-\subsection{Problems with $\eta$}
+\subsection{Problems with $\eta$}\label{sec:eta}
 
@@
 
-\subsection{Problems with $\beta$}
+\subsection{Problems with $\beta$}\label{sec:beta}
 

### 2024-03-22 paper/main.tex
address -> addr + cleanup

@@
   \begin{minipage}{0.21\textwidth}
-  \begin{elpicodetab}
-kind fm type.
-type fapp list fm -> fm.
-type flam (fm -> fm) -> fm.
-type fcon string -> fm.
-type fuva nat -> fm.
-\end{elpicodetab}
+   \inputrawelpicode{code/fo_tm}
   \end{minipage}
@@
   \begin{minipage}{0.24\textwidth}
-  \begin{elpicodetab}
-kind tm type.
-type app list tm -> tm.
-type lam (tm -> tm) -> tm.
-type con string -> tm.
-type uva nat -> list tm -> tm.
-  \end{elpicodetab}
+   \inputrawelpicode{code/ho_tm}
   \end{minipage}
-  \end{tabular}
+  \end{tabular}\vspace{4pt}
   \caption{\Fo{} and \Ho language}\vspace{0.3em}
@@
 
-\begin{elpicode}
-typeabbrev memory A (list (option A)).
-type set?   nat -> memory A -> A -> o.
-type unset? nat -> memory A -> o.
-type assign nat -> memory A -> A -> memory A -> o.
-\end{elpicode}
+\input{code/mem_types}
 
@@
 
-\begin{elpicode}
-  typeabbrev fsubst (memory ftm).
-  kind assmt type.
-  type abs (tm -> assmt) -> assmt.
-  type val tm -> assmt.
-  typeabbrev subst (memory assmt).
-\end{elpicode}
+\input{code/fo_subst}
+\input{code/ho_subst}
 
@@
 
+\subsection{Compilation}
+
 The objective of the compilation is to recognize the higher-order variables
@@
 
-The auxiliary function \elpiIn{close-liks}
+The auxiliary function \elpiIn{close-links} cannot be explained here,
+che palle. \marginpar{fix}
 
@@
 
-\noindent
-Note that \elpiIn{link} carries the arity (number of expected arguments) of
-the variable. \marginpar{say when this is needed}
-
-\begin{elpicode}
-type solve-links links -> links -> subst -> subst -> o.
-solve-links L L S S.
-\end{elpicode}
-
-Then decomp
-
-\begin{elpicode}
-type decompile links -> subst -> fsubst -> o.
-decompile L S O :-
-  map (_\r\r = none) S O1, % allocate empty fsubst
-  (pi N X\ knil N X :- mem L (link X N _) ; N = X) =>
-    decompl S L O1 O.
-type knil nat -> nat -> o.
-
-type decompl links -> subst -> fsubst -> o.
-decompl S [] [].
-decompl S [link _ N _|L] O P :- unset? N S X,
-  decompl S L O P.
-decompl S [link M N _|L] O P :- set? N S X,
-  decomp-assignment S X T, assign M O (some T) O1,
-  decompl S L O1 P.
-
-type decomp-assignment subst -> assignment -> fm -> o.
-decomp-assignment S (abs F) (flam G) :-
-  pi x y\ decomp-tm S x y => decomp-assignment S (F x) (G y).
-decomp-assignment S (val T) T1 :- decomp S T T1.
-
-type decomp subst -> tm -> fm.
-decomp _ (con C) (fcon C).
-decomp S (app A) (app B) :- map (decomp S) A B.
-decomp S (lam F) (flam G) :-
-  pi x y\ decomp S x y => decomp S (F x) (G y).
-decomp S (uva N A) R :- set? N S F,
-  move F A T, decomp S T R.
-decomp S (uva N A) R :- unset? N S,
-  map (decomp S) A B, knil N M, napp (fapp[fuva M|B]) R.
-\end{elpicode}
-
-\marginpar{TODO link}
-\marginpar{TODO nuove subst}
-
-Now unif\marginpar{TODO: code unif}
-
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Uo)} \PYG{k+kt}{fm -> fm -> subst -> subst -> o}~.
-(X ~\Uo~Y) S S1 :-
-  fderef S X X0, fderef S Y Y0,                    ~($\customlabel{step:norm}{norm}$)~
-  comp X0 X1 [] S0 [] L0,                        ~($\customlabel{step:compile}{compile}$)~
-  comp Y0 Y1 S0 S1 L0 L1,
-  (X1 ~\Ue~Y1) [] HS0,                              ~($\customlabel{step:unify}{unify}$)~
-  solve-links L1 L2 HS0 HS1,                        ~($\customlabel{step:link}{link}$)~
-  decompile L2 HS1  S1.                        ~($\customlabel{step:decompile}{decompile}$)~
-\end{elpicode}
-
-\subsection{Prolog simulation}
-Allows us to express the properties.
-we take all terms involved in a search (if a rule is used twice we simply
-take a copy of it), we compile all of them, and then we pick the unification
-prblems among these terms and step trough them.
-
-\begin{elpicode}
-type pick list A -> (pair nat nat) -> (pair A A) -> o.
-pick L (pr X Y) (pr TX TY) :- nth X L TX, nth Y L TY.
-
-type prolog-fo list fm -> list (pair nat nat) -> subst -> o.
-prolog-fo Terms Problems S :-
-  map (pick Terms) Problems FoProblems,
-  fold4 ~(\Uo)~ FoProblems [] S.
-
-type step-ho (pair tm tm) -> links -> links -> subst -> subst -> o.
-step-ho (pr X Y) L0 L1 S0 S2 :-
-  (X1 ~\Ue~Y1) S0 S1,
-  solve-links L0 L1 S1 S2.
-
-type prolog-ho list fm -> list (pair nat nat) -> subst -> o.
-prolog-ho Terms Problems S :-
-  fold4 comp Terms HoTerms [] L0 [] HS0,
-  map (pick HoTerms) Problems HoProblems,
-  fold4 step-ho HoProblems L0 L HS0 HS,
-  decompile L HS S.
-\end{elpicode}
-
-\noindent
-the proprty is that if a step for Fo succeds then the Ho one does, and
-if Fo fails then the Ho fails ()
+\subsection{Execution}
+
+\subsection{Decompilation}
 

### 2024-03-22 paper/code/generate_tex.py

@@ -10,9 +10,10 @@ def get_file_cnt(lines):
         return res
 
 
-def print_tex(lines, fout):
+def print_tex(lines, fout, raw = False):
     with open(fout, "w") as f:
-        f.write("\\begin{elpicode}\n")
+        if not raw:
+            f.write("\\begin{elpicode}\n")
         for l in lines:
             l = re.sub("^ *% +.*\n","",l)   
             l = re.sub("^ *%SNIP.*\n","",l)   
@@ -24,7 +25,8 @@ def print_tex(lines, fout):
             l = re.sub("type \(~\$([^ ]+)\$~\) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~.",l)
             l = re.sub("type (\([^ ]+\)) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~.",l)
             f.write(l)
-        f.write("\\end{elpicode}\n")
+        if not raw:
+            f.write("\\end{elpicode}\n")
 
 def mk_fname(fname):
     return fname.split("/")[-1][:-4] + "tex"
@@ -60,6 +62,7 @@ def read_file(fname):
         for fname in snippets:
             lines = snippets[fname]
             print_tex(lines, fname + ".tex")
+            print_tex(lines, fname + "_raw.tex", True)
 
         
 if __name__ == "__main__":

### 2024-03-22 paper/main.tex
less duplicate code

@@
 
+\input{code/comp_base_types}
+
+\newtheorem{invariant}{Invariant}
+
+\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
+in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
+\elpiIn{(uvar A L)} is such that \elpiIn{(len L N)} holds
+\end{invariant}
+
+\noindent The arity of a variable in \Ho (a \elpiIn{hvariable} is stored in
+the mapping. In particular \elpiIn{m-alloc}
+bla bla explain. Multiple mappings for the same \elpiIn{fvariable}
+are handled in section \ref{sec:eta}.
+
+\input{code/comp_links}
+
 \subsection{Notations}
@@
 
-\begin{elpicode}
-type fder fsubst -> fm -> fm -> o.
-fder S (fapp A) (fapp B) :- map (fder S) A B.
-fder S (flam F) (flam G) :- 
-  pi x\ fder S x x => fder S (F x) (G x).
-fder _ (fcon C) (fcon C).
-fder S (fuva N) R :- set? N S T, fder S T R.
-fder S (fuva N) (fuva N) :- unset? N S.
-
-type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fder}{\rho s}$)~
-fderef S T R :- fder S T T', napp T' R.
-
-type napp fm -> fm -> o.
-\end{elpicode}
+\input{code/fderef}
 
@@
 
-\begin{elpicode}
-type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
-deref S (app A) (app B) :- map (deref S) A B.
-deref S (lam F) (lam G) :- 
-  pi x\ deref S x x => deref S (F x) (G x).
-deref _ (con C) (con C).
-deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
-deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
-type move assignment -> list tm -> tm -> o.
-move (abs Bo)        [H|L] R :- move (Bo H) L R.
-move (val A)         []    A :- !.
-move (val (uva N A)) L     (uva N X) :- append A L X.
-\end{elpicode}
+\input{code/deref}
 
@@
 list, we will explain why in \cref{sec:compilation}
-
-\begin{elpicode}
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
-  fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
-  flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
-  fcon C ~\Eo~fcon C.
-  fuva N ~\Eo~fuva N.
-  flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
-    pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
-  T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
-    pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
-  fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
-  T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
-\end{elpicode}
-
-\begin{figure}
-  \begin{elpicode}
-    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-    app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
-    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
-    con C ~\Ee~fcon C.
-    uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
-  \end{elpicode}
-  \caption{Equal predicate ML}
-  \label{code:eq-pred-meta}
-  \Description[code:equal-pred-meta]{code:equal-pred-meta}
-\end{figure}
 
@@
 rules.
+
+\input{code/feq}
+
+\begin{elpicode}
+  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
+  app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
+  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
+  con C ~\Ee~fcon C.
+  uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
+\end{elpicode}
 
@@
 
-\begin{elpicode}
-~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
-\end{elpicode}
+\input{code/ue_type}
 
@@
 $\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
-\subseteq \rho_2$. The first three rules unify terms with same rigid heads, and
-call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
-assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
-unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
-terms are unification variables, we test that their arguments are in the pattern
-fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
-pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
-$w$ and return the new mapping $\rho_2$ containing all the new variable
-assignment. Finally, if only one of the two terms is an unification variable
-$v$, after having verified that $v$ does not occur in the other term $t$, we
-bind $v$ to $t$ and return the new substitution mapping.
-
-\old
-
-A key property needed in unification is being able to verify if two terms are
-equal wrt a given equational theory. This relation allow to compare terms under
-a certain substitution mapping, so that any time a variable $v$ is assigned in a
-subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
-test for equality is continued on the new-created subterm.
-
-The base equality function over terms can be defined as follows:
-
-The solution we are proposing aim to overcome these unification issues by 1)
-compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
-in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
-variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
-the variables of the OL with the decompiled version of their corresponding meta
-variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
-and that the substitution in the object language is the same as the one returned
-by the ML. \todo{same or $\supseteq$ or $\subseteq$}
-
-In the following section we explain how we deal with term (de)compilation and
-links between unification variables.
+\subseteq \rho_2$.
+% The first three rules unify terms with same rigid heads, and
+% call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
+% assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
+% unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
+% terms are unification variables, we test that their arguments are in the pattern
+% fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
+% pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
+% $w$ and return the new mapping $\rho_2$ containing all the new variable
+% assignment. Finally, if only one of the two terms is an unification variable
+% $v$, after having verified that $v$ does not occur in the other term $t$, we
+% bind $v$ to $t$ and return the new substitution mapping.
+
+% \old
+
+% A key property needed in unification is being able to verify if two terms are
+% equal wrt a given equational theory. This relation allow to compare terms under
+% a certain substitution mapping, so that any time a variable $v$ is assigned in a
+% subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
+% test for equality is continued on the new-created subterm.
+
+% The base equality function over terms can be defined as follows:
+
+% The solution we are proposing aim to overcome these unification issues by 1)
+% compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
+% in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
+% variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
+% the variables of the OL with the decompiled version of their corresponding meta
+% variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
+% and that the substitution in the object language is the same as the one returned
+% by the ML. \todo{same or $\supseteq$ or $\subseteq$}
+
+% In the following section we explain how we deal with term (de)compilation and
+% links between unification variables.
 
@@
 \Ho{} to \Fo{}.
-
-\newtheorem{invariant}{Invariant}
-
-\input{code/comp_base_types}
-
-\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
-in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
-\elpiIn{(uvar A L)} is such that \elpiIn{(len L N)} holds
-\end{invariant}
-
-\noindent The arity of a variable in \Ho (a \elpiIn{hvariable} is stored in
-the mapping. In particular \elpiIn{m-alloc}
-bla bla explain. Multiple mappings for the same \elpiIn{fvariable}
-are handled in section \ref{sec:eta}.
 

### 2024-03-22 paper/code/generate_tex.py

@@ -16,14 +16,15 @@ def print_tex(lines, fout, raw = False):
             f.write("\\begin{elpicode}\n")
         for l in lines:
             l = re.sub("^ *% +.*\n","",l)   
+            l = re.sub("%~(.*)",r"~\g<1>",l)   
             l = re.sub("^ *%SNIP.*\n","",l)   
             l = re.sub("^ *%ENDSNIP.*\n","",l)   
             l = re.sub("^ *%%%.*\n","",l)   
             l = re.sub("==l",r"~$\\Ue$~",l) 
             l = re.sub("==o",r"~$\\Eo$~",l)
             l = re.sub(".*% *HIDE.*\n","",l)
-            l = re.sub("type \(~\$([^ ]+)\$~\) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~.",l)
-            l = re.sub("type (\([^ ]+\)) (.*)\.$",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~.",l)
+            l = re.sub("type \(~\$([^ ]+)\$~\) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~",l)
+            l = re.sub("type (\([^ ]+\)) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~",l)
             f.write(l)
         if not raw:
             f.write("\\end{elpicode}\n")

### 2024-03-22 paper/main.tex
wip

@@
 applied to a (possibly empty) duplicate free list of names (i.e.
-\elpiIn{pattern-fragment} detects variables in \llambda).
-
-The auxiliary function \elpiIn{close-links} cannot be explained here,
-che palle. \marginpar{fix}
+\elpiIn{pattern-fragment} detects variables in \llambda). Note tha compiling
+\elpiIn{Ag} cannot create new mappings nor links, see the comp-lam hyp rule.
+
+The auxiliary function \elpiIn{close-links}
 
 \input{code/comp_base2}
+
+since we want links to bubble up we use the abs constructor of the inctx data
+type to bind back the variable just crossed, and we do so only if the variable
+v occurs in L.
 

### 2024-03-22 paper/main.tex
notatations

@@
 
-We use math mode for \Ho.
+When we write \Ho terms outside code blocks we prefer to use the
+usual and more compact mathematical notation. However we need to
+distinguish between the ``application'' of a unification variable
+to its scope and the application of a term (possibly a unification
+variable) to a list of arguments. We write the scope in subscript
+and we use juxtaposition for application. Here a few examples:
+\vspace{5pt}
 
@@
 \end{tabular}
+\vspace{5pt}
+
+\noindent
+When detailing example we prefer to write links in mathematical notation,
+as equations between terms in a context. The equality sign is subscripted with
+kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
+\begin{elpicode}
+abs x\ val (link-beta (uva A []) (app[uva F [x],con "a"]))
+\end{elpicode}
 

### 2024-03-23 paper/main.tex
url
w

@@
 
-\author{Enrico Tassi}
-\email{enrico.tassi@inria.fr}
+\author{Davide Fissore}
+\email{davide.fissore@inria.fr}
 \affiliation{%
@@
 
-\author{Davide Fissore}
-\email{davide.fissore@inria.fr}
+\author{Enrico Tassi}
+\email{enrico.tassi@inria.fr}
 \affiliation{%
@@
 bla bla.. TODO citare Teyjus.
+The code discussed in the paper can be accessed at the URL:
+\url{https://github.com/FissoreD/paper-ho}.
 
@@
 \label{sec:nutshell}
-A term $s$ is compiled in a term $t$ where any
-``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
+A term $s$ is compiled in a term $t$ where every
+``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
 and an accessory link that represent a suspended unification problem
-$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, meaning it does not
-contradict \Eo (as it would do on ``problematic'' terms).
+$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
+contradict \Eo as it would otherwise do on ``problematic'' terms.
 We now define ``problematic'' and ``well behaved'' more formally.
@@
 makes $\rho t = \lambda x.\lambda y.f x y$
-that is the eta long form of $f$.
+that is the eta long form of $f$. This term is problematic since
+its rigid part, the $\lambda$-abstractions, cannot justify a
+unification failure against, say, a constant.
 
@@
 \newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
-\begin{definition}[Subterm \subterm{t}]
+\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
+  largest set $\mathcal{P(t)}$ that can be obtained by the following rules.
 $$
@@
 \noindent
-We write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$.
+We write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.
 
@@
 \maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
-by compilation.
+during compilation.
 
@@
 \noindent
-In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
-the arguments of an higher order unification variable are via the \elpiIn{fapp}
-constructor. For example in the statement of the instance \coqIn{forall_dec} the
-term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is
-a memory address and \elpiIn{x} is a bound variable.
-
-In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]}.
-We say that the unification variable \elpiIn{uva N L} is in \llambda iff
-\elpiIn{distinct L} holds.
-
-\begin{elpicode}
-type distinct list A -> o.
-distinct [].
-distinct [X|XS] :- name X, not(mem X XS), distinct XS.
-\end{elpicode}
+Unification variables (\elpiIn{fuva} term constructor)
+in \Fo{} have no explicit scope:
+the arguments of an higher order variable are given via the \elpiIn{fapp}
+constructor. For example the term \coqIn{P x} is represented as
+\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
+\elpiIn{x} is a bound variable.\\
+In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
+since unification variables come equipped with an explicit scope.
+We say that the unification variable occurrence \elpiIn{uva N L} is in
+\llambda if and only if \elpiIn{L} is made of distinct names. The
+predicate to test this condition is called \elpiIn{pattern-fragment}:
+
+\input{code/pattern_fragment}
 
@@
 \footnote{one could always load name x for every x under a pi and get rid of the name builtin}
-The
-compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
-reasons, so we don't assume all out terms are in \llambda but rather test. \textbf{\textcolor{red}{what?? }}
-
-In both languages unification variables are identified by a natural number,
-which can be seen as a memory address. The memory and its associated
-operations are described below:
+
+In both languages unification variables are identified by a natural number
+representing a memory address.
+The memory and its associated operations are described below:
 
@@
 If a memory cell is \elpiIn{none}, then the corresponding unification variable
-is not set. \elpiIn{assign} sets an unset cell to the given value.
+is not set. \elpiIn{assign} sets an unset cell to the given value, while
+\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\marginpar{is new used?}
 
 Since in \Ho unification variables have a scope, their solution needs to be
-abstracted over it in order to enable the instantiation of a single
-solution to different scopes. On the contrary solutions to \Fo variables
-are plain terms.
+abstracted over it to enable the instantiation of a single
+solution to different scopes.
+On the contrary a solution to a \Fo variable is a plain term.
 
@@
 in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
-\elpiIn{(uvar A L)} is such that \elpiIn{(len L N)} holds
+\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
 \end{invariant}

### 2024-03-23 paper/main.tex
section4

@@
 \section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
-\label{sec:encodings}
+\label{sec:problem-statement}
 
@@
   \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
-  \unifcomplete{\Ee}{\Ue}{t}
+  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
 \end{gather}
@@
   \end{tabular}\vspace{4pt}
-  \caption{\Fo{} and \Ho language}\vspace{0.3em}
+  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
   \label{code:common-terms}
@@
 abstracted over it to enable the instantiation of a single
-solution to different scopes.
+solution to different scopes. This is obtained via the \elpiIn{inctx}
+container, and in particular via its \elpiIn{abs} binding constructor. 
 On the contrary a solution to a \Fo variable is a plain term.
@@
 
-\begin{invariant}[Unification variable arity] Each variable \elpiIn{A}
+\begin{invariant}[Unification variable arity]\label{inv:uvaarity}
+Each variable \elpiIn{A}
 in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
@@
 
-\noindent The arity of a variable in \Ho (a \elpiIn{hvariable} is stored in
-the mapping. In particular \elpiIn{m-alloc}
-bla bla explain. Multiple mappings for the same \elpiIn{fvariable}
-are handled in section \ref{sec:eta}.
+\noindent
+The compiler establishes a mapping between variables of the two languages.
+In order to preserve invariant \ref{inv:uvaarity} we store the
+arity of each \elpiIn{hvariable} in the mapping and we reuse an existing
+mapping only if the arity matches.
+
+\input{code/alloc_mapping}
+
+\noindent
+When a single \elpiIn{fvariable} occurs multiple times with different numbers
+of arguments the compiler generates multiple mappings for it, on a first
+approximation, and then makes the mapping bijective by introducing
+\linketa; this detail is discussed in section \ref{sec:eta}.
+
+As we mentioned in section~\ref{sec:nutshell} the compiler
+replaces terms in \maybebeta and \maybebeta with fresh variables
+linked to the problematic terms. Each class of problematic terms
+has a dedicated link.
 
@@
 
-\subsection{Notations}
-
-When we write \Ho terms outside code blocks we prefer to use the
-usual and more compact mathematical notation. However we need to
+\begin{invariant}[Link left hand side] The left hand side of a new link
+  is a flexible term.
+\end{invariant}
+
+\noindent
+The right hand side of a link, the problematic term, can occur under binders.
+To accommodate this situation the compiler wraps \elpiIn{baselink} using
+the \elpiIn{inctx} container.
+
+\subsection{Notational conventions}
+
+When we write \Ho terms outside code blocks we follow the
+usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
+$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
+However we need to
 distinguish between the ``application'' of a unification variable
-to its scope and the application of a term (possibly a unification
-variable) to a list of arguments. We write the scope in subscript
-and we use juxtaposition for application. Here a few examples:
+to its scope and the application of a term to a list of arguments.
+We write the scope of unification variables in subscript
+while we use juxtaposition for regular application.
+Here a few examples:
 \vspace{5pt}
@@
 \begin{tabular}{ll}
-  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
   $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
   $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
+  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
   $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
@@
 \noindent
-When detailing example we prefer to write links in mathematical notation,
-as equations between terms in a context. The equality sign is subscripted with
+When detailing examples we write links as equations between terms under a context.
+The equality sign is subscripted with
 kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
@@
 
-here we give the functions/signatures to express the properties 3-5 in the problem
-statement
-
-Together with the description of the terms of the language, we need some
-auxiliary functions to perform operations like term equality, unification,
-dereferencing. There predicates are supposed to be implemented in the OL and the
-ML following respectively their specification. In the following few paragraphs
-we give a brief proposition of them.
-
-\paragraph{Term dereferencing: $\rho s$ and $\sigma t$}
-
-Since in our encoding we explicitly carry a substitution we need to
-define the operation that applies it to a term. Its proposition
-takes care to normalize (flatten) applications, for example
-it turns
-\elpiIn{(app [app [con"f",con"a"],con"b"])}
-into
-\elpiIn{(app [con"f",con"a",con"b"])}.
-
-dereference variables of the two languages. This is particularly useful to check
-if two terms in the OL (resp. in the ML) are equal. The constructors
-representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
-make the dereferencing procedure to recurse over the their subterms. The code
-below display this behavior for the dereferencing performed by the OL:
+In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
+we need to equip \Fo and \Ho with term equality,
+substitution application and unification.
+
+\paragraph{Term equality: \Eo vs. \Ee} We extend the equational theory
+over ground terms to the full languages by adding the reflexivity of
+unification variables (a variable is equal to itself).
+
+The first four rules are common to both equalities and correspond to $\alpha$-equivalence.
+In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
+
+\input{code/feq}
+
+\begin{elpicode}
+  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
+  con C ~\Ee~fcon C.
+  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
+  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
+  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
+\end{elpicode}
+
+\noindent
+The main point in showing these equality tests is to remark how weaker \Ee is,
+and to identify the four rules that need special treatment in
+the implementation of \Ue.
+
+For reference, \elpiIn{(beta T A R)} reduces away \elpiIn{lam} nodes in head
+position in \elpiIn{T} whenever the list \elpiIn{A} provides a corresponding
+argument.
+
+\input{code/beta_fo}
+
+\noindent
+The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
+variables). Elpi provides it as a builtin, but one could implement it by
+systematically loading the hypothetical rule \elpiIn{name x} every time
+a nominal constant is postulated via \elpiIn{pi x\ }.
+
+\paragraph{Substitution application: $\rho s$ and $\sigma t$}
+
+Applying the substitution corresponds to dereferencing a term with respect to
+the memory. To ease the comparison we split \Fo dereferencing into a
+\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
+to memory cells that are set with their values, ans has a corresponding
+operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp},
+in charge of ``flattening'' \elpiIn{fapp} nodes, has no corresponding
+operation in \Ho. The reasons for this asymmetry is that an \elpiIn{fapp}
+node with a flexible head is always mapped to a \elpiIn{uva} (as per
+sections \ref{sec:compilation,sec:beta}), preventing nested applications to
+materialize.
 
@@
 \noindent
-TODO explain napp.
-We use the cut operator to keep the code compact. It is possible to rewrite
-the rule for application traversal so that it is mutually exclusive with the
-first one, but that requires a rather verbose analysis of the head of \elpiIn{A}.
-
-The corresponding code for \Ho is similar, we only show the last two rules that
-differ in a substantial way:
+TODO: about the cut\marginpar{should we rename forall2 into map}
 
@@
 \noindent
-TODO: no need to napp, see the beta section.
-Note that when the substitution \elpiIn{S} maps a unification variable
-\elpiIn{N} to an assignment \elpiIn{F} we ....
-
-TODO: invariant: variables in subst are always fully applied, and length of scope is the
-arity of the HO variable.
-
-Important!!!
-A different reasoning is to be addressed to the variables of the ML. Firstly, a
-meta variable cannot appear in the \elpiIn{app} node as the first element of the
-list, we will explain why in \cref{sec:compilation}
-
-\paragraph{Term equality: \Eo vs. \Ee} We can test if two terms are equal
-following the equational theory of the language being considered. In
-\cref{code:eq-pred} we provide an implementation of the \Eo predicate. The first
-four rules check if the two terms are equal regarding the structure of the
-current node, that is, two terms are equal if they have same head and if
-recursively each subterm is two by two equal. Moreover, since the theory of the
-OL accepts $\eta$- and $\beta$-equivalence, then we implement the corresponding
-rules.
-
-\input{code/feq}
-
-\begin{elpicode}
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-  app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
-  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
-  con C ~\Ee~fcon C.
-  uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
-\end{elpicode}
-
-\noindent
-The equality relation for the ML, accepts $\eta\beta$-equivalence between terms
-of the ML. Recall that \elpiIn{abs x\ f x}, is a valid $\eta$ expansion of the
-function \elpiIn{f} and that \elpiIn{lam x\ app[f, x]} is not that equivalent to
-\elpiIn{f} at meta level. However, since we are interested in using the
-unification procure of the ML, by \cref{prop:correct-ml}, we can use the
-\Ue relation to test, when needed if two terms are equal in the ML.
+Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
+of the arguments of all occurrences of a unification variable and the
+number of abstractions in its assignment have to match.
 
 \paragraph{Term unification: \Uo vs. \Ue}
-The last but not least important relation we should take care of before
-presenting our full algorithm aiming to unify terms of the OL in the ML and
-provide the substitution produced in the ML to the OL, is term unification. This
-procedure is a more powerful version of the equal predicate, since unification
-checks if two terms can be equal be assigning unification variables. In our
-representation, variable assignment (or refinement) is performed by modifying
-the corresponding substitution mapping. We will not give an implementation of
-\Uo, since we are giving an implementation of it using
-our algorithm, see \cref{XX}.
+
+In this paper we assume to have an implementation of \Ue that satisfies
+properties \ref{prop:correct-ml,prop:complete-ml}. Although we provide an
+implementation in the appendix (that we used for testing purposes) we only
+describe its signature. Elpi is expected to provide this brick, as well as
+any other implementation of $\lambda$Prolog.
 
@@
 
-On the other hand, unification in the ML needs to be defined. In
-\cref{code:ue-pred}, we give an implementation of \Ue but
-that is actually what our meta language provides as a builtin.
-
-This predicate has four
-arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
-$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
-\subseteq \rho_2$.
+\noindent
+The only detail worth discussing is the fact that the procedure updates a
+substitution, rather than just crafting one as presented in
+section~\ref{sec:problem-statement}. The reason is that the algorithm folds
+over a term, updating a substitution while it traverses it.\marginpar{explain better}
+
 % The first three rules unify terms with same rigid heads, and

### 2024-03-23 paper/appendix.tex
url
use map

@@
 \section*{Appendix}
+
+This appendix contains the entire code described in this paper.
+The code can also be accessed at the URL: \url{https://github.com/FissoreD/paper-ho}
 
@@
     pi x y\ tm->fm _ x y => tm->fm L (B1 x) (B2 y).
-  tm->fm L (ho.app L1) T :- forall2 (tm->fm L) L1 [Hd|Tl], 
+  tm->fm L (ho.app L1) T :- map (tm->fm L) L1 [Hd|Tl], 
     fo.mk-app Hd Tl T.
   tm->fm L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
-    forall2 (tm->fm L) TL T1, fo.mk-app (fo.fuva VO) T1 T.
+    map (tm->fm L) TL T1, fo.mk-app (fo.fuva VO) T1 T.
 

### 2024-03-23 paper/main.tex
use map

@@
 \noindent
-TODO: about the cut\marginpar{should we rename forall2 into map}
+TODO: about the cut
 
@@
 In this paper we assume to have an implementation of \Ue that satisfies
-properties \ref{prop:correct-ml,prop:complete-ml}. Although we provide an
+properties~\ref{prop:correct-ml} and~\ref{prop:complete-ml}. Although we provide an
 implementation in the appendix (that we used for testing purposes) we only
-describe its signature. Elpi is expected to provide this brick, as well as
+describe its signature here. Elpi is expected to provide this brick, as well as
 any other implementation of $\lambda$Prolog.

### 2024-03-23 paper/appendix.tex
unmess renaming

@@
   type decompl-subst map -> map -> ho.subst -> 
-    fo.subst -> fo.subst -> o.
+    fo.fsubst -> fo.fsubst -> o.
   decompl-subst _ [A|_] _ _ _ :- fail.
@@
   type add-new-map-aux ho.subst -> list ho.tm -> map -> 
-        map ->  fo.subst -> fo.subst -> o.
+        map ->  fo.fsubst -> fo.fsubst -> o.
   add-new-map-aux _ [] _ [] S S.
@@
   type add-new-map ho.subst -> ho.tm -> map -> 
-      map ->  fo.subst -> fo.subst -> o.
+      map ->  fo.fsubst -> fo.fsubst -> o.
   add-new-map _ (ho.uva N _) Map [] F1 F1 :- 
@@
   type complete-mapping-under-ass ho.subst -> ho.assignment -> 
-    map -> map ->  fo.subst -> fo.subst -> o.
+    map -> map ->  fo.fsubst -> fo.fsubst -> o.
   complete-mapping-under-ass H (ho.val Val) Map1 Map2 F1 F2 :- 
@@
   type complete-mapping ho.subst -> ho.subst ->   
-    map -> map -> fo.subst -> fo.subst -> o.
+    map -> map -> fo.fsubst -> fo.fsubst -> o.
   complete-mapping _ [] L L F F.
@@
   type decompile map -> links -> ho.subst -> 
-    fo.subst -> fo.subst -> o.
+    fo.fsubst -> fo.fsubst -> o.
   decompile Map1 L HO FO FO2 :- 

### 2024-03-23 paper/main.tex
update tex

@@
 %
-\newcommand{\progress}{\ensuremath{\mathrm{check}}\xspace}
+\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
 \newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
@@
 
-\newcommand{\wellb}{\ensuremath{\mathcal{W}}}
+\newcommand{\wellb}{\ensuremath{\mathcal{W}}\xspace}
 \begin{definition}[Well behaved set]
@@
 \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
+\label{invariant:arity}
 \end{invariant}
@@
 
+TODO: add ref to \cref{sec:invariant1}
+
 \input{code/alloc_mapping}
@@
 \end{invariant}
+
+\marginpar{WARNING: not true if we delay beta}
 
@@
 
-\section{Use of multivars}
-
-Se il termine initziale è della forma
-
-\begin{elpicode}
-  app[con"xxx", (lam x\ lam y\ Y y x), (lam x\ f)]
-  =
-  app[con"xxx",X,X]
-\end{elpicode}
-
-allora se non uso due X diverse non ho modo di recuperare il quoziente che mi manca.
-
-a sto punto consideriamo liste di problemi e così da eliminare sta xxx senza
-perdità di generalità (e facciamo problemi più corti, e modellizziamo anche la
-sequenza)
-
-\subsection{Problems with $\eta$}\label{sec:eta}
+% \section{Use of multivars}
+
+% Se il termine initziale è della forma
+
+% \begin{elpicode}
+%   app[con"xxx", (lam x\ lam y\ Y y x), (lam x\ f)]
+%   =
+%   app[con"xxx",X,X]
+% \end{elpicode}
+
+% allora se non uso due X diverse non ho modo di recuperare il quoziente che mi manca.
+
+% a sto punto consideriamo liste di problemi e così da eliminare sta xxx senza
+% perdità di generalità (e facciamo problemi più corti, e modellizziamo anche la
+% sequenza)
+
+\section{Handling of \maybeeta}\label{sec:eta}
+
+Even though the unification process explained in the previous sections is able
+to solve a large number of unification problems, it remains still incomplete:
+\wellb is only a subset of terms in \Ho. In order to capture all the unification
+properties of \Eo, we need ad-hoc compilation strategies over those subterms
+that have been defined as ``problematic''.
+
+\subsection{Compilation}
+\subsection{Progress}
 
@@
 
-\subsection{Problems with $\beta$}\label{sec:beta}
+\section{Enforcing Invariant~\ref{invariant:arity}}
+\label{sec:invariant1}
+Deduplicate mapping code etc...
+
+\section{Handling of \maybebeta}\label{sec:beta}
 
@@
 and the right hand side (\rhs) of the \linkbeta.
+
+\subsection{Compilation}
 
@@
 
-One created, there exist two main situations waking up a suspended \linkbeta.
+\subsection{Progress}
+
+Once created, there exist two main situations waking up a suspended \linkbeta.
 The former is strictly connected to the definition of $\beta$-redex and occurs
@@
 \elpiIn{Scope1} and \elpiIn{Extra1} is the same as the concatenation of
-\elpiIn{Scope2} and \elpiIn{Extra2}.
-
-% and some of the arguments in the \elpiIn{NPF} list become names. This is
-% possible after the resolution of other links. In this case, the list \elpiIn{L}
-% obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
-% again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
-% same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
-% new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
-% create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.
-
-\marginpar{A bit too fast, we first compile, then unify, then the oracle, then the manip...}
-An example justifying this last link manipulation is given by the following
+\elpiIn{Scope2} and \elpiIn{Extra2}. Finally, two cases should be considered: 1)
+\elpiIn{Extra2} is the empty list, \lhs and rhs can be unified: we have two
+terms in \llambda; otherwise 2) the \linkbeta in question is replaced with a
+refined version where the \rhs is  \elpiIn{app[uva C Scope2 | Extra2]} and a new
+\linketa is added between the \lhs and the new-added variable \elpiIn{C}.
+\marginpar{dire che si adatta bene nelle approximation}
+
+An example justifying this second  link manipulation is given by the following
 unification problem:
 
-\begin{elpicode}
-  f = flam x\ fapp[F, (X x), a] % $f = \lambda x.F (X x) a$
-\end{elpicode}
-
-\noindent
-under the subsitution $\rho = \{X \mapsto \lambda x.x\}$.
-
-\noindent
-The links generated from this unification problem are:
-
+\def\varF{\ensuremath{\textbf{F}}\xspace}
+\def\varA{\ensuremath{\textbf{A}}\xspace}
+\def\varB{\ensuremath{\textbf{B}}\xspace}
+
+\begin{elpicode}
+  f = flam x\ fapp[F, fapp[A, x]].
+\end{elpicode}
+
+% \begin{elpicode};
+%   f = flam x\ flam y\ fapp[F, fapp[A, x], fapp[B, y]].
+% \end{elpicode}
+
+The compilation of these terms produces the new unification problem: $f = X0$
+
+We obtain the mappings $\mapping{F}{\varF}{0}, \mapping{A}{\varA}{1}$ and the links:
+%
 \begin{gather}
-  \mapping{X}{X1}{1}; \mapping{F}{X2}{0} (\mathrm{The map})\\
-  \linketaM{}{X0}{\lambda x.X3_x}\\
-  \linkbetaM{x}{X3_x}{X2 ~ X1_x ~ a}
+  \linkbetaM{c0}{X3_{c0}}{X2~X1_{c0}}\\
+  \linketaM{}{X0}{\lambda c0.X3_{c0}}
 \end{gather}

### 2024-03-24 paper/main.tex
more work on 5

@@
 
-\begin{invariant}[Link left hand side] The left hand side of a new link
-  is a flexible term.
-\end{invariant}
-
-\marginpar{WARNING: not true if we delay beta}
-
 \noindent
@@
 the \elpiIn{inctx} container.
+
+\begin{invariant}[Link left hand side] The left hand side of a new link
+  is a variable.
+\end{invariant}
+
+\noindent
+If the variable is assigned during a run the link is considered for
+progress and possibly eliminated. This is discussed in
+section~\ref{sec:eta}.\marginpar{is this the right section?}
 
@@
 In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
-we need to equip \Fo and \Ho with term equality,
+we need to equip \Fo{} and \Ho with term equality,
 substitution application and unification.
@@
 
-The objective of the compilation is to recognize the higher-order variables
-available in \Ho{} when expressed in a first order way in \Fo{}. The compiler
-also generates a map to bring back the substitution from
-\Ho{} to \Fo{}.
-
-\input{code/alloc_mapping}
+The main task of the compiler is to recognize \Fo{} variables standing
+for functions and map them to higher order variables in \Ho.
+In order to bring back the substitution from \Ho{} to \Fo{} the compiler
+builds a ``memory map'' connecting the the kind of variables using routine
+\ref{clause:malloc}.
 
@@
 
-This preliminary version of \elpiIn{comp} simply recognizes \Fo variables
+\noindent
+This preliminary version of \elpiIn{comp} recognizes \Fo{} variables
 applied to a (possibly empty) duplicate free list of names (i.e.
 \elpiIn{pattern-fragment} detects variables in \llambda). Note tha compiling
-\elpiIn{Ag} cannot create new mappings nor links, see the comp-lam hyp rule.
-
-The auxiliary function \elpiIn{close-links}
+\elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
+is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
+(see below) grants this property.
 
@@
 
-since we want links to bubble up we use the abs constructor of the inctx data
-type to bind back the variable just crossed, and we do so only if the variable
-v occurs in L.
+\noindent
+The auxiliary function \elpiIn{close-links} tests if the bound variable
+\elpiIn{v} really occurs in the link. If it is the case the link is wrapped into
+an additional \elpiIn{abs} node binding \elpiIn{v}. In this way links generated
+deep inside the compiled terms can be moved outside their original context
+of binders.
+
+Note that \elpiIn{pi x y\..} is syntactic sugar for \elpiIn{pi x\ pi y\..}.
 
@@
 
+XXX links are update unlike section 2
+\input{code/hstep}
+
+note \elpiIn{(a = b) c d} is \elpiIn{(=) a b c d}.
+
 \subsection{Decompilation}
+
+\input{code/decomp}
+\input{code/decompm}
 

### 2024-03-25 paper/appendix.tex
more on 5

@@
 
-  type solve-links links -> links -> ho.subst -> ho.subst -> o.
-  solve-links [] [] X X.
-  solve-links [A|L1] [A|L3] S S2 :- take-link A L1 B L2, !,
+  type progress1 links -> links -> ho.subst -> ho.subst -> o.
+  progress1 [] [] X X.
+  progress1 [A|L1] [A|L3] S S2 :- take-link A L1 B L2, !,
     same-link-eta A B S S1, 
-    solve-links L2 L3 S1 S2.
-  solve-links [L0|L1] L3 S S2 :- deref-link S L0 L,
+    progress1 L2 L3 S1 S2.
+  progress1 [L0|L1] L3 S S2 :- deref-link S L0 L,
     solve-link-abs L R S S1, !,
-    solve-links L1 L2 S1 S2, append R L2 L3.
+    progress1 L1 L2 S1 S2, append R L2 L3.
 \end{elpicode}

### 2024-03-25 paper/main.tex
more on 5

@@
 
-\begin{proposition}[Simulation]\label{prop:sumul}
-$\forall \mathcal{S}, \forall \mathcal{N}$
+\begin{proposition}[Simulation]\label{prop:simulation}
+$\forall \mathcal{S}, \forall \mathcal{N},$
 $$
@@
 In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
-$\forall p \in 1 \ldots \mathcal{N}$
+$\forall p \in 1 \ldots \mathcal{N},$
 $$
@@
 \begin{gather}
-  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
-\unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}\\
+  \unifcorrect{\Eo}{\Uo}{s}\customlabel{prop:correct}{(correct)}\\
+\unifcomplete{\Eo}{\Uo}{s}\customlabel{prop:complete}{(complete)}\\
 % \forall \rho'\rho,
@@
   \rho's_i \in \llambda \Rightarrow
-  \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
+  \rho' s_1 \Uo \rho' s_2 %\label{prop:simulation}
 \end{gather}
@@
 
-Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
+\marginpar{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
 is resilient to unification problems outside \llambda solved by
@@
 
+\noindent
+When it is clear from the context we shall use the same syntax for \Fo{} terms
+(although we never subscripts unification variables).
+
 \subsection{Equational theory and Unification}
@@
 \subsection{Compilation}
+\marginpar{manca beta normal in entrata}
 
@@
 applied to a (possibly empty) duplicate free list of names (i.e.
-\elpiIn{pattern-fragment} detects variables in \llambda). Note tha compiling
-\elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
+\elpiIn{pattern-fragment} detects variables in \llambda).
+Note tha compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
 is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
@@
 
-\input{code/comp_base2}
-
-\noindent
+\input{code/comp_lam}
+
+\noindent
+In the code above the syntax \elpiIn{pi x y\..} is syntactic sugar for
+iterated \elpiIn{pi} abstraction, as in \elpiIn{pi x\ pi y\..}.
+
 The auxiliary function \elpiIn{close-links} tests if the bound variable
@@
 
-Note that \elpiIn{pi x y\..} is syntactic sugar for \elpiIn{pi x\ pi y\..}.
+\input{code/comp_close_links}
+
+\noindent
+Note that we could remove the second rule, whose purpose is to make
+links more readable by pruning unneeded abstractions (unused context entries).
 
@@
 \input{code/hstep}
-
-note \elpiIn{(a = b) c d} is \elpiIn{(=) a b c d}.
+\input{code/progress}
+
+\noindent
+Note thar \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
+\elpiIn{((~\Ue~) A B C D)}.
 
@@
 
-\subsection{Example}
-
-OK
-
-\begin{elpicode}
-Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
-      , flam x\ fapp[fcon"g", fcon"a"] ]
-Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
-lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
-link z z (s z)
-HS = [some (abs x\con"a")]
-S = [some (flam x\fcon a)]
-\end{elpicode}
-
-KO
-
-\begin{elpicode}
-  Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
-  , flam x\ fapp[fcon"g", fcon"a"] ]
-Problems = [ pr 0 1   % $A = \lambda x.x$
-           , pr 2 3 ] % $A a = a$
-lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
-link z z (s z)
-HS = [some (abs x\con"a")]
-S = [some (flam x\fcon a)]
-lam x\ app[f, app[X, x]] = Y,
-  lam x\ x) = X.
-\end{elpicode}
-
-\otext{Goal: $s_1 \Uo s_2$ is compiled into $t_1 \Ue t_2$}
-\otext{What is done: uvars \elpiIn{fo_uv} of OL are replaced into uvars \elpiIn{ho_uv} of the ML}
-\otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
-\otext{Example needing the compiler v0 (tra l'altro lo scope è ignorato):\\ \elpiIn{lam x\ app[con"g",app[uv 0, x]] ~\Uo~lam x\ app[con"g", c"a"]}}
-\otext{Links used to instantiate vars of elpi}
-\otext{After all links, the solution in links are compacted and given to coq}
-\otext{It is not so simple, see next sections (multi-vars, eta, beta)}
-
-The compilation step is meant to recover the higher-order variables of the OL,
-expressed in a first order way, by replacing them with higher-order variables in
-the ML. In particular, every time a variable of the OL is encountered in the
-original term, it is replaced with a meta variable, and if the OL variable is
-applied to a list of distinct names $L$, then this list becomes the scope of the variable.
-For all the other constructors of
-\elpiIn{tm}, the same term constructor is returned and its arguments are
-recursively compiled. The predicate in charge for term compilation is:
-
-\elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
-
-\noindent
-where, we take the term of the OL, produce the term of the ML, take a list
-of link and produce a list of new links, take a substitution and return a
-new substitution.
-
-In particular, due to programming constraints, we need to drag the old subst and
-return a new one extended, if needed, with the new declared meta-variables.
-
-The following code
-%
-\begin{elpicode}
-  kind link type.
-  type link nat -> nat -> nat -> subst.
-\end{elpicode}
-%
-\noindent
-defines a link, which is a relation between to variables indexes, the first
-being the index of a OL variable and the second being the index of a ML
-variable. The third integer\todo{integer or nat?} is the number of term in the
-scope of the two variables, or equivalently, in a typed language, their arity.
-
-As an example, let's study the following unification problem (a slightly
-modified version from \cref{sec:intro}):
-
-\begin{elpicode}
-  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
-    lam x\ app [c"decision", app[uv 0, x]]
-\end{elpicode}
-
-\noindent
-we have the main unification problem where the nested \elpiIn{app} nodes have
-lists of different lengths making the unification to fail. The compilation of
-these terms produces a new unification problem with the following shape:
-
-\begin{elpicode}
-  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~
-    lam x\ app [c"decision", uv 1 [x]]
-\end{elpicode}
-
-\noindent
-The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
-the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
-ML, that is, the index \elpiIn{0} for that unification variable of the OL term
-has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
-different substitution mapping, one for the OL and one for the ML and the indexes
-of variable point to the respective substitution.
-
-decomp che mappa abs verso lam
-\noindent
-\otext{An other example: \\
-  \elpiIn{lam x\ app[f, app[X, x]] = Y, (lam x\ x) = X.}}
+\subsection{Definition of \Uo and its properties}
+
+\input{code/unif_fo}
+
+The code given so far applies to terms in $\beta\eta$-normal form where
+unification variables in \Fo{} can occur non linearly but always with
+the same number of arguments, and where their arguments are distinct names
+(as per \llambda).
+
+\begin{lemma}[Compilation round trip] If
+  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
+\end{lemma}
+\begin{proof}[Proof sketch]
+trivial, since the terms are beta normal beta just builds an app.
+\end{proof}
+
+\begin{lemma}
+Properties \ref{prop:correct} and
+\ref{prop:complete} hold for the implementation of \Uo above
+\end{lemma}
+\begin{proof}[Proof sketch]
+ In this setting \Ee is as strong as
+\Eo on ground terms. What we have to show is that whenever two different \Fo
+terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
+and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
+a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
+If we look at the \Fo{} terms, the are two interesting cases:
+\begin{itemize}
+\item \elpiIn{fuva X ~\Uo~ s}. In this case after \elpiIn{comp} we have
+  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
+  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
+\item \elpiIn{fall[fuva X|L] ~\Uo~ s}. In this case
+ we have $Y_{\vec{x}} \Ue t$ that succeeds with
+ $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
+ is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
+ Thanks to \ref{clause:beta1}
+ $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
+\end{itemize}
+Since the mapping is a bijection occur check in \Ho{} corresponds to occur
+check in \Fo{}.
+\end{proof}
+
+\begin{lemma} Properties simulation (\ref{prop:simulation}) and
+fidelity (\ref{prop:fidelity}) hold
+\end{lemma}
+\begin{proof}[Proof sketch]
+Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
+in this
+context where input terms are $\beta\eta$-normal and we disregard $\eta$-equivalence
+\Ue is equivalent to \Uo.
+\end{proof}
+
+\subsection{Limitations of by this basic scheme}
+
+\begin{gather}
+\lambda x y.F~y~x = \lambda x y.x\\
+\lambda x. f~(F~x)~x = f~(\lambda y.y)
+\end{gather}
+Note that here $F$ is used with different arities, moreover
+in the second problem the left hand side happens to be an
+eta expansion (of $f (\lambda y.y)$) only after we discover that
+$F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
+Both problems are addressed in the next section.
+
+% \subsection{Example}
+
+% OK
+
+% \begin{elpicode}
+% Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
+%       , flam x\ fapp[fcon"g", fcon"a"] ]
+% Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
+% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
+% link z z (s z)
+% HS = [some (abs x\con"a")]
+% S = [some (flam x\fcon a)]
+% \end{elpicode}
+
+% KO
+
+% \begin{elpicode}
+%   Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
+%   , flam x\ fapp[fcon"g", fcon"a"] ]
+% Problems = [ pr 0 1   % $A = \lambda x.x$
+%            , pr 2 3 ] % $A a = a$
+% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
+% link z z (s z)
+% HS = [some (abs x\con"a")]
+% S = [some (flam x\fcon a)]
+% lam x\ app[f, app[X, x]] = Y,
+%   lam x\ x) = X.
+% \end{elpicode}
+
+% \otext{Goal: $s_1 \Uo s_2$ is compiled into $t_1 \Ue t_2$}
+% \otext{What is done: uvars \elpiIn{fo_uv} of OL are replaced into uvars \elpiIn{ho_uv} of the ML}
+% \otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
+% \otext{Example needing the compiler v0 (tra l'altro lo scope è ignorato):\\ \elpiIn{lam x\ app[con"g",app[uv 0, x]] ~\Uo~lam x\ app[con"g", c"a"]}}
+% \otext{Links used to instantiate vars of elpi}
+% \otext{After all links, the solution in links are compacted and given to coq}
+% \otext{It is not so simple, see next sections (multi-vars, eta, beta)}
+
+% The compilation step is meant to recover the higher-order variables of the OL,
+% expressed in a first order way, by replacing them with higher-order variables in
+% the ML. In particular, every time a variable of the OL is encountered in the
+% original term, it is replaced with a meta variable, and if the OL variable is
+% applied to a list of distinct names $L$, then this list becomes the scope of the variable.
+% For all the other constructors of
+% \elpiIn{tm}, the same term constructor is returned and its arguments are
+% recursively compiled. The predicate in charge for term compilation is:
+
+% \elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
+
+% \noindent
+% where, we take the term of the OL, produce the term of the ML, take a list
+% of link and produce a list of new links, take a substitution and return a
+% new substitution.
+
+% In particular, due to programming constraints, we need to drag the old subst and
+% return a new one extended, if needed, with the new declared meta-variables.
+
+% The following code
+% %
+% \begin{elpicode}
+%   kind link type.
+%   type link nat -> nat -> nat -> subst.
+% \end{elpicode}
+% %
+% \noindent
+% defines a link, which is a relation between to variables indexes, the first
+% being the index of a OL variable and the second being the index of a ML
+% variable. The third integer\todo{integer or nat?} is the number of term in the
+% scope of the two variables, or equivalently, in a typed language, their arity.
+
+% As an example, let's study the following unification problem (a slightly
+% modified version from \cref{sec:intro}):
+
+% \begin{elpicode}
+%   lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
+%     lam x\ app [c"decision", app[uv 0, x]]
+% \end{elpicode}
+
+% \noindent
+% we have the main unification problem where the nested \elpiIn{app} nodes have
+% lists of different lengths making the unification to fail. The compilation of
+% these terms produces a new unification problem with the following shape:
+
+% \begin{elpicode}
+%   lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~
+%     lam x\ app [c"decision", uv 1 [x]]
+% \end{elpicode}
+
+% \noindent
+% The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
+% the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
+% ML, that is, the index \elpiIn{0} for that unification variable of the OL term
+% has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
+% different substitution mapping, one for the OL and one for the ML and the indexes
+% of variable point to the respective substitution.
+
+% decomp che mappa abs verso lam
+% \noindent
+% \otext{An other example: \\
+%   \elpiIn{lam x\ app[f, app[X, x]] = Y, (lam x\ x) = X.}}
 

### 2024-03-25 paper/code/generate_tex.py

@@ -21,6 +21,7 @@ def print_tex(lines, fout, raw = False):
             l = re.sub("^ *%ENDSNIP.*\n","",l)   
             l = re.sub("^ *%%%.*\n","",l)   
             l = re.sub("==l",r"~$\\Ue$~",l) 
+            l = re.sub("===o",r"~$\\Uo$~",l)
             l = re.sub("==o",r"~$\\Eo$~",l)
             l = re.sub(".*% *HIDE.*\n","",l)
             l = re.sub("type \(~\$([^ ]+)\$~\) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~",l)

### 2024-03-25 paper/main.tex
progress can change the link store

@@
 \ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
-\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
+\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
 \newcommand{\stepHD}[6]{\ensuremath{%
-#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
+#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
 \newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
@@
 \newcommand{\runHD}[3]{\ensuremath{%
-\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
+\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
@@
 \begin{array}{l}
-\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''} \deff\vspace{2pt}\\
+\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
   \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
 \runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
   \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
-  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore}{\rho_{\mathcal{N}}}
+  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
 \end{array}
@@
 \Leftrightarrow
-\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}
+\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
 $$
@@
 \quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \progress~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
-\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
+    \progress~(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
+\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
 \end{array}
@@
 \noindent
-TODO: about the cut
+Note that the cut operator is inessential, it could be
+removed at the cost of a verbose test on the head of \elpiIn{L}
+in the last rule (\elpiIn{L} head
+can be \elpiIn{fcon}, \elpiIn{flam} or a name).
+
+Applying the substitution in \Ho{} is very similar, with
+the caveat that assignments have to be moved to the
+current scope, i.e. renaming the \elpiIn{abs}-bound variables
+with the names in the scope of the unification variable occurrence.
 
@@
 of the arguments of all occurrences of a unification variable and the
-number of abstractions in its assignment have to match.
+number of abstractions in its assignment have to match. In turn
+this grants that \elpiIn{move} never fails.
+
+\input{code/move}
 
@@
 but play a major role in Sections \ref{sec:eta} and \ref{sec:beta}.
+With respect to \ref{sec:problem-statement} the signature also allows
+for updates to the substitution.  The code below only allocates space
+for the variables, i.e. sets their memory address to \elpiIn{none},
+a details not worth mentioning in the previous discussion.
 
@@
 \input{code/hstep}
+
+TODOD: discuss signature of progress
+
 \input{code/progress}
@@
 
-\subsection{Decompilation}
+TODO: why it terminates
+
+\subsection{Substitution decompilation}
 

### 2024-03-25 paper/main.tex
more5

@@
 \begin{elpicode}
-decision (all A x\ Pm x) :- link Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
+decision (all A x\ Pm x) :- decomp Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
   pi x\ decision (app[P, x]).
@@
 After unifying the head of rule \ref{clause:r3a} with the goal, Elpi runs
-the premise <<\elpiIn{link Pm A P}>> that is in charge of bringing the
+the premise <<\elpiIn{decomp Pm A P}>> that is in charge of bringing the
 assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:
@@
 
-\begin{invariant}[Link left hand side] The left hand side of a new link
+\begin{invariant}[Link left hand side]\ref{inv:linklhs}
+  The left hand side of a new link
   is a variable.
@@
 
-XXX links are update unlike section 2
+A step in \Ho consists in unifying two terms and reconsidering all
+links for progress. If any of the two tasks fail we say that the entire step
+fails, and it is at this granularity that we can relate steps in the
+two languages.
+
 \input{code/hstep}
 
-TODOD: discuss signature of progress
+\noindent
+Note tha the notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
+\elpiIn{((~\Ue~) A B C D)}.
+
+Reconsidering links is a fixpoint, since the progress of a link can update the
+substitution and in turn enable another link to progress.
 
@@
 \noindent
-Note thar \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
-\elpiIn{((~\Ue~) A B C D)}.
-
-TODO: why it terminates
+In the base compilation scheme \elpiIn{progress1} is the identity
+on both the links and the substitution, so the fixpoint trivially terminates.
+Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
+and justify why the don't hinder termination.
+
+TODO: discuss occur check
 
@@
 
+Decompiling the substitution requires to first force the
+progress of links and then allocating new unassigned variables
+in the substitution for \Fo{} and finally decompiling all
+assignments. Note that \ref{inv:linklhs} and the
+occur check allows us to update the subst.
+
+\input{code/decompile}
+
+\noindent
+Decompiling an assignment requires to turn abstractions into
+lambdas. For aestetic purposes we also eta-contract the result
+(not needed since Fo equality can do that)
+
+\input{code/decompm}
+
+\noindent
+Finally decompiling a term is trivial, now that we have an extended
+mapping containing all unassigned variables \Ue may have introduced.
+
 \input{code/decomp}
-\input{code/decompm}
+
+\noindent
+Note that we use beta to build fapp nodes when needed (if Ag is empty
+no \elpiIn{fapp} node should appear).
 

### 2024-03-25 paper/code/generate_tex.py
code: update

@@ -24,6 +24,7 @@ def print_tex(lines, fout, raw = False):
             l = re.sub("===o",r"~$\\Uo$~",l)
             l = re.sub("==o",r"~$\\Eo$~",l)
             l = re.sub(".*% *HIDE.*\n","",l)
+            l = re.sub("% label: (.*).* cnt: (.*)",r"~\\customlabel{\g<1>}{(\g<2>)}~",l)
             l = re.sub("type \(~\$([^ ]+)\$~\) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~",l)
             l = re.sub("type (\([^ ]+\)) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~",l)
             f.write(l)

### 2024-03-26 paper/main.tex
wip: eta and beta

@@
 
+\newcommand{\appsep}{\ensuremath{\textcolor{lightgray}{\cdot}}}
 \newcommand{\EqualRel}{\ensuremath{=}}
@@
 As we mentioned in section~\ref{sec:nutshell} the compiler
-replaces terms in \maybebeta and \maybebeta with fresh variables
+replaces terms in \maybeeta and \maybebeta with fresh variables
 linked to the problematic terms. Each class of problematic terms
@@
 
+% TODO: io sposteri l'invariante sopra nella maybe-eta, perchè  
+% la proprietà è vera solo per i link eta
+% per i link beta, questo non è vero se sospendiamo i link
 \noindent
@@
 \begin{itemize}
-\item \elpiIn{fuva X ~\Uo~ s}. In this case after \elpiIn{comp} we have
+\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
   $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
   $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
-\item \elpiIn{fall[fuva X|L] ~\Uo~ s}. In this case
+\item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
  we have $Y_{\vec{x}} \Ue t$ that succeeds with
@@
 
-Even though the unification process explained in the previous sections is able
-to solve a large number of unification problems, it remains still incomplete:
-\wellb is only a subset of terms in \Ho. In order to capture all the unification
-properties of \Eo, we need ad-hoc compilation strategies over those subterms
-that have been defined as ``problematic''.
+$\eta$-reduction is a property by which a term $t$ is of the form $\lambda x. T
+x$ and $x$ is not free in $T$. In this case we case say that $T$ is equal to its
+$\eta$-expanded version represented by the term $t$. Both the meta and the
+object language accept terms unification up to $\eta$ equivalence, but as
+sketched in the small code snippet in \cref{sec:problem-statement}, \Ue is not
+able to capture this equivalence property when using terms in \Fo. The main
+reason is that the unification of two terms in \Fo{} appears to be syntactical
+on the constructors of these terms. As a very concise example, \Ue can't unify
+the \Fo{} terms \elpiIn{lam x\ app[con "f", x]} and \elpiIn{con "f"} since the
+rigid constructor \elpiIn{lam} is different from the rigid constructor
+\elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
+order to address this problem.
 
 \subsection{Compilation}
+
+\paragraph{Detection of \maybeeta} The main modification of the compiler to solve this
+unification issue consists in identifying all the subterms of a term $t$ that
+are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
+$\eta$-expansion under a certain substitution $\sigma$, the code verifying this
+property is given by the following code:
+
+\input{code/maybe_eta}
+
+The entry point is depicted by the rule~\ref{rule:maybeeta} which takes a name
+$n$, a term $t$ and a list of bound variables $L$ (originally it is the
+singleton containing $n$). This rule checks if $t$ is a term of the form $T
+\appsep n$ (for a term $T$), together with the auxiliary predicate
+\elpiIn{reducible-to} which ensures if a term $t$ can reduce to a name $n$.
+\marginpar{explain better: we start from $\lambda x.F x$}
+The \elpiIn{maybe-eta} predicate dispatches the calls to \elpiIn{reducible-to};
+three cases should be considered: 1) $t$ is a variable $v$, then $t$ can be an
+$\eta$-expansion if at least one of the terms in the scope of $v$ is a
+$\maybeeta$ of $n$; 2) $t$ is a lambda-term, then we recursively call
+\elpiIn{maybe-eta} on the body of $t$ under a local name $x$ which is added to
+the list $L$; 3) $t$ is an application, then $t$ is an $\eta$ expansion if i)
+the last arguments of $t$ can be reduced one by one to the binders in the list
+$L$ (we reverse the list in rule, since, by construction, this list is built in
+reversed order) and ii) none of the first arguments of the application contain a
+rigid occurrence of name in $L$.
+
+As rapidly said before, \elpiIn{reducible-to} tells if a term $t$ reduce to a
+name $n$, or equivalently if $\exists \sigma, \sigma t = n$. This predicate also
+takes the list of all the binders explored (this list is originally empty). A
+term $t$ reduces to a name $n$ if 1) $n = t$; 2) $t$ is a variable $v$, then $t$
+reduce to a $n$ if it exists an arugment in the scope of $v$ reducing to $n$ and
+forall name $n'$ in $L$, there is an argument reducing to $n'$ \marginpar{Note
+that this function is partial: the arg in the scope should be distinct}; 3) $t$
+is a lambda abstraction, then we call recursively \elpiIn{reducible-to} on the
+body of the abstraction with a new local name added to the list $L$; 4) $t$ is
+an application of $n$ to a list of arguments $L'$, then all the arguments should
+reduce to the respective name in the list $L$.
+
+Finally, a name $n$ occurs rigidly in a term $t$ if $n$ occurs in a subterm $t'$
+of $t$ such that $t'$ does not appear in the scope of a variable.
+
+An example of \maybeeta detection over the bound variable $x$ is the following:
+\marginpar{maybe more examples in appendix}
+%
+\begin{gather}
+  T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
+\end{gather}
+
+The correct call to \elpiIn{maybe-eta} is \elpiIn{maybe-eta x T L} with
+\elpiIn{L = [x]}. At first we go under the abstraction $\lambda y$ adding $y$ to
+$L$. Then we find an application, where we verify that 1) $A_{xy}$ does not
+contain $x$ and $y$ rigidly which is the case; 2) $B \appsep a \appsep (\lambda z.y\appsep (C_z))$
+and $D_x$ can respectively reduce to $y$ and $x$. The latter reduction is
+evident, since $D$ has $x$ in scope; the former subterm can reduce to $y$ since
+$B$ is a variable and it exists an argument ($\lambda z.y\appsep C_z$) reducible to
+$y$: under the binder $z$, we have the application of $y$ with a variable with
+$z$ in scope. Note that \maybeeta, only tells if it exists a substitution making
+a term an eta expansion on any term $t$, i.e, $t$ can be in $\maybeeta \cup
+\maybebeta$ without the constraint of being in \llambda, as our example shows. A
+possible substitution making the term $T$ in the example an $\eta$-expansion is
+$\sigma = \{A \mapsto \lambda x.\lambda y.a, B \mapsto \lambda x.\lambda y.y,
+C \mapsto \lambda x.x, D \mapsto \lambda x.x\}$.
+
+\paragraph{Compilation with \linketa} Thanks to the \elpiIn{maybe-eta}
+predicate, we can detect ``$\eta$-problematic'' terms and, consequently replace
+them with a fresh \Ho unification variable at compilation time. The code
+below illustrate this dedicated compilation:
+% TODO: maybe ref to \cref{sec:nutshell}
+
+\input{code/comp_eta}
+
+\marginpar{Where This rule is put in the code ?}
+This rule is applied on \Fo{} lambda-abstractions, tests if their body $F$ is an
+\maybeeta wrt to a local fresh binder $x$, and if so, it compiles $F$ to the \Ho
+term \elpiIn{F1} and returns a \Ho fresh variable $A$ having in scope the free
+names occurring in \elpiIn{F1}. As sketch at the very end of
+\cref{sec:lang-spec}, each time a subterm $t$ in \Fo is replaced with a \Ho
+variable $v$, we build a \elpiIn{link-eta} between $v$ and and the term $t'$
+obtained by the compilation of $t$.
+
+A \elpiIn{link-eta}\footnote{\elpiIn{@val-link-eta A B} is syntactic sugar for
+\elpiIn{val (link-eta A B)}}
+is defined as
+
+\elpiIn{type link-eta tm -> tm -> baselink}
+
+\noindent
+We call the two terms carried by the link respectively left and right hand side.
+
+\begin{invariant}[\linketa left hand side]
+  The left hand side of a \linketa is a variable.
+  \label{invariant:link-eta-left}
+\end{invariant}
+
+\begin{invariant}[\linketa right hand side]
+  The right hand side of a \linketa having the shape $\lambda x.F \appsep x$
+  where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
+  \label{invariant:link-eta-right}
+\end{invariant}
+
 \subsection{Progress}
 
-\input{code/comp_eta}
-
-and aux
-
-\input{code/maybe_eta}
-
-\otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
-\otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
-\otext{It is not doable, with the same elpi var}
-
-Invarianti:
-A destra della eta abbiamo sempre un termine che comincia per $\lambda x. bla$
+We want maintain invariants~\ref{invariant:link-eta-left} and \ref{invariant:link-eta-right}
 

### 2024-03-26 paper/main.tex
w

--- 2024-03-25 paper/main.tex
@@
 
+\newtheorem{invariant}{Invariant}
+\crefname{invariant}{invariant}{invariants}
+
 \def\ol{\acronym{ol}} % object language
@@
 \newcommand{\mapping}[3]{\ensuremath{#1 \mapsto #2^#3}}
+
+\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
+\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
 
@@
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
-\def\linkStore{\ensuremath{\mathbb{L}}}
+\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
 
@@
 
-\newtheorem{invariant}{Invariant}
-
 \begin{invariant}[Unification variable arity]\label{inv:uvaarity}
@@
 
-\begin{invariant}[Link left hand side]\ref{inv:linklhs}
+\begin{invariant}[Link left hand side]\label{inv:linklhs}
   The left hand side of a new link
@@
 \subsection{Execution}
+\label{sec:execution}
 
@@
 \noindent
-Note tha the notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
-\elpiIn{((~\Ue~) A B C D)}.
+Note tha the infix notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
+\elpiIn{((~\Ue\!\!\!~) A B C D)}.
 
@@
 
-Even though the unification process explained in the previous sections is able
-to solve a large number of unification problems, it remains still incomplete:
-\wellb is only a subset of terms in \Ho. In order to capture all the unification
-properties of \Eo, we need ad-hoc compilation strategies over those subterms
-that have been defined as ``problematic''.
+$\eta$-reduction is an equivalence relation over terms where a term of the form
+$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
+free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
+version of $t$. The implementation of the \elpiIn{comp} relation given in
+\cref{sec:compilation} compiles the \Fo{} terms $t_1 = $\elpiIn{flam x\ fapp
+[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
+$\elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
+\mapping{A}{A'}{1}. However, if the oracle sets \elpiIn{A'} to the constant
+\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
+fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
+fact that $t'_1 =$\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
+=$\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
+unification problem by adapting the \elpiIn{comp} relation such that it
+recognizes \maybeeta subterms $s$ and replaces them with fresh \Ho variables
+$v$. This link between the variable $v$ and the subterm $t$ is stored in what we
+call \linketa which is an object with the following type
+
+\elpiIn{type link-eta tm -> tm -> baselink}
+
+\noindent
+where, as sketched in \cref{sec:lang-spec}, the term on the left hand side
+(\lhs) is linked with its left counterpart (\rhs).
+
+\linketa are added in the link store (\linkStore) and activated when special
+conditions are satisfied on \lhs or \rhs. These link activations are managed by
+extending the \elpiIn{progress1} predicate (see \cref{sec:execution}).
+We claim that \linketa progression does not contradict \cref{inv:linklhs} and we
+add the following invariant:
+
+\begin{invariant}[\linketa \rhs]
+  The \rhs of a \linketa having the shape $\lambda x.F_x$
+  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
+  \label{inv:link-eta-right}
+\end{invariant}
+
+In the next three subsections we explain how we detect \maybeeta terms, how we
+compile them and how \linketa are activated during the execution of the program
+and provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.
+
+\subsection{Detection of \maybeeta}
+
+Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
+for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
+given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
+\maybeeta detection is not a trivial operation as it may seems.
+%
+\begin{gather}
+  \lambda x. f \appsep A_x\\
+  \lambda x. f \appsep x \appsep A_x\\
+  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x}\\ \label{eq:lamlam}
+\end{gather}
+
+In the examples above, the first expression is a \maybeeta since $A_x$ can
+reduce to $x$, the second one is not a \maybeeta since for any substitution for
+$A_x$, $x$ is not free in $f \appsep x$. The third equation is a bit more
+complicated since, we have a spine of lambdas, this means that the whole term
+is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
+$t$, and $t$ can be reduced to a term on the form $t' x$ where $x$ is not free
+in $t'$. Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
+$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.
+
+TODO: clarify this
+As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
+since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
+subterm does not use $x$, for example if $A$ is a function on the form $\lambda
+x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
+consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
+the given subterm.
+
+We can now define more formally the two auxiliary relation we need for \maybeeta
+detection:
+
+\begin{definition}[reduce-to]
+  A term $t$ reduce to a name $x$, if $\exists \sigma, \sigma t = x$. In
+  particular, for any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$
+  reduces to a bound variable $x$ if one of the three following cases is
+  satisfied: 1) $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
+  of term $l$ and each $l_i$ reduces to $x_i$; 3) $t$ is a variable with scope
+  $s$, and for any $bv \in [x|\vec{x}]$, it exists a $s_i$, such that $s_i$
+  reduces to $bv$
+\end{definition}
+
+\begin{definition}[occurs-rigidly]
+  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, \exists t' \in
+  \subterm{t}, t' = x$, i.e. $x$ does not appear in the scope of any variable
+  in $t$.
+\end{definition}
+
+Finally, we can derive the implementation for \maybeeta detection:
+
+\begin{definition}[\maybeeta detection]
+  For any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$ is a \maybeeta if
+  1) either $t$ is the application of a term with argument $l_1 \ldots l_m$ such that 
+  $m \geq n$, every $l_{m-n+i}$ reduce to $x_i$, and no $l_{1\ldots m-n-1}$
+  reduces to a name in $\vec{x}$; 2) or $t$ is a variable with scope $s$ and
+  each $x_i$ it exists $s_j$ such that $s_j$ reduces $x_i$.
+\end{definition}
+
+As a final remark, the \maybeeta detection defined just before is an
+over-approximation, in the sense that there exists some terms $t$ considered as
+\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
+$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$. On
+the other hand, we also point out that the \maybeeta detection spot out
+potential $\eta$-expansion for terms that are not in \llambda. For example,
+$\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
+application of term whose argument can reduce to $x$.
+
+The implementation we propose for the \maybeeta relation is given below.
+
+\input{code/maybe_eta}
+
+% Here a complex maybeeta example
+% \begin{gather}
+%   T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
+% \end{gather}
 
 \subsection{Compilation}
+
+Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
+terms and, consequently replace them with fresh \Ho unification variables at
+compilation time. The code below illustrate how this relation is used to for
+term compilation.
+
+\input{code/comp_eta}
+
+This rule, to be inserted just before rule~\ref{rule:complam} from the code in
+\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
+\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
+variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
+in $t'$. The critical part of this compilation is the creation of the \linketa,
+which links the variable \elpiIn{A} with $t$. This link creation enforce
+\cref{inv:linklhs} and \cref{inv:link-eta-right}, since clearly the \lhs is a
+variable and the \rhs is a term starting with the \elpiIn{lam} constructor. We
+can also note that for any name $x$, $tx = t' \appsep x$ never start with
+\elpiIn{lam}, since by contradiction, if 1) $tx$ is an $\eta$-expansion, then
+the recursive call to \elpiIn{comp} would have replaced $tx$ with a fresh
+variable, which contradict the hypothesis and if 2) $t'$ is not an
+$\eta$-expansion, then neither $\lambda t'$ is.
+
 \subsection{Progress}
 
-\input{code/comp_eta}
-
-and aux
-
-\input{code/maybe_eta}
-
-\otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
-\otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
-\otext{It is not doable, with the same elpi var}
-
-Invarianti:
-A destra della eta abbiamo sempre un termine che comincia per $\lambda x. bla$
-
-\begin{textcode}
-  La deduplicate eta:
-  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2]
-    (a destra non c'è mai un termine con testa rigida)
-  - i due termini a dx vengono unificati con la unif e uno dei due link viene buttato
-    NOTA!! A dx abbiamo sempre un termine della forma lam x.VAR x!!!
-    Altrimenti il link sarebbe stato risolto!!
-  - dopo l'unificazione rimane un link [variabile] -> [etaX]
-  - nella progress-eta, se a sx abbiamo una constante o un'app, allora eta-espandiamo
-    di uno per poter unificare con il termine di dx.
-\end{textcode}
+\linketa are meant to suspend the unification of two terms, that is, if we have
+a \maybeeta term $t$ which should be unified with a term $t'$ we don't want to
+unify them with \Ue. As said before this would break unification but it can also
+introduce \maybeeta in the substitution of \Ho.
+
+In order to activate a \linketa, we need to implement new rules for the
+\elpiIn{progress1} predicate. There are two cases making a \linketa to progress,
+1) \lhs is instantiated to a rigid term, in this case \lhs is unified with \rhs.
+TODO: the right eta hand side is eta-expanbded if it is an app/con; 2) \rhs can
+be $\eta$-reduced to a term with rigid head, in this case \lhs and \rhs are
+again unified. If one of these two condition is satisfied, the link has
+fulfilled is task and can be removed from the list of suspended links; if none
+the condition succeeds, the link is kept for a further iteration of
+\elpiIn{progress}.
+
+TODO: dire che scendiamo sotto i vari abs che formano il contesto? 
+TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
+TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$
+
+A second way to progress \linketa, that we call \linketa deduplication, is to
+check if \linkStore contains two \linketa $l_1$ and $l_2$ with same \lhs. This
+situation occurs if two \maybeeta terms are unified with a same unification
+variable. In this case, we can unify the $l_1$ and $l_2$ \rhs (that, by
+construction are both on the form $\lambda x. T_x$) and remove one of the two
+links.
+
+TODO: example for this: $\lambda x. \lambda y. F \appsep y \appsep x = X,
+\lambda x. \lambda y. F \appsep y \appsep x = Y$
+
+We can note that the 1) insertion of these rules for \elpiIn{progress1} do not
+prevent the termination of \elpiIn{progress}, since, a link activation runs
+terminating operations (such as term unification and link-removal); 2) \linketa
+deduplication runs again terminating operations and 3) if none of these two 
+situations is performed, then the substitution and \linkStore
+Note that if the link remain suspended \elpiIn{progress} continues to terminate,
+since, it would mean that the condition of the \elpiIn{if}-branchement succeds
+and the \elpiIn{progress} terminates its execution.
+
+\begin{lemma}
+  We never add eta-expansions in the substitution
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+\end{proof}
+
+TODO: we can however have $\lambda x. F_x$ if we know that $F$ does not reduce
+to $T x$ where x is not free in $T$.
 
@@
 \subsection{Compilation}
+
+\paragraph{Detection of \maybebeta} TODO: \dots
+
+\paragraph{Compilation with \linkbeta}
 

### 2024-03-26 paper/main.tex
update tex

--- 2024-03-26 paper/main.tex
@@
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
-\def\linkStore{\ensuremath{\mathbb{L}}}
+\def\linkStore{\ensuremath{\mathbb{L}}\xspace}
 
@@
 \elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
-order to address this problem.
+order to address this problem. In the sections below, we explain how we detect
+and compile \maybeeta terms with the corresponding \linketa and how these links
+are used to make unification progress. The same explanation structure will be 
+used for \maybebeta terms.
 
@@
 are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
-$\eta$-expansion under a certain substitution $\sigma$, the code verifying this
-property is given by the following code:
+$\eta$-expansion under a certain substitution $\sigma$. The code verifying this
+property is given below:
 
@@
 
-\noindent
-We call the two terms carried by the link respectively left and right hand side.
-
-\begin{invariant}[\linketa left hand side]
-  The left hand side of a \linketa is a variable.
+\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
+\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
+
+\noindent
+We call the two terms carried by the link respectively left and right hand side,
+called respectively \lhs and \rhs.
+
+\begin{invariant}[\linketa \lhs]
+  The \lhs of a \linketa is a variable.
   \label{invariant:link-eta-left}
@@
 
-\begin{invariant}[\linketa right hand side]
-  The right hand side of a \linketa having the shape $\lambda x.F \appsep x$
+TODO: motivate this?
+
+\begin{invariant}[\linketa \rhs]
+  The \rhs of a \linketa having the shape $\lambda x.F \appsep x$
   where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
@@
 
+TODO: motivate this?
+
 \subsection{Progress}
 
-We want maintain invariants~\ref{invariant:link-eta-left} and \ref{invariant:link-eta-right}
+\linketa are meant to suspend the unification of two terms, that is, if we have
+a \maybeeta term $t$ which should be unified with a term $t'$ we don't want to
+unify them with \Ue. As said before this would break unification but it can also
+introduce \maybeeta in the substitution of \Ho.
+
+In order to activate a \linketa, we need to implement new rules for the
+\elpiIn{progress1} predicate. There are two cases making a \linketa to progress,
+1) \lhs is instantiated to a rigid term, in this case \lhs is unified with \rhs.
+TODO: the right eta hand side is eta-expanbded if it is an app/con; 2) \rhs can
+be $\eta$-reduced to a term with rigid head, in this case \lhs and \rhs are
+again unified. If one of these two condition is satisfied, the link has
+fulfilled is task and can be removed from the list of suspended links; if none
+the condition succeeds, the link is kept for a further iteration of
+\elpiIn{progress}.
+
+TODO: dire che scendiamo sotto i vari abs che formano il contesto?
+
+We can note that the insertion of these rules for \elpiIn{progress1} do not
+prevent the termination of \elpiIn{progress}, since, if the link is woken up,
+the unification of the terms and the suppression of the link form \linkStore
+are terminating operation. Note that if the link remain suspend
+\elpiIn{progress} continues to terminate, since, it would mean that the
+condition of the \elpiIn{if}-branchement succeds and the \elpiIn{progress}
+terminates its execution.
+
+A second way to progress \linketa is to check if \linkStore contains two
+\linketa $l_1$ and $l_2$ with same \lhs. In this case we are allowed to unify
+
+TODO: dire che termina
+
+\begin{lemma}
+  We never add eta-expansions in the substitution
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+\end{proof}
+
+TODO: we can however have $\lambda x. F_x$ if we know that $F$ does not reduce
+to $T x$ where x is not free in $T$.
 

### 2024-03-27 paper/main.tex
rework eta intro

@@
 
+\newtheorem{invariant}{Invariant}
+
 \def\ol{\acronym{ol}} % object language
@@
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
-\def\linkStore{\ensuremath{\mathbb{L}}\xspace}
+\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
 
@@
 
-% TODO: io sposteri l'invariante sopra nella maybe-eta, perchè  
-% la proprietà è vera solo per i link eta
-% per i link beta, questo non è vero se sospendiamo i link
 \noindent
@@
 
-$\eta$-reduction is a property by which a term $t$ is of the form $\lambda x. T
-x$ and $x$ is not free in $T$. In this case we case say that $T$ is equal to its
-$\eta$-expanded version represented by the term $t$. Both the meta and the
-object language accept terms unification up to $\eta$ equivalence, but as
-sketched in the small code snippet in \cref{sec:problem-statement}, \Ue is not
-able to capture this equivalence property when using terms in \Fo. The main
-reason is that the unification of two terms in \Fo{} appears to be syntactical
-on the constructors of these terms. As a very concise example, \Ue can't unify
-the \Fo{} terms \elpiIn{lam x\ app[con "f", x]} and \elpiIn{con "f"} since the
-rigid constructor \elpiIn{lam} is different from the rigid constructor
-\elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
-order to address this problem. In the sections below, we explain how we detect
-and compile \maybeeta terms with the corresponding \linketa and how these links
-are used to make unification progress. The same explanation structure will be 
-used for \maybebeta terms.
+$\eta$-reduction is an equivalence relation over terms where a term of the form
+$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
+free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
+version of $t$. The implementation of the \elpiIn{comp} relation given in
+\cref{sec:compilation} compiles the \Fo{} terms $t_1 = $\elpiIn{flam x\ fapp
+[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
+$\elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
+\mapping{A}{A'}{1}. However, if the oracle sets \elpiIn{A'} to the constant
+\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
+fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
+fact that $t'_1 =$\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
+=$\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
+unification problem by adapting the \elpiIn{comp} relation such that it
+recognizes \maybeeta subterms $s$ and replaces them with fresh \Ho variables
+$v$. This link between the variable $v$ and the subterm $t$ is stored in what we
+call \linketa which is an object with the following type
+
+\elpiIn{type link-eta tm -> tm -> baselink}
+
+\noindent
+where, as sketched in \cref{sec:lang-spec}, the term on the left hand side
+(\lhs) is linked with its left counterpart (\rhs).
+
+\linketa are added in the link store (\linkStore) and activated when special
+conditions are satisfied on \lhs or \rhs. These link activations are managed by
+extending the \elpiIn{progress1} predicate (see \cref{sec:execution}).
+We claim that \linketa progression does not contradict \cref{inv:linklhs} and we
+add the following invariant:
+
+\begin{invariant}[\linketa \rhs]
+  The \rhs of a \linketa having the shape $\lambda x.F_x$
+  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
+  \label{invariant:link-eta-right}
+\end{invariant}
+
+In the next three subsections we explain how we detect \maybeeta terms, how we
+compile them and how \linketa are activated during the execution of the program
+and provide justification for why these two invariants remain true.
+
+\subsection{Detection of \maybeeta}
 
@@
 
-\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
-\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
-
 \noindent
@@
 called respectively \lhs and \rhs.
-
-\begin{invariant}[\linketa \lhs]
-  The \lhs of a \linketa is a variable.
-  \label{invariant:link-eta-left}
-\end{invariant}
-
-TODO: motivate this?
-
-\begin{invariant}[\linketa \rhs]
-  The \rhs of a \linketa having the shape $\lambda x.F \appsep x$
-  where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
-  \label{invariant:link-eta-right}
-\end{invariant}
-
-TODO: motivate this?
 
@@
 TODO: dire che scendiamo sotto i vari abs che formano il contesto? 
-
-We can note that the insertion of these rules for \elpiIn{progress1} do not
-prevent the termination of \elpiIn{progress}, since, if the link is woken up,
-the unification of the terms and the suppression of the link form \linkStore
-are terminating operation. Note that if the link remain suspend
-\elpiIn{progress} continues to terminate, since, it would mean that the
-condition of the \elpiIn{if}-branchement succeds and the \elpiIn{progress}
-terminates its execution.
-
-A second way to progress \linketa is to check if \linkStore contains two
-\linketa $l_1$ and $l_2$ with same \lhs. In this case we are allowed to unify
-
-TODO: dire che termina
+TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
+TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$
+
+A second way to progress \linketa, that we call \linketa deduplication, is to
+check if \linkStore contains two \linketa $l_1$ and $l_2$ with same \lhs. This
+situation occurs if two \maybeeta terms are unified with a same unification
+variable. In this case, we can unify the $l_1$ and $l_2$ \rhs (that, by
+construction are both on the form $\lambda x. T_x$) and remove one of the two
+links.
+
+TODO: example for this: $\lambda x. \lambda y. F \appsep y \appsep x = X,
+\lambda x. \lambda y. F \appsep y \appsep x = Y$
+
+We can note that the 1) insertion of these rules for \elpiIn{progress1} do not
+prevent the termination of \elpiIn{progress}, since, a link activation runs
+terminating operations (such as term unification and link-removal); 2) \linketa
+deduplication runs again terminating operations and 3) if none of these two 
+situations is performed, then the substitution and \linkStore
+Note that if the link remain suspended \elpiIn{progress} continues to terminate,
+since, it would mean that the condition of the \elpiIn{if}-branchement succeds
+and the \elpiIn{progress} terminates its execution.
 
@@
 to $T x$ where x is not free in $T$.
-
-\begin{textcode}
-  La deduplicate eta:
-  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2]
-    (a destra non c'è mai un termine con testa rigida)
-  - i due termini a dx vengono unificati con la unif e uno dei due link viene buttato
-    NOTA!! A dx abbiamo sempre un termine della forma lam x.VAR x!!!
-    Altrimenti il link sarebbe stato risolto!!
-  - dopo l'unificazione rimane un link [variabile] -> [etaX]
-  - nella progress-eta, se a sx abbiamo una constante o un'app, allora eta-espandiamo
-    di uno per poter unificare con il termine di dx.
-\end{textcode}
 

### 2024-03-27 paper/main.tex
eta add some def

@@
 
+Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
+for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
+given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
+\maybeeta detection is not a trivial operation as it may seems.
+%
+\begin{gather}
+  \lambda x. f \appsep A_x\\
+  \lambda x. f \appsep x \appsep A_x\\
+  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x}\\ \label{eq:lamlam}
+\end{gather}
+
+In the examples above, the first expression is a \maybeeta since $A_x$ can
+reduce to $x$, the second one is not a \maybeeta since for any substitution for
+$A_x$, $x$ is not free in $f \appsep x$. The third equation is a bit more
+complicated since, we have a vector of lambdas, this means that the whole term
+is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
+$t$, and $t$ can be reduced to a term on the form $t' x$ where $x$ is not free
+in $t'$. Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
+$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.
+
+As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
+since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
+subterm does not use $x$, for example if $A$ is a function on the form $\lambda
+x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
+consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
+the given subterm.
+
+We can now define more formally the two auxiliary relation we need for \maybeeta
+detection:
+
+\begin{definition}[reduce-to]
+  For any term $t$, $\lambda x_1 \ldots x_n.t_{x_0\ldots x_n}$ reduces to a bound
+  variable $x$ if one of the three following cases is satisfied: 1)
+  $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
+  of term $l$ and each $l_i$ reduces to $x_i$; 3) $t$ is a variable with scope
+  $s$, and for any $bv \in [x|\vec{x}]$, it exists a $s_i$, such that $s_i$
+  reduces to $bv$
+\end{definition}
+
+\begin{definition}[occurs-rigidly]
+  Given a term $t$, a bund variable ``rigidly'' occurs in \subterm{t} this
+  term does not appear in the scope of a variable
+\end{definition}
+
+TODO: maybe eta is over-approximation
+
 \subsection{Compilation}

### 2024-03-27 paper/main.tex
update maybeeta detection

@@
 $A_x$, $x$ is not free in $f \appsep x$. The third equation is a bit more
-complicated since, we have a vector of lambdas, this means that the whole term
+complicated since, we have a spine of lambdas, this means that the whole term
 is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
@@
 \begin{definition}[reduce-to]
-  For any term $t$, $\lambda x_1 \ldots x_n.t_{x_0\ldots x_n}$ reduces to a bound
-  variable $x$ if one of the three following cases is satisfied: 1)
-  $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
+  A term $t$ reduce to a name $x$, if $\exists \sigma, \sigma t = x$. In
+  particular, for any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$
+  reduces to a bound variable $x$ if one of the three following cases is
+  satisfied: 1) $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
   of term $l$ and each $l_i$ reduces to $x_i$; 3) $t$ is a variable with scope
@@
 \begin{definition}[occurs-rigidly]
-  Given a term $t$, a bund variable ``rigidly'' occurs in \subterm{t} this
-  term does not appear in the scope of a variable
+  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, \exists t' \in
+  \subterm{t}, t' = x$, i.e. $x$ does not appear in the scope of any variable
+  in $t$.
 \end{definition}
 
-TODO: maybe eta is over-approximation
+Finally, we can derive the implementation for \maybeeta detection:
+
+\begin{definition}[\maybeeta detection]
+  For any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$ is a \maybeeta if
+  1) either $t$ is the application of a term with argument $l_1 \ldots l_m$ such that 
+  $m \geq n$, every $l_{m-n+i}$ reduce to $x_i$, and no $l_{1\ldots m-n-1}$
+  reduces to a name in $\vec{x}$; 2) or $t$ is a variable with scope $s$ and
+  each $x_i$ it exists $s_j$ such that $s_j$ reduces $x_i$.
+\end{definition}
+
+As a final remark, the \maybeeta detection defined just before is an
+over-approximation, in the sense that there exists some terms $t$ considered as
+\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
+$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$. On
+the other hand, we also point out that the \maybeeta detection spot out
+potential $\eta$-expansion for terms that are not in \llambda. For example,
+$\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
+application of term whose argument can reduce to $x$.
+
+The implementation we propose for the \maybeeta relation is given below.
+
+\input{code/maybe_eta}
+
+% Here a complex maybeeta example
+% \begin{gather}
+%   T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
+% \end{gather}
 
@@
 property is given below:
-
-\input{code/maybe_eta}
 

### 2024-03-27 paper/main.tex
wip

@@
 \newtheorem{invariant}{Invariant}
+\crefname{invariant}{invariant}{invariants}
 
@@
 \begin{invariant}[Link left hand side]\label{inv:linklhs}
-  The left hand side of a new link
+  The left hand side of a suspended link
   is a variable.
@@
 \noindent
-If the variable is assigned during a run the link is considered for
-progress and possibly eliminated. This is discussed in
-section~\ref{sec:eta}.\marginpar{is this the right section?}
+New links are suspended by construction.
+If the left hand side variable is assigned during a step, then 
+the link is considered for progress and possibly eliminated.
+This is discussed in \cref{sec:eta} and \cref{sc:beta}.
 
@@
 
-TODO: discuss occur check
+Since compilation moves problematic terms out of the compiled terms,
+\Ue{} can only perform a partial occur check. For example the
+unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
+but should be disallowed if a $\linkStore$ contains a link like
+$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
+a lambda in head position, but we surely know it contains $X$, hence
+$f~Y$ and that triggers occur check.
+The procedure \elpiIn{occur-check-links} is in charge of ensuring that
+each link does not represent a (suspended) unification problem doomed
+to fail because of occur check. This check is needed in order to
+guarantee \cref{prop:fidelity} (simulation fidelity).
 
@@
   where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
-  \label{invariant:link-eta-right}
+  \label{inv:link-eta-right}
 \end{invariant}
@@
 compile them and how \linketa are activated during the execution of the program
-and provide justification for why these two invariants remain true.
+and provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.
 
@@
 
+TODO: clarify this
 As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
@@
 
-\paragraph{Detection of \maybeeta} The main modification of the compiler to solve this
-unification issue consists in identifying all the subterms of a term $t$ that
-are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
-$\eta$-expansion under a certain substitution $\sigma$. The code verifying this
-property is given below:
-
-The entry point is depicted by the rule~\ref{rule:maybeeta} which takes a name
-$n$, a term $t$ and a list of bound variables $L$ (originally it is the
-singleton containing $n$). This rule checks if $t$ is a term of the form $T
-\appsep n$ (for a term $T$), together with the auxiliary predicate
-\elpiIn{reducible-to} which ensures if a term $t$ can reduce to a name $n$.
-\marginpar{explain better: we start from $\lambda x.F x$}
-The \elpiIn{maybe-eta} predicate dispatches the calls to \elpiIn{reducible-to};
-three cases should be considered: 1) $t$ is a variable $v$, then $t$ can be an
-$\eta$-expansion if at least one of the terms in the scope of $v$ is a
-$\maybeeta$ of $n$; 2) $t$ is a lambda-term, then we recursively call
-\elpiIn{maybe-eta} on the body of $t$ under a local name $x$ which is added to
-the list $L$; 3) $t$ is an application, then $t$ is an $\eta$ expansion if i)
-the last arguments of $t$ can be reduced one by one to the binders in the list
-$L$ (we reverse the list in rule, since, by construction, this list is built in
-reversed order) and ii) none of the first arguments of the application contain a
-rigid occurrence of name in $L$.
-
-As rapidly said before, \elpiIn{reducible-to} tells if a term $t$ reduce to a
-name $n$, or equivalently if $\exists \sigma, \sigma t = n$. This predicate also
-takes the list of all the binders explored (this list is originally empty). A
-term $t$ reduces to a name $n$ if 1) $n = t$; 2) $t$ is a variable $v$, then $t$
-reduce to a $n$ if it exists an arugment in the scope of $v$ reducing to $n$ and
-forall name $n'$ in $L$, there is an argument reducing to $n'$ \marginpar{Note
-that this function is partial: the arg in the scope should be distinct}; 3) $t$
-is a lambda abstraction, then we call recursively \elpiIn{reducible-to} on the
-body of the abstraction with a new local name added to the list $L$; 4) $t$ is
-an application of $n$ to a list of arguments $L'$, then all the arguments should
-reduce to the respective name in the list $L$.
-
-Finally, a name $n$ occurs rigidly in a term $t$ if $n$ occurs in a subterm $t'$
-of $t$ such that $t'$ does not appear in the scope of a variable.
-
-An example of \maybeeta detection over the bound variable $x$ is the following:
-\marginpar{maybe more examples in appendix}
-%
-\begin{gather}
-  T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
-\end{gather}
-
-The correct call to \elpiIn{maybe-eta} is \elpiIn{maybe-eta x T L} with
-\elpiIn{L = [x]}. At first we go under the abstraction $\lambda y$ adding $y$ to
-$L$. Then we find an application, where we verify that 1) $A_{xy}$ does not
-contain $x$ and $y$ rigidly which is the case; 2) $B \appsep a \appsep (\lambda z.y\appsep (C_z))$
-and $D_x$ can respectively reduce to $y$ and $x$. The latter reduction is
-evident, since $D$ has $x$ in scope; the former subterm can reduce to $y$ since
-$B$ is a variable and it exists an argument ($\lambda z.y\appsep C_z$) reducible to
-$y$: under the binder $z$, we have the application of $y$ with a variable with
-$z$ in scope. Note that \maybeeta, only tells if it exists a substitution making
-a term an eta expansion on any term $t$, i.e, $t$ can be in $\maybeeta \cup
-\maybebeta$ without the constraint of being in \llambda, as our example shows. A
-possible substitution making the term $T$ in the example an $\eta$-expansion is
-$\sigma = \{A \mapsto \lambda x.\lambda y.a, B \mapsto \lambda x.\lambda y.y,
-C \mapsto \lambda x.x, D \mapsto \lambda x.x\}$.
-
-\paragraph{Compilation with \linketa} Thanks to the \elpiIn{maybe-eta}
-predicate, we can detect ``$\eta$-problematic'' terms and, consequently replace
-them with a fresh \Ho unification variable at compilation time. The code
-below illustrate this dedicated compilation:
-% TODO: maybe ref to \cref{sec:nutshell}
+Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
+terms and, consequently replace them with fresh \Ho unification variables at
+compilation time. The code below illustrate how this relation is used to for
+term compilation.
 
@@
 
-\marginpar{Where This rule is put in the code ?}
-This rule is applied on \Fo{} lambda-abstractions, tests if their body $F$ is an
-\maybeeta wrt to a local fresh binder $x$, and if so, it compiles $F$ to the \Ho
-term \elpiIn{F1} and returns a \Ho fresh variable $A$ having in scope the free
-names occurring in \elpiIn{F1}. As sketch at the very end of
-\cref{sec:lang-spec}, each time a subterm $t$ in \Fo is replaced with a \Ho
-variable $v$, we build a \elpiIn{link-eta} between $v$ and and the term $t'$
-obtained by the compilation of $t$.
-
-A \elpiIn{link-eta}\footnote{\elpiIn{@val-link-eta A B} is syntactic sugar for
-\elpiIn{val (link-eta A B)}}
-is defined as
-
-\elpiIn{type link-eta tm -> tm -> baselink}
-
-\noindent
-We call the two terms carried by the link respectively left and right hand side,
-called respectively \lhs and \rhs.
+This rule, to be inserted just before rule~\ref{rule:complam} from the code in
+\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
+\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
+variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
+in $t'$. The critical part of this compilation is the creation of the \linketa,
+which links the variable \elpiIn{A} with $t$. This link creation enforce
+\cref{inv:linklhs} and \cref{inv:link-eta-right}, since clearly the \lhs is a
+variable and the \rhs is a term starting with the \elpiIn{lam} constructor. We
+can also note that for any name $x$, $tx = t' \appsep x$ never start with
+\elpiIn{lam}, since by contradiction, if 1) $tx$ is an $\eta$-expansion, then
+the recursive call to \elpiIn{comp} would have replaced $tx$ with a fresh
+variable, which contradict the hypothesis and if 2) $t'$ is not an
+$\eta$-expansion, then neither $\lambda t'$ is.
 

### 2024-03-27 paper/main.tex
update compil of eta

@@
 \begin{invariant}[Link left hand side]\label{inv:linklhs}
-  The left hand side of a suspended link
+  The left hand side of a new link
   is a variable.
@@
 \noindent
-New links are suspended by construction.
-If the left hand side variable is assigned during a step, then 
-the link is considered for progress and possibly eliminated.
-This is discussed in \cref{sec:eta} and \cref{sc:beta}.
+If the variable is assigned during a run the link is considered for
+progress and possibly eliminated. This is discussed in
+section~\ref{sec:eta}.\marginpar{is this the right section?}
 
@@
 \noindent
-Note tha the infix notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
-\elpiIn{((~\Ue\!\!\!~) A B C D)}.
+Note tha the notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
+\elpiIn{((~\Ue~) A B C D)}.
 
@@
 
-Since compilation moves problematic terms out of the compiled terms,
-\Ue{} can only perform a partial occur check. For example the
-unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
-but should be disallowed if a $\linkStore$ contains a link like
-$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
-a lambda in head position, but we surely know it contains $X$, hence
-$f~Y$ and that triggers occur check.
-The procedure \elpiIn{occur-check-links} is in charge of ensuring that
-each link does not represent a (suspended) unification problem doomed
-to fail because of occur check. This check is needed in order to
-guarantee \cref{prop:fidelity} (simulation fidelity).
+TODO: discuss occur check
 

### 2024-03-27 paper/main.tex
6

--- 2024-03-27 paper/main.tex
@@
   satisfied: 1) $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
-  of term $l$ and each $l_i$ reduces to $x_i$; 3) $t$ is a variable with scope
-  $s$, and for any $bv \in [x|\vec{x}]$, it exists a $s_i$, such that $s_i$
-  reduces to $bv$
+  of terms $l$ and each $l_i$ reduces to $x_i$
+  (since $\lambda x_1 \ldots x_n.x_{x_1\ldots x_n} = x$) ; 3) $t$ is a unification
+  variable in \llambda with scope
+  $s$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
+  there exists a $s_i \in s$, such that $s_i$ reduces to $v$
 \end{definition}
@@
 \begin{definition}[occurs-rigidly]
-  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, \exists t' \in
-  \subterm{t}, t' = x$, i.e. $x$ does not appear in the scope of any variable
-  in $t$.
+  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, x \in
+  \subterm{\sigma t}$
 \end{definition}
+In other words $x$ \emph{occurs-rigidly} in $t$ if it occurs in $t$
+outside of the scope of unification variables since theirs instantiations
+are allowed to discard their scope.
 
@@
 \begin{definition}[\maybeeta detection]
-  For any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$ is a \maybeeta if
-  1) either $t$ is the application of a term with argument $l_1 \ldots l_m$ such that 
-  $m \geq n$, every $l_{m-n+i}$ reduce to $x_i$, and no $l_{1\ldots m-n-1}$
-  reduces to a name in $\vec{x}$; 2) or $t$ is a variable with scope $s$ and
-  each $x_i$ it exists $s_j$ such that $s_j$ reduces $x_i$.
+  A term $\lambda x_1 \ldots x_n.t$ where each $x_i$ occurs in $t$ is a
+  \maybeeta if
+  either: 1) $t$ is a constant applied to arguments
+  $l_1 \ldots l_m$ such that 
+  $m \geq n$ and every $l_{m-n+i}$ reduces-to $x_i$ and
+  no $x_i$ occurs-rigidly in $l_{1\ldots m-n-1}$; or 2) $t$ is a
+  unification variable with scope $s$ and
+  for each $x_i$ there exists a $s_j \in s$ such that $s_j$ reduces-to $x_i$.
 \end{definition}

### 2024-03-27 paper/main.tex
update tex

@@
 the one of \Ho.
-Both have the invariant that they are not cyclic, TODO explain.
-Other invariant: the terms in ho\_subst never contains eta and beta expansion
+Both have the invariant that they are not cyclic, TODO: explain.
+
+\begin{invariant}
+  Any term in the substitution of \Ho is in normal form.
+  TODO: move this to the right place
+  TODO: definire normal form, maybe in definition 2.7
+  \label{inv:subst-nf}
+\end{invariant}
+\begin{invariant}
+
+  Any variable in the substitution with arity \elpiIn{N} is always fully applied
+  TODO: move this to the right place
+  TODO: definire normal form, maybe in definition 2.7
+  \label{inv:fully-applied-vars}
+\end{invariant}
 
@@
 in the substitution for \Fo{} and finally decompiling all
-assignments. Note that \ref{inv:linklhs} and the
+assignments. Note that \cref{inv:linklhs} and the
 occur check allows us to update the subst.
@@
 \input{code/decompile}
+
+TODO: What is commit-links and complete-mapping?, maybe complete-mapping can be
+hidden in the code rendering?
 
@@
 lambdas. For aestetic purposes we also eta-contract the result
-(not needed since Fo equality can do that)
+(not needed since \Fo{} equality can do that)
 
@@
 \begin{gather}
-\lambda x y.F~y~x = \lambda x y.x\\
-\lambda x. f~(F~x)~x = f~(\lambda y.y)
+\lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
+\lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
 \end{gather}
@@
 
-\linketa are meant to suspend the unification of two terms, that is, if we have
-a \maybeeta term $t$ which should be unified with a term $t'$ we don't want to
-unify them with \Ue. As said before this would break unification but it can also
-introduce \maybeeta in the substitution of \Ho.
-
-In order to activate a \linketa, we need to implement new rules for the
-\elpiIn{progress1} predicate. There are two cases making a \linketa to progress,
-1) \lhs is instantiated to a rigid term, in this case \lhs is unified with \rhs.
-TODO: the right eta hand side is eta-expanbded if it is an app/con; 2) \rhs can
-be $\eta$-reduced to a term with rigid head, in this case \lhs and \rhs are
-again unified. If one of these two condition is satisfied, the link has
-fulfilled is task and can be removed from the list of suspended links; if none
-the condition succeeds, the link is kept for a further iteration of
-\elpiIn{progress}.
-
-TODO: dire che scendiamo sotto i vari abs che formano il contesto? 
+\linketa are meant to delay the unification of ``problematic'' terms. In the
+following, we call \linkStore the list of suspended links.
+
+In order to activate a \linketa $l$, we need to implement new rules for the
+\elpiIn{progress1} predicate. After passing under all the \elpiIn{abs}
+constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
+is instantiated to a rigid term 2) \rhs can be $\eta$-reduced to a term with
+rigid head. If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
+we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
+if $t$ is a constant or a function application, then, $t'$, which by
+construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
+therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
+of $t$) with $t'$. Finally, if \lhs starts has \elpiIn{lam} as constructor, then
+$t$ is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.
+
+The second way to activate a \linketa is when the \rhs can $\eta$-reduced to a
+term $t$ with rigid head, i.e. $t'$ is not a \maybeeta. This means that we can
+unify \lhs with $t'$.
+
+Once a \linketa is activated, it can be removed from \linkStore,
+otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
+that this link progression enforce
+\cref{inv:subst-nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
+the \Ho substitution, since we make unification only when we know that the 
+terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
+no more a \maybeeta, the link is removed from \linkStore.
+
 TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
@@
 
-A second way to progress \linketa, that we call \linketa deduplication, is to
-check if \linkStore contains two \linketa $l_1$ and $l_2$ with same \lhs. This
-situation occurs if two \maybeeta terms are unified with a same unification
-variable. In this case, we can unify the $l_1$ and $l_2$ \rhs (that, by
-construction are both on the form $\lambda x. T_x$) and remove one of the two
-links.
+A second way to progress \linketa, that we call \linketa deduplication, is when
+\linkStore contains two \linketa $l_1$ and $l_2$ with a \lhs having the same
+variable address. Let the \lhs of $l_1$ be \elpiIn{uva U R} and the \lhs of
+$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:fully-applied-vars}, \elpiIn{R}
+and \elpiIn{S} have same scope. Let $t$ be the term obtained by replacing all
+each name \elpiIn{S}$_i$ in the \rhs of $l_1$ with \elpiIn{R}$_i$, $t$ is
+unified with the \rhs of $l_2$ and one of the two links between $l_1$ and $l_2$
+is removed from \linkStore.
 
@@
 
-We can note that the 1) insertion of these rules for \elpiIn{progress1} do not
-prevent the termination of \elpiIn{progress}, since, a link activation runs
-terminating operations (such as term unification and link-removal); 2) \linketa
-deduplication runs again terminating operations and 3) if none of these two 
-situations is performed, then the substitution and \linkStore
-Note that if the link remain suspended \elpiIn{progress} continues to terminate,
-since, it would mean that the condition of the \elpiIn{if}-branchement succeds
-and the \elpiIn{progress} terminates its execution.
-
-\begin{lemma}
-  We never add eta-expansions in the substitution
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-\end{proof}
-
-TODO: we can however have $\lambda x. F_x$ if we know that $F$ does not reduce
-to $T x$ where x is not free in $T$.
+We can note that the insertion of these rules for \elpiIn{progress1} do not
+prevent the termination of \elpiIn{progress}, since: if a link is activated it
+is removed from \linkStore and the recursive call to \elpiIn{progress} will have
+a smaller list of links to recurse on. Moreover, link activation only runs
+unification which are terminating instructions. If a link is deduplicated, the
+termination of \elpiIn{progress} is still guaranteed since again we reduce
+\linkStore and the instructions run by link deduplications are all terminating.
+Finally, if a link is neither activated nor deduplicated, i.e. it remains
+suspended, then the condition of the \elpiIn{if}-statement of \elpiIn{progress}
+succeeds making it to terminate.
+
+TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
+not reduce to $T x$ where x is not free in $T$.
+
+\subsection{Example}
+
+In this subsection we show two complete runs of the extended version of \Ue.
+
+\paragraph{Example 1}
+\begin{center}
+  \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
+\end{center}
+% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
+\begin{itemize}
+  \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
+        $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
+        c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
+  \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
+  \item This unification wakes up the second link, since its \lhs is now
+        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
+        \lambda c1.c0$
+  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
+        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
+        pruning.
+  \item All the links have been solved and the decompilation, assigns the \Fo{}
+        variable \elpiIn{A} to $\lambda xy.y$ 
+\end{itemize}
+
+\paragraph{Example 2}
+\begin{center}
+  \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
+\end{center}
+% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
+\begin{itemize}
+  \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
+        $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
+        $\mapping{C0}{X1}{1}$
+  \item After unification $X0$ is set to $f~\lambda c0.c0$
+  \item This unification wakes up the link, since its \lhs is now
+        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
+        \lambda c1.c0$
+  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
+        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
+        pruning.
+  \item All the links have been solved and the decompilation, assigns the \Fo{}
+        variable \elpiIn{A} to $\lambda xy.y$ 
+\end{itemize}
 

### 2024-03-28 paper/main.tex
tex updates

@@
 \end{invariant}
-\begin{invariant}
-
-  Any variable in the substitution with arity \elpiIn{N} is always fully applied
-  TODO: move this to the right place
-  TODO: definire normal form, maybe in definition 2.7
-  \label{inv:fully-applied-vars}
-\end{invariant}
 
@@
 
-\begin{invariant}[Unification variable arity]\label{inv:uvaarity}
+\begin{invariant}[Unification variable arity]
   Each variable \elpiIn{A}
@@
   \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
-\label{invariant:arity}
+  \label{inv:uvaarity}
 \end{invariant}
@@
 no \elpiIn{fapp} node should appear).
+
+\begin{invariant}
+
+  TODO: dire che il mapping è bijective
+  \label{inv:map-bijective}
+\end{invariant}
 
@@
 \subsection{Limitations of by this basic scheme}
+\label{sec:basic-comp-limitations}
 
@@
   \lambda x. f \appsep x \appsep A_x\\
-  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x}\\ \label{eq:lamlam}
+  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}
 \end{gather}
@@
 variable address. Let the \lhs of $l_1$ be \elpiIn{uva U R} and the \lhs of
-$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:fully-applied-vars}, \elpiIn{R}
+$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:uvaarity}, \elpiIn{R}
 and \elpiIn{S} have same scope. Let $t$ be the term obtained by replacing all
@@
 a smaller list of links to recurse on. Moreover, link activation only runs
-unification which are terminating instructions. If a link is deduplicated, the
+terminating instructions (such as unification). If a link is deduplicated, the
 termination of \elpiIn{progress} is still guaranteed since again we reduce
@@
 
-\subsection{Example}
-
-In this subsection we show two complete runs of the extended version of \Ue.
-
-\paragraph{Example 1}
-\begin{center}
-  \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
-\end{center}
-% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
-\begin{itemize}
-  \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
-        $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
-        c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
-  \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
-  \item This unification wakes up the second link, since its \lhs is now
-        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
-        \lambda c1.c0$
-  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
-        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
-        pruning.
-  \item All the links have been solved and the decompilation, assigns the \Fo{}
-        variable \elpiIn{A} to $\lambda xy.y$ 
-\end{itemize}
-
-\paragraph{Example 2}
-\begin{center}
-  \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
-\end{center}
-% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
-\begin{itemize}
-  \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
-        $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
-        $\mapping{C0}{X1}{1}$
-  \item After unification $X0$ is set to $f~\lambda c0.c0$
-  \item This unification wakes up the link, since its \lhs is now
-        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
-        \lambda c1.c0$
-  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
-        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
-        pruning.
-  \item All the links have been solved and the decompilation, assigns the \Fo{}
-        variable \elpiIn{A} to $\lambda xy.y$ 
-\end{itemize}
+% \subsection{Example}
+
+% In this subsection we show two complete runs of the extended version of \Ue.
+
+% \paragraph{Example 1}
+% \begin{center}
+%   \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
+% \end{center}
+% % $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
+% \begin{itemize}
+%   \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
+%         $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
+%         c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
+%   \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
+%   \item This unification wakes up the second link, since its \lhs is now
+%         instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
+%         \lambda c1.c0$
+%   \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
+%         unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
+%         pruning.
+%   \item All the links have been solved and the decompilation, assigns the \Fo{}
+%         variable \elpiIn{A} to $\lambda xy.y$ 
+% \end{itemize}
+
+% \paragraph{Example 2}
+% \begin{center}
+%   \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
+% \end{center}
+% % $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
+% \begin{itemize}
+%   \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
+%         $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
+%         $\mapping{C0}{X1}{1}$
+%   \item After unification $X0$ is set to $f~\lambda c0.c0$
+%   \item This unification wakes up the link, since its \lhs is now
+%         instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
+%         \lambda c1.c0$
+%   \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
+%         unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
+%         pruning.
+%   \item All the links have been solved and the decompilation, assigns the \Fo{}
+%         variable \elpiIn{A} to $\lambda xy.y$ 
+% \end{itemize}
 
@@
 \label{sec:invariant1}
+
+In \cref{sec:basic-comp-limitations}, we have given two unification problems to
+be run one after the other. The compilation of each term would generate the
+following result:
+
+\begin{description}
+  \item[Unif problems:] $X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)$
+  \item[Mappings:] $\mapping{F}{G}{1}, \mapping{F}{H}{2}$  
+  \item[Links:] $\linketaM{x}{Y_{x}}{\lambda y.H_{y~x}}$,\\
+                $\linketaM{}{X}{\lambda x.Y_{x}}$,\\
+                $\linketaM{}{Z}{\lambda x.f \appsep G_{x}~x}$
+\end{description}
+
+We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
+x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
+respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
+has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
+is linked to the term $\lambda x.f \appsep G_{x}~x$. However, we can remark that
+this compilation breaks \cref{inv:map-bijective}, since
+
 Deduplicate mapping code etc...

### 2024-03-28 paper/main.tex
wip

--- 2024-03-27 paper/main.tex
@@
 in the second problem the left hand side happens to be an
-eta expansion (of $f (\lambda y.y)$) only after we discover that
-$F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
+eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
+that $F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
 Both problems are addressed in the next section.
@@
   of terms $l$ and each $l_i$ reduces to $x_i$
-  (since $\lambda x_1 \ldots x_n.x_{x_1\ldots x_n} = x$) ; 3) $t$ is a unification
+  (e.g. $\lambda x_1 \ldots x_n.x_{x_1\ldots x_n} = x$) ; 3) $t$ is a unification
   variable in \llambda with scope

### 2024-03-28 paper/main.tex
w
x

@@
 
-The equational theory of the meta language Elpi that we want to use to
-implement a form of proof automation is strikingly similar, since it
-it comprises $\eta\beta$ (for the meta language functions), and the
+The equational theory of the meta language Elpi is strikingly similar,
+since it it comprises $\eta\beta$ (for the meta language functions), and the
 unification procedure \Ue{} solves higher-order problems in
@@
 \quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \progress~(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
+    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
 \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
@@
 \newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
-\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
+\newcommand{\maybebeta}{\ensuremath{\overline{\llambda}}\xspace}
+%\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
 \begin{definition}[\maybeeta]
@@
 an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
-that makes the resulting term fall outside of \maybebeta.
+that makes the resulting term fall back in \llambda.
 
@@
 $$
+\begin{array}{l}
 \wellb(\sigma\mathcal{T}) \land
-\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma'}
+  \sigma\mathcal{T}_{p_l} \Ue \sigma\mathcal{T}_{p_r} \mapsto {\sigma'}
+  \Rightarrow \wellb(\sigma' \mathcal{T})\\
+\wellb(\sigma\mathcal{T}) \land
+  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
   \Rightarrow \wellb(\sigma' \mathcal{T})
+\end{array}
 $$
@@
 \noindent
-A less formal way to state \ref{prop:nf} is that \hstep{} never
-``commits'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
+A less formal way to state \ref{prop:nf} is that \hstep{} and \progress never
+``commit'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
 that could be erased by an $\eta$-contraction),
-nor puts in $\sigma$ a flexible application outside \llambda{} (an application
-node that could be erased by a $\beta$-reduction).
-
-Note that proposition \ref{prop:nf} does not hold for \Uo{} since
-decompilation can introduce (actually restore) terms in
-\maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
-during compilation.
+nor put in $\sigma$ a flexible application outside \llambda{}
+(an application node that could be erased by a $\beta$-reduction).
+
+Note that proposition \ref{prop:nf} does not hold for \Uo{} as a whole
+since decompilation can introduce (actually restore) terms in
+\maybeeta or \maybebeta that were move out of the way
+(put in $\linkStore$) during compilation.
 
@@
 \noindent
-The \elpiIn{name} builtin predicate tests if a term is a bound variable.
-\footnote{one could always load name x for every x under a pi and get rid of the name builtin}
-
 In both languages unification variables are identified by a natural number
@@
 To accommodate this situation the compiler wraps \elpiIn{baselink} using
-the \elpiIn{inctx} container.
+the \elpiIn{inctx} container (see, \ref{data:inctx}).
 

### 2024-03-28 paper/main.tex
update tex

@@
 \newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
+\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
 
@@
 \runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \mathcal{T} \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
   \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
-  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
+  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
 \end{array}
@@
 
-\section{Enforcing Invariant~\ref{invariant:arity}}
+\section{Enforcing Invariant~\ref{inv:uvaarity}}
 \label{sec:invariant1}
@@
 has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
-is linked to the term $\lambda x.f \appsep G_{x}~x$. However, we can remark that
-this compilation breaks \cref{inv:map-bijective}, since
-
-Deduplicate mapping code etc...
+is linked to the term $\lambda x.f \appsep G_{x}~x$. However, the mapping
+returned by the compilation, does not breaks \cref{inv:map-bijective}: the \Fo{}
+variable $F$ is mapped to two different \Ho variables. To address this problem
+and enforce \cref{inv:map-bijective}, we clean the mapping with a second phase
+after the compilation. This phase is called \elpiIn{map-deduplication}.
+
+\newcommand{\cv}{\ensuremath{\mathrm{cv}}}
+\newcommand{\ev}{\ensuremath{\mathrm{ev}}}
+\newcommand{\ar}{\ensuremath{\mathrm{ar}}}
+
+Before formally defining this procedure, we need to define some auxiliary
+relations. Let \mapStore be the list of mapping and let $m \in \mapStore$, such
+that $m = \mapping{A}{\textbf{A}}{n}$, we define $\cv, \ev, ar$ such that
+$\cv(m) = A$, $\ev(m) = \textbf{A}$ and $\ar(m) = n$. Let $\langle m_1, m_2
+\rangle \in \mapStore$ such that $\ar(m_1) < \ar(m_2)$ and $n = ar(m_1) -
+ar(m_2) - 1$. We also let $A^i$ be a fresh \Ho variable. We define the
+\elpiIn{make-eta-link} relation taking two mappings $\langle m_1, m_2 \rangle$
+and returning  \linketa such that:
+
+$$ 
+  \forall i \in [0 \dots n], \left\{ 
+    \begin{array}{ll}
+      \linketaM{\phantom{\Gamma:x_0\dots x_{i-1}}}
+              {\ev(m_1)}
+              {\lambda x.A^0_x} & if~ i = 0 \\
+      \linketaM{\Gamma:x_0\dots x_{i-1}}
+               {A^i_{\Gamma}}
+               {\lambda x.A^{i+1}_{\Gamma\cup x}} & if~ 0 < i < n\\
+      \linketaM{\Gamma:x_0\dots x_{i}\phantom{_{-1}}}
+               {A^{i-1}_{\Gamma}}
+               {\lambda x.\ev(m_2)_{\Gamma\cup x}} & if~ i = n\\
+    \end{array} 
+  \right. 
+$$
+
+More concretely, we are saying that for any two mappings, we build as many
+\linketa as the difference of the arities between the two mappings. This links
+are constructed in such a way that the \Ho variable $v$ with lowest arity is
+linked to a fresh variable eta-expended variable $A^0$ having the scope of $v$.
+This variable $A^0$ is then linked to a fresh variable $A^{1}$ with same scope
+of $A^0$ and so on. The last link is built between the $A^{n-1}$ (where $n$ is
+the difference of arities between the two mappings) and the \Ho variable $u$
+with higher arity in the two mappings being considered.
+
+\begin{definition}[map-deduplication]
+  Forall mappings $\langle m_1, m_2 \rangle \in \mapStore$, such that $\cv(m_1) =
+  \cv(m_2)$, the list of \linketa $L$ is created thanks to
+  \elpiIn{make-eta-link}$m_1~m_2~L$ and is added to \linkStore. Then $m_1$ is
+  removed from \mapStore.
+\end{definition}
+
+If we take back the example give at the beginning of this section, we can
+deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
+and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}.
+
+The complete problem to run for resolution is now:
+
+\begin{description}
+  \item[Unif problems:] $X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)$
+  \item[Mappings:] $\mapping{F}{H}{2}$  
+  \item[Links:] $l_1 = \linketaM{x}{Y_{x}}{\lambda y.H_{y~x}}$,\\
+                $l_2 = \linketaM{}{X}{\lambda x.Y_{x}}$,\\
+                $l_3 = \linketaM{}{Z}{\lambda x.f \appsep G_{x}~x}$\\
+                $l_4 = \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}$
+\end{description}
+
+After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
+This assignment makes $l_2$ to progress since the \lhs is materialized and by
+unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
+y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
+After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
+\elpiIn{progress} fixpoint terminates. Next, the second unification problem is
+run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
+and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
+$Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
+As last step, we can note that $G$
+
+TODO: dire che preserviamo l'invariante che tutte le variable sono fully-applied
 

### 2024-03-28 paper/main.tex
tex: end example

@@
 $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
-As last step, we can note that $G$
+As last step, the last link is progressed and the final \Ho substitution is
+$\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
+G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
+H _{x} \mapsto \lambda y.y\}$.
+
+The decompilation phase is only charged, in this example to solve the mappings,
+since no suspended links remain. The only mapping in the list is
+\mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$
 

### 2024-03-28 paper/main.tex
wip

--- 2024-03-28 paper/main.tex
@@
 \noindent
-In both languages unification variables are identified by a natural number
-representing a memory address.
+Natural numbers represent the memory addresses that
+identify unification variables in both languages.
 The memory and its associated operations are described below:
@@
   $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
+  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
   $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
-  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
   $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
@@
 \noindent
-When detailing examples we write links as equations between terms under a context.
+When detailing examples we write links as equations between two
+terms under a context.
 The equality sign is subscripted with
-kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
-\begin{elpicode}
-abs x\ val (link-beta (uva A []) (app[uva F [x],con "a"]))
+kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A_x}{F_x~a}$ corresponds to:
+\begin{elpicode}
+abs x\ val (link-beta (uva A [x]) (app[uva F [x],con "a"]))
 \end{elpicode}
@@
 
-The first four rules are common to both equalities and correspond to $\alpha$-equivalence.
+The first four rules are common to both equalities
+and just define the usual congruence over terms, and since
+we use an HOAS encoding they also capture $\alpha$-equivalence.
 In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
@@
 and to identify the four rules that need special treatment in
-the implementation of \Ue.
+the implementation of \Uo.
 
@@
 The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
-variables). Elpi provides it as a builtin, but one could implement it by
+variables).\footnote{Elpi provides it as a builtin, but one could implement it by
 systematically loading the hypothetical rule \elpiIn{name x} every time
-a nominal constant is postulated via \elpiIn{pi x\ }.
-
-\paragraph{Substitution application: $\rho s$ and $\sigma t$}
-
-Applying the substitution corresponds to dereferencing a term with respect to
-the memory. To ease the comparison we split \Fo dereferencing into a
-\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
-to memory cells that are set with their values, ans has a corresponding
-operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp},
-in charge of ``flattening'' \elpiIn{fapp} nodes, has no corresponding
-operation in \Ho. The reasons for this asymmetry is that an \elpiIn{fapp}
-node with a flexible head is always mapped to a \elpiIn{uva} (as per
-sections \ref{sec:compilation,sec:beta}), preventing nested applications to
-materialize.
-
-\input{code/fderef}
-
-\noindent
+a nominal constant is postulated via \elpiInFN{pi x\ }}
+The choice of using n-ary application, rather than binary, is to make it
+easy to access the application's head. The price we pay is that substituting
+an application in the head of an application should be amended by
+``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
+\footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
+we do in this presentation and, as we explain later,
+no equivalent of \elpiIn{napp} is needed in \Ho.}
+Finally note
 Note that the cut operator is inessential, it could be
@@
 
+\paragraph{Substitution application: $\rho s$ and $\sigma t$}
+
+Applying the substitution corresponds to dereferencing a term with respect to
+the memory. To ease the comparison we split \Fo dereferencing into a
+\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
+to memory cells that are set with their values, and has a corresponding
+operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
+has no corresponding operation in \Ho. The reasons for this asymmetry is
+that an \elpiIn{fapp} node with a flexible head is always mapped
+to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
+preventing nested applications to materialize.
+
+\input{code/fderef}
+
+\noindent
 Applying the substitution in \Ho{} is very similar, with

### 2024-03-28 paper/main.tex

--- 2024-03-28 paper/main.tex
@@
 \noindent
-When detailing examples we write links as equations between terms under a context.
+When detailing examples we write links as equations between two
+terms under a context.
 The equality sign is subscripted with
-kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
+kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A_x}{F_x~a}$ corresponds to:
 \begin{elpicode}
-abs x\ val (link-beta (uva A []) (app[uva F [x],con "a"]))
+abs x\ val (link-beta (uva A [x]) (app[uva F [x],con "a"]))
 \end{elpicode}

### 2024-03-28 paper/main.tex
deduplicate-map: better formula

--- 2024-03-28 paper/main.tex
@@
 
-\newcommand{\cv}{\ensuremath{\mathrm{cv}}}
-\newcommand{\ev}{\ensuremath{\mathrm{ev}}}
-\newcommand{\ar}{\ensuremath{\mathrm{ar}}}
-
 Before formally defining this procedure, we need to define some auxiliary
-relations. Let \mapStore be the list of mapping and let $m \in \mapStore$, such
-that $m = \mapping{A}{\textbf{A}}{n}$, we define $\cv, \ev, ar$ such that
-$\cv(m) = A$, $\ev(m) = \textbf{A}$ and $\ar(m) = n$. Let $\langle m_1, m_2
-\rangle \in \mapStore$ such that $\ar(m_1) < \ar(m_2)$ and $n = ar(m_1) -
-ar(m_2) - 1$. We also let $A^i$ be a fresh \Ho variable. We define the
-\elpiIn{make-eta-link} relation taking two mappings $\langle m_1, m_2 \rangle$
-and returning  \linketa such that:
-
+relations. Let \mapStore be the list of mapping and $\langle m_1, m_2 \rangle
+\in \mapStore$ such that the arity of the \Ho variable in $m_1$ is smaller than
+the one in $m_2$. Let $X$ (resp. $Y$) the \Ho variable of $m_1$ (resp. $m_2$)
+and $n = ar(m_1) - ar(m_2)$. We also let $A^i$ be a fresh \Ho variable. We
+define the \elpiIn{make-eta-link} relation taking two mappings $\langle m_1, m_2
+\rangle$ and returning the following list of link \linketa: $\forall i \in [1..n]$,
+{
+\def\scopeA{\ensuremath{x_1\dots x_{i-1}}}
+\def\scopeB{\ensuremath{x_1\dots x_{i}}}
+%
 $$ 
-  \forall i \in [0 \dots n], \left\{ 
+  \left\{ 
     \begin{array}{ll}
-      \linketaM{\phantom{\Gamma:x_0\dots x_{i-1}}}
-              {\ev(m_1)}
-              {\lambda x.A^0_x} & if~ i = 0 \\
-      \linketaM{\Gamma:x_0\dots x_{i-1}}
-               {A^i_{\Gamma}}
-               {\lambda x.A^{i+1}_{\Gamma\cup x}} & if~ 0 < i < n\\
-      \linketaM{\Gamma:x_0\dots x_{i}\phantom{_{-1}}}
-               {A^{i-1}_{\Gamma}}
-               {\lambda x.\ev(m_2)_{\Gamma\cup x}} & if~ i = n\\
+      \linketaM{\phantom{\scopeA}}{X}      {\lambda x.  A^1_x}         & if~ i = 1 \\
+      \linketaM{\scopeA}{A^{i-1}_{\scopeA}}{\lambda x_i.A^i_{\scopeB}} & if~ 1 < i < n\\
+      \linketaM{\scopeA}{A^{i-1}_{\scopeA}}{\lambda x_i.Y  _{\scopeB}} & if~ i = n
     \end{array} 
@@
 $$
-
+}
+\noindent
 More concretely, we are saying that for any two mappings, we build as many
@@
 are constructed in such a way that the \Ho variable $v$ with lowest arity is
-linked to a fresh variable eta-expended variable $A^0$ having the scope of $v$.
-This variable $A^0$ is then linked to a fresh variable $A^{1}$ with same scope
-of $A^0$ and so on. The last link is built between the $A^{n-1}$ (where $n$ is
-the difference of arities between the two mappings) and the \Ho variable $u$
-with higher arity in the two mappings being considered.
+linked to a fresh variable eta-expended variable $A^1$ having the scope of $v$.
+This variable $A^1$ is then linked to a an $\eta$-expanded fresh variable $A^2$
+with same scope of $A^1$ and so on. The last link is built between the $A^{n-1}$
+(where $n$ is the difference of arities between the two mappings) and the \Ho
+variable $u$ with higher arity in the two mappings being considered.
 
 \begin{definition}[map-deduplication]
-  Forall mappings $\langle m_1, m_2 \rangle \in \mapStore$, such that $\cv(m_1) =
-  \cv(m_2)$, the list of \linketa $L$ is created thanks to
+  Forall mappings $\langle m_1, m_2 \rangle \in \mapStore$, sharing the same
+  \Fo{} variable, the list of \linketa $L$ is created thanks to
   \elpiIn{make-eta-link}$m_1~m_2~L$ and is added to \linkStore. Then $m_1$ is

### 2024-03-28 paper/main.tex
minor

@@
 no equivalent of \elpiIn{napp} is needed in \Ho.}
-Finally note
-Note that the cut operator is inessential, it could be
+Finally note that the cut operator is inessential, it could be
 removed at the cost of a verbose test on the head of \elpiIn{L}
-in the last rule (\elpiIn{L} head
-can be \elpiIn{fcon}, \elpiIn{flam} or a name).
+in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
+can be \elpiIn{fcon}, \elpiIn{flam} or a name.
 
@@
 links (suspended unification problems) that play no role in this section
-but play a major role in Sections \ref{sec:eta} and \ref{sec:beta}.
-With respect to \ref{sec:problem-statement} the signature also allows
-for updates to the substitution.  The code below only allocates space
-for the variables, i.e. sets their memory address to \elpiIn{none},
-a details not worth mentioning in the previous discussion.
+but play a major role in \cref{sec:eta} and \cref{sec:beta}.
+With respect to \cref{sec:problem-statement} the signature also allows
+for updates to the substitution.  The code below uses that possibility
+in order to allocate space for the variables, i.e. sets their memory
+address to \elpiIn{none} (a details not worth mentioning in the
+previous discussion).
 
@@
 This preliminary version of \elpiIn{comp} recognizes \Fo{} variables
-applied to a (possibly empty) duplicate free list of names (i.e.
-\elpiIn{pattern-fragment} detects variables in \llambda).
-Note tha compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
+applied to a (possibly empty) duplicate free list of names.
+Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
 is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
@@
 \noindent
-Note that we could remove the second rule, whose purpose is to make
-links more readable by pruning unneeded abstractions (unused context entries).
+Note that we could remove the first rule, whose solve purpose is to make
+links more readable by pruning unused context entries.
 
@@
 
-Since compilation moves problematic terms out of the compiled terms,
-\Ue{} can only perform a partial occur check. For example the
+Since compilation moves problematic terms out of the sigh of \Ue{},
+that procedure can only perform a partial occur check. For example the
 unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
@@
 a lambda in head position, but we surely know it contains $X$, hence
-$f~Y$ and that triggers occur check.
+$f~Y$ and that fails the occur check.
 The procedure \elpiIn{occur-check-links} is in charge of ensuring that

### 2024-03-28 paper/main.tex
rewrite handling eta

@@
 \newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
-
+%
 $$
@@
 version of $t$. The implementation of the \elpiIn{comp} relation given in
-\cref{sec:compilation} compiles the \Fo{} terms $t_1 = $\elpiIn{flam x\ fapp
+\cref{sec:compilation} compiles the \Fo{} terms $t_1 =$ \elpiIn{flam x\ fapp
 [fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
-$\elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
-\mapping{A}{A'}{1}. However, if the oracle sets \elpiIn{A'} to the constant
+$ \elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
+\mapping{\elpiIn{A}}{\elpiIn{A'}}{1}. However, if the oracle sets \elpiIn{A'} to the constant
 \elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
 fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
-fact that $t'_1 =$\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
-=$\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
+fact that $t'_1 =$~\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
+=$~\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
 unification problem by adapting the \elpiIn{comp} relation such that it
-recognizes \maybeeta subterms $s$ and replaces them with fresh \Ho variables
-$v$. This link between the variable $v$ and the subterm $t$ is stored in what we
+recognizes \maybeeta subterms and replaces them with fresh \Ho variables. 
+This link between the variable $v$ and the subterm $t$ is stored in what we
 call \linketa which is an object with the following type
@@
 \maybeeta detection is not a trivial operation as it may seems.
-%
+\\
+\noindent
+\begin{minipage}{.5\linewidth}
   \begin{gather}
     \lambda x. f \appsep A_x\\
-  \lambda x. f \appsep x \appsep A_x\\
-  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}
+    \lambda x. f \appsep x \appsep A_x
   \end{gather}
+\end{minipage}% This must go next to `\end{minipage}`
+\begin{minipage}{.5\linewidth}
+  \begin{gather}
+    \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
+    \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
+  \end{gather}
+\end{minipage}
 
 In the examples above, the first expression is a \maybeeta since $A_x$ can
-reduce to $x$, the second one is not a \maybeeta since for any substitution for
-$A_x$, $x$ is not free in $f \appsep x$. The third equation is a bit more
+reduce to $x$, the second one is not a \maybeeta since it exists no substitution
+for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
+Eq.~\ref{eq:lamlam} is a bit more
 complicated since, we have a spine of lambdas, this means that the whole term
@@
 
-TODO: clarify this
-As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
-since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
-subterm does not use $x$, for example if $A$ is a function on the form $\lambda
-x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
-consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
-the given subterm.
-
-We can now define more formally the two auxiliary relation we need for \maybeeta
+As a final remark, note that eq.~\ref{eq:lamdisc} is a \maybeeta even if $x$
+occurs in the subterm $t' = f \appsep A_x$. It exists in fact a substitution
+$\sigma = \{A \mapsto \lambda x.a\}$, for any constant $a$, such that $\sigma t'
+= f \appsep a$ is a term where $x$ does not appears as a free variable.
+
+We can now define more formally the two auxiliary relations needed for \maybeeta
 detection:
@@
 deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
-and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}.
+and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x y}}.
 

### 2024-03-28 paper/main.tex
update proof for progress termination

@@
 
-As a final remark, the \maybeeta detection defined just before is an
+Finally, the \maybeeta detection defined just before is an
 over-approximation, in the sense that there exists some terms $t$ considered as
 \maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
-$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$. On
-the other hand, we also point out that the \maybeeta detection spot out
-potential $\eta$-expansion for terms that are not in \llambda. For example,
-$\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
-application of term whose argument can reduce to $x$.
+$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$.
 
@@
 
-We can note that the insertion of these rules for \elpiIn{progress1} do not
-prevent the termination of \elpiIn{progress}, since: if a link is activated it
-is removed from \linkStore and the recursive call to \elpiIn{progress} will have
-a smaller list of links to recurse on. Moreover, link activation only runs
-terminating instructions (such as unification). If a link is deduplicated, the
+\begin{lemma}
+  Forall list of links \elpiIn{L} and \elpiIn{S}, \elpiIn{progress L _ S _}
+  terminates
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  The addition of rules for \elpiIn{progres1} complicates the function
+  \elpiIn{progress}. We can note, however, that they do not prevent the
+  termination of \elpiIn{progress}. 1) If a link is activated it is removed from
+  \linkStore and the recursive call to \elpiIn{progress} will have a smaller
+  list of links to recurse on. Moreover, link activation only runs terminating
+  instructions (such as unification). 2) If a link is deduplicated, the
   termination of \elpiIn{progress} is still guaranteed since again we reduce
-\linkStore and the instructions run by link deduplications are all terminating.
-Finally, if a link is neither activated nor deduplicated, i.e. it remains
-suspended, then the condition of the \elpiIn{if}-statement of \elpiIn{progress}
-succeeds making it to terminate.
+  \linkStore and the instructions run by link deduplications are all
+  terminating. 3) If a link is neither activated nor deduplicated, i.e. it
+  remains suspended, then \linkStore remains unchanged like the substitution;
+  therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
+  terminates.
+\end{proof}
 
@@
 
-$\beta$-reduction problems (\maybebeta) appears any time we deal with a subterm $t
-= X t_1 \dots t_n$, where $X$ is flexible and the list $[t_1 \dots t_n]$ in not
-in \llambda. This unification problem is not solvable without loss of
-generality, since there is not a most general unifier. If we take back the
-example given in \cref{sec:nutshell}, the unification $F a = a$ admits two solutions for $F$:
-$\rho_1 = \{F \mapsto \lambda x.x\}$ and $\rho_2 = \{F \mapsto \lambda \_.a\}$.
-Despite this, it is possible to work with $\maybebeta$ if an oracle provides a
-substitution $\rho$ such that $\rho t$ falls again in the \llambda.
+TODO: say that maybe-eta also work in not(llambda)
+
+% On
+% the other hand, we also point out that the \maybeeta detection spot out
+% potential $\eta$-expansion for terms that are not in \llambda. For example,
+% $\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
+% application of term whose argument can reduce to $x$.
+
+All the previous sections we have dealt with terms in \llambda, however, it is
+often possible to work with terms in \maybebeta and wish to unify them. There
+are situtation, for example, where the oracle has given  which $\beta$-reduction problems (\maybebeta) appears any
+time we deal with a subterm $t = X t_1 \dots t_n$, where $X$ is flexible and the
+list $[t_1 \dots t_n]$ in not in \llambda. This unification problem is not
+solvable without loss of generality, since there is not a most general unifier.
+If we take back the example given in \cref{sec:nutshell}, the unification $F a =
+a$ admits two solutions for $F$: $\rho_1 = \{F \mapsto \lambda x.x\}$ and
+$\rho_2 = \{F \mapsto \lambda \_.a\}$. Despite this, it is possible to work with
+$\maybebeta$ if an oracle provides a substitution $\rho$ such that $\rho t$
+falls again in the \llambda.
 

### 2024-03-28 paper/main.tex
macro for unif pb, mappings and links display

@@
 \usepackage{macros}
+\usepackage{enumitem}
 
@@
 constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
-is instantiated to a rigid term 2) \rhs can be $\eta$-reduced to a term with
-rigid head. If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
+is instantiated to a rigid term 2) \rhs is no more a \maybeeta or it is a term
+which can be reduced to a term with rigid head. 
+If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
 we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
@@
 
-\begin{description}
-  \item[Unif problems:] $X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)$
-  \item[Mappings:] $\mapping{F}{H}{2}$  
-  \item[Links:] $l_1 = \linketaM{x}{Y_{x}}{\lambda y.H_{y~x}}$,\\
-                $l_2 = \linketaM{}{X}{\lambda x.Y_{x}}$,\\
-                $l_3 = \linketaM{}{Z}{\lambda x.f \appsep G_{x}~x}$\\
-                $l_4 = \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}$
+\newcommand{\printMinipage}[1]{
+  \listlength{#1}
+  \foreach \x [count=\i] in #1
+    {
+      \begin{minipage}{2.8cm}
+        $\x$ %
+      \end{minipage}
+      \ifthenelse{\equal{\i}{\thellength}}{}{
+        \ifthenelse{\isodd{\i}}{|}{\\}
+      }
+    }
+}
+
+\newcommand{\printLinks}[3]{
+  \noindent
+  \begin{description}[leftmargin=!,labelwidth=\widthof{\bfseries Unif problems:}]
+    \item[Unif problems:] \printMinipage{#1}
+    \item[Mappings:] \printMinipage{#2}
+    \item[Links:] \printMinipage{#3}
   \end{description}
+}
+
+\printLinks{{
+  X = \lambda x.\lambda y.x, 
+  Z = f\appsep(\lambda x.x)}}
+  {{\mapping{F}{H}{2}}}{{
+  \linketaM{x}{Y_{x}}{\lambda y.H_{y x}},
+  \linketaM{\phantom{x}}{X}{\lambda x.Y_{x}},
+  \linketaM{\phantom{x}}{Z}{\lambda x.f \appsep G_{x}~x},
+  \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}}}
 

### 2024-03-29 paper/main.tex
better display
mv printLink to printCompPb
some updates
update macro display pb

@@
 \newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
+\def\phantx{\phantom{x}}
+
+\newcommand{\printMinipage}[2]{
+  \listlength{#2}
+  \foreach \x [count=\i] in #2
+    {
+      \begin{minipage}{3cm}
+        $#1_\i : \x$ %
+      \end{minipage}
+      \ifthenelse{\equal{\i}{\thellength}}{}{
+        \ifthenelse{\isodd{\i}}{|}{\\}
+      }
+    }
+}
+
+\newcommand{\printEntry}[1]{
+  \listlength{#1}%
+  \ifthenelse{\equal{0}{\thellength}}{}{
+    \item[Entry pb:] 
+    \foreach \x [count=\i] in #1
+      {%
+        \begin{minipage}{8cm}%
+          $\mathcal{P}_\i : \x$ %
+        \end{minipage}%
+        \ifthenelse{\equal{\i}{\thellength}}{}{\\}%
+      }%
+  }%
+}
+\newcommand{\printItemComp}[3]{
+  \listlength{#3}%
+  \ifthenelse{\equal{0}{\thellength}}{}{
+    \item[#1] \printMinipage{#2}{#3}
+  }%
+}
+
+\newcommand{\printAll}[4]{
+  \noindent%
+  \begin{description}[leftmargin=!,labelwidth=\widthof{\bfseries Mapping:}]
+    \printEntry{#1}%
+    \printItemComp{Compil:}{\mathcal{T}}{#2}%
+    \printItemComp{Mapping:}{m}{#3}%
+    \printItemComp{Links:}{l}{#4}
+  \end{description}
+}
 
@@
 In \cref{sec:basic-comp-limitations}, we have given two unification problems to
-be run one after the other. The compilation of each term would generate the
-following result:
-
-\begin{description}
-  \item[Unif problems:] $X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)$
-  \item[Mappings:] $\mapping{F}{G}{1}, \mapping{F}{H}{2}$  
-  \item[Links:] $\linketaM{x}{Y_{x}}{\lambda y.H_{y~x}}$,\\
-                $\linketaM{}{X}{\lambda x.Y_{x}}$,\\
-                $\linketaM{}{Z}{\lambda x.f \appsep G_{x}~x}$
-\end{description}
+be run one after the other. In the following table we present the entry problem,
+its compiled version, with the corresponding list of mapping and links.
+%
+\printAll
+  {{\elpiIn{flam x\flam y\fapp[fuva F,y,x] = flam x\flam y\x},
+    \elpiIn{flam x\fapp[fcon"f", fapp[fuva F, x], x] = fapp[fcon"f", flam x\x]}}}
+  {{X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)}}
+  {{\mapping{F}{G}{1}, \mapping{F}{H}{2}}}
+  {{\linketaM{x}{Y_{x}}{\lambda y.H_{y~x}},
+    \linketaM{\phantx}{X}{\lambda x.Y_{x}},
+    \linketaM{\phantx}{Z}{\lambda x.f \appsep G_{x}~x}}}
 
@@
 The complete problem to run for resolution is now:
-
-\newcommand{\printMinipage}[1]{
-  \listlength{#1}
-  \foreach \x [count=\i] in #1
-    {
-      \begin{minipage}{2.8cm}
-        $\x$ %
-      \end{minipage}
-      \ifthenelse{\equal{\i}{\thellength}}{}{
-        \ifthenelse{\isodd{\i}}{|}{\\}
-      }
-    }
-}
-
-\newcommand{\printLinks}[3]{
-  \noindent
-  \begin{description}[leftmargin=!,labelwidth=\widthof{\bfseries Unif problems:}]
-    \item[Unif problems:] \printMinipage{#1}
-    \item[Mappings:] \printMinipage{#2}
-    \item[Links:] \printMinipage{#3}
-  \end{description}
-}
-
-\printLinks{{
-  X = \lambda x.\lambda y.x, 
+%
+\printAll
+  {{}}
+  {{X = \lambda x.\lambda y.x, 
     Z = f\appsep(\lambda x.x)}}
-  {{\mapping{F}{H}{2}}}{{
-  \linketaM{x}{Y_{x}}{\lambda y.H_{y x}},
+  {{\mapping{F}{H}{2}}}
+  {{\linketaM{x}{Y_{x}}{\lambda y.H_{y x}},
     \linketaM{\phantom{x}}{X}{\lambda x.Y_{x}},

### 2024-03-29 paper/main.tex
tex update

@@
 
-$\eta$-reduction is an equivalence relation over terms where a term of the form
+$\eta$-reduction is an equivalence relation where a term of the form
 $\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
-free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
-version of $t$. The implementation of the \elpiIn{comp} relation given in
-\cref{sec:compilation} compiles the \Fo{} terms $t_1 =$ \elpiIn{flam x\ fapp
-[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
-$ \elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
-\mapping{\elpiIn{A}}{\elpiIn{A'}}{1}. However, if the oracle sets \elpiIn{A'} to the constant
-\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
-fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
-fact that $t'_1 =$~\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
-=$~\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
-unification problem by adapting the \elpiIn{comp} relation such that it
-recognizes \maybeeta subterms and replaces them with fresh \Ho variables. 
-This link between the variable $v$ and the subterm $t$ is stored in what we
+free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
+\printAll
+  {{\elpiIn{flam x\ fapp [fuva A, x]} \Uo \elpiIn{fcon "f"}}}
+  {{\lambda x. A'_x \Ue f}}
+  {{\mapping{\elpiIn{A}}{A'}{1}}}
+  {{}}
+Following the implementation of \elpiIn{comp} given in
+\cref{sec:compilation}, the unification problem $\mathcal{P}_1$ given 
+above, is compiled into the unification problem $\mathcal{T}_1$. 
+We can note that all the terms appearing in $\mathcal{P}_1$ are
+unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathcal{T}_1$:
+its left and right hand sides have both different rigid head:
+$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
+and $f$ corresponds to \elpiIn{con"f"}. 
+This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
+is a term belonging to \maybeeta.
+
+In order to guarantee \cref{prop:simulation}, we need to modify the way terms are
+compiled. The goal is to recognize
+every \maybeeta subterm $t$ and replace it with fresh \Ho variables $v$. 
+This connection between the variable $v$ and the subterm $t$ is stored in what we
 call \linketa which is an object with the following type
@@
 \begin{invariant}[\linketa \rhs]
-  The \rhs of a \linketa having the shape $\lambda x.F_x$
-  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
+  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t_x$
+  where $t_x$ is a \maybeeta term and $x$ is free in $t$.
   \label{inv:link-eta-right}
@@
 In the next three subsections we explain how we detect \maybeeta terms, how we
-compile them and how \linketa are activated during the execution of the program
-and provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.
+compile them and how the generated \linketa are activated during the \elpiIn{progress}.
+Moreover, we provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.
 
@@
 
-Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
-for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
-given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
+Compiling a term $t$ forces us to determine if there exists a $t' \in \subterm{t}$,
+such that $t' = \lambda x. t''_x$,
+for any term $t''$, can be a $\eta$-expansion, i.e. under a
+given substitution $\sigma$, we have $\sigma (\lambda x.t''_x) = t'$. This
 \maybeeta detection is not a trivial operation as it may seems.
@@
 \noindent
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{.5\linewidth}%
   \begin{gather}

### 2024-03-29 paper/main.tex
update tex

@@
 
-In the examples above, the first expression is a \maybeeta since $A_x$ can
-reduce to $x$, the second one is not a \maybeeta since it exists no substitution
+In the examples above, the first term is a \maybeeta since $A_x$ can
+reduce to $x$ by setting $A_x = \lambda x.x$, 
+the second one is not a \maybeeta since it exists no substitution
 for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
-Eq.~\ref{eq:lamlam} is a bit more
-complicated since, we have a spine of lambdas, this means that the whole term
-is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
-$t$, and $t$ can be reduced to a term on the form $t' x$ where $x$ is not free
-in $t'$. Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
+It is a bit more complicated to determine if eq.~\ref{eq:lamlam} a \maybeeta since,
+we have a spine of lambdas, this means that the whole term
+is a \maybeeta, if the inner $\lambda$-term $\lambda y.f\appsep A_x\appsep B_yx$,
+is an $\eta$-expansion and, if so, we need to verify if the $\eta$-reduced 
+subterm, i.e. $f \appsep A_x$, can make $\lambda x.f \appsep A_x$ an $\eta$-expansion.
+Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
 $\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.
-
-As a final remark, note that eq.~\ref{eq:lamdisc} is a \maybeeta even if $x$
-occurs in the subterm $t' = f \appsep A_x$. It exists in fact a substitution
+Lastly, eq.~\ref{eq:lamdisc} is a \maybeeta even if $x$
+occurs in the subterm $t' = f \appsep A_x$, since it exists a substitution
 $\sigma = \{A \mapsto \lambda x.a\}$, for any constant $a$, such that $\sigma t'
@@
 
-We can now define more formally the two auxiliary relations needed for \maybeeta
-detection:
+We can now define more formally how \maybeeta terms are detected together with
+its auxiliary functions:
 

### 2024-03-29 paper/main.tex
update tex

@@
 
-\begin{invariant}
-  Any term in the substitution of \Ho is in normal form.
-  TODO: move this to the right place
-  TODO: definire normal form, maybe in definition 2.7
-  \label{inv:subst-nf}
-\end{invariant}
-
 \input{code/comp_base_types}
@@
 
-Finally, we can derive the implementation for \maybeeta detection:
+We can now derive the implementation for \maybeeta detection:
 
@@
 
-Finally, the \maybeeta detection defined just before is an
+As a final remark, the \maybeeta detection defined just before is an
 over-approximation, in the sense that there exists some terms $t$ considered as
 \maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
-$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$.
-
-The implementation we propose for the \maybeeta relation is given below.
-
-\input{code/maybe_eta}
+$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$
+which is considered a \maybeeta. Thi is because, we suppose at the same time
+that the second $A_x$ can reduce to $x$, to the first $A_x$ is not reducible to
+$A_x$. This is however not considered as a problem, since, adding holes to 
+the compiled term does not break \cref{prop:simulation}, since each hole is
+connected to a \linketa which is activated exactly when the hole is instantiated
+the \linketa, as explain below is activated.
+
+% The implementation we propose for the \maybeeta relation is given below.
+
+% \input{code/maybe_eta}
 
@@
 which links the variable \elpiIn{A} with $t$. This link creation enforce
-\cref{inv:linklhs} and \cref{inv:link-eta-right}, since clearly the \lhs is a
-variable and the \rhs is a term starting with the \elpiIn{lam} constructor. We
-can also note that for any name $x$, $tx = t' \appsep x$ never start with
-\elpiIn{lam}, since by contradiction, if 1) $tx$ is an $\eta$-expansion, then
-the recursive call to \elpiIn{comp} would have replaced $tx$ with a fresh
-variable, which contradict the hypothesis and if 2) $t'$ is not an
-$\eta$-expansion, then neither $\lambda t'$ is.
+\cref{inv:linklhs} and \cref{inv:link-eta-right}, since \lhs is a trivially a
+variable and the \rhs is a term $t$ starting with the \elpiIn{lam} constructor where
+$x$ is free in $t$ otherwise $t$ would not have been detected as a \maybeeta.
+
+\begin{corollary}
+  The \rhs of any \linketa has exactly one lambda abstraction.
+\end{corollary}
+
+\begin{proof}[Proof sketch]
+  By contradiction, suppose that a \linketa $l$ has $t$ as \rhs and $t = \lambda
+  x.\lambda y.t'_{xy}$. Two cases are to be analyses: 1) $\lambda y.t'_{xy}$ is
+  a $\maybeeta$, then, by construction, \rhs would have been replaced with a the
+  $\eta$-expansion of fresh variable $v$, which is a contradiction, since
+  $\lambda y.t'_{xy} \neq \lambda x.v_{x}$; 2) $\lambda y.t'_{xy}$ is not an a
+  \maybeeta, then neither $t$ is, which is a contradiction since \rhs is always
+  a \maybeeta by construction.
+\end{proof}
 
@@
 
-In order to activate a \linketa $l$, we need to implement new rules for the
-\elpiIn{progress1} predicate. After passing under all the \elpiIn{abs}
+In order to activate a \linketa $l$, we need to extend the \elpiIn{progress1} predicate
+by adding new rules. After passing under all the \elpiIn{abs}
 constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
@@
 therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
-of $t$) with $t'$. Finally, if \lhs starts has \elpiIn{lam} as constructor, then
-$t$ is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.
-
-The second way to activate a \linketa is when the \rhs can $\eta$-reduced to a
-term $t$ with rigid head, i.e. $t'$ is not a \maybeeta. This means that we can
-unify \lhs with $t'$.
+of $t$) with $t'$. Finally, if $t$ is a term with \elpiIn{lam} as head, then
+it is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.
+
+The second way to activate a \linketa is when the \rhs is no more a \maybeeta or
+\rhs can be $\eta$-reduced to a term $t$ with rigid head. In both cases, \lhs is
+unified with $t'$.
 
@@
 that this link progression enforce
-\cref{inv:subst-nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
+\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
 the \Ho substitution, since we make unification only when we know that the 
@@
 no more a \maybeeta, the link is removed from \linkStore.
+
+An example
 

### 2024-03-29 paper/main.tex
update elpi printer for tex
update example eta

@@
 \printAll
-  {{\elpiIn{flam x\ fapp [fuva A, x]} \Uo \elpiIn{fcon "f"}}}
-  {{\lambda x. A'_x \Ue f}}
-  {{\mapping{\elpiIn{A}}{A'}{1}}}
+  {{\elpiIn{flam x\ fapp [fuva X, x]} \Uo \elpiIn{fcon"f"}}}
+  {{\lambda x. X'_x \Ue f}}
+  {{\mapping{\elpiIn{X}}{X'}{1}}}
   {{}}
@@
 which can be reduced to a term with rigid head. 
-If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
+If \lhs is instantiated to a rigid term $t$, by \cref{prop:nf},
 we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
@@
 
-An example
-
-TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
+\printAll
+  {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
+  {{A \Ue f}}
+  {{\mapping{X}{X'}{1}}}
+  {{\linketaM{}{A}{\lambda x.X'_{x}}}}
+
+The example above shows the new compilation of the unification problem given at
+the beginning of \cref{sec:eta}. This time, we see that the the left hand side
+$t$ of $\mathcal{P}_1$ has been detected as a \maybeeta and replaced with the
+fresh variable $A$. Moreover, \linkStore contains the link $l_1$ connecting $A$
+with $\lambda x.X'_{x}$, the compiled version of $t$. After the resolution of
+$\mathcal{T}_1$, $A$ is assigned to $f$. Therefore \lhs of $l_1$ is a term with
+\elpiIn{con} as constructor. This means that \rhs of $l_1$ is a $\eta$-expansion
+and therefore we can unify $\lambda x.f x$ with \rhs, which instantiate $X'_x$
+to $f \appsep x$. $l_1$ is removed from \linkStore which is now empty,
+\elpiIn{progress} terminate, and the decompilation, will instantiate the $X$
+variable to $f$ (which is the $\eta$-contracted version of $\lambda x.f \appsep
+x$).
+
+A second example, showing the activation of a link when the \rhs is no more a
+\maybeeta, is given in \cref{sec:invariant1}, since we need to work with
+variables used with different arities.
+
 TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$

### 2024-03-29 paper/main.tex
update elpi printe + some tex

@@
 \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
-\lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
+\lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
+% \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
 \end{gather}
@@
 \printAll
-  {{\elpiIn{flam x\flam y\fapp[fuva F,y,x] = flam x\flam y\x},
-    \elpiIn{flam x\fapp[fcon"f", fapp[fuva F, x], x] = fapp[fcon"f", flam x\x]}}}
-  {{X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)}}
-  {{\mapping{F}{G}{1}, \mapping{F}{H}{2}}}
-  {{\linketaM{x}{Y_{x}}{\lambda y.H_{y~x}},
-    \linketaM{\phantx}{X}{\lambda x.Y_{x}},
-    \linketaM{\phantx}{Z}{\lambda x.f \appsep G_{x}~x}}}
+  {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
+    \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
+  {{E0 == \lambda x.\lambda y.x,
+    E3 == E5}}
+  {{\mapping{Y}{E5}{0},
+    \mapping{X}{E2}{2}}}
+  {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
+    \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
+    \linketaM{}{E0}{\lambda x.E1_{x}},
+    \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
 

### 2024-03-30 paper/main.tex
add notations
alternative list of maybe-eta examples
tex wip

@@
 \noindent
+When variables $x$ and $y$ can occur in term $t$ we shall write
+$t_{xy}$ to stress this fact.
+
+We write $\sigma = \{ A_{xy} \mapsto y\}$ for the assignment
+\elpiIn{abs x\abs y\y} and $\sigma = \{ A \mapsto \lambda x.\lambda y.y\}$
+for \elpiIn{lam x\lam y\y}.
+
 When detailing examples we write links as equations between two
@@
 
+\begin{center}
+  \begin{tabular}{lll}
+    Term & Status & Evidence \\\hline
+    $\lambda x. f \appsep A_x$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto x ~\}$ \\
+    $\lambda x. f \appsep x \appsep A_x$ & $\not\in\maybeeta$ &\\
+    $\lambda x. \lambda y. f \appsep A_x \appsep B_{y x}$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto x,~ B_{yx} \mapsto y ~\}$\\
+    $\lambda x. f \appsep A_x \appsep x$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto a ~\}$
+  \end{tabular}
+\end{center}
+
 In the examples above, the first term is a \maybeeta since $A_x$ can
@@
 \maybeeta, is given in \cref{sec:invariant1}, since we need to work with
-variables used with different arities.
-
-TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$
-
-A second way to progress \linketa, that we call \linketa deduplication, is when
+variables used with different arities. This example represent the run of the
+unification problems proposed at \cref{sec:basic-comp-limitations}
+
+Another, way to progress \linketa, that we call \linketa deduplication, is when
 \linkStore contains two \linketa $l_1$ and $l_2$ with a \lhs having the same
@@
 
-TODO: example for this: $\lambda x. \lambda y. F \appsep y \appsep x = X,
-\lambda x. \lambda y. F \appsep y \appsep x = Y$
+A very basic example of \linketa deduplication, is given below:
+
+\printAll
+  {{\elpiIn{flam x\ fapp [fuva X, x] == flam x\ fapp [fuva X, x]}}}
+  {{E0 == E2}}
+  {{\mapping{X}{X'}{1}}}
+  {{\linketaM{}{E2}{\lambda x.X'_{x}},
+    \linketaM{}{E0}{\lambda x.X'_{x}}}}
+
+TODO: explain
 
@@
 not reduce to $T x$ where x is not free in $T$.
-
-% \subsection{Example}
-
-% In this subsection we show two complete runs of the extended version of \Ue.
-
-% \paragraph{Example 1}
-% \begin{center}
-%   \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
-% \end{center}
-% % $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
-% \begin{itemize}
-%   \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
-%         $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
-%         c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
-%   \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
-%   \item This unification wakes up the second link, since its \lhs is now
-%         instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
-%         \lambda c1.c0$
-%   \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
-%         unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
-%         pruning.
-%   \item All the links have been solved and the decompilation, assigns the \Fo{}
-%         variable \elpiIn{A} to $\lambda xy.y$ 
-% \end{itemize}
-
-% \paragraph{Example 2}
-% \begin{center}
-%   \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
-% \end{center}
-% % $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
-% \begin{itemize}
-%   \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
-%         $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
-%         $\mapping{C0}{X1}{1}$
-%   \item After unification $X0$ is set to $f~\lambda c0.c0$
-%   \item This unification wakes up the link, since its \lhs is now
-%         instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
-%         \lambda c1.c0$
-%   \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
-%         unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
-%         pruning.
-%   \item All the links have been solved and the decompilation, assigns the \Fo{}
-%         variable \elpiIn{A} to $\lambda xy.y$ 
-% \end{itemize}
 

### 2024-03-30 pre-commit.sh
add pre-commit file

@@ -1,0 +1,20 @@
+#!/usr/bin/env bash
+
+PRECOMMIT_PATH=".git/hooks/pre-commit"
+PRECOMMIT_TEXT="#!/usr/bin/env bash\n./pre-commit.sh"
+
+touch $PRECOMMIT_PATH & 
+chmod u+x $PRECOMMIT_PATH & 
+echo -e $PRECOMMIT_TEXT > $PRECOMMIT_PATH
+
+make test >& /tmp/unif-log
+RESULT=$?
+
+cat /tmp/unif-log | grep -A100 "SOME.*FAIL"
+
+if [[ ${RESULT} -ne 0 ]]
+then
+    RED='\033[0;31m'
+    printf "${RED}Unit tests failed. Commit rejected\n"
+fi
+exit ${RESULT}

### 2024-03-30 paper/main.tex
rework maybeeta example

@@
 
-Compiling a term $t$ forces us to determine if there exists a $t' \in \subterm{t}$,
-such that $t' = \lambda x. t''_x$,
-for any term $t''$, can be a $\eta$-expansion, i.e. under a
-given substitution $\sigma$, we have $\sigma (\lambda x.t''_x) = t'$. This
-\maybeeta detection is not a trivial operation as it may seems.
-\\
-\noindent
-\begin{minipage}{.5\linewidth}%
-  \begin{gather}
-    \lambda x. f \appsep A_x\\
-    \lambda x. f \appsep x \appsep A_x
-  \end{gather}
-\end{minipage}% This must go next to `\end{minipage}`
-\begin{minipage}{.5\linewidth}
-  \begin{gather}
-    \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
-    \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
-  \end{gather}
-\end{minipage}
-
+When compiling a term $t$ we need to determine if any
+subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
+where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
+there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
+The detection of lambda abstractions that can ``disappear''
+is not as trivial as it may seems, here a few examples:
+% \\
+% \noindent
+% \begin{minipage}{.5\linewidth}%
+%   \begin{gather}
+%     \lambda x. f \appsep A_x\\
+%     \lambda x. f \appsep x \appsep A_x
+%   \end{gather}
+% \end{minipage}% This must go next to `\end{minipage}`
+% \begin{minipage}{.5\linewidth}
+%   \begin{gather}
+%     \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
+%     \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
+%   \end{gather}
+% \end{minipage}
+%
 \begin{center}
   \begin{tabular}{lll}
-    Term & Status & Evidence \\\hline
-    $\lambda x. f \appsep A_x$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto x ~\}$ \\
-    $\lambda x. f \appsep x \appsep A_x$ & $\not\in\maybeeta$ &\\
-    $\lambda x. \lambda y. f \appsep A_x \appsep B_{y x}$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto x,~ B_{yx} \mapsto y ~\}$\\
-    $\lambda x. f \appsep A_x \appsep x$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto a ~\}$
+    %Term & Status & Evidence \\\hline
+    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
+    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
+    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
+    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
   \end{tabular}
 \end{center}
-
-In the examples above, the first term is a \maybeeta since $A_x$ can
-reduce to $x$ by setting $A_x = \lambda x.x$, 
-the second one is not a \maybeeta since it exists no substitution
-for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
-It is a bit more complicated to determine if eq.~\ref{eq:lamlam} a \maybeeta since,
-we have a spine of lambdas, this means that the whole term
-is a \maybeeta, if the inner $\lambda$-term $\lambda y.f\appsep A_x\appsep B_yx$,
-is an $\eta$-expansion and, if so, we need to verify if the $\eta$-reduced 
-subterm, i.e. $f \appsep A_x$, can make $\lambda x.f \appsep A_x$ an $\eta$-expansion.
-Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
-$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.
-Lastly, eq.~\ref{eq:lamdisc} is a \maybeeta even if $x$
-occurs in the subterm $t' = f \appsep A_x$, since it exists a substitution
-$\sigma = \{A \mapsto \lambda x.a\}$, for any constant $a$, such that $\sigma t'
-= f \appsep a$ is a term where $x$ does not appears as a free variable.
+\vspace{4pt}
+
+% In the examples above, the first term is a \maybeeta since $A_x$ can
+% reduce to $x$ by setting $A_x = \lambda x.x$, 
+% the second one is not a \maybeeta since it exists no substitution
+% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
+\noindent
+The first two examples are easy, and show how a unification variable can expose
+or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
+The third example shows when a variable occurs outside the scope of a unification
+variable cannot be erased and hence prevent a term from being an $\eta$-expansion.\\
+The last example shows the recursive nature of the check we need to implement.
+The term starts with a spine of two lambdas hence the whole term
+is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
+is in \maybeeta itself, that means it could $\eta$-contract to
+$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
+$\eta$-expansion.\\
 

### 2024-03-30 paper/main.tex
minor

@@
 
-We represent a logic program \emph{run} in \Fo as
-a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
+We represent a logic program \emph{run} in \Fo{} as
+a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
 unification problem between terms $\mathcal{S}_{p_l}$ and

### 2024-03-30 paper/main.tex
rework notation

@@
 Here a few examples:
-\vspace{5pt}
+\vspace{4pt}
 
 \begin{tabular}{ll}
-  $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
+  $f\appsep a$ &  \elpiIn{app[con "f", con "a"]}\\
   $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
-  $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
-  $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
+  $\lambda x.F_{x} \appsep a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
+  $\lambda x.F_{x} \appsep x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
 \end{tabular}
-\vspace{5pt}
+\vspace{4pt}
 
@@
 
-We write $\sigma = \{ A_{xy} \mapsto y\}$ for the assignment
-\elpiIn{abs x\abs y\y} and $\sigma = \{ A \mapsto \lambda x.\lambda y.y\}$
+We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
+\elpiIn{abs x\abs y\y} and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
 for \elpiIn{lam x\lam y\y}.

### 2024-03-30 paper/main.tex
alternative syntax for examples
rename bound vars
indentation

@@
 no more a \maybeeta, the link is removed from \linkStore.
-
-\printAll
-  {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
-  {{A \Ue f}}
-  {{\mapping{X}{X'}{1}}}
-  {{\linketaM{}{A}{\lambda x.X'_{x}}}}
+% \printAll
+%   {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
+%   {{A \Ue f}}
+%   {{\mapping{X}{X'}{1}}}
+%   {{\linketaM{}{A}{\lambda x.X'_{x}}}}
+$$
+\begin{array}{l}
+  \mathcal{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
+  \mathcal{T} = \{~ A \Ue f ~\}\\
+  \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
+  \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
+\end{array}
+$$
 
@@
 its compiled version, with the corresponding list of mapping and links.
-%
+
 \printAll
@@
     \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
-
+$$
+{\arraycolsep=2pt \begin{array}{rrclrcl}
+    \mathcal{P} = \{ &
+       \lambda x.\lambda y.X \appsep y \appsep x & \Uo & \lambda x.\lambda y. x & \quad
+       \lambda x. f \appsep (X \appsep x) \appsep x & \Uo&  Y ~\}\\
+    \mathcal{T} = \{ &
+       A &\Ue& \lambda x.\lambda y.x & \quad
+       D &\Ue& F ~\}\\
+    \mathbb{M} = \{ & \multicolumn{6}{l}{
+      \mapping{Y}{F}{0} \quad
+      \mapping{X}{C}{2} ~\}}\\
+    \linkStore \;= \Big\{  &
+      \multicolumn{6}{l}{ \begin{array}{ll}
+          \linketaM{}{A}{\lambda x.B_{x}} & \quad
+          \linketaM{x}{B_{x}}{\lambda y.C_{y x}}\\
+          \linketaM{}{D}{\lambda x.(f~E_{x}~x)} & \quad
+          \linketaM{y}{E_{y}}{\lambda w.C_{y w}}
+      \end{array} ~\Big\}}
+\end{array}}
+$$
+\marginpar{aggiustare i nomi delle variabili nel testo}
 We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep

### 2024-03-30 paper/main.tex
6.1

@@
 %\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
-\begin{definition}[\maybeeta]
+\begin{definition}[\maybeeta]\label{def:maybeeta}
   $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
@@
 
-\begin{definition}[\maybebeta]
+\begin{definition}[\maybebeta]\label{def:maybebeta}
   $\maybebeta = \{ X t_1 \ldots t_n ~|~ X t_1 \ldots t_n \not\in \llambda \}$.
@@
 \section{Handling of \maybeeta}\label{sec:eta}
-
+\marginpar{TODO: fix syntax}
 $\eta$-reduction is an equivalence relation where a term of the form
@@
 
-\begin{definition}[reduce-to]
-  A term $t$ reduce to a name $x$, if $\exists \sigma, \sigma t = x$. In
-  particular, for any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$
-  reduces to a bound variable $x$ if one of the three following cases is
-  satisfied: 1) $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
-  of terms $l$ and each $l_i$ reduces to $x_i$
-  (e.g. $\lambda x_1 \ldots x_n.x_{x_1\ldots x_n} = x$) ; 3) $t$ is a unification
-  variable in \llambda with scope
-  $s$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
-  there exists a $s_i \in s$, such that $s_i$ reduces to $v$
+\marginpar{the compiler must do beta normal!!!!}
+\newcommand{\reduceto}{\emph{may-contract-to}}
+\begin{definition}[\reduceto]
+  A term $s$ \reduceto{} a name $x$ if there exists a
+  substitution $\rho$ such that $\rho s \Eo{} x$.
 \end{definition}
 
-\begin{definition}[occurs-rigidly]
-  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, x \in
-  \subterm{\sigma t}$
+\begin{lemma}\label{lem:reduceto}
+A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
+%, where $x, x_1\ldots x_n$ can occur in $t$, 
+\reduceto{} $x$ only if one of the following three conditions holds:
+\begin{enumerate}
+  \item $n = 0$ and $t = x$;
+  \item $t$ is the application of $x$ to a list
+     of terms $l$ and each $l_i$ \reduceto{} $x_i$
+     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
+  \item $t$ is a unification variable with scope
+    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
+    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
+    (if $n = 0$ this is equivalent to $x \in W$).
+\end{enumerate}
+\end{lemma}
+\begin{proof}[Proof sketch]
+Since our terms are in $\beta$-normal form there is only
+one rule that can play a role (namely \ref{clause:eta1}),
+hence if the term $s$ is not exactly $x$ (case 1)
+it can only be an $\eta$-expansion of it, or a unification
+variable that can be assigned to it, or a combination of both.
+If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
+contraction. In that case the term $t$ under the spine of binders for
+$x_1\ldots x_n$ can either be $x$ itself applied to terms that can
+\reduceto{} these variables (case 2), or a unification variable
+that can be assigned to that application (case 3).
+\end{proof}
+
+\noindent
+Note that this condition does not require the term to be in \llambda.
+\marginpar{Is this relevant}
+
+\newcommand{\occursrigid}{\emph{occurs-rigidly}}
+\begin{definition}[\occursrigid]\label{def:occrigid}
+  A name $x$ \occursrigid{} in a $\beta$-normal term $t$, if $\forall \rho, x \in
+  \subterm{\rho t}$
 \end{definition}
-In other words $x$ \emph{occurs-rigidly} in $t$ if it occurs in $t$
-outside of the scope of unification variables since theirs instantiations
-are allowed to discard their scope.
+In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
+outside of the scope of unification variables since an instantiation
+is allowed to discard $x$ from the scope of the unification variable.
+Note that $\eta$-contraction cannot make $x$ disappear, since the
+variables being erased by $\eta$-contraction are locally bound inside $t$.
 
@@
 
-\begin{definition}[\maybeeta detection]
-  A term $\lambda x_1 \ldots x_n.t$ where each $x_i$ occurs in $t$ is a
-  \maybeeta if
-  either: 1) $t$ is a constant applied to arguments
+\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
+  A $\beta$-normal term
+  $\lambda x_1 \ldots x_n.t \in \maybeeta$ (as per \cref{def:maybeeta})
+  % where each $x_i$ occurs in $t$ is a
+  %\maybeeta if
+  %either: 
+  if one of the following holds:
+  \begin{enumerate}
+    \item $t$ is a constant or variable applied to arguments
       $l_1 \ldots l_m$ such that 
-  $m \geq n$ and every $l_{m-n+i}$ reduces-to $x_i$ and
-  no $x_i$ occurs-rigidly in $l_{1\ldots m-n-1}$; or 2) $t$ is a
-  unification variable with scope $s$ and
-  for each $x_i$ there exists a $s_j \in s$ such that $s_j$ reduces-to $x_i$.
-\end{definition}
-
-As a final remark, the \maybeeta detection defined just before is an
-over-approximation, in the sense that there exists some terms $t$ considered as
-\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
-$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$
-which is considered a \maybeeta. Thi is because, we suppose at the same time
-that the second $A_x$ can reduce to $x$, to the first $A_x$ is not reducible to
-$A_x$. This is however not considered as a problem, since, adding holes to 
-the compiled term does not break \cref{prop:simulation}, since each hole is
-connected to a \linketa which is activated exactly when the hole is instantiated
-the \linketa, as explain below is activated.
+      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
+      the term  $l_i$
+      \reduceto{} $x_i$, and
+      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
+    \item $t$ is a
+      unification variable with scope $W$ and
+      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
+      \reduceto{} $x_i$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}[Proof sketch]
+Follows from \cref{def:occrigid} and \cref{lem:reduceto}
+\end{proof}
+
+\noindent
+Remark that the converse of \cref{lem:maybeeta} does not hold: 
+there exists a term $t \in \maybeeta$
+such that forall substitution $\rho$, $\rho t$ is not an
+$\eta$-expansion. A simple counter example is
+$\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
+which is in \maybeeta{} since $x$ does not \occursrigid{} in the first argument
+of $f$,
+and the second argument of $f$ \reduceto{} $x$. In other words $A \appsep x$
+may either use $x$ or discard it.\\
+This is however not a problem in the sense that it does not break
+\cref{prop:simulation}. Terms in \maybeeta{} are compiled to
+unification variables and a link (see \cref{sec:etacomp}):
+the link makes progress (see \cref{sec:etaprogress})
+in the same step in which the
+variable is instantiated, and that compensates for the
+potential misclassification.
 
@@
 
-\subsection{Compilation}
+\subsection{Compilation}\label{sec:etacomp}
 
@@
 
-\subsection{Progress}
+\subsection{Progress}\label{sec:etaprogress}
 

### 2024-03-30 paper/main.tex
bli bla

@@
 \newcommand{\occursrigid}{\emph{occurs-rigidly}}
+\newcommand{\occurrigid}{\emph{occur-rigidly}}
 \begin{definition}[\occursrigid]\label{def:occrigid}
@@
 $\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
-which is in \maybeeta{} since $x$ does not \occursrigid{} in the first argument
+which is in \maybeeta{} since $x$ does not \occurrigid{} in the first argument
 of $f$,
 and the second argument of $f$ \reduceto{} $x$. In other words $A \appsep x$
-may either use $x$ or discard it.\\
+may either use or discard $x$, but our analysis does not know that \emph{the same
+term} cannot have two contrasting behaviors.\\
 This is however not a problem in the sense that it does not break
@@
 in the same step in which the
-variable is instantiated, and that compensates for the
-potential misclassification.
+variable is instantiated, and that compensates for this coarse analysis.
 

### 2024-03-30 paper/main.tex
free mathcal P for subterm

@@
         \begin{minipage}{8cm}%
-          $\mathcal{P}_\i : \x$ %
+          $\mathbb{P}_\i : \x$ %
         \end{minipage}%
@@
     \printEntry{#1}%
-    \printItemComp{Compil:}{\mathcal{T}}{#2}%
+    \printItemComp{Compil:}{\mathbb{T}}{#2}%
     \printItemComp{Mapping:}{m}{#3}%
@@
 a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
-unification problem between terms $\mathcal{S}_{p_l}$ and
-$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
+unification problem between terms $\mathbb{S}_{p_l}$ and
+$\mathbb{S}_{p_r}$ taken from the set of all terms $\mathbb{S}$.
 The composition of these steps starting from the
@@
 \begin{array}{l}
-\stepF{\mathcal{S}}{p}{\rho}{\rho''}
+\stepF{\mathbb{S}}{p}{\rho}{\rho''}
 \deff
-\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
-\runF{\mathcal{S}}{\mathcal{N}}{\rho}
+\stepFD{\mathbb{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
+\runF{\mathbb{S}}{\mathcal{N}}{\rho}
 \deff
-\runFD{\mathcal{S}}{\mathcal{N}}
+\runFD{\mathbb{S}}{\mathcal{N}}
 \end{array}
@@
 \begin{array}{l}
-\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
-  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
-\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \mathcal{T} \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
-  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
+\stepH{\mathbb{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
+  \qquad\stepHD{\mathbb{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
+\runH{\mathbb{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
+  \qquad \mathbb{T} \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathbb{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \runHD{\mathbb{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
   \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
@@
 \begin{proposition}[Simulation]\label{prop:simulation}
-$\forall \mathcal{S}, \forall \mathcal{N},$
-$$
-  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
+$\forall \mathbb{S}, \forall \mathcal{N},$
+$$
+  \runF{\mathbb{S}}{\mathcal{N}}{\rho}
   \Leftrightarrow
-  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
+  \runH{\mathbb{S}}{\mathcal{N}}{\rho}
 $$
@@
 \begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
+In the context of~ \hrun, if $~\mathbb{T} \subseteq \llambda$ we have that
 $\forall p \in 1 \ldots \mathcal{N},$
 $$
-\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
+\stepF{\mathbb{S}}{p}{\rho_{p-1}}{\rho_p}
 \Leftrightarrow
-\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
+\stepH{\mathbb{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
 $$
@@
 We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
-$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
+$\mathbb{S} = \{ s_1, s_2 \}$ as follows:
 %
@@
 \begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
-  largest set $\mathcal{P(t)}$ that can be obtained by the following rules.
+  largest set $\\subterm{t}$ that can be obtained by the following rules.
 $$
@@
 \begin{proposition}[\wellb{}-preservation]\label{prop:nf}
-$\forall \mathcal{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
+$\forall \mathbb{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
 $$
 \begin{array}{l}
-\wellb(\sigma\mathcal{T}) \land
-  \sigma\mathcal{T}_{p_l} \Ue \sigma\mathcal{T}_{p_r} \mapsto {\sigma'}
-  \Rightarrow \wellb(\sigma' \mathcal{T})\\
-\wellb(\sigma\mathcal{T}) \land
+\wellb(\sigma\mathbb{T}) \land
+  \sigma\mathbb{T}_{p_l} \Ue \sigma\mathbb{T}_{p_r} \mapsto {\sigma'}
+  \Rightarrow \wellb(\sigma' \mathbb{T})\\
+\wellb(\sigma\mathbb{T}) \land
   \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
-  \Rightarrow \wellb(\sigma' \mathcal{T})
+  \Rightarrow \wellb(\sigma' \mathbb{T})
 \end{array}
@@
 Following the implementation of \elpiIn{comp} given in
-\cref{sec:compilation}, the unification problem $\mathcal{P}_1$ given 
-above, is compiled into the unification problem $\mathcal{T}_1$. 
-We can note that all the terms appearing in $\mathcal{P}_1$ are
-unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathcal{T}_1$:
+\cref{sec:compilation}, the unification problem $\mathbb{P}_1$ given 
+above, is compiled into the unification problem $\mathbb{T}_1$. 
+We can note that all the terms appearing in $\mathbb{P}_1$ are
+unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathbb{T}_1$:
 its left and right hand sides have both different rigid head:
@@
 \begin{array}{l}
-  \mathcal{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
-  \mathcal{T} = \{~ A \Ue f ~\}\\
+  \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
+  \mathbb{T} = \{~ A \Ue f ~\}\\
   \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
@@
 the beginning of \cref{sec:eta}. This time, we see that the the left hand side
-$t$ of $\mathcal{P}_1$ has been detected as a \maybeeta and replaced with the
+$t$ of $\mathbb{P}_1$ has been detected as a \maybeeta and replaced with the
 fresh variable $A$. Moreover, \linkStore contains the link $l_1$ connecting $A$
 with $\lambda x.X'_{x}$, the compiled version of $t$. After the resolution of
-$\mathcal{T}_1$, $A$ is assigned to $f$. Therefore \lhs of $l_1$ is a term with
+$\mathbb{T}_1$, $A$ is assigned to $f$. Therefore \lhs of $l_1$ is a term with
 \elpiIn{con} as constructor. This means that \rhs of $l_1$ is a $\eta$-expansion
@@
 {\arraycolsep=2pt \begin{array}{rrclrcl}
-    \mathcal{P} = \{ &
+    \mathbb{P} = \{ &
        \lambda x.\lambda y.X \appsep y \appsep x & \Uo & \lambda x.\lambda y. x & \quad
        \lambda x. f \appsep (X \appsep x) \appsep x & \Uo&  Y ~\}\\
-    \mathcal{T} = \{ &
+    \mathbb{T} = \{ &
        A &\Ue& \lambda x.\lambda y.x & \quad

### 2024-03-31 paper/main.tex
printer generator for unif problems in tex

@@
 
-\newcommand{\printMinipage}[2]{
-  \listlength{#2}
-  \foreach \x [count=\i] in #2
-    {
-      \begin{minipage}{3cm}
-        $#1_\i : \x$ %
-      \end{minipage}
-      \ifthenelse{\equal{\i}{\thellength}}{}{
-        \ifthenelse{\isodd{\i}}{|}{\\}
-      }
-    }
-}
-
-\newcommand{\printEntry}[1]{
-  \listlength{#1}%
-  \ifthenelse{\equal{0}{\thellength}}{}{
-    \item[Entry pb:] 
-    \foreach \x [count=\i] in #1
-      {%
-        \begin{minipage}{8cm}%
-          $\mathbb{P}_\i : \x$ %
-        \end{minipage}%
-        \ifthenelse{\equal{\i}{\thellength}}{}{\\}%
-      }%
-  }%
-}
-\newcommand{\printItemComp}[3]{
-  \listlength{#3}%
-  \ifthenelse{\equal{0}{\thellength}}{}{
-    \item[#1] \printMinipage{#2}{#3}
-  }%
-}
-
-\newcommand{\printAll}[4]{
-  \noindent%
-  \begin{description}[leftmargin=!,labelwidth=\widthof{\bfseries Mapping:}]
-    \printEntry{#1}%
-    \printItemComp{Compil:}{\mathbb{T}}{#2}%
-    \printItemComp{Mapping:}{m}{#3}%
-    \printItemComp{Links:}{l}{#4}
-  \end{description}
-}
+\include{pb_printer}
 
@@
 free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
-\printAll
-  {{\elpiIn{flam x\ fapp [fuva X, x]} \Uo \elpiIn{fcon"f"}}}
-  {{\lambda x. X'_x \Ue f}}
-  {{\mapping{\elpiIn{X}}{X'}{1}}}
+\printAlll
+  {{{\lambda x. X \appsep x, f}}}
+  {{{\lambda x. A_x, f}}}
+  {{{X,A,1}}}
   {{}}
@@
 
+\printAlll
+  {{{\lambda x.X \appsep x,f}}}
+  {{{A,f}}}
+  {{{X,B,1}}}
+  {{{,A,\lambda x.B_x}}}
+
 The example above shows the new compilation of the unification problem given at
@@
 
-\printAll
-  {{\elpiIn{flam x\ fapp [fuva X, x] == flam x\ fapp [fuva X, x]}}}
-  {{E0 == E2}}
-  {{\mapping{X}{X'}{1}}}
-  {{\linketaM{}{E2}{\lambda x.X'_{x}},
-    \linketaM{}{E0}{\lambda x.X'_{x}}}}
+\printAlll
+  {{{\lambda x. X \appsep x, \lambda x. X \appsep x}}}
+  {{{E0, E2}}}
+  {{{X,A1,1}}}
+  {{{,E2,\lambda x.A1_{x}},
+    {,E0,\lambda x.A1_{x}}}}
 
@@
 
-\printAll
-  {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
-    \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
-  {{E0 == \lambda x.\lambda y.x,
-    E3 == E5}}
-  {{\mapping{Y}{E5}{0},
-    \mapping{X}{E2}{2}}}
-  {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
-    \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
-    \linketaM{}{E0}{\lambda x.E1_{x}},
-    \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
+% \printAll
+%   {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
+%     \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
+%   {{E0 == \lambda x.\lambda y.x,
+%     E3 == E5}}
+%   {{\mapping{Y}{E5}{0},
+%     \mapping{X}{E2}{2}}}
+%   {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
+%     \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
+%     \linketaM{}{E0}{\lambda x.E1_{x}},
+%     \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
 $$
@@
 $$
+
+\printAlll
+  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
+    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
+  {{{E0,\lambda x.\lambda y.x},
+    {E3,E5}}}
+  {{{Y,E5,0},
+    {X,E4,1},
+    {X,E2,2}}}
+  {{{,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
+    {,E0,\lambda x.E1_{x}},
+    {x,E1_{x},\lambda y.E2_{y x}}}}
+
 \marginpar{aggiustare i nomi delle variabili nel testo}
@@
 %
-\printAll
-  {{}}
-  {{X = \lambda x.\lambda y.x, 
-    Z = f\appsep(\lambda x.x)}}
-  {{\mapping{F}{H}{2}}}
-  {{\linketaM{x}{Y_{x}}{\lambda y.H_{y x}},
-    \linketaM{\phantom{x}}{X}{\lambda x.Y_{x}},
-    \linketaM{\phantom{x}}{Z}{\lambda x.f \appsep G_{x}~x},
-    \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}}}
+\printAlll
+  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
+    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
+  {{{E0,\lambda x.\lambda y.x},
+    {E3,E5}}}
+  {{{Y,E5,0},
+    {X,E2,2}}}
+  {{{x,E4_{x},\lambda y.E2_{x y}},
+    {,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
+    {,E0,\lambda x.E1_{x}},
+    {x,E1_{x},\lambda y.E2_{y x}}}}
 

### 2024-03-31 paper/main.tex
fix english 6

@@
 Remark that the converse of \cref{lem:maybeeta} does not hold: 
-there exists a term $t \in \maybeeta$
-such that forall substitution $\rho$, $\rho t$ is not an
+there exists a term $t$ satisfying the criteria (1) above
+that is not in $\maybeeta$, i.e.
+there exists no substitution $\rho$ such that $\rho t$ is an
 $\eta$-expansion. A simple counter example is
 $\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
-which is in \maybeeta{} since $x$ does not \occurrigid{} in the first argument
+since $x$ does not \occurrigid{} in the first argument
 of $f$,
-and the second argument of $f$ \reduceto{} $x$. In other words $A \appsep x$
-may either use or discard $x$, but our analysis does not know that \emph{the same
-term} cannot have two contrasting behaviors.\\
-This is however not a problem in the sense that it does not break
-\cref{prop:simulation}. Terms in \maybeeta{} are compiled to
-unification variables and a link (see \cref{sec:etacomp}):
-the link makes progress (see \cref{sec:etaprogress})
-in the same step in which the
-variable is instantiated, and that compensates for this coarse analysis.
+and the second argument of $f$ \reduceto{} $x$.
+In other words $A \appsep x$
+may either use or discard $x$, but our analysis does not
+take into account that \emph{the same
+term} cannot have two contrasting behaviors.
+
+As we will see in the rest of this section this is not a problem
+since it does not break
+\cref{prop:simulation} nor \cref{prop:fidelity}.
+% A term in \maybeeta{} is compiled to a
+% unification variable and a link (see \cref{sec:etacomp}):
+% the link makes progress (see \cref{sec:etaprogress})
+% in the same step in which the
+% variable is instantiated, and that compensates
+% for this coarse analysis.
 

### 2024-03-31 paper/main.tex
update PBs 6 and 7
reword 6.0 and 6.1

@@
 \section{Handling of \maybeeta}\label{sec:eta}
-\marginpar{TODO: fix syntax}
 $\eta$-reduction is an equivalence relation where a term of the form
 $\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
-free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
+free variable in $t$. We call $t$ the $\eta$-contraction of
+$\lambda x. t \appsep x$.
+
+Following the compilation scheme of \cref{sec:compilation} the
+unification problem $\mathbb{P}$ is compiled as follows:
+%
 \printAlll
@@
   {{}}
-Following the implementation of \elpiIn{comp} given in
-\cref{sec:compilation}, the unification problem $\mathbb{P}_1$ given 
-above, is compiled into the unification problem $\mathbb{T}_1$. 
-We can note that all the terms appearing in $\mathbb{P}_1$ are
-unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathbb{T}_1$:
-its left and right hand sides have both different rigid head:
-$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
-and $f$ corresponds to \elpiIn{con"f"}. 
-This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
-is a term belonging to \maybeeta.
-
-In order to guarantee \cref{prop:simulation}, we need to modify the way terms are
-compiled. The goal is to recognize
-every \maybeeta subterm $t$ and replace it with fresh \Ho variables $v$. 
-This connection between the variable $v$ and the subterm $t$ is stored in what we
-call \linketa which is an object with the following type
+
+\noindent
+While the original problem should be solved by \Uo{} (by assigning $f$ to $X$)
+\Ue{} is not able to solve the problem in $\mathbb{T}$:
+its left and right hand sides have different rigid heads:
+$\lambda x.A_x$ corresponds to the \Ho{} term
+\elpiIn{lam x\ uva A [x]} 
+while $f$ corresponds to
+\elpiIn{con"f"}.
+
+In order to guarantee \cref{prop:simulation} we detect
+lambdas that can disappear by eta contraction (the \maybeeta set
+introduced in \cref{sec:nutshell}); then 
+we modify the way terms are compiled and move terms in \maybeeta
+from  $\mathbb{T}$ to the \linkStore and put fresh unification variables
+in their place.
 
@@
 \noindent
-where, as sketched in \cref{sec:lang-spec}, the term on the left hand side
-(\lhs) is linked with its left counterpart (\rhs).
-
-\linketa are added in the link store (\linkStore) and activated when special
-conditions are satisfied on \lhs or \rhs. These link activations are managed by
-extending the \elpiIn{progress1} predicate (see \cref{sec:execution}).
-We claim that \linketa progression does not contradict \cref{inv:linklhs} and we
-add the following invariant:
+As per \cref{inv:linklhs} the term on the left hand side
+(\lhs) is a variable, and its right counterpart (\rhs) is the
+term in \maybeeta and it has the following property:
 
 \begin{invariant}[\linketa \rhs]
-  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t_x$
-  where $t_x$ is a \maybeeta term and $x$ is free in $t$.
+  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t$
+  and $t$ is not a lambda. 
+  %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
   \label{inv:link-eta-right}
@@
 
-In the next three subsections we explain how we detect \maybeeta terms, how we
-compile them and how the generated \linketa are activated during the \elpiIn{progress}.
-Moreover, we provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.
+\linketa are kept in the link store (\linkStore) during execution
+and activated when some conditions hold on \lhs or \rhs.
+Link activation is implemented by extending the \elpiIn{progress1}
+predicate (defined in \cref{sec:execution}).
 
@@
 
-\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
-  A $\beta$-normal term
-  $\lambda x_1 \ldots x_n.t \in \maybeeta$ (as per \cref{def:maybeeta})
-  % where each $x_i$ occurs in $t$ is a
-  %\maybeeta if
-  %either: 
-  if one of the following holds:
+\newcommand{\testmaybeeta}{\emph{maybe-eta}}
+\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
+  Given a $\beta$-normal term
+  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
+  of the following holds:
   \begin{enumerate}
@@
   \end{enumerate}
+\end{definition}
+\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
+  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
+  then $t \in \maybeeta$.
 \end{lemma}
@@
 Remark that the converse of \cref{lem:maybeeta} does not hold: 
-there exists a term $t$ satisfying the criteria (1) above
-that is not in $\maybeeta$, i.e.
+there exists a term $t$ satisfying the criteria (1) of
+\cref{def:testmaybeeta} that is not in $\maybeeta$, i.e.
 there exists no substitution $\rho$ such that $\rho t$ is an
@@
 
-Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
-terms and, consequently replace them with fresh \Ho unification variables at
-compilation time. The code below illustrate how this relation is used to for
-term compilation.
+% Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
+% terms and, consequently replace them with fresh \Ho unification variables at
+% compilation time. The code below illustrate how this relation is used to for
+% term compilation.
+The following rule is inserted just before rule~\ref{rule:complam} from the code in
+\cref{sec:compilation}.
 
@@
 
-This rule, to be inserted just before rule~\ref{rule:complam} from the code in
-\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
-\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
+\noindent
+The rule verifies if the \Fo{} term $\lambda t$ in input is in
+\maybeeta. Let $\lambda t'$ be the compiled version of $\lambda t$, then the fresh
 variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
@@
 %   {{\linketaM{}{A}{\lambda x.X'_{x}}}}
-$$
-\begin{array}{l}
-  \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
-  \mathbb{T} = \{~ A \Ue f ~\}\\
-  \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
-  \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
-\end{array}
-$$
+% $$
+% \begin{array}{l}
+%   \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
+%   \mathbb{T} = \{~ A \Ue f ~\}\\
+%   \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
+%   \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
+% \end{array}
+% $$
 
@@
 its compiled version, with the corresponding list of mapping and links.
-
-% \printAll
-%   {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
-%     \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
-%   {{E0 == \lambda x.\lambda y.x,
-%     E3 == E5}}
-%   {{\mapping{Y}{E5}{0},
-%     \mapping{X}{E2}{2}}}
-%   {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
-%     \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
-%     \linketaM{}{E0}{\lambda x.E1_{x}},
-%     \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
-$$
-{\arraycolsep=2pt \begin{array}{rrclrcl}
-    \mathbb{P} = \{ &
-       \lambda x.\lambda y.X \appsep y \appsep x & \Uo & \lambda x.\lambda y. x & \quad
-       \lambda x. f \appsep (X \appsep x) \appsep x & \Uo&  Y ~\}\\
-    \mathbb{T} = \{ &
-       A &\Ue& \lambda x.\lambda y.x & \quad
-       D &\Ue& F ~\}\\
-    \mathbb{M} = \{ & \multicolumn{6}{l}{
-      \mapping{Y}{F}{0} \quad
-      \mapping{X}{C}{2} ~\}}\\
-    \linkStore \;= \Big\{  &
-      \multicolumn{6}{l}{ \begin{array}{ll}
-          \linketaM{}{A}{\lambda x.B_{x}} & \quad
-          \linketaM{x}{B_{x}}{\lambda y.C_{y x}}\\
-          \linketaM{}{D}{\lambda x.(f~E_{x}~x)} & \quad
-          \linketaM{y}{E_{y}}{\lambda w.C_{y w}}
-      \end{array} ~\Big\}}
-\end{array}}
-$$
-
+% make test ONLY=7001 TEX=tex
 \printAlll
@@
     {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
-  {{{E0,\lambda x.\lambda y.x},
-    {E3,E5}}}
-  {{{Y,E5,0},
-    {X,E4,1},
-    {X,E2,2}}}
-  {{{,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
-    {,E0,\lambda x.E1_{x}},
-    {x,E1_{x},\lambda y.E2_{y x}}}}
+  {{{A,\lambda x.\lambda y.x},
+    {D,F}}}
+  {{{Y,F,0},
+    {X,E,1},
+    {X,C,2}}}
+  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
+    {,A,\lambda x.B_{x}},
+    {x,B_{x},\lambda y.C_{y x}}}}
 
@@
 The complete problem to run for resolution is now:
-%
+% make test ONLY=7001 TEX=tex
 \printAlll
@@
     {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
-  {{{E0,\lambda x.\lambda y.x},
-    {E3,E5}}}
-  {{{Y,E5,0},
-    {X,E2,2}}}
-  {{{x,E4_{x},\lambda y.E2_{x y}},
-    {,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
-    {,E0,\lambda x.E1_{x}},
-    {x,E1_{x},\lambda y.E2_{y x}}}}
+  {{{A,\lambda x.\lambda y.x},
+    {D,F}}}
+  {{{Y,F,0},
+    {X,C,2}}}
+  {{{x,E_{x},\lambda y.C_{x y}},
+    {,D,\lambda x.(f\appsep E_{x}\appsep x)},
+    {,A,\lambda x.B_{x}},
+    {x,B_{x},\lambda y.C_{y x}}}}
 

### 2024-04-01 paper/main.tex
flip pb ot ease reading
rework 6 up to 6.2

@@
 \noindent
-While the original problem should be solved by \Uo{} (by assigning $f$ to $X$)
-\Ue{} is not able to solve the problem in $\mathbb{T}$:
-its left and right hand sides have different rigid heads:
-$\lambda x.A_x$ corresponds to the \Ho{} term
-\elpiIn{lam x\ uva A [x]} 
-while $f$ corresponds to
-\elpiIn{con"f"}.
+While $\lambda x.X \appsep x \Uo{} f$ does admit the solution
+$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
+$\mathbb{T}$ does not:
+\elpiIn{lam x\ uva A [x]} and
+\elpiIn{con"f"} start with different, rigid, term constructors hence
+\Ue{} fails.
 
 In order to guarantee \cref{prop:simulation} we detect
-lambdas that can disappear by eta contraction (the \maybeeta set
-introduced in \cref{sec:nutshell}); then 
-we modify the way terms are compiled and move terms in \maybeeta
-from  $\mathbb{T}$ to the \linkStore and put fresh unification variables
-in their place.
-
-\elpiIn{type link-eta tm -> tm -> baselink}
-
-\noindent
-As per \cref{inv:linklhs} the term on the left hand side
-(\lhs) is a variable, and its right counterpart (\rhs) is the
-term in \maybeeta and it has the following property:
+lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
+we modify the term compiler put fresh unification variables
+in their place: the problematic term is moved from 
+from  $\mathbb{T}$ to \linkStore (\cref{sec:etacomp}). The compilation
+of the problem $\mathbb{P}$ above is refined to: 
+%
+\printAlll
+  {{{\lambda x.X \appsep x,f}}}
+  {{{A,f}}}
+  {{{X,B,1}}}
+  {{{,A,\lambda x.B_x}}}
+
+\noindent
+As per \cref{inv:linklhs} the term on the left is a variable, and its
+right counterpart is the
+term in \maybeeta, and that term has the following property:
 
 \begin{invariant}[\linketa \rhs]
-  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t$
+  The \rhs of any \linketa %in \linkStore 
+  has the shape $\lambda x.t$
   and $t$ is not a lambda. 
@@
 
-\linketa are kept in the link store (\linkStore) during execution
+\linketa are kept in the link store \linkStore during execution
 and activated when some conditions hold on \lhs or \rhs.
@@
 
-\subsection{Detection of \maybeeta}
+\subsection{Detection of \maybeeta}\label{sec:etadetection}
 
@@
 is not as trivial as it may seems, here a few examples:
-% \\
-% \noindent
-% \begin{minipage}{.5\linewidth}%
-%   \begin{gather}
-%     \lambda x. f \appsep A_x\\
-%     \lambda x. f \appsep x \appsep A_x
-%   \end{gather}
-% \end{minipage}% This must go next to `\end{minipage}`
-% \begin{minipage}{.5\linewidth}
-%   \begin{gather}
-%     \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
-%     \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
-%   \end{gather}
-% \end{minipage}
 %
@@
 or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
-The third example shows when a variable occurs outside the scope of a unification
-variable cannot be erased and hence prevent a term from being an $\eta$-expansion.\\
+The third example shows that when a variable occurs outside the scope of a unification
+variable it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
 The last example shows the recursive nature of the check we need to implement.
@@
 is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
-is in \maybeeta itself, that means it could $\eta$-contract to
+is in \maybeeta itself. If it is, it could $\eta$-contract to
 $f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
@@
 \noindent
-The rule verifies if the \Fo{} term $\lambda t$ in input is in
-\maybeeta. Let $\lambda t'$ be the compiled version of $\lambda t$, then the fresh
-variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
-in $t'$. The critical part of this compilation is the creation of the \linketa,
-which links the variable \elpiIn{A} with $t$. This link creation enforce
-\cref{inv:linklhs} and \cref{inv:link-eta-right}, since \lhs is a trivially a
-variable and the \rhs is a term $t$ starting with the \elpiIn{lam} constructor where
-$x$ is free in $t$ otherwise $t$ would not have been detected as a \maybeeta.
+The rule triggers when the input term \elpiIn{flam F} is in
+\maybeeta. It compiles it to \elpiIn{lam F1} but puts the fresh
+variable \elpiIn{A} in its place. The variable sees all the names free in
+\elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
+which relates the variable \elpiIn{A} with \elpiIn{lam F1}.
+This link clearly validates \cref{inv:linklhs}.
 
 \begin{corollary}
-  The \rhs of any \linketa has exactly one lambda abstraction.
+  The \rhs of any \linketa has exactly one lambda abstraction, hence
+  the rule above respects \cref{inv:link-eta-right}.
 \end{corollary}
@@
 \begin{proof}[Proof sketch]
-  By contradiction, suppose that a \linketa $l$ has $t$ as \rhs and $t = \lambda
-  x.\lambda y.t'_{xy}$. Two cases are to be analyses: 1) $\lambda y.t'_{xy}$ is
-  a $\maybeeta$, then, by construction, \rhs would have been replaced with a the
-  $\eta$-expansion of fresh variable $v$, which is a contradiction, since
-  $\lambda y.t'_{xy} \neq \lambda x.v_{x}$; 2) $\lambda y.t'_{xy}$ is not an a
-  \maybeeta, then neither $t$ is, which is a contradiction since \rhs is always
-  a \maybeeta by construction.
+  By contradiction, suppose that the rule above triggered and that
+  the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
+  If $\testmaybeeta{}~\lambda y.t_{xy}$ holds the recursive call to
+  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
+  in its place, so this case is impossible.
+  Otherwise, if $\testmaybeeta{}~\lambda y.t_{xy}$ does not hold, also
+  $\testmaybeeta{}~\lambda x.\lambda y.t_{xy}$ does not hold, contradicting
+  the assumption that the rule triggered.
 \end{proof}
@@
 
-\linketa are meant to delay the unification of ``problematic'' terms. In the
-following, we call \linkStore the list of suspended links.
-
-In order to activate a \linketa $l$, we need to extend the \elpiIn{progress1} predicate
+\linketa are meant to delay the unification of ``problematic'' terms until
+we know for sure if the head lambda has to be $\eta$-contracted or not.
+
+We extend the \elpiIn{progress1} predicate
 by adding new rules. After passing under all the \elpiIn{abs}
@@
 no more a \maybeeta, the link is removed from \linkStore.
-% \printAll
-%   {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
-%   {{A \Ue f}}
-%   {{\mapping{X}{X'}{1}}}
-%   {{\linketaM{}{A}{\lambda x.X'_{x}}}}
-% $$
-% \begin{array}{l}
-%   \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
-%   \mathbb{T} = \{~ A \Ue f ~\}\\
-%   \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
-%   \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
-% \end{array}
-% $$
-
-\printAlll
-  {{{\lambda x.X \appsep x,f}}}
-  {{{A,f}}}
-  {{{X,B,1}}}
-  {{{,A,\lambda x.B_x}}}
 
@@
     {D,F}}}
-  {{{Y,F,0},
-    {X,E,1},
+  {{{X,E,1},
+    {Y,F,0},
     {X,C,2}}}

### 2024-04-01 paper/main.tex
begin 6.3

@@
 
+TODO: W preservation \cref{prop:nf}
+
 \subsection{Progress}\label{sec:etaprogress}
@@
 we know for sure if the head lambda has to be $\eta$-contracted or not.
+
+\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}}
+\begin{definition}[\progressetaleft]\label{def:progressetaleft}
+A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
+$X$ becomes rigid. There are two cases:
+\begin{enumerate}
+  \item if $X = a$ or $X = f \appsep a_1\ldots a_n$
+    we unify the $\eta$-expansion of the $X$ with $T$, that is we run
+    $\lambda x.X \appsep x\Ue{} T$  (under
+    the context $\Gamma$)
+  \item if $X = \lambda x.t$ we run $X \Ue{} T$.
+\end{enumerate}
+\end{definition}
+
+\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
+\begin{definition}[\progressetaright]\label{def:progressetaright}
+A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
+$\testmaybeeta{} T$ does not hold (anymore) and by $\eta$-contracting
+$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
+the context $\Gamma$).
+\end{definition}
+
+There is a third case in which a link is removed from \linkStore, namely
+when the \lhs is assigned to another variable that is the \lhs of another
+\linketa.
+
+\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}}
+\begin{definition}[\progressetadedup]\label{def:progressetadedup}
+  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
+  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
+  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
+  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
+  bound variables, i.e. $T'' = T[\vec{r}/\vec{s}]$.
+  We then run $X_S \Ue{} T''$ (under the context $\Gamma$).
+\end{definition}
 

### 2024-04-01 paper/main.tex
examples 6

@@
 \begin{enumerate}
-  \item if $X = a$ or $X = f \appsep a_1\ldots a_n$
+  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
     we unify the $\eta$-expansion of the $X$ with $T$, that is we run
@@
   \item if $X = \lambda x.t$ we run $X \Ue{} T$.
+  \marginpar{do we have the name case in the code?}
 \end{enumerate}
@@
 the context $\Gamma$).
+\marginpar{there was a note about T' having a rigid head, TO BE FIXED}
 \end{definition}
@@
   hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
-  bound variables, i.e. $T'' = T[\vec{r}/\vec{s}]$.
-  We then run $X_S \Ue{} T''$ (under the context $\Gamma$).
+  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
+  We then run $T \Ue{} T''$ (under the context $\Gamma$).
 \end{definition}
 
-We extend the \elpiIn{progress1} predicate
-by adding new rules. After passing under all the \elpiIn{abs}
-constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
-is instantiated to a rigid term 2) \rhs is no more a \maybeeta or it is a term
-which can be reduced to a term with rigid head. 
-If \lhs is instantiated to a rigid term $t$, by \cref{prop:nf},
-we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
-if $t$ is a constant or a function application, then, $t'$, which by
-construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
-therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
-of $t$) with $t'$. Finally, if $t$ is a term with \elpiIn{lam} as head, then
-it is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.
-
-The second way to activate a \linketa is when the \rhs is no more a \maybeeta or
-\rhs can be $\eta$-reduced to a term $t$ with rigid head. In both cases, \lhs is
-unified with $t'$.
-
-Once a \linketa is activated, it can be removed from \linkStore,
-otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
-that this link progression enforce
-\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
-the \Ho substitution, since we make unification only when we know that the 
+TODO: prove \cref{prop:nf}: we never commit a \maybeeta{} term in
+$\sigma$ since we run \Ue{} only when we know that the 
 terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
 no more a \maybeeta, the link is removed from \linkStore.
-
-The example above shows the new compilation of the unification problem given at
-the beginning of \cref{sec:eta}. This time, we see that the the left hand side
-$t$ of $\mathbb{P}_1$ has been detected as a \maybeeta and replaced with the
-fresh variable $A$. Moreover, \linkStore contains the link $l_1$ connecting $A$
-with $\lambda x.X'_{x}$, the compiled version of $t$. After the resolution of
-$\mathbb{T}_1$, $A$ is assigned to $f$. Therefore \lhs of $l_1$ is a term with
-\elpiIn{con} as constructor. This means that \rhs of $l_1$ is a $\eta$-expansion
-and therefore we can unify $\lambda x.f x$ with \rhs, which instantiate $X'_x$
-to $f \appsep x$. $l_1$ is removed from \linkStore which is now empty,
-\elpiIn{progress} terminate, and the decompilation, will instantiate the $X$
-variable to $f$ (which is the $\eta$-contracted version of $\lambda x.f \appsep
-x$).
-
-A second example, showing the activation of a link when the \rhs is no more a
-\maybeeta, is given in \cref{sec:invariant1}, since we need to work with
-variables used with different arities. This example represent the run of the
-unification problems proposed at \cref{sec:basic-comp-limitations}
-
-Another, way to progress \linketa, that we call \linketa deduplication, is when
-\linkStore contains two \linketa $l_1$ and $l_2$ with a \lhs having the same
-variable address. Let the \lhs of $l_1$ be \elpiIn{uva U R} and the \lhs of
-$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:uvaarity}, \elpiIn{R}
-and \elpiIn{S} have same scope. Let $t$ be the term obtained by replacing all
-each name \elpiIn{S}$_i$ in the \rhs of $l_1$ with \elpiIn{R}$_i$, $t$ is
-unified with the \rhs of $l_2$ and one of the two links between $l_1$ and $l_2$
-is removed from \linkStore.
-
-A very basic example of \linketa deduplication, is given below:
-
-\printAlll
-  {{{\lambda x. X \appsep x, \lambda x. X \appsep x}}}
-  {{{E0, E2}}}
-  {{{X,A1,1}}}
-  {{{,E2,\lambda x.A1_{x}},
-    {,E0,\lambda x.A1_{x}}}}
-
-TODO: explain
 
@@
 \end{proof}
+
+\paragraph{example 1}
+
+The example above shows the new compilation of the unification problem given at
+the beginning of \cref{sec:eta}. This time, we see that the the left hand side
+$t$ of $\mathbb{P}_1$ has been detected as a \maybeeta and replaced with the
+fresh variable $A$. Moreover, \linkStore contains the link $l_1$ connecting $A$
+with $\lambda x.X'_{x}$, the compiled version of $t$. After the resolution of
+$\mathbb{T}_1$, $A$ is assigned to $f$. Therefore \lhs of $l_1$ is a term with
+\elpiIn{con} as constructor. This means that \rhs of $l_1$ is a $\eta$-expansion
+and therefore we can unify $\lambda x.f x$ with \rhs, which instantiate $X'_x$
+to $f \appsep x$. $l_1$ is removed from \linkStore which is now empty,
+\elpiIn{progress} terminate, and the decompilation, will instantiate the $X$
+variable to $f$ (which is the $\eta$-contracted version of $\lambda x.f \appsep
+x$).
+
+\paragraph{example 2}
+
+A second example, showing the activation of a link when the \rhs is no more a
+\maybeeta, is given in \cref{sec:invariant1}, since we need to work with
+variables used with different arities. This example represent the run of the
+unification problems proposed at \cref{sec:basic-comp-limitations}
+
+\paragraph{example 3}
+
+A very basic example of \linketa deduplication, is given below:
+% make test ONLY=7002 TEX=tex
+\printAlll
+  {{{\lambda x.(X\appsep x),\lambda x.(Y\appsep x)}}}
+  {{{A,C}}}
+  {{{X,B,1},
+    {Y,D,1}}}
+  {{{,A,\lambda x.B_{x}},
+    {,C,\lambda x.D_{x}}}}
+
+\noindent
+The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
+By unifying the two \rhs we get
+$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
+In turn, given the map $\mathbb{M}$, this second assignment is decompiled to
+$\rho = \{~ X \mapsto Y ~\}$ as expected.
 

### 2024-04-01 paper/main.tex
happy with 6, 1 example to double check

@@
 Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
-and justify why the don't hinder termination.
+and justify why the don't hinder termination. For brevity we omit the code
+that applies the substitution \elpiIn{S1} to all terms in \linkStore.
 
@@
 \begin{lemma}
-  Forall list of links \elpiIn{L} and \elpiIn{S}, \elpiIn{progress L _ S _}
-  terminates
+  \elpiIn{progress} terminates.
 \end{lemma}
@@
 \begin{proof}[Proof sketch]
-  The addition of rules for \elpiIn{progres1} complicates the function
-  \elpiIn{progress}. We can note, however, that they do not prevent the
-  termination of \elpiIn{progress}. 1) If a link is activated it is removed from
-  \linkStore and the recursive call to \elpiIn{progress} will have a smaller
-  list of links to recurse on. Moreover, link activation only runs terminating
-  instructions (such as unification). 2) If a link is deduplicated, the
-  termination of \elpiIn{progress} is still guaranteed since again we reduce
-  \linkStore and the instructions run by link deduplications are all
-  terminating. 3) If a link is neither activated nor deduplicated, i.e. it
-  remains suspended, then \linkStore remains unchanged like the substitution;
-  therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
-  terminates.
+  Rules \cref{def:progressetaleft,def:progressetaright} and
+  \cref{def:progressetadedup} remove one link from \linkStore, hence they
+  cannot be applied indefinitely.
+  Moreover each rule only relies on terminating operations such as \Ue,
+  $\eta$-contraction, $\eta$-expansion, relocation (a recursive copy of a
+  finite term).
+  % The addition of rules for \elpiIn{progres1} complicates the function
+  % \elpiIn{progress}. We can note, however, that they do not prevent the
+  % termination of \elpiIn{progress}. 1) If a link is activated it is removed from
+  % \linkStore and the recursive call to \elpiIn{progress} will have a smaller
+  % list of links to recurse on. Moreover, link activation only runs terminating
+  % instructions (such as unification). 2) If a link is deduplicated, the
+  % termination of \elpiIn{progress} is still guaranteed since again we reduce
+  % \linkStore and the instructions run by link deduplications are all
+  % terminating. 3) If a link is neither activated nor deduplicated, i.e. it
+  % remains suspended, then \linkStore remains unchanged like the substitution;
+  % therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
+  % terminates.
 \end{proof}
 
-\paragraph{example 1}
-
-The example above shows the new compilation of the unification problem given at
-the beginning of \cref{sec:eta}. This time, we see that the the left hand side
-$t$ of $\mathbb{P}_1$ has been detected as a \maybeeta and replaced with the
-fresh variable $A$. Moreover, \linkStore contains the link $l_1$ connecting $A$
-with $\lambda x.X'_{x}$, the compiled version of $t$. After the resolution of
-$\mathbb{T}_1$, $A$ is assigned to $f$. Therefore \lhs of $l_1$ is a term with
-\elpiIn{con} as constructor. This means that \rhs of $l_1$ is a $\eta$-expansion
-and therefore we can unify $\lambda x.f x$ with \rhs, which instantiate $X'_x$
-to $f \appsep x$. $l_1$ is removed from \linkStore which is now empty,
-\elpiIn{progress} terminate, and the decompilation, will instantiate the $X$
-variable to $f$ (which is the $\eta$-contracted version of $\lambda x.f \appsep
-x$).
-
-\paragraph{example 2}
+\paragraph{Example of \progressetaleft}
+
+The example at the beginning of \cref{sec:eta}, once
+$\sigma = \{~ A \mapsto f ~\}$, triggers this rule since the link
+becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
+In turn the rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
+resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
+Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
+$X$ is mapped to $B$ and
+$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.
+
+\paragraph{Example of \progressetaright}\marginpar{TODO}
 
@@
 
-\paragraph{example 3}
+\paragraph{Example of \progressetadedup}
 
@@
 
+\marginpar{I don't get this TODO}
 TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does

### 2024-04-01 paper/main.tex
shorten 7

@@
 \label{sec:basic-comp-limitations}
-
-\begin{gather}
-\lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
-\lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
-% \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
-\end{gather}
-Note that here $F$ is used with different arities, moreover
+The basic compilation scheme is not about to
+deal wit the following problem:
+\printAlll
+  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,G}}}
+  {{}}
+  {{}}
+  {{}}
+\marginpar{TODO: remove empty T}
+% \begin{gather}
+% \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
+% \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
+% % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
+% \end{gather}
+
+noindent Note that here $X$ is used with different arities, moreover
 in the second problem the left hand side happens to be an
 eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
-that $F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
-Both problems are addressed in the next section.
+that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
+Both problems are addressed in the next two sections.
 
@@
 
-In \cref{sec:basic-comp-limitations}, we have given two unification problems to
-be run one after the other. In the following table we present the entry problem,
-its compiled version, with the corresponding list of mapping and links.
+We report here the problem given in \cref{sec:basic-comp-limitations} where
+$X$ is used with two different arities and the output of the compilation does
+not respect\cref{inv:map-bijective} (merging the two mappings
+for $s$ would break \ref{inv:uvaarity}).
+In this section we explain how to replace the duplicate mapping with some
+\linketa{} in order to restore the invariants.
 % make test ONLY=7001 TEX=tex
@@
 
-\marginpar{aggiustare i nomi delle variabili nel testo}
-We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
-x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
-respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
-has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
-is linked to the term $\lambda x.f \appsep G_{x}~x$. However, the mapping
-returned by the compilation, does not breaks \cref{inv:map-bijective}: the \Fo{}
-variable $F$ is mapped to two different \Ho variables. To address this problem
-and enforce \cref{inv:map-bijective}, we clean the mapping with a second phase
-after the compilation. This phase is called \elpiIn{map-deduplication}.
-
-Before formally defining this procedure, we need to define some auxiliary
-relations. Let \mapStore be the list of mapping and $\langle m_1, m_2 \rangle
-\in \mapStore$ such that the arity of the \Ho variable in $m_1$ is smaller than
-the one in $m_2$. Let $X$ (resp. $Y$) the \Ho variable of $m_1$ (resp. $m_2$)
-and $n = ar(m_1) - ar(m_2)$. We also let $A^i$ be a fresh \Ho variable. We
-define the \elpiIn{make-eta-link} relation taking two mappings $\langle m_1, m_2
-\rangle$ and returning the following list of link \linketa: $\forall i \in [1..n]$,
-{
-\def\scopeA{\ensuremath{x_1\dots x_{i-1}}}
-\def\scopeB{\ensuremath{x_1\dots x_{i}}}
-%
-$$ 
-  \left\{ 
-    \begin{array}{ll}
-      \linketaM{\phantom{\scopeA}}{X}      {\lambda x.  A^1_x}         & if~ i = 1 \\
-      \linketaM{\scopeA}{A^{i-1}_{\scopeA}}{\lambda x_i.A^i_{\scopeB}} & if~ 1 < i < n\\
-      \linketaM{\scopeA}{A^{i-1}_{\scopeA}}{\lambda x_i.Y  _{\scopeB}} & if~ i = n
-    \end{array} 
-  \right. 
-$$
-}
-\noindent
-More concretely, we are saying that for any two mappings, we build as many
-\linketa as the difference of the arities between the two mappings. This links
-are constructed in such a way that the \Ho variable $v$ with lowest arity is
-linked to a fresh variable eta-expended variable $A^1$ having the scope of $v$.
-This variable $A^1$ is then linked to a an $\eta$-expanded fresh variable $A^2$
-with same scope of $A^1$ and so on. The last link is built between the $A^{n-1}$
-(where $n$ is the difference of arities between the two mappings) and the \Ho
-variable $u$ with higher arity in the two mappings being considered.
-
-\begin{definition}[map-deduplication]
-  Forall mappings $\langle m_1, m_2 \rangle \in \mapStore$, sharing the same
-  \Fo{} variable, the list of \linketa $L$ is created thanks to
-  \elpiIn{make-eta-link}$m_1~m_2~L$ and is added to \linkStore. Then $m_1$ is
-  removed from \mapStore.
+We see that the \elpiIn{maybe-eta} as identified $\lambda xy.X \appsep y \appsep
+x$ and $\lambda x.f\appsep (X \appsep x) \appsep x$ and the compiler has
+replaced them with $A$ and $D$ respectively.
+% $A$ is linked with $\lambda x.B_x$, $B$
+% has arity $1$ and is $\eta$-linked with $\lambda y.C\appsep y \appsep x$ and $D$
+% is linked to the term $\lambda x.f \appsep E_{x}~x$. 
+However, the mapping \mapStore breaks \cref{inv:map-bijective}: the \Fo{}
+variable $X$ is mapped to two different \Ho variables. To address this problem
+we adjust the compiler's output with a \elpiIn{map-deduplication} procedure.
+
+\newcommand{\alignarity}{\emph{align-arity}}
+\begin{definition}[\alignarity] Given two mappings
+  $m_1 : X \mapsto A^m$ and
+  $m_2 : X \mapsto C^n$ where $m < n$ and $d = n - m$,
+  $\alignarity{}~ m_1 ~ m_2$ generates the following $d$ links, one
+  for each $i$ such that $0 \leq i < d$,
+\[%
+  \linketaM{x_0 \ldots x_{m+i}}
+           {B^i_{x_0 \ldots x_{m+i}}}
+           {\lambda x_{m+i+1}.B^{i+1}_{x_0 \ldots x_{m+i+1}}}
+\]%
+where $B^i$ is a fresh variable of arity $m+i$, and $B^0 = A$ as well as $B^d = C$.
 \end{definition}
 
-If we take back the example give at the beginning of this section, we can
-deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
-and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x y}}.
-
-The complete problem to run for resolution is now:
+The intuition is that we $\eta$-expand the occurrence of the variable
+with lower arity to match the higher arity. Since each \linketa can
+add exactly one lambda, we need as many links as the difference between the
+two arities.
+
+\newcommand{\mapdeduplication}{\emph{map-deduplication}}
+
+\begin{definition}[\mapdeduplication]
+  Forall mappings $m_1, m_2\in \mapStore$ such that 
+  $m_1 : X \mapsto A^m$ and
+  $m_2 : X \mapsto C^n$ and $m < n$
+  we remove $m_1$ from \mapStore and
+  add to \linkStore the result of $\alignarity{}~m_1~m_2$.
+\end{definition}
+
+If we look back the example give at the beginning of this section, we can
+deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
+and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
+After deduplication the compiler output is as follows:
 % make test ONLY=7001 TEX=tex
@@
     {x,B_{x},\lambda y.C_{y x}}}}
-
-After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
-This assignment makes $l_2$ to progress since the \lhs is materialized and by
-unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
-y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
-After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
-\elpiIn{progress} fixpoint terminates. Next, the second unification problem is
-run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
-and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
-$Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
-As last step, the last link is progressed and the final \Ho substitution is
-$\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
-G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
-H _{x} \mapsto \lambda y.y\}$.
-
-The decompilation phase is only charged, in this example to solve the mappings,
-since no suspended links remain. The only mapping in the list is
-\mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$
+\marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
+% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
+% This assignment makes $l_2$ to progress since the \lhs is materialized and by
+% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
+% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
+% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
+% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
+% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
+% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
+% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
+% As last step, the last link is progressed and the final \Ho substitution is
+% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
+% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
+% H _{x} \mapsto \lambda y.y\}$.
+% The decompilation phase is only charged, in this example to solve the mappings,
+% since no suspended links remain. The only mapping in the list is
+% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$
 

### 2024-04-01 paper/main.tex
more \def, adjust printer_pb (links on 3 col + empty T)

@@
 \usepackage{enumitem}
+\usepackage{intcalc}
 
@@
 \newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
-\def\phantx{\phantom{x}}
+
+\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
+\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
+\newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
+\newcommand{\hoUnifPb}{\ensuremath{\mathbb{T}}\xspace}
 
@@
 a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
-unification problem between terms $\mathbb{S}_{p_l}$ and
-$\mathbb{S}_{p_r}$ taken from the set of all terms $\mathbb{S}$.
+unification problem between terms $\foUnifPb_{p_l}$ and
+$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
 The composition of these steps starting from the
@@
 \newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
-\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
-\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
 %
@@
 \begin{array}{l}
-\stepF{\mathbb{S}}{p}{\rho}{\rho''}
+\stepF{\foUnifPb}{p}{\rho}{\rho''}
 \deff
-\stepFD{\mathbb{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
-\runF{\mathbb{S}}{\mathcal{N}}{\rho}
+\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
+\runF{\foUnifPb}{\mathcal{N}}{\rho}
 \deff
-\runFD{\mathbb{S}}{\mathcal{N}}
+\runFD{\foUnifPb}{\mathcal{N}}
 \end{array}
@@
 \begin{array}{l}
-\stepH{\mathbb{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
-  \qquad\stepHD{\mathbb{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
-\runH{\mathbb{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \mathbb{T} \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathbb{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
-  \qquad \runHD{\mathbb{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
+\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
+  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
+\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
+  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
   \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
@@
 \begin{proposition}[Simulation]\label{prop:simulation}
-$\forall \mathbb{S}, \forall \mathcal{N},$
+$\forall \foUnifPb, \forall \mathcal{N},$
 $$
-  \runF{\mathbb{S}}{\mathcal{N}}{\rho}
+  \runF{\foUnifPb}{\mathcal{N}}{\rho}
   \Leftrightarrow
-  \runH{\mathbb{S}}{\mathcal{N}}{\rho}
+  \runH{\foUnifPb}{\mathcal{N}}{\rho}
 $$
@@
 \begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of~ \hrun, if $~\mathbb{T} \subseteq \llambda$ we have that
+In the context of~ \hrun, if $~\hoUnifPb \subseteq \llambda$ we have that
 $\forall p \in 1 \ldots \mathcal{N},$
 $$
-\stepF{\mathbb{S}}{p}{\rho_{p-1}}{\rho_p}
+\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
 \Leftrightarrow
-\stepH{\mathbb{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
+\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
 $$
@@
 We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
-$\mathbb{S} = \{ s_1, s_2 \}$ as follows:
+$\foUnifPb = \{ s_1, s_2 \}$ as follows:
 %
@@
 \begin{proposition}[\wellb{}-preservation]\label{prop:nf}
-$\forall \mathbb{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
+$\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
 $$
 \begin{array}{l}
-\wellb(\sigma\mathbb{T}) \land
-  \sigma\mathbb{T}_{p_l} \Ue \sigma\mathbb{T}_{p_r} \mapsto {\sigma'}
-  \Rightarrow \wellb(\sigma' \mathbb{T})\\
-\wellb(\sigma\mathbb{T}) \land
+\wellb(\sigma\hoUnifPb) \land
+  \sigma\hoUnifPb_{p_l} \Ue \sigma\hoUnifPb_{p_r} \mapsto {\sigma'}
+  \Rightarrow \wellb(\sigma' \hoUnifPb)\\
+\wellb(\sigma\hoUnifPb) \land
   \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
-  \Rightarrow \wellb(\sigma' \mathbb{T})
+  \Rightarrow \wellb(\sigma' \hoUnifPb)
 \end{array}
@@
 \printAlll
-  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,G}}}
+  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,Y}}}
   {{}}
@@
   {{}}
-\marginpar{TODO: remove empty T}
 % \begin{gather}
@@
 
-noindent Note that here $X$ is used with different arities, moreover
+\noindent Note that here $X$ is used with different arities, moreover
 in the second problem the left hand side happens to be an
@@
 Following the compilation scheme of \cref{sec:compilation} the
-unification problem $\mathbb{P}$ is compiled as follows:
+unification problem \foUnifPb is compiled as follows:
 %
@@
 $\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
-$\mathbb{T}$ does not:
+\hoUnifPb does not:
 \elpiIn{lam x\ uva A [x]} and
@@
 lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
-we modify the term compiler put fresh unification variables
+we modify the term compiler by putting fresh unification variables
 in their place: the problematic term is moved from 
-from  $\mathbb{T}$ to \linkStore (\cref{sec:etacomp}). The compilation
-of the problem $\mathbb{P}$ above is refined to: 
+from  \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
+of the problem \foUnifPb above is refined to: 
 %
@@
 
-\newcommand{\testmaybeeta}{\emph{maybe-eta}}
+\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
 \begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
@@
 A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
-$\testmaybeeta{} T$ does not hold (anymore) and by $\eta$-contracting
+$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
 $T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
@@
 $X$ is used with two different arities and the output of the compilation does
-not respect\cref{inv:map-bijective} (merging the two mappings
-for $s$ would break \ref{inv:uvaarity}).
+not respect \cref{inv:map-bijective} (merging the two mappings
+for $s$ would break \cref{inv:uvaarity}).
 In this section we explain how to replace the duplicate mapping with some

### 2024-04-01 paper/main.tex
update intro of maybe-beta

@@
 $\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
-In turn, given the map $\mathbb{M}$, this second assignment is decompiled to
+In turn, given the map \mapStore, this second assignment is decompiled to
 $\rho = \{~ X \mapsto Y ~\}$ as expected.
@@
 
-TODO: say that maybe-eta also work in not(llambda)
-
 % On
@@
 
-All the previous sections we have dealt with terms in \llambda, however, it is
-often possible to work with terms in \maybebeta and wish to unify them. There
-are situtation, for example, where the oracle has given  which $\beta$-reduction problems (\maybebeta) appears any
-time we deal with a subterm $t = X t_1 \dots t_n$, where $X$ is flexible and the
-list $[t_1 \dots t_n]$ in not in \llambda. This unification problem is not
-solvable without loss of generality, since there is not a most general unifier.
-If we take back the example given in \cref{sec:nutshell}, the unification $F a =
-a$ admits two solutions for $F$: $\rho_1 = \{F \mapsto \lambda x.x\}$ and
-$\rho_2 = \{F \mapsto \lambda \_.a\}$. Despite this, it is possible to work with
-$\maybebeta$ if an oracle provides a substitution $\rho$ such that $\rho t$
-falls again in the \llambda.
-
-On the other hand, the \Ue is not designed to understand how the $\beta$-redexes
-work in the object language. Therefore, even if we know that $F$ is assigned
-to $\lambda x.x$, \Ue is not able to unify $F a$ with $a$. On the other hand,
-the problem $F a = G$ is solvable by \Ue, but the final result is that $G$ is
-assigned to $(\lambda x.x) a$ which breaks the invariant saying that the
-substitution of the meta language does not generate terms outide \wellb{} (Property \ref{prop:nf}).
-
-The solution to this problem is to modify the compiler such that any sub-term $t$
-considered as a potential $\beta$-redex is replaced with a hole $h$ and a new dedicated
-link, called \linkbeta.
-
-\begin{elpicode}
-  type link-beta tm -> tm -> link.
-\end{elpicode}
-
-\def\rhs{\ensuremath{rhs}\xspace}
-\def\lhs{\ensuremath{lhs}\xspace}
-
-This link carries two terms, the former representing the variable $h$ for the
-new created hole and the latter containing the subterm $t$. As for the \linketa,
-we will call $h$ and $t$ respectively the left hand side (\lhs)
-and the right hand side (\rhs) of the \linkbeta.
+Until now, we have only dealt we unification of terms in \llambda. However, we
+want the unification relation to be more robust so that it can work with terms
+in \maybebeta. Unification in \maybebeta is in general a non-decidable
+procedure, e.g. $X \appsep a \Uo a$ is a unification problem in \maybebeta,
+since we have $X a$ has the shape $X t_1\dots t_n$ where $X$ is a unification
+variable and $t_1\dots t_n$ is not a list of distinct names (in our example, $a$
+is a constant, hence not a name). We also point out that this unification
+problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
+and $\rho_2 = \{X \mapsto \lambda \_.a\}$.
+
+It is the case, however, given a list of unification $n$ problems \foUnifPb, if
+$\foUnifPb_i$ is in \maybebeta, it is possible that the resolution of the
+problems $\bigwedge_{j=0}^{i-1}\foUnifPb_j$ gives a partial substitution
+$\rho$, such that $\rho\foUnifPb_i$ falls again in \llambda.
+%
+\printAlll
+  {{{X,\lambda x.Y},
+    {(X\appsep a),a}}}
+  {{{A,\lambda x.B},
+    {(A\appsep a),a}}}
+  {{{Y,B,0},
+    {X,A,0}}}
+  {{}}
+
+In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
+$\foUnifPb_2$, can be solved in \llambda. On the other hand, even though \Uo is
+able to solve \foUnifPb, \Ue can't solve the compiled problems \hoUnifPb. In
+fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
+\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
+not unifable problem $(\lambda x. B) \appsep a \Ue a$.
+
+In order to encompass this unification, we capture terms $t$ in \maybebeta and
+replace them with fresh varaibles $v$. As per \maybebeta, the variables $v$ and
+the terms $t$ are linked through a \linkbeta.
+
+\linkbeta guarantee \cref{inv:linklhs} and the term on the \rhs has the
+following property:
+
+\begin{invariant}[\linkbeta \rhs]
+  The \rhs of any \linkbeta has the shape $X t_1\dots t_n$ such that $X$ is a
+  flexible variable and $t_1\dots t_n$ is not in \llambda.
+\end{invariant}
+
+\linkbeta are put in \linkStore and activated when \rhs falls in \llambda.
+
+% This property can be relaxed, to accept approximations: a dedicated
+% section is given in a future section
+\begin{corollary}
+  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
+  counterpart is still in \maybebeta, the current unification problem is not in
+  \llambda and the unification fails.
+\end{corollary}
+
+\begin{proof}[Proof sketch]
+  Given $X \appsep t_1\dots t_n \Ue t$ where $t$ is a rigid term and
+  $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
+  is replaced with a variable $V$, and the \linkbeta \linkbetaM{}{V}{X
+  \appsep t_1\dots t_n} is created. The unification instantiates $V$ to $t$,
+  making the \lhs of the link a rigid term, while \rhs is still in \maybebeta.
+  The original problem is in fact outside \llambda and unification fails.
+\end{proof}
 
@@
 
-\paragraph{Detection of \maybebeta} TODO: \dots
-
-\paragraph{Compilation with \linkbeta}
-
-In order to build a \linkbeta, we need to adapt the compiler so that it can
-recognize these ``problematic'' subterms. The following code snippet illustrate
-such behavior, we suppose the rule to be added just after \cref{rule:comp-pf}.
+Detection of \maybebeta is quite simple to implement in the compiler, it is 
+sufficient to
+
+Dire che maybebeta is simple to detect e che maybe eta funziona bene and su 
+termini che non sono il llambda
+
+\begin{corollary}
+  Right hand side is $X_{scope} t_1 \dots t_n$ and $t_1$ is in $scope$ or
+  $t_1$ is not a name
+\end{corollary}
 

### 2024-04-01 paper/main.tex
adapt tex to print beta links
Update maybebeta compilation section

@@
 \subsection{Compilation}
+\label{sec:compilation}
 \marginpar{manca beta normal in entrata}
@@
   {{{X,B,1}}}
-  {{{,A,\lambda x.B_x}}}
+  {{{\eta,,A,\lambda x.B_x}}}
 
@@
     {Y,D,1}}}
-  {{{,A,\lambda x.B_{x}},
-    {,C,\lambda x.D_{x}}}}
+  {{{\eta,,A,\lambda x.B_{x}},
+    {\eta,,C,\lambda x.D_{x}}}}
 
@@
     {X,C,2}}}
-  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
-    {,A,\lambda x.B_{x}},
-    {x,B_{x},\lambda y.C_{y x}}}}
+  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
+    {\eta,,A,\lambda x.B_{x}},
+    {\eta,x,B_{x},\lambda y.C_{y x}}}}
 
@@
   {{{x,E_{x},\lambda y.C_{x y}},
-    {,D,\lambda x.(f\appsep E_{x}\appsep x)},
-    {,A,\lambda x.B_{x}},
-    {x,B_{x},\lambda y.C_{y x}}}}
+    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
+    {\eta,,A,\lambda x.B_{x}},
+    {\eta,x,B_{x},\lambda y.C_{y x}}}}
 \marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
@@
 
-Detection of \maybebeta is quite simple to implement in the compiler, it is 
-sufficient to
-
-Dire che maybebeta is simple to detect e che maybe eta funziona bene and su 
-termini che non sono il llambda
+Detection of \maybebeta is quite simple to implement in the compiler, since it
+is sufficient to capture applications with flexible head and argument that
+are not in \llambda.
+
+\input{code/comp_beta}
+
+The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
+that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
+\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
+the application of a fresh variable \elpiIn{C} having in scope all the compiled
+list of terms obtained from \elpiIn{Pf} and \elpiIn{Extra1} as arguments. The
+variable \elpiIn{B} returned has compiled terms is a fresh variable having in
+scope all the free variables used in \rhs.
+
+\begin{invariant}
+  The \rhs of a \linkbeta has the shape \elpiIn{app[uva N S|L]}.
+\end{invariant}
 
 \begin{corollary}
-  Right hand side is $X_{scope} t_1 \dots t_n$ and $t_1$ is in $scope$ or
-  $t_1$ is not a name
+  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then \elpiIn{L} is not
+  empty.
 \end{corollary}
 
-\input{code/comp_beta}
-
-A term is \maybebeta if it has the shape \elpiIn{fapp[fuva A|Ag]} and
-\elpiIn{distinct Ag} does not hold. In that case, \elpiIn{Ag} is split in two
-sublist \elpiIn{Pf} and \elpiIn{Extra} such that former is the longest prefix of
-\elpiIn{Ag} such that \elpiIn{distinct Pf} holds. \elpiIn{Extra} is the list
-such that \elpiIn{append Pf Extra Ag}. Next important step is to compile
-recursively the terms of these lists and allocate a memory adress \elpiIn{B}
-from the substitution in order to map the \Fo variable \elpiIn{fuva A} to
-the \Ho variable \elpiIn{uva B}. The \linkbeta to return in the end is given
-by the term \elpiIn{Beta = app[uva B Scope1 | Extra1]} constituting the \rhs,
-and a fresh variable \elpiIn{C} having in scope all the free variables occurring
-in \elpiIn{Beta} (this is \lhs). We point out that the \rhs is intentionally
-built as an \elpiIn{uva} where \elpiIn{Extra1} are not in scope, since by
-invariant, we want all the variables appearing in \Ho to be in \llambda.
+\begin{proof}[Proof sketch]
+  By contradiction, if \elpiIn{L} is a empty list, then the original \Fo{}
+  term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of 
+  disting names. This case is however captured by rule~\ref{rule:complam} (from
+  \cref{sec:compilation}) and no \linkbeta is produced: a contradiction.
+\end{proof}
+
+\begin{corollary}
+  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then the 
+  first argument $t$ in \elpiIn{L} either appears in \elpiIn{S} or it is 
+  not a name.
+\end{corollary}
+
+\begin{proof}[Proof sketch]
+  By construction, the lists \elpiIn{S} and \elpiIn{L} are built by splitting
+  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
+  \elpiIn{S} is the longest prefix of the compiled terms in \elpiIn{Ag} which is
+  in \llambda. Therefore, by definition of \llambda, either the first element of
+  \elpiIn{L} is a name appearing in \elpiIn{S} or it a term with a constructor
+  of \elpiIn{tm} as functor.  
+\end{proof}
+
+TODO: Dire che maybe eta fa il detect anche su 
+termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
+esempio?
 

### 2024-04-02 paper/main.tex
type in call of printAlll
Update beta sections

@@
 lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
-we modify the term compiler by putting fresh unification variables
-in their place: the problematic term is moved from 
+we modify the compiled terms by putting fresh unification variables
+in their place: the problematic term is moved 
 from  \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
@@
 hence if the term $s$ is not exactly $x$ (case 1)
-it can only be an $\eta$-expansion of it, or a unification
-variable that can be assigned to it, or a combination of both.
+it can only be an $\eta$-expansion of $x$, or a unification
+variable that can be assigned to $x$, or a combination of both.
 If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
@@
   \begin{enumerate}
-    \item $t$ is a constant or variable applied to arguments
+    \item $t$ is a constant or a variable applied to the arguments
       $l_1 \ldots l_m$ such that 
@@
     {X,C,2}}}
-  {{{x,E_{x},\lambda y.C_{x y}},
+  {{{\eta,x,E_{x},\lambda y.C_{x y}},
     {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
@@
 %
+% make test ONLY=7003 TEX=TEX + deactivate beta tex
 \printAlll
@@
 In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
-$\foUnifPb_2$, can be solved in \llambda. On the other hand, even though \Uo is
-able to solve \foUnifPb, \Ue can't solve the compiled problems \hoUnifPb. In
+$\foUnifPb_2$, can be solved in \llambda. On the other hand, we see that, 
+\Ue can't solve the compiled problems \hoUnifPb. In
 fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
 \mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
-not unifable problem $(\lambda x. B) \appsep a \Ue a$.
-
-In order to encompass this unification, we capture terms $t$ in \maybebeta and
-replace them with fresh varaibles $v$. As per \maybebeta, the variables $v$ and
-the terms $t$ are linked through a \linkbeta.
-
-\linkbeta guarantee \cref{inv:linklhs} and the term on the \rhs has the
+non-unifable problem $(\lambda x. B) \appsep a \Ue a$.
+
+In order to encompass this unification, term compilation should capture the
+terms $t$ in \maybebeta and replace them with fresh varaibles $v$. As per
+\maybebeta, the variables $v$ and the terms $t$ are linked through a \linkbeta.
+
+\linkbeta guarantees \cref{inv:linklhs} and the term on the \rhs has the
 following property:
@@
 \begin{invariant}[\linkbeta \rhs]
-  The \rhs of any \linkbeta has the shape $X t_1\dots t_n$ such that $X$ is a
+  The \rhs of any \linkbeta has the shape $X~t_1\dots t_n$ such that $X$ is a
   flexible variable and $t_1\dots t_n$ is not in \llambda.
@@
   $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
-  is replaced with a variable $V$, and the \linkbeta \linkbetaM{}{V}{X
+  is replaced with a variable $V$, and the \linkbeta \linkbetaM{\Gamma}{V}{X
   \appsep t_1\dots t_n} is created. The unification instantiates $V$ to $t$,
@@
 \elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
-the application of a fresh variable \elpiIn{C} having in scope all the compiled
-list of terms obtained from \elpiIn{Pf} and \elpiIn{Extra1} as arguments. The
-variable \elpiIn{B} returned has compiled terms is a fresh variable having in
-scope all the free variables used in \rhs.
+the application of a fresh variable \elpiIn{C} having in scope all the free
+variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
+variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
+scope all the free variables occurring in \elpiIn{Pf1}.
+
+\newcommand{\rhsBeta}{\ensuremath{N_S\appsep L}\xspace}
 
 \begin{invariant}
-  The \rhs of a \linkbeta has the shape \elpiIn{app[uva N S|L]}.
+  The \rhs of a \linkbeta has the shape \rhsBeta (equivalent to
+  \elpiIn{app[uva N S|L]}).
 \end{invariant}
@@
 \begin{proof}[Proof sketch]
-  By contradiction, if \elpiIn{L} is a empty list, then the original \Fo{}
-  term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of 
-  disting names. This case is however captured by rule~\ref{rule:complam} (from
-  \cref{sec:compilation}) and no \linkbeta is produced: a contradiction.
+  By contradiction, if \elpiIn{L} is a empty list, then the original \Fo{} term
+  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
+  distinct names (i.e. there list \elpiIn{Extra} is empty). This case is however
+  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
+  \linkbeta is produced: a contradiction.
 \end{proof}
@@
 \begin{corollary}
-  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then the 
+  Let \rhsBeta be the \rhs of a \linkbeta, then the 
   first argument $t$ in \elpiIn{L} either appears in \elpiIn{S} or it is 
@@
 
-Once created, there exist two main situations waking up a suspended \linkbeta.
-The former is strictly connected to the definition of $\beta$-redex and occurs
-when the head of \rhs is materialized by the oracle (see
-\cref{prop:simulation}). In this case \rhs is safely\marginpar{explain why}
-$\beta$-reduced to a new term $t'$ and the result can be unified with \lhs. In
-this scenario the \linkbeta has accomplished its goal and can be removed from
-\linkStore.
-
-The second circumstance making the \linkbeta to progress is the instantiation of
-the variables in the \elpiIn{Extra1} making the corresponding arguments to
-reduce to names. In this case, we want to take the list \elpiIn{Scope1} and
-append to it the largest prefix of \elpiIn{Extra1} in a new variable
-\elpiIn{Scope2} such that \elpiIn{Scope2} remains in \llambda; we call
-\elpiIn{Extra2} the suffix of \elpiIn{Extra1} such that the concatenation of
-\elpiIn{Scope1} and \elpiIn{Extra1} is the same as the concatenation of
-\elpiIn{Scope2} and \elpiIn{Extra2}. Finally, two cases should be considered: 1)
-\elpiIn{Extra2} is the empty list, \lhs and rhs can be unified: we have two
-terms in \llambda; otherwise 2) the \linkbeta in question is replaced with a
-refined version where the \rhs is  \elpiIn{app[uva C Scope2 | Extra2]} and a new
-\linketa is added between the \lhs and the new-added variable \elpiIn{C}.
-\marginpar{dire che si adatta bene nelle approximation}
-
-An example justifying this second  link manipulation is given by the following
-unification problem:
-
-\def\varF{\ensuremath{\textbf{F}}\xspace}
-\def\varA{\ensuremath{\textbf{A}}\xspace}
-\def\varB{\ensuremath{\textbf{B}}\xspace}
-
-\begin{elpicode}
-  f = flam x\ fapp[F, fapp[A, x]].
-\end{elpicode}
-
-% \begin{elpicode};
-%   f = flam x\ flam y\ fapp[F, fapp[A, x], fapp[B, y]].
-% \end{elpicode}
-
-The compilation of these terms produces the new unification problem: $f = X0$
-
-We obtain the mappings $\mapping{F}{\varF}{0}, \mapping{A}{\varA}{1}$ and the links:
+\newcommand{\progBetaLL}{progress-beta-llambda}
+\newcommand{\progBetaRH}{progress-beta-rigid-head}
+
+The activation of a \linkbeta is performed when its \rhs falls under \llambda
+under a given substitution.
+
+\begin{definition}[\progBetaLL]
+  A link \linkbetaM{\Gamma}{X}{N_S\appsep [t|L]} is removed from \linkStore, if
+  it exists a substitution $\gamma$, such that the first term $t$ is a name no
+  occurring in $S$. In this case, let $M$ a fresh variable, then the new links
+  \linkbetaM{\Gamma,x}{X}{M_{S,n}\appsep L} and \linketaM{\Gamma}{N}{\lambda
+  x.M_x} are added to \linkStore.
+\end{definition}
+
+\begin{definition}[\progBetaRH]
+  A link \linkbetaM{\Gamma}{X}{\rhsBeta} is removed from
+  \linkStore if $N_S$ is instantiated to a term $t$ and the
+  $\beta$-reduced term $t'$ obtained from the application of $t$ to
+  $L$ is in \llambda. Moreover, $X$ is unified to $t$.
+\end{definition}
+
+\begin{lemma}
+  \elpiIn{progress} terminates
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  TODO: bla
+\end{proof}
+
+\begin{corollary}
+  Given \linkbetaM{\Gamma}{X}{\rhsBeta}, then \elpiIn{distinct-name} $S$ holds.  
+\end{corollary}
+
+\paragraph{Example of \progBetaLL} 
+A simple example of \linkbeta progression due to \progBetaLL is given below:
+
+% make test ONLY=7004 TEX=tex
+\printAlll
+  {{{X,\lambda x.x},
+    {\lambda x.(Y\appsep (X\appsep x)),a}}}
+  {{{A,\lambda x.x},
+    {B,a}}}
+  {{{Y,D,0},
+    {X,A,0}}}
+  {{{\eta,,A,\lambda x.E_{x}},
+    {\eta,,B,\lambda x.C_{x}},
+    {\beta,x,C_{x},(D\appsep E_{x})}}}
+
+The \linkbeta in \linkStore has a variable $D$ applied to the variable $E_x$.
+The first unification problem, $\hoUnifPb_1$ produces the substitution $\sigma =
+\{A \mapsto \lambda x.x\}$. This instantiation triggers $\linkStore_1$ which
+before its removal, assignes $E$ to $\lambda x.x$. Now, the \linkbeta is
+\linkbetaM{x}{C_x}{(D \appsep x)}. This link is replaced with the couple of
+links: \linkbetaM{x}{C_x}{F_x}, \linketaM{}{E}{\lambda x.D_x}. $\hoUnifPb_2$
+assignes $B$ which activates $\linkStore_2$, and then all the remaining links
+are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
+C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
+Y \mapsto a\}$.
+
+\paragraph{Example of \progBetaRH}
+We can take the example provided in \cref{sec:beta}. The problem is compiled
+into:
 %
-\begin{gather}
-  \linkbetaM{c0}{X3_{c0}}{X2~X1_{c0}}\\
-  \linketaM{}{X0}{\lambda c0.X3_{c0}}
-\end{gather}
-
-\noindent
-where the first link is a \linketa between the variable \texttt{X0}, representing
-the right side of the unification problem (it is a \maybeeta) and
-\texttt{X3}; and a \linkbeta between the variable \texttt{X3} and the subterm
-$\lambda x.X1_x ~ a$ (it is a \maybebeta).
-The substitution tells that \substCell{x}{X1_x}{x}.
-
-We can now represent the \hrun execution from this configuration which will, at
-first, dereference all the links, and then try to solve them. The only link
-being modified is the second one, which is set to \linkbetaM{x}{X3}{X2 x a}. The
-\rhs of the link has now a variable which is partially in the PF, we can
-therefore remove the original \linkbeta and replace it with the following couple
-on links:
-
-\begin{textcode}
-  ~$\vdash$~ X1   =~$\eta$~= x\ `X4 x'
-x ~$\vdash$~ X3 x =~$\beta$~= x\ `X4 x' a
-\end{textcode}
-
-By these links we say that \texttt{X1} is now $\eta$-linked to a fresh variable
-\texttt{X4} with arity one. This new variable is used in the new \linkbeta where
-the name \texttt{x} is in its scope. This allows
+% make test ONLY=7003 TEX=tex
+\printAlll
+  {{{X,\lambda x.Y},
+    {(X\appsep a),a}}}
+  {{{A,\lambda x.B},
+    {C,a}}}
+  {{{Y,B,0},
+    {X,A,0}}}
+  {{{\beta,,C,(A\appsep a)}}}
+
+The first unification, $\hoUnifPb_1$, gives the substitution $\sigma = \{A
+\mapsto \lambda x.B\}$. The \linkbeta, $\linkStore_1$, becomes
+\linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
+$B$. $B$ is in \llambda and is unified with $C$. The resolution of $\hoUnifPb_2$
+gives the final substitution $\sigma = \{A \mapsto \lambda x.B, B \mapsto C, C
+\mapsto a\}$ which is decompiled into $\rho = \{X \mapsto \lambda x.a, Y \mapsto
+a\}$.
 

### 2024-04-02 paper/main.tex
impl of progress-beta consistent with tex

@@
   it exists a substitution $\gamma$, such that the first term $t$ is a name no
-  occurring in $S$. In this case, let $M$ a fresh variable, then the new links
-  \linkbetaM{\Gamma,x}{X}{M_{S,n}\appsep L} and \linketaM{\Gamma}{N}{\lambda
-  x.M_x} are added to \linkStore.
+  occurring in $S$. In this case, let $M$ a fresh variable, then the new link
+  \linkbetaM{\Gamma,x}{X}{M_{S,n}\appsep L} is added to \linkStore and if $L =
+  []$, then $X$ is unified with $M_{S,n}$, otherwise, the
+  \linketaM{\Gamma}{N}{\lambda x.M_x} is added to \linkStore.
+  \label{def:progBetaLL}
 \end{definition}
@@
   $L$ is in \llambda. Moreover, $X$ is unified to $t$.
+  \label{def:progBetaRH}
 \end{definition}
@@
 \begin{proof}[Proof sketch]
-  TODO: bla
+  By \cref{def:progBetaRH}, the \linkbeta is removed from \linkStore, hence 
+  they cannot be applied indefinitely. If the \linkbeta is progressed due to
+  \progBetaLL, then two links are added, the first 
 \end{proof}
@@
 \begin{corollary}
-  Given \linkbetaM{\Gamma}{X}{\rhsBeta}, then \elpiIn{distinct-name} $S$ holds.  
+  Given a \linkbeta, the scope of its \rhs variables in \llambda.
 \end{corollary}
+
+\begin{proof}[Proof sketch]
+  By construction, the \rhs of \linkbeta is of the form \rhsBeta and $S$ is in
+  \llambda. If a \linkbeta is triggered by \progBetaRH, then, by
+  \cref{def:progBetaRH} that link is removed by \linkStore, and the property is
+  satisfied. If the \linketa is activated by \progBetaLL, then, by
+  \cref{def:progBetaLL}, $L = [n|L']$ and the scope `$S,x$', is in \llambda. 
+\end{proof}
 

### 2024-04-02 paper/main.tex
visible TODO

@@
 
-\marginpar{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
+\todo[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
 is resilient to unification problems outside \llambda solved by
@@
 is not set. \elpiIn{assign} sets an unset cell to the given value, while
-\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\marginpar{is new used?}
+\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\todo[inline]{is new used?}
 
@@
 section~\ref{sec:problem-statement}. The reason is that the algorithm folds
-over a term, updating a substitution while it traverses it.\marginpar{explain better}
+over a term, updating a substitution while it traverses it.\todo[inline]{explain better}
 
@@
 % and that the substitution in the object language is the same as the one returned
-% by the ML. \todo{same or $\supseteq$ or $\subseteq$}
+% by the ML. \todo[inline]{same or $\supseteq$ or $\subseteq$}
 
@@
 \label{sec:compilation}
-\marginpar{manca beta normal in entrata}
+\todo[inline]{manca beta normal in entrata}
 
@@
 % being the index of a OL variable and the second being the index of a ML
-% variable. The third integer\todo{integer or nat?} is the number of term in the
+% variable. The third integer\todo[inline]{integer or nat?} is the number of term in the
 % scope of the two variables, or equivalently, in a typed language, their arity.
@@
 
-\marginpar{the compiler must do beta normal!!!!}
+\todo[inline]{the compiler must do beta normal!!!!}
 \newcommand{\reduceto}{\emph{may-contract-to}}
@@
 Note that this condition does not require the term to be in \llambda.
-\marginpar{Is this relevant}
+\todo[inline]{Is this relevant}
 
@@
   \item if $X = \lambda x.t$ we run $X \Ue{} T$.
-  \marginpar{do we have the name case in the code?}
+  \todo[inline]{do we have the name case in the code?}
 \end{enumerate}
@@
 the context $\Gamma$).
-\marginpar{there was a note about T' having a rigid head, TO BE FIXED}
+\todo[inline]{there was a note about T' having a rigid head, TO BE FIXED}
 \end{definition}
@@
 
-\paragraph{Example of \progressetaright}\marginpar{TODO}
+\paragraph{Example of \progressetaright}\todo[inline]{TODO}
 
@@
 
-\marginpar{I don't get this TODO}
+\todo[inline]{I don't get this TODO}
 TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
@@
     {\eta,x,B_{x},\lambda y.C_{y x}}}}
-\marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
+\todo[inline]{TODO: non so se serva spiegare l'esecuzione (in commento)}
 % After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
@@
 
+\todo[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
+Third order, functions that take functions, is undecidable, proved by Dowek.
+IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point.
+This paragraph needs fixing.}
 Until now, we have only dealt we unification of terms in \llambda. However, we
@@
 In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
-$\foUnifPb_2$, can be solved in \llambda. On the other hand, we see that, 
+$\foUnifPb_2$, can be solved in \llambda.
+\todo[inline]{it is even a ground term, there is no unification left to perform actually}
+On the other hand, we see that, 
 \Ue can't solve the compiled problems \hoUnifPb. In
@@
 
-In order to encompass this unification, term compilation should capture the
-terms $t$ in \maybebeta and replace them with fresh varaibles $v$. As per
+In order to
+encompass\todo[inline]{not sure has the right meaning}
+this unification, term compilation should capture the
+terms $t$ in \maybebeta and replace them with fresh varaibles $v$.
+\todo[inline]{we use capitals for unif variables}
+As per
 \maybebeta, the variables $v$ and the terms $t$ are linked through a \linkbeta.

### 2024-04-02 paper/main.tex
notes on 8

@@
 
-TODO: dire che preserviamo l'invariante che tutte le variable sono fully-applied
-
+\todo[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
+}
 \section{Handling of \maybebeta}\label{sec:beta}
@@
   flexible variable and $t_1\dots t_n$ is not in \llambda.
+  \todo[inline]{si dice: flexible term; oppure unification variable}
 \end{invariant}
@@
 \linkbeta are put in \linkStore and activated when \rhs falls in \llambda. 
+\todo[inline]{un po' ridondante con 8.2, non so se serva}
 
@@
   If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
-  counterpart is still in \maybebeta, the current unification problem is not in
-  \llambda and the unification fails.
+  counterpart is still in \maybebeta,
+  the current unification problem is not in \llambda
+  \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
+  and the unification fails.
 \end{corollary}
@@
 variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
-scope all the free variables occurring in \elpiIn{Pf1}.
+scope all the free variables occurring in \elpiIn{Pf1}
+\todo[inline]{anche in Extra1 no?}.
 
 \newcommand{\rhsBeta}{\ensuremath{N_S\appsep L}\xspace}
+\todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
+... where S = [s1 .. sn] e L = [l1 .. lm]}
 
@@
 
-TODO: Dire che maybe eta fa il detect anche su 
+\todo[inline]{Dire che maybe eta fa il detect anche su 
 termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
-esempio?
+esempio?}
 
@@
 
-\newcommand{\progBetaLL}{progress-beta-llambda}
-\newcommand{\progBetaRH}{progress-beta-rigid-head}
+\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}}
+\newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}}
 
@@
   A link \linkbetaM{\Gamma}{X}{N_S\appsep [t|L]} is removed from \linkStore, if
-  it exists a substitution $\gamma$, such that the first term $t$ is a name no
+  it exists a substitution $\gamma$, such that the first term $t$ is a name not
   occurring in $S$. In this case, let $M$ a fresh variable, then the new link
@@
   \label{def:progBetaLL}
+  \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
 \end{definition}
@@
   \progBetaLL, then two links are added, the first
+  \todo[inline]{finire}
 \end{proof}
@@
   Given a \linkbeta, the scope of its \rhs variables in \llambda.
+  \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
 \end{corollary}

### 2024-04-03 paper/main.tex
tex: update beta section

@@
   \elpiIn{progress} terminates.
+  \label{lemma:prog-eta-terminates}
 \end{lemma}
@@
   The \rhs of any \linkbeta has the shape $X~t_1\dots t_n$ such that $X$ is a
-  flexible variable and $t_1\dots t_n$ is not in \llambda.
-  \todo[inline]{si dice: flexible term; oppure unification variable}
+  flexible term and $t_1\dots t_n$ is not in \llambda.
 \end{invariant}
@@
   counterpart is still in \maybebeta,
-  the current unification problem is not in \llambda
-  \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
+  the original unification problem is not in \llambda
+  % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
+  \todo[inline]{Is it clearer?}
   and the unification fails.
@@
 variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
-scope all the free variables occurring in \elpiIn{Pf1}
-\todo[inline]{anche in Extra1 no?}.
-
-\newcommand{\rhsBeta}{\ensuremath{N_S\appsep L}\xspace}
-\todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
-... where S = [s1 .. sn] e L = [l1 .. lm]}
+scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}
+
+\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
+\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
+% \todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
+% ... where S = [s1 .. sn] e L = [l1 .. lm]}
+\todo[inline]{Is it clearer?}
+
+In the following, we pose $X_{s_1\dots s_n}$ to represent a variable $X$ with
+the list of distinct names $s_1\dots s_n$ as scope. Moreover, $X_{s_1\dots
+s_n}\appsep t_1\dots t_m$ is the application of that variable to the list of
+terms $t_1\dots t_m$. Note that \rhsBeta is equivalent to \elpiIn{app[uva N S |
+T]}.
 
 \begin{invariant}
-  The \rhs of a \linkbeta has the shape \rhsBeta (equivalent to
-  \elpiIn{app[uva N S|L]}).
+  The \rhs of a \linkbeta has the shape \rhsBeta.
 \end{invariant}
@@
 \begin{corollary}
-  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then \elpiIn{L} is not
-  empty.
+  Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
 \end{corollary}
@@
 \begin{proof}[Proof sketch]
-  By contradiction, if \elpiIn{L} is a empty list, then the original \Fo{} term
+  By contradiction, if $m = 0$, then the original \Fo{} term
   has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
-  distinct names (i.e. there list \elpiIn{Extra} is empty). This case is however
+  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
   captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
@@
 \begin{corollary}
-  Let \rhsBeta be the \rhs of a \linkbeta, then the 
-  first argument $t$ in \elpiIn{L} either appears in \elpiIn{S} or it is 
+  Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
   not a name.
@@
 \begin{proof}[Proof sketch]
-  By construction, the lists \elpiIn{S} and \elpiIn{L} are built by splitting
+  By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
   the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
-  \elpiIn{S} is the longest prefix of the compiled terms in \elpiIn{Ag} which is
-  in \llambda. Therefore, by definition of \llambda, either the first element of
-  \elpiIn{L} is a name appearing in \elpiIn{S} or it a term with a constructor
-  of \elpiIn{tm} as functor.  
+  $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
+  in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
+  in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
+  \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
 \end{proof}
@@
 \begin{definition}[\progBetaLL]
-  A link \linkbetaM{\Gamma}{X}{N_S\appsep [t|L]} is removed from \linkStore, if
-  it exists a substitution $\gamma$, such that the first term $t$ is a name not
-  occurring in $S$. In this case, let $M$ a fresh variable, then the new link
-  \linkbetaM{\Gamma,x}{X}{M_{S,n}\appsep L} is added to \linkStore and if $L =
-  []$, then $X$ is unified with $M_{S,n}$, otherwise, the
-  \linketaM{\Gamma}{N}{\lambda x.M_x} is added to \linkStore.
+  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
+  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
+  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
+  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
+  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
+  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
+  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
   \label{def:progBetaLL}
-  \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
+  % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
+  \todo[inline]{Is it clearer?}
 \end{definition}
@@
   A link \linkbetaM{\Gamma}{X}{\rhsBeta} is removed from
-  \linkStore if $N_S$ is instantiated to a term $t$ and the
+  \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
   $\beta$-reduced term $t'$ obtained from the application of $t$ to
-  $L$ is in \llambda. Moreover, $X$ is unified to $t$.
+  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified to $t$.
   \label{def:progBetaRH}
@@
 \begin{proof}[Proof sketch]
-  By \cref{def:progBetaRH}, the \linkbeta is removed from \linkStore, hence 
-  they cannot be applied indefinitely. If the \linkbeta is progressed due to
-  \progBetaLL, then two links are added, the first
-  \todo[inline]{finire}
+  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
+  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
+  creates each time a new \linketa and replace the old \linkbeta with a new one.
+  This progression can however triggered at most $m$ times, since each new
+  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
+  the \linkbeta is removed from \linkStore which is now filled by $m$ new
+  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
+  terminates if \linkStore is made by only \linketa and therefore
+  \elpiIn{progress} still terminates.
+  \todo[inline]{is it ok?}
 \end{proof}
@@
 \begin{corollary}
-  Given a \linkbeta, the scope of its \rhs variables in \llambda.
-  \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
+  Given a \linkbeta, the variables occurring in its \rhs is in \llambda.
+  % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
+  \todo[inline]{is it clearer?}
 \end{corollary}
@@
 \begin{proof}[Proof sketch]
-  By construction, the \rhs of \linkbeta is of the form \rhsBeta and $S$ is in
-  \llambda. If a \linkbeta is triggered by \progBetaRH, then, by
-  \cref{def:progBetaRH} that link is removed by \linkStore, and the property is
-  satisfied. If the \linketa is activated by \progBetaLL, then, by
-  \cref{def:progBetaLL}, $L = [n|L']$ and the scope `$S,x$', is in \llambda. 
+  By construction, the \rhs of \linkbeta is of the form \rhsBeta, $s_1\dots s_n$
+  is in \llambda and all the terms $t_1\dots t_n$ are in \llambda, too. If a
+  \linkbeta is triggered by \progBetaRH, then, by \cref{def:progBetaRH}, that
+  link is removed by \linkStore, and the property is satisfied. If the \linketa
+  is activated by \progBetaLL, then, by \cref{def:progBetaLL}, the new \linkbeta
+  as a variable as a scope which is still in \llambda. 
 \end{proof}

### 2024-04-03 paper/main.tex
tex: rework maybebeta

@@
 want the unification relation to be more robust so that it can work with terms
-in \maybebeta. Unification in \maybebeta is in general a non-decidable
-procedure, e.g. $X \appsep a \Uo a$ is a unification problem in \maybebeta,
-since we have $X a$ has the shape $X t_1\dots t_n$ where $X$ is a unification
-variable and $t_1\dots t_n$ is not a list of distinct names (in our example, $a$
-is a constant, hence not a name). We also point out that this unification
+in \maybebeta. In general, unification in \maybebeta admits more then one
+solution and commiting one of them in the substitution does not guarantee
+prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
 problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
-and $\rho_2 = \{X \mapsto \lambda \_.a\}$.
-
-It is the case, however, given a list of unification $n$ problems \foUnifPb, if
-$\foUnifPb_i$ is in \maybebeta, it is possible that the resolution of the
-problems $\bigwedge_{j=0}^{i-1}\foUnifPb_j$ gives a partial substitution
-$\rho$, such that $\rho\foUnifPb_i$ falls again in \llambda.
+and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
+future unifications.
+
+It is the case that, given a list of unification problems, $\foUnifPb_1\dots
+\foUnifPb_n$ with $\foUnifPb_n$ in \maybebeta, the resolution of
+$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
+that $\rho\foUnifPb_n$ falls again in \llambda.
 %

### 2024-04-03 paper/main.tex
tex: beta update

@@
 
-In order to
-encompass\todo[inline]{not sure has the right meaning}
-this unification, term compilation should capture the
-terms $t$ in \maybebeta and replace them with fresh varaibles $v$.
-\todo[inline]{we use capitals for unif variables}
-As per
-\maybebeta, the variables $v$ and the terms $t$ are linked through a \linkbeta.
+To address
+this unification problem, term compilation should capture the
+terms $t$ in \maybebeta and replace them with fresh varaibles $X$.
+The variables $X$ and the terms $t$ are linked through a \linkbeta.
 
@@
 
+\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
+\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
+% \todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
+% ... where S = [s1 .. sn] e L = [l1 .. lm]}
+\todo[inline]{Is it clearer?}
+
 \begin{invariant}[\linkbeta \rhs]
-  The \rhs of any \linkbeta has the shape $X~t_1\dots t_n$ such that $X$ is a
-  flexible term and $t_1\dots t_n$ is not in \llambda.
+  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
+  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
+  terms. This is equivalent to \elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
+  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
 \end{invariant}
 
-\linkbeta are put in \linkStore and activated when \rhs falls in \llambda. 
-\todo[inline]{un po' ridondante con 8.2, non so se serva}
+% \linkbeta are put in \linkStore and activated when \rhs falls in \llambda. 
+% \todo[inline]{un po' ridondante con 8.2, non so se serva}
 
@@
   $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
-  is replaced with a variable $V$, and the \linkbeta \linkbetaM{\Gamma}{V}{X
-  \appsep t_1\dots t_n} is created. The unification instantiates $V$ to $t$,
+  is replaced with a variable $Y$, and the \linkbeta \linkbetaM{\Gamma}{Y}{X
+  \appsep t_1\dots t_n} is created. The unification instantiates $Y$ to $t$,
   making the \lhs of the link a rigid term, while \rhs is still in \maybebeta.
-  The original problem is in fact outside \llambda and unification fails.
+  The original problem is in fact outside \llambda.
 \end{proof}
@@
 Detection of \maybebeta is quite simple to implement in the compiler, since it
-is sufficient to capture applications with flexible head and argument that
+is sufficient to detect applications with flexible head and argument that
 are not in \llambda.
@@
 
-\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
-\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
-% \todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
-% ... where S = [s1 .. sn] e L = [l1 .. lm]}
-\todo[inline]{Is it clearer?}
-
-In the following, we pose $X_{s_1\dots s_n}$ to represent a variable $X$ with
-the list of distinct names $s_1\dots s_n$ as scope. Moreover, $X_{s_1\dots
-s_n}\appsep t_1\dots t_m$ is the application of that variable to the list of
-terms $t_1\dots t_m$. Note that \rhsBeta is equivalent to \elpiIn{app[uva N S |
-T]}.
+% In the following, we pose $X_{s_1\dots s_n}$ to represent a variable $X$ with
+% the list of distinct names $s_1\dots s_n$ as scope. Moreover, $X_{s_1\dots
+% s_n}\appsep t_1\dots t_m$ is the application of that variable to the list of
+% terms $t_1\dots t_m$. Note that \rhsBeta is equivalent to \elpiIn{app[uva N S |
+% T]}.
 

### 2024-04-03 paper/main.tex
notes

@@
   % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
-  \todo[inline]{Is it clearer?}
+  %\todo[inline]{Is it clearer?}
   and the unification fails.
+  \todo[inline]{it is clear.
+Quello che manca è percè serve farlo, e che cosa si guadagna.
+Qui dimostri che fallire "ha senso" se siamo solo interessati a llambda, ma allo stesso tempo sti beta link sono lì per gestire roba fuori.}
 \end{corollary}
@@
 \begin{proof}[Proof sketch]
-  By contradiction, if $m = 0$, then the original \Fo{} term
+  By contradiction,\todo{we assume to have a link} if $m = 0$, then the original \Fo{} term
   has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
@@
   captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
-  \linkbeta is produced: a contradiction.
+  \linkbeta is produced: a contradiction.\todo{since we started by assuming we had a link}
 \end{proof}
@@
   % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
-  \todo[inline]{Is it clearer?}
+  %\todo[inline]{Is it clearer?}
+  \todo[inline]{l1 or t1?
+Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via.
+In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
 \end{definition}
@@
   \elpiIn{progress} still terminates.
-  \todo[inline]{is it ok?}
+  %\todo[inline]{is it ok?}
+  \todo[inline]{funziona.
+per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione).
+Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
 \end{proof}
@@
 \begin{corollary}
-  Given a \linkbeta, the variables occurring in its \rhs is in \llambda.
+  Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
   % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
@@
 \paragraph{Example of \progBetaLL}
-A simple example of \linkbeta progression due to \progBetaLL is given below:
+Consider the \linkbeta below:
 
@@
 
-The \linkbeta in \linkStore has a variable $D$ applied to the variable $E_x$.
-The first unification problem, $\hoUnifPb_1$ produces the substitution $\sigma =
-\{A \mapsto \lambda x.x\}$. This instantiation triggers $\linkStore_1$ which
-before its removal, assignes $E$ to $\lambda x.x$. Now, the \linkbeta is
-\linkbetaM{x}{C_x}{(D \appsep x)}. This link is replaced with the couple of
-links: \linkbetaM{x}{C_x}{F_x}, \linketaM{}{E}{\lambda x.D_x}. $\hoUnifPb_2$
-assignes $B$ which activates $\linkStore_2$, and then all the remaining links
+\noindent
+Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
+The first unification problem results in $\sigma =
+\{A \mapsto \lambda x.x\}$. In turn this instantiation
+triggers progress \todo{which rule?} in first \linketa that is 
+removed and $E$ is assigned to $\lambda x.x$.
+Under this substitution the \linkbeta becomes
+\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
+it is replaced with the following two links:
+\linkbetaM{x}{C_x}{F_x} and
+\todo{ma allora il link ora sparisce perhè è in llambda?}
+\linketaM{}{E}{\lambda x.D_x}. The second unification
+problem assignes $a$ to $B$, that in turn activates
+the second \linketa ($a$ is assigned to $C$), and then all the remaining links
 are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
 C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
-Y \mapsto a\}$.
+Y \mapsto a\}$.\todo{assegnament per D incompleto}
+\todo{alla fine non si usa anche la regola beta-rigid-head?}
 
@@
 
-The first unification, $\hoUnifPb_1$, gives the substitution $\sigma = \{A
-\mapsto \lambda x.B\}$. The \linkbeta, $\linkStore_1$, becomes
+The first unification problems is solved by the substitution $\sigma = \{A
+\mapsto \lambda x.B\}$. The \linkbeta becomes
 \linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
-$B$. $B$ is in \llambda and is unified with $C$. The resolution of $\hoUnifPb_2$
+$B$. $B$ is in \llambda and is unified with $C$. The resolution of the second
+unification problem
 gives the final substitution $\sigma = \{A \mapsto \lambda x.B, B \mapsto C, C

### 2024-04-03 paper/main.tex
todo in diff colors

@@
 the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
-bla bla.. TODO citare Teyjus.
+bla bla.. \noteE[inline]{citare Teyjus}.
 The code discussed in the paper can be accessed at the URL:
@@
 \Ue the one provided by the meta language.
-TODO extend \Eo and \Ee with reflexivity on uvars.
+\noteE[inline]{extend \Eo and \Ee with reflexivity on uvars.}
 
@@
 are linked to the unification variable that stands in their place in the
-compiled term. These links are checked for or progress XXX improve....
+compiled term. These links are checked for or progress \noteE[inline]{XXX improve....}
 
@@
 
-\todo[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
+\noteE[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
 is resilient to unification problems outside \llambda solved by
@@
 is not set. \elpiIn{assign} sets an unset cell to the given value, while
-\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\todo[inline]{is new used?}
+\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\noteE[inline]{is new used?}
 
@@
 
-TODO: add ref to \cref{sec:invariant1}
+\noteD{add ref to \cref{sec:invariant1}}
 
@@
 section~\ref{sec:problem-statement}. The reason is that the algorithm folds
-over a term, updating a substitution while it traverses it.\todo[inline]{explain better}
+over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
 
@@
 % and that the substitution in the object language is the same as the one returned
-% by the ML. \todo[inline]{same or $\supseteq$ or $\subseteq$}
+% by the ML. \noteE[inline]{same or $\supseteq$ or $\subseteq$}
 
@@
 \label{sec:compilation}
-\todo[inline]{manca beta normal in entrata}
+\noteE[inline]{manca beta normal in entrata}
 
@@
 % being the index of a OL variable and the second being the index of a ML
-% variable. The third integer\todo[inline]{integer or nat?} is the number of term in the
+% variable. The third integer\noteE[inline]{integer or nat?} is the number of term in the
 % scope of the two variables, or equivalently, in a typed language, their arity.
@@
 
-\todo[inline]{the compiler must do beta normal!!!!}
+\noteE[inline]{the compiler must do beta normal!!!!}
 \newcommand{\reduceto}{\emph{may-contract-to}}
@@
 Note that this condition does not require the term to be in \llambda.
-\todo[inline]{Is this relevant}
+\noteE[inline]{Is this relevant}
 
@@
     the context $\Gamma$)
+    \noteE[inline]{where y comes from? $X = y$: $y$ is in the ctx of X}
     \item if $X = \lambda x.t$ we run $X \Ue{} T$.
-  \todo[inline]{do we have the name case in the code?}
 \end{enumerate}
@@
 the context $\Gamma$).
-\todo[inline]{there was a note about T' having a rigid head, TO BE FIXED}
+\noteE[inline]{there was a note about T' having a rigid head, TO BE FIXED}
 \end{definition}
@@
 
-\paragraph{Example of \progressetaright}\todo[inline]{TODO}
-
-A second example, showing the activation of a link when the \rhs is no more a
-\maybeeta, is given in \cref{sec:invariant1}, since we need to work with
-variables used with different arities. This example represent the run of the
-unification problems proposed at \cref{sec:basic-comp-limitations}
-
 \paragraph{Example of \progressetadedup}
@@
 
-\todo[inline]{I don't get this TODO}
-TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
-not reduce to $T x$ where x is not free in $T$.
+We delay in next section a \linketa progression due to \progressetaright
 
@@
     {\eta,x,B_{x},\lambda y.C_{y x}}}}
-\todo[inline]{TODO: non so se serva spiegare l'esecuzione (in commento)}
+
+\noteD[inline]{ci vuole l'esempio per il \progressetaright}
 % After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
@@
 
-\todo[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
+\noteE[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
 }
@@
 
-\todo[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
+\noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
 Third order, functions that take functions, is undecidable, proved by Dowek.
@@
 $\foUnifPb_2$, can be solved in \llambda.
-\todo[inline]{it is even a ground term, there is no unification left to perform actually}
+\noteE[inline]{it is even a ground term, there is no unification left to perform actually}
 On the other hand, we see that, 
@@
 \newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
-% \todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
+% \noteE[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
 % ... where S = [s1 .. sn] e L = [l1 .. lm]}
-\todo[inline]{Is it clearer?}
+\noteE[inline]{Is it clearer?}
 
@@
 % \linkbeta are put in \linkStore and activated when \rhs falls in \llambda. 
-% \todo[inline]{un po' ridondante con 8.2, non so se serva}
+% \noteE[inline]{un po' ridondante con 8.2, non so se serva}
 
@@
   the original unification problem is not in \llambda
-  % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
-  %\todo[inline]{Is it clearer?}
+  % \noteE[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
+  %\noteE[inline]{Is it clearer?}
   and the unification fails.
-  \todo[inline]{it is clear.
+  \noteE[inline]{it is clear.
 Quello che manca è percè serve farlo, e che cosa si guadagna.
@@
 \begin{proof}[Proof sketch]
-  By contradiction,\todo{we assume to have a link} if $m = 0$, then the original \Fo{} term
+  By contradiction,\noteE{we assume to have a link} if $m = 0$, then the original \Fo{} term
   has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
@@
   captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
-  \linkbeta is produced: a contradiction.\todo{since we started by assuming we had a link}
+  \linkbeta is produced: a contradiction.\noteE{since we started by assuming we had a link}
 \end{proof}
@@
 
-\todo[inline]{Dire che maybe eta fa il detect anche su 
+\noteE[inline]{Dire che maybe eta fa il detect anche su 
 termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
@@
   \label{def:progBetaLL}
-  % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
-  %\todo[inline]{Is it clearer?}
-  \todo[inline]{l1 or t1?
+  % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
+  %\noteE[inline]{Is it clearer?}
+  \noteE[inline]{l1 or t1?
 Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via.
@@
   \elpiIn{progress} still terminates.
-  %\todo[inline]{is it ok?}
-  \todo[inline]{funziona.
+  %\noteE[inline]{is it ok?}
+  \noteE[inline]{funziona.
 per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione).
@@
   Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
-  % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
-  \todo[inline]{is it clearer?}
+  % \noteE[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
+  \noteE[inline]{is it clearer?}
 \end{corollary}
@@
 \{A \mapsto \lambda x.x\}$. In turn this instantiation
-triggers progress \todo{which rule?} in first \linketa that is 
+triggers progress \noteE{which rule?} in first \linketa that is 
 removed and $E$ is assigned to $\lambda x.x$.
@@
 \linkbetaM{x}{C_x}{F_x} and
-\todo{ma allora il link ora sparisce perhè è in llambda?}
+\noteE{ma allora il link ora sparisce perhè è in llambda?}
 \linketaM{}{E}{\lambda x.D_x}. The second unification
@@
 C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
-Y \mapsto a\}$.\todo{assegnament per D incompleto}
-\todo{alla fine non si usa anche la regola beta-rigid-head?}
+Y \mapsto a\}$.\noteE{assegnament per D incompleto}
+\noteE{alla fine non si usa anche la regola beta-rigid-head?}
 
@@
 
-\subsection{Tricky examples}
-
+\subsection{Heuristics}
+
+Possiamo perdere OC e rimanere sospesi o commit di soluzione approx
 \begin{elpicode}
@@
 
-\section{First order approximation}
-
-\otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
-\otext{We can re-use part of the algo for $\beta$ given before}
-
-\section{Unif encoding in real life}
-\otext{Il ML presentato qui è esattamente elpi}
-\otext{Il OL presentato qui è esattamente coq}
-\otext{Come implementatiamo tutto ciò nel solver}
-
-\section{Results: stdpp and tlc}
-\otext{How may rule are we solving?}
-\otext{Can we do some perf test}
+% \section{First order approximation}
+
+% \otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
+% \otext{We can re-use part of the algo for $\beta$ given before}
+
+% \section{Unif encoding in real life}
+% \otext{Il ML presentato qui è esattamente elpi}
+% \otext{Il OL presentato qui è esattamente coq}
+% \otext{Come implementatiamo tutto ciò nel solver}
+
+% \section{Results: stdpp and tlc}
+% \otext{How may rule are we solving?}
+% \otext{Can we do some perf test}
+
+\section{Practical example}
 

### 2024-04-03 paper/main.tex
some rework

@@
 
-TODO: W preservation \cref{prop:nf}
-
 \subsection{Progress}\label{sec:etaprogress}
@@
 \newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
-\begin{definition}[\progressetaright]\label{def:progressetaright}
-A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
-$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
-$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
-the context $\Gamma$).
-\noteE[inline]{there was a note about T' having a rigid head, TO BE FIXED}
+\begin{definition}[\progressetaright]\label{def:progressetaright} A link
+\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
+$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
+$T'$, $T'$ is a term not starting with the \elpiIn{lam} constructor. In the
+first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
+$T'$ (under the context $\Gamma$).
 \end{definition}
@@
 There is a third case in which a link is removed from \linkStore, namely
-when the \lhs is assigned to another variable that is the \lhs of another
+when the \lhs is assigned to a variable that is the \lhs of another
 \linketa.

### 2024-04-03 paper/main.tex
add example of pregress-eta-right from example of sec7

@@
 
-\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}}
+\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}\xspace}
 \begin{definition}[\progressetaleft]\label{def:progressetaleft}
@@
 
-\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
+\newcommand{\progressetaright}{\emph{progress-$\eta$-right}\xspace}
 \begin{definition}[\progressetaright]\label{def:progressetaright} A link
@@
 
-We delay in next section a \linketa progression due to \progressetaright
+We delay at the end of  next section an example of \linketa progression due to
+\progressetaright
 
@@
 
-\noteD[inline]{ci vuole l'esempio per il \progressetaright}
+In this example, $\hoUnifPb_1$ assigns $A$ which triggers $\linkStore_3$ and
+then $\linkStore_4$ by \cref{def:progressetaleft}. $C_{yx}$ is therefore
+assigned to $x$ (the second variable of its scope). We can finally see the
+\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ ($C_{xy}$
+gives $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
+$E_x$. Moreover, $\linkStore_2$ is also triggered due to
+\cref{def:progressetaright}: $\lambda x.(f \appsep (lambda y.y) \appsep x)$ is
+$\eta$-reducible to $f \appsep (lambda y.y)$ which is a term not starting with
+the \elpiIn{lam} constructor.
+
 % After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.

### 2024-04-03 paper/main.tex
wip: small update

@@
   and the unification fails.
-  \noteE[inline]{it is clear.
-Quello che manca è percè serve farlo, e che cosa si guadagna.
-Qui dimostri che fallire "ha senso" se siamo solo interessati a llambda, ma allo stesso tempo sti beta link sono lì per gestire roba fuori.}
+  \label{cor:keep-fidelity}
 \end{corollary}
@@
 \end{proof}
+
+\begin{lemma}
+  Given a \hoUnifPb and a substitution $\sigma$ then the resolution of
+  $\sigma\hoUnifPb$ guarantees \cref{prop:fidelity} 
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  If $\sigma\hoUnifPb$ is in \llambda, then by \cref{DA FARE}, the problem 
+  unifies iff its corresponding \Fo{} unifies too. If $\sigma\hoUnifPb$ is in 
+  \maybebeta, then, by \cref{cor:keep-fidelity}, the unification fails, as per
+  the corresponding unification in \Fo{}.
+\end{proof}
+
+\noteD[inline]{manca una ref che dice che se \hoUnifPb is not in $\llambda$ but
+$\sigma\hoUnifPb$ is in \llambda, then \hoUnifPb succeeds iff its corresponding
+\Fo{} pb succeeds}.
 

### 2024-04-03 paper/main.tex
typo

@@
 $E_x$. Moreover, $\linkStore_2$ is also triggered due to
-\cref{def:progressetaright}: $\lambda x.(f \appsep (lambda y.y) \appsep x)$ is
-$\eta$-reducible to $f \appsep (lambda y.y)$ which is a term not starting with
+\cref{def:progressetaright}: $\lambda x.(f \appsep (\lambda y.y) \appsep x)$ is
+$\eta$-reducible to $f \appsep (\lambda y.y)$ which is a term not starting with
 the \elpiIn{lam} constructor.

### 2024-04-03 paper/main.tex
corollary->lemma

@@
 % section is given in a future section
-\begin{corollary}
+\begin{lemma}
   If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
@@
   and the unification fails.
-  \label{cor:keep-fidelity}
-\end{corollary}
+  \label{lemma:keep-fidelity}
+\end{lemma}
 
@@
   unifies iff its corresponding \Fo{} unifies too. If $\sigma\hoUnifPb$ is in 
-  \maybebeta, then, by \cref{cor:keep-fidelity}, the unification fails, as per
+  \maybebeta, then, by \cref{lemma:keep-fidelity}, the unification fails, as per
   the corresponding unification in \Fo{}.

### 2024-04-03 paper/main.tex
update proof cor 8.3

@@
 \begin{proof}[Proof sketch]
-  By contradiction,\noteE{we assume to have a link} if $m = 0$, then the original \Fo{} term
-  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
-  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
-  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
-  \linkbeta is produced: a contradiction.\noteE{since we started by assuming we had a link}
+  Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
+  \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
+  list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
+  however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
+  no \linkbeta is produced which contradicts our initial assumption.
 \end{proof}

### 2024-04-03 paper/main.tex
Proof of lemma 2.8 for eta links

@@
   the rule above respects \cref{inv:link-eta-right}.
+  \label{cor:rhs-eta-onelamb}
 \end{corollary}
@@
 
-TODO: prove \cref{prop:nf}: we never commit a \maybeeta{} term in
-$\sigma$ since we run \Ue{} only when we know that the 
-terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
-no more a \maybeeta, the link is removed from \linkStore.
+\begin{lemma}
+  Given a \linketa $l$, the unification done by \progressetaleft is between
+  terms in $\wellb$  
+  \label{lemma:unif-eta-1}
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  Let $\sigma$ be the substitution, such that $\wellb(\sigma)$. $\lhs \in
+  \sigma$, therefore $\wellb(\lhs)$. By \cref{def:progressetaleft}, if 1) \lhs is
+  a name, a constant of an application, then, \lhs is unified with the
+  $\eta$-reduced term $t$ obtain from \rhs. By \cref{cor:rhs-eta-onelamb}, \rhs
+  has one lambda, therefore $\wellb(t)$. Otherwise, 2) \lhs has \elpiIn{lam} as
+  functor, \rhs should not be an $\eta$-expansion ans, so, $\wellb(\rhs)$. In
+  both cases, unification is performed between terms in \wellb.
+\end{proof}
+
+\begin{lemma}
+  Given a \linketa $l$, the unification done by \progressetaright is between
+  terms in \wellb.
+  \label{lemma:unif-eta-2}
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  \lhs is variable, and, by \cref{def:progressetaright}, \rhs is either no more
+  a \maybeeta, i.e. \rhs is not a $\eta$-expansion and, so, $\wellb(\rhs)$.
+  Otherwise, \rhs can reduce to a term which cannot be a $\eta$-expansion, and,
+  so, $\wellb(\rhs)$. In both cases, unification is done between terms in
+  \wellb.
+\end{proof}
+
+\begin{lemma}
+  Given a \linketa $l$, the unification done by \progressetadedup is between
+  terms in \wellb.
+  \label{lemma:unif-eta-3}
+\end{lemma}
+
+\begin{proof}
+  Trivial, since the unification is done between unification variables, which
+  are by definition in \wellb.
+\end{proof}
+
+\begin{lemma}
+  \Cref{prop:nf} holds, i.e., given a substitution $\sigma$ and a \linketa $l$,
+  after the activation of $l$, $\wellb(\sigma)$ holds.
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
+  unification performed by the activation of a \linketa is performed between
+  terms in \wellb, therefore, the substitution remains \wellb.
+\end{proof}
 

### 2024-04-03 paper/main.tex
add some notes

@@
 
+\noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
+forse si può compattare in una prova più piccola?}
 \begin{lemma}
@@
 
+\noteD[inline]{Should we proove simulation fidelity for \linketa insertion?}
+
 \paragraph{Example of \progressetaleft}
@@
 
-\noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
+% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
 Third order, functions that take functions, is undecidable, proved by Dowek.
@@
 This paragraph needs fixing.}
+\noteD[inline]{I've rewritten it, it is clearer?}
 Until now, we have only dealt we unification of terms in \llambda. However, we
@@
 \noteE[inline]{it is even a ground term, there is no unification left to perform actually}
+\noteD[inline]{i don't understand the note}
 On the other hand, we see that, 
@@
 % ... where S = [s1 .. sn] e L = [l1 .. lm]}
-\noteE[inline]{Is it clearer?}
+\noteD[inline]{Is it clearer?}
 

### 2024-04-03 paper/main.tex
moved lemma

@@
 
-\begin{lemma}
-  Given a \hoUnifPb and a substitution $\sigma$ then the resolution of
-  $\sigma\hoUnifPb$ guarantees \cref{prop:fidelity} 
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-  If $\sigma\hoUnifPb$ is in \llambda, then by \cref{DA FARE}, the problem 
-  unifies iff its corresponding \Fo{} unifies too. If $\sigma\hoUnifPb$ is in 
-  \maybebeta, then, by \cref{lemma:keep-fidelity}, the unification fails, as per
-  the corresponding unification in \Fo{}.
-\end{proof}
-
-\noteD[inline]{manca una ref che dice che se \hoUnifPb is not in $\llambda$ but
-$\sigma\hoUnifPb$ is in \llambda, then \hoUnifPb succeeds iff its corresponding
-\Fo{} pb succeeds}.
-
 \subsection{Compilation}
@@
 \begin{proof}[Proof sketch]
-  By construction, the \rhs of \linkbeta is of the form \rhsBeta, $s_1\dots s_n$
+  By construction, the \rhs of \linkbeta has the shape \rhsBeta, $s_1\dots s_n$
   is in \llambda and all the terms $t_1\dots t_n$ are in \llambda, too. If a
@@
   as a variable as a scope which is still in \llambda. 
+\end{proof}
+
+\begin{lemma}
+  Given a \hoUnifPb and a substitution $\sigma$ then the resolution of
+  $\sigma\hoUnifPb$ guarantees \cref{prop:fidelity} 
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  If $\sigma\hoUnifPb$ is in \llambda, then by
+  \cref{def:progBetaLL,def:progBetaRH}, then \linkbeta desappear and the
+  unification done between terms in \llambda. This problem unifies iff
+  its corresponding \Fo{} problem unifies too. If $\sigma\hoUnifPb$ is in \maybebeta,
+  then, by \cref{lemma:keep-fidelity}, the unification fails, as per the
+  corresponding unification in \Fo{}.
 \end{proof}

### 2024-04-04 paper/main.tex
add paragraph for beta approximation

@@
 
-\subsection{Heuristics}
-
-Possiamo perdere OC e rimanere sospesi o commit di soluzione approx
-\begin{elpicode}
-  triple ok (@lam x\ @app[@f, @app[@X, x]]) @Y,
-  triple ok @X (@lam x\ x),
-  triple ok @Y @f
-\end{elpicode}
-
-\begin{elpicode}
-% @okl 22 [
-%   triple ok (@lam x\ @lam y\ @app[@Y, y, x]) @X,
-%   triple ok (@lam x\ @f) @X,
-% ].
-\end{elpicode}
-
-% \section{First order approximation}
-
-% \otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
-% \otext{We can re-use part of the algo for $\beta$ given before}
-
-% \section{Unif encoding in real life}
-% \otext{Il ML presentato qui è esattamente elpi}
-% \otext{Il OL presentato qui è esattamente coq}
-% \otext{Come implementatiamo tutto ciò nel solver}
-
-% \section{Results: stdpp and tlc}
-% \otext{How may rule are we solving?}
-% \otext{Can we do some perf test}
+\subsection{Relaxing \cref{lemma:keep-fidelity}}
+
+\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\maybebeta}}
+
+% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
+Working with terms in \llambda is sometime too restrictive. There exists systems
+such as $\lambda$Prolog~\cite{lamProlog}, Abella~\cite{gacek2008abella}, which 
+delay the resolution of \maybebeta unification problems if the substitution
+is not able to put them in \llambda.
+
+\printAlll
+  {{{(X\appsep a),a},
+    {X,\lambda x.Y}}}
+  {{}}
+  {{}}
+  {{}}
+
+In the example above, $\foUnifPb_1$ is in \maybebeta and the object language
+cannot solve it, and, by \cref{prop:fidelity}, the meta language neither.
+However, we can be more permissive, and relax \cref{lemma:keep-fidelity}. This
+modification is quite simple to manage: we are introducing a new \maybebeta
+progress rule, say \progBetaNoLLWait, by which, if \lhs is rigid and \rhs is
+flexible, the considered \linkbeta is kept in the store and no progression is
+done\footnote{This new rule trivially guarantees the termination of progress}.
+\progBetaNoLLWait makes \elpiIn{occur-check-links} partial, since the check is
+possible only on links with a variable on the \lhs. This means that we can have
+two links \linkbetaM{}{X}{Y\appsep a} and \linkbetaM{}{f\appsep X}{Y\appsep a}
+where the occur check does not thow an error. Note however, that the
+decompilation of the two links will force the unification of $X$ to $Y\appsep a$
+and then the unification of $f\appsep (Y \appsep a)$ to $Y\appsep a$, which
+fails by the occur check of \Ue.
+
+A second strategy to deal with problem that are in \maybebeta is to make some
+approximation. This is the case for example of the unification algorithm of Coq
+used in its type class solver~\cite{sozeau08}. The approximation consists in
+forcing a choice (among the others) when the unification problem is in
+\maybebeta. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
+argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
+that this is of course an approximation, since $\sigma = \{X = \lambda x.Y, Y =
+\_\}$ is another valid substitution for the original problem. This approximation
+can be easily introduced in our unification procedure, by adding new custom 
+\linkbeta progress rules.
 

### 2024-04-04 paper/main.tex
tex : solved a todo

@@
   {{{X,\lambda x.x},
-    {\lambda x.(Y\appsep (X\appsep x)),a}}}
+    {\lambda x.(Y\appsep (X\appsep x)),f}}}
   {{{A,\lambda x.x},
-    {B,a}}}
+    {B,f}}}
   {{{Y,D,0},
@@
 \{A \mapsto \lambda x.x\}$. In turn this instantiation
-triggers progress \noteE{which rule?} in first \linketa that is 
-removed and $E$ is assigned to $\lambda x.x$.
+triggers $\linkStore_1$ by \progressetaleft 
+and $E$ is assigned to $\lambda x.x$.
 Under this substitution the \linkbeta becomes
@@
 \linketaM{}{E}{\lambda x.D_x}. The second unification
-problem assignes $a$ to $B$, that in turn activates
-the second \linketa ($a$ is assigned to $C$), and then all the remaining links
-are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
-C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
-Y \mapsto a\}$.\noteE{assegnament per D incompleto}
+problem assignes $f$ to $B$, that in turn activates
+the second \linketa ($f$ is assigned to $C$), and then all the remaining links
+are solved. The final \Ho substitution is $\sigma = \{A _{} \mapsto \lambda x.x, 
+B _{} \mapsto f, C _{x} \mapsto (f\appsep x), 
+D _{} \mapsto f, E _{x} \mapsto x, F _{x} \mapsto C_{x} \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
+Y \mapsto f\}$.
 \noteE{alla fine non si usa anche la regola beta-rigid-head?}

### 2024-04-04 paper/main.tex
tex solved a todo
tex: rewrite lemma progress-beta-L𝜆

@@
 \begin{definition}[\progBetaLL]
-  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
-  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
-  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
+  Given a substitution $\sigma$ and a \linkbeta \linkbetaM{\Gamma}{T}{\rhsBeta}
+  such that $\sigma t_1$ is a name, say $t$, and $t\notin s_1\dots s_n$. If $m =
+  0$, then the \linkbeta is removed and \lhs is unified with \rhsBetaHead.
+  If $m > 0$, then the \linkbeta is replaced by a refined version
+  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots t_m}$ with reduced
+  list of arguments and $Y$ being a fresh variable. Moreover, the new \linkbeta,
   \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to 
-  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
-  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
-  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
-  \label{def:progBetaLL}
-  % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
-  %\noteE[inline]{Is it clearer?}
-  \noteE[inline]{l1 or t1?
+  \linkStore.
+
+  % A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
+  % the substitution $\sigma$, $\sigma t_1$ is a name, say $t$, such that, $t
+  % \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
+  % \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
+  % \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
+  % 2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
+  % s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
+  % \label{def:progBetaLL}
+  % % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
+  % %\noteE[inline]{Is it clearer?}
+  % \noteE[inline]{l1 or t1?
 Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via.
 In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
+  \noteD{L'ho riformulato}
 \end{definition}
@@
 \linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
-it is replaced with the following two links:
-\linkbetaM{x}{C_x}{F_x} and
-\noteE{ma allora il link ora sparisce perhè è in llambda?}
-\linketaM{}{E}{\lambda x.D_x}. The second unification
+it is replaced with the link:
+\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
 problem assignes $f$ to $B$, that in turn activates
@@
 \noteE{alla fine non si usa anche la regola beta-rigid-head?}
+\noteD{nope. Si risolve tutto by \progBetaLL}
 

### 2024-04-04 paper/main.tex
rework proof termination progress with betas

@@
   % s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
-  % \label{def:progBetaLL}
+  \label{def:progBetaLL}
   % % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
@@
 \begin{proof}[Proof sketch]
-  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
-  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
-  creates each time a new \linketa and replace the old \linkbeta with a new one.
-  This progression can however triggered at most $m$ times, since each new
-  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
-  the \linkbeta is removed from \linkStore which is now filled by $m$ new
-  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
-  terminates if \linkStore is made by only \linketa and therefore
-  \elpiIn{progress} still terminates.
+  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
+  \progBetaRH, then it desappears from \linkStore and \elpiIn{progress}
+  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
+  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
+  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
+  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
+  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
+  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
+  terminates.
+  \noteD{L'ho riscritto}
+
+  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
+  % \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
+  % creates each time a new \linketa and replace the old \linkbeta with a new one.
+  % This progression can however triggered at most $m$ times, since each new
+  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
+  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
+  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
+  % terminates if \linkStore is made by only \linketa and therefore
+  % \elpiIn{progress} still terminates.
   %\noteE[inline]{is it ok?}
@@
   % \noteE[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
-  \noteE[inline]{is it clearer?}
+  \noteD[inline]{is it clearer?}
 \end{corollary}
@@
 is not able to put them in \llambda.
-
+%
+% make test ONLY=7005 TEX=tex
 \printAlll

### 2024-04-05 paper/main.tex
typos

@@
   Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
-  Constuctions~\cite{felty93lics}.
+  Constructions~\cite{felty93lics}.
 
@@
 Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
-Constuctions~\cite{felty93lics}.
+Constructions~\cite{felty93lics}.
 
@@
 
-XXX permuting hrun does not change the final result if check dooes not fail eagerly
+XXX permuting hrun does not change the final result if check does not fail eagerly
 
@@
 the link is considered for progress and possibly eliminated.
-This is discussed in \cref{sec:eta} and \cref{sc:beta}.
+This is discussed in \cref{sec:eta} and \cref{sec:beta}.
 
@@
 \section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
-\label{sec:compilation}
+% \label{sec:compilation}
 
@@
 Decompiling an assignment requires to turn abstractions into
-lambdas. For aestetic purposes we also eta-contract the result
+lambdas. For aesthetic purposes we also eta-contract the result
 (not needed since \Fo{} equality can do that)
@@
 in \maybebeta. In general, unification in \maybebeta admits more then one
-solution and commiting one of them in the substitution does not guarantee
+solution and committing one of them in the substitution does not guarantee
 prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
@@
 \mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
-non-unifable problem $(\lambda x. B) \appsep a \Ue a$.
+non-unifiable problem $(\lambda x. B) \appsep a \Ue a$.
 
@@
 this unification problem, term compilation should capture the
-terms $t$ in \maybebeta and replace them with fresh varaibles $X$.
+terms $t$ in \maybebeta and replace them with fresh variables $X$.
 The variables $X$ and the terms $t$ are linked through a \linkbeta.
@@
   Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
-  \progBetaRH, then it desappears from \linkStore and \elpiIn{progress}
+  \progBetaRH, then it disappears from \linkStore and \elpiIn{progress}
   terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
@@
   % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
-  % \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
+  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
   % creates each time a new \linketa and replace the old \linkbeta with a new one.
@@
   If $\sigma\hoUnifPb$ is in \llambda, then by
-  \cref{def:progBetaLL,def:progBetaRH}, then \linkbeta desappear and the
+  \cref{def:progBetaLL,def:progBetaRH}, then \linkbeta disappear and the
   unification done between terms in \llambda. This problem unifies iff
@@
 \linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
-problem assignes $f$ to $B$, that in turn activates
+problem assigns $f$ to $B$, that in turn activates
 the second \linketa ($f$ is assigned to $C$), and then all the remaining links
@@
 two links \linkbetaM{}{X}{Y\appsep a} and \linkbetaM{}{f\appsep X}{Y\appsep a}
-where the occur check does not thow an error. Note however, that the
+where the occur check does not throw an error. Note however, that the
 decompilation of the two links will force the unification of $X$ to $Y\appsep a$

### 2024-04-05 paper/main.tex
small fix

@@
 right counterpart is the
-term in \maybeeta, and that term has the following property:
+term in \maybeeta. That term has the following property:
 
@@
 The first two examples are easy, and show how a unification variable can expose
-or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
+or erase a variable in their scope and turn the resulting term in an $\eta$-expansion or not.\\
 The third example shows that when a variable occurs outside the scope of a unification
@@
 $\eta$-expansion.\\
-
 We can now define more formally how \maybeeta terms are detected together with
@@
 
-\noteE[inline]{the compiler must do beta normal!!!!}
 \newcommand{\reduceto}{\emph{may-contract-to}}
 \begin{definition}[\reduceto]
-  A term $s$ \reduceto{} a name $x$ if there exists a
+  A $\beta$-normal term $s$ \reduceto{} a name $x$ if there exists a
   substitution $\rho$ such that $\rho s \Eo{} x$.
@@
 If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
-contraction. In that case the term $t$ under the spine of binders for
-$x_1\ldots x_n$ can either be $x$ itself applied to terms that can
+contraction. In that case the term $t$ is under the spine of binders
+$x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
 \reduceto{} these variables (case 2), or a unification variable
@@
 
-\noindent
-Note that this condition does not require the term to be in \llambda.
-\noteE[inline]{Is this relevant}
+% \noindent
+% Note that this condition does not require the term to be in \llambda.
+% \noteE[inline]{Is this relevant}
 
@@
 \end{definition}
+
+\noteD{Not too clear?}
 In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
@@
   \begin{enumerate}
-    \item $t$ is a constant or a variable applied to the arguments
+    \item $t$ is a constant or a name applied to the arguments
       $l_1 \ldots l_m$ such that 
-      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
+      $m \geq n$ and for every $i$ such that $m - n < i \leq m$
       the term  $l_i$
       \reduceto{} $x_i$, and
-      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
+      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n}$;
     \item $t$ is a
@@
 The rule triggers when the input term \elpiIn{flam F} is in
-\maybeeta. It compiles it to \elpiIn{lam F1} but puts the fresh
-variable \elpiIn{A} in its place. The variable sees all the names free in
+\maybeeta. It compiles \elpiIn{flam F} to \elpiIn{lam F1} but puts the fresh
+variable \elpiIn{A} in its place. This variable sees all the names free in
 \elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
@@
 \begin{proof}[Proof sketch]
-  By contradiction, suppose that the rule above triggered and that
+  By contradiction, suppose that the rule above is triggered and that
   the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
@@
 \linketa are meant to delay the unification of ``problematic'' terms until
-we know for sure if the head lambda has to be $\eta$-contracted or not.
+we know for sure if the term has to be $\eta$-contracted or not.
 

### 2024-04-05 paper/main.tex
typo

@@
 
-\newcommand{\linkbeta}{\texttt{link-}\ensuremath{\beta}\xspace}
-\newcommand{\linketa}{\texttt{link-}\ensuremath{\eta}\xspace}
+\newcommand{\linkMacro}[1]{\ensuremath{#1}\texttt{-link}\xspace}
+
+\newcommand{\linkbeta}{\linkMacro{\beta}}
+\newcommand{\linketa} {\linkMacro{\eta}}
 
@@
 
-TODO: What is commit-links and complete-mapping?, maybe complete-mapping can be
-hidden in the code rendering?
+\noteE{What is commit-links and complete-mapping?}
 
@@
 
-\subsection{Compilation}\label{sec:etacomp}
+\subsection{Compilation and decompilation}\label{sec:etacomp}
 
@@
 
+\noteD[inline]{Dire della commit-links: intuizione fa round trip}
+
 \subsection{Progress}\label{sec:etaprogress}
@@
 A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
-$X$ becomes rigid. There are two cases:
+$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
 \begin{enumerate}
   \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
-    we unify the $\eta$-expansion of the $X$ with $T$, that is we run
-    $\lambda x.X \appsep x\Ue{} T$  (under
-    the context $\Gamma$)
-    \noteE[inline]{where y comes from? $X = y$: $y$ is in the ctx of X}
+    we unify the $\eta$-expansion of $X$ with $T$, that is we run
+    $\lambda x.X \appsep x\Ue{} T$
     \item if $X = \lambda x.t$ we run $X \Ue{} T$.
@@
 
-\noteD[inline]{Should we proove simulation fidelity for \linketa insertion?}
+\noteD[inline]{Proove simulation fidelity, dicendo che \progressetaright è inutile}
 
@@
 
-\noteE[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
-}
+% \noteE[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
+% }
 \section{Handling of \maybebeta}\label{sec:beta}
@@
 
-\subsection{Compilation}
+\subsection{Compilation and decompilation}
 
@@
 
+\noteD[inline]{Dire della commit-links: intuizione non devono esserci beta per corollary}
+
 \subsection{Progress}
@@
   $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots t_m}$ with reduced
-  list of arguments and $Y$ being a fresh variable. Moreover, the new \linkbeta,
+  list of arguments and $Y$ being a fresh variable. Moreover, the new link
   \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to 
@@
 triggers $\linkStore_1$ by \progressetaleft 
-and $E$ is assigned to $\lambda x.x$.
+and $E_x$ is assigned to $x$.
 Under this substitution the \linkbeta becomes
@@
 Y \mapsto f\}$.
-\noteE{alla fine non si usa anche la regola beta-rigid-head?}
-\noteD{nope. Si risolve tutto by \progBetaLL}
 
@@
 
-\subsection{Relaxing \cref{lemma:keep-fidelity}}
+\subsection{Relaxing \cref{lemma:keep-fidelity} : tobe renamed}
 
@@
 
+\noteD[inline]{Dire qualcosa sulla commit-links}
+
 \section{Practical example}

### 2024-04-05 paper/main.tex
add todo after discussion of this morning
change macro for link names
rewrok title, abstract and intro

@@
 
-\title{HO unification from object language to meta language}
+% \title{HO unification from object language to meta language}
+\title{Higher order unification for free!}
+\subtitle{Reusing the meta-language unification for the object language}
 
@@
 \begin{abstract}
-  Specifying and implementing a logic from scratch requires significant effort.
+  Specifying and implementing a proof system from scratch requires significant effort.
   Logical Frameworks and Higher Order Logic Programming Languages provide
@@
   The object logic we are interested in is Coq's~\cite{Coq-refman}
-  Dependent Type Theory (DTT),
+  Calculus of Inductive Constructions (CIC),
   for which we aim to implement a unification procedure \Uo using the ML
@@
   fragment~\cite{miller92jsc}.
-  We want \Uo to be as powerful as \Ue but on the object logic DTT.
-  Elpi also comes with an encoding for DTT that works well
+  We want \Uo to be as powerful as \Ue but on the object logic CiC.
+  Elpi also comes with an encoding for CiC that works well
   for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
@@
 
-  We apply this technique to the implementation of a type-class~\cite{wadler89}
-  solver for Coq~\cite{Coq-refman}.
-  Type-class solvers are proof search procedures based on
-  unification that back-chain designated lemmas, providing essential
-  automation to widely used
-  Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-  and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
-  constitute our test bed.
+  % We apply this technique to the implementation of a type-class~\cite{wadler89}
+  % solver for Coq~\cite{Coq-refman}.
+  % Type-class solvers are proof search procedures based on
+  % unification that back-chain designated lemmas, providing essential
+  % automation to widely used
+  % Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
+  % and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
+  % constitute our test bed.
 \end{abstract}
@@
 
-Specifying and implementing a logic from scratch requires significant effort.
+Specifying and implementing a logic or a proof system from scratch requires
+significant effort.
 Logical Frameworks and Higher Order Logic Programming Languages provide
 dedicated, high-level Meta Languages (ML) to facilitate this task in two
-key ways: 1) variable binding and substitution are simplified when ML binders
-represent object logic ones; 2) proof construction, and even proof search, is
-greatly simplified by leveraging the unification procedure provided by the ML.
+key ways. The first and most well know one is that variable binding and
+substitution can be taken for granted when ML binders represent object
+logic ones. The second one that comes to mind is unification, the cornerstone
+for proof construction and proof search, however in this paper we describe how
+reusing that brick may not be as easy at is seems.
+
 Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
@@
 Constructions~\cite{felty93lics}.
-
 The object logic we are interested in is Coq's~\cite{Coq-refman}
-Dependent Type Theory (DTT), and we want to code a type-class~\cite{wadler89}
-solver for Coq~\cite{Coq-refman} using the Coq-Elpi~\cite{tassi:hal-01637063}
+Calculus of inductive Constructions (Cic), and we want to code a
+type-class~\cite{wadler89} solver for Coq~\cite{Coq-refman} using
+the Coq-Elpi~\cite{tassi:hal-01637063}
 meta programming framework.
@@
 
-As the running example we take the \coqIn{Decide} type class,
+As an example we take the \coqIn{Decide} type class
 from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
@@
 type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
-smaller than \coqIn{n} is finite; 2) the predicate \coqIn{nfact n nf},
+smaller than \coqIn{n}, is finite; 2) the predicate \coqIn{nfact n nf},
 linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
-3) the universal closure of a predicate has a decision procedure if the
-predicate has and if its domain is finite.
+3) the universal closure of a predicate has a decision procedure if
+its domain is finite and if the predicate is decidable.
 
@@
 \noindent Under this context of instances a type-class solver is able to prove
-the following statement automatically by back-chaining.
+the following statement automatically by back-chaining the lemmas above:
 
@@
 \noindent
-The encoding of DTT provided by Elpi, that we will discuss at length later in
-section~\ref{sec:encodings,sec:lang-spec} and ~\ref{}, is an Higher Order Abstract
+The encoding of CiC provided by Elpi, that we will discuss at length later in
+\cref{sec:encodings,sec:lang-spec}, is an Higher Order Abstract
 Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
@@
 %The root cause is that
-%\ref{problem:p} is an higher order in DTT, but becomes
+%\ref{problem:p} is an higher order in CiC, but becomes
 %first order in the meta language due to the ``naive'' encoding.
@@
 us to rephrase the problematic rule \ref{clause:r3} as follows:
+\todo{why not link?}
 
@@
 \noindent
-After unifying the head of rule \ref{clause:r3a} with the goal, Elpi runs
-the premise <<\elpiIn{decomp Pm A P}>> that is in charge of bringing the
+After unifying the head of rule \ref{clause:r3a} with the goal, Elpi would run
+the premise <<\elpiIn{decomp Pm A P}>> in charge of bringing the
 assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:
@@
 (Coq) $\beta$-redex in the second premise (the predicate
-under the \elpiIn{pi w\ }\hspace{-0.4em}):
+under the \elpiIn{pi w\ }\hspace{-0.4em}). We show below the premise before and
+after the instantiation of \elpiIn{P}:
 
 \begin{elpicode}
-decision (app[lam A (a\ app[con"nfact", a, con"3"]), w])
+decision (app[                    P                  , w])
+decision (app[ lam A (a\ app[con"nfact", a, con"3"]) , w])
 \end{elpicode}
@@
 \begin{elpicode}
-app[lam A (a\ app[con"nfact", a, con"3"]), x] =
-app[con"nfact", N, NF]
+app[ lam A (a\ app[con"nfact", a, con"3"]) , x] =
+app[ con"nfact"                            , N, NF]
 \end{elpicode}
@@
 ~\\
-The root cause of the problems we sketched in the running example
+The root cause of the problems we sketched in this example
 is that the unification procedure \Ue of the meta language is not aware
@@
 To state precisely the problem we solve we need a \Fo{} representation
-of DTT terms and a \Ho one.
+of CiC terms and a \Ho one.
 We call \Eo the equality over ground terms in \Fo,
@@
 
-\section{Alternative encodings and related work}
+\section{Alternative encodings and related work}\label{sec:encodings}
 
@@
 
-Our encoding of DTT may look ``semi shallow'' since we use the meta-language
+Our encoding of CiC may look ``semi shallow'' since we use the meta-language
 lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
@@
 There are two reasons for dismissing this encoding. The first one is that
-in DTT it is not always possible to adopt it since the type system
+in CiC it is not always possible to adopt it since the type system
 of the meta language is too weak to accommodate terms with a variable arity,
@@
 
-In the literature we could find a few related encoding of DTT.
+In the literature we could find a few related encoding of CiC.
 TODO In~\cite{felty93lics} is related and make the
-discrepancy between the types of ML and DTT visible. In this case
+discrepancy between the types of ML and CiC visible. In this case
 one needs 4 application nodes. Moreover the objective is an encoding
@@
 are first class terms, i.e. they have a concrete syntax. We keep these languages
-minimal, for example, we omit the \elpiIn{all} quantifier of DTT we used
+minimal, for example, we omit the \elpiIn{all} quantifier of CiC we used
 in the example in Section~\ref{sec:intro} together with the type notation of

### 2024-04-05 paper/main.tex
sketch conclusion

@@
 
-\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
+\section{Problem statement and solution} %%%%%%%%%%%%%%%%%%%%%%
 \label{sec:problem-statement}
@@
 This choice would underuse the logic programming engine provided by
-the metalanguage since by removing any datum from the head of rules
-indexing degenerates. Moreover the unification procedure built in the
-meta language is likely to be faster than one implemented in it,
-especially if the meta language is interpreted as Elpi is.
+the meta-language since, by removing any datum from the head of rules,
+indexing degenerates. Moreover the unification procedure \elpiIn{unif}
+programmed in the meta-language is likely to be n order of magnitude
+slower than one that is built-in.
 
@@
 
-\section{Alternative encodings and related work}\label{sec:encodings}
+\section{Other encodings and related work}\label{sec:encodings}
 
@@
 
-\section{Practical example}
+\section{Actual implementation}
+
+In this paper we show a minimized example. The full code is there.
+But we also have to code things in Coq.
+In Elpi we don't have a main loop, we rely on the interpreter one.
+So links are constraints and progress are CHR rules.
+
+\begin{elpicode}
+link-eta bla :- suspend-condition, !, declare_constraint (link-eta bla) Vars.
+link-eta bla :- progress. % solves the goal, hence the constraint is not more in L.
+constraints link-beta {
+  % undup
+  rule (G1 ?- link-eta (uvar X LX1) T1) /
+       (G2 ?- link-eta (uvar X LX2) T2) |
+       relocate LX1 LX2 T2 T2' <=>
+       (G1 ?- T1 = T2').
+}
+\end{elpicode}
 
@@
 
+Benefits: less work, reuse efficient ho unif (3x faster), indexing,
+
+Future: tabling and static analysis (reuse for ML again).
+
+Very little is Coq specific. Applies to all OL that are not a subsystem of 
+HOL, or for ML that are used for meta programming.
+
 \printbibliography

### 2024-04-06 paper/main.tex
some work on the intro

@@
   fragment~\cite{miller92jsc}.
-  We want \Uo to be as powerful as \Ue but on the object logic CiC.
-  Elpi also comes with an encoding for CiC that works well
+  We want \Uo to be as powerful as \Ue but on the object logic CIC.
+  Elpi also comes with an encoding for CIC that works well
   for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
@@
 
-Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
+Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
 $\lambda$Prolog~\cite{miller_nadathur_2012} and
 Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-which have been utilized to implement various formal systems such as
-First Order Logic~\cite{felty88cade},
-Set Theory~\cite{10.1007/BF00881873},
-Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
-Constructions~\cite{felty93lics}.
+have been utilized to specify various
+logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
+In some cases, the most notable one being Higher Order Logic~\cite{books/sp/NipkowPW02},
+the ML Isabelle is such a good fit that it implements an interactive proof
+system for HOL, and not just a specification.
+
 The object logic we are interested in is Coq's~\cite{Coq-refman}
-Calculus of inductive Constructions (Cic), and we want to code a
+Calculus of Inductive Constructions (CIC), and we want to implement a
 type-class~\cite{wadler89} solver for Coq~\cite{Coq-refman} using
-the Coq-Elpi~\cite{tassi:hal-01637063}
-meta programming framework.
+the Coq-Elpi~\cite{tassi:hal-01637063} meta programming language,
+a dialect of $\lambda$Prolog already used to extend
+Coq~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
 Type-class solvers are unification based proof search procedures
-that combine a set of designated lemmas in order to providing essential
-automation to widely used Coq libraries.
-
-As an example we take the \coqIn{Decide} type class
+that  provide essential automation to widely used Coq libraries.
+These solvers are reminiscent of Prolog: they back-chain lemmas taken
+from a designated database of type class instances.
+
+As an example we take the \coqIn{Decision} type class
 from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-library. The class identifies predicates equipped with a decision procedure.
-The following three designated lemmas (called \coqIn{Instances} in the
-type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
-smaller than \coqIn{n}, is finite; 2) the predicate \coqIn{nfact n nf},
-linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
+library. The class
+identifies predicates equipped with a decision procedure.
+The following three designated \coqIn{Instances} state that:
+1) the type \coqIn{fin n}, of natural numbers
+smaller than \coqIn{n}, is \coqIn{Finite} (another type class);
+2) the predicate \coqIn{nfact n nf}, linking a natural number
+\coqIn{n} to its prime factors \coqIn{nf}, is decidable;
 3) the universal closure of a predicate has a decision procedure if
@@
 
-\noindent Under this context of instances a type-class solver is able to prove
-the following statement automatically by back-chaining the lemmas above:
+\noindent Given this database of instances a type-class solver is
+able to prove the following statement automatically,
+by back-chaining the lemmas above:
 
 \begin{coqcode}
-  Check _ : Decision (forall y: fin 7, nfact y 3).       ~\customlabel{goal:g}{(g)}~
+  Check _ : Decision (forall n: fin 7, nfact n 3).       ~\customlabel{goal:g}{(g)}~
 \end{coqcode}
@@
 \noindent
-The encoding of CiC provided by Elpi, that we will discuss at length later in
+The encoding of CIC provided by Elpi, that we will discuss at length later in
 \cref{sec:encodings,sec:lang-spec}, is an Higher Order Abstract
@@
 %The root cause is that
-%\ref{problem:p} is an higher order in CiC, but becomes
+%\ref{problem:p} is an higher order in CIC, but becomes
 %first order in the meta language due to the ``naive'' encoding.
@@
 To state precisely the problem we solve we need a \Fo{} representation
-of CiC terms and a \Ho one.
+of CIC terms and a \Ho one.
 We call \Eo the equality over ground terms in \Fo,
@@
 
-Our encoding of CiC may look ``semi shallow'' since we use the meta-language
+Our encoding of CIC may look ``semi shallow'' since we use the meta-language
 lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
@@
 There are two reasons for dismissing this encoding. The first one is that
-in CiC it is not always possible to adopt it since the type system
+in CIC it is not always possible to adopt it since the type system
 of the meta language is too weak to accommodate terms with a variable arity,
@@
 
-In the literature we could find a few related encoding of CiC.
+In the literature we could find a few related encoding of CIC.
 TODO In~\cite{felty93lics} is related and make the
-discrepancy between the types of ML and CiC visible. In this case
+discrepancy between the types of ML and CIC visible. In this case
 one needs 4 application nodes. Moreover the objective is an encoding
@@
 are first class terms, i.e. they have a concrete syntax. We keep these languages
-minimal, for example, we omit the \elpiIn{all} quantifier of CiC we used
+minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
 in the example in Section~\ref{sec:intro} together with the type notation of
@@
 
-\section{Actual implementation}
+\section{Actual implementation in Elpi}
 
 In this paper we show a minimized example. The full code is there.
-But we also have to code things in Coq.
+But we also have to code things in Coq-Elpi.
+
 In Elpi we don't have a main loop, we rely on the interpreter one.
 So links are constraints and progress are CHR rules.
+Constraints are suspended goals that are resumend when some unif variable
+is assigned.
 
 \begin{elpicode}
-link-eta bla :- suspend-condition, !, declare_constraint (link-eta bla) Vars.
+link-eta bla :- suspend-condition Holes, !, declare_constraint (link-eta bla) Holes.
 link-eta bla :- progress. % solves the goal, hence the constraint is not more in L.
-constraints link-beta {
-  % undup
-  rule (G1 ?- link-eta (uvar X LX1) T1) /
-       (G2 ?- link-eta (uvar X LX2) T2) |
-       relocate LX1 LX2 T2 T2' <=>
-       (G1 ?- T1 = T2').
+\end{elpicode}
+
+This matches with all progress rules but for the ones considering
+two links. For these we need to resort to CHR to manipulate L. For example
+deduplicate eta
+
+\begin{elpicode}
+constraints link-eta {
+  rule (N1 : G1 ?- link-eta (uvar X LX1) T1) % match
+    /  (N2 : G2 ?- link-eta (uvar X LX2) T2) % remove
+    |  (relocate LX1 LX2 T2 T2')             % condition
+   <=> (N1 : G1 ?- T1 = T2').                % new goal
 }
@@
 
+Remark how the invariant about uvar arity makes this easy, since LX1 and LX2
+have the same length. Also note that N1 only contains the names of the first
+link (while relocate runs in the disjoint union) and Elpi ensures that
+T2' can live in N1.
+
 \section{Conclusion}

### 2024-04-08 paper/main.tex
wip tex: Add decompilation

@@
 \begin{proof}[Proof sketch]
-trivial, since the terms are beta normal beta just builds an app.
+trivial, since the terms are beta normal beta just builds an app. \noteD{Reformulate?}
 \end{proof}
@@
 % term compilation.
+
+\paragraph{Compilation}
 The following rule is inserted just before rule~\ref{rule:complam} from the code in
@@
 
-\noteD[inline]{Dire della commit-links: intuizione fa round trip}
+\paragraph{Decompilation}
+Decompilation of \linketa is performed by adding new rules to the
+\elpiIn{commit-link} predicate. In particular, given \linketaM{\Gamma}{X}{t}, we
+can note that this unification never fails, since $X$ is a flexible term and no
+other \linketa has $X$ has \lhs (by \cref{def:progressetadedup}). The link is
+remove from \linkStore and \elpiIn{commit-links} terminates.
 
@@
   terms in \wellb, therefore, the substitution remains \wellb.
+  \noteD[inline]{Bisogna aggiungere un lemma nella \cref{sec:nutshell} che dice che
+  unificare due termini in \wellb, in una $\sigma$, tale che $\wellb(\sigma)$, 
+  non invalida \wellb}
 \end{proof}
@@
   s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
+  \label{inv:beta-rhs}
 \end{invariant}
@@
 
+\paragraph{Compilation}
 Detection of \maybebeta is quite simple to implement in the compiler, since it
@@
 
-\noteD[inline]{Dire della commit-links: intuizione non devono esserci beta per corollary}
+\paragraph{Decompilation}
+During \elpiIn{progress}, as claimed in \cref{inv:beta-rhs}, the decompilation
+can only have \linkbeta with not instantiated \lhs. In this case, \lhs is
+unified with \rhs.
+\noteD[inline]{not really sure of this, we can have $F\appsep a = \lambda x.G x$.
+In this case when do we fail: for sure in decompile. But to respect fidelity,
+we should fail immediately: we have a \linkbeta and a \linketa with same \lhs}
 
@@
 \newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}}
+\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}}
 
@@
 
+\noteD{added the def below to respect fidelity}
+% Example of this `X a = Y a'
+\begin{definition}[\progBetaDedup]
+  Given two \linkbeta $l1$ and $l2$, if they share the same \lhs, unification
+  fails.
+  \label{def:progBetaDedup}
+\end{definition}
+
 \begin{lemma}
@@
 
-\noteD[inline]{Dire qualcosa sulla commit-links}
+\noteD{Paragraph for decompile in case of \maybebeta}
+The \elpiIn{commit-link} predicate can be extended to add heuristics if during
+the decompilation phase \linkbeta remain. For example, the same approximation
+explained above can be delayed and applied only if the terms in \maybebeta never
+falls in \llambda after the execution of all the unification problems. We want
+to point out, that we call this \emph{approximation}, since we are
+making a choice among all the possible unifiers and therefore, we can pick
+the wrong one.
 

### 2024-04-08 paper/main.tex
more work on itro

@@
 %\setcctype{by}
-\copyrightyear{2024}
-\acmYear{XXXX 2024}
-\acmBooktitle{YYY}
-\acmDOI{ZZZZZZZZZZZZ}
+%\copyrightyear{2024}
+%\acmYear{XXXX 2024}
+%\acmBooktitle{YYY}
+%\acmDOI{ZZZZZZZZZZZZ}
 
@@
 key ways. The first and most well know one is that variable binding and
-substitution can be taken for granted when ML binders represent object
-logic ones. The second one that comes to mind is unification, the cornerstone
-for proof construction and proof search, however in this paper we describe how
-reusing that brick may not be as easy at is seems.
+substitution come for free is one uses the ML binders to represent the object
+language ones. The second one that comes to mind is unification, a cornerstone
+for proof construction and proof search. However reusing that brick may require
+some additional work that we carry out in this paper.
 
@@
 logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
-In some cases, the most notable one being Higher Order Logic~\cite{books/sp/NipkowPW02},
+In the notable of Higher Order Logic~\cite{books/sp/NipkowPW02},
 the ML Isabelle is such a good fit that it implements an interactive proof
-system for HOL, and not just a specification.
-
+system for the object logic, and not just a specification.
 The object logic we are interested in is Coq's~\cite{Coq-refman}
-Calculus of Inductive Constructions (CIC), and we want to implement a
-type-class~\cite{wadler89} solver for Coq~\cite{Coq-refman} using
-the Coq-Elpi~\cite{tassi:hal-01637063} meta programming language,
+Calculus of Inductive Constructions (CIC) and we want to implement a
+form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
+We want to use the Elpi~\cite{tassi:hal-01637063} meta programming language,
 a dialect of $\lambda$Prolog already used to extend
-Coq~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
+Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
 Type-class solvers are unification based proof search procedures
-that  provide essential automation to widely used Coq libraries.
-These solvers are reminiscent of Prolog: they back-chain lemmas taken
-from a designated database of type class instances.
-
-As an example we take the \coqIn{Decision} type class
+reminiscent of Prolog: they back-chain lemmas taken
+from a designated database of ``type class instances''.
+For this reason we believe that $\lambda$Prolog, hence Elpi, is a good fit for
+implementing such as form of automation. In this paper we focus on one aspect of
+this work, namely \emph{how to reuse the unification of the ML Elpi in order to
+implement the one used by the type-class solver}.
+
+We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
 from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-library. The class
-identifies predicates equipped with a decision procedure.
+library. The class  \coqIn{Decision}
+identifies predicates equipped with a decision procedure, while
+\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
 The following three designated \coqIn{Instances} state that:
 1) the type \coqIn{fin n}, of natural numbers
-smaller than \coqIn{n}, is \coqIn{Finite} (another type class);
+smaller than \coqIn{n}, is finite;
 2) the predicate \coqIn{nfact n nf}, linking a natural number
-\coqIn{n} to its prime factors \coqIn{nf}, is decidable;
+\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
 3) the universal closure of a predicate has a decision procedure if
@@
 \begin{coqcode}
-Instance fin_fin n : Finite (fin n).              (* r1 *)
-Instance nfact_dec n nf : Decision (nfact n nf).  (* r2 *)
-Instance forall_dec A P : Finite A ~$\to$~             (* r3 *)
+Instance fin_fin: ~$\forall$~n, Finite (fin n).             (* r1 *)
+Instance nfact_dec: ~$\forall$~n nf, Decision (nfact n nf). (* r2 *)
+Instance forall_dec: ~$\forall$~A P, Finite A ~$\to$~            (* r3 *)
   ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
@@
 
-\noindent Given this database of instances a type-class solver is
-able to prove the following statement automatically,
-by back-chaining the lemmas above:
+\noindent Given this database a type-class solver is expected to
+prove the following statement automatically:
 
 \begin{coqcode}
-  Check _ : Decision (forall n: fin 7, nfact n 3).       ~\customlabel{goal:g}{(g)}~
+  Decision (~$\forall$~x: fin 7, nfact x 3)                   (* g *)
 \end{coqcode}
@@
 \noindent
-The encoding of CIC provided by Elpi, that we will discuss at length later in
-\cref{sec:encodings,sec:lang-spec}, is an Higher Order Abstract
-Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
+The proof found by the solver back-chains on rule 3 (the only rule
+about the $\forall$ quantifier), and then solves the premises with rules
+rules 1 and 2 respectively.
+Note that rule 3 features a second order parameter \coqIn{P} that stands for
+a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
+The solver has to infer a value for \coqIn{P} by unifying the conclusion
+of rule 3 with the goal, and in particular by solving the unification
+problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
+called pattern-fragment \llambda~\cite{miller92jsc} and hence admits a unique
+solution \coqIn{P = ~$\lambda$~x.nfact x 3}.
+
+In order to implement such a search in Elpi we shall describe the encoding
+of CIC terms and then the encoding of rules. Elpi comes with
+an Higher Order Abstract Syntax~\cite{10.1145/53990.54010} datatype of CIC
+terms, called \elpiIn{tm}, that features (among others) the following
 constructors:
@@
 \noindent
-Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
-the concrete syntax to abstract, at the meta level, an expression
-\elpiIn{e} over a variable \elpiIn{x}
-is <<\elpiIn{x\ e}>>, and square brackets denote a list of
-terms separated by comma. As an example we show the encoding of the Coq term
-<<\coqIn{~$\forall$~y:t, nfact y 3}>>:
+Following $\lambda$Prolog~\cite{miller_nadathur_2012}'s standard syntax,
+the meta level binding of a variable \elpiIn{x} in an expression
+\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets denote a
+list of terms separated by comma. For example the term
+<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:
 
@@
 logic-programming rules: capital letters denote rule
-parameters; \elpiIn{:-} separates the rule's head from the premises;
+parameters; \elpiIn{:-} separates the rule's head from the premises and
 \elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
@@
 \begin{elpicode}
-finite (app[con"fin", N]).                            ~\customlabel{clause:r1}{(r1)}~
+finite   (app [con"fin", N]).                         ~\customlabel{clause:r1}{(r1)}~
 decision (app [con"nfact", N, NF]).                   ~\customlabel{clause:r2}{(r2)}~
-decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
-  pi w\ decision (app[P, w]).
+decision (all A x\ app [P, x]) :- finite A,           ~\customlabel{clause:r3}{(r3)}~
+  pi w\ decision (app [P, w]).
 \end{elpicode}
@@
 \noindent
-Unfortunately this translation of rule \ref{clause:r3} uses the
-predicate \coqIn{P} as a first order term: for the meta
-language its type is \elpiIn{tm}.
-If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
-\ref{goal:g} given below
+Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work,
+since it uses the predicate \coqIn{P} as a first order term: for the meta
+language its type is \elpiIn{tm}. If we try to back-chain the rule
+\ref{clause:r3} on the encoding of the goal g given below
 
 \begin{elpicode}
-decision (all (app[con"fin", con"7"]) y\
-  app[con"nfact", y, con"3"]).
+decision (all (app [con"fin", con"7"]) y\              ~\customlabel{goal:g}{(g)}~
+  app [con"nfact", y, con"3"]).
 \end{elpicode}
@@
 \begin{elpicode}
-app[con"nfact", y, con"3"] = app[P, y]                 ~\customlabel{problem:p}{(p)}~
+app [con"nfact", y, con"3"] = app [P, y]               ~\customlabel{problem:p}{(p)}~
 \end{elpicode}
@@
 \noindent
-In this paper we study a more sophisticated encoding of Coq terms allowing
-us to rephrase the problematic rule \ref{clause:r3} as follows:
-\todo{why not link?}
+In this paper we study a more sophisticated encoding of rules that, on a first
+approximation, would shape \ref{clause:r3} as follows:
 
 \begin{elpicode}
-decision (all A x\ Pm x) :- decomp Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
-  pi x\ decision (app[P, x]).
+decision (all A x\ Pm x) :- link Pm P A, finite A,    ~\customlabel{clause:r3a}{(r3')}~
+  pi x\ decision (app [P, x]).
 \end{elpicode}
@@
 \begin{elpicode}
-app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(p')}~
-Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
-A = app[con"fin", con"7"]              % assignment for A
+app [con"nfact", y, con"3"] = Pm y                    ~\customlabel{problem:pa}{(p')}~
+Pm = x\ app [con"nfact", x, con"3"]                   ~\customlabel{solution:pm}{(\rho)}~
 \end{elpicode}
@@
 \noindent
-After unifying the head of rule \ref{clause:r3a} with the goal, Elpi would run
-the premise <<\elpiIn{decomp Pm A P}>> in charge of bringing the
-assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:
+Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g}
+the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
+back to the domain \elpiIn{tm} of Coq terms:
 
 \begin{elpicode}
-P = lam A a\ app[con"nfact", a, con"3"]
+P = lam A a\ app [con"nfact", a, con"3"]
 \end{elpicode}
@@
 decision (app[                    P                   , w])
-decision (app[ lam A (a\ app[con"nfact", a, con"3"]) , w])
+decision (app[ lam A (a\ app [con"nfact", a, con"3"]) , w])
 \end{elpicode}
@@
 \begin{elpicode}
-app[ lam A (a\ app[con"nfact", a, con"3"]) , x] =
+app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
 app[ con"nfact"                             , N, NF]
@@
 include $\eta\beta$-conversion and admit most general
-unifiers for unification problems in the pattern fragment \llambda~\cite{miller92jsc}.
+unifiers for unification problems in the pattern-fragment.
 
 \paragraph{Contributions}
-In this paper we discuss alternative encodings of Coq in
-Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
-in which the problems sketched here can be fully described.
-We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
-the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
-bla bla.. \noteE[inline]{citare Teyjus}.
-The code discussed in the paper can be accessed at the URL:
-\url{https://github.com/FissoreD/paper-ho}.
+In this paper we identify a minimal language \Fo{} in which the problems
+sketched in the introduction can be formally described.
+We detail an encoding of a logic program on \Fo{} to a strongly related
+logic program in \Ho (the language of the meta language) and we show that
+the unification procedure of the meta language \Ue{} can be effectively
+used to simulate a unification procedure \Uo for the object language that
+features $\eta\beta$-conversion in the pattern-fragment.
+
+\cref{sec:problem-statement} formally states the problem and gives the
+intuition behind our solution. \cref{sec:encodings} discusses alternative
+term encodings and related works. \cref{sec:lang-spec} introduces
+the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
+basic simulation of higher order logic programs.
+\cref{sec:eta,sec:invariant1} completes its equational theory
+with support for $\eta$-conversion. \cref{sec:beta} deals with the
+practical necessity of ``tolerating'' terms outside of the
+pattern-fragment and discusses how heuristic can be applied.
+Finally \cref{sec:implementation} discusses the implementation in Elpi.
+
+The $\lambda$Prolog code discussed in the paper can be accessed at the
+URL: \url{https://github.com/FissoreD/paper-ho}.
 
@@
 \label{sec:problem-statement}
-
-The equational theory of Coq's Dependent Type Theory is very rich. In
-addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
-are compared up to proposition unfolding and fixpoint unrolling. Still,
-for efficiency and predictability reasons, most form of automatic proof search
-employ a unification procedure that captures a simpler one,
-just $\eta\beta$, and that solves higher-order problems
-restricted to the pattern fragment $\llambda$~\cite{miller92jsc}.
-We call this unification procedure \Uo{}.
-
-The equational theory of the meta language Elpi is strikingly similar,
-since it it comprises $\eta\beta$ (for the meta language functions), and the
-unification procedure \Ue{} solves higher-order problems in
-$\llambda$.
-
-In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
-since the abstraction and application term constructors
-the two unification procedures deal with are different. For example
-
-\begin{tabular}{lcl}
-\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
-\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"}\\
-\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"} \\
-\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
-\end{tabular}
-
-\noindent
-One could ignore this similarity, and ``just'' describe the object language
-unification procedure in the meta language, that is crafting a \elpiIn{unif}
-predicate to be used as follows in rule \ref{clause:r3}:
-
-\begin{elpicode}
-decision X :- unif X (all A x\ app[P, x]), finite A,
-  pi x\ decision (app[P, x]).
-\end{elpicode}
-
-\noindent
-This choice would underuse the logic programming engine provided by
-the meta-language since, by removing any datum from the head of rules,
-indexing degenerates. Moreover the unification procedure \elpiIn{unif}
-programmed in the meta-language is likely to be n order of magnitude
-slower than one that is built-in.
-
-To state precisely the problem we solve we need a \Fo{} representation
-of CIC terms and a \Ho one.
-We call \Eo the equality over ground terms in \Fo,
-\Ee the equality over ground terms in \Ho,
-\Uo the unification procedure we want to implement and
-\Ue the one provided by the meta language.
-\noteE[inline]{extend \Eo and \Ee with reflexivity on uvars.}
 
@@
 }
+\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
+\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
+\newcommand{\notllambda}{\ensuremath{\overline{\llambda}}\xspace}
+%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}
+
+The equational theory of Coq's Dependent Type Theory is very rich. In
+addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
+are compared up to proposition unfolding and fixpoint unrolling. Still,
+for efficiency and predictability reasons, most form of automatic proof search
+employ a unification procedure that captures a simpler one,
+just $\eta\beta$, and that solves higher-order problems
+restricted to the pattern fragment \llambda.
+We call this unification procedure \Uo{}.
+
+The equational theory of the meta language Elpi is strikingly similar,
+since it it comprises $\eta\beta$ (for the meta language functions), and the
+unification procedure \Ue{} solves problems in
+\llambda as well.
+
+In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
+since the abstraction and application term constructors
+the two unification procedures deal with are different. For example:
+%
+\setlength{\tabcolsep}{2pt}
+\vspace{4pt}
+\begin{tabular}{lcl|}
+\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
+\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
+\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
+\end{tabular}
+\phantom{mn}
+\begin{tabular}{lcl}
+\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
+\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
+\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
+\end{tabular}
+\vspace{4pt}
+
+To state precisely the problem we solve we need a \Fo{} representation
+of CIC terms and a \Ho one. We extend the equality over ground
+terms to open ones with the reflexivity of unification variables (that is
+unification variable X is equal to itself, not to another one such a Y).
+We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
+We call \Uo the unification procedure we want to implement and
+\Ue the one provided by the meta language.
+We assume that the unification of our meta
+language is correct and complete in \llambda.
+%
+%\begin{gather}
+%  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
+%  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
+%\end{gather}
 
@@
 with substitution $\sigma$; we write $\sigma t$ for the application of
-the substitution to $t$, and $\sigma X = \{ \sigma t | t \in X\}$ when
+the substitution to $t$, and $\sigma X = \{ \sigma t ~|~ t \in X\}$ when
 $X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
-general than $\sigma'$. We assume that the unification of our meta
-language is correct:
-%
-\begin{gather}
-  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
-  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
-\end{gather}
+general than $\sigma'$ and we use set union to denote the concatenation of
+two substitutions. We shall use $\rho$ for \Fo{} substitutions, and
+$\sigma$ for the \Ho ones.
 
@@
 We illustrate a compilation $\C{s}{t}{m}{l}$ that
-maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
-list of links $l$.
+maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
+a variable mapping $m$ and list of links $l$.
 The variable map connects unification variables in \Ho with variables
 in \Fo{} and is used to ``decompile'' the assignment,
-$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
-are linked to the unification variable that stands in their place in the
-compiled term. These links are checked for or progress \noteE[inline]{XXX improve....}
+$\D{\sigma}{m}{l}{\rho}$.
+Links are an accessory piece of information whose
+description is deferred to \cref{sec:nutshell}.
 
 We represent a logic program \emph{run} in \Fo{} as
-a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
-unification problem between terms $\foUnifPb_{p_l}$ and
+a list \emph{steps} $p$ of length $\mathcal{N}$.
+Each step $p$ represents 
+a unification problem between two terms $\foUnifPb_{p_l}$ and
 $\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
@@
 rules, with fresh unification variables each time}
-The initial here $\rho_0$ is the empty substitution
 %
@@
 
-We simulate each run in \Fo{} with a run in \Ho as follows.
-Note that $\sigma_0$ is the empty substitution.
+\noindent We simulate each run in \Fo{} with a run in \Ho as follows.
+%Note that $\sigma_0$ is the empty substitution.
 $$
@@
 \noindent
-Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
-terms) and a call to \progress{} on the set of links. We claim the following:
+Here \hstep{} is made of two sub-steps: a call to the meta language
+unification and a check for \progress{} on the set of links.
+We claim:
 
 \begin{proposition}[Simulation]\label{prop:simulation}
-$\forall \foUnifPb, \forall \mathcal{N},$
+$\forall \foUnifPb, \forall \mathcal{N},$ if $~\foUnifPb \subseteq \llambda$
 $$
@@
 \begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of~ \hrun, if $~\hoUnifPb \subseteq \llambda$ we have that
+In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
 $\forall p \in 1 \ldots \mathcal{N},$
@@
 \Leftrightarrow
-\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
+\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
 $$
@@
 
-XXX permuting hrun does not change the final result if check does not fail eagerly
-
-XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to
-a non MGU too early
-
-We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
-$\foUnifPb = \{ s_1, s_2 \}$ as follows:
+%\cref{sec:beta}
+%XXX permuting hrun does not change the final result if check does not fail eagerly
+% XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to a non MGU too early
+
+We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
+%$\foUnifPb = \{ s_1, s_2 \}$ 
+as follows:
 %
@@
 
-\begin{proposition}[Properties of \Uo{}]
+\begin{proposition}[Properties of \Uo{} in \llambda]
 \begin{gather}
-  \unifcorrect{\Eo}{\Uo}{s}\customlabel{prop:correct}{(correct)}\\
-\unifcomplete{\Eo}{\Uo}{s}\customlabel{prop:complete}{(complete)}\\
-% \forall \rho'\rho,
-  \rho s_1 \Eo \rho s_2 \Rightarrow
-  \rho' \subseteq \rho \Rightarrow
-  \rho's_i \in \llambda \Rightarrow
-  \rho' s_1 \Uo \rho' s_2 %\label{prop:simulation}
+  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
+  \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
 \end{gather}
@@
 
-Properties \ref{prop:correct} and \ref{prop:complete} state, respectively, that
-in \llambda the implementation of \Uo is correct, complete and returns the most
-general unifier.
-
-\noteE[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
-is resilient to unification problems outside \llambda solved by
-a third party. We believe this property is of practical interest since we
-want the user to be able to add heuristics via hand written rules
-to the ones obtained by our compilation scheme. A Typical example
-is the following problem \ref{problem:q} that is outside \llambda:
-
-\begin{elpicode}
-app [F, con"a"] = app[con"f", con"a", con"a"]          ~\customlabel{problem:q}{(q)}~
-F = lam x\ app[con"f",x,x]                             ~\customlabel{heuristic:h}{(h)}~
-\end{elpicode}
-
-\noindent
-Instead of rejecting it our scheme accepts it and guarantees that if
-\ref{heuristic:h} is given (after the compilation part of the scheme, as
-a run time hint) then ...
+\noindent
+In addition to correctness and completeness in \llambda, we claim that \hrun
+handles terms outside \llambda in the following sense:
+
+\begin{proposition}[Properties of \hrun outside \llambda]
+\begin{gather}
+  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
+  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
+  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
+  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow
+  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \mapsto \sigma \label{prop:step-nllam}
+\end{gather}
+\end{proposition}
+
+\noindent
+Property \ref{prop:compilation-nllam} states that two terms for which
+there is a unifier (given by an oracle, not necessarily a most general one),
+then the compiler generates two terms that unify in \Ho. In other words
+the compiler moves out of the way all problematic terms.
+
+Property \ref{prop:step-nllam} says that if the two terms involved in a step
+re-enter \llambda, then \hstep succeeds. This is a typical example
+in which the order of the unification problems in a logic program run
+do matter. The simplest example is the sequence $F = \lambda x.a$ and
+$F \appsep a = a$: the second problem is not in \llambda and has two unifiers,
+namely $F = \lambda x.x$ and $F = \lambda x.a$. The first problem picks one
+of the two, making the second problem re-enter \llambda.
 
@@
 ``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
-and an accessory link that represent a suspended unification problem
+and an accessory \emph{link} that represent a suspended unification problem
 $h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
@@
 
-\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
-\newcommand{\maybebeta}{\ensuremath{\overline{\llambda}}\xspace}
-%\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
 \begin{definition}[\maybeeta]\label{def:maybeeta}
@@
 An example of term $t$ in \maybeeta{} is
-$\lambda x.\lambda y.F~y~x$
+$\lambda x.\lambda y.F \appsep y \appsep x$
 since the substitution
@@
 its rigid part, the $\lambda$-abstractions, cannot justify a
-unification failure against, say, a constant.
+unification failure against a constant or an application.
 
 \begin{definition}[\maybebeta]\label{def:maybebeta}
-  $\maybebeta = \{ X t_1 \ldots t_n ~|~ X t_1 \ldots t_n \not\in \llambda \}$.
+  $\maybebeta = \{ X \appsep x_1 \ldots x_n ~|~ x_1 \ldots x_n ~\mathrm{are~distinct~names} \}$
 \end{definition}
@@
 \noindent
-An example of $t$ in \maybebeta{} is $F a$ for a constant $a$. Note however tha
-an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
-that makes the resulting term fall back in \llambda.
+An example of term $t$ in \maybebeta{} is the application $F \appsep x$.
+This term is problematic since the application node of
+its syntax tree cannot be used to justify a
+unification failure, i.e. by properly instantiating $F$ the term
+head constructor may become a $\lambda$, or a constant or stay an application.
+
+\begin{definition}[\notllambda]\label{def:notllambda}
+  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n \not\in \llambda \}$.
+\end{definition}
+
+\noindent
+These terms are problematic for the very same reason terms in \maybebeta are,
+but cannot be handled directly by the unification of the meta language, that
+is only required to handle terms in \llambda.
+
+% An example of $t$ in \notllambda{} is $F \appsep a$ for a constant $a$.
+% Note however tha
+% an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
+% that makes the resulting term fall back in \llambda.
+% \todo{said before}
 
 \newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
-\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
-  largest set $\\subterm{t}$ that can be obtained by the following rules.
-$$
-\begin{array}{l}
-t \in \subterm{t}\\
-t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
-t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
-\end{array}
-$$
-\end{definition}
-
-\noindent
-We write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.
+We write $\subterm{t}$ the set of sub-terms of $t$, and
+%\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
+%  largest set $\subterm{t}$ that can be obtained by the following rules.
+%$$
+%\begin{array}{l}
+%t \in \subterm{t}\\
+%t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
+%t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
+%\end{array}
+%$$
+%\end{definition}
+%\noindent
+we write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.
 
@@
 $$
-\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
+\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\notllambda{} ~\cup~ \maybeeta{} ~\cup~ \maybebeta{})
 $$
@@
 
-\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
+\begin{proposition}[\wellb{}-preservation of \hstep]\label{prop:nf}
 $\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
@@
 that could be erased by an $\eta$-contraction),
-nor put in $\sigma$ a flexible application outside \llambda{}
+nor put in $\sigma$ an application whose head is flexible
 (an application node that could be erased by a $\beta$-reduction).
@@
 since decompilation can introduce (actually restore) terms in
-\maybeeta or \maybebeta that were move out of the way
-(put in $\linkStore$) during compilation.
+\maybeeta, \maybebeta or \notllambda that were moved out of the way
+during compilation.
 
@@
 
-Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
-
-Our encoding of CIC may look ``semi shallow'' since we use the meta-language
-lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
-A fully shallow encoding unfortunately does not fit our use case, although
-it would make the running example work:
+\noindent
+One could ignore the similarity between \Uo{} and \Ue{} and ``just''
+describe the object language unification procedure in the meta language by
+crafting a \elpiIn{unif} routine and using it as follows in
+rule \ref{clause:r3}:
+
+\begin{elpicode}
+decision X :- unif X (all A x\ app [P, x]), finite A,
+  pi x\ decision (app [P, x]).
+\end{elpicode}
+
+\noindent
+This choice would underuse the logic programming engine provided by
+the meta language since, by removing any datum from the head of rules,
+indexing degenerates. Moreover the unification procedure \elpiIn{unif}
+programmed in the meta language is likely to be an order of magnitude
+slower than one that is built-in.
+
+% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
+Another possibility is to avoid having the application and abstraction nodes
+in the syntax tree, and use the ones of meta language, as in the following:
 
@@
 in CIC it is not always possible to adopt it since the type system
-of the meta language is too weak to accommodate terms with a variable arity,
-like the following example:
+of the meta language is too weak to accommodate for the one of the 
+object language. In CIC the lambda abstraction has to carry a type in order to
+make type checking decidable. Moreover CIC allows for functions with a
+variable arity, like the following example:
 
@@
 \noindent
-The second reason is the encoding for Coq is used for meta programming the
-system, hence it must accommodate the manipulation of terms that are now
-know in advance (not even defined in Coq) without using introspection
-primitives such as Prologs's \texttt{functor} and \texttt{arg}.
-
-In the literature we could find a few related encoding of CIC.
-TODO In~\cite{felty93lics} is related and make the
-discrepancy between the types of ML and CIC visible. In this case
-one needs 4 application nodes. Moreover the objective is an encoding
-of terms, proofs, not proof search. Also note the conv predicate,
-akin to the unif we rule out.
-
-TODO This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
-
-None of the encodings above provide a solution to our problem.
+The type system of the $\lambda$Prolog is too stringent to accept this terms.\\
+The second reason is that the CIC encoding provided by Elpi is used for meta
+programming (extending) the Coq system, hence it must accommodate the
+manipulation of terms that are now know in advance (not even defined in Coq)
+without using introspection primitives such as Prologs's \texttt{functor}
+and \texttt{arg}. In this sense constants have to live in an open world, like
+the \elpiIn{string} data type used in the examples so far.
+
+In the literature we could find related encoding of the Calculus of
+Constructions~\cite{felty93lics}. The goal of that work was to exhibit
+a logic program performing proof checking in CC and hence relate the
+proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
+programs) with the Calculus of Constructions. The encoding is hence tailored
+toward a different goal, and utilizes three relations to represent the
+equational theory of CC. Section 6 contains a discussion about the use of the
+unification procedure of the meta language in presence of non ground goals, but
+the authors do not aim at exploiting it to the degree we want.\todo{sucks}
 
@@
 As we mentioned in section~\ref{sec:nutshell} the compiler
-replaces terms in \maybeeta and \maybebeta with fresh variables
+replaces terms in \maybeeta and \notllambda with fresh variables
 linked to the problematic terms. Each class of problematic terms
@@
 \section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
-% \label{sec:compilation}
+\label{sec:simulation}
 
@@
 With respect to \cref{sec:problem-statement} the signature also allows
-for updates to the substitution.  The code below uses that possibility
+for updates to the substitution.
+
+\input{code/comp_base}
+
+\noindent
+The code above uses that possibility
 in order to allocate space for the variables, i.e. sets their memory
 address to \elpiIn{none} (a details not worth mentioning in the
-previous discussion).
-
-\input{code/comp_base}
-
-\noindent
-This preliminary version of \elpiIn{comp} recognizes \Fo{} variables
-applied to a (possibly empty) duplicate free list of names.
-Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
-is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
-(see below) grants this property.
+previous sections).
 
@@
 links more readable by pruning unused context entries.
+
+\subsubsection{Detection of \maybebeta}
+
+The following rule is inserted just before rule~\ref{rule:compapp}.
+
+\input{code/comp_base_beta}
+
+\noindent
+Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
+is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
+(see below) grants this property. Also note that this rule generates no links.
 
@@
 % }
-\section{Handling of \maybebeta}\label{sec:beta}
+\section{Handling of \notllambda}\label{sec:beta}
 
@@
 want the unification relation to be more robust so that it can work with terms
-in \maybebeta. In general, unification in \maybebeta admits more then one
+in \notllambda. In general, unification in \notllambda admits more then one
 solution and committing one of them in the substitution does not guarantee
@@
 It is the case that, given a list of unification problems, $\foUnifPb_1\dots
-\foUnifPb_n$ with $\foUnifPb_n$ in \maybebeta, the resolution of
+\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, the resolution of
 $\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
@@
 this unification problem, term compilation should capture the
-terms $t$ in \maybebeta and replace them with fresh variables $X$.
+terms $t$ in \notllambda and replace them with fresh variables $X$.
 The variables $X$ and the terms $t$ are linked through a \linkbeta.
@@
   If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
-  counterpart is still in \maybebeta,
+  counterpart is still in \notllambda,
   the original unification problem is not in \llambda
@@
   \appsep t_1\dots t_n} is created. The unification instantiates $Y$ to $t$,
-  making the \lhs of the link a rigid term, while \rhs is still in \maybebeta.
+  making the \lhs of the link a rigid term, while \rhs is still in \notllambda.
   The original problem is in fact outside \llambda.
@@
 
-\paragraph{Compilation}
-Detection of \maybebeta is quite simple to implement in the compiler, since it
+Detection of \notllambda is quite simple to implement in the compiler, since it
 is sufficient to detect applications with flexible head and argument that
@@
   unification done between terms in \llambda. This problem unifies iff
-  its corresponding \Fo{} problem unifies too. If $\sigma\hoUnifPb$ is in \maybebeta,
+  its corresponding \Fo{} problem unifies too. If $\sigma\hoUnifPb$ is in \notllambda,
   then, by \cref{lemma:keep-fidelity}, the unification fails, as per the
@@
 
-\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\maybebeta}}
+\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}
 
@@
 such as $\lambda$Prolog~\cite{lamProlog}, Abella~\cite{gacek2008abella}, which 
-delay the resolution of \maybebeta unification problems if the substitution
+delay the resolution of \notllambda unification problems if the substitution
 is not able to put them in \llambda.
@@
 
-In the example above, $\foUnifPb_1$ is in \maybebeta and the object language
+In the example above, $\foUnifPb_1$ is in \notllambda and the object language
 cannot solve it, and, by \cref{prop:fidelity}, the meta language neither.
 However, we can be more permissive, and relax \cref{lemma:keep-fidelity}. This
-modification is quite simple to manage: we are introducing a new \maybebeta
+modification is quite simple to manage: we are introducing a new \notllambda
 progress rule, say \progBetaNoLLWait, by which, if \lhs is rigid and \rhs is
@@
 
-A second strategy to deal with problem that are in \maybebeta is to make some
+A second strategy to deal with problem that are in \notllambda is to make some
 approximation. This is the case for example of the unification algorithm of Coq
@@
 forcing a choice (among the others) when the unification problem is in
-\maybebeta. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
+\notllambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
 argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
@@
 
-\section{Actual implementation in Elpi}
+\section{Actual implementation in Elpi}\label{sec:implementation}
 
@@
 
-In Elpi we don't have a main loop, we rely on the interpreter one.
-So links are constraints and progress are CHR rules.
-Constraints are suspended goals that are resumend when some unif variable
-is assigned.
+The main difference between the presentation in the previous sections and
+the actual implementation for Coq is that the main loop \hrun is replaced by
+the one of Prolog that chains calls to the unification procedure. In order
+implement the store of links we resort to Elpi's CLP engine and
+use constraints (suspended goals) to represent links, and constraint
+handling rules to implement progress operations involving more than one link.
+
+\todo{finish}
+about the progress of 1 link:
 
 \begin{elpicode}
-link-eta bla :- suspend-condition Holes, !, declare_constraint (link-eta bla) Holes.
-link-eta bla :- progress. % solves the goal, hence the constraint is not more in L.
+link-eta L R :- suspend-condition L R Holes, !,
+  declare_constraint (link-eta L R) Holes.
+link-eta L R :-
+  progress. % e.g. L = R.
 \end{elpicode}
 
-This matches with all progress rules but for the ones considering
-two links. For these we need to resort to CHR to manipulate L. For example
-deduplicate eta
+about the progress of 2 links:
 
 \begin{elpicode}
-constraints link-eta {
-  rule (N1 : G1 ?- link-eta (uvar X LX1) T1) % match
-    /  (N2 : G2 ?- link-eta (uvar X LX2) T2) % remove
+constraint link-eta {
+  rule (N1 ~$\triangleright$~ G1 ?- link-eta (uvar X LX1) T1) % match
+    /  (N2 ~$\triangleright$~ G2 ?- link-eta (uvar X LX2) T2) % remove
     |  (relocate LX1 LX2 T2 T2')             % condition
-   <=> (N1 : G1 ?- T1 = T2').                % new goal
+   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                % new goal
 }

### 2024-04-08 paper/main.tex
update tex

@@
 \usepackage{intcalc}
+\usepackage{nameref}
 
@@
 % section is given in a future section
-\begin{lemma}
+\begin{lemma}[\linkbeta with rigid \lhs]
   If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
@@
 
+% From \cref{lemma:keep-fidelity}, we can deduce the following corollary
+
+% \begin{corollary}
+%   Given a \linkbeta \linkbetaM{\Gamma}{X}{T}, if it exists a second link
+%   $l\in\linkStore$, then unification fails.   
+% \end{corollary}
+
 \subsection{Compilation and decompilation}
@@
 
-\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}}
-\newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}}
-\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}}
+\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}\xspace}
+\newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}\xspace}
+\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}\xspace}
 
@@
 \begin{definition}[\progBetaDedup]
-  Given two \linkbeta $l1$ and $l2$, if they share the same \lhs, unification
-  fails.
+  Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
+  share the same \lhs. In this case, the two \rhs are unified and a $l_2$ is
+  removed from \linkStore.
   \label{def:progBetaDedup}
@@
   \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
-  terminates.
-  \noteD{L'ho riscritto}
+  terminates. Finally \progBetaDedup also guarantees termination since it 
+  makes a unification $\mathbb{U}$ and if $\mathbb{U}$ fails, then 
+  \elpiIn{progress} terminates and if $\mathbb{U}$ succeeds, 
+  the recursive calls to \elpiIn{progress} have a specialized \linkbeta\dots
 
@@
 Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
+  \noteD{da finire}
 \end{proof}
@@
 
-\subsection{Relaxing \cref{lemma:keep-fidelity} : tobe renamed}
+\subsection{Relaxing lemma \nameref{lemma:keep-fidelity}}
 

### 2024-04-08 paper/main.tex
tex: fidelity linketa

@@
 
+\begin{lemma}[Fidelity with \linketa]
+  Let \hoUnifPb a unification problem and $\sigma$ a substitution. The introduction 
+  of \linketa guarantees \cref{prop:fidelity} 
+  \label{lemma:fidelity-maybeeta}
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  If 1) a \linketa is activated by \progressetaleft, then unification is done
+  between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
+  activation performs a unification which succeeds iff the original problem in
+  \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
+  unification is done between two \maybeeta terms, and again this unification
+  succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
+  \progressetaright, the unification, done between a variable and a term,
+  always succeeds, this is what we expect to guarantee fidelity, since we
+  are essentially removing a hole added by the compilation in the place of a
+  \maybeeta subterm. In all the cases fidelity is respected.
+\end{proof}
+
 \paragraph{Example of \progressetaleft}
@@
 
-\begin{lemma}
-  Given a \hoUnifPb and a substitution $\sigma$ then the resolution of
-  $\sigma\hoUnifPb$ guarantees \cref{prop:fidelity} 
+\begin{lemma}[Fidelity with \linkbeta]
+  The introduction of \linkbeta guarantees \cref{prop:fidelity} 
 \end{lemma}
@@
 \begin{proof}[Proof sketch]
+  Let \hoUnifPb a unification problem and $\sigma$ a substitution
+  such that $\hoUnifPb \in \notllambda$. 
   If $\sigma\hoUnifPb$ is in \llambda, then by
-  \cref{def:progBetaLL,def:progBetaRH}, then \linkbeta disappear and the
-  unification done between terms in \llambda. This problem unifies iff
-  its corresponding \Fo{} problem unifies too. If $\sigma\hoUnifPb$ is in \notllambda,
+  \cref{def:progBetaLL,def:progBetaRH}, the \linkbeta associated to the subterm of
+  \hoUnifPb have been solved and removed. The
+  unification is done between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
+  fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
   then, by \cref{lemma:keep-fidelity}, the unification fails, as per the

### 2024-04-08 paper/main.tex
rework intro

@@
 
-Specifying and implementing a logic or a proof system from scratch requires
-significant effort.
-Logical Frameworks and Higher Order Logic Programming Languages provide
-dedicated, high-level Meta Languages (ML) to facilitate this task in two
-key ways. The first and most well know one is that variable binding and
-substitution come for free is one uses the ML binders to represent the object
-language ones. The second one that comes to mind is unification, a cornerstone
-for proof construction and proof search. However reusing that brick may require
-some additional work that we carry out in this paper.
-
 Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
@@
 logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
-In the notable of Higher Order Logic~\cite{books/sp/NipkowPW02},
-the ML Isabelle is such a good fit that it implements an interactive proof
-system for the object logic, and not just a specification.
+The use of these meta languages facilitated this task in two
+key ways. The first and most well know one is that variable binding and
+substitution come for free. %, exploited in all works mentioned above.
+The second one is that these meta languages come equipped with some form
+of unification, a cornerstone in proof construction and proof search.
+
+% exploited only
+%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
+%the meta language Isabelle is such a good match for HOL that it could used to
+%implement an interactive proof system for the object logic, in addition to
+%it specification.\\
 The object logic we are interested in is Coq's~\cite{Coq-refman}
@@
 form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
-We want to use the Elpi~\cite{tassi:hal-01637063} meta programming language,
+In particular we want to leverage the Elpi~\cite{tassi:hal-01637063} meta programming language,
 a dialect of $\lambda$Prolog already used to extend
@@
 Type-class solvers are unification based proof search procedures
-reminiscent of Prolog: they back-chain lemmas taken
-from a designated database of ``type class instances''.
-For this reason we believe that $\lambda$Prolog, hence Elpi, is a good fit for
+reminiscent of Prolog that back-chain lemmas taken
+from a designated database of ``type class instances'', hence
+we can expect that Elpi is a good fit for
 implementing such as form of automation. In this paper we focus on one aspect of
-this work, namely \emph{how to reuse the unification of the ML Elpi in order to
-implement the one used by the type-class solver}.
+this work, namely \emph{how to reuse the higher order unification procedure
+of the meta language in order to implement a type-class solver for the object
+language}. As it turns out, re-using the unification of the meta language is
+not a trivial task.
 
@@
 
+\todo{cite isabelle's TC, that are baked in}
+
 \section{Preliminaries: \Fo{} and \Ho}

### 2024-04-08 paper/main.tex
move related works to the end
major reorganization of 2,3,4... still a wip

@@
 The root cause of the problems we sketched in this example
-is that the unification procedure \Ue of the meta language is not aware
-of the equational theory of the object logic, even if both theories
-include $\eta\beta$-conversion and admit most general
-unifiers for unification problems in the pattern-fragment.
-
-\paragraph{Contributions}
-In this paper we identify a minimal language \Fo{} in which the problems
-sketched in the introduction can be formally described.
-We detail an encoding of a logic program on \Fo{} to a strongly related
-logic program in \Ho (the language of the meta language) and we show that
-the unification procedure of the meta language \Ue{} can be effectively
-used to simulate a unification procedure \Uo for the object language that
-features $\eta\beta$-conversion in the pattern-fragment.
-
-\cref{sec:problem-statement} formally states the problem and gives the
-intuition behind our solution. \cref{sec:encodings} discusses alternative
-term encodings and related works. \cref{sec:lang-spec} introduces
-the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
-basic simulation of higher order logic programs.
-\cref{sec:eta,sec:invariant1} completes its equational theory
-with support for $\eta$-conversion. \cref{sec:beta} deals with the
-practical necessity of ``tolerating'' terms outside of the
-pattern-fragment and discusses how heuristic can be applied.
-Finally \cref{sec:implementation} discusses the implementation in Elpi.
-
-The $\lambda$Prolog code discussed in the paper can be accessed at the
-URL: \url{https://github.com/FissoreD/paper-ho}.
-
-\section{Problem statement and solution} %%%%%%%%%%%%%%%%%%%%%%
-\label{sec:problem-statement}
-
-\newcommand{\specunif}[3]{
-  #3_i \in \llambda \Rightarrow %
-    \exists \rho, %
-      \rho #3_1 #1 \rho #3_2  %
-        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho
-}
-
-\newcommand{\unifcorrect}[3]{
-  % \forall \rho #3_1 #3_2, %
-    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
-    #3_i \in \llambda \Rightarrow
-      #3_1 #2 #3_2 \mapsto \rho
-        \Rightarrow
-          \rho #3_1 #1 \rho #3_2  %
-}
-
-\newcommand{\unifcomplete}[3]{
-  % \forall #3_1 #3_2, %
-    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
-    #3_i \in \llambda \Rightarrow
-      % \forall \rho, %
-        \rho #3_1 #1 \rho #3_2  %
-          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
-}
-\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
-\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
-\newcommand{\notllambda}{\ensuremath{\overline{\llambda}}\xspace}
-%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}
-
-The equational theory of Coq's Dependent Type Theory is very rich. In
+is a subtle mismatch between the equational theories and
+unification procedures of the meta language and the object language.
+
+The equational theory of CIC is very rich. In
 addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
@@
 since the abstraction and application term constructors
-the two unification procedures deal with are different. For example:
+the two unification procedures deal with are different. For example:\\
 %
@@
 \vspace{4pt}
-
-To state precisely the problem we solve we need a \Fo{} representation
-of CIC terms and a \Ho one. We extend the equality over ground
-terms to open ones with the reflexivity of unification variables (that is
-unification variable X is equal to itself, not to another one such a Y).
-We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
-We call \Uo the unification procedure we want to implement and
-\Ue the one provided by the meta language.
-We assume that the unification of our meta
-language is correct and complete in \llambda.
 %
-%\begin{gather}
-%  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
-%  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
-%\end{gather}
-
+\paragraph{Contributions}
+In this paper we identify a minimal language \Fo{} in which the problems
+sketched in the introduction can be formally described.
+We detail an encoding of a logic program on \Fo{} to a strongly related
+logic program in \Ho (the language of the meta language) and we show that
+the unification procedure of the meta language \Ue{} can be effectively
+used to simulate a unification procedure \Uo for the object language that
+features $\eta\beta$-conversion in the pattern-fragment.
+
+\cref{sec:problem-statement} formally states the problem and gives the
+intuition behind our solution. \cref{sec:encodings} discusses alternative
+term encodings and related works. \cref{sec:lang-spec} introduces
+the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
+basic simulation of higher order logic programs.
+\cref{sec:eta,sec:invariant1} completes its equational theory
+with support for $\eta$-conversion. \cref{sec:beta} deals with the
+practical necessity of ``tolerating'' terms outside of the
+pattern-fragment and discusses how heuristic can be applied.
+Finally \cref{sec:implementation} discusses the implementation in Elpi.
+
+The $\lambda$Prolog code discussed in the paper can be accessed at the
+URL: \url{https://github.com/FissoreD/paper-ho}.
+
+\section{Problem statement and solution} %%%%%%%%%%%%%%%%%%%%%%
+\label{sec:problem-statement}
+
+\newcommand{\specunif}[3]{
+  #3_i \in \llambda \Rightarrow %
+    \exists \rho, %
+      \rho #3_1 #1 \rho #3_2  %
+        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho
+}
+
+\newcommand{\unifcorrect}[3]{
+  % \forall \rho #3_1 #3_2, %
+    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    #3_i \in \llambda \Rightarrow
+      #3_1 #2 #3_2 \mapsto \rho
+        \Rightarrow
+          \rho #3_1 #1 \rho #3_2  %
+}
+
+\newcommand{\unifcomplete}[3]{
+  % \forall #3_1 #3_2, %
+    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    #3_i \in \llambda \Rightarrow
+      % \forall \rho, %
+        \rho #3_1 #1 \rho #3_2  %
+          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
+}
+\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
+\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
+\newcommand{\notllambda}{\ensuremath{\overline{\llambda}}\xspace}
+%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}
+
+Even if we encountered the problem in CIC one does not need such a
+complex system in order to state it. The minimal setting we could find
+is one where a \Fo{} language (for first order) with a rich(er) equational
+theory is phrased in a \Ho{} meta language with a simpler one, and one
+needs to build the unification of \Fo{} on top of the one of \Ho{}.
+
+In order to reason about unification we provide a description of the
+\Fo{} and \Ho languages where unification variables
+are first class terms, i.e. they have a concrete syntax. We keep these languages
+minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
+in the example in Section~\ref{sec:intro} together with the type notation of
+terms carried by the \elpiIn{lam} constructor.
+%
+\setlength{\abovecaptionskip}{0pt}
+\setlength{\belowcaptionskip}{-13pt}
+
+\begin{figure}[H]
+  \begin{tabular}{ll}
+  \begin{minipage}{0.21\textwidth}
+   \inputrawelpicode{code/fo_tm}
+  \end{minipage}
+  &
+  \begin{minipage}{0.24\textwidth}
+   \inputrawelpicode{code/ho_tm}
+  \end{minipage}
+  \end{tabular}\vspace{4pt}
+  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
+  \label{code:common-terms}
+  \Description[code:common-terms]{code:common-terms}
+\end{figure}
+
+\noindent
+Unification variables (\elpiIn{fuva} term constructor)
+in \Fo{} have no explicit scope:
+the arguments of an higher order variable are given via the \elpiIn{fapp}
+constructor. For example the term \coqIn{P x} is represented as
+\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
+\elpiIn{x} is a bound variable.\\
+In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
+since unification variables come equipped with an explicit scope.
+\todo{use faddr for Fo}
+
+faddr and addr are memory addresses, the details are given in bla,
+here they are just unique identifiers for unif variables and their types
+do now allow to mistake one for the other.
+\todo{cleanup}
+
+When we write \Ho terms outside code blocks we follow the
+usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
+$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
+However we need to
+distinguish between the ``application'' of a unification variable
+to its scope and the application of a term to a list of arguments.
+We write the scope of unification variables in subscript
+while we use juxtaposition for regular application.
+Here a few examples:
+\vspace{4pt}
+
+\begin{tabular}{ll}
+  $f\appsep a$ &  \elpiIn{app[con "f", con "a"]}\\
+  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
+  $\lambda x.F_{x} \appsep a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
+  $\lambda x.F_{x} \appsep x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
+\end{tabular}
+\vspace{4pt}
+
+\noindent
+When it is clear from the context we shall use the same syntax for \Fo{} terms
+(although we never subscripts unification variables).
+
+\subsection{eq and unif}
+
+In order to specify the two unification procedures we need to
+define the equational theories of \Fo{} and \Ho{} and define
+what substitutions are.
+
+\paragraph{Term equality: \Eo vs. \Ee} We extend the equational theory
+over ground terms to the full languages by adding the reflexivity of
+unification variables (a variable is equal to itself).
+
+The first four rules are common to both equalities
+and just define the usual congruence over terms, and since
+we use an HOAS encoding they also capture $\alpha$-equivalence.
+In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
+
+\input{code/feq}
+
+\begin{elpicode}
+  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
+  con C ~\Ee~fcon C.
+  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
+  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
+  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
+\end{elpicode}
+
+\noindent
+The main point in showing these equality tests is to remark how weaker \Ee is,
+and to identify the four rules that need special treatment in
+the implementation of \Uo.
+
+For reference, \elpiIn{(beta T A R)} reduces away \elpiIn{lam} nodes in head
+position in \elpiIn{T} whenever the list \elpiIn{A} provides a corresponding
+argument.
+
+\input{code/beta_fo}
+
+\noindent
+The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
+variables).\footnote{Elpi provides it as a builtin, but one could implement it by
+systematically loading the hypothetical rule \elpiIn{name x} every time
+a nominal constant is postulated via \elpiInFN{pi x\ }}
+The choice of using n-ary application, rather than binary, is to make it
+easy to access the application's head. The price we pay is that substituting
+an application in the head of an application should be amended by
+``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
+\footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
+we do in this presentation and, as we explain later,
+no equivalent of \elpiIn{napp} is needed in \Ho.}
+Finally note that the cut operator is inessential, it could be
+removed at the cost of a verbose test on the head of \elpiIn{L}
+in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
+can be \elpiIn{fcon}, \elpiIn{flam} or a name.
+
+\paragraph{Substitution application: $\rho s$ and $\sigma t$}
+
+Applying the substitution corresponds to dereferencing a term with respect to
+the memory. To ease the comparison we split \Fo dereferencing into a
+\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
+to memory cells that are set with their values, and has a corresponding
+operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
+has no corresponding operation in \Ho. The reasons for this asymmetry is
+that an \elpiIn{fapp} node with a flexible head is always mapped
+to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
+preventing nested applications to materialize.
+
+\input{code/fderef}
+
+\noindent
+Applying the substitution in \Ho{} is very similar, with
+the caveat that assignments have to be moved to the
+current scope, i.e. renaming the \elpiIn{abs}-bound variables
+with the names in the scope of the unification variable occurrence.
+
+\input{code/deref}
+
+\noindent
+Note that move strongly relies on invariant \ref{inv:uvaarity}
+that we shall define later.\todo{fixup}
+
+\input{code/move}
+
+\paragraph{Term unification: \Uo vs. \Ue}
+
+In this paper we assume to have an implementation of \Ue that satisfies
+properties~\ref{prop:correct-ml} and~\ref{prop:complete-ml}. Although we provide an
+implementation in the appendix (that we used for testing purposes) we only
+describe its signature here. Elpi is expected to provide this brick, as well as
+any other implementation of $\lambda$Prolog.
+
+\input{code/ue_type}
+%
+% To state precisely the problem we solve we need a \Fo{} representation
+% of CIC terms and a \Ho one. We extend the equality over ground
+% terms to open ones with the reflexivity of unification variables (that is
+% unification variable X is equal to itself, not to another one such a Y).
+% We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
+% We call \Uo the unification procedure we want to implement and
+% \Ue the one provided by the meta language.
 We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
@@
 $\sigma$ for the \Ho ones.
+We assume that the unification of our meta
+language is correct and complete in \llambda.
+
+\begin{gather}
+  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
+  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
+\end{gather}
+
+\subsection{Logic program simulation}
 
@@
 \newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
-
-We illustrate a compilation $\C{s}{t}{m}{l}$ that
-maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
-a variable mapping $m$ and list of links $l$.
-The variable map connects unification variables in \Ho with variables
-in \Fo{} and is used to ``decompile'' the assignment,
-$\D{\sigma}{m}{l}{\rho}$.
-Links are an accessory piece of information whose
-description is deferred to \cref{sec:nutshell}.
 
@@
 
-\noindent We simulate each run in \Fo{} with a run in \Ho as follows.
+\noindent 
+\todo{later in section..}
+We illustrate a compilation $\C{s}{t}{m}{l}$ that
+maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
+a variable mapping $m$ and list of links $l$.
+The variable map connects unification variables in \Ho with variables
+in \Fo{} and is used to ``decompile'' the assignment,
+$\D{\sigma}{m}{l}{\rho}$.
+Links are an accessory piece of information whose
+description is deferred to \cref{sec:nutshell}.
+
+We simulate each run in \Fo{} with a run in \Ho as follows.
 %Note that $\sigma_0$ is the empty substitution.
@@
 
-We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
-%$\foUnifPb = \{ s_1, s_2 \}$ 
-as follows:
-%
-$$
-\begin{array}{l}
-s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
-\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
-\quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
-\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
-\end{array}
-$$
-
-\begin{proposition}[Properties of \Uo{} in \llambda]
-\begin{gather}
-  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
-  \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
-\end{gather}
-\end{proposition}
-
-\noindent
-In addition to correctness and completeness in \llambda, we claim that \hrun
+% We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
+% %$\foUnifPb = \{ s_1, s_2 \}$ 
+% as follows:
+% %
+% $$
+% \begin{array}{l}
+% s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
+% \quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
+% \quad    t_1 \Ue t_2 \mapsto \sigma' \land
+%     \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
+% \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
+% \end{array}
+% $$
+
+% \begin{proposition}[Properties of \Uo{} in \llambda]
+% \begin{gather}
+%   \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
+%   \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
+% \end{gather}
+% \end{proposition}
+
+\noindent
+%In addition to correctness and completeness in \llambda,
+We claim that \hrun
 handles terms outside \llambda in the following sense:
@@
 
-\subsection{The intuition in a nutshell}
+\subsection{The solution in a nutshell}
 \label{sec:nutshell}
@@
 and an accessory \emph{link} that represent a suspended unification problem
-$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
-contradict \Eo as it would otherwise do on ``problematic'' terms.
+$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
+it does not contradict \Eo as it would otherwise do on ``problematic'' terms.
 We now define ``problematic'' and ``well behaved'' more formally.
@@
 
-Note that proposition \ref{prop:nf} does not hold for \Uo{} as a whole
-since decompilation can introduce (actually restore) terms in
-\maybeeta, \maybebeta or \notllambda that were moved out of the way
-during compilation.
-
-\section{Other encodings and related work}\label{sec:encodings}
-
-\noindent
-One could ignore the similarity between \Uo{} and \Ue{} and ``just''
-describe the object language unification procedure in the meta language by
-crafting a \elpiIn{unif} routine and using it as follows in
-rule \ref{clause:r3}:
-
-\begin{elpicode}
-decision X :- unif X (all A x\ app [P, x]), finite A,
-  pi x\ decision (app [P, x]).
-\end{elpicode}
-
-\noindent
-This choice would underuse the logic programming engine provided by
-the meta language since, by removing any datum from the head of rules,
-indexing degenerates. Moreover the unification procedure \elpiIn{unif}
-programmed in the meta language is likely to be an order of magnitude
-slower than one that is built-in.
-
-% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
-Another possibility is to avoid having the application and abstraction nodes
-in the syntax tree, and use the ones of meta language, as in the following:
-
-\begin{elpicode}
-finite (fin N).
-decision (nfact N NF).
-decision (all A x\ P x) :- finite A, pi x\ decision (P x).
-\end{elpicode}
-
-\noindent
-There are two reasons for dismissing this encoding. The first one is that
-in CIC it is not always possible to adopt it since the type system
-of the meta language is too weak to accommodate for the one of the 
-object language. In CIC the lambda abstraction has to carry a type in order to
-make type checking decidable. Moreover CIC allows for functions with a
-variable arity, like the following example:
-
-\begin{coqcode}
-Fixpoint arr T n := if n is S m then T -> arr T m else T.
-Definition sum n : arr nat n := ...
-Check sum 2   7 8   : nat.
-Check sum 3   7 8 9 : nat.
-\end{coqcode}
-
-\noindent
-The type system of the $\lambda$Prolog is too stringent to accept this terms.\\
-The second reason is that the CIC encoding provided by Elpi is used for meta
-programming (extending) the Coq system, hence it must accommodate the
-manipulation of terms that are now know in advance (not even defined in Coq)
-without using introspection primitives such as Prologs's \texttt{functor}
-and \texttt{arg}. In this sense constants have to live in an open world, like
-the \elpiIn{string} data type used in the examples so far.
-
-In the literature we could find related encoding of the Calculus of
-Constructions~\cite{felty93lics}. The goal of that work was to exhibit
-a logic program performing proof checking in CC and hence relate the
-proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
-programs) with the Calculus of Constructions. The encoding is hence tailored
-toward a different goal, and utilizes three relations to represent the
-equational theory of CC. Section 6 contains a discussion about the use of the
-unification procedure of the meta language in presence of non ground goals, but
-the authors do not aim at exploiting it to the degree we want.\todo{sucks}
-
-\todo{cite isabelle's TC, that are baked in}
-
-\section{Preliminaries: \Fo{} and \Ho}
+% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
+% since decompilation can introduce (actually restore) terms in
+% \maybeeta, \maybebeta or \notllambda that were moved out of the way
+% during compilation.
+
+\section{Preliminaries for \Fo{} and \Ho}
 \label{sec:lang-spec}
-
-In order to reason about unification we provide a description of the
-\Fo{} and \Ho languages where unification variables
-are first class terms, i.e. they have a concrete syntax. We keep these languages
-minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
-in the example in Section~\ref{sec:intro} together with the type notation of
-terms carried by the \elpiIn{lam} constructor.
-%
-\setlength{\abovecaptionskip}{0pt}
-\setlength{\belowcaptionskip}{-13pt}
-
-\begin{figure}[H]
-  \begin{tabular}{ll}
-  \begin{minipage}{0.21\textwidth}
-   \inputrawelpicode{code/fo_tm}
-  \end{minipage}
-  &
-  \begin{minipage}{0.24\textwidth}
-   \inputrawelpicode{code/ho_tm}
-  \end{minipage}
-  \end{tabular}\vspace{4pt}
-  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
-  \label{code:common-terms}
-  \Description[code:common-terms]{code:common-terms}
-\end{figure}
-
-\noindent
-Unification variables (\elpiIn{fuva} term constructor)
-in \Fo{} have no explicit scope:
-the arguments of an higher order variable are given via the \elpiIn{fapp}
-constructor. For example the term \coqIn{P x} is represented as
-\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
-\elpiIn{x} is a bound variable.\\
-In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
-since unification variables come equipped with an explicit scope.
+\todo{maybe move here napp beta deref fderef move}
+
 We say that the unification variable occurrence \elpiIn{uva N L} is in
@@
 
+\noindent
+Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
+of the arguments of all occurrences of a unification variable and the
+number of abstractions in its assignment have to match. In turn
+this grants that \elpiIn{move} never fails.
+
+%\input{code/move}
+
 \subsection{Notational conventions}
-
-When we write \Ho terms outside code blocks we follow the
-usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
-$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
-However we need to
-distinguish between the ``application'' of a unification variable
-to its scope and the application of a term to a list of arguments.
-We write the scope of unification variables in subscript
-while we use juxtaposition for regular application.
-Here a few examples:
-\vspace{4pt}
-
-\begin{tabular}{ll}
-  $f\appsep a$ &  \elpiIn{app[con "f", con "a"]}\\
-  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
-  $\lambda x.F_{x} \appsep a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
-  $\lambda x.F_{x} \appsep x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
-\end{tabular}
-\vspace{4pt}
 
@@
 
-\noindent
-When it is clear from the context we shall use the same syntax for \Fo{} terms
-(although we never subscripts unification variables).
-
-\subsection{Equational theory and Unification}
-
-In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
-we need to equip \Fo{} and \Ho with term equality,
-substitution application and unification.
-
-\paragraph{Term equality: \Eo vs. \Ee} We extend the equational theory
-over ground terms to the full languages by adding the reflexivity of
-unification variables (a variable is equal to itself).
-
-The first four rules are common to both equalities
-and just define the usual congruence over terms, and since
-we use an HOAS encoding they also capture $\alpha$-equivalence.
-In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
-
-\input{code/feq}
-
-\begin{elpicode}
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-  con C ~\Ee~fcon C.
-  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
-  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
-  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
-\end{elpicode}
-
-\noindent
-The main point in showing these equality tests is to remark how weaker \Ee is,
-and to identify the four rules that need special treatment in
-the implementation of \Uo.
-
-For reference, \elpiIn{(beta T A R)} reduces away \elpiIn{lam} nodes in head
-position in \elpiIn{T} whenever the list \elpiIn{A} provides a corresponding
-argument.
-
-\input{code/beta_fo}
-
-\noindent
-The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
-variables).\footnote{Elpi provides it as a builtin, but one could implement it by
-systematically loading the hypothetical rule \elpiIn{name x} every time
-a nominal constant is postulated via \elpiInFN{pi x\ }}
-The choice of using n-ary application, rather than binary, is to make it
-easy to access the application's head. The price we pay is that substituting
-an application in the head of an application should be amended by
-``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
-\footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
-we do in this presentation and, as we explain later,
-no equivalent of \elpiIn{napp} is needed in \Ho.}
-Finally note that the cut operator is inessential, it could be
-removed at the cost of a verbose test on the head of \elpiIn{L}
-in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
-can be \elpiIn{fcon}, \elpiIn{flam} or a name.
-
-\paragraph{Substitution application: $\rho s$ and $\sigma t$}
-
-Applying the substitution corresponds to dereferencing a term with respect to
-the memory. To ease the comparison we split \Fo dereferencing into a
-\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
-to memory cells that are set with their values, and has a corresponding
-operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
-has no corresponding operation in \Ho. The reasons for this asymmetry is
-that an \elpiIn{fapp} node with a flexible head is always mapped
-to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
-preventing nested applications to materialize.
-
-\input{code/fderef}
-
-\noindent
-Applying the substitution in \Ho{} is very similar, with
-the caveat that assignments have to be moved to the
-current scope, i.e. renaming the \elpiIn{abs}-bound variables
-with the names in the scope of the unification variable occurrence.
-
-\input{code/deref}
-
-\noindent
-Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
-of the arguments of all occurrences of a unification variable and the
-number of abstractions in its assignment have to match. In turn
-this grants that \elpiIn{move} never fails.
-
-\input{code/move}
-
-\paragraph{Term unification: \Uo vs. \Ue}
-
-In this paper we assume to have an implementation of \Ue that satisfies
-properties~\ref{prop:correct-ml} and~\ref{prop:complete-ml}. Although we provide an
-implementation in the appendix (that we used for testing purposes) we only
-describe its signature here. Elpi is expected to provide this brick, as well as
-any other implementation of $\lambda$Prolog.
-
-\input{code/ue_type}
-
-\noindent
-The only detail worth discussing is the fact that the procedure updates a
-substitution, rather than just crafting one as presented in
-section~\ref{sec:problem-statement}. The reason is that the algorithm folds
-over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
+\todo{notation for problem compilation}
 
@@
 
-\subsubsection{Detection of \maybebeta}
+\subsubsection{Compilation of terms in \maybebeta}
 
@@
 (see below) grants this property. Also note that this rule generates no links.
+
+\noindent
+The only detail worth discussing is the fact that the procedure updates a
+substitution, rather than just crafting one as presented in
+section~\ref{sec:problem-statement}. The reason is that the algorithm folds
+over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
 
@@
 
+\section{Other encodings and related work}\label{sec:encodings}
+
+\noindent
+One could ignore the similarity between \Uo{} and \Ue{} and ``just''
+describe the object language unification procedure in the meta language by
+crafting a \elpiIn{unif} routine and using it as follows in
+rule \ref{clause:r3}:
+
+\begin{elpicode}
+decision X :- unif X (all A x\ app [P, x]), finite A,
+  pi x\ decision (app [P, x]).
+\end{elpicode}
+
+\noindent
+This choice would underuse the logic programming engine provided by
+the meta language since, by removing any datum from the head of rules,
+indexing degenerates. Moreover the unification procedure \elpiIn{unif}
+programmed in the meta language is likely to be an order of magnitude
+slower than one that is built-in.
+
+% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
+Another possibility is to avoid having the application and abstraction nodes
+in the syntax tree, and use the ones of meta language, as in the following:
+
+\begin{elpicode}
+finite (fin N).
+decision (nfact N NF).
+decision (all A x\ P x) :- finite A, pi x\ decision (P x).
+\end{elpicode}
+
+\noindent
+There are two reasons for dismissing this encoding. The first one is that
+in CIC it is not always possible to adopt it since the type system
+of the meta language is too weak to accommodate for the one of the 
+object language. In CIC the lambda abstraction has to carry a type in order to
+make type checking decidable. Moreover CIC allows for functions with a
+variable arity, like the following example:
+
+\begin{coqcode}
+Fixpoint arr T n := if n is S m then T -> arr T m else T.
+Definition sum n : arr nat n := ...
+Check sum 2   7 8   : nat.
+Check sum 3   7 8 9 : nat.
+\end{coqcode}
+
+\noindent
+The type system of the $\lambda$Prolog is too stringent to accept this terms.\\
+The second reason is that the CIC encoding provided by Elpi is used for meta
+programming (extending) the Coq system, hence it must accommodate the
+manipulation of terms that are now know in advance (not even defined in Coq)
+without using introspection primitives such as Prologs's \texttt{functor}
+and \texttt{arg}. In this sense constants have to live in an open world, like
+the \elpiIn{string} data type used in the examples so far.
+
+In the literature we could find related encoding of the Calculus of
+Constructions~\cite{felty93lics}. The goal of that work was to exhibit
+a logic program performing proof checking in CC and hence relate the
+proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
+programs) with the Calculus of Constructions. The encoding is hence tailored
+toward a different goal, and utilizes three relations to represent the
+equational theory of CC. Section 6 contains a discussion about the use of the
+unification procedure of the meta language in presence of non ground goals, but
+the authors do not aim at exploiting it to the degree we want.\todo{sucks}
+
+\todo{cite isabelle's TC, that are baked in}
+
 \section{Conclusion}

### 2024-04-08 paper/main.tex
headers

@@
 
+\subsection{Preliminaries: \Fo{} and \Ho{}}
+
 In order to reason about unification we provide a description of the
@@
 
+\paragraph{Notational conventions}
+
 When we write \Ho terms outside code blocks we follow the
@@
 
-\subsection{eq and unif}
+\subsection{Equational theories an unification}
 
@@
 
+\todo{move to next section}
 For reference, \elpiIn{(beta T A R)} reduces away \elpiIn{lam} nodes in head
@@
 
-\subsection{Logic program simulation}
+\subsection{The problem: Logic Program Simulation}
 
@@
 
-\subsection{The solution in a nutshell}
+\subsection{The solution (in a nutshell)}
 \label{sec:nutshell}
@@
 
-\section{Preliminaries for \Fo{} and \Ho}
+\section{Ground work for \Fo{}, \Ho and the compiler}
 \label{sec:lang-spec}

### 2024-04-09 paper/main.tex
rephrase of a sentence

--- 2024-04-08 paper/main.tex
@@
 
-\newcommand{\occursrigid}{\emph{occurs-rigidly}}
-\newcommand{\occurrigid}{\emph{occur-rigidly}}
+\newcommand{\occursrigid}{\emph{occurs-rigidly}\xspace}
+\newcommand{\occurrigid}{\emph{occur-rigidly}\xspace}
 \begin{definition}[\occursrigid]\label{def:occrigid}
@@
 
-\noteD{Not too clear?}
-In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
-outside of the scope of unification variables since an instantiation
-is allowed to discard $x$ from the scope of the unification variable.
-Note that $\eta$-contraction cannot make $x$ disappear, since the
-variables being erased by $\eta$-contraction are locally bound inside $t$.
+In other words $x$ \occursrigid in $t$ if it occurs in $t$
+outside of the scope of a unification variable $X$, otherwise an instantiation
+of $X$ can make $x$ disappears from $t$.
+Moreover, note that $\eta$-contracting $t$ cannot make $x$ disappear, since
+$x$ is not a locally bound variable inside $t$.
 

### 2024-04-09 paper/main.tex
rephrase of some sentences

@@
 \paragraph{Decompilation}
-Decompilation of \linketa is performed by adding new rules to the
-\elpiIn{commit-link} predicate. In particular, given \linketaM{\Gamma}{X}{t}, we
-can note that this unification never fails, since $X$ is a flexible term and no
-other \linketa has $X$ has \lhs (by \cref{def:progressetadedup}). The link is
-remove from \linkStore and \elpiIn{commit-links} terminates.
+Decompilation of the remaining \linketa (i.e. the \linketa that have been
+activated) is performed by iterating over them and unifying \lhs and \rhs. Note
+that this unification never fails, since \lhs is a flexible term not appearing
+in any other \linketa (by \cref{def:progressetadedup}).
 
@@
 
-\noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
-forse si può compattare in una prova più piccola?}
+% \noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
+% forse si può compattare in una prova più piccola?}
 \begin{lemma}
@@
 \begin{proof}[Proof sketch]
-  Let $\sigma$ be the substitution, such that $\wellb(\sigma)$. $\lhs \in
-  \sigma$, therefore $\wellb(\lhs)$. By \cref{def:progressetaleft}, if 1) \lhs is
-  a name, a constant of an application, then, \lhs is unified with the
-  $\eta$-reduced term $t$ obtain from \rhs. By \cref{cor:rhs-eta-onelamb}, \rhs
-  has one lambda, therefore $\wellb(t)$. Otherwise, 2) \lhs has \elpiIn{lam} as
-  functor, \rhs should not be an $\eta$-expansion ans, so, $\wellb(\rhs)$. In
-  both cases, unification is performed between terms in \wellb.
+  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
+  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
+  \cref{def:progressetaleft}, if 1) \lhs is a name, a constant of an
+  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
+  \cref{inv:link-eta-right}, $\rhs = \lambda x. t$, therefore $\wellb(t)$.
+  Otherwise, 2) \lhs has \elpiIn{lam} as functor, \rhs should not be an
+  $\eta$-expansion, so, $\wellb(\rhs)$. In both cases, unification is
+  performed between terms in \wellb.
 \end{proof}

### 2024-04-09 paper/main.tex
correction of a proof

--- 2024-04-09 paper/main.tex
@@
 \begin{proof}
-  Trivial, since the unification is done between unification variables, which
-  are by definition in \wellb.
+  The unification is done between the \rhs of two
+  \linketa. Both \rhs has the shape $\lambda x.t$, with $\wellb(t)$, therefore,
+  the unification is done between well-behaved terms.
 \end{proof}
@@
 \begin{lemma}
-  \Cref{prop:nf} holds, i.e., given a substitution $\sigma$ and a \linketa $l$,
-  after the activation of $l$, $\wellb(\sigma)$ holds.
+  The introduction of \linketa guarantees \fullref{prop:nf}
+  \label{lemma:unif-wellb}
 \end{lemma}

### 2024-04-09 paper/main.tex
rework of a proof

@@
 
-\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}}
+\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}\xspace}
 \begin{definition}[\progressetadedup]\label{def:progressetadedup}
@@
   By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
-  unification performed by the activation of a \linketa is performed between
+  unification performed by the activation of a \linketa is done between
   terms in \wellb, therefore, the substitution remains \wellb.
@@
 
-\noteD[inline]{Proove simulation fidelity, dicendo che \progressetaright è inutile}
-
 \begin{lemma}[Fidelity with \linketa]
-  Let \hoUnifPb a unification problem and $\sigma$ a substitution. The introduction 
-  of \linketa guarantees \cref{prop:fidelity} 
+  The introduction 
+  of \linketa guarantees \fullref{prop:fidelity} 
   \label{lemma:fidelity-maybeeta}
@@
 \begin{proof}[Proof sketch]
-  If 1) a \linketa is activated by \progressetaleft, then unification is done
-  between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
-  activation performs a unification which succeeds iff the original problem in
-  \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
-  unification is done between two \maybeeta terms, and again this unification
-  succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
-  \progressetaright, the unification, done between a variable and a term,
-  always succeeds, this is what we expect to guarantee fidelity, since we
-  are essentially removing a hole added by the compilation in the place of a
-  \maybeeta subterm. In all the cases fidelity is respected.
+  \progressetaleft and \progressetadedup activate a \linketa when, in the
+  original unification problem, a \maybeeta term is unified with respectively a
+  well-behaved term or another \maybeeta term. In both cases, the links trigger
+  a unification which succeeds iff the same unification in \Fo{} succeeds,
+  guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
+  this progression refines a variable to a rigid term and plays no role in 
+  \cref{prop:fidelity}.
+  % If 1) a \linketa is activated by \progressetaleft, then unification is done
+  % between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
+  % activation performs a unification which succeeds iff the original problem in
+  % \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
+  % unification is done between two \maybeeta terms, and again this unification
+  % succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
+  % \progressetaright, the unification, done between a variable and a term,
+  % always succeeds, this is what we expect to guarantee fidelity, since we
+  % are essentially removing a hole added by the compilation in the place of a
+  % \maybeeta subterm. In all the cases fidelity is respected.
 \end{proof}

### 2024-04-09 paper/main.tex
some rework

@@
 The example at the beginning of \cref{sec:eta}, once
-$\sigma = \{~ A \mapsto f ~\}$, triggers this rule since the link
+$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
 becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
@@
 This paragraph needs fixing.}
-\noteD[inline]{I've rewritten it, it is clearer?}
-Until now, we have only dealt we unification of terms in \llambda. However, we
-want the unification relation to be more robust so that it can work with terms
-in \notllambda. In general, unification in \notllambda admits more then one
+% \noteD[inline]{I've rewritten it, it is clearer?}
+In general, unification between \notllambda terms admits more then one
 solution and committing one of them in the substitution does not guarantee
-prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
-problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
+\cref{prop:complete}. For instance, $X \appsep a \Uo a$ 
+admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
 and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
@@
 
-It is the case that, given a list of unification problems, $\foUnifPb_1\dots
+It is the case, however, that, given a list of unification problems, $\foUnifPb_1\dots
 \foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, the resolution of
@@
 In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
-$\foUnifPb_2$, can be solved in \llambda.
-\noteE[inline]{it is even a ground term, there is no unification left to perform actually}
-\noteD[inline]{i don't understand the note}
+$\foUnifPb_2$ can be solved in \llambda.
+% \noteE[inline]{it is even a ground term, there is no unification left to perform actually}
+% \noteD[inline]{i don't understand the note}
 On the other hand, we see that, 
@@
 \mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
-non-unifiable problem $(\lambda x. B) \appsep a \Ue a$.
+non-unifiable problem $(\lambda x. B) \appsep a \nUe a$.
 
@@
 this unification problem, term compilation should capture the
-terms $t$ in \notllambda and replace them with fresh variables $X$.
-The variables $X$ and the terms $t$ are linked through a \linkbeta.
+terms in \notllambda and replace them with fresh variables.
+This replacement should produce links that we call \linkbeta.
 
@@
 % ... where S = [s1 .. sn] e L = [l1 .. lm]}
-\noteD[inline]{Is it clearer?}
 
@@
   unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
-  terms. This is equivalent to \elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
+  terms. This is equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
   s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
@@
   is replaced with a variable $Y$, and the \linkbeta \linkbetaM{\Gamma}{Y}{X
-  \appsep t_1\dots t_n} is created. The unification instantiates $Y$ to $t$,
-  making the \lhs of the link a rigid term, while \rhs is still in \notllambda.
-  The original problem is in fact outside \llambda.
+  \appsep t_1\dots t_n} is created. If $Y$ is instantiated to the rigid term $t$,
+  and \rhs is still in \notllambda, then the original \Fo{} unification problem
+  is between two terms in \notllambda. Unification fails.
 \end{proof}
@@
 is sufficient to detect applications with flexible head and argument that
-are not in \llambda.
+are not in \llambda. The following rule for \notllambda compilation is inserted 
+just before rule~\ref{rule:compapp}.
 
@@
 
-\noteD{Paragraph for decompile in case of \maybebeta}
-The \elpiIn{commit-link} predicate can be extended to add heuristics if during
-the decompilation phase \linkbeta remain. For example, the same approximation
-explained above can be delayed and applied only if the terms in \maybebeta never
-falls in \llambda after the execution of all the unification problems. We want
-to point out, that we call this \emph{approximation}, since we are
-making a choice among all the possible unifiers and therefore, we can pick
-the wrong one.
+% \noteD{Paragraph for decompile in case of \maybebeta}
+Decompilation of \linkbeta is possible by extending \elpiIn{commit-link} with
+new heuristics.
 

### 2024-04-09 paper/main.tex
more work

--- 2024-04-09 paper/main.tex
@@
 \vspace{4pt}
-
-To state precisely the problem we solve we need a \Fo{} representation
-of CIC terms and a \Ho one. We extend the equality over ground
-terms to open ones with the reflexivity of unification variables (that is
-unification variable X is equal to itself, not to another one such a Y).
-We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
-We call \Uo the unification procedure we want to implement and
-\Ue the one provided by the meta language.
-We assume that the unification of our meta
-language is correct and complete in \llambda.
 %
-%\begin{gather}
-%  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
-%  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
-%\end{gather}
-
-We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
-with substitution $\sigma$; we write $\sigma t$ for the application of
-the substitution to $t$, and $\sigma X = \{ \sigma t ~|~ t \in X\}$ when
-$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
-general than $\sigma'$ and we use set union to denote the concatenation of
-two substitutions. We shall use $\rho$ for \Fo{} substitutions, and
-$\sigma$ for the \Ho ones.
-
+\paragraph{Contributions}
+In this paper we identify a minimal language \Fo{} in which the problems
+sketched in the introduction can be formally described.
+We detail an encoding of a logic program on \Fo{} to a strongly related
+logic program in \Ho (the language of the meta language) and we show that
+the unification procedure of the meta language \Ue{} can be effectively
+used to simulate a unification procedure \Uo for the object language that
+features $\eta\beta$-conversion in the pattern-fragment.
+
+\cref{sec:problem-statement} formally states the problem and gives the
+intuition behind our solution. \cref{sec:encodings} discusses alternative
+term encodings and related works. \cref{sec:grounwork} introduces
+the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
+basic simulation of higher order logic programs.
+\cref{sec:eta,sec:invariant1} completes its equational theory
+with support for $\eta$-conversion. \cref{sec:beta} deals with the
+practical necessity of ``tolerating'' terms outside of the
+pattern-fragment and discusses how heuristic can be applied.
+Finally \cref{sec:implementation} discusses the implementation in Elpi.
+
+The $\lambda$Prolog code discussed in the paper can be accessed at the
+URL: \url{https://github.com/FissoreD/paper-ho}.
+
+\section{Problem statement and solution} %%%%%%%%%%%%%%%%%%%%%%
+\label{sec:problem-statement}
+
+\newcommand{\specunif}[3]{
+  \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    \exists \rho, %
+      \rho #3_1 #1 \rho #3_2  %
+        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho
+}
+
+\newcommand{\unifcorrect}[3]{
+  % \forall \rho #3_1 #3_2, %
+    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
+      #3_1 #2 #3_2 \mapsto \rho
+        \Rightarrow
+          \rho #3_1 #1 \rho #3_2  %
+}
+
+\newcommand{\unifcomplete}[3]{
+  % \forall #3_1 #3_2, %
+    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
+    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
+      % \forall \rho, %
+        \rho #3_1 #1 \rho #3_2  %
+          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
+}
+\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
+\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
+\newcommand{\notllambda}{\ensuremath{\Diamond\llambda}\xspace}
+%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}
+
+Even if we encountered the problem working on CIC we devise
+a minimal setting to ease its study. In this setting we have
+a \Fo{} language (for first order) with a rich(er) equational
+theory and a \Ho{} meta language with a simpler one, and we reuse
+the unification procedure of \Ho{} in order to implement one for \Fo.
+
+\subsection{Preliminaries: \Fo{} and \Ho{}}
+
+In order to reason about unification we provide a description of the
+\Fo{} and \Ho languages where unification variables
+are first class terms, i.e. they have a concrete syntax.
+% We keep these languages
+%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
+%in the example in Section~\ref{sec:intro} together with the type notation of
+%terms carried by the \elpiIn{lam} constructor.
+%
+{
+\setlength{\abovecaptionskip}{0pt}
+\setlength{\belowcaptionskip}{-13pt}
+
+\begin{figure}[H]
+  \begin{tabular}{ll}
+  \begin{minipage}{0.21\textwidth}
+   \inputrawelpicode{code/fo_tm}
+  \end{minipage}
+  &
+  \begin{minipage}{0.24\textwidth}
+   \inputrawelpicode{code/ho_tm}
+  \end{minipage}
+  \end{tabular}\vspace{4pt}
+  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
+  \label{code:common-terms}
+  \Description[code:common-terms]{code:common-terms}
+\end{figure}
+}
+\noindent
+Unification variables (\elpiIn{fuva} term constructor)
+in \Fo{} have no explicit scope:
+the arguments of an higher order variable are given via the \elpiIn{fapp}
+constructor. For example the term <<\coqIn{P x}>> is represented as
+<<\elpiIn{fapp[fuva N, x]}>>, where \elpiIn{N} is a memory address and
+\elpiIn{x} is a bound variable.\\
+In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
+since unification variables are higher order and come equipped with an
+explicit scope.
+%\todo{use faddr for Fo}
+%
+%faddr and addr are memory addresses, the details are given in bla,
+%here they are just unique identifiers for unif variables and their types
+%do now allow to mistake one for the other.
+%\todo{cleanup}
+
+\paragraph{Notational conventions}
+
+When we write \Ho terms outside code blocks we follow the
+usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
+$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
+However we need to
+distinguish between the ``application'' of a unification variable
+to its scope and the application of a term to a list of arguments.
+We write the scope of unification variables in subscript
+while we use juxtaposition for regular application.
+Here a few examples:\\
+\vspace{4pt}
+{
+\setlength{\tabcolsep}{1em}
+\begin{tabular}{ll}
+  $f\appsep a$                  & \elpiIn{app [con "f", con "a"]}\\
+  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
+  $\lambda x.F_{x} \appsep a$   & \elpiIn{lam x\ app [uva F [x], con "a"]} \\
+  $\lambda x.F_{x} \appsep x$   & \elpiIn{lam x\ app [uva F [x], x]} \\
+\end{tabular}
+}
+\vspace{4pt}
+
+\noindent
+When it is clear from the context we shall use the same syntax for \Fo{} terms
+(although we never subscripts unification variables).
+
+We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
+terms in \Ho{}.
+
+\subsection{Equational theories an unification}
+
+In order to specify unification we need to
+define the equational theory and
+substitution (unification-variable assignment).
+
+\subsubsection{Term equality: \Eo and \Ee}
+For both languages we extend the equational theory
+over ground terms to the full language by adding the reflexivity of
+unification variables (a variable is equal to itself).
+
+The first four rules are common to both equalities
+and just define the usual congruence over terms. Since
+we use an HOAS encoding they also capture $\alpha$-equivalence.
+In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
+
+\input{code/feq}
+
+\begin{elpicode}
+  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
+  con C ~\Ee~fcon C.
+  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
+  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
+  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
+\end{elpicode}
+
+\noindent
+The main point in showing these equality tests is to remark how
+weaker \Ee is, and to identify the four rules that need special
+treatment in the implementation of \Uo.
+We defer the description of \elpiIn{beta} to \cref{sec:grounwork}, for now
+it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
+weak head normal form of <<\elpiIn{app[F|L]}>>.
+
+\paragraph{Substitution: $\rho s$ and $\sigma t$}
+
+We write $\sigma = \{~ X \mapsto t ~\}$ the substitution that assigns
+the term $t$ to the variable $X$.
+We write $\sigma t$ for the application of
+the substitution to a term $t$, and $\sigma X = \{~ \sigma t ~|~ t \in X ~\}$ when
+$X$ is a set of terms.
+We write $\sigma \subseteq \sigma'$ when $\sigma$ is more
+general than $\sigma'$.
+The domain of a substitution is the set of unification variables for which
+it provides an assignment.
+We write $\sigma \cup \sigma'$ set union to denote the concatenation of
+two substitutions whose domains are disjoint.
+We shall use $\rho$ for \Fo{} substitutions,
+and $\sigma$ for the \Ho ones.
+\todo{can be made a little more formal}
+For brevity, in this section we consider
+the substitution for \Fo{} and \Ho{} identical.
+We defer to \cref{sec:grounwork} a more precise description
+pointing out theirs differences.
+
+\paragraph{Term unification: \Uo vs. \Ue}
+
+Although we provide an implementation of \Ue in the supplementary material
+(that we used for testing purposes) we only describe its signature here.
+
+\input{code/ue_type}
+
+The meta language of choice is expected to provide 
+an implementation of \Ue that satisfies
+the following properties:
+%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
+\begin{gather}
+  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
+  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
+\end{gather}
+
+%
+% To state precisely the problem we solve we need a \Fo{} representation
+% of CIC terms and a \Ho one. We extend the equality over ground
+% terms to open ones with the reflexivity of unification variables (that is
+% unification variable X is equal to itself, not to another one such a Y).
+% We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
+% We call \Uo the unification procedure we want to implement and
+% \Ue the one provided by the meta language.
+
+We write 
+$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
+$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
+We write $t_1 \Ue t_2 \mapsto \sigma$ when
+the initial substitution is empty.
+Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
+the domains of $\sigma$ and $\sigma'$ are disjoint.
+
+Although we provide an implementation of \Uo{} in \cref{sec:founif},
+our real goal is the simulation of an entire logic program.
+
+\subsection{The problem: Logic Program Simulation}
+
+We represent a logic program \emph{run} in \Fo{} as
+a list \emph{steps} $p$ of length $\mathcal{N}$.
+At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
+$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
+\footnote{If the same rule is used multiple times in a run we
+just consider as many copies as needed of the terms composing the
+rules, with fresh unification variables each time.}
+The composition of these steps starting from the
+empty substitution $\rho_0$ produces the final
+substitution $\rho_\mathcal{N}$, that is the result of the
+logic-program execution.
+%
 \newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
 \newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
-
-We illustrate a compilation $\C{s}{t}{m}{l}$ that
-maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
-a variable mapping $m$ and list of links $l$.
-The variable map connects unification variables in \Ho with variables
-in \Fo{} and is used to ``decompile'' the assignment,
-$\D{\sigma}{m}{l}{\rho}$.
-Links are an accessory piece of information whose
-description is deferred to \cref{sec:nutshell}.
-
-We represent a logic program \emph{run} in \Fo{} as
-a list \emph{steps} $p$ of length $\mathcal{N}$.
-Each step $p$ represents 
-a unification problem between two terms $\foUnifPb_{p_l}$ and
-$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
-The composition of these steps starting from the
-empty substitution $\rho_0$ produces the final
-substitution $\rho_\mathcal{N}$.
-\footnote{If the same rule is used multiple time in a run we
-just consider as many copies as needed of the terms composing the
-rules, with fresh unification variables each time}
-%
 \newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
@@
 
-\noindent We simulate each run in \Fo{} with a run in \Ho as follows.
+\noindent 
+In order to simulate a \Fo{} logic program in \Ho{} we compile
+each term $s \in \Fo{}$ into a term $t \in \Ho{}$. We write this
+step $\C{s}{t}{m}{l}$. The implementation of the compiler
+is detailed in \cref{sec:compilation,sec:eta,sec:beta}, here we just point
+out that it additionally a variable mapping $m$ and list of links $l$.
+The variable map connects unification variables in \Ho with variables
+in \Fo{} and is used to ``decompile'' the assignment,
+$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
+description is deferred to \cref{sec:nutshell}.
+
+We simulate each run in \Fo{} with a run in \Ho as follows.
 %Note that $\sigma_0$ is the empty substitution.
@@
 \noindent
-Here \hstep{} is made of two sub-steps: a call to the meta language
-unification and a check for \progress{} on the set of links.
+By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
+for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
+for the set $\{ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} \}$.\\
+\hstep{} is made of two sub-steps: a call to the meta language
+unification and a check for \progress{} on the set of links, that intuitively
+will compensate for the weaker equational theory honoured by \Ue.
+\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
+finally decompiles the solution.
 We claim:
@@
 
-We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
-%$\foUnifPb = \{ s_1, s_2 \}$ 
-as follows:
-%
-$$
-\begin{array}{l}
-s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
-\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
-\quad    t_1 \Ue t_2 \mapsto \sigma' \land
-    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
-\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
-\end{array}
-$$
-
-\begin{proposition}[Properties of \Uo{} in \llambda]
-\begin{gather}
-  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
-  \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
-\end{gather}
-\end{proposition}
-
-\noindent
-In addition to correctness and completeness in \llambda, we claim that \hrun
+% We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
+% %$\foUnifPb = \{ s_1, s_2 \}$ 
+% as follows:
+% %
+% $$
+% \begin{array}{l}
+% s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
+% \quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
+% \quad    t_1 \Ue t_2 \mapsto \sigma' \land
+%     \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
+% \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
+% \end{array}
+% $$
+
+% \begin{proposition}[Properties of \Uo{} in \llambda]
+% \begin{gather}
+%   \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
+%   \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
+% \end{gather}
+% \end{proposition}
+
+%In addition to correctness and completeness in \llambda,
+We also claim that \hrun
 handles terms outside \llambda in the following sense:
@@
   t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
-  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow
-  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \mapsto \sigma \label{prop:step-nllam}
+  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow \label{prop:step-nllam}\\
+  \exists \rho,\stepF{\foUnifPb}{p}{\rho}{\rho'} \Leftrightarrow
+  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \nonumber
 \end{gather}
@@
 Property \ref{prop:compilation-nllam} states that two terms for which
-there is a unifier (given by an oracle, not necessarily a most general one),
-then the compiler generates two terms that unify in \Ho. In other words
-the compiler moves out of the way all problematic terms.
-
+there is a unifier (given by an oracle and not necessarily a most general one),
+then the compiler generates two terms that unify in \Ho.\\
+%In other words the compiler moves out of the way all problematic terms.
 Property \ref{prop:step-nllam} says that if the two terms involved in a step
 re-enter \llambda, then \hstep succeeds. This is a typical example
-in which the order of the unification problems in a logic program run
-do matter. The simplest example is the sequence $F = \lambda x.a$ and
-$F \appsep a = a$: the second problem is not in \llambda and has two unifiers,
-namely $F = \lambda x.x$ and $F = \lambda x.a$. The first problem picks one
-of the two, making the second problem re-enter \llambda.
-
-\subsection{The intuition in a nutshell}
+in which the order of the unification problems in a logic-program run
+does matter. The simplest example is the sequence $F \UnifRel \lambda x.a$ and
+$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
+unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
+$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
+making the second problem re-enter \llambda. In other words
+\cref{prop:fidelity} \todo{...}
+
+\subsection{The solution (in a nutshell)}
 \label{sec:nutshell}
@@
 ``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
-and an accessory \emph{link} that represent a suspended unification problem
-$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
-contradict \Eo as it would otherwise do on ``problematic'' terms.
-We now define ``problematic'' and ``well behaved'' more formally.
+and an accessory \emph{link} that represents a suspended unification problem
+$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
+it does not contradict \Eo as it would otherwise do on the
+``problematic'' sub-terms. We now define ``problematic'' and ``well behaved''
+more formally.
 
@@
 \begin{definition}[\notllambda]\label{def:notllambda}
-  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n \not\in \llambda \}$.
+  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
 \end{definition}
@@
 
-Note that proposition \ref{prop:nf} does not hold for \Uo{} as a whole
-since decompilation can introduce (actually restore) terms in
-\maybeeta, \maybebeta or \notllambda that were moved out of the way
-during compilation.
-
-\section{Other encodings and related work}\label{sec:encodings}
-
-\noindent
-One could ignore the similarity between \Uo{} and \Ue{} and ``just''
-describe the object language unification procedure in the meta language by
-crafting a \elpiIn{unif} routine and using it as follows in
-rule \ref{clause:r3}:
-
-\begin{elpicode}
-decision X :- unif X (all A x\ app [P, x]), finite A,
-  pi x\ decision (app [P, x]).
-\end{elpicode}
-
-\noindent
-This choice would underuse the logic programming engine provided by
-the meta language since, by removing any datum from the head of rules,
-indexing degenerates. Moreover the unification procedure \elpiIn{unif}
-programmed in the meta language is likely to be an order of magnitude
-slower than one that is built-in.
-
-% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
-Another possibility is to avoid having the application and abstraction nodes
-in the syntax tree, and use the ones of meta language, as in the following:
-
-\begin{elpicode}
-finite (fin N).
-decision (nfact N NF).
-decision (all A x\ P x) :- finite A, pi x\ decision (P x).
-\end{elpicode}
-
-\noindent
-There are two reasons for dismissing this encoding. The first one is that
-in CIC it is not always possible to adopt it since the type system
-of the meta language is too weak to accommodate for the one of the 
-object language. In CIC the lambda abstraction has to carry a type in order to
-make type checking decidable. Moreover CIC allows for functions with a
-variable arity, like the following example:
-
-\begin{coqcode}
-Fixpoint arr T n := if n is S m then T -> arr T m else T.
-Definition sum n : arr nat n := ...
-Check sum 2   7 8   : nat.
-Check sum 3   7 8 9 : nat.
-\end{coqcode}
-
-\noindent
-The type system of the $\lambda$Prolog is too stringent to accept this terms.\\
-The second reason is that the CIC encoding provided by Elpi is used for meta
-programming (extending) the Coq system, hence it must accommodate the
-manipulation of terms that are now know in advance (not even defined in Coq)
-without using introspection primitives such as Prologs's \texttt{functor}
-and \texttt{arg}. In this sense constants have to live in an open world, like
-the \elpiIn{string} data type used in the examples so far.
-
-In the literature we could find related encoding of the Calculus of
-Constructions~\cite{felty93lics}. The goal of that work was to exhibit
-a logic program performing proof checking in CC and hence relate the
-proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
-programs) with the Calculus of Constructions. The encoding is hence tailored
-toward a different goal, and utilizes three relations to represent the
-equational theory of CC. Section 6 contains a discussion about the use of the
-unification procedure of the meta language in presence of non ground goals, but
-the authors do not aim at exploiting it to the degree we want.\todo{sucks}
-
-\todo{cite isabelle's TC, that are baked in}
-
-\section{Preliminaries: \Fo{} and \Ho}
-\label{sec:lang-spec}
-
-In order to reason about unification we provide a description of the
-\Fo{} and \Ho languages where unification variables
-are first class terms, i.e. they have a concrete syntax. We keep these languages
-minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
-in the example in Section~\ref{sec:intro} together with the type notation of
-terms carried by the \elpiIn{lam} constructor.
-%
-\setlength{\abovecaptionskip}{0pt}
-\setlength{\belowcaptionskip}{-13pt}
-
-\begin{figure}[H]
-  \begin{tabular}{ll}
-  \begin{minipage}{0.21\textwidth}
-   \inputrawelpicode{code/fo_tm}
-  \end{minipage}
-  &
-  \begin{minipage}{0.24\textwidth}
-   \inputrawelpicode{code/ho_tm}
-  \end{minipage}
-  \end{tabular}\vspace{4pt}
-  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
-  \label{code:common-terms}
-  \Description[code:common-terms]{code:common-terms}
-\end{figure}
-
-\noindent
-Unification variables (\elpiIn{fuva} term constructor)
-in \Fo{} have no explicit scope:
-the arguments of an higher order variable are given via the \elpiIn{fapp}
-constructor. For example the term \coqIn{P x} is represented as
-\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
-\elpiIn{x} is a bound variable.\\
-In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
-since unification variables come equipped with an explicit scope.
+% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
+% since decompilation can introduce (actually restore) terms in
+% \maybeeta, \maybebeta or \notllambda that were moved out of the way
+% during compilation.
+
+\section{Ground work for \Fo{}, \Ho and the compiler}
+\label{sec:grounwork}
+\todo{maybe move here napp beta deref fderef move}
+
 We say that the unification variable occurrence \elpiIn{uva N L} is in
@@
 
-\subsection{Notational conventions}
-
-When we write \Ho terms outside code blocks we follow the
-usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
-$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
-However we need to
-distinguish between the ``application'' of a unification variable
-to its scope and the application of a term to a list of arguments.
-We write the scope of unification variables in subscript
-while we use juxtaposition for regular application.
-Here a few examples:
-\vspace{4pt}
-
-\begin{tabular}{ll}
-  $f\appsep a$ &  \elpiIn{app[con "f", con "a"]}\\
-  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
-  $\lambda x.F_{x} \appsep a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
-  $\lambda x.F_{x} \appsep x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
-\end{tabular}
-\vspace{4pt}
-
-\noindent
-When variables $x$ and $y$ can occur in term $t$ we shall write
-$t_{xy}$ to stress this fact.
-
-We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
-\elpiIn{abs x\abs y\y} and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
-for \elpiIn{lam x\lam y\y}.
-
-When detailing examples we write links as equations between two
-terms under a context.
-The equality sign is subscripted with
-kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A_x}{F_x~a}$ corresponds to:
-\begin{elpicode}
-abs x\ val (link-beta (uva A [x]) (app[uva F [x],con "a"]))
-\end{elpicode}
-
-\noindent
-When it is clear from the context we shall use the same syntax for \Fo{} terms
-(although we never subscripts unification variables).
-
-\subsection{Equational theory and Unification}
-
-In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
-we need to equip \Fo{} and \Ho with term equality,
-substitution application and unification.
-
-\paragraph{Term equality: \Eo vs. \Ee} We extend the equational theory
-over ground terms to the full languages by adding the reflexivity of
-unification variables (a variable is equal to itself).
-
-The first four rules are common to both equalities
-and just define the usual congruence over terms, and since
-we use an HOAS encoding they also capture $\alpha$-equivalence.
-In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
-
-\input{code/feq}
-
-\begin{elpicode}
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-  con C ~\Ee~fcon C.
-  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
-  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
-  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
-\end{elpicode}
-
-\noindent
-The main point in showing these equality tests is to remark how weaker \Ee is,
-and to identify the four rules that need special treatment in
-the implementation of \Uo.
-
-For reference, \elpiIn{(beta T A R)} reduces away \elpiIn{lam} nodes in head
-position in \elpiIn{T} whenever the list \elpiIn{A} provides a corresponding
-argument.
+\noindent
+Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
+of the arguments of all occurrences of a unification variable and the
+number of abstractions in its assignment have to match. In turn
+this grants that \elpiIn{move} never fails.
+
+%\input{code/move}
 
@@
 
-\paragraph{Substitution application: $\rho s$ and $\sigma t$}
-
 Applying the substitution corresponds to dereferencing a term with respect to
-the memory. To ease the comparison we split \Fo dereferencing into a
+the memory.
+
+To ease the comparison we split \Fo dereferencing into a
 \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
@@
 \noindent
-Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
-of the arguments of all occurrences of a unification variable and the
-number of abstractions in its assignment have to match. In turn
-this grants that \elpiIn{move} never fails.
+Note that move strongly relies on invariant \ref{inv:uvaarity}
+that we shall define later.\todo{fixup}
 
@@
 
-\paragraph{Term unification: \Uo vs. \Ue}
-
-In this paper we assume to have an implementation of \Ue that satisfies
-properties~\ref{prop:correct-ml} and~\ref{prop:complete-ml}. Although we provide an
-implementation in the appendix (that we used for testing purposes) we only
-describe its signature here. Elpi is expected to provide this brick, as well as
-any other implementation of $\lambda$Prolog.
-
-\input{code/ue_type}
-
-\noindent
-The only detail worth discussing is the fact that the procedure updates a
-substitution, rather than just crafting one as presented in
-section~\ref{sec:problem-statement}. The reason is that the algorithm folds
-over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
+\subsection{Notational conventions}
+
+\noindent
+When variables $x$ and $y$ can occur in term $t$ we shall write
+$t_{xy}$ to stress this fact.
+
+We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
+\elpiIn{abs x\abs y\y} and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
+for \elpiIn{lam x\lam y\y}.
+
+When detailing examples we write links as equations between two
+terms under a context.
+The equality sign is subscripted with
+kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A_x}{F_x~a}$ corresponds to:
+\begin{elpicode}
+abs x\ val (link-beta (uva A [x]) (app[uva F [x],con "a"]))
+\end{elpicode}
+
+\todo{notation for problem compilation}
 
@@
 
-\subsection{Definition of \Uo and its properties}
+\subsection{Definition of \Uo and its properties}\label{sec:founif}
 
@@
 
-\newcommand{\occursrigid}{\emph{occurs-rigidly}\xspace}
-\newcommand{\occurrigid}{\emph{occur-rigidly}\xspace}
+\newcommand{\occursrigid}{\emph{occurs-rigidly}}
+\newcommand{\occurrigid}{\emph{occur-rigidly}}
 \begin{definition}[\occursrigid]\label{def:occrigid}
@@
 
-In other words $x$ \occursrigid in $t$ if it occurs in $t$
-outside of the scope of a unification variable $X$, otherwise an instantiation
-of $X$ can make $x$ disappears from $t$.
-Moreover, note that $\eta$-contracting $t$ cannot make $x$ disappear, since
-$x$ is not a locally bound variable inside $t$.
+\noteD{Not too clear?}
+In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
+outside of the scope of unification variables since an instantiation
+is allowed to discard $x$ from the scope of the unification variable.
+Note that $\eta$-contraction cannot make $x$ disappear, since the
+variables being erased by $\eta$-contraction are locally bound inside $t$.
 
@@
 \paragraph{Decompilation}
-Decompilation of the remaining \linketa (i.e. the \linketa that have been
-activated) is performed by iterating over them and unifying \lhs and \rhs. Note
-that this unification never fails, since \lhs is a flexible term not appearing
-in any other \linketa (by \cref{def:progressetadedup}).
+Decompilation of \linketa is performed by adding new rules to the
+\elpiIn{commit-link} predicate. In particular, given \linketaM{\Gamma}{X}{t}, we
+can note that this unification never fails, since $X$ is a flexible term and no
+other \linketa has $X$ has \lhs (by \cref{def:progressetadedup}). The link is
+remove from \linkStore and \elpiIn{commit-links} terminates.
 
@@
 
-% \noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
-% forse si può compattare in una prova più piccola?}
+\noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
+forse si può compattare in una prova più piccola?}
 \begin{lemma}
@@
 \begin{proof}[Proof sketch]
-  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
-  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
-  \cref{def:progressetaleft}, if 1) \lhs is a name, a constant of an
-  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
-  \cref{inv:link-eta-right}, $\rhs = \lambda x. t$, therefore $\wellb(t)$.
-  Otherwise, 2) \lhs has \elpiIn{lam} as functor, \rhs should not be an
-  $\eta$-expansion, so, $\wellb(\rhs)$. In both cases, unification is
-  performed between terms in \wellb.
+  Let $\sigma$ be the substitution, such that $\wellb(\sigma)$. $\lhs \in
+  \sigma$, therefore $\wellb(\lhs)$. By \cref{def:progressetaleft}, if 1) \lhs is
+  a name, a constant of an application, then, \lhs is unified with the
+  $\eta$-reduced term $t$ obtain from \rhs. By \cref{cor:rhs-eta-onelamb}, \rhs
+  has one lambda, therefore $\wellb(t)$. Otherwise, 2) \lhs has \elpiIn{lam} as
+  functor, \rhs should not be an $\eta$-expansion ans, so, $\wellb(\rhs)$. In
+  both cases, unification is performed between terms in \wellb.
 \end{proof}
@@
 \begin{proof}
-  The unification is done between the \rhs of two
-  \linketa. Both \rhs has the shape $\lambda x.t$, with $\wellb(t)$, therefore,
-  the unification is done between well-behaved terms.
+  Trivial, since the unification is done between unification variables, which
+  are by definition in \wellb.
 \end{proof}
@@
 \begin{lemma}
-  The introduction of \linketa guarantees \fullref{prop:nf}
-  \label{lemma:unif-wellb}
+  \Cref{prop:nf} holds, i.e., given a substitution $\sigma$ and a \linketa $l$,
+  after the activation of $l$, $\wellb(\sigma)$ holds.
 \end{lemma}

### 2024-04-09 paper/main.tex
rename properties

@@
 
-\begin{proposition}[Properties of \hrun outside \llambda]
+\begin{proposition}[Fidelity recovery]
 \begin{gather}
@@
 
-\begin{proposition}[\wellb{}-preservation of \hstep]\label{prop:nf}
+\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
 $\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
@@
 (an application node that could be erased by a $\beta$-reduction).
+\todo{dire che W preservation is key for our claims}
 

### 2024-04-09 paper/main.tex
delenda est

@@
 % \title{HO unification from object language to meta language}
-\title{Higher order unification for free!}
+\title{Higher-Order unification for free!}
 \subtitle{Reusing the meta-language unification for the object language}
@@
 %
+\vspace{4pt}
+{
 \setlength{\tabcolsep}{2pt}
-\vspace{4pt}
 \begin{tabular}{lcl|}
@@
 \end{tabular}
+}
 \vspace{4pt}
@@
 treatment in the implementation of \Uo.
-We defer the description of \elpiIn{beta} to \cref{sec:grounwork}, for now
+For brevity we omit the code of \elpiIn{beta}:
 it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
@@
 \begin{definition}[\notllambda]\label{def:notllambda}
-  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
+  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
 \end{definition}
@@
 \noindent
-A less formal way to state \ref{prop:nf} is that \hstep{} and \progress never
-``commit'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
-that could be erased by an $\eta$-contraction),
-nor put in $\sigma$ an application whose head is flexible
-(an application node that could be erased by a $\beta$-reduction).
-\todo{dire che W preservation is key for our claims}
-
-% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
-% since decompilation can introduce (actually restore) terms in
-% \maybeeta, \maybebeta or \notllambda that were moved out of the way
-% during compilation.
-
-\section{Ground work for \Fo{}, \Ho and the compiler}
+\cref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
+the problematic terms moved on the side by the compiler are not
+put back by \hstep, hence \Ue{} can operate properly.
+In \cref{sec:compilation,sec:eta,sec:beta}
+we describe how the compiler recognizes terms in \maybeeta, \maybebeta and
+\notllambda and how progress takes care of them preserving \wellb
+and granting \cref{prop:simulation,prop:fidelity}.
+
+Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
+since decompilation can introduce (actually restore) problematic terms.
+
+\section{Ground work for the compiler}
 \label{sec:grounwork}
-\todo{maybe move here napp beta deref fderef move}
-
-We say that the unification variable occurrence \elpiIn{uva N L} is in
-\llambda if and only if \elpiIn{L} is made of distinct names. The
-predicate to test this condition is called \elpiIn{pattern-fragment}:
-
-\input{code/pattern_fragment}
-
-\noindent
-Natural numbers represent the memory addresses that
-identify unification variables in both languages.
+
+% The predicate to test this condition is called \elpiIn{pattern-fragment}:
+
+% \input{code/pattern_fragment}
+
+Unification variables are identified by a
+natural number, that represents a memory addresses
 The memory and its associated operations are described below:
@@
 is not set. \elpiIn{assign} sets an unset cell to the given value, while
-\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\noteE[inline]{is new used?}
-
-Since in \Ho unification variables have a scope, their solution needs to be
-abstracted over it to enable the instantiation of a single
-solution to different scopes. This is obtained via the \elpiIn{inctx}
-container, and in particular via its \elpiIn{abs} binding constructor.
+\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.
+
+Since each occurrence of a \Ho unification variables has a scope,
+its solution needs to be abstracted over it to enable the
+instantiation of a single assignment to different scopes.
+This is expressed by the \elpiIn{inctx} container, and in particular
+its \elpiIn{abs} binding constructor.
 On the contrary a solution to a \Fo variable is a plain term.
@@
 the one of \Ho.
-Both have the invariant that they are not cyclic, TODO: explain.
+The compiler establishes a mapping between variables of the two languages.
 
 \input{code/comp_base_types}
+
+Each \elpiIn{hvariable} is stored in the mapping together with
+its arity so that the code of \ref{clause:malloc} below can preserve:
 
@@
 
-\noindent
-The compiler establishes a mapping between variables of the two languages.
-In order to preserve invariant \ref{inv:uvaarity} we store the
-arity of each \elpiIn{hvariable} in the mapping and we reuse an existing
-mapping only if the arity matches.
-
-\noteD{add ref to \cref{sec:invariant1}}
-
 \input{code/alloc_mapping}
@@
 approximation, and then makes the mapping bijective by introducing
-\linketa; this detail is discussed in section \ref{sec:eta}.
-
+\linketa; this detail is discussed in section \ref{sec:invariant1}.
+
+\noindent
 As we mentioned in section~\ref{sec:nutshell} the compiler
-replaces terms in \maybeeta and \notllambda with fresh variables
-linked to the problematic terms. Each class of problematic terms
-has a dedicated link.
+replaces terms in \maybeeta, \maybebeta and \notllambda with fresh
+variables linked to the problematic terms. Terms in \maybebeta do not
+need a link since \Ho{} variables faithfully represent
+the problematic term thanks to their scope.
 
@@
 
+% \input{code/beta_fo}
+
+% \noindent
+% The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
+% variables).\footnote{Elpi provides it as a builtin, but one could implement it by
+% systematically loading the hypothetical rule \elpiIn{name x} every time
+% a nominal constant is postulated via \elpiInFN{pi x\ }}
+% The choice of using n-ary application, rather than binary, is to make it
+% easy to access the application's head. The price we pay is that substituting
+% an application in the head of an application should be amended by
+% ``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
+% \footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
+% we do in this presentation and, as we explain later,
+% no equivalent of \elpiIn{napp} is needed in \Ho.}
+% Finally note that the cut operator is inessential, it could be
+% removed at the cost of a verbose test on the head of \elpiIn{L}
+% in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
+% can be \elpiIn{fcon}, \elpiIn{flam} or a name.
+
+Applying the substitution corresponds to dereferencing a term with respect to
+the memory. To ease the comparison we split \Fo{} dereferencing into a
+\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
+to memory cells that are set with their values, and has a corresponding
+operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
+has no corresponding operation in \Ho, and only ensures that
+terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
+<<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
+and \elpiIn{L2}. The reasons for this asymmetry is
+that an \elpiIn{fapp} node with a flexible head is always mapped
+to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
+preventing nested applications to materialize.
+
+\input{code/fderef}
+
+\noindent
+Applying the substitution in \Ho{} is very similar, with
+the caveat that assignments have to be moved to the
+current scope, i.e. renaming the \elpiIn{abs}-bound variables
+with the names in the scope of the unification variable occurrence.
+
+\input{code/deref}
+
 \noindent
@@
 this grants that \elpiIn{move} never fails.
-
-%\input{code/move}
-
-\input{code/beta_fo}
-
-\noindent
-The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
-variables).\footnote{Elpi provides it as a builtin, but one could implement it by
-systematically loading the hypothetical rule \elpiIn{name x} every time
-a nominal constant is postulated via \elpiInFN{pi x\ }}
-The choice of using n-ary application, rather than binary, is to make it
-easy to access the application's head. The price we pay is that substituting
-an application in the head of an application should be amended by
-``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
-\footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
-we do in this presentation and, as we explain later,
-no equivalent of \elpiIn{napp} is needed in \Ho.}
-Finally note that the cut operator is inessential, it could be
-removed at the cost of a verbose test on the head of \elpiIn{L}
-in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
-can be \elpiIn{fcon}, \elpiIn{flam} or a name.
-
-Applying the substitution corresponds to dereferencing a term with respect to
-the memory.
-
-To ease the comparison we split \Fo dereferencing into a
-\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
-to memory cells that are set with their values, and has a corresponding
-operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
-has no corresponding operation in \Ho. The reasons for this asymmetry is
-that an \elpiIn{fapp} node with a flexible head is always mapped
-to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
-preventing nested applications to materialize.
-
-\input{code/fderef}
-
-\noindent
-Applying the substitution in \Ho{} is very similar, with
-the caveat that assignments have to be moved to the
-current scope, i.e. renaming the \elpiIn{abs}-bound variables
-with the names in the scope of the unification variable occurrence.
-
-\input{code/deref}
-
-\noindent
-Note that move strongly relies on invariant \ref{inv:uvaarity}
-that we shall define later.\todo{fixup}
 

### 2024-04-09 paper/main.tex
typesetting

@@
 since the abstraction and application term constructors
-the two unification procedures deal with are different. For example:\\
-%
-\vspace{4pt}
-{
-\setlength{\tabcolsep}{2pt}
-\begin{tabular}{lcl|}
-\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
-\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
-\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
-\end{tabular}
-\phantom{mn}
-\begin{tabular}{lcl}
-\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
-\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
-\end{tabular}
-}
-\vspace{4pt}
+the two unification procedures deal with are different.
+% For example:\\
+% %
+% \vspace{4pt}
+% {
+% \setlength{\tabcolsep}{2pt}
+% \begin{tabular}{lcl|}
+% \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
+% \elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
+% \elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
+% \end{tabular}
+% \phantom{mn}
+% \begin{tabular}{lcl}
+% \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
+% \elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
+% \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
+% \end{tabular}
+% }
+% \vspace{4pt}
 %
@@
 
-\cref{sec:problem-statement} formally states the problem and gives the
+\todo{fix}\cref{sec:problem-statement} formally states the problem and gives the
 intuition behind our solution. \cref{sec:encodings} discusses alternative
@@
 \Fo{} and \Ho languages where unification variables
-are first class terms, i.e. they have a concrete syntax.
+are first class terms, i.e. they have a concrete syntax as
+per \cref{code:common-terms}.
 % We keep these languages
@@
 \setlength{\belowcaptionskip}{-13pt}
-
-\begin{figure}[H]
+\begin{figure} %[H]
   \begin{tabular}{ll}
@@
 }
-\noindent
 Unification variables (\elpiIn{fuva} term constructor)
@@
   \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
+\end{array}
+$$
+$$
+\begin{array}{l}
   \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
@@
 
-\begin{definition}[\maybeeta]\label{def:maybeeta}
-  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
-\end{definition}
-
-\noindent
-An example of term $t$ in \maybeeta{} is
-$\lambda x.\lambda y.F \appsep y \appsep x$
-since the substitution
-$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
-makes $\rho t = \lambda x.\lambda y.f x y$
-that is the eta long form of $f$. This term is problematic since
-its rigid part, the $\lambda$-abstractions, cannot justify a
-unification failure against a constant or an application.
-
 \begin{definition}[\maybebeta]\label{def:maybebeta}
-  $\maybebeta = \{ X \appsep x_1 \ldots x_n ~|~ x_1 \ldots x_n ~\mathrm{are~distinct~names} \}$
+  \maybebeta is the set of terms of the form $X \appsep x_1 \ldots x_n$
+  such that $x_1 \ldots x_n$ are distinct names (of bound variables).
 \end{definition}
@@
 
+\begin{definition}[\maybeeta]\label{def:maybeeta}
+  \maybeeta is the set of terms $t$ such that $\exists \rho, \rho t$
+  is an eta expansion.
+\end{definition}
+
+\noindent
+An example of term $t$ in \maybeeta{} is
+$\lambda x.\lambda y.F \appsep y \appsep x$
+since the substitution
+$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
+makes $\rho t = \lambda x.\lambda y.f \appsep x \appsep y$
+that is the eta long form of $f$. This term is problematic since
+its leading $\lambda$ abstraction cannot justify a
+unification failure against a constant or an application.
+
 \begin{definition}[\notllambda]\label{def:notllambda}
-  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
+  \notllambda is the set of terms of the form $X \appsep t_1 \ldots t_n$
+  such that $t_1 \ldots t_n$ are not distinct names.
 \end{definition}
@@
 \noindent
-\cref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
-the problematic terms moved on the side by the compiler are not
-put back by \hstep, hence \Ue{} can operate properly.
+Proposition \ref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
+informally it says that the problematic terms moved on the side by the compiler
+are not put back by \hstep, hence \Ue{} can operate properly.
 In \cref{sec:compilation,sec:eta,sec:beta}
-we describe how the compiler recognizes terms in \maybeeta, \maybebeta and
-\notllambda and how progress takes care of them preserving \wellb
+we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
+\notllambda and how \progress takes care of them preserving \wellb
 and granting \cref{prop:simulation,prop:fidelity}.
 
-Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
-since decompilation can introduce (actually restore) problematic terms.
+% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
+% since decompilation can introduce (actually restore) problematic terms.
 
@@
 later, in the following sections. This scheme is sufficient to implement
-an \Uo{} that respects $\beta$-conversion for terms in \llambda. The extension to
-$\eta\beta$-conversion is described in Section \ref{sec:eta} and the support
-for terms outside \llambda in Section \ref{sec:beta}.
+a \hstep that respects $\beta$-conversion for terms in \llambda.
+The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
+the support for terms outside \llambda in \cref{sec:beta}.
 

### 2024-04-09 paper/main.tex
small changes

@@
 % section is given in a future section
-\begin{lemma}[\linkbeta with rigid \lhs]
-  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
-  counterpart is still in \notllambda,
-  the original unification problem is not in \llambda
-  % \noteE[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
-  %\noteE[inline]{Is it clearer?}
-  and the unification fails.
-  \label{lemma:keep-fidelity}
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-  Given $X \appsep t_1\dots t_n \Ue t$ where $t$ is a rigid term and
-  $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
-  is replaced with a variable $Y$, and the \linkbeta \linkbetaM{\Gamma}{Y}{X
-  \appsep t_1\dots t_n} is created. If $Y$ is instantiated to the rigid term $t$,
-  and \rhs is still in \notllambda, then the original \Fo{} unification problem
-  is between two terms in \notllambda. Unification fails.
-\end{proof}
 
@@
 \newcommand{\progBetaLL}{\emph{progress-beta-\llambda}\xspace}
-\newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}\xspace}
+\newcommand{\progBetaRH}{\emph{progress-beta-rigid-\rhs}\xspace}
 \newcommand{\progBetaDedup}{\emph{progress-beta-dedup}\xspace}
+\newcommand{\progBetaLH}{\emph{progress-rigid-\lhs}\xspace}
 
@@
 
-\noteD{added the def below to respect fidelity}
 % Example of this `X a = Y a'
@@
   Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
-  share the same \lhs. In this case, the two \rhs are unified and a $l_2$ is
+  share the same \lhs. The two \rhs are unified and a $l_2$ is
   removed from \linkStore.
   \label{def:progBetaDedup}
+\end{definition}
+
+\begin{definition}[\progBetaLH]
+  Given a \linkbeta with rigid \lhs, the unification fails.
+  \label{def:progBetaLH}
 \end{definition}
@@
   \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
-  terminates. Finally \progBetaDedup also guarantees termination since it 
-  makes a unification $\mathbb{U}$ and if $\mathbb{U}$ fails, then 
-  \elpiIn{progress} terminates and if $\mathbb{U}$ succeeds, 
-  the recursive calls to \elpiIn{progress} have a specialized \linkbeta\dots
+  terminates. Finally \progBetaDedup (resp. \progBetaLH) also guarantees termination since 
+  it removes a link from \linkStore (resp. immediately fails).
 
@@
 
-\begin{corollary}
-  Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
-  % \noteE[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
-  \noteD[inline]{is it clearer?}
-\end{corollary}
-
-\begin{proof}[Proof sketch]
-  By construction, the \rhs of \linkbeta has the shape \rhsBeta, $s_1\dots s_n$
-  is in \llambda and all the terms $t_1\dots t_n$ are in \llambda, too. If a
-  \linkbeta is triggered by \progBetaRH, then, by \cref{def:progBetaRH}, that
-  link is removed by \linkStore, and the property is satisfied. If the \linketa
-  is activated by \progBetaLL, then, by \cref{def:progBetaLL}, the new \linkbeta
-  as a variable as a scope which is still in \llambda. 
-\end{proof}
+% \begin{corollary}
+%   Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
+%   % \noteE[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
+%   \noteD[inline]{is it clearer?}
+% \end{corollary}
+
+% \begin{proof}[Proof sketch]
+%   By construction, the \rhs of \linkbeta has the shape \rhsBeta, $s_1\dots s_n$
+%   is in \llambda and all the terms $t_1\dots t_n$ are in \llambda, too. If a
+%   \linkbeta is triggered by \progBetaRH, then, by \cref{def:progBetaRH}, that
+%   link is removed by \linkStore, and the property is satisfied. If the \linketa
+%   is activated by \progBetaLL, then, by \cref{def:progBetaLL}, the new \linkbeta
+%   as a variable as a scope which is still in \llambda. 
+% \end{proof}
 
 \begin{lemma}[Fidelity with \linkbeta]
-  The introduction of \linkbeta guarantees \cref{prop:fidelity} 
+  The introduction of \linkbeta guarantees \fullref{prop:fidelity} 
 \end{lemma}
@@
   fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
-  then, by \cref{lemma:keep-fidelity}, the unification fails, as per the
+  then, by \cref{def:progBetaLH}, the unification fails, as per the
   corresponding unification in \Fo{}.

### 2024-04-10 paper/main.tex
typo
update ref

@@
 \begin{lemma}
-Properties \ref{prop:correct} and
-\ref{prop:complete} hold for the implementation of \Uo above
+Properties \ref{prop:correct-ml} and
+\ref{prop:complete-ml} hold for the implementation of \Uo above
 \end{lemma}
@@
  In this setting \Ee is as strong as
-\Eo on ground terms. What we have to show is that whenever two different \Fo
+\Eo on ground terms. What we have to show is that whenever two different \Fo{}
 terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
@@
 
-\subsection{Relaxing lemma \nameref{lemma:keep-fidelity}}
+\subsection{Relaxing \fullref{def:progBetaLH}}
 
@@
 argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
-that this is of course an approximation, since $\sigma = \{X = \lambda x.Y, Y =
+that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
 \_\}$ is another valid substitution for the original problem. This approximation

### 2024-04-10 paper/main.tex
update ref
clarify proofs

@@
 \crefname{invariant}{invariant}{invariants}
+\crefname{equation}{property}{properties}
 
@@
 \begin{lemma}
-Properties \ref{prop:correct-ml} and
-\ref{prop:complete-ml} hold for the implementation of \Uo above
+\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
 \end{lemma}
@@
 
-\begin{lemma} Properties simulation (\ref{prop:simulation}) and
-fidelity (\ref{prop:fidelity}) hold
+\begin{lemma} \Fullref{prop:simulation} and
+\fullref{prop:fidelity} hold
 \end{lemma}
@@
 
+\begin{lemma}
+  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb{t}$.
+  \label{lemma:unif-eta-aux}
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  By construction, every ``problematic'' term in \Fo{} is replaced with a variable
+  in the corresponding \Ho term. Therefore, $t$ is \wellb.
+\end{proof}
+
 % \noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
@@
   \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
-  \cref{def:progressetaleft}, if 1) \lhs is a name, a constant of an
+  \progressetaleft, if 1) \lhs is a name, a constant or an
   application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
-  \cref{inv:link-eta-right}, $\rhs = \lambda x. t$, therefore $\wellb(t)$.
-  Otherwise, 2) \lhs has \elpiIn{lam} as functor, \rhs should not be an
-  $\eta$-expansion, so, $\wellb(\rhs)$. In both cases, unification is
+  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux}, $\rhs = \lambda x. t$ and $\wellb(t)$.
+  Otherwise, 2) \lhs has \elpiIn{lam} as functor. In both cases, unification is
   performed between terms in \wellb.
@@
   \lhs is variable, and, by \cref{def:progressetaright}, \rhs is either no more
-  a \maybeeta, i.e. \rhs is not a $\eta$-expansion and, so, $\wellb(\rhs)$.
-  Otherwise, \rhs can reduce to a term which cannot be a $\eta$-expansion, and,
-  so, $\wellb(\rhs)$. In both cases, unification is done between terms in
-  \wellb.
+  a \maybeeta, i.e. \rhs is not a $\eta$-expansion and, so, $\wellb(\rhs)$,
+  otherwise, \rhs can reduce to a term which cannot be a $\eta$-expansion, and,
+  so, $\wellb(\rhs)$. In both cases, the unification between \rhs and \lhs
+  is done between terms that are in \wellb.
 \end{proof}
@@
 \begin{proof}
-  The unification is done between the \rhs of two
-  \linketa. Both \rhs has the shape $\lambda x.t$, with $\wellb(t)$, therefore,
-  the unification is done between well-behaved terms.
+  The unification is done between the \rhs of two \linketa. Both \rhs has the
+  shape $\lambda x.t$, and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
+  Therefore, the unification is done between well-behaved terms.
 \end{proof}
@@
   terms in \wellb, therefore, the substitution remains \wellb.
-  \noteD[inline]{Bisogna aggiungere un lemma nella \cref{sec:nutshell} che dice che
-  unificare due termini in \wellb, in una $\sigma$, tale che $\wellb(\sigma)$, 
-  non invalida \wellb}
 \end{proof}

### 2024-04-10 paper/main.tex
update lemma+proofs

@@
 
-\begin{invariant}[Unification variable arity]
+\begin{invariant}[Unification-variable arity]
   Each variable \elpiIn{A}
@@
 
-\section{Enforcing Invariant~\ref{inv:uvaarity}}
+\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
 \label{sec:invariant1}
@@
 $X$ is used with two different arities and the output of the compilation does
-not respect \cref{inv:map-bijective} (merging the two mappings
-for $s$ would break \cref{inv:uvaarity}).
-In this section we explain how to replace the duplicate mapping with some
-\linketa{} in order to restore the invariants.
+not respect \cref{inv:map-bijective}. Moreover, merging the two mappings\noteD{TODO: da cambiare 488 in tex}
+for $s$ would break \fullref{inv:uvaarity}.
+In this section we explain how to replace the duplicate mapping with
+\linketa in order to restore the invariants.
 % make test ONLY=7001 TEX=tex
@@
 In this example, $\hoUnifPb_1$ assigns $A$ which triggers $\linkStore_3$ and
-then $\linkStore_4$ by \cref{def:progressetaleft}. $C_{yx}$ is therefore
+then $\linkStore_4$ by \progressetaleft. $C_{yx}$ is therefore
 assigned to $x$ (the second variable of its scope). We can finally see the
-\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ ($C_{xy}$
-gives $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
-$E_x$. Moreover, $\linkStore_2$ is also triggered due to
-\cref{def:progressetaright}: $\lambda x.(f \appsep (\lambda y.y) \appsep x)$ is
-$\eta$-reducible to $f \appsep (\lambda y.y)$ which is a term not starting with
-the \elpiIn{lam} constructor.
+\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ (the term $C_{xy}$
+reduces to $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
+$E_x$. After the execution of the remaining \hstep, we obtain the 
+following \Fo{} substitution $\rho = \{X := \lambda x.\lambda y.y, Y := (f\appsep \lambda x.x)\}$.
 
@@
 solution and committing one of them in the substitution does not guarantee
-\cref{prop:complete}. For instance, $X \appsep a \Uo a$ 
+\cref{prop:complete-ml}. For instance, $X \appsep a \Uo a$ 
 admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
@@
 
-It is the case, however, that, given a list of unification problems, $\foUnifPb_1\dots
-\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, the resolution of
+Given a list of unification problems, $\foUnifPb_1\dots
+\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, it is often the case that the resolution of
 $\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
@@
 To address
-this unification problem, term compilation should capture the
-terms in \notllambda and replace them with fresh variables.
-This replacement should produce links that we call \linkbeta.
-
-\linkbeta guarantees \cref{inv:linklhs} and the term on the \rhs has the
+this unification problem, term compilation must recognize and replace
+\notllambda terms with fresh variables.
+This replacement produces links that we call \linkbeta.
+
+\linkbeta respects \cref{inv:linklhs} and the term on the \rhs has the
 following property:
@@
   The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
-  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
-  terms. This is equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
-  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
+  unification variable with scope $s_1\dots s_n$\footnote{with $s_1\dots s_n$
+  that are distinct names} and $t_1\dots t_m$ is a list of terms. This is
+  equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$= s_1\dots s_n$
+  and \elpiIn{L}~$=t_1\dots t_m$.
   \label{inv:beta-rhs}
@@
 variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
-scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}
+scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}.
+Note that this construction enforce \cref{inv:beta-rhs}.
 
@@
 % T]}.
-
-\begin{invariant}
-  The \rhs of a \linkbeta has the shape \rhsBeta.
-\end{invariant}
 

### 2024-04-10 paper/main.tex
section 1 and 2

@@
 logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
-The use of these meta languages facilitated this task in two
+The use of these meta languages facilitates this task in two
 key ways. The first and most well know one is that variable binding and
@@
 The second one is that these meta languages come equipped with some form
-of unification, a cornerstone in proof construction and proof search.
+of unification, a cornerstone of proof construction and proof search.
 
@@
 form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
-In particular we want to leverage the Elpi~\cite{tassi:hal-01637063} meta programming language,
-a dialect of $\lambda$Prolog already used to extend
-Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
 Type-class solvers are unification based proof search procedures
 reminiscent of Prolog that back-chain lemmas taken
-from a designated database of ``type class instances'', hence
-we can expect that Elpi is a good fit for
-implementing such as form of automation. In this paper we focus on one aspect of
-this work, namely \emph{how to reuse the higher order unification procedure
-of the meta language in order to implement a type-class solver for the object
-language}. As it turns out, re-using the unification of the meta language is
-not a trivial task.
+from a designated database of ``type class instances''. Given this
+analogy with Logic Programming we want to leverage the
+Elpi~\cite{tassi:hal-01637063} meta programming language,
+a dialect of $\lambda$Prolog, already used to extend
+Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
+In this paper we focus on one aspect of
+this work, precisely \emph{how to reuse the higher order unification procedure
+of the meta language in order to simulate a higher order logic program
+for the object language}.
 
@@
 \coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
-The following three designated \coqIn{Instances} state that:
-1) the type \coqIn{fin n}, of natural numbers
-smaller than \coqIn{n}, is finite;
-2) the predicate \coqIn{nfact n nf}, linking a natural number
+The following three designated type-class instances state that:
+1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
+is finite;
+2) the predicate \coqIn{nfact n nf}, relating a natural number
 \coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
@@
 The proof found by the solver back-chains on rule 3 (the only rule
-about the $\forall$ quantifier), and then solves the premises with rules
+about the $\forall$ quantifier), and then solves the premises with
 rules 1 and 2 respectively.
@@
 The solver has to infer a value for \coqIn{P} by unifying the conclusion
-of rule 3 with the goal, and in particular by solving the unification
+of rule 3 with the goal, and in particular it has to solve the unification
 problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
-called pattern-fragment \llambda~\cite{miller92jsc} and hence admits a unique
-solution \coqIn{P = ~$\lambda$~x.nfact x 3}.
+called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
+solution $\customlabel{solution:intro}{\sigma}$ that assigns
+the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
 
 In order to implement such a search in Elpi we shall describe the encoding
-of CIC terms and then the encoding of rules. Elpi comes with
-an Higher Order Abstract Syntax~\cite{10.1145/53990.54010} datatype of CIC
+of CIC terms and then the encoding of instances as rules (a.k.a. clauses).
+Elpi comes equipped with
+an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
 terms, called \elpiIn{tm}, that features (among others) the following
@@
 the meta level binding of a variable \elpiIn{x} in an expression
-\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets denote a
+\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets delimit a
 list of terms separated by comma. For example the term
@@
 \begin{elpicode}
-all (con"t") y\ app[con"nfact", y, con"3"]
+all (con"t") y\ app [con"nfact", y, con"3"]
 \end{elpicode}
@@
 language its type is \elpiIn{tm}. If we try to back-chain the rule
-\ref{clause:r3} on the encoding of the goal g given below
+\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below
 
@@
 \noindent
-In this paper we study a more sophisticated encoding of rules that, on a first
-approximation, would shape \ref{clause:r3} as follows:
+In this paper we study a more sophisticated encoding of CIC terms and rules
+that, on a first approximation, would reshape \ref{clause:r3} as follows:
 
@@
 the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
-back to the domain \elpiIn{tm} of Coq terms:
+back to the domain \elpiIn{tm} of Coq terms, obtaining the expected solution
+\ref{solution:intro}:
 
 \begin{elpicode}
-P = lam A a\ app [con"nfact", a, con"3"]
+P = lam A x\ app [con"nfact", x, con"3"]
 \end{elpicode}
@@
 \begin{elpicode}
-decision (app[                    P                   , w])
-decision (app[ lam A (a\ app [con"nfact", a, con"3"]) , w])
+decision (app [                   P                  , w])
+decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
 \end{elpicode}
@@
 The root cause of the problems we sketched in this example
-is a subtle mismatch between the equational theories and
-unification procedures of the meta language and the object language.
-
-The equational theory of CIC is very rich. In
-addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
-are compared up to proposition unfolding and fixpoint unrolling. Still,
-for efficiency and predictability reasons, most form of automatic proof search
-employ a unification procedure that captures a simpler one,
-just $\eta\beta$, and that solves higher-order problems
-restricted to the pattern fragment \llambda.
-We call this unification procedure \Uo{}.
-
-The equational theory of the meta language Elpi is strikingly similar,
-since it it comprises $\eta\beta$ (for the meta language functions), and the
-unification procedure \Ue{} solves problems in
-\llambda as well.
-
-In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
-since the abstraction and application term constructors
-the two unification procedures deal with are different.
+is a subtle mismatch between the equational theories of the meta language
+and the object language, that in turns makes the 
+unification procedures of the meta language weak.\\
+The equational theory of the meta language Elpi encompasses
+$\eta\beta$-equivalence and its unification procedure can solve higher order
+problems in the pattern fragment. Although 
+the equational theory of CIC is much richer, for efficiency and predictability
+reasons automatic proof search procedure typically employ a unification
+procedure that only captures a $\eta\beta$-equivalence and only operates
+in \llambda. The similarity is striking, but one needs some care in order to
+simulate a logic program in CIC using the unification of Elpi.
+
 % For example:\\
@@
 \setlength{\belowcaptionskip}{-13pt}
-\begin{figure} %[H]
+\begin{figure}[b] %[H]
   \begin{tabular}{ll}
@@
 }
-Unification variables (\elpiIn{fuva} term constructor)
-in \Fo{} have no explicit scope:
+Unification variables
+in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
 the arguments of an higher order variable are given via the \elpiIn{fapp}
 constructor. For example the term <<\coqIn{P x}>> is represented as
-<<\elpiIn{fapp[fuva N, x]}>>, where \elpiIn{N} is a memory address and
-\elpiIn{x} is a bound variable.\\
+<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
+of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
 In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
@@
 The first four rules are common to both equalities
-and just define the usual congruence over terms. Since
+and define the usual congruence over terms. Since
 we use an HOAS encoding they also capture $\alpha$-equivalence.
@@
 
-We write $\sigma = \{~ X \mapsto t ~\}$ the substitution that assigns
+We write $\sigma = \{~ X \mapsto t ~\}$ for the substitution that assigns
 the term $t$ to the variable $X$.
@@
 it provides an assignment.
-We write $\sigma \cup \sigma'$ set union to denote the concatenation of
+We write $\sigma \cup \sigma'$ to denote the concatenation of
 two substitutions whose domains are disjoint.
@@
 and $\sigma$ for the \Ho ones.
-\todo{can be made a little more formal}
 For brevity, in this section we consider
@@
 
-Although we provide an implementation of \Ue in the supplementary material
-(that we used for testing purposes) we only describe its signature here.
+Although we provide an implementation of the meta-language unification \Ue
+in the supplementary material (that we used for testing purposes) we only
+describe its signature here.
 
 \input{code/ue_type}
+
+We write 
+$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
+$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
+We write $t_1 \Ue t_2 \mapsto \sigma$ when
+the initial substitution is empty.
+Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
+the domains of $\sigma$ and $\sigma'$ are disjoint.
 
@@
 
-%
-% To state precisely the problem we solve we need a \Fo{} representation
-% of CIC terms and a \Ho one. We extend the equality over ground
-% terms to open ones with the reflexivity of unification variables (that is
-% unification variable X is equal to itself, not to another one such a Y).
-% We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
-% We call \Uo the unification procedure we want to implement and
-% \Ue the one provided by the meta language.
-
-We write 
-$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
-$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
-We write $t_1 \Ue t_2 \mapsto \sigma$ when
-the initial substitution is empty.
-Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
-the domains of $\sigma$ and $\sigma'$ are disjoint.
-
-Although we provide an implementation of \Uo{} in \cref{sec:founif},
-our real goal is the simulation of an entire logic program.
-
-\subsection{The problem: Logic Program Simulation}
+Even if we provide an implementation of the object-language unification
+\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
+logic program.
+
+\subsection{The problem: logic-program Simulation}
 
@@
 \noindent
+\todo{say that backtracking is not important} 
 In order to simulate a \Fo{} logic program in \Ho{} we compile
-each term $s \in \Fo{}$ into a term $t \in \Ho{}$. We write this
-step $\C{s}{t}{m}{l}$. The implementation of the compiler
-is detailed in \cref{sec:compilation,sec:eta,sec:beta}, here we just point
-out that it additionally a variable mapping $m$ and list of links $l$.
-The variable map connects unification variables in \Ho with variables
+each \Fo{} term in \foUnifPb into a \Ho{} term $t$.
+We write this step $\C{s}{t}{m}{l}$. The implementation of the compiler
+is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
+out that it additionally produce a variable mapping $m$ and list of links $l$.
+The variable map connects unification variables in \Ho to variables
 in \Fo{} and is used to ``decompile'' the assignment,
@@
 
-%\cref{sec:beta}
-%XXX permuting hrun does not change the final result if check does not fail eagerly
-% XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to a non MGU too early
-
-% We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
-% %$\foUnifPb = \{ s_1, s_2 \}$ 
-% as follows:
-% %
-% $$
-% \begin{array}{l}
-% s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
-% \quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
-% \quad    t_1 \Ue t_2 \mapsto \sigma' \land
-%     \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
-% \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
-% \end{array}
-% $$
-
-% \begin{proposition}[Properties of \Uo{} in \llambda]
-% \begin{gather}
-%   \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
-%   \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
-% \end{gather}
-% \end{proposition}
-
-%In addition to correctness and completeness in \llambda,
-We also claim that \hrun
-handles terms outside \llambda in the following sense:
-
-\begin{proposition}[Fidelity recovery]
-\begin{gather}
-  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
-  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
-  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
-  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow \label{prop:step-nllam}\\
-  \exists \rho,\stepF{\foUnifPb}{p}{\rho}{\rho'} \Leftrightarrow
-  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \nonumber
-\end{gather}
+We also claim that \hrun handles terms outside \llambda in the following sense:
+
+\begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
+In the context of \hrun, if $\sigma_{p-1} \hoUnifPb_{p} \in \llambda$ then
+$$
+\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p} \Leftrightarrow
+\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p}
+$$
 \end{proposition}
 
-\noindent
-Property \ref{prop:compilation-nllam} states that two terms for which
-there is a unifier (given by an oracle and not necessarily a most general one),
-then the compiler generates two terms that unify in \Ho.\\
-%In other words the compiler moves out of the way all problematic terms.
-Property \ref{prop:step-nllam} says that if the two terms involved in a step
-re-enter \llambda, then \hstep succeeds. This is a typical example
-in which the order of the unification problems in a logic-program run
-does matter. The simplest example is the sequence $F \UnifRel \lambda x.a$ and
+In other words if the two terms involved in a step
+re-enter \llambda, then \hstep and \fstep are again related, even if
+$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
+
+This property has a practical relevance since in many logic programming
+implementations, including Elpi, the order in which unification problems
+are tackled does matter.
+The simplest example is the sequence $F \UnifRel \lambda x.a$ and
 $F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
@@
 $\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
-making the second problem re-enter \llambda. In other words
-\cref{prop:fidelity} \todo{...}
+making the second problem re-enter \llambda.
 
@@
 \label{sec:nutshell}
-A term $s$ is compiled in a term $t$ where every
+A term $s$ is compiled to a term $t$ where every
 ``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
-and an accessory \emph{link} that represents a suspended unification problem
+with an accessory \emph{link} that represents a suspended unification problem
 $h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
 it does not contradict \Eo as it would otherwise do on the
-``problematic'' sub-terms. We now define ``problematic'' and ``well behaved''
-more formally.
+``problematic'' sub-terms.\\
+We now define ``problematic'' and ``well behaved'' more formally.
 
@@
 \noindent
-An example of term $t$ in \maybebeta{} is the application $F \appsep x$.
+An example of term \maybebeta{} is the application $F \appsep x$.
 This term is problematic since the application node of
@@
 \begin{definition}[\maybeeta]\label{def:maybeeta}
-  \maybeeta is the set of terms $t$ such that $\exists \rho, \rho t$
+  \maybeeta is the set of terms $s$ such that $\exists \rho, \rho s$
   is an eta expansion.
@@
 \noindent
-An example of term $t$ in \maybeeta{} is
+An example of term $s$ in \maybeeta{} is
 $\lambda x.\lambda y.F \appsep y \appsep x$
@@
 $\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
-makes $\rho t = \lambda x.\lambda y.f \appsep x \appsep y$
+makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$
 that is the eta long form of $f$. This term is problematic since
 its leading $\lambda$ abstraction cannot justify a
-unification failure against a constant or an application.
+unification failure against a constant $f$.
 
@@
 but cannot be handled directly by the unification of the meta language, that
-is only required to handle terms in \llambda.
+is only required to handle terms in \llambda. Still, there exists a
+substitution $\rho$ such that $\rho s \in \llambda$.
 
@@
 $$
-\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\notllambda{} ~\cup~ \maybeeta{} ~\cup~ \maybebeta{})
+\wellb(X) \Leftrightarrow \forall t \in \subterm{X},
+t \not\in (\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{})
 $$
@@
 \noindent
-
-\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
+We claim our compiler validates the following property:
+
+\begin{proposition}[\wellb-enforcing]\ref{prop:w-enforcing}
+  Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
+  then 
+  $$
+    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
+    t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-w}
+  $$
+\end{proposition}
+
+\noindent
+Note that the property holds for any substitution, it could be given by an
+oracle and not necessarily a most general one, and for any terms,
+in particular terms in $\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{}$.
+
+\begin{proposition}[\wellb{}-preservation]\label{prop:w-preservation}
 $\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
@@
 \noindent
-Proposition \ref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
+Proposition \ref{prop:w-preservation} is key to prove \cref{prop:simulation,prop:fidelity}:
 informally it says that the problematic terms moved on the side by the compiler
-are not put back by \hstep, hence \Ue{} can operate properly.
-In \cref{sec:compilation,sec:eta,sec:beta}
+are not put back by \hstep, hence \Ue{} can continue to operate properly.
+In \cref{sec:simulation,sec:eta,sec:beta}
 we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
 \notllambda and how \progress takes care of them preserving \wellb
-and granting \cref{prop:simulation,prop:fidelity}.
-
-% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
+and granting \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.
+
+% Note that proposition \ref{prop:w-preservation} does not hold for \hrun as a whole
 % since decompilation can introduce (actually restore) problematic terms.
@@
 
-% \noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
+% \noteD[inline]{Below the proof of \cref{prop:w-preservation}, ho usato 3 lemmi ausiliari,
 % forse si può compattare in una prova più piccola?}
@@
   Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
-  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
+  \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
   \progressetaleft, if 1) \lhs is a name, a constant or an
@@
 \begin{lemma}
-  The introduction of \linketa guarantees \fullref{prop:nf}
+  The introduction of \linketa guarantees \fullref{prop:w-preservation}
   \label{lemma:unif-wellb}

### 2024-04-10 paper/main.tex
fix

@@
 
-\section{Ground work for the compiler}
+\section{Basic compilation and simulation}
+\label{sec:simulation}
+
+\subsection{Memory (\mapStore) and substitution ($\rho$ and $\sigma$)}
 \label{sec:grounwork}
@@
 Unification variables are identified by a
-natural number, that represents a memory addresses
+natural number that represents a memory addresses.
 The memory and its associated operations are described below:
@@
 Since each occurrence of a \Ho unification variables has a scope,
-its solution needs to be abstracted over it to enable the
-instantiation of a single assignment to different scopes.
+its assignment needs to be abstracted over it to enable the
+instantiation of the same assignment to different scopes.
 This is expressed by the \elpiIn{inctx} container, and in particular
@@
 the one of \Ho.
+
 The compiler establishes a mapping between variables of the two languages.
@@
 
-\begin{invariant}[Unification-variable arity]
+\begin{invariant}[Unification-variable arity]\label{inv:uvaarity}
   Each variable \elpiIn{A}
-  in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
-  \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
-  \label{inv:uvaarity}
+  in \Ho has a (unique) arity \elpiIn{N} and each occurrence
+  \elpiIn{(uva A L)} is such that \elpiIn{L} has length \elpiIn{N}.
 \end{invariant}
@@
 \linketa; this detail is discussed in section \ref{sec:invariant1}.
+
+Applying the substitution corresponds to dereferencing a term with respect to
+the memory.
+%  To ease the comparison we split \Fo{} dereferencing into a
+% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
+% to memory cells that are set with their values, and has a corresponding
+% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
+% has no corresponding operation in \Ho, and only ensures that
+% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
+% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
+% and \elpiIn{L2}. The reasons for this asymmetry is
+% that an \elpiIn{fapp} node with a flexible head is always mapped
+% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
+% preventing nested applications to materialize.
+% \input{code/fderef}
+It is worth looking at the code fpr \Ho{} to remark how assignments are moved
+to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
+with the names in the scope of the unification variable occurrence.
+
+\input{code/deref}
+
+\noindent
+Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
+of the arguments of all occurrences of a unification are the same. Hence
+they have the same type in the meta-level and the number of \elpiIn{abs}
+nodes in the assignment matches that length. In turn this grants that
+\elpiIn{move} never fails.\todo{convoluted}
+
+\input{code/move}
+
+We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
+<<\elpiIn{abs x\abs y\y}>> and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
+for <<\elpiIn{lam x\lam y\y}>>.
+
+\subsection{Links (\linkStore)}
 
@@
 \input{code/comp_links}
+\todo{rename link beta to link llam?}
 
@@
 This is discussed in \cref{sec:eta} and \cref{sec:beta}.
-
-% \input{code/beta_fo}
-
-% \noindent
-% The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
-% variables).\footnote{Elpi provides it as a builtin, but one could implement it by
-% systematically loading the hypothetical rule \elpiIn{name x} every time
-% a nominal constant is postulated via \elpiInFN{pi x\ }}
-% The choice of using n-ary application, rather than binary, is to make it
-% easy to access the application's head. The price we pay is that substituting
-% an application in the head of an application should be amended by
-% ``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
-% \footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
-% we do in this presentation and, as we explain later,
-% no equivalent of \elpiIn{napp} is needed in \Ho.}
-% Finally note that the cut operator is inessential, it could be
-% removed at the cost of a verbose test on the head of \elpiIn{L}
-% in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
-% can be \elpiIn{fcon}, \elpiIn{flam} or a name.
-
-Applying the substitution corresponds to dereferencing a term with respect to
-the memory. To ease the comparison we split \Fo{} dereferencing into a
-\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
-to memory cells that are set with their values, and has a corresponding
-operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
-has no corresponding operation in \Ho, and only ensures that
-terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
-<<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
-and \elpiIn{L2}. The reasons for this asymmetry is
-that an \elpiIn{fapp} node with a flexible head is always mapped
-to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
-preventing nested applications to materialize.
-
-\input{code/fderef}
-
-\noindent
-Applying the substitution in \Ho{} is very similar, with
-the caveat that assignments have to be moved to the
-current scope, i.e. renaming the \elpiIn{abs}-bound variables
-with the names in the scope of the unification variable occurrence.
-
-\input{code/deref}
-
-\noindent
-Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
-of the arguments of all occurrences of a unification variable and the
-number of abstractions in its assignment have to match. In turn
-this grants that \elpiIn{move} never fails.
-
-\input{code/move}
-
-\subsection{Notational conventions}
-
-\noindent
-When variables $x$ and $y$ can occur in term $t$ we shall write
-$t_{xy}$ to stress this fact.
-
-We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
-\elpiIn{abs x\abs y\y} and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
-for \elpiIn{lam x\lam y\y}.
 
@@
 
-\todo{notation for problem compilation}
+\subsection{Notational conventions}\todo{move away}
+
+When variables $x$ and $y$ can occur in term $t$ we shall write
+$t_{xy}$ to stress this fact. 
+\todo{add notation for problem compilation}
 
@@
 
-\section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
-\label{sec:simulation}
-
-In this section we describe a basic compilation scheme that we refine
-later, in the following sections. This scheme is sufficient to implement
-a \hstep that respects $\beta$-conversion for terms in \llambda.
-The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
-the support for terms outside \llambda in \cref{sec:beta}.
+% \section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
+
+% In this section we describe a basic compilation scheme that we refine
+% later, in the following sections. This scheme is sufficient to implement
+% a \hstep that respects $\beta$-conversion for terms in \llambda.
+% The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
+% the support for terms outside \llambda in \cref{sec:beta}.
 
@@
 
-\subsection{Limitations of by this basic scheme}
-\label{sec:basic-comp-limitations}
-The basic compilation scheme is not about to
-deal wit the following problem:
-\printAlll
-  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,Y}}}
-  {{}}
-  {{}}
-  {{}}
-% \begin{gather}
-% \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
-% \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
-% % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
-% \end{gather}
-
-\noindent Note that here $X$ is used with different arities, moreover
-in the second problem the left hand side happens to be an
-eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
-that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
-Both problems are addressed in the next two sections.
+% \subsection{Limitations of by this basic scheme}
+% \label{sec:basic-comp-limitations}
+% The basic compilation scheme is not about to
+% deal wit the following problem:
+% \printAlll
+%   {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,Y}}}
+%   {{}}
+%   {{}}
+%   {{}}
+% % \begin{gather}
+% % \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
+% % \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
+% % % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
+% % \end{gather}
+
+% \noindent Note that here $X$ is used with different arities, moreover
+% in the second problem the left hand side happens to be an
+% eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
+% that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
+% Both problems are addressed in the next two sections.
+
+\section{Handling of \maybebeta}\label{sec:llam}
+
+\subsection{Detection}
+\subsection{Compilation and decompilation}
+\subsection{Progress}
 

### 2024-04-10 paper/main.tex
add todo

@@
 
-We write 
+We write \todo{meglio dire che sigma' estende sigma, togliere unione e quindi dominio da sopra}
 $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
@@
 
+\todo{spiegare llam (first use after intro)}
 Even if we provide an implementation of the object-language unification
@@
 \subsection{The problem: logic-program Simulation}
-
+\todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
 We represent a logic program \emph{run} in \Fo{} as
-a list \emph{steps} $p$ of length $\mathcal{N}$.
+a list \emph{steps} of length $\mathcal{N}$.
 At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
@@
 each \Fo{} term in \foUnifPb into a \Ho{} term $t$.
-We write this step $\C{s}{t}{m}{l}$. The implementation of the compiler
+We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
 is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
@@
   \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
-  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
+  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
   \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
@@
 \begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
-In the context of \hrun, if $\sigma_{p-1} \hoUnifPb_{p} \in \llambda$ then
+In the context of \hrun, if 
+$\rho_{p-1} \foUnifPb_{p} \in \llambda$ 
+(even if $\;\foUnifPb_{p} \not\in \llambda$)
+then
 $$
@@
 making the second problem re-enter \llambda.
+\todo{say that this is runtime, while the other fidelity is compile time}
 
@@
 
-\begin{proposition}[\wellb-enforcing]\ref{prop:w-enforcing}
+\begin{proposition}[\wellb-enforcing]\label{prop:w-enforcing}
   Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
@@
 \noindent
+\todo{dire perchè W(T), dire che il converse non vale (alcuni check li fa la progress)}
 Note that the property holds for any substitution, it could be given by an

### 2024-04-10 paper/main.tex
w
remove trivial decomp

@@
 
-\subsection{Memory (\mapStore) and substitution ($\rho$ and $\sigma$)}
+\subsection{Memory map (\mapStore) and substitution ($\rho$ and $\sigma$)}
 \label{sec:grounwork}
@@
 
-\noteE{What is commit-links and complete-mapping?}
-
-\noindent
-Decompiling an assignment requires to turn abstractions into
-lambdas. For aesthetic purposes we also eta-contract the result
-(not needed since \Fo{} equality can do that)
-
-\input{code/decompm}
-
-\noindent
-Finally decompiling a term is trivial, now that we have an extended
-mapping containing all unassigned variables \Ue may have introduced.
-
-\input{code/decomp}
-
-\noindent
-Note that we use beta to build fapp nodes when needed (if Ag is empty
-no \elpiIn{fapp} node should appear).
+\todo{commit links + complete-map + abs to lam}
+% \noteE{What is commit-links and complete-mapping?}
+
+% \noindent
+% Decompiling an assignment requires to turn abstractions into
+% lambdas. For aesthetic purposes we also eta-contract the result
+% (not needed since \Fo{} equality can do that)
+
+% \input{code/decompm}
+
+% \noindent
+% Finally decompiling a term is trivial, now that we have an extended
+% mapping containing all unassigned variables \Ue may have introduced.
+
+% \input{code/decomp}
+
+% \noindent
+% Note that we use beta to build fapp nodes when needed (if Ag is empty
+% no \elpiIn{fapp} node should appear).
 

### 2024-04-10 paper/main.tex
rework 6 title

@@
 \noindent
+We write $\wellb(t)$ as a short for $\wellb(\{t\})$.
 We claim our compiler validates the following property:
@@
 to fail because of occur check. This check is needed in order to
-guarantee \cref{prop:fidelity} (simulation fidelity).
+guarantee \fullref{prop:fidelity}.
 
@@
 \begin{lemma}
-  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb{t}$.
+  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb(t)$.
   \label{lemma:unif-eta-aux}
@@
 
-\begin{lemma}[Fidelity with \linketa]
+\begin{theorem}[Fidelity in \maybeeta]
   The introduction 
@@
   \label{lemma:fidelity-maybeeta}
-\end{lemma}
+\end{theorem}
 
@@
 
-\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
+%\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
+\section{Making \mapStore a bijection}
 \label{sec:invariant1}
-
+\todo{l'esempio è sparito}
 We report here the problem given in \cref{sec:basic-comp-limitations} where
 $X$ is used with two different arities and the output of the compilation does
-not respect \cref{inv:map-bijective}. Moreover, merging the two mappings\noteD{TODO: da cambiare 488 in tex}
+not respect \cref{inv:map-bijective}. Moreover, merging the two mappings
 for $s$ would break \fullref{inv:uvaarity}.

### 2024-04-10 paper/main.tex
todo

@@
 \paragraph{Decompilation}
-During \elpiIn{progress}, as claimed in \cref{inv:beta-rhs}, the decompilation
+During \elpiIn{progress}, as claimed in \fullref{inv:beta-rhs}, the decompilation
 can only have \linkbeta with not instantiated \lhs. In this case, \lhs is
 unified with \rhs.
-\noteD[inline]{not really sure of this, we can have $F\appsep a = \lambda x.G x$.
-In this case when do we fail: for sure in decompile. But to respect fidelity,
-we should fail immediately: we have a \linkbeta and a \linketa with same \lhs}
+\todo{qui falliamo, nella 7.3 dire che LHS è in llambda quindi LSH = Ugly può funzionare,
+in Elpi funziona però in questi casi il termine in llam deve essere a sinistra}.
 

### 2024-04-10 paper/main.tex
explain diamond

@@
 ``problematic'' sub-terms.\\
-We now define ``problematic'' and ``well behaved'' more formally.
+We now define ``problematic'' and ``well behaved'' more formally
+and we pick the $\Diamond$ symbol since it
+stands for ``possibly'' in modal logic and all problematic terms
+are characterized by some ``uncertainty''.
 
@@
 new heuristics.
+\todo{cita paper su estensioni a llam via sospensione etc}
+\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
 

### 2024-04-10 paper/main.tex
sections 3 and 4 are a draft

@@
 previous sections).
+\todo{explain fold6}
 
@@
 
-\subsubsection{Compilation of terms in \maybebeta}
-
-The following rule is inserted just before rule~\ref{rule:compapp}.
-
-\input{code/comp_base_beta}
-
-\noindent
-Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
-is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
-(see below) grants this property. Also note that this rule generates no links.
-
-\noindent
-The only detail worth discussing is the fact that the procedure updates a
-substitution, rather than just crafting one as presented in
-section~\ref{sec:problem-statement}. The reason is that the algorithm folds
-over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
-
 \subsection{Execution}
@@
 
-\noindent
+\subsubsection{Progress}
 In the base compilation scheme \elpiIn{progress1} is the identity
@@
 
+\subsubsection{Occur check}
 Since compilation moves problematic terms out of the sigh of \Ue{},
@@
 
-Decompiling the substitution requires to first force the
-progress of links and then allocating new unassigned variables
-in the substitution for \Fo{} and finally decompiling all
-assignments. Note that \cref{inv:linklhs} and the
-occur check allows us to update the subst.
-
-\input{code/decompile}
-
-\todo{commit links + complete-map + abs to lam}
+Decompiling the substitution involves three steps.
+
+First and foremost problematic terms stored in
+\linkStore have to be moved back into the game.
+Since links are of the form uvar = term (\fullref{inv:linklhs})
+and are duplicate free (see dedup-beta dedup-eta),
+one can turn a link X = t into an assignment $X \mapsto t$.
+This can in general be achieved by unifying X with t. The case
+where t is not in \llambda (link beta/llam) is discussed in section xx.
+
+The second step amounts at allocating new variables in the
+memory of \Fo{}. In particular some unif problems
+such as $F x y = F x z$ requires to allocate a 
+variable $G$ so that the assignment
+$F_{ab} \mapsto G_a$ can be used to perform
+required pruning.
+
+The last step amounts at decompiling each assignment.
+Decompiling a term is trivial. An assignment has an
+abs node, as in move, can be eliminated by replacing
+the bound variable by the actual term in scope.
+In order to do this, one needs the \mapStore to be
+a bijection. This is the job of \cref{sec:invariant1}.
+\todo{polish}
+
+\todo[inline]{dire che però si passa per una subst in cui ste abs
+le cambio in lam. Nel codice Coq ci scrivevamo il tipo nella
+arity, e quindi sappiamo fare i lambda bene, senza perdita di
+informazione. Qui i lam non hanno info, facile. Ma in generale
+bisogna spiegare come ci si salva. Ci dormo su: o non generiamo
+la subst ma solo il primo termine (la query iniziale) istanziato
+(funziona sempre, la prova è quella sopra) oppure bisogna siegare
+tutto sto casino e serve un po' di spazio.}
+
+%\input{code/decompile_type}
+
 % \noteE{What is commit-links and complete-mapping?}
@@
 
-\begin{invariant}
-
-  TODO: dire che il mapping è bijective
-  \label{inv:map-bijective}
-\end{invariant}
-
 \subsection{Definition of \Uo and its properties}\label{sec:founif}
@@
 
-The code given so far applies to terms in $\beta\eta$-normal form where
-unification variables in \Fo{} can occur non linearly but always with
-the same number of arguments, and where their arguments are distinct names
-(as per \llambda).
+The code given so far still makes no use
+of the higher order nature of the ML unif language,
+indeed the scope of unif variables generated by the
+compiler is always empty, so \Ue{} is first order.
+
+Still, if \foUnifPb is already \wellb, we can set up a proof
+that will also work when comp enforces \wellb and  hstep
+preserves it, and when terms in \llambda are mapped to ho variables
+with a scope.
 
@@
 \begin{proof}[Proof sketch]
-trivial, since the terms are beta normal beta just builds an app. \noteD{Reformulate?}
+trivial if the mapping is a bijection and 
+the terms are beta normal.
+some discussion about commit maybellam to be done later.
 \end{proof}
@@
 a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
-If we look at the \Fo{} terms, the are two interesting cases:
+If we look at the \Fo{} terms is only one interesting cases:
 \begin{itemize}
@@
   $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
-\item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
- we have $Y_{\vec{x}} \Ue t$ that succeeds with
- $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
- is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
- Thanks to \ref{clause:beta1}
- $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
 \end{itemize}
@@
 
-\begin{lemma} \Fullref{prop:simulation} and
+\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
 \fullref{prop:fidelity} hold
-\end{lemma}
+\end{theorem}
 \begin{proof}[Proof sketch]
@@
 in this
-context where input terms are $\beta\eta$-normal and we disregard $\eta$-equivalence
+context where input terms are $\beta\eta$-normal and all is \wellb,
 \Ue is equivalent to \Uo.
@@
 
-\subsection{Detection}
+\paragraph{Detection} trivial, pattern-fragment.
 \subsection{Compilation and decompilation}
-\subsection{Progress}
-
+
+The following rule is inserted just before rule~\ref{rule:compapp}.
+
+\input{code/comp_base_beta}
+
+\noindent
+Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
+is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
+(see below) grants this property. Also note that this rule generates no links.
+
+\noindent
+The only detail worth discussing is the fact that the procedure updates a
+substitution, rather than just crafting one as presented in
+section~\ref{sec:problem-statement}. The reason is that the algorithm folds
+over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
+
+\paragraph{Decompilation} since no link, nothing new
+
+\paragraph{Progress} since ML unif is complete here, no
+need to move terms aside, just use uva is enough.
+
+\begin{lemma}
+  \Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo
+  in \cref{sec:founif}
+  \end{lemma}
+  \begin{proof}[Proof sketch]
+  If we look at the \Fo{} terms, the is one more case interesting cases:
+  \begin{itemize}
+  \item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
+   we have $Y_{\vec{x}} \Ue t$ that succeeds with
+   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
+\todo{ste x escono da L}   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
+   Thanks to \ref{clause:beta1}
+   $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
+  \end{itemize}
+  \end{proof}
+
+\begin{lemma}[\wellb-enforcement]\label{lem:w-enforcement-maybebeta} Even if $\foUnifPb \cap \maybebeta \ne \emptyset$,
+  $\hoUnifPb \cup \maybebeta = \emptyset$
+\end{lemma}
+\begin{proof}[Proof sketch]
+  problematic terms are mapped to uva by comp, the problematic fapp node is gone.
+\end{proof}
+
+\begin{theorem}[Fidelity in \maybebeta] \Fullref{prop:simulation} and
+\fullref{prop:fidelity} hold
+\end{theorem}
+\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
+  it is the same as in \cref{sec:simulation}, even if now
+  we really need \Ue{} to deal with \llambda, while before a FO unif
+  would have done.
+\end{proof}
 % \subsection{Example}
@@
 \end{lemma}
+\todo{is this W-enforcing?}
 

### 2024-04-11 paper/main.tex
wip

@@
 To accommodate this situation the compiler wraps \elpiIn{baselink} using
-the \elpiIn{inctx} container (see, \ref{data:inctx}).
+the \elpiIn{inctx} container (see rule~\ref{data:inctx}).
 
@@
 % \noteD[inline]{i don't understand the note}
+\noteD{Cambiare esempio riprendendo quello di enrico}
 On the other hand, we see that, 
@@
 
-\noteE[inline]{Dire che maybe eta fa il detect anche su 
-termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
-esempio?}
-
 \paragraph{Decompilation}
@@
 \newcommand{\progBetaDedup}{\emph{progress-beta-dedup}\xspace}
-\newcommand{\progBetaLH}{\emph{progress-rigid-\lhs}\xspace}
-
-The activation of a \linkbeta is performed when its \rhs falls under \llambda
-under a given substitution.
+\newcommand{\progBetaLH}{\emph{progress-beta-rigid-\lhs}\xspace}
+
+We give $4$ different definitions of \linkbeta progression for a given \linkbeta
+\linkbetaM{\Gamma}{T}{\rhsBeta}, say $l$.
 
 \begin{definition}[\progBetaLL]
-  Given a substitution $\sigma$ and a \linkbeta \linkbetaM{\Gamma}{T}{\rhsBeta}
-  such that $\sigma t_1$ is a name, say $t$, and $t\notin s_1\dots s_n$. If $m =
-  0$, then the \linkbeta is removed and \lhs is unified with \rhsBetaHead.
-  If $m > 0$, then the \linkbeta is replaced by a refined version
-  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots t_m}$ with reduced
-  list of arguments and $Y$ being a fresh variable. Moreover, the new link
-  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to 
-  \linkStore.
+  Given a substitution $\sigma$, where $\sigma t_1$ is a name, say $t$, and
+  $t\notin s_1\dots s_n$. If $m = 0$, then $l$ is removed and \lhs is
+  unified with \rhsBetaHead. If $m > 0$, then $l$ is replaced by a
+  refined version $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots
+  t_m}$ with reduced list of arguments and $Y$ being a fresh variable. Moreover,
+  the new link \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}}
+  is added to \linkStore.
 
@@
 In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
-  \noteD{L'ho riformulato}
+  % \noteD{L'ho riformulato}
 \end{definition}
@@
 \begin{definition}[\progBetaRH]
-  A link \linkbetaM{\Gamma}{X}{\rhsBeta} is removed from
+  $l$ is removed from
   \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
   $\beta$-reduced term $t'$ obtained from the application of $t$ to
-  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified to $t$.
+  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified with $t$.
   \label{def:progBetaRH}
@@
   Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
-  share the same \lhs. The two \rhs are unified and a $l_2$ is
+  share the same \lhs. The two \rhs are unified and $l_2$ is
   removed from \linkStore.
@@
 Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
-  \noteD{da finire}
 \end{proof}

### 2024-04-11 paper/main.tex
update section map bijection

@@
 \begin{theorem}[Fidelity in \maybeeta]
-  The introduction 
-  of \linketa guarantees \fullref{prop:fidelity} 
+  Given a list of unification problems \foUnifPb, such that 
+  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$,
+  the introduction 
+  of \linketa guarantees \fullref{prop:fidelity}.
+  \footnote{We also suppose that any higher-order variable is always applied with
+  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}
   \label{lemma:fidelity-maybeeta}
@@
 \label{sec:invariant1}
-\todo{l'esempio è sparito}
-We report here the problem given in \cref{sec:basic-comp-limitations} where
-$X$ is used with two different arities and the output of the compilation does
-not respect \cref{inv:map-bijective}. Moreover, merging the two mappings
-for $s$ would break \fullref{inv:uvaarity}.
-In this section we explain how to replace the duplicate mapping with
-\linketa in order to restore the invariants.
+
+In \cref{sec:grounwork}, we introduced the definition of ``memory map'' (\mapStore).
+This memory allows to decompile the \Ho terms back to the object language.
+It is the case that, while solving unification problems, a same unification
+variable $X$ is used multiple times with different arities.
+%
 % make test ONLY=7001 TEX=tex
@@
 
-We see that the \elpiIn{maybe-eta} as identified $\lambda xy.X \appsep y \appsep
-x$ and $\lambda x.f\appsep (X \appsep x) \appsep x$ and the compiler has
-replaced them with $A$ and $D$ respectively.
-% $A$ is linked with $\lambda x.B_x$, $B$
-% has arity $1$ and is $\eta$-linked with $\lambda y.C\appsep y \appsep x$ and $D$
-% is linked to the term $\lambda x.f \appsep E_{x}~x$. 
-However, the mapping \mapStore breaks \cref{inv:map-bijective}: the \Fo{}
-variable $X$ is mapped to two different \Ho variables. To address this problem
-we adjust the compiler's output with a \elpiIn{map-deduplication} procedure.
+In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
+$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
+By \fullref{inv:uvaarity}, we are not allowed to use a same \Ho variable to 
+represent the two occurrences of $X$.
+If we execute \hrun, we remark that the unification fails.
+There is in fact a major problem:
+\hstep is not conscious of the connection between the variables $C$ and $E$ (both corresponding to $X$),
+since no link in \linkStore puts $C$ and $E$ in relation and decompilation
+does not work properly if a \Fo{} variable is mapped to two distinct \Ho variables. 
+The two main drawbacks connected to this situation are 
+firstly the lost of \fullref{prop:fidelity} and
+secondly, if we want to guarantee at least \fullref{prop:simulation}, we should
+overcomplicate the decompilation phase. In order to ease the second
+drawback, we pose the following property:
+
+\begin{proposition}[\mapStore is a bijection]
+  Given a list of unification problems \foUnifPb, then the memory map \mapStore
+  compiled from \foUnifPb is a bijection relating the \Fo{} and the \Ho{} 
+  variables.
+\end{proposition}
+
+We finally adjust the compiler's output with a \elpiIn{map-deduplication} procedure.
 
@@
 
-\newcommand{\mapdeduplication}{\emph{map-deduplication}}
+\newcommand{\mapdeduplication}{\emph{map-deduplication}\xspace}
 
@@
 \end{definition}
+
+\begin{theorem}[Fidelity with \mapdeduplication]
+  Given a list of unification problems \foUnifPb, such that 
+  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
+  if \foUnifPb contains two same \Fo{} variables with different arities,
+  then \mapdeduplication guarantees \fullref{prop:fidelity}
+\end{theorem}
+
+\begin{proof}[Proof sketch]
+  By the definition of \mapdeduplication, any two same \Fo{} variables $X_1, X_2$
+  with different arities are related with \linketa. 
+  If one of the two variables is instantiated, the corresponding
+  \linketa is triggered instantiating the related variable.
+  This allows to make unification fail if $X_1$ and $X_2$ are unified with 
+  different terms. Finally, since \foUnifPb contains only terms thar are
+  either \wellb or \maybeeta, by \cref{lemma:fidelity-maybeeta},
+  we can conclude the proof.
+\end{proof}
 
@@
 
-% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
-% This assignment makes $l_2$ to progress since the \lhs is materialized and by
-% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
-% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
-% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
-% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
-% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
-% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
-% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
-% As last step, the last link is progressed and the final \Ho substitution is
-% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
-% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
-% H _{x} \mapsto \lambda y.y\}$.
-% The decompilation phase is only charged, in this example to solve the mappings,
-% since no suspended links remain. The only mapping in the list is
-% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$
-
-% \noteE[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
-% }
 \section{Handling of \notllambda}\label{sec:beta}

### 2024-04-11 paper/appendix.tex
link-beta -> link-llam

@@
     (T1 ==l T2) S1 S2.
-  progress-beta-link-aux T1 T2 S S [@val-link-beta T1 T2] :- !.
+  progress-beta-link-aux T1 T2 S S [@val-link-llam T1 T2] :- !.
 
@@
         ho.subst -> links -> o.
-  progress-beta-link T (ho.app[ho.uva V Scope | L] as T2) S S2 [@val-link-beta T T2] :- 
+  progress-beta-link T (ho.app[ho.uva V Scope | L] as T2) S S2 [@val-link-llam T T2] :- 
     arity T Arity, len L ArgsNb, ArgsNb >n Arity, !,
@@
       (T2 = ho.app [ho.uva Ad2 Scope2 | L2], 
-      NewLinks = [@val-link-beta T T2 | LinkEta]).
+      NewLinks = [@val-link-llam T T2 | LinkEta]).
 
@@
 
-  progress-beta-link T1 (ho.app[ho.uva _ _ | _] as T2) H H [@val-link-beta T1 T2] :- !.
+  progress-beta-link T1 (ho.app[ho.uva _ _ | _] as T2) H H [@val-link-llam T1 T2] :- !.
 
@@
 
-  solve-link-abs (@val-link-beta A B) NewLinks S S1 :- !,
+  solve-link-abs (@val-link-llam A B) NewLinks S S1 :- !,
     progress-beta-link A B S S1 NewLinks.
@@
     (T1' ==l T2') H1 H2.
-  commit-links-aux (@val-link-beta T1 T2) H1 H2 :- 
+  commit-links-aux (@val-link-llam T1 T2) H1 H2 :- 
     ho.deref H1 T1 T1', ho.deref H1 T2 T2',

### 2024-04-11 paper/main.tex
link-beta -> link-llam
other rename beta to llam

@@
 
-\newcommand{\linkbeta}{\linkMacro{\beta}}
+\newcommand{\linkbeta}{\linkMacro{\llambda}}
 \newcommand{\linketa} {\linkMacro{\eta}}
@@
 \newcommand{\linketaM}[3]{\ensuremath{#1 \vdash #2 =_\eta #3}}
-\newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_\beta #3}}
+\newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_{\llambda} #3}}
 \newcommand{\substCell}[3]{\ensuremath{#1 \vdash #2 = #3}}
@@
 \input{code/comp_links}
-\todo{rename link beta to link llam?}
 
@@
 \begin{elpicode}
-abs x\ val (link-beta (uva A [x]) (app[uva F [x],con "a"]))
+abs x\ val (link-llam (uva A [x]) (app[uva F [x],con "a"]))
 \end{elpicode}
@@
 \printAlll
-  {{{X,\lambda x.Y},
+  {{{X,\lambda x.a},
     {(X\appsep a),a}}}
-  {{{A,\lambda x.B},
+  {{{A,\lambda x.a},
     {(A\appsep a),a}}}
-  {{{Y,B,0},
-    {X,A,0}}}
+  {{{X,A,0}}}
   {{}}
@@
 $\foUnifPb_2$ can be solved in \llambda.
-% \noteE[inline]{it is even a ground term, there is no unification left to perform actually}
-% \noteD[inline]{i don't understand the note}
-\noteD{Cambiare esempio riprendendo quello di enrico}
 On the other hand, we see that, 
@@
 fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
-\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
-non-unifiable problem $(\lambda x. B) \appsep a \nUe a$.
+\mapsto \lambda x. a\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
+non-unifiable problem $(\lambda x. a) \appsep a \nUe a$.
 
@@
 
-\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}\xspace}
-\newcommand{\progBetaRH}{\emph{progress-beta-rigid-\rhs}\xspace}
-\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}\xspace}
-\newcommand{\progBetaLH}{\emph{progress-beta-rigid-\lhs}\xspace}
+\newcommand{\progBetaLL}{\emph{progress-\llambda-refine}\xspace}
+\newcommand{\progBetaRH}{\emph{progress-\llambda-\rhs}\xspace}
+\newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
+\newcommand{\progBetaLH}{\emph{progress-\llambda-\lhs}\xspace}
 

### 2024-04-11 paper/main.tex
wip: maybebeta section

@@
 
+\noteD{Si puo togliere se serve spazio}
 \begin{corollary}
@@
 \paragraph{Decompilation}
-During \elpiIn{progress}, as claimed in \fullref{inv:beta-rhs}, the decompilation
-can only have \linkbeta with not instantiated \lhs. In this case, \lhs is
-unified with \rhs.
-\todo{qui falliamo, nella 7.3 dire che LHS è in llambda quindi LSH = Ugly può funzionare,
-in Elpi funziona però in questi casi il termine in llam deve essere a sinistra}.
+We suppose that the unification of the object language rejects any unification
+between two terms where at least one of them is not in \llambda 
+(this includes unifications like $X = Y\appsep Z$ and $X\appsep Y = X \appsep Y$).
+Therefore, a failure is thrown if any \linkbeta remains in \linkStore at the
+begin of decompilation, i.e. all \linkbeta should be solved before decompilation.
 
@@
 \newcommand{\progBetaRH}{\emph{progress-\llambda-\rhs}\xspace}
-\newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
-\newcommand{\progBetaLH}{\emph{progress-\llambda-\lhs}\xspace}
-
-We give $4$ different definitions of \linkbeta progression for a given \linkbeta
-\linkbetaM{\Gamma}{T}{\rhsBeta}, say $l$.
+% \newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
+\newcommand{\progBetaFail}{\emph{progress-\llambda-fail}\xspace}
+
+Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
+$4$ different activation rules:
 
@@
 % Example of this `X a = Y a'
-\begin{definition}[\progBetaDedup]
-  Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
-  share the same \lhs. The two \rhs are unified and $l_2$ is
-  removed from \linkStore.
-  \label{def:progBetaDedup}
-\end{definition}
-
-\begin{definition}[\progBetaLH]
-  Given a \linkbeta with rigid \lhs, the unification fails.
-  \label{def:progBetaLH}
+% \begin{definition}[\progBetaDedup]
+%   Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
+%   share the same \lhs. The two \rhs are unified and $l_2$ is
+%   removed from \linkStore.
+%   \label{def:progBetaDedup}
+% \end{definition}
+
+\begin{definition}[\progBetaFail]
+  If it exists a link $l' \in \linkStore$ with same \lhs as $l$, or the \lhs 
+  of $l$ become rigid, then unification fail.
+  \footnote{This is a consequence of what claimed in \linkbeta decompilation}
+  \label{def:progBetaFail}
 \end{definition}
@@
   \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
-  terminates. Finally \progBetaDedup (resp. \progBetaLH) also guarantees termination since 
-  it removes a link from \linkStore (resp. immediately fails).
+  terminates. Finally \progBetaFail also guarantees termination since 
+  it makes \elpiIn{progress} immediately fails.
 
@@
 
-\begin{lemma}[Fidelity with \linkbeta]
+\begin{theorem}[Fidelity with \linkbeta]
   The introduction of \linkbeta guarantees \fullref{prop:fidelity}
-\end{lemma}
+\end{theorem}
 
@@
 
-\subsection{Relaxing \fullref{def:progBetaLH}}
+\subsection{Relaxing \fullref{def:progBetaFail}}
 
@@
 % from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
-Working with terms in \llambda is sometime too restrictive. There exists systems
-such as $\lambda$Prolog~\cite{lamProlog}, Abella~\cite{gacek2008abella}, which 
+Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}. 
+There exists systems
+such as $\lambda$Prolog~\cite{lamProlog} which 
 delay the resolution of \notllambda unification problems if the substitution
-is not able to put them in \llambda.
+is not able to put them in \llambda. For example in 
+\noteD{Teijus}
 %

### 2024-04-11 paper/main.tex
add a citation to teyjus

@@
 There exists systems
-such as $\lambda$Prolog~\cite{lamProlog} which 
+such as Teyjus\cite{Nadathur2001} and $\lambda$Prolog~\cite{lamProlog} which 
 delay the resolution of \notllambda unification problems if the substitution
-is not able to put them in \llambda. For example in 
-\noteD{Teijus}
+is not able to put them in \llambda.
 %

### 2024-04-11 paper/main.tex
Update relaxing progress-llam-fail

--- 2024-04-11 paper/main.tex
@@
 
-\newcommand{\Uo}{\ensuremath{\UnifRel_o}\xspace}
+\newcommand{\Uo}{\texorpdfstring{\ensuremath{\UnifRel_o}\xspace}{unif\_o}}
 \newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
@@
 
-\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}} % space non va
-\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}
+\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
+\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
 
@@
 
-\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
-\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
+\newcommand{\linkStore}{\texorpdfstring{\ensuremath{\mathbb{L}}\xspace}{L}}
+\newcommand{\mapStore}{\texorpdfstring{\ensuremath{\mathbb{M}}\xspace}{M}}
 \newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
@@
 }
-\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
-\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
-\newcommand{\notllambda}{\ensuremath{\Diamond\llambda}\xspace}
+\newcommand{\maybeeta}{\texorpdfstring{\ensuremath{\Diamond\eta}\xspace}{maybeeta}}
+\newcommand{\maybebeta}{\texorpdfstring{\ensuremath{\Diamond\beta_0}\xspace}{maybebeta}}
+\newcommand{\notllambda}{\texorpdfstring{\ensuremath{\Diamond\llambda}\xspace}{maybellam}}
 %\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}
@@
 
-\subsection{The problem: logic-program Simulation}
+\subsection{The problem: logic-program simulation}
 \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
@@
 
-\subsection{Memory map (\mapStore) and substitution ($\rho$ and $\sigma$)}
+\subsection{Memory map (\mapStore) and substitution (\texorpdfstring{$\rho$ and $\sigma$}{rho and sigma})}
 \label{sec:grounwork}
@@
 
-\subsection{Definition of \Uo and its properties}\label{sec:founif}
+\subsection{Definition of \Uo{} and its properties}\label{sec:founif}
 
@@
 \section{Handling of \notllambda}\label{sec:beta}
+\noteD[inline]{Dire che OL fallisce tutte le unif not in llambda, e poi modificare 
+le progress e la decomp}
 
@@
 
-\subsection{Relaxing \fullref{def:progBetaFail}}
+\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
 
@@
 There exists systems
-such as $\lambda$Prolog~\cite{lamProlog} which 
+such as Teyjus~\cite{Nadathur2001} and $\lambda$Prolog~\cite{lamProlog} which 
 delay the resolution of \notllambda unification problems if the substitution
-is not able to put them in \llambda. For example in 
-\noteD{Teijus}
+is not able to put them in \llambda.
+
+In this section we want to show how we can adapt the unification of the object
+language in the meta language by simply adding (or removing) rules to the
+\elpiIn{progress} predicate.
 %
@@
   {{{(X\appsep a),a},
-    {X,\lambda x.Y}}}
+    {X,\lambda x.a}}}
   {{}}
@@
 
-In the example above, $\foUnifPb_1$ is in \notllambda and the object language
-cannot solve it, and, by \cref{prop:fidelity}, the meta language neither.
-However, we can be more permissive, and relax \cref{lemma:keep-fidelity}. This
-modification is quite simple to manage: we are introducing a new \notllambda
-progress rule, say \progBetaNoLLWait, by which, if \lhs is rigid and \rhs is
-flexible, the considered \linkbeta is kept in the store and no progression is
-done\footnote{This new rule trivially guarantees the termination of progress}.
-\progBetaNoLLWait makes \elpiIn{occur-check-links} partial, since the check is
-possible only on links with a variable on the \lhs. This means that we can have
-two links \linkbetaM{}{X}{Y\appsep a} and \linkbetaM{}{f\appsep X}{Y\appsep a}
-where the occur check does not throw an error. Note however, that the
-decompilation of the two links will force the unification of $X$ to $Y\appsep a$
-and then the unification of $f\appsep (Y \appsep a)$ to $Y\appsep a$, which
-fails by the occur check of \Ue.
-
-A second strategy to deal with problem that are in \notllambda is to make some
-approximation. This is the case for example of the unification algorithm of Coq
+In the example above, $\foUnifPb_1$ is in \notllambda. \emph{If} the object
+language delays the first unification problem waiting $X$ to be
+be instantiated in a future unification, we can relax \cref{def:progBetaFail}. 
+Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
+we keep it in \linkStore until the head of its \rhs also become rigid.
+In this case, since \lhs and \rhs have rigid heads, they can be unified
+just before removing $l$ from \linkStore. We can note that this rule trivially
+guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
+partial: there exists \linkbeta with a non-flexible \lhs.
+
+A second strategy to deal with problem that are in \notllambda is to make
+approximations. This is the case for example of the unification algorithm of Coq
 used in its type class solver~\cite{sozeau08}. The approximation consists in
-forcing a choice (among the others) when the unification problem is in
-\notllambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
+forcing a choice (among the others) when the unification problem is outside
+\llambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
 argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
 that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
-\_\}$ is another valid substitution for the original problem. This approximation
-can be easily introduced in our unification procedure, by adding new custom 
-\linkbeta progress rules.
+\_\}$ is another valid substitution for the original problem. We stress the 
+fact that, again, our unification procedure in the meta language can be 
+accommodated for this new behavior: given a \linkbeta, if \lhs is not in \llambda,
+then \elpiIn{progress} can try to align the rightmost arguments and unify
+the resulting heads.
+
+Note that delaying unification outside \llambda can leave \linkbeta during the
+decompilation phase. Therefore, new rules to \elpiIn{commit-links} should
+be added accordingly.
 
 % \noteD{Paragraph for decompile in case of \maybebeta}
-Decompilation of \linkbeta is possible by extending \elpiIn{commit-link} with
-new heuristics.
-\todo{cita paper su estensioni a llam via sospensione etc}
-\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
+% \todo{cita paper su estensioni a llam via sospensione etc}
+\todo[inline]{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
 

### 2024-04-11 paper/main.tex
small update

@@
 \section{Handling of \notllambda}\label{sec:beta}
-\noteD[inline]{Dire che OL fallisce tutte le unif not in llambda, e poi modificare 
-le progress e la decomp}
+
+In this section we suppose the unification of the object language between two
+terms $t_1$ and $t_2$ to fail each time at least one of the between $t_1$ or
+$t_2$ is outside $\llambda$. This means for instance that $X \nUo Y \appsep Z$ and 
+$X\appsep Y \nUo X\appsep Y$.
 
@@
 \paragraph{Decompilation}
-We suppose that the unification of the object language rejects any unification
-between two terms where at least one of them is not in \llambda 
-(this includes unifications like $X = Y\appsep Z$ and $X\appsep Y = X \appsep Y$).
-Therefore, a failure is thrown if any \linkbeta remains in \linkStore at the
+A failure is thrown if any \linkbeta remains in \linkStore at the
 begin of decompilation, i.e. all \linkbeta should be solved before decompilation.
@@
   of $l$ become rigid, then unification fail.
-  \footnote{This is a consequence of what claimed in \linkbeta decompilation}
   \label{def:progBetaFail}

### 2024-04-11 paper/main.tex
update sec maybebeta0

@@
 
-\paragraph{Detection} trivial, pattern-fragment.
+A first problem we encounter when making unification between terms that are
+well behaved is the need to treat higher-order variables. 
+%
+% make test ONLY=7006 TEX=tex
+\printAlll
+  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
+  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
+  {{{X,A,0}}}
+  {{}}
+
+In the example above, we can note that the very basilar compilation given in 
+the previous section is not able to make the \Ho unification problem succeeds.
+The unification of $T_1$ fails while trying to unifying $A \appsep x$ and $x$.
+This is due to the fact that $A \appsep x$ (equivalent to \elpiIn{app[uva A [], x]})
+is represented as the application of the variable $A$ to the name $x$.
+In order to exploit the higher-order unification algorithm of the meta language,
+we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.
+
 \subsection{Compilation and decompilation}
 
-The following rule is inserted just before rule~\ref{rule:compapp}.
+In order to address this problem, we add the following rule before rule~\ref{rule:compapp}.
 
@@
 
-\paragraph{Decompilation} since no link, nothing new
-
-\paragraph{Progress} since ML unif is complete here, no
-need to move terms aside, just use uva is enough.
+\paragraph{Decompilation} Since no link is created by the compilation of 
+\maybebeta terms, no modification should be done
+to the \elpiIn{commit-link} predicate.
+
+\paragraph{Progress} Similarly to decompilation, since no link is produced,
+no modification to the \elpiIn{progress} predicate is needed.
 

### 2024-04-12 paper/main.tex
more work on abstract

@@
   dedicated, high-level Meta Languages (ML) to facilitate this task in two
-  key ways: 1) variable binding and substitution are simplified when ML binders
+  key ways: 1) variable binding and substitution are for free when ML binders
   represent object logic ones; 2) proof construction, and even proof search, is
@@
 
-  The object logic we are interested in is Coq's~\cite{Coq-refman}
-  Calculus of Inductive Constructions (CIC),
-  for which we aim to implement a unification procedure \Uo using the ML
+  The object logic we are interested in is % an extension of the latter,
+  Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
+  we want to implement a unification based 
+  proof search procedure using the ML
   Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
@@
   fragment~\cite{miller92jsc}.
-  We want \Uo to be as powerful as \Ue but on the object logic CIC.
   Elpi also comes with an encoding for CIC that works well
@@
   ``underuses'' \Ue by restricting it to first-order unification problems only.
-  To address this issue, we propose a better-behaved encoding, \Ho,
-  demonstrate how to map unification problems in \Fo{}
-  to related problems in \Ho, and illustrate
-  how to map back the unifiers found by \Ue, effectively implementing
-   \Uo on top of \Ue for the encoding \Fo.
-
-  % We apply this technique to the implementation of a type-class~\cite{wadler89}
-  % solver for Coq~\cite{Coq-refman}.
-  % Type-class solvers are proof search procedures based on
-  % unification that back-chain designated lemmas, providing essential
-  % automation to widely used
-  % Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-  % and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
-  % constitute our test bed.
+
+  In this paper we propose a better-behaved encoding, \Ho,
+  and show how to map unification problems in \Fo{}
+  to related problems in \Ho.
+    As a result we can lift the ML
+  higher-order unification procedure to \Fo{} in a way that it
+  honours $\eta\beta$-equivalence (for CIC functions), solves
+  problems in the pattern fragment and finally allows
+  for the addition of heuristics to deal with problems outside the
+  pattern fragment. The unification procedure we build can be used to
+  effectively simulate a logic program in \Fo{} by re-using optimizations
+  provided by the ML such as clause indexing.
+
 \end{abstract}

### 2024-04-12 paper/main.tex
reword

@@
   Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
-  we want to implement a unification based 
+  we want to implement a higher-order unification-based 
   proof search procedure using the ML
@@
   $\eta\beta$ equivalence and comes equipped with a
-  higher order unification procedure \Ue restricted to the pattern
+  higher-order unification procedure \Ue restricted to the pattern
   fragment~\cite{miller92jsc}.
-  Elpi also comes with an encoding for CIC that works well
-  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
-  Unfortunately this encoding, which we refer to as \Fo,
-  ``underuses'' \Ue by restricting it to first-order unification problems only.
-
-  In this paper we propose a better-behaved encoding, \Ho,
+  Elpi comes with an encoding of CIC that works well
+  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
+  but restricts \Ue to roughly first-order unification problems only.
+  We call this basic encoding \Fo.
+
+  In this paper we propose a better-behaved encoding \Ho,
   and show how to map unification problems in \Fo{}
@@
   pattern fragment. The unification procedure we build can be used to
-  effectively simulate a logic program in \Fo{} by re-using optimizations
+  efficiently simulate a logic program in \Fo{} by re-using optimizations
   provided by the ML such as clause indexing.
@@
 In this paper we focus on one aspect of
-this work, precisely \emph{how to reuse the higher order unification procedure
-of the meta language in order to simulate a higher order logic program
+this work, precisely \emph{how to reuse the higher-order unification procedure
+of the meta language in order to simulate a higher-order logic program
 for the object language}.
@@
 
+In this paper we show how to lift the meta language higher-order
+unification procedure to the object language. The resulting
+unification honours $\eta\beta$-equivalence for
+
 Benefits: less work, reuse efficient ho unif (3x faster), indexing,

### 2024-04-12 paper/main.tex
if only I was a native speaker...

@@
   to related problems in \Ho.
-    As a result we can lift the ML
-  higher-order unification procedure to \Fo{} in a way that it
-  honours $\eta\beta$-equivalence (for CIC functions), solves
-  problems in the pattern fragment and finally allows
-  for the addition of heuristics to deal with problems outside the
-  pattern fragment. The unification procedure we build can be used to
-  efficiently simulate a logic program in \Fo{} by re-using optimizations
-  provided by the ML such as clause indexing.
+    As a result we obtain \Uo{}, a higher-order unification procedure for \Fo{}
+    that honours $\eta\beta$-equivalence (for CIC functions), solves
+  problems in the pattern fragment and allows
+  for the use of heuristics to deal with problems outside the
+  pattern fragment. Moreover, since \Uo{} delegates most of the work to \Ue,
+  it can be used to efficiently simulate a logic program in \Fo{} by
+  taking advantage of
+  unification-related optimizations of the ML, such as clause indexing.
 

### 2024-04-12 paper/main.tex
finish intro

@@
 reminiscent of Prolog that back-chain lemmas taken
-from a designated database of ``type class instances''. Given this
+from a database of ``type-class instances''. Given this
 analogy with Logic Programming we want to leverage the
@@
 \coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
-The following three designated type-class instances state that:
+The following three type-class instances state that:
 1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
@@
 called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
-solution $\customlabel{solution:intro}{\sigma}$ that assigns
+solution $\customlabel{solution:intro}{\rho}$ that assigns
 the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
@@
 In order to implement such a search in Elpi we shall describe the encoding
-of CIC terms and then the encoding of instances as rules (a.k.a. clauses).
+of CIC terms and then the encoding of instances as rules.
 Elpi comes equipped with
@@
 \noindent
-Following $\lambda$Prolog~\cite{miller_nadathur_2012}'s standard syntax,
+Following the standard syntax of $\lambda$Prolog~\cite{miller_nadathur_2012}
 the meta level binding of a variable \elpiIn{x} in an expression
-\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets delimit a
+\elpiIn{e} is written <<\elpiIn{x\ e}>>, while square brackets delimit a
 list of terms separated by comma. For example the term
@@
 \noindent
-Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work,
+Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work
 since it uses the predicate \coqIn{P} as a first order term: for the meta
@@
 \begin{elpicode}
-decision (all (app [con"fin", con"7"]) y\              ~\customlabel{goal:g}{(g)}~
-  app [con"nfact", y, con"3"]).
+decision (all (app [con"fin", con"7"]) x\              ~\customlabel{goal:g}{(g)}~
+  app [con"nfact", x, con"3"]).
 \end{elpicode}
@@
 \begin{elpicode}
-app [con"nfact", y, con"3"] = app [P, y]               ~\customlabel{problem:p}{(p)}~
+app [con"nfact", x, con"3"] = app [P, x]               ~\customlabel{problem:p}{(p)}~
 \end{elpicode}
@@
 \begin{elpicode}
-app [con"nfact", y, con"3"] = Pm y                    ~\customlabel{problem:pa}{(p')}~
-Pm = x\ app [con"nfact", x, con"3"]                   ~\customlabel{solution:pm}{(\rho)}~
+app [con"nfact", x, con"3"] = Pm x                    ~\customlabel{problem:pa}{(p')}~
+Pm = x\ app [con"nfact", x, con"3"]                   ~\customlabel{solution:pm}{(\sigma)}~
 \end{elpicode}
@@
 (Coq) $\beta$-redex in the second premise (the predicate
-under the \elpiIn{pi w\ }\hspace{-0.4em}). We show below the premise before and
-after the instantiation of \elpiIn{P}:
-
-\begin{elpicode}
-decision (app [                   P                  , w])
-decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
-\end{elpicode}
-
-\noindent
+under the \elpiIn{pi w\ }\hspace{-0.4em}).
+% We show below the premise before and
+% after the instantiation of \elpiIn{P}:
+
+% \begin{elpicode}
+% decision (app [                   P                  , w])
+% decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
+% \end{elpicode}
+
+% \noindent
 In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
@@
 \begin{elpicode}
-app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
-app[ con"nfact"                             , N, NF]
+app [ lam A (a\ app [con"nfact", a, con"3"]) , x] =
+app [ con"nfact"                             , N, NF]
 \end{elpicode}
-\noindent
-~\\
+
+\noindent
 The root cause of the problems we sketched in this example
@@
 and the object language, that in turns makes the 
-unification procedures of the meta language weak.\\
+unification procedures of the meta language weak.
 The equational theory of the meta language Elpi encompasses
-$\eta\beta$-equivalence and its unification procedure can solve higher order
+$\eta\beta$-equivalence and its unification procedure can solve higher-order
 problems in the pattern fragment. Although 
@@
 
-% For example:\\
-% %
-% \vspace{4pt}
-% {
-% \setlength{\tabcolsep}{2pt}
-% \begin{tabular}{lcl|}
-% \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
-% \elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
-% \elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
-% \end{tabular}
-% \phantom{mn}
-% \begin{tabular}{lcl}
-% \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
-% \elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
-% \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
-% \end{tabular}
-% }
-% \vspace{4pt}
-%
 \paragraph{Contributions}
@@
 sketched in the introduction can be formally described.
-We detail an encoding of a logic program on \Fo{} to a strongly related
-logic program in \Ho (the language of the meta language) and we show that
-the unification procedure of the meta language \Ue{} can be effectively
-used to simulate a unification procedure \Uo for the object language that
-features $\eta\beta$-conversion in the pattern-fragment.
-
-\todo{fix}\cref{sec:problem-statement} formally states the problem and gives the
-intuition behind our solution. \cref{sec:encodings} discusses alternative
-term encodings and related works. \cref{sec:grounwork} introduces
-the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
-basic simulation of higher order logic programs.
-\cref{sec:eta,sec:invariant1} completes its equational theory
-with support for $\eta$-conversion. \cref{sec:beta} deals with the
-practical necessity of ``tolerating'' terms outside of the
-pattern-fragment and discusses how heuristic can be applied.
-Finally \cref{sec:implementation} discusses the implementation in Elpi.
-
+We detail an encoding of a logic program in \Fo{} to a strongly related
+logic program in \Ho (the language of the meta-language) and we show that
+the higher-order unification procedure of the meta language \Ue{} can be
+efficiently used to simulate a higher-order unification procedure \Uo for
+the object language that features $\eta\beta$-conversion. We show how \Uo
+can be extended with heuristics to deal with problems outside the pattern
+fragment.\\
+Section~\ref{sec:problem-statement} formally states the problem and gives the
+intuition behind our solution; \cref{sec:simulation} sets up a basic
+simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
+extend it to higher-order logic programs in the pattern fragment
+while \cref{sec:beta} goes beyond the pattern fragment.
+Section \ref{sec:implementation} discusses the implementation in Elpi.\\
 The $\lambda$Prolog code discussed in the paper can be accessed at the
-URL: \url{https://github.com/FissoreD/paper-ho}.
+address \url{https://github.com/FissoreD/paper-ho}.
 
@@
 a minimal setting to ease its study. In this setting we have
-a \Fo{} language (for first order) with a rich(er) equational
-theory and a \Ho{} meta language with a simpler one, and we reuse
-the unification procedure of \Ho{} in order to implement one for \Fo.
+a \Fo{} language (for first order) with a richer equational
+theory and a \Ho{} meta language with a simpler one.
 

### 2024-04-12 paper/main.tex
sec 2.2

@@
 general than $\sigma'$.
-The domain of a substitution is the set of unification variables for which
-it provides an assignment.
-We write $\sigma \cup \sigma'$ to denote the concatenation of
-two substitutions whose domains are disjoint.
+% The domain of a substitution is the set of unification variables for which
+% it provides an assignment.
+% We write $\sigma \cup \sigma'$ to denote the concatenation of
+% two substitutions whose domains are disjoint.
 We shall use $\rho$ for \Fo{} substitutions,
@@
 
-We write \todo{meglio dire che sigma' estende sigma, togliere unione e quindi dominio da sopra}
+We write 
 $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
 $\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
-We write $t_1 \Ue t_2 \mapsto \sigma$ when
-the initial substitution is empty.
-Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
-the domains of $\sigma$ and $\sigma'$ are disjoint.
+Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$.
+We write $t_1 \Ue t_2 \mapsto \sigma'$ when
+the initial substitution $\sigma$ is empty.
+% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
+% the domains of $\sigma$ and $\sigma'$ are disjoint.
+We write \llambda as the set of terms that are in the pattern-fragment, i.e.
+every higher-order variable is applied to a list of distinct names.
 
@@
 
-\todo{spiegare llam (first use after intro)}
 Even if we provide an implementation of the object-language unification
@@
   $\hoUnifPb \cup \maybebeta = \emptyset$
-\end{lemma}
+\end{lemma}\noteD{Non capisco}
 \begin{proof}[Proof sketch]

### 2024-04-12 paper/main.tex
sec 2.3

@@
 \subsection{The problem: logic-program simulation}
-\todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
+% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
 We represent a logic program \emph{run} in \Fo{} as
 a list \emph{steps} of length $\mathcal{N}$.
-At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
-$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
+\foUnifPb is a list of pair of terms, each pair correspond to a unification
+problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
+$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
+elements extracted from $P_p$).
 \footnote{If the same rule is used multiple times in a run we

### 2024-04-12 paper/main.tex
git url update

@@
 %\acmDOI{ZZZZZZZZZZZZ}
+
+\def\githubUrl{\url{https://github.com/FissoreD/ho-unif-for-free}}
 
@@
 The $\lambda$Prolog code discussed in the paper can be accessed at the
-address \url{https://github.com/FissoreD/paper-ho}.
+address \githubUrl.
 

### 2024-04-12 paper/main.tex
sec 3.3

@@
 
-When variables $x$ and $y$ can occur in term $t$ we shall write
-$t_{xy}$ to stress this fact. 
-\todo{add notation for problem compilation}
+In \cref{sec:eta,sec:invariant1,sec:beta,sec:llam},
+we use the following schema to represent the compilation of the list of \Fo{}
+problems \foUnifPb into the \Ho problems \hoUnifPb. \mapStore and \linkStore are
+respectively the mapping and the link store.
+%
+\printAlll
+  {{{p_1,p_2},{p_3,p_4}}}
+  {{{t_1,t_2},{t_3,t_4}}}
+  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
+  {{{\eta,\Gamma,a,b}}}
+
+We index each sub-problem, sub-mapping, sub-link with its position in the image
+starting from $1$ and counting from left to right, top to bottom. For example,
+$\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
+The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.
+
+Moreover, to indicate the scope of a \Ho variable, we use that scope as
+subscript of the considered variable. For example, $X_{xy}$ is the variable $X$ 
+having in scope $x$ and $y$.
 

### 2024-04-12 paper/main.tex
sec 10

@@
 
-In this paper we show how to lift the meta language higher-order
-unification procedure to the object language. The resulting
-unification honours $\eta\beta$-equivalence for
+In this paper we show how to lift the meta language higher-order unification
+procedure to the object language. Our proposed approach is highly adaptable to
+align closely with the behavior of the object language. It is not tightly
+coupled with the Coq system but can serve as a flexible framework for meta
+programming in any ML.
+
+Our encoding leverages the advantage of not needing to recode the unification
+algorithm of the object language. Instead, it utilizes the unification of the
+meta language facilitated by the various links we establish to handle
+``problematic'' subterms. Additionally, our encoding benefits from the
+application of indexing algorithms for static clause filtering.
+
+Furthermore, the unification process we propose is tailored for potential future
+implementations of tabled search, incorporating memoization to retrieve
+solutions from previous searches.
+
+---
 

### 2024-04-12 paper/main.tex
sec 2.3: corrected notation
sec 2.3: notation

@@
 \newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
-\newcommand{\stepFD}[5]{%
-\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
+\newcommand{\stepFD}[4]{%
+\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
 \newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
 \newcommand{\stepHD}[6]{\ensuremath{%
-#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
+#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
 \newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
@@
 \begin{array}{l}
-\stepF{\foUnifPb}{p}{\rho}{\rho''}
+\stepF{\foUnifPb}{p}{\rho}{\rho'}
 \deff
-\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
+\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
 \runF{\foUnifPb}{\mathcal{N}}{\rho}
@@
 In order to simulate a \Fo{} logic program in \Ho{} we compile
-each \Fo{} term in \foUnifPb into a \Ho{} term $t$.
+each pair of \Fo{} terms in \foUnifPb into a pair of \Ho{} terms $t$.
 We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
@@
 for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
-for the set $\{ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} \}$.\\
+for the pair $(\hoUnifPb_{p_l}, \hoUnifPb_{p_r})$.\\
 \hstep{} is made of two sub-steps: a call to the meta language

### 2024-04-12 paper/main.tex
typos

@@
 a minimal setting to ease its study. In this setting we have
-a \Fo{} language (for first order) with a richer equational
+a \Fo{} language (for first order) with a rich equational
 theory and a \Ho{} meta language with a simpler one.
@@
 When it is clear from the context we shall use the same syntax for \Fo{} terms
-(although we never subscripts unification variables).
-
+(although we never subscripts unification variables).\\
 We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
@@
 it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
-weak head normal form of <<\elpiIn{app[F|L]}>>.
+weak head normal form of <<\elpiIn{app [F|L]}>>.
+Note that the symbol \elpiIn{|} separates the head of a list from the tail.
 
@@
 
-Although we provide an implementation of the meta-language unification \Ue
-in the supplementary material (that we used for testing purposes) we only
-describe its signature here.
+% Although we provide an implementation of the meta-language unification \Ue
+% in the supplementary material (that we used for testing purposes) we only
+% describe its signature here.
+\Ho{}'s unification signature is:
 
@@
 
+\noindent
 We write 
@@
 $\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
-Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$.
+Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
+reflected by signature above that relates two substitutions.
 We write $t_1 \Ue t_2 \mapsto \sigma'$ when
@@
 
+\noindent
 Even if we provide an implementation of the object-language unification
@@
 a list \emph{steps} of length $\mathcal{N}$.
-\foUnifPb is a list of pair of terms, each pair correspond to a unification
+\foUnifPb is a list of pair of terms, each pair corresponds to a unification
 problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
 $\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
-elements extracted from $P_p$).
+elements extracted from $\foUnifPb_p$).
 \footnote{If the same rule is used multiple times in a run we

### 2024-04-12 paper/main.tex
sec 7.2: typo
wip

@@
 \newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
-\newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}
+\newcommand{\llambda}{\ensuremath{\mathcal{L}}\xspace}
 
@@
 
-\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
-\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
+\newcommand{\lhs}{lhs\xspace}
+\newcommand{\rhs}{rhs\xspace}
 
@@
 
-\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}\xspace}
+\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
 \begin{definition}[\progressetaleft]\label{def:progressetaleft}
@@
 
-\newcommand{\progressetaright}{\emph{progress-$\eta$-right}\xspace}
+\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
 \begin{definition}[\progressetaright]\label{def:progressetaright} A link
@@
 
-\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}\xspace}
+\newcommand{\progressetadedup}{\emph{$\eta$-progress-deduplicate}\xspace}
 \begin{definition}[\progressetadedup]\label{def:progressetadedup}
@@
 \begin{theorem}[Fidelity in \maybeeta]
-  Given a list of unification problems \foUnifPb, such that 
-  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$,
+  % F x = z, F x y = z y
+  Given a list of unification problems \foUnifPb, such that \noteD{riprendere sec 3.1}
+  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$, \noteD{dire che M e biiettivo}
   the introduction 
@@
   \footnote{We also suppose that any higher-order variable is always applied with
-  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}
+  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}\noteD{modificare accordingly}
   \label{lemma:fidelity-maybeeta}
@@
 \begin{proof}[Proof sketch]
-  By the definition of \mapdeduplication, any two same \Fo{} variables $X_1, X_2$
+  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
+  variables $X_1, X_2$
   with different arities are related with \linketa. 
@@
 
+\begin{comment}
 \begin{proof}[Proof sketch]
@@
 \end{proof}
+\end{comment}
 
@@
 
-\newcommand{\progBetaLL}{\emph{progress-\llambda-refine}\xspace}
-\newcommand{\progBetaRH}{\emph{progress-\llambda-\rhs}\xspace}
+\newcommand{\progBetaLL}{\emph{\llambda-progress-refine}\xspace}
+\newcommand{\progBetaRH}{\emph{\llambda-progress-\rhs}\xspace}
 % \newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
-\newcommand{\progBetaFail}{\emph{progress-\llambda-fail}\xspace}
+\newcommand{\progBetaFail}{\emph{\llambda-progress-fail}\xspace}
 
 Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
-$4$ different activation rules:
+$3$ different activation rules:
 
@@
   Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
-  \progBetaRH, then it disappears from \linkStore and \elpiIn{progress}
+  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
   terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
@@
   %\noteE[inline]{is it ok?}
-  \noteE[inline]{funziona.
+  % \noteE[inline]{funziona.
 per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione).
@@
 \begin{theorem}[Fidelity with \linkbeta]
-  The introduction of \linkbeta guarantees \fullref{prop:fidelity}
+  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
 \end{theorem}
@@
   fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
-  then, by \cref{def:progBetaLH}, the unification fails, as per the
+  then, by \cref{def:progBetaFail}, the unification fails, as per the
   corresponding unification in \Fo{}.
@@
     {\eta,,B,\lambda x.C_{x}},
-    {\beta,x,C_{x},(D\appsep E_{x})}}}
+    {\llambda,x,C_{x},(D\appsep E_{x})}}}
 
@@
     {X,A,0}}}
-  {{{\beta,,C,(A\appsep a)}}}
+  {{{\llambda,,C,(A\appsep a)}}}
+
+% \todo{Rinominare i print dei problemi}
 
@@
 
-\section{Conclusion}
+\section{Related work and conclusion}
 

### 2024-04-12 paper/main.tex
typo
section 2 done (can be spellchecked)

@@
 Note that the symbol \elpiIn{|} separates the head of a list from the tail.
+\todo{explain forall2}
 
@@
 \subsection{The problem: logic-program simulation}
-% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
 We represent a logic program \emph{run} in \Fo{} as
-a list \emph{steps} of length $\mathcal{N}$.
-\foUnifPb is a list of pair of terms, each pair corresponds to a unification
-problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
-$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
-elements extracted from $\foUnifPb_p$).
-\footnote{If the same rule is used multiple times in a run we
-just consider as many copies as needed of the terms composing the
-rules, with fresh unification variables each time.}
+a sequence of \emph{steps} of length $\mathcal{N}$.
+At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
+$\foUnifPb_{p_r}$, taken from the list of all unification
+problems \foUnifPb.
+% \footnote{If the same rule is used multiple times in a run we
+% consider as many copies as needed of the terms composing the
+% rules, with fresh unification variables each time. Also, we do not
+% model backtracking directly, but rather the fact that a step in a run may
+% fail, and in that case the entire run is said to fail. A regular logic program
+% that succeeds is modeled as a many runs, with possibly overlapping starts,
+% where at least one run does not end in a failure. }
 The composition of these steps starting from the
@@
 In order to simulate a \Fo{} logic program in \Ho{} we compile
-each pair of \Fo{} terms in \foUnifPb into a pair of \Ho{} terms $t$.
+each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
 We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
@@
 description is deferred to \cref{sec:nutshell}.
-
-We simulate each run in \Fo{} with a run in \Ho as follows.
+We write $\hoUnifPb_p = \{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$
+and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.
+
+We simulate each run in \Fo{} with a run in \Ho as follows:
+
 %Note that $\sigma_0$ is the empty substitution.
@@
   \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
-\end{array}
-$$
-$$
-\begin{array}{l}
   \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
@@
 for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
-for the pair $(\hoUnifPb_{p_l}, \hoUnifPb_{p_r})$.\\
+for the set $\{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$.\\
 \hstep{} is made of two sub-steps: a call to the meta language
@@
 \noindent
-That is, the two executions give the same result. Moreover:
+That is, the two executions give the same result if all terms in
+\foUnifPb are in the pattern fragment. Moreover:
 
@@
 $\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
+Indeed, the main difference between \cref{prop:fidelity} and \cref{prop:fidelity-recovery}
+is that the assumption of the former is purely static, it can be checked upfront. 
+When this assumption is not satisfied one can still simulate a logic program and
+have guarantees of fidelity if, at run time, decidability of higher-order
+unification is restored.
 
@@
 making the second problem re-enter \llambda.
-\todo{say that this is runtime, while the other fidelity is compile time}
+
+\paragraph{Backtracking} We omit it from our model of a logic programs execution
+since it pays a very minor role, orthogonal to higher-order unification.
+We point out that each \emph{run} corresponds to a (proof search) branch in the
+logic program that either fails at some point, or succeeds. A computation that
+succeeds by backtracking, exploring multiple branches, could be
+modeled as set of runs with (possibly non empty) common prefixes.
 
@@
 We now define ``problematic'' and ``well behaved'' more formally
-and we pick the $\Diamond$ symbol since it
-stands for ``possibly'' in modal logic and all problematic terms
-are characterized by some ``uncertainty''.
+We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
+and all problematic terms are characterized by some ``uncertainty''.
 
@@
 \noindent
-\todo{dire perchè W(T), dire che il converse non vale (alcuni check li fa la progress)}
-Note that the property holds for any substitution, it could be given by an
-oracle and not necessarily a most general one, and for any terms,
-in particular terms in $\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{}$.
+In other words the compiler outputs terms in \wellb, even if its
+input is not.
+Note that the property holds for any substitution. $\rho$ could be given by an
+oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
+does not contradict \Eo.
 

### 2024-04-12 paper/main.tex
up to 3.5

@@
 
-Unification variables are identified by a
-natural number that represents a memory addresses.
+Unification variables are identified by a (unique) memory address.
 The memory and its associated operations are described below:
@@
 
-Since each occurrence of a \Ho unification variables has a scope,
+Since each \Ho unification variables occurs together with a scope,
 its assignment needs to be abstracted over it to enable the
@@
 its \elpiIn{abs} binding constructor.
-On the contrary a solution to a \Fo variable is a plain term.
-
-\input{code/fo_subst}
+
 \input{code/ho_subst}
@@
 \noindent
-We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
-the one of \Ho.
+A solution to a \Fo{} variable is a plain term, that is \elpiIn{fsubst}
+is an abbreviation for \elpiIn{mem fm}.
+%We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
+%the one of \Ho.
 
@@
 
+\noindent
 Each \elpiIn{hvariable} is stored in the mapping together with
-its arity so that the code of \ref{clause:malloc} below can preserve:
+its arity (a number) so that the code of \ref{clause:malloc} below can preserve:
 
@@
 
-Applying the substitution corresponds to dereferencing a term with respect to
-the memory.
+%Applying the substitution corresponds to dereferencing a term with respect to
+%the memory.
 %  To ease the comparison we split \Fo{} dereferencing into a
@@
 % \input{code/fderef}
-It is worth looking at the code fpr \Ho{} to remark how assignments are moved
+It is worth looking at the code of \elpiIn{deref} that
+applies the substitution to a \Ho{} term. Remark how assignments are moved
 to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
@@
 of the arguments of all occurrences of a unification are the same. Hence
-they have the same type in the meta-level and the number of \elpiIn{abs}
-nodes in the assignment matches that length. In turn this grants that
-\elpiIn{move} never fails.\todo{convoluted}
+they have the same simple type for the meta-level and hence the number of
+\elpiIn{abs} nodes in the assignment matches that length.
+In turn this grants that \elpiIn{move} never fails.
 
@@
 To accommodate this situation the compiler wraps \elpiIn{baselink} using
-the \elpiIn{inctx} container (see rule~\ref{data:inctx}).
+the \elpiIn{inctx} container (see~\ref{data:inctx} also used for \elpiIn{subst}).
 
@@
 New links are suspended by construction.
-If the left hand side variable is assigned during a step, then 
+If the left hand side is assigned during a step, then 
 the link is considered for progress and possibly eliminated.
@@
 \label{sec:compilation}
-\noteE[inline]{manca beta normal in entrata}
-
-The main task of the compiler is to recognize \Fo{} variables standing
-for functions and map them to higher order variables in \Ho.
+
+The simple compiler described in this section serves as a base for the
+extensions in \cref{sec:llam,sec:eta,sec:beta}.
+Its main task is to beta normalize the term and map one syntax tree to
+the other.
 In order to bring back the substitution from \Ho{} to \Fo{} the compiler
@@
 links (suspended unification problems) that play no role in this section
-but play a major role in \cref{sec:eta} and \cref{sec:beta}.
+but play a major role in \cref{sec:llam,sec:eta,sec:beta}.
 With respect to \cref{sec:problem-statement} the signature also allows
@@
 \input{code/comp_base}
+\todo{use the real code}
+\begin{elpicode}
+type compile fm -> tm -> mmap -> mmap -> links -> links ->
+  subst -> subst -> o.
+compile Fo0 Ho M0 M1 L0 M1 S0 S1 :-
+  beta-normal Fo0 Fo, comp Fo Ho M0 M1 L0 M1 S0 S1.
+\end{elpicode}
 
@@
 \noindent
-Note that we could remove the first rule, whose solve purpose is to make
+Note that we could remove the first rule, whose sole purpose is to make
 links more readable by pruning unused context entries.
@@
 A step in \Ho consists in unifying two terms and reconsidering all
-links for progress. If any of the two tasks fail we say that the entire step
+links for progress. If any of the two tasks fails we say that the entire step
 fails, and it is at this granularity that we can relate steps in the
@@
 Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
-and justify why the don't hinder termination. For brevity we omit the code
-that applies the substitution \elpiIn{S1} to all terms in \linkStore.
+and justify why the don't hinder termination.
+% For brevity we omit the code
+% that applies the substitution \elpiIn{S1} to all terms in \linkStore.
 
 \subsubsection{Occur check}
-Since compilation moves problematic terms out of the sigh of \Ue{},
+Since compilation moves problematic terms out of the sight of \Ue{},
 that procedure can only perform a partial occur check. For example the
@@
 but should be disallowed if a $\linkStore$ contains a link like
-$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
+$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
 a lambda in head position, but we surely know it contains $X$, hence
 $f~Y$ and that fails the occur check.
-The procedure \elpiIn{occur-check-links} is in charge of ensuring that
-each link does not represent a (suspended) unification problem doomed
-to fail because of occur check. This check is needed in order to
+The procedure \elpiIn{occur-check-links} is in charge of
+performing this check that is needed in order to
 guarantee \fullref{prop:fidelity}.

### 2024-04-12 paper/main.tex
rework last sec

@@
 
-\section{Other encodings and related work}\label{sec:encodings}
-
-\noindent
-One could ignore the similarity between \Uo{} and \Ue{} and ``just''
-describe the object language unification procedure in the meta language by
-crafting a \elpiIn{unif} routine and using it as follows in
-rule \ref{clause:r3}:
+\noindent
+\section{Related work and conclusion}
+
+Different strategies can be used to unify terms of the object language in the
+meta langauge. One initial approach involves integrating \Uo as a procedure
+within the ML, outlined as follows:
 
 \begin{elpicode}
-decision X :- unif X (all A x\ app [P, x]), finite A,
+  decision X :- X ~\Uo~(all A x\ app [P, x]), finite A,
     pi x\ decision (app [P, x]).
@@
 
-\noindent
-This choice would underuse the logic programming engine provided by
-the meta language since, by removing any datum from the head of rules,
-indexing degenerates. Moreover the unification procedure \elpiIn{unif}
-programmed in the meta language is likely to be an order of magnitude
-slower than one that is built-in.
+Opting for this method would result in a suboptimal utilization of the logic
+programming engine provided by the meta language, as it degrades indexing by
+eliminating data from rule heads. Additionally, embedding \Uo within the meta
+language is likely to be significantly slower compared to a built-in
+solution.
 
@@
 \begin{elpicode}
-finite (fin N).
-decision (nfact N NF).
 decision (all A x\ P x) :- finite A, pi x\ decision (P x).
@@
 \noindent
-There are two reasons for dismissing this encoding. The first one is that
-in CIC it is not always possible to adopt it since the type system
-of the meta language is too weak to accommodate for the one of the 
-object language. In CIC the lambda abstraction has to carry a type in order to
-make type checking decidable. Moreover CIC allows for functions with a
-variable arity, like the following example:
-
-\begin{coqcode}
-Fixpoint arr T n := if n is S m then T -> arr T m else T.
-Definition sum n : arr nat n := ...
-Check sum 2   7 8   : nat.
-Check sum 3   7 8 9 : nat.
-\end{coqcode}
-
-\noindent
-The type system of the $\lambda$Prolog is too stringent to accept this terms.\\
-The second reason is that the CIC encoding provided by Elpi is used for meta
-programming (extending) the Coq system, hence it must accommodate the
-manipulation of terms that are now know in advance (not even defined in Coq)
-without using introspection primitives such as Prologs's \texttt{functor}
-and \texttt{arg}. In this sense constants have to live in an open world, like
-the \elpiIn{string} data type used in the examples so far.
+However, two reasons justify rejecting this encoding. Firstly, in CIC, it is
+not always feasible to adopt it due to the meta language's type system being
+too limited to accommodate that of the object language. \\
+Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
+programming, in order to extend the Coq system. Consequently, it must be able to
+manipulate terms that are not known in advance
+without relying on introspection primitives such as Prolog's
+\texttt{functor} and \texttt{arg}. In this context, constants need to live in
+an open world, akin to the \elpiIn{string} data type used in the preceding
+examples.
 
@@
 
-\section{Related work and conclusion}
-
-In this paper we show how to lift the meta language higher-order unification
-procedure to the object language. Our proposed approach is highly adaptable to
+Our encoding provides a third option that addresses all the considerations
+mentioned earlier. It capitalizes on the benefit of not requiring the recoding
+of the unification algorithm of the object language. Instead, it employs the
+unification capabilities of the meta language, facilitated by the various links
+we establish to manage ``problematic'' subterms. Moreover, our encoding takes
+advantage of indexing algorithms for static clause filtering. It's noteworthy
+that we only replace the minimum necessary information with variables,
+specifically targeting \maybeeta and \notllambda subterms.
+
+Our proposed approach is highly adaptable to
 align closely with the behavior of the object language. It is not tightly
@@
 
-Our encoding leverages the advantage of not needing to recode the unification
-algorithm of the object language. Instead, it utilizes the unification of the
-meta language facilitated by the various links we establish to handle
-``problematic'' subterms. Additionally, our encoding benefits from the
-application of indexing algorithms for static clause filtering.
-
 Furthermore, the unification process we propose is tailored for potential future

### 2024-04-12 paper/main.tex
small change
commented some text

@@
 we establish to manage ``problematic'' subterms. Moreover, our encoding takes
-advantage of indexing algorithms for static clause filtering. It's noteworthy
+advantage of indexing algorithms and mode analysis for static clause filtering. It's noteworthy
 that we only replace the minimum necessary information with variables,
@@
 
----
-
-Benefits: less work, reuse efficient ho unif (3x faster), indexing,
-
-Future: tabling and static analysis (reuse for ML again).
-
-Very little is Coq specific. Applies to all OL that are not a subsystem of 
-HOL, or for ML that are used for meta programming.
+% ---
+
+% Benefits: less work, reuse efficient ho unif (3x faster), indexing,
+
+% Future: tabling and static analysis (reuse for ML again).
+
+% Very little is Coq specific. Applies to all OL that are not a subsystem of 
+% HOL, or for ML that are used for meta programming.
 

### 2024-04-12 paper/main.tex
add bib
done 4 (can be spellchecked)

@@
 
-\subsubsection{Occur check}
+\subsubsection{Occur check}\label{sec:oc}
 Since compilation moves problematic terms out of the sight of \Ue{},
@@
 First and foremost problematic terms stored in
-\linkStore have to be moved back into the game.
-Since links are of the form uvar = term (\fullref{inv:linklhs})
-and are duplicate free (see dedup-beta dedup-eta),
-one can turn a link X = t into an assignment $X \mapsto t$.
-This can in general be achieved by unifying X with t. The case
-where t is not in \llambda (link beta/llam) is discussed in section xx.
+\linkStore have to be moved back into the game:
+a suspended link must be turned into a valid assignment.
+This operation is possible thanks to \fullref{inv:linklhs},
+thanks to the fact that no link causes an occur-check~(\ref{sec:oc})
+and the fact that \linkStore is duplicate free (\cref{TODO}).
 
 The second step amounts at allocating new variables in the
-memory of \Fo{}. In particular some unif problems
-such as $F x y = F x z$ requires to allocate a 
-variable $G$ so that the assignment
-$F_{ab} \mapsto G_a$ can be used to perform
-required pruning.
+memory of \Fo{}. This technicality is required because some
+higher-order unification may need to prune a variable. For example
+$F\appsep x\appsep y = F\appsep x\appsep z$ requires to allocate a 
+variable $G$ in order to express the assignment $F_{ab} \mapsto G_a$.
 
 The last step amounts at decompiling each assignment.
-Decompiling a term is trivial. An assignment has an
-abs node, as in move, can be eliminated by replacing
-the bound variable by the actual term in scope.
-In order to do this, one needs the \mapStore to be
-a bijection. This is the job of \cref{sec:invariant1}.
-\todo{polish}
-
-\todo[inline]{dire che però si passa per una subst in cui ste abs
-le cambio in lam. Nel codice Coq ci scrivevamo il tipo nella
-arity, e quindi sappiamo fare i lambda bene, senza perdita di
-informazione. Qui i lam non hanno info, facile. Ma in generale
-bisogna spiegare come ci si salva. Ci dormo su: o non generiamo
-la subst ma solo il primo termine (la query iniziale) istanziato
-(funziona sempre, la prova è quella sopra) oppure bisogna siegare
-tutto sto casino e serve un po' di spazio.}
-
-%\input{code/decompile_type}
-
-% \noteE{What is commit-links and complete-mapping?}
-
-% \noindent
-% Decompiling an assignment requires to turn abstractions into
-% lambdas. For aesthetic purposes we also eta-contract the result
-% (not needed since \Fo{} equality can do that)
-
-% \input{code/decompm}
-
-% \noindent
-% Finally decompiling a term is trivial, now that we have an extended
-% mapping containing all unassigned variables \Ue may have introduced.
-
-% \input{code/decomp}
-
-% \noindent
-% Note that we use beta to build fapp nodes when needed (if Ag is empty
-% no \elpiIn{fapp} node should appear).
+Decompiling a term is trivial since \mapStore is a bijection.
+The only tricky part concerns the \elpiIn{abs} node. In out simple setting
+the \elpiIn{flam} node carries no extra info (other then the function body),
+so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
+In the case of CIC, where lambdas carry the type of the bound variable, one
+has to store it somewhere. Note how this piece of information is akin
+to the arity of variables, that is CIC's unification variables have a
+(function) type, and that type can be used to annotate the lambdas
+needed in order to express their assignment.
+
+\begin{lemma}[Compilation round trip]\label{prop:comprt} If
+  \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
+\end{lemma}
+% \begin{proof}[Proof sketch]
+% Trivial since \mapStore is a bijection and 
+% the terms are beta normal.
+% some discussion about commit maybellam to be done later.
+% \end{proof}
 
@@
 
+Even if the compiler, so far, makes no use 
+of the higher-order nature of the meta language (all variables have
+an empty scope so far) we can already show the code of \Ue{}.
+
 \input{code/unif_fo}
-
-The code given so far still makes no use
-of the higher order nature of the ML unif language,
-indeed the scope of unif variables generated by the
-compiler is always empty, so \Ue{} is first order.
-
-Still, if \foUnifPb is already \wellb, we can set up a proof
-that will also work when comp enforces \wellb and  hstep
-preserves it, and when terms in \llambda are mapped to ho variables
-with a scope.
-
-\begin{lemma}[Compilation round trip] If
-  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
-\end{lemma}
-\begin{proof}[Proof sketch]
-trivial if the mapping is a bijection and 
-the terms are beta normal.
-some discussion about commit maybellam to be done later.
-\end{proof}
+\todo{Fix real code calling compile (not comp), rename decompm to decomp}
+
+If \wellb{\foUnifPb} then we can prove that \Uo is a good
+``first order'' unification algorithm. Later, when the compiler will
+enforce it by \cref{prop:w-enforcing} and the proof will be adjusted.
 
@@
  In this setting \Ee is as strong as
-\Eo on ground terms. What we have to show is that whenever two different \Fo{}
+\Eo. What we have to show is that whenever two different \Fo{}
 terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
-and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
+and \ref{clause:beta2} if needed) we can produce this $\rho$ by finding
 a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
-If we look at the \Fo{} terms is only one interesting cases:
-\begin{itemize}
-\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
+If we look at the syntax of \Fo{} terms the only interesting case is
+\elpiIn{fuva X ~\Uo~s}. In this case after compilation we have
 $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
-  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
-\end{itemize}
-Since the mapping is a bijection occur check in \Ho{} corresponds to occur
-check in \Fo{}.
+$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
 \end{proof}
@@
 \begin{proof}[Proof sketch]
-Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
-in this
-context where input terms are $\beta\eta$-normal and all is \wellb,
-\Ue is equivalent to \Uo.
+Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same.
+Since all terms are $\beta$-normal and \wellb, \Ue is equivalent to \Uo.
 \end{proof}
@@
 approximations. This is the case for example of the unification algorithm of Coq
-used in its type class solver~\cite{sozeau08}. The approximation consists in
+used in its type-class solver~\cite{sozeau08}. The approximation consists in
 forcing a choice (among the others) when the unification problem is outside

### 2024-04-12 paper/main.tex
solve todo code
remove todo
abstract with spellchecker

@@
   key ways: 1) variable binding and substitution are for free when ML binders
-  represent object logic ones; 2) proof construction, and even proof search, is
+  represent object logic ones; 2) proof construction, and even proof search, are
   greatly simplified by leveraging the unification procedure provided by the ML.
@@
 
-  The object logic we are interested in is % an extension of the latter,
-  Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
-  we want to implement a higher-order unification-based 
-  proof search procedure using the ML
+  The object logic we are interested in is
+  Coq's Calculus of Inductive Constructions (CIC)\cite{Coq-refman}. We aim to
+  develop a higher-order unification-based 
+  proof search procedure for it using the ML
   Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
-  Elpi's equational theory comprises
-  $\eta\beta$ equivalence and comes equipped with a
+  Elpi's equational theory includes
+  $\eta\beta$ equivalence and features a
   higher-order unification procedure \Ue restricted to the pattern
   fragment~\cite{miller92jsc}.
-  Elpi comes with an encoding of CIC that works well
+  Elpi offers an encoding of CIC suitable
   for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
   but restricts \Ue to roughly first-order unification problems only.
-  We call this basic encoding \Fo.
-
-  In this paper we propose a better-behaved encoding \Ho,
-  and show how to map unification problems in \Fo{}
-  to related problems in \Ho.
-    As a result we obtain \Uo{}, a higher-order unification procedure for \Fo{}
-    that honours $\eta\beta$-equivalence (for CIC functions), solves
-  problems in the pattern fragment and allows
+  We refer to this basic encoding as \Fo.
+
+  In this paper we propose a more well-behaved encoding called \Ho,
+  and show how to translate unification problems from \Fo{}
+  to corresponding ones in \Ho.
+  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
+    that honours $\eta\beta$-equivalence (for CIC functions), addresses
+  problems within the pattern fragment, and allows
   for the use of heuristics to deal with problems outside the
-  pattern fragment. Moreover, since \Uo{} delegates most of the work to \Ue,
+  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
   it can be used to efficiently simulate a logic program in \Fo{} by
@@
 \input{code/comp_base}
-\todo{use the real code}
-\begin{elpicode}
-type compile fm -> tm -> mmap -> mmap -> links -> links ->
-  subst -> subst -> o.
-compile Fo0 Ho M0 M1 L0 M1 S0 S1 :-
-  beta-normal Fo0 Fo, comp Fo Ho M0 M1 L0 M1 S0 S1.
-\end{elpicode}
+\input{code/compile.tex}
 
@@
 \input{code/unif_fo}
-\todo{Fix real code calling compile (not comp), rename decompm to decomp}
 

### 2024-04-12 paper/main.tex
intro with spellchecker

@@
 Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
-$\lambda$Prolog~\cite{miller_nadathur_2012} and
+$\lambda$Prolog~\cite{miller_nadathur_2012}, and
 Isabelle~\cite{10.1007/978-3-540-71067-7_7}
@@
 The use of these meta languages facilitates this task in two
-key ways. The first and most well know one is that variable binding and
+key ways. The first and most well-know one is that variable binding and
 substitution come for free. %, exploited in all works mentioned above.
@@
 The object logic we are interested in is Coq's~\cite{Coq-refman}
-Calculus of Inductive Constructions (CIC) and we want to implement a
+Calculus of Inductive Constructions (CIC), and we aim to implement a
 form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
-Type-class solvers are unification based proof search procedures
-reminiscent of Prolog that back-chain lemmas taken
+Type-class solvers are unification based on proof search procedures
+reminiscent of Prolog, which back-chain lemmas taken
 from a database of ``type-class instances''. Given this
 analogy with Logic Programming we want to leverage the
-Elpi~\cite{tassi:hal-01637063} meta programming language,
+Elpi~\cite{tassi:hal-01637063} meta-programming language,
 a dialect of $\lambda$Prolog, already used to extend
 Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
-In this paper we focus on one aspect of
+In this paper, we focus on one aspect of
 this work, precisely \emph{how to reuse the higher-order unification procedure
@@
 identifies predicates equipped with a decision procedure, while
-\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
+\coqIn{Finite} identifies types whose inhabitants can be enumerated in a (finite) list.
 The following three type-class instances state that:
@@
 
-\noindent Given this database a type-class solver is expected to
+\noindent Given this database, a type-class solver is expected to
 prove the following statement automatically:
@@
 rules 1 and 2 respectively.
-Note that rule 3 features a second order parameter \coqIn{P} that stands for
+Note that rule 3 features a second-order parameter \coqIn{P} that represents
 a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
 The solver has to infer a value for \coqIn{P} by unifying the conclusion
-of rule 3 with the goal, and in particular it has to solve the unification
-problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
-called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
+of rule 3 with the goal, and in particular, it has to solve the unification
+problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
+pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
 solution $\customlabel{solution:intro}{\rho}$ that assigns
@@
 
-In order to implement such a search in Elpi we shall describe the encoding
+In order to implement such a search in Elpi, we shall describe the encoding
 of CIC terms and then the encoding of instances as rules.
@@
 an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
-terms, called \elpiIn{tm}, that features (among others) the following
+terms, called \elpiIn{tm}, that includes (among others) the following
 constructors:
@@
 \noindent
-Following the standard syntax of $\lambda$Prolog~\cite{miller_nadathur_2012}
-the meta level binding of a variable \elpiIn{x} in an expression
-\elpiIn{e} is written <<\elpiIn{x\ e}>>, while square brackets delimit a
-list of terms separated by comma. For example the term
+Following the standard syntax of $\lambda$Prolog~\cite{miller_nadathur_2012},
+the meta-level binding of a variable \elpiIn{x} in an expression
+\elpiIn{e} is written as <<\elpiIn{x\ e}>>, while square brackets delimit a
+list of terms separated by comma. For example, the term
 <<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:
@@
 We now illustrate the encoding of the three instances above as higher-order
-logic-programming rules: capital letters denote rule
-parameters; \elpiIn{:-} separates the rule's head from the premises and
+logic programming rules: capital letters denote rule
+parameters; \elpiIn{:-} separates the rule's head from the premises, and
 \elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
@@
 \noindent
-Since \elpiIn{Pm} is an higher-order unification variable
+Since \elpiIn{Pm} is a higher-order unification variable
 of type \elpiIn{tm -> tm},
@@
 \noindent
-Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g}
+Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g},
 the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
@@
 % \noindent
-In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
+In turn, this redex prevents rule \ref{clause:r2} from backchaining properly since
 the following unification problem has no solution:
@@
 \noindent
-The root cause of the problems we sketched in this example
+The root cause of the problems we outlined in this example
 is a subtle mismatch between the equational theories of the meta language
-and the object language, that in turns makes the 
+and the object language, which in turn makes the 
 unification procedures of the meta language weak.
@@
 the equational theory of CIC is much richer, for efficiency and predictability
-reasons automatic proof search procedure typically employ a unification
+reasons, automatic proof search procedures typically employ a unification
 procedure that only captures a $\eta\beta$-equivalence and only operates
-in \llambda. The similarity is striking, but one needs some care in order to
+in \llambda. The similarity is striking, but one needs to exercise some caution in order to
 simulate a logic program in CIC using the unification of Elpi.

### 2024-04-12 paper/main.tex
sec 2.1 with spellchecker

@@
 
-Even if we encountered the problem working on CIC we devise
-a minimal setting to ease its study. In this setting we have
+Even if we encountered the problem working on CIC, we devise
+a minimal setting to ease its study. In this setting, we have
 a \Fo{} language (for first order) with a rich equational
@@
 
-In order to reason about unification we provide a description of the
+To reason about unification, we provide a description of the
 \Fo{} and \Ho languages where unification variables
-are first class terms, i.e. they have a concrete syntax as
-per \cref{code:common-terms}.
+are first-class terms, i.e. they have a concrete syntax as
+shown in \cref{code:common-terms}.
 % We keep these languages
@@
 in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
-the arguments of an higher order variable are given via the \elpiIn{fapp}
+the arguments of a higher-order variable are given via the \elpiIn{fapp}
 constructor. For example the term <<\coqIn{P x}>> is represented as
@@
 of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
-In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
-since unification variables are higher order and come equipped with an
+In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
+since unification variables are higher-order and come equipped with an
 explicit scope.
@@
 $x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
-However we need to
+However, we need to
 distinguish between the ``application'' of a unification variable
@@
 while we use juxtaposition for regular application.
-Here a few examples:\\
+Here are few examples:\\
 \vspace{4pt}
@@
 \noindent
-When it is clear from the context we shall use the same syntax for \Fo{} terms
+When it is clear from the context, we shall use the same syntax for \Fo{} terms
 (although we never subscripts unification variables).\\
-We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
+We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$, \ldots for
 terms in \Ho{}.

### 2024-04-12 paper/main.tex
sec 2.2 with spellchecker

@@
 
-In order to specify unification we need to
+In order to specify unification, we need to
 define the equational theory and
@@
 \subsubsection{Term equality: \Eo and \Ee}
-For both languages we extend the equational theory
-over ground terms to the full language by adding the reflexivity of
+For both languages, we extend the equational theory
+over ground terms to the full language by adding the reflexivity for
 unification variables (a variable is equal to itself).
@@
 and define the usual congruence over terms. Since
-we use an HOAS encoding they also capture $\alpha$-equivalence.
-In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.
+we use an HOAS encoding, they also capture $\alpha$-equivalence.
+In addition to that, \Eo has rules for $\eta$ and $\beta$-equivalence.
 
@@
 treatment in the implementation of \Uo.
-For brevity we omit the code of \elpiIn{beta}:
+For brevity, we omit the code of \elpiIn{beta}:
 it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
@@
 and $\sigma$ for the \Ho ones.
-For brevity, in this section we consider
+For brevity, in this section, we consider
 the substitution for \Fo{} and \Ho{} identical.
 We defer to \cref{sec:grounwork} a more precise description
-pointing out theirs differences.
+pointing out their differences.
 
@@
 Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
-reflected by signature above that relates two substitutions.
+reflected by the signature above that relates two substitutions.
 We write $t_1 \Ue t_2 \mapsto \sigma'$ when

### 2024-04-12 paper/main.tex
sec 2.3 with spellchecker

@@
 a sequence of \emph{steps} of length $\mathcal{N}$.
-At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
+At each step, $p$ we unify two terms, $\foUnifPb_{p_l}$ and
 $\foUnifPb_{p_r}$, taken from the list of all unification
@@
 empty substitution $\rho_0$ produces the final
-substitution $\rho_\mathcal{N}$, that is the result of the
-logic-program execution.
+substitution $\rho_\mathcal{N}$, which is the result of the
+logic program execution.
 %
@@
 \noindent
-\todo{say that backtracking is not important} 
-In order to simulate a \Fo{} logic program in \Ho{} we compile
+In order to simulate a \Fo{} logic program in \Ho{}, we compile
 each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
-We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
+We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
 is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
-out that it additionally produce a variable mapping $m$ and list of links $l$.
+out that it additionally produces a variable mapping $m$ and a list of links $l$.
 The variable map connects unification variables in \Ho to variables
@@
 unification and a check for \progress{} on the set of links, that intuitively
-will compensate for the weaker equational theory honoured by \Ue.
-\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
+will compensate for the weaker equational theory honored by \Ue.
+\hrun{} compiles all terms in \foUnifPb{}, then executes each step, and
 finally decompiles the solution.
@@
 \noindent
-In particular this property guarantees that a \emph{failure} in the \Fo{} run
+In particular, this property guarantees that a \emph{failure} in the \Fo{} run
 is matched by a failure in \Ho{} \emph{at the same step}. We consider this
 property very important from a practical point of view since it guarantees
-that the execution traces are strongly related and in turn this enables a user
+that the execution traces are strongly related, and in turn, this enables a user
 to debug a logic program in \Fo{} by looking at its execution trace in
@@
 
-In other words if the two terms involved in a step
+In other words, if the two terms involved in a step
 re-enter \llambda, then \hstep and \fstep are again related, even if
@@
 is that the assumption of the former is purely static, it can be checked upfront. 
-When this assumption is not satisfied one can still simulate a logic program and
+When this assumption is not satisfied, one can still simulate a logic program and
 have guarantees of fidelity if, at run time, decidability of higher-order
@@
 
-This property has a practical relevance since in many logic programming
+This property has practical relevance since in many logic programming
 implementations, including Elpi, the order in which unification problems
@@
 unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
-$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
+$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$,
 making the second problem re-enter \llambda.
 
-\paragraph{Backtracking} We omit it from our model of a logic programs execution
-since it pays a very minor role, orthogonal to higher-order unification.
+\paragraph{Backtracking} We omit it from our model of a logic program's execution
+since it plays a very minor role, orthogonal to higher-order unification.
 We point out that each \emph{run} corresponds to a (proof search) branch in the
@@
 succeeds by backtracking, exploring multiple branches, could be
-modeled as set of runs with (possibly non empty) common prefixes.
+modeled as a set of runs with (possibly non-empty) common prefixes.
 

### 2024-04-12 paper/main.tex
sec 2.4 with spellchecker

@@
 \noindent
-An example of term \maybebeta{} is the application $F \appsep x$.
+An example of a \maybebeta{} term is the application $F \appsep x$.
 This term is problematic since the application node of
@@
 unification failure, i.e. by properly instantiating $F$ the term
-head constructor may become a $\lambda$, or a constant or stay an application.
+head constructor may become a $\lambda$, or a constant, or remain an application.
 
@@
 \noindent
-An example of term $s$ in \maybeeta{} is
+An example of a term $s$ in \maybeeta{} is
 $\lambda x.\lambda y.F \appsep y \appsep x$
@@
 $\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
-makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$
-that is the eta long form of $f$. This term is problematic since
+makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
+which is the eta long form of $f$. This term is problematic since
 its leading $\lambda$ abstraction cannot justify a
@@
 These terms are problematic for the very same reason terms in \maybebeta are,
-but cannot be handled directly by the unification of the meta language, that
+but they cannot be handled directly by the unification of the meta language, which
 is only required to handle terms in \llambda. Still, there exists a
@@
 \noindent
-Proposition \ref{prop:w-preservation} is key to prove \cref{prop:simulation,prop:fidelity}:
-informally it says that the problematic terms moved on the side by the compiler
-are not put back by \hstep, hence \Ue{} can continue to operate properly.
+Proposition \ref{prop:w-preservation} is key to proving \cref{prop:simulation,prop:fidelity}.
+Informally, it says that the problematic terms moved on the side by the compiler
+are not reintroduced by \hstep, hence \Ue{} can continue to operate properly.
 In \cref{sec:simulation,sec:eta,sec:beta}
@@
 \notllambda and how \progress takes care of them preserving \wellb
-and granting \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.
+and ensuring \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.
 

### 2024-04-12 paper/main.tex
done 4 (can be spellchecked)

@@
 
-\newcommand{\unifcorrect}[3]{
+\newcommand{\unifcorrectX}[4]{
   % \forall \rho #3_1 #3_2, %
     % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
-    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
+   #4 \Rightarrow
       #3_1 #2 #3_2 \mapsto \rho
@@
 }
-
-\newcommand{\unifcomplete}[3]{
+\newcommand{\unifcorrect}[3]{\unifcorrectX{#1}{#2}{#3}{ \{#3_1, #3_2\} \subseteq \llambda}}
+
+\newcommand{\unifcompleteX}[4]{
   % \forall #3_1 #3_2, %
     % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
-    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
+    #4 \Rightarrow
       % \forall \rho, %
@@
 }
+\newcommand{\unifcomplete}[3]{\unifcompleteX{#1}{#2}{#3}{\{#3_1, #3_2\} \subseteq \llambda}}
+
 \newcommand{\maybeeta}{\texorpdfstring{\ensuremath{\Diamond\eta}\xspace}{maybeeta}}
-\newcommand{\maybebeta}{\texorpdfstring{\ensuremath{\Diamond\beta_0}\xspace}{maybebeta}}
+\newcommand{\maybebeta}{\texorpdfstring{\ensuremath{\Diamond\beta}\xspace}{maybebeta}}
 \newcommand{\notllambda}{\texorpdfstring{\ensuremath{\Diamond\llambda}\xspace}{maybellam}}
@@
 
-\subsection{Notational conventions}\todo{move away}
-
-In \cref{sec:eta,sec:invariant1,sec:beta,sec:llam},
-we use the following schema to represent the compilation of the list of \Fo{}
-problems \foUnifPb into the \Ho problems \hoUnifPb. \mapStore and \linkStore are
-respectively the mapping and the link store.
-%
-\printAlll
-  {{{p_1,p_2},{p_3,p_4}}}
-  {{{t_1,t_2},{t_3,t_4}}}
-  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
-  {{{\eta,\Gamma,a,b}}}
-
-We index each sub-problem, sub-mapping, sub-link with its position in the image
-starting from $1$ and counting from left to right, top to bottom. For example,
-$\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
-The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.
-
-Moreover, to indicate the scope of a \Ho variable, we use that scope as
-subscript of the considered variable. For example, $X_{xy}$ is the variable $X$ 
-having in scope $x$ and $y$.
-
 % The first three rules unify terms with same rigid heads, and
@@
 
-Even if the compiler, so far, makes no use 
-of the higher-order nature of the meta language (all variables have
-an empty scope so far) we can already show the code of \Ue{}.
+We already have all the pieces to show the code of \Ue{}.
 
@@
 
-If \wellb{\foUnifPb} then we can prove that \Uo is a good
-``first order'' unification algorithm. Later, when the compiler will
-enforce it by \cref{prop:w-enforcing} and the proof will be adjusted.
-
-\begin{lemma}
-\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
+\noindent
+So far the compiler is very basic, it does not really
+enforce the terms passed to \hstep are in \wellb,
+and indeed makes no use of the higher-order capabilities
+of the meta language (all generated
+variables have an empty scope).
+Still, we can prove that \Uo is a good
+``first order'' unification algorithm if the input
+already happens to be in \wellb.
+Later, when the compiler will enforce \cref{prop:w-enforcing}
+and the proof will be adjusted to cover for thew cases.
+
+\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
+\begin{gather}
+  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
+  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
+\end{gather}
 \end{lemma}
 \begin{proof}[Proof sketch]
- In this setting \Ee is as strong as
-\Eo. What we have to show is that whenever two different \Fo{}
+ In this setting \Ee is as strong as \Eo on ground terms.What we have to show is that whenever two different \Fo{}
 terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
@@
 \begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
-\fullref{prop:fidelity} hold
+\fullref{prop:fidelity} hold if $\;\wellb{(\foUnifPb)}$.
 \end{theorem}
 \begin{proof}[Proof sketch]
-Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same.
-Since all terms are $\beta$-normal and \wellb, \Ue is equivalent to \Uo.
+Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same,
+and by \cref{prop:uniffow} \Ue is equivalent to \Uo.
 \end{proof}
+
+\subsection{Notational conventions}
+
+In the following sections we adopt this notation to talk about the compiler's
+output: \foUnifPb is the input set of problems that compiled to
+\hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
+%
+\printAlll
+  {{{p_1,p_2},{p_3,p_4}}}
+  {{{t_1,t_2},{t_3,t_4}}}
+  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
+  {{{\eta,\Gamma,a,b}}}
+
+We index each sub-problem, sub-mapping, sub-link with its position
+starting from $1$ and counting from left to right, top to bottom.
+For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
+%The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.
 
@@
 
-A first problem we encounter when making unification between terms that are
-well behaved is the need to treat higher-order variables. 
+In order to make \Uo higher-order we need to
+take care of terms in \maybebeta.
+In the example below, we can see that the basic compilation given in 
+the previous section is not able to make the \Ho unification problem succeeds.
 %
@@
   {{}}
-
-In the example above, we can note that the very basilar compilation given in 
-the previous section is not able to make the \Ho unification problem succeeds.
-The unification of $T_1$ fails while trying to unifying $A \appsep x$ and $x$.
-This is due to the fact that $A \appsep x$ (equivalent to \elpiIn{app[uva A [], x]})
-is represented as the application of the variable $A$ to the name $x$.
+\todo{fix indentation}
+The unification problem $\hoUnifPb_1$ fails while trying to unifying
+$A \appsep x$ and $x$, equivalent to \elpiIn{app [uva A [], x]}
+versus \elpiIn{x}.
 In order to exploit the higher-order unification algorithm of the meta language,
-we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.
+we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$
+that is \elpiIn{uva A [x]}.
 
@@
 
-In order to address this problem, we add the following rule before rule~\ref{rule:compapp}.
+We add the following rule before rule~\ref{rule:compapp}, where
+\elpiIn{pattern-fragment} is a predicate checking is a list of
+terms is a list of distinct names.
 
@@
 Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
-is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
-(see below) grants this property. Also note that this rule generates no links.
-
-\noindent
-The only detail worth discussing is the fact that the procedure updates a
-substitution, rather than just crafting one as presented in
-section~\ref{sec:problem-statement}. The reason is that the algorithm folds
-over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
+is made of bound variables and the hypothetical rule 
+\cref{rule:complamh}
+loaded by \elpiIn{comp-lam}
+grants this property.
+
+% The only detail worth discussing is the fact that the procedure updates a
+% substitution, rather than just crafting one as presented in
+% section~\ref{sec:problem-statement}. The reason is that the algorithm folds
+% over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}
 
 \paragraph{Decompilation} Since no link is created by the compilation of 
-\maybebeta terms, no modification should be done
-to the \elpiIn{commit-link} predicate.
+\maybebeta terms, no modification to the \elpiIn{commit-link}
+is needed.
 
@@
 
-\begin{lemma}
-  \Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo
-  in \cref{sec:founif}
+\begin{definition}[$\wellb/_\beta$]
+$
+\wellb/_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
+$
+\end{definition}
+
+\begin{lemma}[Properties of \Uo]\label{prop:uniffowb} The following properties hold for \Uo
+  where
+  \begin{gather}
+    \unifcorrectX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}\\
+    \unifcompleteX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}
+  \end{gather}
+\end{lemma}
+
+  \begin{proof}[Proof sketch]
+  If we look at the \Fo{} terms the is one more interesting case,
+  namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
+  are distinct names compiled to $\vec{w}$. In this case
+  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
+   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$ that in turn
+   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
+   Thanks to \ref{clause:beta1}
+   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
+  \end{proof}
+
+\begin{lemma}[$\wellb/_\beta$-enforcement]\label{lem:w-enforcement-maybebeta}
+  Given two terms $s_1$ and $s_2$ in \maybebeta, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
+  then 
+  $$
+    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
+    t_1 \Ue t_2 \mapsto \sigma
+  $$ 
 \end{lemma}
 \begin{proof}[Proof sketch]
-  If we look at the \Fo{} terms, the is one more case interesting cases:
-  \begin{itemize}
-  \item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
-   we have $Y_{\vec{x}} \Ue t$ that succeeds with
-   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
-\todo{ste x escono da L}   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
-   Thanks to \ref{clause:beta1}
-   $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
-  \end{itemize}
+  problematic terms are mapped to uva by comp,
+  the problematic fapp node is gone.
+  if F x = t then  F = lam x.t[]
+  \todo{finish/maybe W enforcement is not needed here}
 \end{proof}
 
-\begin{lemma}[\wellb-enforcement]\label{lem:w-enforcement-maybebeta} Even if $\foUnifPb \cap \maybebeta \ne \emptyset$,
-  $\hoUnifPb \cup \maybebeta = \emptyset$
-\end{lemma}\noteD{Non capisco}
-\begin{proof}[Proof sketch]
-  problematic terms are mapped to uva by comp, the problematic fapp node is gone.
-\end{proof}
-
-\begin{theorem}[Fidelity in \maybebeta] \Fullref{prop:simulation} and
-\fullref{prop:fidelity} hold
+\begin{theorem}[Fidelity in $\wellb/_\beta$] \Fullref{prop:simulation} and
+\fullref{prop:fidelity} hold if $\;\wellb/_\beta(\foUnifPb)$
 \end{theorem}
 \begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
-  it is the same as in \cref{sec:simulation}, even if now
-  we really need \Ue{} to deal with \llambda, while before a FO unif
-  would have done.
+  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
+  \cref{prop:uniffowb}.
 \end{proof}
+
 % \subsection{Example}

### 2024-04-12 paper/main.tex
a pass on 7 and 9

@@
 % from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
-Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}. 
-There exists systems
-such as Teyjus~\cite{Nadathur2001} and $\lambda$Prolog~\cite{lamProlog} which 
-delay the resolution of \notllambda unification problems if the substitution
-is not able to put them in \llambda.
-
-In this section we want to show how we can adapt the unification of the object
-language in the meta language by simply adding (or removing) rules to the
+Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}
+and a there exist a few strategies to go bejond \llambda without 
+implementing Huet's algorithm~\ref{REF}.
+
+Some implementations
+of $\lambda$Prolog~\cite{lamProlog} 
+such as Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification problems until the substitution
+ put them in \llambda. Other systems,
+for example of the unification algorithm of Coq
+used in its type-class solver~\cite{sozeau08},
+ apply heuristics like preferring projection over mimic, and commit
+ to that solution.
+
+In this section we show how we can implement these strategies
+by simply adding (or removing) rules to the
 \elpiIn{progress} predicate.
+In the example below $\foUnifPb_1$ is in \notllambda.
 %
@@
 
-In the example above, $\foUnifPb_1$ is in \notllambda. \emph{If} the object
-language delays the first unification problem waiting $X$ to be
-be instantiated in a future unification, we can relax \cref{def:progBetaFail}. 
+If we want the object language unification to delay the first unification problem
+(waiting for $X$ to be be instantiated by a future unification),
+we can relax \cref{def:progBetaFail}. 
 Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
@@
 guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
-partial: there exists \linkbeta with a non-flexible \lhs.
-
-A second strategy to deal with problem that are in \notllambda is to make
-approximations. This is the case for example of the unification algorithm of Coq
-used in its type-class solver~\cite{sozeau08}. The approximation consists in
-forcing a choice (among the others) when the unification problem is outside
-\llambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
-argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
-that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
-\_\}$ is another valid substitution for the original problem. We stress the 
-fact that, again, our unification procedure in the meta language can be 
-accommodated for this new behavior: given a \linkbeta, if \lhs is not in \llambda,
-then \elpiIn{progress} can try to align the rightmost arguments and unify
-the resulting heads.
-
-Note that delaying unification outside \llambda can leave \linkbeta during the
-decompilation phase. Therefore, new rules to \elpiIn{commit-links} should
-be added accordingly.
+partial since we break \fullref{inv:linklhs}. %: there exists \linkbeta with a non-flexible \lhs.
+
+If we want to follow the second strategy and pick an
+arbitrary solution. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
+argument of the two terms is the same and unification can succeed by assigning
+$X a$ to $Y$. Note
+that this is not the only solution, since $\sigma = \{X \mapsto \lambda x.Y \}$
+is another valid substitution. Our unification procedure
+can be modified to accommodate this behavior by adding a
+rule to \elpiIn{progress} that tries to
+align the rightmost arguments and unify
+the resulting heads whenever \lhs of a \linkbeta is not in \llambda.
+
+Note that delaying unification outside \llambda can leave \linkbeta for the
+decompilation phase. Therefore \elpiIn{commit-links} should
+be modified accordingly.
 
@@
 % \todo{cita paper su estensioni a llam via sospensione etc}
-\todo[inline]{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
+\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
 
@@
 
-Different strategies can be used to unify terms of the object language in the
-meta langauge. One initial approach involves integrating \Uo as a procedure
-within the ML, outlined as follows:
+Different strategies can be used to unify terms of the object language.
+% in the meta langauge. 
+The first approach that comes to mind consist in implementing \Uo as a
+regular routine in the ML as follows:
 
@@
 
+\noindent
 Opting for this method would result in a suboptimal utilization of the logic
 programming engine provided by the meta language, as it degrades indexing by
-eliminating data from rule heads. Additionally, embedding \Uo within the meta
-language is likely to be significantly slower compared to a built-in
-solution.
+eliminating all data from rule heads. Additionally,
+implementing an unification procedure in the meta
+language is likely to be significantly slower compared to the built-in one.
 
 % Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
-Another possibility is to avoid having the application and abstraction nodes
+Another possibility is to avoid having application and abstraction nodes
 in the syntax tree, and use the ones of meta language, as in the following:
@@
 \noindent
-However, two reasons justify rejecting this encoding. Firstly, in CIC, it is
+However, this encoding has two big limitations. Firstly, in CIC, it is
 not always feasible to adopt it due to the meta language's type system being
-too limited to accommodate that of the object language. \\
+too limited to accommodate for that one the object language. E.g. CIC
+can typecheck variadic functions, see for example~\ref{CPDT chlipala}. \\
 Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
@@
 
-In the literature we could find related encoding of the Calculus of
-Constructions~\cite{felty93lics}. The goal of that work was to exhibit
-a logic program performing proof checking in CC and hence relate the
+In the literature we could find a related encoding of the Calculus of
+Constructions (CC)~\cite{felty93lics}. The goal of that work was to exhibit
+a logic program performing proof checking for CC and hence relate the
 proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
-programs) with the Calculus of Constructions. The encoding is hence tailored
-toward a different goal, and utilizes three relations to represent the
+programs) with the one of CC. The encoding is hence tailored
+toward a different goal, for example it utilizes three relations to represent the
 equational theory of CC. Section 6 contains a discussion about the use of the
@@
 
-Our encoding provides a third option that addresses all the considerations
-mentioned earlier. It capitalizes on the benefit of not requiring the recoding
-of the unification algorithm of the object language. Instead, it employs the
+\emph{The approach presented in this paper}
+provides a third option that addresses all the concerns
+mentioned earlier. It capitalizes on the benefit of not requiring to fully
+implement
+the unification algorithm of the object language. Instead, it employs the
 unification capabilities of the meta language, facilitated by the various links
-we establish to manage ``problematic'' subterms. Moreover, our encoding takes
-advantage of indexing algorithms and mode analysis for static clause filtering. It's noteworthy
-that we only replace the minimum necessary information with variables,
-specifically targeting \maybeeta and \notllambda subterms.
-
-Our proposed approach is highly adaptable to
-align closely with the behavior of the object language. It is not tightly
-coupled with the Coq system but can serve as a flexible framework for meta
-programming in any ML.
-
-Furthermore, the unification process we propose is tailored for potential future
-implementations of tabled search, incorporating memoization to retrieve
-solutions from previous searches.
+%we establish
+to manage ``problematic'' subterms. As a result of this choice our encoding takes
+advantage of indexing data structures and mode analysis for clause filtering.
+It it worth mentioning that we only replace terms with variables only
+when it is strictly needed, leaving the rest of the term structure intact
+and hence indexable by the meta language logic programming engine.
+Moreover, the unification process we obtain is ready to take
+advantage of potential improvements to the programming engine,
+such as tabled search, and apply forms of static analysis for
+the meta language, such as determinacy, to the object language. 
+Finally, our approach is flexible enough to accommodate different strategies
+and heuristics to handle terms outside the pattern fragment, and
+it is not tightly coupled with CIC.
 

### 2024-04-12 paper/main.tex
the abstract has to fit 1 column

@@
 
-\keywords{Logic Programming, Meta-Programming, Higher-Order Unification, Proof Automation}
+\keywords{Logic Programming, Meta-Programming, Higher-Order Unification}
 
@@
 \paragraph{Decompilation}
-A failure is thrown if any \linkbeta remains in \linkStore at the
-begin of decompilation, i.e. all \linkbeta should be solved before decompilation.
+All \linkbeta should be solved before decompilation.
+If any \linkbeta remains in \linkStore, decompilation fails.
 

### 2024-04-13 paper/main.tex
sec 3.1 with spellchecker

@@
 We simulate each run in \Fo{} with a run in \Ho as follows:
-
+%
 %Note that $\sigma_0$ is the empty substitution.
@@
 
-Since each \Ho unification variables occurs together with a scope,
+Since each \Ho unification variable occurs together with a scope,
 its assignment needs to be abstracted over it to enable the
@@
 When a single \elpiIn{fvariable} occurs multiple times with different numbers
-of arguments the compiler generates multiple mappings for it, on a first
-approximation, and then makes the mapping bijective by introducing
+of arguments, the compiler generates multiple mappings for it, on a first
+approximation, and then ensures the mapping are bijective by introducing
 \linketa; this detail is discussed in section \ref{sec:invariant1}.
@@
 % \input{code/fderef}
-It is worth looking at the code of \elpiIn{deref} that
-applies the substitution to a \Ho{} term. Remark how assignments are moved
-to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
+It is worth examining the code of \elpiIn{deref}, which
+applies the substitution to a \Ho{} term. Notice how assignments are moved
+to the current scope, i.e. the \elpiIn{abs}-bound variables are renamed
 with the names in the scope of the unification variable occurrence.
@@
 \noindent
-Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
-of the arguments of all occurrences of a unification are the same. Hence
-they have the same simple type for the meta-level and hence the number of
+Note that \elpiIn{move} strongly relies on invariant \ref{inv:uvaarity}: the length
+of the arguments of all occurrences of a unification variable is the same. Hence,
+they have the same simple type for the meta-level, and therefore the number of
 \elpiIn{abs} nodes in the assignment matches that length.
-In turn this grants that \elpiIn{move} never fails.
+This guarantees that \elpiIn{move} never fails.
 

### 2024-04-13 paper/main.tex
sec 3.2 and 3.3 with spellchecker

@@
 \noindent
-As we mentioned in section~\ref{sec:nutshell} the compiler
-replaces terms in \maybeeta, \maybebeta and \notllambda with fresh
+As mentioned in section~\ref{sec:nutshell}, the compiler
+replaces terms in \maybeeta, \maybebeta, and \notllambda with fresh
 variables linked to the problematic terms. Terms in \maybebeta do not
@@
 \noindent
-The right hand side of a link, the problematic term, can occur under binders.
-To accommodate this situation the compiler wraps \elpiIn{baselink} using
+The right-hand side of a link, the problematic term, can occur under binders.
+To accommodate this situation, the compiler wraps \elpiIn{baselink} using
 the \elpiIn{inctx} container (see~\ref{data:inctx} also used for \elpiIn{subst}).
@@
 \begin{invariant}[Link left hand side]\label{inv:linklhs}
-  The left hand side of a suspended link
+  The left-hand side of a suspended link
   is a variable.
@@
 New links are suspended by construction.
-If the left hand side is assigned during a step, then 
+If the left-hand side is assigned during a step, then 
 the link is considered for progress and possibly eliminated.
@@
 
-When detailing examples we write links as equations between two
+When detailing examples, we represent links as equations between two
 terms under a context.
-The equality sign is subscripted with
+The equality sign is subscripted with the
 kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A_x}{F_x~a}$ corresponds to:
@@
 In order to bring back the substitution from \Ho{} to \Fo{} the compiler
-builds a ``memory map'' connecting the the kind of variables using routine
-\ref{clause:malloc}.
+builds a ``memory map'' connecting the kind of variables using routine
+\ref{clause:malloc}.\noteD{non mi piace}
 
 The signature of the \elpiIn{comp} predicate below allows for the generation of
-links (suspended unification problems) that play no role in this section
+links (suspended unification problems), which play no role in this section
 but play a major role in \cref{sec:llam,sec:eta,sec:beta}.
-With respect to \cref{sec:problem-statement} the signature also allows
+With respect to \cref{sec:problem-statement}, the signature also allows
 for updates to the substitution.
@@
 The code above uses that possibility
-in order to allocate space for the variables, i.e. sets their memory
+in order to allocate space for the variables, i.e. it sets their memory
 address to \elpiIn{none} (a details not worth mentioning in the
@@
 \noindent
-In the code above the syntax \elpiIn{pi x y\..} is syntactic sugar for
+In the code above, the syntax \elpiIn{pi x y\..} is syntactic sugar for
 iterated \elpiIn{pi} abstraction, as in \elpiIn{pi x\ pi y\..}.
@@
 The auxiliary function \elpiIn{close-links} tests if the bound variable
-\elpiIn{v} really occurs in the link. If it is the case the link is wrapped into
+\elpiIn{v} really occurs in the link. If it does, the link is wrapped into
 an additional \elpiIn{abs} node binding \elpiIn{v}. In this way links generated

### 2024-04-13 paper/main.tex
sec 3.4 with spellchecker

@@
 
-A step in \Ho consists in unifying two terms and reconsidering all
-links for progress. If any of the two tasks fails we say that the entire step
-fails, and it is at this granularity that we can relate steps in the
+A step in \Ho consists of unifying two terms and reconsidering all
+links for progress. If either of these tasks fails, we consider the entire step to
+fail. It is at this granularity that we can relate steps in the
 two languages.
@@
 \noindent
-Note tha the infix notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
+Note that the infix notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
 \elpiIn{((~\Ue\!\!\!~) A B C D)}.
 
-Reconsidering links is a fixpoint, since the progress of a link can update the
-substitution and in turn enable another link to progress.
+Reconsidering links is a fixpoint process because the progress of a link
+can update the substitution, which may then enable another link to progress.
 
@@
 \subsubsection{Progress}
-In the base compilation scheme \elpiIn{progress1} is the identity
+In the base compilation scheme, \elpiIn{progress1} is the identity function
 on both the links and the substitution, so the fixpoint trivially terminates.
 Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
-and justify why the don't hinder termination.
+and explain why the don't hinder termination.
 % For brevity we omit the code
@@
 Since compilation moves problematic terms out of the sight of \Ue{},
-that procedure can only perform a partial occur check. For example the
+that procedure can only perform a partial occur check. For example, the
 unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,

### 2024-04-13 paper/main.tex
sec 3.5 with spellchecker + some todo

@@
 
-First and foremost problematic terms stored in
+First and foremost, problematic terms stored in
 \linkStore have to be moved back into the game:
@@
 This operation is possible thanks to \fullref{inv:linklhs},
-thanks to the fact that no link causes an occur-check~(\ref{sec:oc})
-and the fact that \linkStore is duplicate free (\cref{TODO}).
-
-The second step amounts at allocating new variables in the
+which ensures that no link causes an occur-check~(\ref{sec:oc})
+and the fact that \linkStore is duplicate-free (\cref{TODO}).
+
+The second step involves allocating new variables in the
 memory of \Fo{}. This technicality is required because some
-higher-order unification may need to prune a variable. For example
-$F\appsep x\appsep y = F\appsep x\appsep z$ requires to allocate a 
+higher-order unifications may require pruning a variable. For example,
+$F\appsep x\appsep y = F\appsep x\appsep z$ requires allocating a 
 variable $G$ in order to express the assignment $F_{ab} \mapsto G_a$.
 
-The last step amounts at decompiling each assignment.
-Decompiling a term is trivial since \mapStore is a bijection.
-The only tricky part concerns the \elpiIn{abs} node. In out simple setting
-the \elpiIn{flam} node carries no extra info (other then the function body),
+The final step is to decompile each assignment.
+Decompiling a term is straightforward since \mapStore is a bijection.
+The only complex part concerns the \elpiIn{abs} node. In our simple setting,
+the \elpiIn{flam} node carries no additional information (other than the function body),
 so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
-In the case of CIC, where lambdas carry the type of the bound variable, one
-has to store it somewhere. Note how this piece of information is akin
-to the arity of variables, that is CIC's unification variables have a
-(function) type, and that type can be used to annotate the lambdas
-needed in order to express their assignment.
+However, in the case of CIC, where lambdas carry the type of the bound variable, one
+must store this information somewhere. Note that this information is similar
+to the arity of variables; that in CIC, unification variables have a
+(function) type, and this type can be used to annotate the lambdas
+needed to express their assignment.
 
@@
   \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
-\end{lemma}
+\end{lemma}\noteD{Should also decompile links\dots}
 % \begin{proof}[Proof sketch]
@@
   Given a list of unification problems \foUnifPb, such that \noteD{riprendere sec 3.1}
-  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$, \noteD{dire che M e biiettivo}
+  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$ and the 
+  map memory is bijective,
   the introduction 
   of \linketa guarantees \fullref{prop:fidelity}.
-  \footnote{We also suppose that any higher-order variable is always applied with
-  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}\noteD{modificare accordingly}
+  \footnote{A map memory is bijective if any higher-order variable is applied
+  with the same number of arguments. This problem is addressed in
+  \cref{sec:invariant1}}
   \label{lemma:fidelity-maybeeta}

### 2024-04-13 paper/main.tex
5.0 reworded and shrinked, can be spell checked.

@@
 \section{Handling of \maybeeta}\label{sec:eta}
-$\eta$-reduction is an equivalence relation where a term of the form
-$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
-free variable in $t$. We call $t$ the $\eta$-contraction of
-$\lambda x. t \appsep x$.
-
-Following the compilation scheme of \cref{sec:compilation} the
-unification problem \foUnifPb is compiled as follows:
+A term 
+$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
+$x$ does not occur free in $t$, and conversely $t$ is the $\eta$-contraction of
+$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
+but the current compilation scheme does not,
+as shown by the following example:
 %
@@
 In order to guarantee \cref{prop:simulation} we detect
-lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
-we modify the compiled terms by putting fresh unification variables
-in their place: the problematic term is moved 
+lambda abstractions that can disappear by $\eta$-contraction
+(\cref{sec:etadetection}) and we modify the compiler so that it
+generates fresh unification variables
+in their place and moves the problematic term 
 from \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
@@
 \noindent
-As per \cref{inv:linklhs} the term on the left is a variable, and its
-right counterpart is the
-term in \maybeeta. That term has the following property:
+As per \cref{inv:linklhs} the \linketa left hand side is a variable
+while the right hand side is a term in \maybeeta that has the following property:
 
@@
 
-\linketa are kept in the link store \linkStore during execution
-and activated when some conditions hold on \lhs or \rhs.
-Link activation is implemented by extending the \elpiIn{progress1}
-predicate (defined in \cref{sec:execution}).
+Each \linketa is kept in the link store \linkStore during execution
+and is activated under some conditions.
+Activation is implemented by extending the \elpiIn{progress1}
+predicate defined in \cref{sec:execution}.
 

### 2024-04-14 paper/main.tex
add todo to section 6

@@
 
-We can now derive the implementation for \maybeeta detection:
+We can now describe the implementation of \maybeeta detection:
 
@@
 \begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
-  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
-  then $t \in \maybeeta$.
+  If $t$ is a $\beta$-normal term
+  and
+  $t \in \maybeeta$ then \testmaybeeta{} $t$ holds.
 \end{lemma}
@@
 \noindent
-The rule triggers when the input term \elpiIn{flam F} is in
-\maybeeta. It compiles \elpiIn{flam F} to \elpiIn{lam F1} but puts the fresh
-variable \elpiIn{A} in its place. This variable sees all the names free in
-\elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
-which relates the variable \elpiIn{A} with \elpiIn{lam F1}.
-This link clearly validates \cref{inv:linklhs}.
+TWhenever \elpiIn{flam F} is detected to be in
+\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
+variable \elpiIn{A}. This variable sees all the names free in
+\elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
+Invariant~\fullref{inv:linklhs} holds for this link, moreover:
 
@@
   The \rhs of any \linketa has exactly one lambda abstraction, hence
-  the rule above respects \cref{inv:link-eta-right}.
+  the rule above respects \fullref{inv:link-eta-right}.
   \label{cor:rhs-eta-onelamb}
@@
 \begin{proof}[Proof sketch]
-  By contradiction, suppose that the rule above is triggered and that
-  the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
-  If $\testmaybeeta{}~\lambda y.t_{xy}$ holds the recursive call to
+  By contradiction, suppose that the rule above is applied and that
+  the \rhs of the link is $\lambda x.\lambda y.t$, where $x$ and $y$ occur in $t$.
+  If $\testmaybeeta{}~\lambda y.t$ holds then the recursive call to
   \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
   in its place, so this case is impossible.
-  Otherwise, if $\testmaybeeta{}~\lambda y.t_{xy}$ does not hold, also
-  $\testmaybeeta{}~\lambda x.\lambda y.t_{xy}$ does not hold, contradicting
-  the assumption that the rule triggered.
+  Otherwise, if $\testmaybeeta{}~\lambda y.t$ does not hold, also
+  $\testmaybeeta{}~\lambda x.\lambda y.t$ does not hold, contradicting
+  the assumption that the rule was applied.
 \end{proof}
@@
 \paragraph{Decompilation}
-Decompilation of the remaining \linketa (i.e. the \linketa that have been
-activated) is performed by iterating over them and unifying \lhs and \rhs. Note
-that this unification never fails, since \lhs is a flexible term not appearing
-in any other \linketa (by \cref{def:progressetadedup}).
+The decompilation of a \linketa is performed by unifying
+the \lhs with the \rhs. Note that this unification never fails, since \lhs is a
+flexible term not appearing in any other \linketa
+(by \cref{def:progressetadedup}).
 
@@
 
-\linketa are meant to delay the unification of ``problematic'' terms until
+\linketa{}s are meant to delay the unification of ``problematic'' terms until
 we know for sure if the term has to be $\eta$-contracted or not.
@@
 $\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
-$T'$, $T'$ is a term not starting with the \elpiIn{lam} constructor. In the
+$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
 first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
@@
   Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
+  \todo{is w sigmaa defined?}
   \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
-  \progressetaleft, if 1) \lhs is a name, a constant or an
-  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
-  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux}, $\rhs = \lambda x. t$ and $\wellb(t)$.
-  Otherwise, 2) \lhs has \elpiIn{lam} as functor. In both cases, unification is
+  \progressetaleft, if 1) the \lhs is a name, a constant or an
+  application, then, $\lambda x.\lhs \appsep x$ is unified with the \rhs. By
+  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux},
+  $\rhs = \lambda x. t$ and $\wellb(t)$.
+  Otherwise, 2) the \lhs has \elpiIn{lam} as functor.
+  \todo{what does as a functor mean?}
+  In both cases, unification is
   performed between terms in \wellb.
@@
 \begin{proof}
-  The unification is done between the \rhs of two \linketa. Both \rhs has the
-  shape $\lambda x.t$, and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
-  Therefore, the unification is done between well-behaved terms.
+  We unify the \rhs{} of two \linketa. Both have the
+  shape $\lambda x.t$ and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
 \end{proof}
@@
 \begin{lemma}
-  The introduction of \linketa guarantees \fullref{prop:w-preservation}
+  The introduction of \linketa guarantees \fullref{prop:w-enforcing}
+  %\label{lemma:unif-wellb}
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  By \cref{def:testmaybeeta}.\todo{shall we move the proof here?}
+\end{proof}
+
+\begin{lemma}
+  The progress of \linketa guarantees \fullref{prop:w-preservation}
   \label{lemma:unif-wellb}
 \end{lemma}
-\todo{is this W-enforcing?}
 
@@
   % F x = z, F x y = z y
-  Given a list of unification problems \foUnifPb, such that \noteD{riprendere sec 3.1}
-  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$ and the 
-  map memory is bijective,
+  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
+  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow t \notin \notllambda$ and if the 
+  memory map is bijective, then
   the introduction 
   of \linketa guarantees \fullref{prop:fidelity}.
-  \footnote{A map memory is bijective if any higher-order variable is applied
-  with the same number of arguments. This problem is addressed in
-  \cref{sec:invariant1}}
+  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
   \label{lemma:fidelity-maybeeta}
@@
   a unification which succeeds iff the same unification in \Fo{} succeeds,
+  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule}
   guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
@@
 becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
-In turn the rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
+In turn the rule \todo{which rule} runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
 resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
@@
 $f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.
+\todo{explain why decompile does so and which eta?}
 
@@
 
-We delay at the end of  next section an example of \linketa progression due to
-\progressetaright
+We give an example for rule
+\progressetaright in the next section.
 

### 2024-04-15 paper/main.tex
sec 3.6, 3.6 and all 4 spellchecked

@@
 \noindent
-So far the compiler is very basic, it does not really
-enforce the terms passed to \hstep are in \wellb,
+So far the compiler is very basic. It does not really
+enforce that the terms passed to \hstep are in \wellb,
 and indeed makes no use of the higher-order capabilities
@@
 Still, we can prove that \Uo is a good
-``first order'' unification algorithm if the input
+``first-order'' unification algorithm if the input
 already happens to be in \wellb.
 Later, when the compiler will enforce \cref{prop:w-enforcing}
-and the proof will be adjusted to cover for thew cases.
+the proof will be adjusted to cover for the new cases.
 
@@
 \begin{proof}[Proof sketch]
- In this setting \Ee is as strong as \Eo on ground terms.What we have to show is that whenever two different \Fo{}
+ In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
 terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
-and \ref{clause:beta2} if needed) we can produce this $\rho$ by finding
+and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
 a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
-If we look at the syntax of \Fo{} terms the only interesting case is
-\elpiIn{fuva X ~\Uo~s}. In this case after compilation we have
+If we look at the syntax of \Fo{}, terms the only interesting case is
+\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
 $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
@@
 
-In the following sections we adopt this notation to talk about the compiler's
-output: \foUnifPb is the input set of problems that compiled to
+In the following sections, we adopt this notation discuss the compiler's
+output: \foUnifPb represents the input set of problems that compiled to
 \hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
@@
 
-We index each sub-problem, sub-mapping, sub-link with its position
+We index each sub-problem, sub-mapping, and sub-link with its position
 starting from $1$ and counting from left to right, top to bottom.
@@
 
-In order to make \Uo higher-order we need to
+In order to make \Uo higher-order, we need to
 take care of terms in \maybebeta.
@@
 \todo{fix indentation}
-The unification problem $\hoUnifPb_1$ fails while trying to unifying
-$A \appsep x$ and $x$, equivalent to \elpiIn{app [uva A [], x]}
+The unification problem $\hoUnifPb_1$ fails while trying to unify
+$A \appsep x$ and $x$, which is equivalent to \elpiIn{app [uva A [], x]}
 versus \elpiIn{x}.
 In order to exploit the higher-order unification algorithm of the meta language,
-we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$
-that is \elpiIn{uva A [x]}.
+we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$,
+which is \elpiIn{uva A [x]}.
 
@@
 We add the following rule before rule~\ref{rule:compapp}, where
-\elpiIn{pattern-fragment} is a predicate checking is a list of
+\elpiIn{pattern-fragment} is a predicate checking if a list of
 terms is a list of distinct names.
@@
 Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
-is made of bound variables and the hypothetical rule 
-\cref{rule:complamh}
+is made of bound variables, and the hypothetical rule 
+\ref{rule:complamh}
 loaded by \elpiIn{comp-lam}
@@
   \begin{proof}[Proof sketch]
-  If we look at the \Fo{} terms the is one more interesting case,
+  If we look at the \Fo{} terms, there is one more interesting case,
   namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
@@
   the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
-   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$ that in turn
+   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$, which in turn
    is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.

### 2024-04-15 paper/main.tex
sec 5.1 spellcheck

@@
 $\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
-$x$ does not occur free in $t$, and conversely $t$ is the $\eta$-contraction of
+$x$ does not occur free in $t$, and conversely, $t$ is the $\eta$-contraction of
 $\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
@@
 
-In order to guarantee \cref{prop:simulation} we detect
+In order to guarantee \cref{prop:simulation}, we detect
 lambda abstractions that can disappear by $\eta$-contraction
-(\cref{sec:etadetection}) and we modify the compiler so that it
+(\cref{sec:etadetection}), and we modify the compiler so that it
 generates fresh unification variables
@@
 \noindent
-As per \cref{inv:linklhs} the \linketa left hand side is a variable
-while the right hand side is a term in \maybeeta that has the following property:
+As per \cref{inv:linklhs} the \linketa left-hand side is a variable
+while the right-hand side is a term in \maybeeta that has the following property:
 
@@
 
-When compiling a term $t$ we need to determine if any
+When compiling a term $t$, we need to determine if any
 subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
@@
 The detection of lambda abstractions that can ``disappear''
-is not as trivial as it may seems, here a few examples:
+is not as trivial as it may seems. Here a few examples:
 %
@@
 The first two examples are easy, and show how a unification variable can expose
-or erase a variable in their scope and turn the resulting term in an $\eta$-expansion or not.\\
+or erase a variable in its scope, turning the resulting term into an $\eta$-expansion or not.\\
 The third example shows that when a variable occurs outside the scope of a unification
-variable it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
+variable, it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
 The last example shows the recursive nature of the check we need to implement.
-The term starts with a spine of two lambdas hence the whole term
+The term starts with a spine of two lambdas, hence the whole term
 is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
@@
 If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
-contraction. In that case the term $t$ is under the spine of binders
+contraction. In that case, the term $t$ is under the spine of binders
 $x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
@@
 
-In other words $x$ \occursrigid in $t$ if it occurs in $t$
-outside of the scope of a unification variable $X$, otherwise an instantiation
+In other words, $x$ \occursrigid in $t$ if it occurs in $t$
+outside of the scope of a unification variable $X$; otherwise, an instantiation
 of $X$ can make $x$ disappears from $t$.
@@
 
-As we will see in the rest of this section this is not a problem
+As we will see in the rest of this section, this is not a problem
 since it does not break

### 2024-04-15 paper/main.tex
sec 5.2 and 5.3 until lemma 5.11 + example eta-progress-lhs

@@
 \noindent
-TWhenever \elpiIn{flam F} is detected to be in
+Whenever \elpiIn{flam F} is detected to be in
 \maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
@@
   in its place, so this case is impossible.
-  Otherwise, if $\testmaybeeta{}~\lambda y.t$ does not hold, also
-  $\testmaybeeta{}~\lambda x.\lambda y.t$ does not hold, contradicting
+  Otherwise, if $\testmaybeeta{}~\lambda y.t$ does not hold, then also
+  $\testmaybeeta{}~\lambda x.\lambda y.t$ does not hold either, contradicting
   the assumption that the rule was applied.
@@
 $T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
-first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
+first case, $X$ is unified with $T$, and in the second one, $X$ is unified with
 $T'$ (under the context $\Gamma$).
@@
   another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
-  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
+  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same;
   hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
@@
 becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
-In turn the rule \todo{which rule} runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
+This rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
 resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
-Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
-$X$ is mapped to $B$ and
-$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.
-\todo{explain why decompile does so and which eta?}
+By the definition of \linketa decompilation,
+decompilation generates $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$, since:
+$X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$.
+% \todo{explain why decompile does so and which eta?}
 

### 2024-04-15 paper/main.tex
6 done, some todo

@@
 
-\subsection{Substitution decompilation}
+\subsection{Substitution decompilation}\label{sec:decompilation}
 
@@
 
-In \cref{sec:grounwork}, we introduced the definition of ``memory map'' (\mapStore).
-This memory allows to decompile the \Ho terms back to the object language.
-It is the case that, while solving unification problems, a same unification
+In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
+This map allows to decompile the \Ho substitution to the object language.
+Given the higher order nature of CIC the same unification
 variable $X$ is used multiple times with different arities.
@@
 $\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
-By \fullref{inv:uvaarity}, we are not allowed to use a same \Ho variable to 
-represent the two occurrences of $X$.
-If we execute \hrun, we remark that the unification fails.
-There is in fact a major problem:
-\hstep is not conscious of the connection between the variables $C$ and $E$ (both corresponding to $X$),
-since no link in \linkStore puts $C$ and $E$ in relation and decompilation
-does not work properly if a \Fo{} variable is mapped to two distinct \Ho variables. 
-The two main drawbacks connected to this situation are 
-firstly the lost of \fullref{prop:fidelity} and
-secondly, if we want to guarantee at least \fullref{prop:simulation}, we should
-overcomplicate the decompilation phase. In order to ease the second
-drawback, we pose the following property:
+In order to preserve \fullref{inv:uvaarity} the compiler
+did generate two entries in \mapStore for $X$, and in turn this
+breaks \fullref{prop:fidelity} since two incompatible assignements to
+the variables $X$ maps to, namely $E$ and $C$, would not be detected.
+We remedy by post-processing the compiler's output by adding
+additional \linketa to ensure the folllowgin property:
 
 \begin{proposition}[\mapStore is a bijection]
-  Given a list of unification problems \foUnifPb, then the memory map \mapStore
-  compiled from \foUnifPb is a bijection relating the \Fo{} and the \Ho{} 
-  variables.
+  After compilation, for each \Fo{} variable $X$ in \foUnifPb{} and for each \Ho{}
+  variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
+  (for any arity $n$).
 \end{proposition}
 
-We finally adjust the compiler's output with a \elpiIn{map-deduplication} procedure.
+\noindent
+Note that the execution in \Ho{} may require allocating new variables as
+explained in~\cref{sec:decompilation}. The property above does not
+concern these.
+
+The core procedure of this post-processing step is \emph{align-arity} that
+is iterated by \elpiIn{map-deduplication}:
 
@@
   $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
-  if \foUnifPb contains two same \Fo{} variables with different arities,
+  if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
   then \mapdeduplication guarantees \fullref{prop:fidelity}
@@
   By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
-  variables $X_1, X_2$
-  with different arities are related with \linketa. 
-  If one of the two variables is instantiated, the corresponding
-  \linketa is triggered instantiating the related variable.
-  This allows to make unification fail if $X_1$ and $X_2$ are unified with 
-  different terms. Finally, since \foUnifPb contains only terms thar are
-  either \wellb or \maybeeta, by \cref{lemma:fidelity-maybeeta},
-  we can conclude the proof.
+  variables $X_1, X_2$ with different arities are compiled to two
+  variables $A^n$ and $C^m$ related by a chain of \linketa.
+  We show that whenever one of the two is instantiated with a rigid term the 
+  the chain becomes shorter and hence, eventually, the
+  other variable is instantiated too.
+  Without loss of generality we assume $m < n$ and $A$ is instantiated.
+  The corresponding \linketa is triggered by \progressetaleft instantiating
+  the next variable in the chain of links.
 \end{proof}
+\todo{Not sure I fullfy fixed the proof, but the old oen was not ok}
+
+\noindent
+Note that shortening the chain of links allows unification to fail if
+$X_1$ and $X_2$ are unified with different terms.
 
@@
 
-In this example, $\hoUnifPb_1$ assigns $A$ which triggers $\linkStore_3$ and
-then $\linkStore_4$ by \progressetaleft. $C_{yx}$ is therefore
-assigned to $x$ (the second variable of its scope). We can finally see the
-\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ (the term $C_{xy}$
-reduces to $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
-$E_x$. After the execution of the remaining \hstep, we obtain the 
-following \Fo{} substitution $\rho = \{X := \lambda x.\lambda y.y, Y := (f\appsep \lambda x.x)\}$.
+\noindent
+The resolution of $\hoUnifPb_1$ assigns $\lambda x.\lambda y.x$ to $A$
+that in turn triggers $\linkStore_3$ and then $\linkStore_4$ by \progressetaleft,
+one per $\lambda$ in the solution of $A$.
+The unification variable $C_{yx}$ is therefore unified with $x$
+(the second variable of its scope).
+As a result $\linkStore_1$ becomes \linketaM{x}{E_x}{\lambda y.y},
+and the \rhs is no more in \maybeeta so \progressetaright fires and
+$E_x$ is unified with $\lambda y.y$, resulting in
+$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ E_x \mapsto \lambda y.y~\}$.\\
+The remaining simplify step $\hoUnifPb_2$ identified $D$ with $F$,
+hence the resulting \Fo{} substitution is
+$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto \lambda x.(f\appsep (\lambda y.y))\appsep x\}$.
+\todo{verificare, la vecchia soluzione era sballata}
 

### 2024-04-15 paper/main.tex
minor

@@
 hence the resulting \Fo{} substitution is
+\todo{verificare, la vecchia soluzione era sballata}
 $\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto \lambda x.(f\appsep (\lambda y.y))\appsep x\}$.
-\todo{verificare, la vecchia soluzione era sballata}
 

### 2024-04-15 paper/main.tex
minor

@@
 $\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ E_x \mapsto \lambda y.y~\}$.\\
-The remaining simplify step $\hoUnifPb_2$ identified $D$ with $F$,
+The remaining step $\hoUnifPb_2$ identifies $D$ with $F$,
 hence the resulting \Fo{} substitution is

### 2024-04-15 paper/main.tex
7.1

@@
 
-In this section we suppose the unification of the object language between two
-terms $t_1$ and $t_2$ to fail each time at least one of the between $t_1$ or
-$t_2$ is outside $\llambda$. This means for instance that $X \nUo Y \appsep Z$ and 
-$X\appsep Y \nUo X\appsep Y$.
-
-% On
-% the other hand, we also point out that the \maybeeta detection spot out
-% potential $\eta$-expansion for terms that are not in \llambda. For example,
-% $\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
-% application of term whose argument can reduce to $x$.
-
-% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
-Third order, functions that take functions, is undecidable, proved by Dowek.
-IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point.
-This paragraph needs fixing.}
-% \noteD[inline]{I've rewritten it, it is clearer?}
-In general, unification between \notllambda terms admits more then one
-solution and committing one of them in the substitution does not guarantee
-\cref{prop:complete-ml}. For instance, $X \appsep a \Uo a$ 
+Even if we do not requrie \Ue to handle problems outside of \llambda it is
+worth handling terms in \notllambda since, in practice, these terms often
+re-enter \llambda at runtime, as observed in~\cite{Nadathur2001} for example,
+or to accommodate for the introduction of heuristics.
+
+In the following example problem $\foUnifPb_2$ is $X \appsep a \Uo a$  that
 admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
-and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
-future unifications.
-
-Given a list of unification problems, $\foUnifPb_1\dots
-\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, it is often the case that the resolution of
-$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
-that $\rho\foUnifPb_n$ falls again in \llambda.
+and $\rho_2 = \{X \mapsto \lambda x.a\}$. The unification algorithm alone
+cannot chose, since no solution is more general than the other,
+although the first problem solve this dilemma at run-time making the second
+problem well behaved:
 %
-% make test ONLY=7003 TEX=TEX + deactivate beta tex
 \printAlll
@@
 
-In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
-$\foUnifPb_2$ can be solved in \llambda.
-On the other hand, we see that, 
-\Ue can't solve the compiled problems \hoUnifPb. In
-fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
-\mapsto \lambda x. a\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
-non-unifiable problem $(\lambda x. a) \appsep a \nUe a$.
-
-To address
-this unification problem, term compilation must recognize and replace
-\notllambda terms with fresh variables.
-This replacement produces links that we call \linkbeta.
-
-\linkbeta respects \cref{inv:linklhs} and the term on the \rhs has the
-following property:
+% In this section we suppose the unification of the object language between two
+% terms $t_1$ and $t_2$ to fail each time at least one of the between $t_1$ or
+% $t_2$ is outside $\llambda$. This means for instance that $X \nUo Y \appsep Z$ and 
+% $X\appsep Y \nUo X\appsep Y$.
+
+% On
+% the other hand, we also point out that the \maybeeta detection spot out
+% potential $\eta$-expansion for terms that are not in \llambda. For example,
+% $\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
+% application of term whose argument can reduce to $x$.
+
+% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions.
+Third order, functions that take functions, is undecidable, proved by Dowek.
+IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point.
+This paragraph needs fixing.}
+% \noteD[inline]{I've rewritten it, it is clearer?}
+% In general, unification between \notllambda terms admits more then one
+% solution and committing one of them in the substitution does not guarantee
+% \cref{prop:complete-ml}. For instance, $X \appsep a \Uo a$ 
+% admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
+% and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
+% future unifications.
+
+% Given a list of unification problems, $\foUnifPb_1\dots
+% \foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, it is often the case that the resolution of
+% $\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
+% that $\rho\foUnifPb_n$ falls again in \llambda.
+% %
+% % make test ONLY=7003 TEX=TEX + deactivate beta tex
+% \printAlll
+%   {{{X,\lambda x.a},
+%     {(X\appsep a),a}}}
+%   {{{A,\lambda x.a},
+%     {(A\appsep a),a}}}
+%   {{{X,A,0}}}
+%   {{}}
+
+\noindent
+We want to support this scenario but we have to improve a little our compiler
+and \progress routines.
+In particular $\hoUnifPb_1$ generates $\sigma = \{ A
+\mapsto \lambda x. a\}$ making $\sigma\hoUnifPb_2$ equal to
+$(\lambda x. a) \appsep a \Ue a$ that \Ue{} cannot solve since it
+lacks rules \ref{clause:beta1} and \ref{clause:beta2}.
+
+To address this problem the compiler must recognize and replace
+\notllambda terms with fresh variables and generate a new kind of links
+that we call \linkbeta.
+
+In addition to \fullref{inv:linklhs}, the term on the \rhs of a \linkbeta
+has the following property:
 
@@
 \begin{invariant}[\linkbeta \rhs]
-  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
-  unification variable with scope $s_1\dots s_n$\footnote{with $s_1\dots s_n$
-  that are distinct names} and $t_1\dots t_m$ is a list of terms. This is
-  equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$= s_1\dots s_n$
-  and \elpiIn{L}~$=t_1\dots t_m$.
+  The \rhs of any \linkbeta has the shape $X\appsep t_1\dots t_m$
+  where $X$ is a unification variable in \llambda (with scope $s_1\dots s_n$)
+  and $t_1\dots t_m$ is a list of terms such that $m>0$ and
+  $t_1$ is either a variable occurring in $s_1\dots s_n$ or
+  a term other than a variable.
   \label{inv:beta-rhs}
 \end{invariant}
+
+\noindent Note that the shape of such as \rhs is
+  <<\elpiIn{app [uva X S | L]}>>, where \elpiIn{S}~$= s_1\dots s_n$
+  and \elpiIn{L}~$=t_1\dots t_m$.
 
@@
 
-Detection of \notllambda is quite simple to implement in the compiler, since it
-is sufficient to detect applications with flexible head and argument that
-are not in \llambda. The following rule for \notllambda compilation is inserted 
-just before rule~\ref{rule:compapp}.
+For the detection of \notllambda it
+is sufficient to identify applications with a flexible head and arguments that
+are not in distinct names. The following rule for \notllambda compilation
+is inserted  just before rule~\ref{rule:compapp}.
 
@@
 
-The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
-that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
-\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
-the application of a fresh variable \elpiIn{C} having in scope all the free
-variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
-variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
-scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}.
-Note that this construction enforce \cref{inv:beta-rhs}.
+\noindent
+The list \elpiIn{Ag} is split into the two parts: 
+\elpiIn{Pf} is in \llambda, and can be empty; 
+\elpiIn{Extra} cannot be empty and is such that
+\elpiIn{append Pf Extra Ag}.
+The \rhs of the \linkbeta is
+the application of a fresh variable \elpiIn{C} having
+in scope all names in \elpiIn{Pf1} (the compilation of \elpiIn{Pf}).
+The variable \elpiIn{B}, returned has the compiled term, is a fresh variable
+having in scope all the free variables occurring in \elpiIn{Pf1} and
+\elpiIn{Extra1} (the compilation of \elpiIn{Extra}).
+This construction enforces \cref{inv:beta-rhs}.
 
@@
 
-\noteD{Si puo togliere se serve spazio}
-\begin{corollary}
-  Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
-\end{corollary}
-
-\begin{comment}
-\begin{proof}[Proof sketch]
-  Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
-  \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
-  list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
-  however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
-  no \linkbeta is produced which contradicts our initial assumption.
-\end{proof}
-\end{comment}
-
-\noteD{Si puo togliere se serve spazio}
-\begin{corollary}
-  Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
-  not a name.
-\end{corollary}
-
-\begin{comment}
-\begin{proof}[Proof sketch]
-  By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
-  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
-  $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
-  in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
-  in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
-  \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
-\end{proof}
-\end{comment}
+% \noteD{Si puo togliere se serve spazio}
+% \begin{corollary}
+%   Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
+% \end{corollary}
+
+% \begin{comment}
+% \begin{proof}[Proof sketch]
+%   Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
+%   \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
+%   list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
+%   however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
+%   no \linkbeta is produced which contradicts our initial assumption.
+% \end{proof}
+% \end{comment}
+
+% \noteD{Si puo togliere se serve spazio}
+% \begin{corollary}
+%   Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
+%   not a name.
+% \end{corollary}
+
+% \begin{comment}
+% \begin{proof}[Proof sketch]
+%   By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
+%   the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
+%   $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
+%   in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
+%   in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
+%   \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
+% \end{proof}
+% \end{comment}
 

### 2024-04-15 paper/main.tex
section 7

@@
 Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
-$3$ different activation rules:
+the following activation rules:
 
 \begin{definition}[\progBetaLL]
-  Given a substitution $\sigma$, where $\sigma t_1$ is a name, say $t$, and
-  $t\notin s_1\dots s_n$. If $m = 0$, then $l$ is removed and \lhs is
-  unified with \rhsBetaHead. If $m > 0$, then $l$ is replaced by a
-  refined version $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots
-  t_m}$ with reduced list of arguments and $Y$ being a fresh variable. Moreover,
-  the new link \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}}
+  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
+  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
+  unified with $X_{s_1\ldots s_n~s}$.
+  \todo{I changed this part, it is OK?}
+  If $m > 1$, then $l$ is replaced by the link
+  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
+  (where $Y$ is a fresh variable of arity $n+1$) and link
+  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
   is added to \linkStore.
@@
 \begin{definition}[\progBetaRH]
-  $l$ is removed from
-  \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
-  $\beta$-reduced term $t'$ obtained from the application of $t$ to
-  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified with $t$.
+  Link $l$ is removed from
+  \linkStore if \rhsBetaHead is instantiated to a term $t$ and
+  $t \appsep t_1 \ldots t_m$ $\beta$-reduces to a $t' \in \llambda$.
   \label{def:progBetaRH}
@@
 \begin{definition}[\progBetaFail]
-  If it exists a link $l' \in \linkStore$ with same \lhs as $l$, or the \lhs 
-  of $l$ become rigid, then unification fail.
+  \progress fails whenever:
+  there exists another link $l' \in \linkStore$ with same \lhs as $l$;
+  or the \lhs  of $l$ become rigid.
   \label{def:progBetaFail}
 \end{definition}
+
+\noindent In order to accommodate for the implementation of heuristics we
+shall relax these failure conditions in~\cref{sec:heuristics}.
 
@@
   The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
+  if \Uo{} fails outside \llambda.
 \end{theorem}
@@
 \begin{proof}[Proof sketch]
-  Let \hoUnifPb a unification problem and $\sigma$ a substitution
-  such that $\hoUnifPb \in \notllambda$. 
-  If $\sigma\hoUnifPb$ is in \llambda, then by
-  \cref{def:progBetaLL,def:progBetaRH}, the \linkbeta associated to the subterm of
-  \hoUnifPb have been solved and removed. The
-  unification is done between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
-  fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
-  then, by \cref{def:progBetaFail}, the unification fails, as per the
-  corresponding unification in \Fo{}.
+  Let $\foUnifPb_i$ be the first problem in \foUnifPb not in \llambda and
+  le $\sigma$ be the substitution obtained solving $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$.
+  If $\sigma\hoUnifPb_i$ is in \llambda, then each variable that the
+  compiler used to replace a problemat term is assigned and hence
+  by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
+  been solved and removed. Note that in this case all
+  calls to \Ue{} are between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
+  fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,
+  then by \cref{def:progBetaFail} unification fails as the
+  corresponding unification in \Fo{} would.
 \end{proof}
@@
 the second \linketa ($f$ is assigned to $C$), and then all the remaining links
-are solved. The final \Ho substitution is $\sigma = \{A _{} \mapsto \lambda x.x, 
-B _{} \mapsto f, C _{x} \mapsto (f\appsep x), 
-D _{} \mapsto f, E _{x} \mapsto x, F _{x} \mapsto C_{x} \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
-Y \mapsto f\}$.
+are solved. The final \Ho substitution is $\sigma = \{~A _{} \mapsto \lambda x.x ~;~
+B _{} \mapsto f ~;~ C _{x} \mapsto (f\appsep x) ~;~
+D _{} \mapsto f ~;~ E _{x} \mapsto x ~;~
+F _{x} \mapsto C_{x} ~\}$ and is decompiled into $\rho = \{~X \mapsto \lambda x.x ~;~
+Y \mapsto f~\}$.
 
 \paragraph{Example of \progBetaRH}
-We can take the example provided in \cref{sec:beta}. The problem is compiled
-into:
+We give a second look at the opening example in \cref{sec:beta}.
+\todo{this is a little more complex that the opening problem}
+The problem is compiled into:
 %
@@
 
-The first unification problems is solved by the substitution $\sigma = \{A
-\mapsto \lambda x.B\}$. The \linkbeta becomes
+The first unification problems is solved by the substitution
+$\sigma = \{~A \mapsto \lambda x.B~\}$. The \linkbeta becomes
 \linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
@@
 unification problem
-gives the final substitution $\sigma = \{A \mapsto \lambda x.B, B \mapsto C, C
-\mapsto a\}$ which is decompiled into $\rho = \{X \mapsto \lambda x.a, Y \mapsto
-a\}$.
+gives the final substitution $\sigma = \{~A \mapsto \lambda x.B ~;~ B \mapsto C ~;~ C
+\mapsto a ~\}$ that is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
+a ~\}$.
 
 \subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
-
+\label{sec:heuristics}
 \newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}
@@
 % from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
-Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}
-and a there exist a few strategies to go bejond \llambda without 
-implementing Huet's algorithm~\ref{REF}.
-
-Some implementations
-of $\lambda$Prolog~\cite{lamProlog} 
-such as Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification problems until the substitution
- put them in \llambda. Other systems,
-for example of the unification algorithm of Coq
-used in its type-class solver~\cite{sozeau08},
- apply heuristics like preferring projection over mimic, and commit
- to that solution.
+Working with terms in \llambda is sometime too restrictive~\cite{Abel2018ExtensionsTM}
+and we could find in literature a few strategies to go beyond \llambda without 
+implementing Huet's algorithm~\cite{Huet75}.
+Some implementations of $\lambda$Prolog~\cite{lamProlog} such as
+Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification
+problems until the substitution makes them reenter \llambda. Other systems, for
+example of the unification algorithm of Coq used in its type-class
+solver~\cite{sozeau08}, apply heuristics like preferring projection over mimic,
+and commit to that solution.
 
@@
   {{{(X\appsep a),a},
-    {X,\lambda x.a}}}
-  {{}}
-  {{}}
-  {{}}
-
+    {X,\lambda x.Y}}}
+  {{{A,a},
+    {B,\lambda x.C}}}
+  {{{Y,C,0},
+    {X,B,0}}}
+  {{{\llambda,,A,(B\appsep a)}}}
+
+\noindent
 If we want the object language unification to delay the first unification problem
-(waiting for $X$ to be be instantiated by a future unification),
+(waiting for $X$ to be be instantiated),
 we can relax \cref{def:progBetaFail}. 
-Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
+Instead of failing when the \lhs of $\linkStore_1$ becomes rigid (equal to $a$),
 we keep it in \linkStore until the head of its \rhs also become rigid.
-In this case, since \lhs and \rhs have rigid heads, they can be unified
-just before removing $l$ from \linkStore. We can note that this rule trivially
-guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
-partial since we break \fullref{inv:linklhs}. %: there exists \linkbeta with a non-flexible \lhs.
-
-If we want to follow the second strategy and pick an
-arbitrary solution. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
-argument of the two terms is the same and unification can succeed by assigning
-$X a$ to $Y$. Note
-that this is not the only solution, since $\sigma = \{X \mapsto \lambda x.Y \}$
-is another valid substitution. Our unification procedure
-can be modified to accommodate this behavior by adding a
-rule to \elpiIn{progress} that tries to
-align the rightmost arguments and unify
-the resulting heads whenever \lhs of a \linkbeta is not in \llambda.
-
-Note that delaying unification outside \llambda can leave \linkbeta for the
+In this case, since both the \lhs and \rhs have rigid heads, they can be unified.
+While this relaxed rule does not break \fullref{prop:fidelity} per se,
+the \elpiIn{occur-check-links} procedure becomes incomplete since
+\fullref{inv:linklhs} is broken. Also note that delaying unification outside \llambda can leave \linkbeta for the
 decompilation phase. Therefore \elpiIn{commit-links} should
@@
 
-% \noteD{Paragraph for decompile in case of \maybebeta}
-% \todo{cita paper su estensioni a llam via sospensione etc}
-\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
+If instead we want \Uo{} to follow the second strategy and pick an
+arbitrary solution we can modify \progress by applying the desired heuristic.
+For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
+argument of the two terms is the same and unification can succeed by assigning
+$X a$ to $Y$. This heuristic can be implemented by adding the following rule:
+
+\input{code/progress_heuristic}
+
+\noindent
+Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}.
 

### 2024-04-15 paper/main.tex
section 8

@@
 
-In this paper we show a minimized example. The full code is there.
-But we also have to code things in Coq-Elpi.
-
-The main difference between the presentation in the previous sections and
-the actual implementation for Coq is that the main loop \hrun is replaced by
-the one of Prolog that chains calls to the unification procedure. In order
-implement the store of links we resort to Elpi's CLP engine and
-use constraints (suspended goals) to represent links, and constraint
-handling rules to implement progress operations involving more than one link.
-
-\todo{finish}
-about the progress of 1 link:
+In this paper we did study a compiler and simulation loop on a minimal language.
+The actual implementation uses the Coq-Elpi meta language to both compile
+the sequence of problems (the rules) and execute them, that is Elpi
+plays the role of \Ho{} as well.
+
+The main difference is that we cannot implement \hrun since it comes as part of
+the runtime of the programming language. In particular the runtime iterates
+\Ue, but \hstep also needs to check for \progress. Luckily Elpi extends~\cite{TASSI_2019}
+$\lambda$Prolog with syntactic constraints (suspended goals) and
+Constraint Handling Rules (CHR)~\cite{fruehwirth2017constraint} to operate on them.
+An important design choice is that constraints are suspended on a user
+computed list of unification variables, and are resumed as soon as one of them
+is assigned, before any other existing goal is considered. In turn this
+property enables a timely activation of links that grants~\fullref{prop:fidelity}.
+For brevity we only provide two pseaudo-code snippets. The first one depicts
+how \linketa are suspended or make \progress.
 
 \begin{elpicode}
-link-eta L R :- suspend-condition L R Holes, !,
-  declare_constraint (link-eta L R) Holes.
-link-eta L R :-
-  progress. % e.g. L = R.
+link-eta L R :- var L, maybe-eta R HolesInR, !,
+  declare_constraint (link-eta L R) [L|HolesInR].
+link-eta L R :- not (var L), !, eta-progress-lhs L R.
+link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
 \end{elpicode}
-
-about the progress of 2 links:
+\todo{improve code}
+
+\noindent
+The second snippet illustrate the deduplication of \linketa. The syntax
+<<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 
+goal \elpiIn{P} under a set \elpiIn{G} of hypothetical rules (introduced
+by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
+operator) \elpi{N} heigenvariables. Sequents are presented to
+CHR with their higher order unification variables replaced by
+``frozen constants'', that is $A_{xyz}$ becomes <<\elpiIn{uvar ~$f_A$~ [x,y,z]}>>
+for a fresh constant $f_A$. Frozen constants are ``defrost'' when
+they are part of a new goal (see \cite[section 4.3]{TASSI_2019}).
 
@@
 
-Remark how the invariant about uvar arity makes this easy, since LX1 and LX2
-have the same length. Also note that N1 only contains the names of the first
-link (while relocate runs in the disjoint union) and Elpi ensures that
-T2' can live in N1.
+\noindent
+The first directive matches a constraint
+whose \lhs is a variable \elpiIn{X}; the second mathes \emph{and removes}
+a constrain on the same variable; the third is a test that accessorily
+relocates \elpiIn{T2} and the latter crafs the new goal that unifies
+\elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
+heigenvariables \elpiIn{N1}.
+
+% Remark that \fullref{inv:uvaarity} makes it possible to relocate \elpiIn{T2}
+% using the name-to-name mapping obtained by zipping
+% the lists (of equal length) \elpiIn{LX1} and \elpiIn{LX2}.
 

### 2024-04-15 paper/main.tex
9

@@
     |  (relocate LX1 LX2 T2 T2')               % condition
-   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                % new goal
-}
+   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                } % new goal
 \end{elpicode}
@@
 The first approach that comes to mind consist in implementing \Uo as a
-regular routine in the ML as follows:
+regular routine in the ML, i.e. write rules as follows:
 
@@
 Opting for this method would result in a suboptimal utilization of the logic
-programming engine provided by the meta language, as it degrades indexing by
+programming engine provided by the ML, as it degrades indexing by
 eliminating all data from rule heads. Additionally,
-implementing an unification procedure in the meta
-language is likely to be significantly slower compared to the built-in one.
-
-% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
+implementing a unification procedure in the ML is likely to be significantly
+slower compared to the built-in one.
+
 Another possibility is to avoid having application and abstraction nodes
-in the syntax tree, and use the ones of meta language, as in the following:
+in the syntax tree, and use the ones of the ML, as in:
 
@@
 \noindent
-However, this encoding has two big limitations. Firstly, in CIC, it is
-not always feasible to adopt it due to the meta language's type system being
-too limited to accommodate for that one the object language. E.g. CIC
-can typecheck variadic functions, see for example~\ref{CPDT chlipala}. \\
-Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
+However, this encoding has two big limitations. First it is
+not always feasible to adopt it for CIC due to the fact that the type system
+of the ML is too limited to accommodate for that one the object language,
+e.g. CIC can typecheck variadic functions~\cite{cpdt}. \\
+Second, the CIC encoding provided by Elpi is primarily utilized for meta
 programming, in order to extend the Coq system. Consequently, it must be able to
@@
 an open world, akin to the \elpiIn{string} data type used in the preceding
-examples.
+examples, they cannot be symbols of the ML.
 
 In the literature we could find a related encoding of the Calculus of
-Constructions (CC)~\cite{felty93lics}. The goal of that work was to exhibit
+Constructions (CC)~\cite{felty93lics}. The goal of that work is to exhibit
 a logic program performing proof checking for CC and hence relate the
@@
 toward a different goal, for example it utilizes three relations to represent the
-equational theory of CC. Section 6 contains a discussion about the use of the
-unification procedure of the meta language in presence of non ground goals, but
+equational theory of CC, and that choice alone makes things arder for us.
+Section 6 contains a discussion about the use of the
+unification procedure of the ML in presence of non ground goals, but
 the authors do not aim at exploiting it to the degree we want.\todo{sucks}
 
-\todo{cite isabelle's TC, that are baked in}
+Another work that is, surprisingly, only superficially related to ours is the
+type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
+classes identify (simple) types, they are not higher order predicates
+as in Coq, hence the solver does not require a higher-order unification
+procedure.
 

### 2024-04-15 paper/main.tex
minor

@@
 \input{code/progress_heuristic}
-
-\noindent
-Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}.
-
+%\noindent
+%Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}. 
+\vspace{-1em}
 \section{Actual implementation in Elpi}\label{sec:implementation}
@@
 \begin{elpicode}
-constraint link-eta {
   rule (N1 ~$\triangleright$~ G1 ?- link-eta (uvar X LX1) T1)   % match
@@
     |  (relocate LX1 LX2 T2 T2')               % condition
-   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                } % new goal
+   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                  % new goal
 \end{elpicode}
@@
 The first directive matches a constraint
-whose \lhs is a variable \elpiIn{X}; the second mathes \emph{and removes}
-a constrain on the same variable; the third is a test that accessorily
+whose \lhs is a variable \elpiIn{X}; the second mathes and removes
+a constraint on the same variable; the third
 relocates \elpiIn{T2} and the latter crafs the new goal that unifies
@@
 toward a different goal, for example it utilizes three relations to represent the
-equational theory of CC, and that choice alone makes things arder for us.
+equational theory of CC, and that choice alone makes things harder for us.
 Section 6 contains a discussion about the use of the
@@
 
-Another work that is, surprisingly, only superficially related to ours is the
+Another work that is, somewhat surprisingly, only superficially related to ours is the
 type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
@@
 
+We considered mechanizing our results with Abella, especially in the early phases.
+Unfortunately we could not. One reason
+is that eliminating the cut operator, although possible, has a high cost in
+verbosity.
+% he same for the \elpiIn{name}
+%\todo{is it sill described?} predicate.
+Another problem is that the logic of Abella does not let one quantify over
+predicates and combinators like \elpiIn{forall} are used pervasively in our
+code.
+\todo{finish}
+
 % ---
@@
 
-\clearpage
-
-\input{appendix.tex}
+% \clearpage
+
+% \input{appendix.tex}
 

### 2024-04-15 paper/main.tex
typos

@@
   ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
-  con C ~\Ee~fcon C.
-  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
-  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
-  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
+  con C ~\Ee~con C.
+  app A ~\Ee~app B :- forall2 ~(\Ee)~ A B.
+  lam F ~\Ee~lam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
+  uva N A ~\Ee~uva N B :- forall2 ~(\Ee)~ A B.
 \end{elpicode}
@@
 a sequence of \emph{steps} of length $\mathcal{N}$.
-At each step, $p$ we unify two terms, $\foUnifPb_{p_l}$ and
+At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
 $\foUnifPb_{p_r}$, taken from the list of all unification
@@
 description is deferred to \cref{sec:nutshell}.
-We write $\hoUnifPb_p = \{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$
+We write $\foUnifPb_p = \{~ \foUnifPb_{p_l}, \foUnifPb_{p_r} ~\}$
 and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.
@@
 
-\paragraph{Backtracking} We omit it from our model of a logic program's execution
+\paragraph{Backtracking} We omit it from our model of a logic-program execution
 since it plays a very minor role, orthogonal to higher-order unification.
@@
 input is not.
-Note that the property holds for any substitution. $\rho$ could be given by an
+Note that the property holds for any substitution, $\rho$ could be given by an
 oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply

### 2024-04-15 paper/main.tex
todos
add todo

@@
 \end{theorem}
+\todo{should we introduce W/eta as prev section W/beta?}
 
@@
 \end{theorem}
+\todo{should we introduce W/llam as prev section W/beta and W/eta?}
 
@@
 the meta language, such as determinacy, to the object language.
+\todo{mention preliminar experiment show that our prolog is 2x faster than the Coq one} 
 Finally, our approach is flexible enough to accommodate different strategies

### 2024-04-15 paper/main.tex
solve a todo

@@
 We give a second look at the opening example in \cref{sec:beta}.
-\todo{this is a little more complex that the opening problem}
+% \todo{this is a little more complex that the opening problem}
 The problem is compiled into:
@@
 \printAlll
-  {{{X,\lambda x.Y},
+  {{{X,\lambda x.a},
     {(X\appsep a),a}}}
-  {{{A,\lambda x.B},
-    {C,a}}}
-  {{{Y,B,0},
-    {X,A,0}}}
-  {{{\llambda,,C,(A\appsep a)}}}
+  {{{A,\lambda x.a},
+    {B,a}}}
+  {{{X,A,0}}}
+  {{{\llambda,,B,(A\appsep a)}}}
 
@@
 
-The first unification problems is solved by the substitution
-$\sigma = \{~A \mapsto \lambda x.B~\}$. The \linkbeta becomes
-\linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
-$B$. $B$ is in \llambda and is unified with $C$. The resolution of the second
+The first unification problem is solved with the substitution
+$\sigma = \{~A \mapsto \lambda x.a~\}$. The \linkbeta becomes
+\linkbetaM{}{B}{((\lambda x.a) \appsep a)} whose \rhs can be $\beta$-reduced to
+$a$. $a$ is in \llambda and is unified with $B$. The resolution of the second
 unification problem
-gives the final substitution $\sigma = \{~A \mapsto \lambda x.B ~;~ B \mapsto C ~;~ C
-\mapsto a ~\}$ that is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
+confirm that $B$ equal $a$. The final substitution $\sigma = \{~A \mapsto \lambda x.a ~;~ B \mapsto a ~\}$ is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
 a ~\}$.

### 2024-04-15 paper/main.tex
solve 1 todo

@@
 % make test ONLY=7006 TEX=tex
-\printAlll
+\printAlllSingle
   {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
@@
   {{}}
-\todo{fix indentation}
 The unification problem $\hoUnifPb_1$ fails while trying to unify

### 2024-04-15 paper/main.tex
attempt

--- 2024-04-15 paper/main.tex
@@
 \begin{proposition}[\wellb-enforcing]\label{prop:w-enforcing}
+  Given a term $s$ in \Fo{}, 
+  $$
+    \C{s}{t}{m}{l} \Rightarrow \wellb(t)
+  $$
+\end{proposition}
+
+\begin{proposition}[compiler-useful]\label{prop:w-noncontra}
   Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
@@
 In other words the compiler outputs terms in \wellb, even if its
-input is not.
+input is not. And if two terms can unify in the source language
+then these terms also unify in the target language. 
 Note that the property holds for any substitution, $\rho$ could be given by an
 oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
-does not contradict \Eo.
+does not contradict \Eo. Also note that a compiler that
+translate all terms to unification variables would satisfy this
+property, but wound fail to verify \cref{prop:fidelity}.
 

### 2024-04-16 paper/main.tex
w-enforcement

@@
   \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
-\end{gather}
+\end{gather}\todo{give a name and reuse it for \Uo in next sections}
 
@@
 
-\begin{lemma}[Compilation round trip]\label{prop:comprt} If
-  \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
-\end{lemma}\noteD{Should also decompile links\dots}
+\begin{proposition}[Compilation round trip]\label{prop:comprt}
+  % \elpiIn{compile F T [] M [] L [] S} then 
+  % if \elpiIn{T} $\in$ S
+  If $\C{s}{t}{m}{l}$, such that $l \in \linkStore$ and $m \in \mapStore$, if
+  $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
+  then $\D{\sigma}{\mapStore}{\linkStore}{\rho}$ and
+  $\rho X \Eo \rho s$.
+\end{proposition}
+\todo{explain no proof, llam link breaks}
+
+% \noteD{Should also decompile links\dots}
 % \begin{proof}[Proof sketch]
@@
 
-\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
+\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow} The following properties hold for \Uo
+  as implemented above: \todo{reuse statement from section 2.4}
 \begin{gather}
-  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
-  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
+  \unifcorrectX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}\\
+  \unifcompleteX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}
 \end{gather}
@@
 $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
-$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
+$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
 \end{proof}
@@
 
-\begin{definition}[$\wellb/_\beta$]
+\begin{definition}[$\wellb_\beta$]
 $
-\wellb/_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
+\wellb_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
 $
@@
 
-\begin{lemma}[Properties of \Uo]\label{prop:uniffowb} The following properties hold for \Uo
-  where
+\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb} The following properties hold for \Uo
+  where \todo{reuse 2.4}
   \begin{gather}
-    \unifcorrectX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}\\
-    \unifcompleteX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}
+    \unifcorrectX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}\\
+    \unifcompleteX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}
   \end{gather}
@@
 
-\begin{lemma}[$\wellb/_\beta$-enforcement]\label{lem:w-enforcement-maybebeta}
-  Given two terms $s_1$ and $s_2$ in \maybebeta, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
-  then 
-  $$
-    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
-    t_1 \Ue t_2 \mapsto \sigma
-  $$ 
+\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
+  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
 \end{lemma}
-\begin{proof}[Proof sketch]
-  problematic terms are mapped to uva by comp,
-  the problematic fapp node is gone.
-  if F x = t then  F = lam x.t[]
-  \todo{finish/maybe W enforcement is not needed here}
-\end{proof}
-
-\begin{theorem}[Fidelity in $\wellb/_\beta$] \Fullref{prop:simulation} and
-\fullref{prop:fidelity} hold if $\;\wellb/_\beta(\foUnifPb)$
+% \begin{proof}[Proof sketch]
+%   problematic terms are mapped to uva by comp,
+%   the problematic fapp node is gone.
+%   if F x = t then  F = lam x.t[]
+%   \todo{finish/maybe W enforcement is not needed here}
+% \end{proof}
+\noindent
+In other word the compiler eliminates all subterms that are in \maybebeta.
+
+\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
+\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$
 \end{theorem}
@@
 
+\begin{definition}[$\wellb_{\beta \eta}$]
+  $
+  \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
+  $
+\end{definition}
+
+\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybeeta}
+  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_{\beta \eta}(s)$  then $\:\wellb(t)$.
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  By \cref{lem:maybeeta}
+\end{proof}
+
 \paragraph{Decompilation}
@@
   shape $\lambda x.t$ and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
-\end{proof}
-
-\begin{lemma}
-  The introduction of \linketa guarantees \fullref{prop:w-enforcing}
-  %\label{lemma:unif-wellb}
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-  By \cref{def:testmaybeeta}.\todo{shall we move the proof here?}
 \end{proof}
@@
 
-\begin{theorem}[Fidelity in \maybeeta]
+\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
   % F x = z, F x y = z y
   Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
-  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow t \notin \notllambda$ and if the 
+  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
   memory map is bijective, then
@@
 
+% \begin{definition}[$\wellb_{\beta \eta \llambda}$]
+%   $
+%   \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
+%   $
+% \end{definition}
+
+\begin{lemma}[$\wellb$-enforcement]\label{lem:w-enforcement-maybellam}
+  $\forall s, \C{s}{t}{m}{l}$, then $\:\wellb(t)$.
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  By \cref{lem:w-enforcement-maybebeta,lem:w-enforcement-maybeeta} and the 
+  rule above.
+\end{proof}
+
 \paragraph{Decompilation}
@@
   unified with $X_{s_1\ldots s_n~s}$.
-  \todo{I changed this part, it is OK?}
   If $m > 1$, then $l$ is replaced by the link
@@
   it makes \elpiIn{progress} immediately fails.
-
+%
   % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
@@
 
-\begin{theorem}[Fidelity with \linkbeta]
+\begin{theorem}[Fidelity in \Fo]
   The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
@@
 \end{theorem}
-\todo{should we introduce W/llam as prev section W/beta and W/eta?}
 

### 2024-04-16 paper/main.tex
update tex

@@
 In order to bring back the substitution from \Ho{} to \Fo{} the compiler
-builds a ``memory map'' connecting the kind of variables using routine
-\ref{clause:malloc}.\noteD{non mi piace}
+builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
+using the routine
+\ref{clause:malloc}.
 
@@
 in order to allocate space for the variables, i.e. it sets their memory
-address to \elpiIn{none} (a details not worth mentioning in the
+address to \elpiIn{none} (a detail not worth mentioning in the
 previous sections).
@@
 \begin{lemma}
-  Given a \linketa $l$, the unification done by \progressetaleft is between
+  Given a \linketa \linketaM{\Gamma}{X}{T}, the unification done by \progressetaleft is between
   terms in $\wellb$  
@@
 \begin{proof}[Proof sketch]
-  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
-  \todo{is w sigmaa defined?}
-  \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
-  \progressetaleft, if 1) the \lhs is a name, a constant or an
-  application, then, $\lambda x.\lhs \appsep x$ is unified with the \rhs. By
+  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb),
+  By\todo{reformulate}
+  \progressetaleft, if 1) the $\sigma X$ is a name, a constant or an
+  application, then, $\lambda x.X \appsep x$ is unified with the \rhs. By
   \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux},
   $\rhs = \lambda x. t$ and $\wellb(t)$.
-  Otherwise, 2) the \lhs has \elpiIn{lam} as functor.
-  \todo{what does as a functor mean?}
+  Otherwise, 2) $\sigma X = \lambda x.t$.
+  % \todo{what does as a functor mean?}
   In both cases, unification is
@@
   % F x = z, F x y = z y
-  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
+  Given a list of unification problems \foUnifPb, if
   $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
@@
 \end{theorem}
-\todo{should we introduce W/eta as prev section W/beta?}
 
@@
   a unification which succeeds iff the same unification in \Fo{} succeeds,
-  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule}
+  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule, citare regolae in =o}
   guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
@@
   Given a list of unification problems \foUnifPb, such that 
-  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
+  $\wellb_{\beta \eta}(\foUnifPb)$,
   if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
@@
   By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
-  variables $X_1, X_2$ with different arities are compiled to two
+  variables with different arities are compiled to two
   variables $A^n$ and $C^m$ related by a chain of \linketa.
-  We show that whenever one of the two is instantiated with a rigid term the 
+  We show that whenever one of the two is instantiated with a rigid term then
   the chain becomes shorter and hence, eventually, the
@@
 and the \rhs is no more in \maybeeta so \progressetaright fires and
-$E_x$ is unified with $\lambda y.y$, resulting in
-$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ E_x \mapsto \lambda y.y~\}$.\\
+$E_x$ is unified with $\lambda y.y$; then $\linkStore_2$ makes progress thanks to the same 
+rule giving
+$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ D \mapsto f\appsep (\lambda y.y) ~;~ E_x \mapsto \lambda y.y ~\}$.\\
 The remaining step $\hoUnifPb_2$ identifies $D$ with $F$,
 hence the resulting \Fo{} substitution is
-\todo{verificare, la vecchia soluzione era sballata}
-$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto \lambda x.(f\appsep (\lambda y.y))\appsep x\}$.
+$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto f\appsep (\lambda y.y)\}$.
 
@@
 by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
-operator) \elpi{N} heigenvariables. Sequents are presented to
+operator) \elpiIn{N} heigenvariables. Sequents are presented to
 CHR with their higher order unification variables replaced by

### 2024-04-16 paper/main.tex
wip

--- 2024-04-16 paper/main.tex
@@
 \noindent
-We now illustrate the encoding of the three instances above as higher-order
+We now illustrate a naive encoding of the three instances above as higher-order
 logic programming rules: capital letters denote rule
@@
 \noindent
-Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work
+Unfortunately this encoding of rule \ref{clause:r3} does not work
 since it uses the predicate \coqIn{P} as a first order term: for the meta
@@
   \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
-\end{gather}\todo{give a name and reuse it for \Uo in next sections}
+\end{gather}
 
@@
 \begin{proposition}[\wellb-enforcing]\label{prop:w-enforcing}
-  Given a term $s$ in \Fo{}, 
-  $$
-    \C{s}{t}{m}{l} \Rightarrow \wellb(t)
-  $$
-\end{proposition}
-
-\begin{proposition}[compiler-useful]\label{prop:w-noncontra}
   Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
@@
 In other words the compiler outputs terms in \wellb, even if its
-input is not. And if two terms can unify in the source language
-then these terms also unify in the target language. 
+input is not.
 Note that the property holds for any substitution, $\rho$ could be given by an
 oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
-does not contradict \Eo. Also note that a compiler that
-translate all terms to unification variables would satisfy this
-property, but wound fail to verify \cref{prop:fidelity}.
+does not contradict \Eo.
 
@@
 
-\begin{proposition}[Compilation round trip]\label{prop:comprt}
-  % \elpiIn{compile F T [] M [] L [] S} then 
-  % if \elpiIn{T} $\in$ S
-  If $\C{s}{t}{m}{l}$, such that $l \in \linkStore$ and $m \in \mapStore$, if
-  $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
-  then $\D{\sigma}{\mapStore}{\linkStore}{\rho}$ and
-  $\rho X \Eo \rho s$.
-\end{proposition}
-\todo{explain no proof, llam link breaks}
-
-% \noteD{Should also decompile links\dots}
+\begin{lemma}[Compilation round trip]\label{prop:comprt} If
+  \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
+\end{lemma}\noteD{Should also decompile links\dots}
 % \begin{proof}[Proof sketch]
@@
 
-\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow} The following properties hold for \Uo
-  as implemented above: \todo{reuse statement from section 2.4}
+\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
 \begin{gather}
-  \unifcorrectX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}\\
-  \unifcompleteX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}
+  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
+  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
 \end{gather}
@@
 $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
-$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
+$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
 \end{proof}
@@
 
-\begin{definition}[$\wellb_\beta$]
+\begin{definition}[$\wellb/_\beta$]
 $
-\wellb_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
+\wellb/_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
 $
@@
 
-\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb} The following properties hold for \Uo
-  where \todo{reuse 2.4}
+\begin{lemma}[Properties of \Uo]\label{prop:uniffowb} The following properties hold for \Uo
+  where
   \begin{gather}
-    \unifcorrectX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}\\
-    \unifcompleteX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}
+    \unifcorrectX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}\\
+    \unifcompleteX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}
   \end{gather}
@@
 
-\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
-  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
+\begin{lemma}[$\wellb/_\beta$-enforcement]\label{lem:w-enforcement-maybebeta}
+  Given two terms $s_1$ and $s_2$ in \maybebeta, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
+  then 
+  $$
+    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
+    t_1 \Ue t_2 \mapsto \sigma
+  $$ 
 \end{lemma}
-% \begin{proof}[Proof sketch]
-%   problematic terms are mapped to uva by comp,
-%   the problematic fapp node is gone.
-%   if F x = t then  F = lam x.t[]
-%   \todo{finish/maybe W enforcement is not needed here}
-% \end{proof}
-\noindent
-In other word the compiler eliminates all subterms that are in \maybebeta.
-
-\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
-\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$
+\begin{proof}[Proof sketch]
+  problematic terms are mapped to uva by comp,
+  the problematic fapp node is gone.
+  if F x = t then  F = lam x.t[]
+  \todo{finish/maybe W enforcement is not needed here}
+\end{proof}
+
+\begin{theorem}[Fidelity in $\wellb/_\beta$] \Fullref{prop:simulation} and
+\fullref{prop:fidelity} hold if $\;\wellb/_\beta(\foUnifPb)$
 \end{theorem}
@@
 
-\begin{definition}[$\wellb_{\beta \eta}$]
-  $
-  \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
-  $
-\end{definition}
-
-\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybeeta}
-  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_{\beta \eta}(s)$  then $\:\wellb(t)$.
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-  By \cref{lem:maybeeta}
-\end{proof}
-
 \paragraph{Decompilation}
@@
   shape $\lambda x.t$ and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
+\end{proof}
+
+\begin{lemma}
+  The introduction of \linketa guarantees \fullref{prop:w-enforcing}
+  %\label{lemma:unif-wellb}
+\end{lemma}
+
+\begin{proof}[Proof sketch]
+  By \cref{def:testmaybeeta}.\todo{shall we move the proof here?}
 \end{proof}
@@
 
-\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
+\begin{theorem}[Fidelity in \maybeeta]
   % F x = z, F x y = z y
   Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
-  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
+  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow t \notin \notllambda$ and if the 
   memory map is bijective, then
@@
 
-% \begin{definition}[$\wellb_{\beta \eta \llambda}$]
-%   $
-%   \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
-%   $
-% \end{definition}
-
-\begin{lemma}[$\wellb$-enforcement]\label{lem:w-enforcement-maybellam}
-  $\forall s, \C{s}{t}{m}{l}$, then $\:\wellb(t)$.
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-  By \cref{lem:w-enforcement-maybebeta,lem:w-enforcement-maybeeta} and the 
-  rule above.
-\end{proof}
-
 \paragraph{Decompilation}
@@
   unified with $X_{s_1\ldots s_n~s}$.
+  \todo{I changed this part, it is OK?}
   If $m > 1$, then $l$ is replaced by the link
@@
   it makes \elpiIn{progress} immediately fails.
-%
+
   % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
@@
 
-\begin{theorem}[Fidelity in \Fo]
+\begin{theorem}[Fidelity with \linkbeta]
   The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}

### 2024-04-16 paper/main.tex
typos plus yve's remark

@@
 \begin{proposition}[compiler-useful]\label{prop:w-noncontra}
-  Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
+  Given two terms $s_1$ and $s_2$, if $\;\exists \rho, \rho s_1 \Eo \rho s_2$,
   then 
@@
 Note that the property holds for any substitution, $\rho$ could be given by an
-oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
-does not contradict \Eo. Also note that a compiler that
+oracle and/or not necessarily be a most general one: when applied to terms in \wellb,
+\Ue{} does not contradict \Eo. Also note that a compiler that
 translate all terms to unification variables would satisfy this

### 2024-04-16 paper/main.tex
lemma 5.13

@@
 \begin{lemma}
-  Given a \linketa \linketaM{\Gamma}{X}{T}, the unification done by \progressetaleft is between
+  Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaleft is between
   terms in $\wellb$  
@@
 \begin{proof}[Proof sketch]
-  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb),
-  By\todo{reformulate}
-  \progressetaleft, if 1) the $\sigma X$ is a name, a constant or an
-  application, then, $\lambda x.X \appsep x$ is unified with the \rhs. By
-  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux},
-  $\rhs = \lambda x. t$ and $\wellb(t)$.
-  Otherwise, 2) $\sigma X = \lambda x.t$.
-  % \todo{what does as a functor mean?}
-  In both cases, unification is
-  performed between terms in \wellb.
+  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
+  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
+  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$: both terms are \wellb.
+  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
+  are both \wellb. This concludes the proof.
 \end{proof}

### 2024-04-16 paper/main.tex
zap ML from the abstract

@@
   Logical Frameworks and Higher Order Logic Programming Languages provide
-  dedicated, high-level Meta Languages (ML) to facilitate this task in two
-  key ways: 1) variable binding and substitution are for free when ML binders
-  represent object logic ones; 2) proof construction, and even proof search, are
-  greatly simplified by leveraging the unification procedure provided by the ML.
-  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
+  dedicated, high-level meta languages to facilitate this task in two
+  % key 
+  ways: 1) variable binding and substitution are for free when meta language binders
+  represent object logic ones; 2) proof construction, and %even 
+  proof search, are
+  greatly simplified by leveraging the unification procedure provided by the meta language.
+  Notable examples of meta languages are Elf~\cite{elf}, Twelf~\cite{twelf},
   $\lambda$Prolog~\cite{miller_nadathur_2012} and
   Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-  which have been utilized to implement various formal systems such as
+  which have been %utilized
+  used to implement %various
+  many formal systems such as
   First Order Logic~\cite{felty88cade},
   Set Theory~\cite{10.1007/BF00881873},
-  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
+  Higher Order Logic~\cite{books/sp/NipkowPW02}, and %even
+  the Calculus of
   Constructions~\cite{felty93lics}.
@@
   develop a higher-order unification-based 
-  proof search procedure for it using the ML
+  proof search procedure for it using the meta language
   Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
   Elpi's equational theory includes
-  $\eta\beta$ equivalence and features a
+  $\eta\beta$-equivalence and features a
   higher-order unification procedure \Ue restricted to the pattern
@@
   taking advantage of
-  unification-related optimizations of the ML, such as clause indexing.
+  unification-related optimizations of the meta language, such as clause indexing.
 
@@
   % if \elpiIn{T} $\in$ S
-  If $\C{s}{t}{m}{l}$, such that $l \in \linkStore$ and $m \in \mapStore$, if
+  If $\C{s}{t}{m}{l}$ and $l \in \linkStore$ and $m \in \mapStore$
+  and
   $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
@@
 \end{proposition}
-\todo{explain no proof, llam link breaks}
+
+\noindent
+The statement above roughly states that the translation from
+\Fo{} to \Ho is not lossy, it can be reversed. The statement
+is made convoluted by the fact that we hid the procedure to
+decompile a term in favor of the one to decompile substitutions.
+We omit to sketch the proof of this property for brevity.
 

### 2024-04-16 paper/main.tex
help by yves

@@
 
-\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
-\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
+%\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
+%\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
+\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}\xspace}}{F}} % space non va
+\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}}\xspace}{H}}
 
@@
 \label{sec:intro}
-
-Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
-$\lambda$Prolog~\cite{miller_nadathur_2012}, and
-Isabelle~\cite{10.1007/978-3-540-71067-7_7}
-have been utilized to specify various
-logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
-The use of these meta languages facilitates this task in two
-key ways. The first and most well-know one is that variable binding and
+We aim to implement a
+form of proof search known as type-class resolution~\cite{wadler89,sozeau08}
+for the Coq's system~\cite{Coq-refman}.
+Type-class solvers are unification based proof search procedures
+reminiscent of Prolog, which back-chain lemmas taken
+from a database of ``type-class instances''. Given this
+analogy with Logic Programming we want to leverage the
+Elpi meta-programming language~\cite{tassi:hal-01637063},
+a dialect of $\lambda$Prolog, already used to extend
+Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
+
+The use of meta languages facilitates the implementation of
+proof systems in two key ways. The first and most well-know one is that variable binding and
 substitution come for free. %, exploited in all works mentioned above.
@@
 of unification, a cornerstone of proof construction and proof search.
-
 % exploited only
@@
 %it specification.\\
-The object logic we are interested in is Coq's~\cite{Coq-refman}
-Calculus of Inductive Constructions (CIC), and we aim to implement a
-form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
-Type-class solvers are unification based on proof search procedures
-reminiscent of Prolog, which back-chain lemmas taken
-from a database of ``type-class instances''. Given this
-analogy with Logic Programming we want to leverage the
-Elpi~\cite{tassi:hal-01637063} meta-programming language,
-a dialect of $\lambda$Prolog, already used to extend
-Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
 In this paper, we focus on one aspect of
@@
 We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
-from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
-library. The class  \coqIn{Decision}
+from the Stdpp
+library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}.
+The class  \coqIn{Decision}
 identifies predicates equipped with a decision procedure, while
@@
 Note that rule 3 features a second-order parameter \coqIn{P} that represents
-a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
+a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
 The solver has to infer a value for \coqIn{P} by unifying the conclusion
@@
 problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
-pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
+pattern-fragment and admits a unique
 solution $\customlabel{solution:intro}{\rho}$ that assigns
 the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
+In the rest of this paper shall refer to the pattern-fragment as
+\llambda~\cite{miller92jsc}.
 
 In order to implement such a search in Elpi, we shall describe the encoding
-of CIC terms and then the encoding of instances as rules.
+of Coq terms and then the encoding of instances as rules.
 Elpi comes equipped with
-an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
+an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
 terms, called \elpiIn{tm}, that includes (among others) the following
@@
 parameters; \elpiIn{:-} separates the rule's head from the premises, and
-\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
+<<\elpiIn{pi w\ p}>> introduces a fresh nominal constant \elpiIn{w}
 for the premise \elpiIn{p}.
@@
 \noindent
-In this paper we study a more sophisticated encoding of CIC terms and rules
+The problem above has no solution in a first-order language
+where \elpiIn{P} can only stand for a predicate symbol, while in a higher-order
+language \elpiIn{P} can stand for any 1-argument predicate.
+
+In this paper we study a more sophisticated encoding of Coq terms and rules
 that, on a first approximation, would reshape \ref{clause:r3} as follows:
@@
 problems in the pattern fragment. Although 
-the equational theory of CIC is much richer, for efficiency and predictability
+the equational theory of Coq is much richer, for efficiency and predictability
 reasons, automatic proof search procedures typically employ a unification
@@
 in \llambda. The similarity is striking, but one needs to exercise some caution in order to
-simulate a logic program in CIC using the unification of Elpi.
+simulate a higher-order logic program in Coq using the unification of Elpi.
 
@@
 fragment.\\
-Section~\ref{sec:problem-statement} formally states the problem and gives the
+Section~\ref{sec:problem-statement} states the problem formally and gives the
 intuition behind our solution; \cref{sec:simulation} sets up a basic
@@
 
-Even if we encountered the problem working on CIC, we devise
+Even if we encountered the problem working on Coq's Calculus of Inductive
+Constructions, we devise
 a minimal setting to ease its study. In this setting, we have
@@
 is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
-out that it additionally produces a variable mapping $m$ and a list of links $l$.
+out that it additionally produces a variable map $m$ and a list of links $l$.
 The variable map connects unification variables in \Ho to variables
@@
 so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
-However, in the case of CIC, where lambdas carry the type of the bound variable, one
+However, in the case of Coq, where lambdas carry the type of the bound variable, one
 must store this information somewhere. Note that this information is similar
-to the arity of variables; that in CIC, unification variables have a
+to the arity of variables; that in Coq, unification variables have a
 (function) type, and this type can be used to annotate the lambdas
@@
 This map allows to decompile the \Ho substitution to the object language.
-Given the higher order nature of CIC the same unification
+Given the higher order nature of Coq the same unification
 variable $X$ is used multiple times with different arities.
@@
 However, this encoding has two big limitations. First it is
-not always feasible to adopt it for CIC due to the fact that the type system
+not always feasible to adopt it for Coq due to the fact that the type system
 of the ML is too limited to accommodate for that one the object language,
-e.g. CIC can typecheck variadic functions~\cite{cpdt}. \\
-Second, the CIC encoding provided by Elpi is primarily utilized for meta
+e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
+Second, the encoding of Coq terms provided by Elpi is primarily utilized for meta
 programming, in order to extend the Coq system. Consequently, it must be able to
@@
 and heuristics to handle terms outside the pattern fragment, and
-it is not tightly coupled with CIC.
+it is not tightly coupled with Coq.
 

### 2024-04-16 paper/main.tex
minor

@@
 problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
-pattern-fragment and admits a unique
+pattern fragment and admits a unique
 solution $\customlabel{solution:intro}{\rho}$ that assigns
 the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
-In the rest of this paper shall refer to the pattern-fragment as
+In the rest of this paper shall refer to the pattern fragment as
 \llambda~\cite{miller92jsc}.
@@
 \noindent
+Note that the symbol \elpiIn{|} separates the head of a list from the tail,
+that \elpiIn{forall2} applies a binary predicate to two lists (of the same
+length). Finally \elpiIn{=>} is $\lambda$Prolog syntax to augment the
+current program with an extra rule, and both procedures use it to
+assume that the fresh nominal constants introduces by \elpiIn{pi} are
+equal to themselves.
+
 The main point in showing these equality tests is to remark how
@@
 weak head normal form of <<\elpiIn{app [F|L]}>>.
-Note that the symbol \elpiIn{|} separates the head of a list from the tail.
-\todo{explain forall2}
 
@@
 
-\begin{proposition}[Simulation]\label{prop:simulation}
+\begin{proposition}[Run equivalence]\label{prop:simulation}
 $\forall \foUnifPb, \forall \mathcal{N},$ if $~\foUnifPb \subseteq \llambda$

### 2024-04-16 paper/main.tex
Update proof th 5.18 and th 6.4

@@
 \begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
-\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$
+\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
+\label{th:fidelity-beta}
 \end{theorem}
@@
   % F x = z, F x y = z y
-  Given a list of unification problems \foUnifPb, if
-  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
+  Given a list of unification problems \foUnifPb, such that
+  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
   memory map is bijective, then
@@
 \begin{proof}[Proof sketch]
-  \progressetaleft and \progressetadedup activate a \linketa when, in the
-  original unification problem, a \maybeeta term is unified with respectively a
-  well-behaved term or another \maybeeta term. In both cases, the links trigger
-  a unification which succeeds iff the same unification in \Fo{} succeeds,
-  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule, citare regolae in =o}
-  guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
-  this progression refines a variable to a rigid term and plays no role in 
-  \cref{prop:fidelity}.
-  % If 1) a \linketa is activated by \progressetaleft, then unification is done
-  % between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
-  % activation performs a unification which succeeds iff the original problem in
-  % \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
-  % unification is done between two \maybeeta terms, and again this unification
-  % succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
-  % \progressetaright, the unification, done between a variable and a term,
-  % always succeeds, this is what we expect to guarantee fidelity, since we
-  % are essentially removing a hole added by the compilation in the place of a
-  % \maybeeta subterm. In all the cases fidelity is respected.
+  We want to prove that for any $\fstep_i$, $\fstep_i$ succeeds iff $\hstep_i$
+  succeeds. An $\hstep_i$ is made by a unification $\mathcal{U}$ and a link
+  progression $\mathcal{P}$.
+  Two cases should be analyzed: 1) $\fstep_i$ succeeds, this means that $\mathcal{U}$
+  succeeds since any \maybeeta term is replaced with a fresh variable, and by
+  \cref{th:fidelity-beta}, the other subterms unify. Moreover, $\mathcal{P}$
+  succeeds, since the $\eta$-progression
+  rules simply imitate the rules~\ref{clause:eta1} and \ref{clause:eta2},
+  performing unification between $\eta$-contracted terms. 
+  If 2) $\fstep_i$ fails, either the failure is 
+  due to the unification of terms that are not in \maybeeta. In this case
+  $\mathcal{U}$ fails thanks to \cref{th:fidelity-beta}. 
+  On the other hand, the failure is
+  due to a unification over \maybeeta terms, making $\mathcal{P}$ to fail.
 \end{proof}
@@
   By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
-  variables with different arities are compiled to two
-  variables $A^n$ and $C^m$ related by a chain of \linketa.
-  We show that whenever one of the two is instantiated with a rigid term then
-  the chain becomes shorter and hence, eventually, the
-  other variable is instantiated too.
-  Without loss of generality we assume $m < n$ and $A$ is instantiated.
-  The corresponding \linketa is triggered by \progressetaleft instantiating
-  the next variable in the chain of links.
+  variable used at different arities are compiled to two
+  variables $A^m$ and $C^n$ related by a chain of \linketa.
+  Whenever the two variables are assigned to the non-unifiable terms,
+  the chain of \linketa is entirely waken up refining little by little
+  the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
+  On the other
+  hand, if the two variables still remain unifiable after their assignement, 
+  then the activation of 
+  the corresponding \linketa just refines the chain, by reducing eventually
+  its length.
 \end{proof}
-\todo{Not sure I fullfy fixed the proof, but the old oen was not ok}
+% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}
 

### 2024-04-16 paper/main.tex
clean proof lemma 5.14

@@
 \begin{lemma}
-  Given a \linketa $l$, the unification done by \progressetaright is between
+  Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaright is between
   terms in \wellb.
@@
 \begin{proof}[Proof sketch]
-  \lhs is variable, and, by \cref{def:progressetaright}, \rhs is either no more
-  a \maybeeta, i.e. \rhs is not a $\eta$-expansion and, so, $\wellb(\rhs)$,
-  otherwise, \rhs can reduce to a term which cannot be a $\eta$-expansion, and,
-  so, $\wellb(\rhs)$. In both cases, the unification between \rhs and \lhs
-  is done between terms that are in \wellb.
+  $X$ is variable, and, by \cref{def:progressetaright}, $\lambda x.t$ is either no more
+  a \maybeeta, i.e. it is not a $\eta$-expansion and, so, $\wellb(\lambda x.t)$, 
+  the unification is performed between $X$ and $\lambda x.t$.
+  Otherwise, $\lambda x.t$ has the shape $\lambda x.t' \appsep x$ with $x$ not free in $t'$ and
+   $\wellb(t')$. In this second case the unification is done between $X$ and $t'$.
+  In both cases, unifications are between \wellb terms.
 \end{proof}

### 2024-04-16 paper/main.tex
clean proof  lemma 5.15

@@
 \begin{lemma}
-  Given a \linketa $l$, the unification done by \progressetadedup is between
+  The unification done by \progressetadedup is between
   terms in \wellb.
@@
 \begin{proof}
-  We unify the \rhs{} of two \linketa. Both have the
-  shape $\lambda x.t$ and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
+  Given two \linketa{}s \linketaM{\Gamma_1}{X}{\lambda x.t} and
+  \linketaM{\Gamma_2}{Y}{\lambda x.t'} 
+  the unification is performed between $t$ and $t'$. Both terms are
+  \wellb and therefore, unification is done between \wellb terms.
 \end{proof}

### 2024-04-16 paper/main.tex
remove unused macros
wip

@@
 
-\def\elpi{\proglang{elpi}}
-\def\coqelpi{\proglang{coq-elpi}}
-\def\lambdaprolog{\proglang{$\lambda$-prolog}}
-\def\coq{\proglang{coq}}
-
-\newcommand{\library}[1]{\textit{#1}\xspace}
-\def\stdpp{\library{stdpp}}
-\def\iris{\library{iris}}
-
-\newcommand*{\acronym}[1]{\texttt{#1}\xspace}
-
 \newtheorem{invariant}{Invariant}
@@
 \crefname{equation}{property}{properties}
-
-\def\ol{\acronym{ol}} % object language
-\def\ml{\acronym{ml}} % meta language
-\def\lf{\acronym{lf}} % logical framework
-\def\ho{\acronym{ho}} % higher order
-\def\Forall{$\forall$}
 
@@
 \newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_{\llambda} #3}}
-\newcommand{\substCell}[3]{\ensuremath{#1 \vdash #2 = #3}}
 \newcommand{\mapping}[3]{\ensuremath{#1 \mapsto #2^#3}}
@@
 
-The use of meta languages facilitates the implementation of
-proof systems in two key ways. The first and most well-know one is that variable binding and
+The use of a meta language facilitates the implementation of a
+proof system in two key ways. The first and most well-know one is that variable binding and
 substitution come for free. %, exploited in all works mentioned above.
@@
 %it specification.\\
-In this paper, we focus on one aspect of
-this work, precisely \emph{how to reuse the higher-order unification procedure
-of the meta language in order to simulate a higher-order logic program
-for the object language}.
+In this paper we focus on this last aspect of the implementation of
+a Prolog-like proof system, precisely \emph{how to reuse the higher-order
+unification procedure of the meta language in order to simulate a
+higher-order logic program for the object language}.
 
@@
 
-\newcommand{\unifcorrectX}[4]{
+\newcommand{\unifcorrectX}[4]{\ensuremath{
   % \forall \rho #3_1 #3_2, %
@@
           \rho #3_1 #1 \rho #3_2  %
-}
+}}
 \newcommand{\unifcorrect}[3]{\unifcorrectX{#1}{#2}{#3}{ \{#3_1, #3_2\} \subseteq \llambda}}
 
-\newcommand{\unifcompleteX}[4]{
+\newcommand{\unifcompleteX}[4]{\ensuremath{
   % \forall #3_1 #3_2, %
@@
           \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
-}
+}}
 \newcommand{\unifcomplete}[3]{\unifcompleteX{#1}{#2}{#3}{\{#3_1, #3_2\} \subseteq \llambda}}
@@
 
+The specification of a ``good'' unification procedure restricted
+to a domain $\mathcal{D}$ is the following:
+
+\begin{proposition}[Good unification]\label{prop:good-unif}
+A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
+$\mathcal{D}$ satisfies the following properties:
+  \begin{gather}
+    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
+    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
+  \end{gather}
+\end{proposition}
+
+\noindent
 The meta language of choice is expected to provide 
-an implementation of \Ue that satisfies
-the following properties:
-%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
-\begin{gather}
-  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
-  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
-\end{gather}\todo{give a name and reuse it for \Uo in next sections}
-
-\noindent
+an implementation of \Ue that is a good unification for \Ee
+in \llambda.
+
 Even if we provide an implementation of the object-language unification
@@
 previous sections).
-\todo{explain fold6}
 
@@
 iterated \elpiIn{pi} abstraction, as in \elpiIn{pi x\ pi y\..}.
+The auxiliary predicate \elpiIn{fold6} folds a predicate
+with three accumulators (the memory map, the links and the substitution)
+over a lists to obtain a new list and the final values of the
+accumulators.
 
@@
 
-\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow} The following properties hold for \Uo
-  as implemented above: \todo{reuse statement from section 2.4}
-\begin{gather}
-  \unifcorrectX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}\\
-  \unifcompleteX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}
-\end{gather}
+\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
+  The implementation of \Uo above is a good unification (as per \cref{prop:good-unif})
+  for \Eo in the domain \wellb.
 \end{lemma}
@@
 
-\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb} The following properties hold for \Uo
-  where \todo{reuse 2.4}
-  \begin{gather}
-    \unifcorrectX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}\\
-    \unifcompleteX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}
-  \end{gather}
+\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb}
+  Given the updated compilation scheme, the \Uo of~\cref{sec:founif}
+  is a good unification (as per \cref{prop:good-unif})
+  for \Eo in the domain $\wellb_\beta$.
 \end{lemma}

### 2024-04-16 paper/main.tex
still rough 5.18

@@
 
-In order to specify unification, we need to
-define the equational theory and
-substitution (unification-variable assignment).
+% In order to specify unification, we need to
+% define the equational theory and
+% substitution (unification-variable assignment).
 
@@
 In \cref{sec:simulation,sec:eta,sec:beta}
-we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
+we describe how to %the compiler
+recognize terms in \maybebeta, \maybeeta and
 \notllambda and how \progress takes care of them preserving \wellb
@@
 builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
-using the routine
+using %the routine
 \ref{clause:malloc}.
@@
 The unification problem $\hoUnifPb_1$ fails while trying to unify
-$A \appsep x$ and $x$, which is equivalent to \elpiIn{app [uva A [], x]}
+$A \appsep x$ and $x$, which is equivalent to <<\elpiIn{app [uva A [], x]}>>
 versus \elpiIn{x}.
 In order to exploit the higher-order unification algorithm of the meta language,
-we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$,
-which is \elpiIn{uva A [x]}.
+we %need to 
+compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.
+%,which is \elpiIn{uva A [x]}.
 
@@
 \begin{proof}[Proof sketch]
-  We want to prove that for any $\fstep_i$, $\fstep_i$ succeeds iff $\hstep_i$
-  succeeds. An $\hstep_i$ is made by a unification $\mathcal{U}$ and a link
-  progression $\mathcal{P}$.
-  Two cases should be analyzed: 1) $\fstep_i$ succeeds, this means that $\mathcal{U}$
-  succeeds since any \maybeeta term is replaced with a fresh variable, and by
-  \cref{th:fidelity-beta}, the other subterms unify. Moreover, $\mathcal{P}$
-  succeeds, since the $\eta$-progression
-  rules simply imitate the rules~\ref{clause:eta1} and \ref{clause:eta2},
-  performing unification between $\eta$-contracted terms. 
-  If 2) $\fstep_i$ fails, either the failure is 
-  due to the unification of terms that are not in \maybeeta. In this case
-  $\mathcal{U}$ fails thanks to \cref{th:fidelity-beta}. 
-  On the other hand, the failure is
-  due to a unification over \maybeeta terms, making $\mathcal{P}$ to fail.
+  We want to prove that for any $i$, $\fstep_i$ succeeds iff $\hstep_i$
+  succeeds. An $\hstep_i$ is made by a unification
+  $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
+  progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
+  The interesting case is when the infolved terms are in \maybeeta.
+  wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
+  we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
+  In all cases $X \Ue t2$ succeeds since X is a variable and
+  hence assigns t2 to X.
+  Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
+  additional care of eta expanding t2 if needed (5.9.1).
+
+  Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
+  $s1$ with $s2$. The interesting case is when fostep succeds by
+  finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
+  rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
+  and that is in exactly the cases in which T (from page 3) is
+  not a lam (see 5.9.1) compensating for the missing rules in \Ue.
+
+  2) If $\fstep_i$ fails to unify $s1 \Uo s2$
+  it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
+  and the one flam-flam we find different terms.
+  (5.9.1) only adds lambdas that are forcibly consumed by
+  lam-lam rule.
   \end{proof}
+
+Note that \progressetadedup always unifies terms starting with exaclty one
+lambda, hence lam-lma, so we can't fail due to that lambda or succed thanks
+to that lambda. Hence it does not impact fidelity.
+
+\begin{corollary}[Properties of \Uo in $\wellb_{\beta\eta}$]\label{prop:unifetawb}
+  Given the updated compilation scheme, the \Uo of~\cref{sec:founif}
+  is a good unification (as per \cref{prop:good-unif})
+  for \Eo in the domain $\wellb_{\beta\eta}$.
+\end{corollary}
+
+Note \progressetaright is only needed in the next section where chains of links
+are added. Hence we give an example for rule \progressetaright in the next section.
+\todo{fix english}
 
@@
 $\rho = \{~ X \mapsto Y ~\}$ as expected.
-
-We give an example for rule
-\progressetaright in the next section.
 

### 2024-04-16 paper/main.tex
w

@@
 \end{definition}
-
 % Example of this `X a = Y a'
@@
 \begin{definition}[\progBetaFail]
-  \progress fails whenever:
-  there exists another link $l' \in \linkStore$ with same \lhs as $l$;
-  or the \lhs  of $l$ become rigid.
+  \progress fails whenever
+  there exists another link $l' \in \linkStore$ with the same \lhs % as $l$;
+  or the \lhs % of $l$
+  become rigid.
   \label{def:progBetaFail}
@@
   \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
-  terminates. Finally \progBetaFail also guarantees termination since 
-  it makes \elpiIn{progress} immediately fails.
+  terminates. Finally \progBetaFail guarantees termination since 
+  %it makes 
+  %\elpiIn{progress}
+  it immediately fails.
 %
@@
 Consider the \linkbeta below:
-
+%
 % make test ONLY=7004 TEX=tex

### 2024-04-16 paper/main.tex
minor

@@
 shown in \cref{code:common-terms}.
+Unification variables
+in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
+the arguments of a higher-order variable are given via the \elpiIn{fapp}
+constructor. For example the term <<\coqIn{P x}>> is represented as
+<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
+of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
+In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
+since unification variables are higher-order and come equipped with an
+explicit scope.
+
 % We keep these languages
@@
 }
-Unification variables
-in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
-the arguments of a higher-order variable are given via the \elpiIn{fapp}
-constructor. For example the term <<\coqIn{P x}>> is represented as
-<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
-of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
-In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
-since unification variables are higher-order and come equipped with an
-explicit scope.
 %\todo{use faddr for Fo}
@@
 \noindent
-The second snippet illustrate the deduplication of \linketa. The syntax
+The second snippet illustrates the deduplication of \linketa. The syntax
 <<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 

### 2024-04-16 paper/main.tex
add todo
rework 5.18

@@
 explicit scope.
-
+%
 % We keep these languages
@@
 
-\begin{proof}[Proof sketch]
-  We want to prove that for any $i$, $\fstep_i$ succeeds iff $\hstep_i$
-  succeeds. An $\hstep_i$ is made by a unification
-  $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
-  progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
-  The interesting case is when the infolved terms are in \maybeeta.
-  wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
-  we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
-  In all cases $X \Ue t2$ succeeds since X is a variable and
-  hence assigns t2 to X.
-  Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
-  additional care of eta expanding t2 if needed (5.9.1).
-
-  Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
-  $s1$ with $s2$. The interesting case is when fostep succeds by
-  finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
-  rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
-  and that is in exactly the cases in which T (from page 3) is
-  not a lam (see 5.9.1) compensating for the missing rules in \Ue.
-
-  2) If $\fstep_i$ fails to unify $s1 \Uo s2$
-  it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
-  and the one flam-flam we find different terms.
-  (5.9.1) only adds lambdas that are forcibly consumed by
-  lam-lam rule.
+\begin{proof}[Proof sketch]\todo{to reformulate}
+  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
+  % succeeds. A $\hstep_i$ is made by a unification
+  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
+  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
+  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
+  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
+  progression $p$.
+  Without loss of generality,\todo{explain why} we consider a \Fo unification problem of the form $s_1 \Uo s_2$
+  where $s_1 \in \maybeeta$, which is compiled into the \linketa $\linketaM{}{X}{\lambda x.t_1}$. 
+  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
+  only have to treat the progression $p$.
+  Two cases should be analyzed:
+  \begin{itemize}
+    \item $s_2 \notin \maybeeta$, in this case, the unification problem becomes
+          $X \Ue t_2$, which assigns $X$ to $t_2$.
+          After the unification, \progressetaleft triggers the \linketa,
+          and $\lambda x.t_1$ is unified with $t_2$ (or its $\eta$-expansion when needed).
+          This unification succeeds iff $s_1$ and $s_2$ unify. \progressetaleft
+          aims essentially to transpose the two rules \ref{clause:eta1} and \ref{clause:eta2}
+          of the object language in the meta one.
+    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
+          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
+          $X$ and $Y$, \progressetadedup is fired and $t_1$ and $t_2$ are unified.
+          This unification succeeds iff $s_1 \Uo s_2$ succeeds.
+  \end{itemize}
+  In this setting, \progressetaright does not play a critical role since \mapStore
+  is a bijection; it's role is to perform variable refinement, i.e. it never fails.
+  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
+  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
+  % In all cases $X \Ue t2$ succeeds since X is a variable and
+  % hence assigns t2 to X.
+  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
+  % additional care of eta expanding t2 if needed (5.9.1).
+% 
+  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
+  % $s1$ with $s2$. The interesting case is when fostep succeds by
+  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
+  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
+  % and that is in exactly the cases in which T (from page 3) is
+  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
+  % 
+  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
+  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
+  % and the one flam-flam we find different terms.
+  % (5.9.1) only adds lambdas that are forcibly consumed by
+  % lam-lam rule.
 \end{proof}
 
-Note that \progressetadedup always unifies terms starting with exaclty one
-lambda, hence lam-lma, so we can't fail due to that lambda or succed thanks
-to that lambda. Hence it does not impact fidelity.
+% Note that \progressetadedup always unifies terms starting with exaclty one
+% lambda, hence lam-lma, so we can't fail due to that lambda or succed thanks
+% to that lambda. Hence it does not impact fidelity.
 

### 2024-04-17 paper/main.tex
update th 6.4

@@
   variables $A^m$ and $C^n$ related by a chain of \linketa.
-  Whenever the two variables are assigned to the non-unifiable terms,
-  the chain of \linketa is entirely waken up refining little by little
-  the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
-  On the other
-  hand, if the two variables still remain unifiable after their assignement, 
-  then the activation of 
-  the corresponding \linketa just refines the chain, by reducing eventually
-  its length.
+  % Attenzione a questo caso:
+  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
+  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
+  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
+%
+  Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
+  $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
+  term, then the chain of links between $A^m$ and $C^n$ is completely triggered
+  so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
+  Note that in this case fidelity is guaranteed, since term unified to $C^n$
+  sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
+  links is shrinked by $q$. Note that if this length is bigger then $n-m$ again
+  the full chain collapse and $A^m$ and $C^n$ are agian tiny coupled. Otherwise,
+  if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
+  with at least $q$ lambdas. In this second case fidelity is also guaranteed,
+  due to the constraint of the number of lambda by which $C^n$ should start.
+%
+  % not starting with $\lambda$, then the entrire chain of link is
+  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
+  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
+  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
+  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
+  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
+  % that if $t$ is rigid, by the previous sentence, the remi
+%
+  % If $A^m$ is instantiated, 
+%
+  % Whenever the two variables are assigned to the non-unifiable terms,
+  % the chain of \linketa is entirely waken up refining little by little
+  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
+  % On the other
+  % hand, if the two variables still remain unifiable after their assignement, 
+  % then the activation of 
+  % the corresponding \linketa just refines the chain, by reducing eventually
+  % its length.
 \end{proof}

### 2024-04-17 paper/main.tex
cite beluga

@@
   greatly simplified by leveraging the unification procedure provided by the meta language.
-  Notable examples of meta languages are Elf~\cite{elf}, Twelf~\cite{twelf},
-  $\lambda$Prolog~\cite{miller_nadathur_2012} and
+  Notable examples of meta languages are Twelf~\cite{twelf},
+  $\lambda$Prolog~\cite{miller_nadathur_2012}, Beluga~\cite{Beluga} and
   Isabelle~\cite{10.1007/978-3-540-71067-7_7}
   which have been %utilized
-  used to implement %various
-  many formal systems such as
+  used to implement or specify %various
+  many
+  formal systems such as
   First Order Logic~\cite{felty88cade},
@@
   The object logic we are interested in is
-  Coq's Calculus of Inductive Constructions (CIC)\cite{Coq-refman}. We aim to
+  Coq's type theory~\cite{Coq-refman}. We aim to
   develop a higher-order unification-based 
-  proof search procedure for it using the meta language
+  proof search procedure %for it 
+  using the meta language
   Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
@@
   $\eta\beta$-equivalence and features a
-  higher-order unification procedure \Ue restricted to the pattern
-  fragment~\cite{miller92jsc}.
-  Elpi offers an encoding of CIC suitable
-  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
-  but restricts \Ue to roughly first-order unification problems only.
+  higher-order unification procedure \Ue 
+  %restricted to the
+  for the
+  pattern fragment~\cite{miller92jsc}.
+  Elpi offers an encoding of Coq terms suitable
+  for meta programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
+  but that restricts \Ue to first-order unification problems only.
   We refer to this basic encoding as \Fo.
@@
   Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
-    that honours $\eta\beta$-equivalence (for CIC functions), addresses
+    that honours $\eta\beta$-equivalence (for Coq functions), addresses
   problems within the pattern fragment, and allows

### 2024-04-17 paper/main.tex
update progress eta

@@
 \begin{definition}[\progressetaleft]\label{def:progressetaleft}
-A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
+A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
 $X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
@@
   \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
-    we unify the $\eta$-expansion of $X$ with $T$, that is we run
-    $\lambda x.X \appsep x\Ue{} T$
-    \item if $X = \lambda x.t$ we run $X \Ue{} T$.
+    we unify the $\eta$-expansion of $X$ with $t$, that is we run
+    $\lambda x.X \appsep x\Ue{} t$
+    \item if $X = \lambda x.t$ we run $X \Ue{} t$.
 \end{enumerate}
@@
 
-\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
-\begin{definition}[\progressetaright]\label{def:progressetaright} A link
-\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
-$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
-$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
-first case, $X$ is unified with $T$, and in the second one, $X$ is unified with
-$T'$ (under the context $\Gamma$).
+\newcommand{\progressetashrink}{\emph{$\eta$-progress-shrink}\xspace}
+\begin{definition}[\progressetashrink]\label{def:progressetashrink}
+Given a link \linketaM{\Gamma}{X}{t}, and let $S$ the scope of $X$
+we pose $T = S \cap \Gamma$. If $T$ is not empty,
+then the original link is replaced by a new link
+\linketaM{S}{X}{t'}, where $t'$ is a term where all the variables in $\Gamma / X$ 
+have been pruned.
 \end{definition}
 
-There is a third case in which a link is removed from \linkStore, namely
-when the \lhs is assigned to a variable that is the \lhs of another
-\linketa.
+% There is a third case in which a link is removed from \linkStore, namely
+% when the \lhs is assigned to a variable that is the \lhs of another
+% \linketa.
 
@@
 
-\begin{lemma}
-  Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaright is between
-  terms in \wellb.
-  \label{lemma:unif-eta-2}
-\end{lemma}
-
-\begin{proof}[Proof sketch]
-  $X$ is variable, and, by \cref{def:progressetaright}, $\lambda x.t$ is either no more
-  a \maybeeta, i.e. it is not a $\eta$-expansion and, so, $\wellb(\lambda x.t)$, 
-  the unification is performed between $X$ and $\lambda x.t$.
-  Otherwise, $\lambda x.t$ has the shape $\lambda x.t' \appsep x$ with $x$ not free in $t'$ and
-   $\wellb(t')$. In this second case the unification is done between $X$ and $t'$.
-  In both cases, unifications are between \wellb terms.
-\end{proof}
+% \begin{lemma}
+%   Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaright is between
+%   terms in \wellb.
+%   \label{lemma:unif-eta-2}
+% \end{lemma}
+
+% \begin{proof}[Proof sketch]
+%   $X$ is variable, and, by \cref{def:progressetaright}, $\lambda x.t$ is either no more
+%   a \maybeeta, i.e. it is not a $\eta$-expansion and, so, $\wellb(\lambda x.t)$, 
+%   the unification is performed between $X$ and $\lambda x.t$.
+%   Otherwise, $\lambda x.t$ has the shape $\lambda x.t' \appsep x$ with $x$ not free in $t'$ and
+%    $\wellb(t')$. In this second case the unification is done between $X$ and $t'$.
+%   In both cases, unifications are between \wellb terms.
+% \end{proof}
 
@@
 \begin{proof}[Proof sketch]
-  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
+  By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
   unification performed by the activation of a \linketa is done between
@@
 \begin{proof}[Proof sketch]
-  Rules \cref{def:progressetaleft,def:progressetaright} and
+  Rules \cref{def:progressetaleft} and
   \cref{def:progressetadedup} remove one link from \linkStore, hence they
@@
   \end{itemize}
-  In this setting, \progressetaright does not play a critical role since \mapStore
-  is a bijection; it's role is to perform variable refinement, i.e. it never fails.
   % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
@@
 
-Note \progressetaright is only needed in the next section where chains of links
-are added. Hence we give an example for rule \progressetaright in the next section.
-\todo{fix english}
-
 \paragraph{Example of \progressetaleft}
@@
   add to \linkStore the result of $\alignarity{}~m_1~m_2$.
+\end{definition}
+
+In order to grant fidelity, we add a new link progression for \linketa{}s.
+
+\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
+\begin{definition}[\progressetaright]\label{def:progressetaright} A link
+\linketaM{\Gamma}{X}{t} is removed from \linkStore when either 1)
+$\testmaybeeta~t$ does not hold (anymore) or 2) by $\eta$-contracting $t$ to
+$t'$ where $t'$ does not start with the \elpiIn{lam} constructor. In the
+first case, $X$ is unified with $t$, and in the second one, $X$ is unified with
+$t'$ (under the context $\Gamma$).
 \end{definition}

### 2024-04-17 paper/main.tex
remove snippet

@@
 If instead we want \Uo{} to follow the second strategy and pick an
-arbitrary solution we can modify \progress by applying the desired heuristic.
+arbitrary solution we can modify \progress by applying the desired heuristic
+instead of failing.
 For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
 argument of the two terms is the same and unification can succeed by assigning
-$X a$ to $Y$. This heuristic can be implemented by adding the following rule:
-
-\input{code/progress_heuristic}
+$X a$ to $Y$. This heuristic is used by~\cite{sozeau08}.
+
+%\input{code/progress_heuristic}
 %\noindent

### 2024-04-17 paper/main.tex
many todo to simplify

@@
 \noindent
-Note that we could remove the first rule, whose sole purpose is to make
-links more readable by pruning unused context entries.
+\todo{remove pruning optimization from code above}
 
@@
 % that applies the substitution \elpiIn{S1} to all terms in \linkStore.
+
+\subsubsection{Scope check}\label{sec:sc}
+The predicate xxxx \todo{add} takes
+a link \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} and replaces it by
+\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
+The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ after
+executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$, that is $t'$ cannot contain
+variables in $\Gamma / \{x_1\ldots x_n\}$.
+This is key for fidelity since suspended unif problem
+must stay solvable.
 
@@
 
-\newcommand{\progressetashrink}{\emph{$\eta$-progress-shrink}\xspace}
-\begin{definition}[\progressetashrink]\label{def:progressetashrink}
-Given a link \linketaM{\Gamma}{X}{t}, and let $S$ the scope of $X$
-we pose $T = S \cap \Gamma$. If $T$ is not empty,
-then the original link is replaced by a new link
-\linketaM{S}{X}{t'}, where $t'$ is a term where all the variables in $\Gamma / X$ 
-have been pruned.
-\end{definition}
-
 % There is a third case in which a link is removed from \linkStore, namely
@@
   progression $p$.
-  Without loss of generality,\todo{explain why} we consider a \Fo unification problem of the form $s_1 \Uo s_2$
-  where $s_1 \in \maybeeta$, which is compiled into the \linketa $\linketaM{}{X}{\lambda x.t_1}$. 
+  Without loss of generality,\todo{explain why} we consider a
+  \Fo unification problem of the form $s_1 \Uo s_2$
+  where $s_1 \in \maybeeta$. $s_1$ is compiled to $X$ with the accessory link
+  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
   The unification $u$ always succeeds, since $X$ is a fresh variable, we 
-  only have to treat the progression $p$.
+  only have to consider the link progression $p$.
   Two cases should be analyzed:
   \begin{itemize}
-    \item $s_2 \notin \maybeeta$, in this case, the unification problem becomes
-          $X \Ue t_2$, which assigns $X$ to $t_2$.
-          After the unification, \progressetaleft triggers the \linketa,
-          and $\lambda x.t_1$ is unified with $t_2$ (or its $\eta$-expansion when needed).
-          This unification succeeds iff $s_1$ and $s_2$ unify. \progressetaleft
-          aims essentially to transpose the two rules \ref{clause:eta1} and \ref{clause:eta2}
-          of the object language in the meta one.
+    \item $s_2 \notin \maybeeta$,
+          in this case $s_2$ is compiler to $t_2$ and $u$ assigns $t_2$ to $X$
+          and triggers \progressetaleft if $t_2$ is rigid.
+          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
+          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
+          $\lambda x.t_2\appsep x$, otherwise they unify thanks to lam-lam\todo{dare nome}
+          and $p$ unifies $\lambda x.t_1$ with $t_2$. Alternatively, if $s_1$ and $s_2$ fails
+          to unify, it cannot be because of the head $\lambda$ constructors,
+          since wither lam-lam or \ref{clause:eta1} move under it and find
+          different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
+          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
+          $\eta$-expands $t_2$.
+
     \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
           with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
-          $X$ and $Y$, \progressetadedup is fired and $t_1$ and $t_2$ are unified.
-          This unification succeeds iff $s_1 \Uo s_2$ succeeds.
+          $X$ and $Y$, \progressetadedup is fired and, once the $\lambda$ are crossed,
+          $t_1$ and $t_2$ are unified. 
+          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
   \end{itemize}
@@
 
+\begin{comment}
 \paragraph{Example of \progressetadedup}
@@
 $\rho = \{~ X \mapsto Y ~\}$ as expected.
+\end{comment}
 
@@
     {\eta,x,B_{x},\lambda y.C_{y x}}}}
-
+\todo{simplify example since no prog right}
 In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
@@
 \end{definition}
+\todo{comment this rule}
 
@@
 
+\begin{comment}
 \paragraph{Example of \progBetaLL}
@@
 a ~\}$.
-
+\end{comment} 
 \subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
@@
 link-eta L R :- not (var L), !, eta-progress-lhs L R.
-link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
 \end{elpicode}
 \todo{improve code}
+%link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
 

### 2024-04-17 paper/main.tex
elf

@@
   greatly simplified by leveraging the unification procedure provided by the meta language.
-  Notable examples of meta languages are Twelf~\cite{twelf},
+  Notable examples of meta languages are Elf~\cite{elf}, Twelf~\cite{twelf},
   $\lambda$Prolog~\cite{miller_nadathur_2012}, Beluga~\cite{Beluga} and

### 2024-04-17 paper/main.tex
abella

--- 2024-04-17 paper/main.tex
@@
   greatly simplified by leveraging the unification procedure provided by the meta language.
-  Notable examples of meta languages are Twelf~\cite{twelf},
-  $\lambda$Prolog~\cite{miller_nadathur_2012}, Beluga~\cite{Beluga} and
+  Notable examples of meta languages are Elf~\cite{elf}, Twelf~\cite{twelf},
+  $\lambda$Prolog~\cite{miller_nadathur_2012}, Beluga~\cite{Beluga}, Abella~\cite{gacek2008abella} and
   Isabelle~\cite{10.1007/978-3-540-71067-7_7}
@@
 
-We considered mechanizing our results with Abella, especially in the early phases.
+We considered mechanizing our results with Abella~\cite{gacek2008abella}, especially in the early phases.
 Unfortunately we could not. One reason

### 2024-04-17 paper/main.tex
update scope-check

@@
 \subsubsection{Scope check}\label{sec:sc}
-The predicate xxxx \todo{add} takes
-a link \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} and replaces it by
+The predicate \elpiIn{scope-check} walks through all links. 
+Each link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} is replaced by
 \linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
-The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ after
-executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$, that is $t'$ cannot contain
-variables in $\Gamma / \{x_1\ldots x_n\}$.
-This is key for fidelity since suspended unif problem
-must stay solvable.
+The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
+executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$. This unification
+ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
+and therefore grants fidelity.
+In fact, a link represents a suspended unification problem between a unification variable $X$
+and a ``problematic'' term $t$. This unification is possible iff 
+the free variables in $t$ are also in the scope of $X$.
 
@@
 
+\begin{comment}
 \newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
@@
 \todo{comment this rule}
+\end{comment}
 
@@
 and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
-After deduplication the compiler output is as follows:
+
+After deduplication the compiler output is as follows:\todo{add simple example}
+\begin{comment}
 % make test ONLY=7001 TEX=tex
@@
 $\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto f\appsep (\lambda y.y)\}$.
+\end{comment}
 
@@
   is added to \linkStore.   
-
+  %
   % A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
@@
 %Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}. 
-\vspace{-1em}
+% \vspace{-1em}
 \section{Actual implementation in Elpi}\label{sec:implementation}

### 2024-04-17 paper/main.tex
solve a todo

@@
           progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
-          $\lambda x.t_2\appsep x$, otherwise they unify thanks to lam-lam\todo{dare nome}
+          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}
           and $p$ unifies $\lambda x.t_1$ with $t_2$. Alternatively, if $s_1$ and $s_2$ fails

### 2024-04-18 paper/main.tex
update tex

@@
   has the shape $\lambda x.t$
-  and $t$ is not a lambda. 
+  and $t$ is \wellb. 
   %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
@@
 \elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
-Invariant~\fullref{inv:linklhs} holds for this link, moreover:
+Invariant~\fullref{inv:linklhs} holds for this link. Moreover:
 
@@
 \begin{lemma}
-  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb(t)$.
+  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb(t)$, enforcing \cref{inv:link-eta-right}.
   \label{lemma:unif-eta-aux}
@@
   By construction, every ``problematic'' term in \Fo{} is replaced with a variable
-  in the corresponding \Ho term. Therefore, $t$ is \wellb.
+  in the corresponding \Ho term. Therefore, $\wellb(t)$.
 \end{proof}
@@
   Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
-  are both \wellb. This concludes the proof.
+  are both \wellb.
 \end{proof}
@@
 This map allows to decompile the \Ho substitution to the object language.
-Given the higher order nature of Coq the same unification
-variable $X$ is used multiple times with different arities.
+Given the higher-order nature of Coq the same unification
+variable $X$ can be used multiple times with different arities.
+\begin{comment}
 %
@@
     {\eta,x,B_{x},\lambda y.C_{y x}}}}
-\todo{simplify example since no prog right}
-In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
-$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
+\end{comment}
+% make test ONLY=7007 TEX=tex
+\printAlll
+  {{{\lambda x.(X\appsep x),a},
+    {X,\lambda x.a}}}
+  {{{A,a},
+    {C,\lambda x.a}}}
+  {{{X,C,0},
+    {X,B,1}}}
+  {{{\eta,,A,\lambda x.B_{x}}}}
+% \todo{simplify example since no prog right}
+In the unification problems \foUnifPb above, we see that $X$ is used with arity $1$ in
+$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
 In order to preserve \fullref{inv:uvaarity} the compiler
-did generate two entries in \mapStore for $X$, and in turn this
-breaks \fullref{prop:fidelity} since two incompatible assignements to
-the variables $X$ maps to, namely $E$ and $C$, would not be detected.
-We remedy by post-processing the compiler's output by adding
-additional \linketa to ensure the folllowgin property:
+did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
+However, there is no connection between these two variables
+and any incompatible assignements to them would not be detected, breaking,
+in turn \cref{prop:fidelity}. This incompatibility will only be detected during the
+decompilation phase.
+To address this issue, we post-processing the compiler's output we want to
+ensure the follwoing property:
 
@@
 The intuition is that we $\eta$-expand the occurrence of the variable
-with lower arity to match the higher arity. Since each \linketa can
+with higher arity to match the lower arity. Since each \linketa can
 add exactly one lambda, we need as many links as the difference between the
@@
 
+\begin{comment}
 In order to grant fidelity, we add a new link progression for \linketa{}s.
 
-\begin{comment}
 \newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
@@
 
-If we look back the example give at the beginning of this section, we can
-deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
-and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
-
-After deduplication the compiler output is as follows:\todo{add simple example}
+\paragraph{Example of \mapdeduplication}
+If we take back the example give at the beginning of this section,
+\mapdeduplication produces the following result:
+% 
+% make test ONLY=7007 TEX=tex
+\printAlll
+  {{{\lambda x.(X\appsep x),a},
+    {X,\lambda x.a}}}
+  {{{A,a},
+    {C,\lambda x.a}}}
+  {{{X,B,1}}}
+  {{{\eta,,C,\lambda x.B_{x}},
+    {\eta,,A,\lambda x.B_{x}}}}
+
+\noindent
+Note that $\mapping{X}{B}{1}$ desappears from \mapStore
+in favour of the auxiliary \linketa \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
+The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
+$\linkStore_2$ by \progressetaleft, assigning <<$a \appsep x$>> to $B_x$.
+The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
+turn triggers $\linkStore_1$ by \progressetaleft. This link perfroms
+$\lambda x.a \Ue \lambda x.(a \appsep x)$, which, as expected, fails.
+
 \begin{comment}

### 2024-04-18 paper/main.tex
remove todo
correct a cref

@@
 
-\noindent
-\todo{remove pruning optimization from code above}
+% \noindent
+% \todo{remove pruning optimization from code above}
 
@@
 which ensures that no link causes an occur-check~(\ref{sec:oc})
-and the fact that \linkStore is duplicate-free (\cref{TODO}).
+and the fact that \linkStore is duplicate-free (see \cref{sec:invariant1}).
 

### 2024-04-18 paper/main.tex
reformulate lemma 7.7 + proof

@@
 
-\noindent In order to accommodate for the implementation of heuristics we
-shall relax these failure conditions in~\cref{sec:heuristics}.
+We will relax this last failure condition to accommodate for the
+implementation of heuristics in~\cref{sec:heuristics}.
+
+Finally, in order to correctly correlate a variable appearing in the \lhs of a
+\linketa and in the \rhs of a \linkbeta, we should introduce the following
+\linketa progression rule.
+
+\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
+\begin{definition}[\progressetaright]\label{def:progressetaright} A link
+\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
+$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
+$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
+first case, $X$ is unified with $T$, and in the second one, $X$ is unified with
+$T'$ (under the context $\Gamma$).
+\end{definition}
+
+The idea behind this is to instantiate the variable of a \linketa whenever its
+\rhs is for sure a \wellb term.
 
@@
   \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
-  terminates. Finally \progBetaFail guarantees termination since 
+  terminates. \progBetaFail guarantees termination since 
   %it makes 
   %\elpiIn{progress}
-  it immediately fails.
+  it immediately fails and \progressetaright terminates since it executes 
+  terminating instructions and if its permises succeed, ther considered
+  \linketa is removed from \linkStore.
 %
@@
   The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
-  if \Uo{} fails outside \llambda.
+  if \Uo{} falls outside \llambda.
 \end{theorem}
@@
 \begin{proof}[Proof sketch]
-  Let $\foUnifPb_i$ be the first problem in \foUnifPb not in \llambda and
-  le $\sigma$ be the substitution obtained solving $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$.
-  If $\sigma\hoUnifPb_i$ is in \llambda, then each variable that the
-  compiler used to replace a problemat term is assigned and hence
+  Let $\foUnifPb_i$ be the first problem in \foUnifPb, such that it is not in \llambda and le
+  $\sigma$ be the substitution obtained solving
+  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of $\hstep_{i-1}$,
+  each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 
+  unification problem, is instantiated if it is known for sure
+  that $X$ is no more a \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
+  % each variable that the
+  % compiler used to replace a ``problematic'' term is assigned and hence
   by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
-  been solved and removed. Note that in this case all
-  calls to \Ue{} are between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
+  is solved and removed. Note that, in this case, all
+  calls to \Ue{} are between terms in \llambda and by \cref{th:fidelity-beta}
   fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,

### 2024-04-21 paper/main.tex
re-add link-eta rule in constraint store
base -> basis (Yves)
update tex
minor nits

@@
 \begin{definition}[\maybeeta]\label{def:maybeeta}
-  \maybeeta is the set of terms $s$ such that $\exists \rho, \rho s$
-  is an eta expansion.
+  \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho \lambda x.s$
+  is an eta expansion. % without the lambda, 5.11 is not true, that
+  % is we assume variables are in W
 \end{definition}
@@
 
-The simple compiler described in this section serves as a base for the
+The simple compiler described in this section serves as a basis for the
 extensions in \cref{sec:llam,sec:eta,sec:beta}.
@@
     $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
-    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
+    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x~;~ B \mapsto \lambda y.\lambda x.y ~\}$
   \end{tabular}
@@
     $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
-    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
+    there exists a $w \in W$, such that $w$ \reduceto{} $v$
     (if $n = 0$ this is equivalent to $x \in W$).
@@
 contraction. In that case, the term $t$ is under the spine of binders
-$x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
+$x_1\ldots x_n$ can either be $x$ applied to terms that can
 \reduceto{} these variables (case 2), or a unification variable
@@
 variable \elpiIn{A}. This variable sees all the names free in
-\elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
+\elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
 Invariant~\fullref{inv:linklhs} holds for this link. Moreover:
@@
 The decompilation of a \linketa is performed by unifying
-the \lhs with the \rhs. Note that this unification never fails, since \lhs is a
+the \lhs with the \rhs. Note that this unification never fails, since the \lhs is a
 flexible term not appearing in any other \linketa
@@
 \linketa{}s are meant to delay the unification of ``problematic'' terms until
-we know for sure if the term has to be $\eta$-contracted or not.
+we know for sure if their head lambdas can be $\eta$-contracted or not.
 
@@
     $\lambda x.X \appsep x\Ue{} t$
-    \item if $X = \lambda x.t$ we run $X \Ue{} t$.
+    \item if $X = \lambda x.s$ we run $X \Ue{} t$.
 \end{enumerate}
@@
   By \cref{inv:link-eta-right}, we have $\wellb(t)$.
-  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$: both terms are \wellb.
+  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
   Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
-  are both \wellb.
+  are both in \wellb.
 \end{proof}
@@
   unification performed by the activation of a \linketa is done between
-  terms in \wellb, therefore, the substitution remains \wellb.
+  terms in \wellb.
 \end{proof}
@@
   of \linketa guarantees \fullref{prop:fidelity}.
-  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
+  \footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
   \label{lemma:fidelity-maybeeta}
@@
   % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
-  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
-  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
+  We want to prove that \fstep succeeds iff \hstep
+  succeeds, where \hstep is made by a unification step $u$ and a link
   progression $p$.
-  Without loss of generality,\todo{explain why} we consider a
+  Without loss of generality we consider a
   \Fo unification problem of the form $s_1 \Uo s_2$
-  where $s_1 \in \maybeeta$. $s_1$ is compiled to $X$ with the accessory link
+  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
   $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
@@
     \item $s_2 \notin \maybeeta$,
-          in this case $s_2$ is compiler to $t_2$ and $u$ assigns $t_2$ to $X$
+          in this case $s_2$ is compilet to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
           and triggers \progressetaleft if $t_2$ is rigid.
@@
           progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
-          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}
-          and $p$ unifies $\lambda x.t_1$ with $t_2$. Alternatively, if $s_1$ and $s_2$ fails
+          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}\todo{name}
+          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
           to unify, it cannot be because of the head $\lambda$ constructors,
-          since wither lam-lam or \ref{clause:eta1} move under it and find
+          since with \cref{lam-lam}\todo{name} or \ref{clause:eta1} move under it and find
           different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
@@
 
-In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
-This map allows to decompile the \Ho substitution to the object language.
-Given the higher-order nature of Coq the same unification
-variable $X$ can be used multiple times with different arities.
+We want to allow for partial application of functions.
+As a consequence
+the same unification variable $X$ can be used multiple times
+with different arities.
+Then it is the case the
+memory map \mapStore generated by the compiler is not
+a bijection. For example:
 \begin{comment}
@@
 % \todo{simplify example since no prog right}
-In the unification problems \foUnifPb above, we see that $X$ is used with arity $1$ in
+in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
 $\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
@@
 and any incompatible assignements to them would not be detected, breaking,
-in turn \cref{prop:fidelity}. This incompatibility will only be detected during the
+in turn \cref{prop:fidelity}. 
+% This incompatibility will only be detected during the
 decompilation phase.
-To address this issue, we post-processing the compiler's output we want to
-ensure the follwoing property:
+To address this issue, we post-process the compiler's output
+to ensure the follwoing property:
 
@@
 explained in~\cref{sec:decompilation}. The property above does not
-concern these.
+concern these variables since they are fresh and have no corresponding
+variable in \Fo.
 
@@
 \end{elpicode}
-\todo{improve code}
-%link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
+% \todo{improve code}
+% link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
 

### 2024-04-21 paper/main.tex
reword

@@
 \noindent
-Opting for this method would result in a suboptimal utilization of the logic
+This method fails to take advantage of the logic
 programming engine provided by the ML, as it degrades indexing by

### 2024-04-22 paper/main.tex
minor rewordings

@@
 
-The second step involves allocating new variables in the
-memory of \Fo{}. This technicality is required because some
-higher-order unifications may require pruning a variable. For example,
+The second step allocates
+in the memory of \Fo{}
+new variables used to implement the pruning of higher-order variables. 
+For example,
 $F\appsep x\appsep y = F\appsep x\appsep z$ requires allocating a 
@@
 
-The final step is to decompile each assignment.
-Decompiling a term is straightforward since \mapStore is a bijection.
-The only complex part concerns the \elpiIn{abs} node. In our simple setting,
-the \elpiIn{flam} node carries no additional information (other than the function body),
-so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
-However, in the case of Coq, where lambdas carry the type of the bound variable, one
-must store this information somewhere. Note that this information is similar
-to the arity of variables; that in Coq, unification variables have a
-(function) type, and this type can be used to annotate the lambdas
-needed to express their assignment.
+The final step is to decompile each assignment. The \elpiIn{val}
+node carries a term that easy to decompile since \mapStore is a bijection.
+Since the \elpiIn{flam} node
+carries no additional information (other than the function body),
+each \elpiIn{abs} node can be decompiled to a \elpiIn{flam} one.
+
+However, if the \elpiIn{flam} node was carrying more information,
+as is the case for Coq where it holds the type of the bound variable,
+one would need to store this piece of information in the memory map,
+were we store the arity (that is a very simple function type).
 
@@
 
-\noindent
-The statement above roughly states that the translation from
-\Fo{} to \Ho is not lossy, it can be reversed. The statement
-is made convoluted by the fact that we hid the procedure to
-decompile a term in favor of the one to decompile substitutions.
+% \noindent
+% The statement above roughly states that the translation from
+% \Fo{} to \Ho is not lossy, it can be reversed. The statement
+% is made convoluted by the fact that we hid the procedure to
+% decompile a term in favor of the one to decompile substitutions.
 We omit to sketch the proof of this property for brevity.
@@
 already happens to be in \wellb.
-Later, when the compiler will enforce \cref{prop:w-enforcing}
-the proof will be adjusted to cover for the new cases.
+Later, when the compiler will unsure \fullref{prop:w-enforcing}
+and
+the proof will be refined to cover for the new cases.
 
 \begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
-  The implementation of \Uo above is a good unification (as per \cref{prop:good-unif})
-  for \Eo in the domain \wellb.
+  The implementation of \Uo above is a good unification
+  for \Eo
+  (as per \cref{prop:good-unif})
+  in the domain \wellb.
 \end{lemma}
@@
 \begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
-\fullref{prop:fidelity} hold if $\;\wellb{(\foUnifPb)}$.
+\fullref{prop:fidelity} hold if $\;\wellb{(\foUnifPb)}$. \label{thm:fidfow}
 \end{theorem}
 \begin{proof}[Proof sketch]
-Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same,
-and by \cref{prop:uniffow} \Ue is equivalent to \Uo.
+Since \elpiIn{progress1} is a no-op then \fstep and \hstep are the same.
+By \cref{prop:uniffow}, \Ue is equivalent to \Uo in \wellb.
 \end{proof}
@@
 
-In the following sections, we adopt this notation discuss the compiler's
-output: \foUnifPb represents the input set of problems that compiled to
-\hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
+In the following sections we use the following notation for input and output
+of the compiler. \foUnifPb represents the input unification problems in \Fo{}
+while \hoUnifPb their corresponding compiled version with memory mapping
+\mapStore and links \linkStore. For example:
 %
@@
 
+  \noindent
 We index each sub-problem, sub-mapping, and sub-link with its position
@@
 For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
+
+As we deal with each category of problematic terms we
+weaken the assumptions of~\cref{thm:fidfow} using the following
+definitions:
+
+\begin{definition}[$\wellb_\beta$]
+  $
+  \wellb_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
+  $
+  \end{definition}
+
+\begin{definition}[$\wellb_{\beta \eta}$]
+  $
+  \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
+  $
+\end{definition}
+
 %The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.
@@
 
-In order to make \Uo higher-order, we need to
-take care of terms in \maybebeta.
-In the example below, we can see that the basic compilation given in 
-the previous section is not able to make the \Ho unification problem succeeds.
+% In order to make \Uo higher-order, we need to
+% take care of terms in \maybebeta.
+% In the example below, we can see that
+The basic compiler given in 
+the previous section is unable to make the following
+higher-order unification problem succeed.
 %
@@
 
-\paragraph{Decompilation} Since no link is created by the compilation of 
-\maybebeta terms, no modification to the \elpiIn{commit-link}
-is needed.
-
-\paragraph{Progress} Similarly to decompilation, since no link is produced,
-no modification to the \elpiIn{progress} predicate is needed.
-
-\begin{definition}[$\wellb_\beta$]
-$
-\wellb_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
-$
-\end{definition}
+\paragraph{Decompilation}
+No change
+to \elpiIn{commit-link}
+since no link is added.
+% by the compilation of 
+%\maybebeta terms, no modification to the \elpiIn{commit-link}
+%is needed.
+
+\paragraph{Progress}
+No change to \elpiIn{progress} since no link is added.
+% Similarly to decompilation, since no link is produced,
+% no modification to the \elpiIn{progress} predicate is needed.
 
@@
 as shown by the following example:
+while $\lambda x.X \appsep x \Uo{} f$ does admit the solution
+$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
+\hoUnifPb does not.
 %
@@
 \noindent
-While $\lambda x.X \appsep x \Uo{} f$ does admit the solution
-$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
-\hoUnifPb does not:
-\elpiIn{lam x\ uva A [x]} and
-\elpiIn{con"f"} start with different, rigid, term constructors hence
-\Ue{} fails.
+The reason is that \elpiIn{lam x\ uva A [x]} and
+\elpiIn{con"f"} start with different, rigid, term constructors.
 
@@
 \end{proof}
-
-\begin{definition}[$\wellb_{\beta \eta}$]
-  $
-  \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
-  $
-\end{definition}
 

### 2024-04-22 paper/main.tex
conclusions

@@
 an open world, akin to the \elpiIn{string} data type used in the preceding
-examples, they cannot be symbols of the ML.
+examples, they cannot be symbols of the meta language.
 
@@
 Section 6 contains a discussion about the use of the
-unification procedure of the ML in presence of non ground goals, but
-the authors do not aim at exploiting it to the degree we want.\todo{sucks}
+unification procedure of the meta language in presence of non ground goals, but
+the authors are not interested in exploiting a decidable fragment of
+higher-order unification but are reather leaning towards an interactive
+system where explicit
+control is given the user.
 
@@
 as in Coq, hence the solver does not require a higher-order unification
-procedure.
-
+procedure.\\
+~\\
 \emph{The approach presented in this paper}
-provides a third option that addresses all the concerns
-mentioned earlier. It capitalizes on the benefit of not requiring to fully
-implement
-the unification algorithm of the object language. Instead, it employs the
-unification capabilities of the meta language, facilitated by the various links
-%we establish
-to manage ``problematic'' subterms. As a result of this choice our encoding takes
-advantage of indexing data structures and mode analysis for clause filtering.
+%provides a third option that 
+addresses all the concerns
+mentioned earlier. It takes advantage of the
+unification capabilities of the meta language
+at the price of handling ``problematic'' sub terms on
+the side.\\
+As a result %of this choice 
+our encoding takes
+advantage of indexing data structures and
+mode analysis for clause filtering.
 It it worth mentioning that we only replace terms with variables only
@@
 and hence indexable by the meta language logic programming engine.
+Some preliminary benchmarks on the Stdpp
+library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
+encouraging results: our type-class solver has some
+overhead on small goal but is consistently faster than
+the Coq one starting from goals made of XXX nodes.\\
+\todo{D: how many?}
+Our approach is flexible enough to accommodate different strategies
+and heuristics to handle terms outside the pattern fragment.
 Moreover, the unification process we obtain is ready to take
 advantage of potential improvements to the programming engine,
-such as tabled search, and apply forms of static analysis for
-the meta language, such as determinacy, to the object language.
-\todo{mention preliminar experiment show that our prolog is 2x faster than the Coq one} 
-Finally, our approach is flexible enough to accommodate different strategies
-and heuristics to handle terms outside the pattern fragment, and
-it is not tightly coupled with Coq.
-
-We considered mechanizing our results with Abella~\cite{gacek2008abella}, especially in the early phases.
-Unfortunately we could not. One reason
-is that eliminating the cut operator, although possible, has a high cost in
-verbosity.
-% he same for the \elpiIn{name}
-%\todo{is it sill described?} predicate.
-Another problem is that the logic of Abella does not let one quantify over
-predicates and combinators like \elpiIn{forall} are used pervasively in our
-code.
-\todo{finish}
+such as tabled search. Finally, our shallow encoding can
+enable to apply forms of static analysis for
+the meta language, such as determinacy, also to the object language.
+
+We considered mechanizing our results with Abella~\cite{gacek2008abella},
+especially in the early phase of this work, but unfortunately we could not.
+The main reason is that the logic of Abella does not let one quantify over
+predicates nor user the cut operator. Eliminating cut or
+combinators like \elpiIn{map} is surely possible, but 
+has a high cost in verbosity given how pervasively they are used.
+Moreover, and more importantly, it creates a distance
+between the verified and the actual code, which we found to
+hinder our exploratory phase.
 

### 2024-04-22 paper/main.tex
gzip: page breaking seems now good

@@
 
-For the detection of \notllambda it
-is sufficient to identify applications with a flexible head and arguments that
-are not in distinct names. The following rule for \notllambda compilation
-is inserted  just before rule~\ref{rule:compapp}.
+By inserting this rule
+before rule~\ref{rule:compapp}.
+we identify terms \notllambda, i.e. flexible applications with arguments that
+are not distinct names:
 
@@
 
-Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
-the following activation rules:
+%We add the followins rules to \elpiIn{progress1}.
+Let $l$ be a \linkbeta of the form \linkbetaM{\Gamma}{T}{\rhsBeta}
 
@@
 
-We will relax this last failure condition to accommodate for the
-implementation of heuristics in~\cref{sec:heuristics}.
+We relax this condition and accommodate for
+% the implementation of 
+heuristics in~\ref{sec:heuristics}.
 
@@
 link-eta L R :- not (var L), !, eta-progress-lhs L R.
+link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
 \end{elpicode}
-% \todo{improve code}
-% link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
+\todo{check progress-rhs is back}
 
@@
 
-Different strategies can be used to unify terms of the object language.
-% in the meta langauge. 
+Object-language terms can be unified using
+different strategies.
+
 The first approach that comes to mind consist in implementing \Uo as a
-regular routine in the ML, i.e. write rules as follows:
+regular routine, i.e. write rules as follows
 
 \begin{elpicode}
-  decision X :- X ~\Uo~(all A x\ app [P, x]), finite A,
+  decision X :- unify X (all A x\ app [P, x]), finite A,
     pi x\ decision (app [P, x]).
@@
 \noindent
+where \elpiIn{unify} is a regular predicate.
 This method fails to take advantage of the logic
-programming engine provided by the ML, as it degrades indexing by
-eliminating all data from rule heads. Additionally,
-implementing a unification procedure in the ML is likely to be significantly
+programming engine provided by the meta language since
+it removes all data from rule's heads and makes
+indexing degrade. Additionally,
+implementing a unification procedure in the meta language is likely to be significantly
 slower compared to the built-in one.
@@
 Another possibility is to avoid having application and abstraction nodes
-in the syntax tree, and use the ones of the ML, as in:
+in the syntax tree, and use the meta language ones, as in:
 
@@
 not always feasible to adopt it for Coq due to the fact that the type system
-of the ML is too limited to accommodate for that one the object language,
+of the meta language is too limited to accommodate for that one the object language,
 e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
-Second, the encoding of Coq terms provided by Elpi is primarily utilized for meta
-programming, in order to extend the Coq system. Consequently, it must be able to
+Second, the encoding of Coq terms provided by Elpi is primarily used for meta
+programming, i.e. to extend the Coq system. Consequently, it must be able to
 manipulate terms that are not known in advance
 without relying on introspection primitives such as Prolog's
-\texttt{functor} and \texttt{arg}. In this context, constants need to live in
-an open world, akin to the \elpiIn{string} data type used in the preceding
-examples, they cannot be symbols of the meta language.
+\texttt{functor/3} and \texttt{arg/3}. To avoid that constants 
+cannot be symbols of the meta language, but must rather
+live in
+an open world, akin to the \elpiIn{string} data type used in the \elpiIn{con}
+constructor.
 
@@
 unification capabilities of the meta language
-at the price of handling ``problematic'' sub terms on
-the side.\\
+at the price of handling problematic sub terms on
+the side.
 As a result %of this choice 
@@
 mode analysis for clause filtering.
-It it worth mentioning that we only replace terms with variables only
+It is worth mentioning that we only replace terms with variables only
 when it is strictly needed, leaving the rest of the term structure intact
-and hence indexable by the meta language logic programming engine.
+and hence indexable. % by the meta language logic programming engine.
 Some preliminary benchmarks on the Stdpp
@@
 encouraging results: our type-class solver has some
-overhead on small goal but is consistently faster than
+overhead on small goals but is consistently faster than
 the Coq one starting from goals made of XXX nodes.\\

### 2024-04-24 paper/main.tex
relecture of tex

@@
 the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
-In the rest of this paper shall refer to the pattern fragment as
+In the rest of this paper, we refer to the pattern fragment as
 \llambda~\cite{miller92jsc}.
@@
 Elpi comes equipped with
-an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
+an Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
 terms, called \elpiIn{tm}, that includes (among others) the following
@@
 unification procedures of the meta language weak.
-The equational theory of the meta language Elpi encompasses
+The equational theory of the meta-language Elpi encompasses
 $\eta\beta$-equivalence and its unification procedure can solve higher-order
@@
 We detail an encoding of a logic program in \Fo{} to a strongly related
-logic program in \Ho (the language of the meta-language) and we show that
+logic program in \Ho (the language of the meta language) and we show that
 the higher-order unification procedure of the meta language \Ue{} can be
@@
 Note that the symbol \elpiIn{|} separates the head of a list from the tail,
-that \elpiIn{forall2} applies a binary predicate to two lists (of the same
+that \elpiIn{forall2} applies a binary predicate to two lists (of same
 length). Finally \elpiIn{=>} is $\lambda$Prolog syntax to augment the
@@
 A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
-$\mathcal{D}$ satisfies the following properties:
+$\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
   \begin{gather}
@@
 A term $s$ is compiled to a term $t$ where every
-``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
+``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
 with an accessory \emph{link} that represents a suspended unification problem
-$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
+$h \Ue p$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
 it does not contradict \Eo as it would otherwise do on the
@@
 its syntax tree cannot be used to justify a
-unification failure, i.e. by properly instantiating $F$ the term
+unification failure, i.e. by properly instantiating $F$, the term\noteD{or a name}
 head constructor may become a $\lambda$, or a constant, or remain an application.
@@
 since the substitution
-$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
+$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$\noteD{a,b are reserved for constants}
 makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
@@
 We write $\wellb(t)$ as a short for $\wellb(\{t\})$.
-We claim our compiler validates the following property:
+We claim our compiler validates the following properties:
 
@@
 \noindent
-A solution to a \Fo{} variable is a plain term, that is \elpiIn{fsubst}
+A solution to a \Fo{} variable is a plain term, that is, \elpiIn{fsubst}
 is an abbreviation for \elpiIn{mem fm}.
@@
 The signature of the \elpiIn{comp} predicate below allows for the generation of
-links (suspended unification problems), which play no role in this section
-but play a major role in \cref{sec:llam,sec:eta,sec:beta}.
+links (suspended unification problems), which plays no role in this section
+but plays a major role in \cref{sec:llam,sec:eta,sec:beta}.
 With respect to \cref{sec:problem-statement}, the signature also allows
@@
 unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
-but should be disallowed if a $\linkStore$ contains a link like
+but should be disallowed if $\linkStore$ contains a link like
 $\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
-a lambda in head position, but we surely know it contains $X$, hence
-$f~Y$ and that fails the occur check.
+a lambda in head position, but we surely know it contains $X$, hence,
+% $f~Y$ and 
+that fails the occur check.
 The procedure \elpiIn{occur-check-links} is in charge of
@@
 The final step is to decompile each assignment. The \elpiIn{val}
-node carries a term that easy to decompile since \mapStore is a bijection.
+node carries a term that is easy to decompile since \mapStore is a bijection.
 Since the \elpiIn{flam} node
@@
 
-We already have all the pieces to show the code of \Ue{}.
+We already have all the ingredients to show the code of \Uo{}.
 
@@
 \begin{proof}[Proof sketch]
- In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
+ In this setting, \Ee is as strong as \Eo on ground terms\noteD{module var reflexivity}. What we have to show is that whenever two different \Fo{}
 terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
@@
 \elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
-$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
+$Y \Ue t$\noteD{shouldn't be uva Y [] \Ue t} that succeeds with $\sigma = \{ Y \mapsto t\}$ and
 $\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
@@
 \end{theorem}
-\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
+\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
   \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
@@
     \item $s_2 \notin \maybeeta$,
-          in this case $s_2$ is compilet to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
-          and triggers \progressetaleft if $t_2$ is rigid.
+          in this case $s_2$ is compile to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
+          and triggers \progressetaleft.
           If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
           progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
-          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}\todo{name}
+          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \ref{clause:lam-lam}
           and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
-          to unify, it cannot be because of the head $\lambda$ constructors,
-          since with \cref{lam-lam}\todo{name} or \ref{clause:eta1} move under it and find
-          different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
+          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
+          since \ref{clause:lam-lam} or \ref{clause:eta1} try to unify under it, eventually finding two
+          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
           \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
@@
 \todo{check progress-rhs is back}
+\noteD{first rule is wrong, since maybe-eta is 1-argument pred}
 

### 2024-04-24 paper/main.tex
minor TODOs

@@
 \begin{proposition}[Good unification]\label{prop:good-unif}
-A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
+A good unification $\UnifRel\;$ for equality $\;\EqualRel\;$ in the domain
 $\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
@@
 its syntax tree cannot be used to justify a
-unification failure, i.e. by properly instantiating $F$, the term\noteD{or a name}
-head constructor may become a $\lambda$, or a constant, or remain an application.
+unification failure, i.e. by properly instantiating $F$, the term
+head constructor may become a $\lambda$, or a constant, or a name, or remain an application.
 
@@
 since the substitution
-$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$\noteD{a,b are reserved for constants}
+$\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
 makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
@@
 If we look at the syntax of \Fo{}, terms the only interesting case is
-\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
-$Y \Ue t$\noteD{shouldn't be uva Y [] \Ue t} that succeeds with $\sigma = \{ Y \mapsto t\}$ and
+<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
+<<\elpiIn{uva Y [] ~\Ue~t}>>
+that succeeds with $\sigma = \{ Y \mapsto t\}$ and
 $\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
@@
 \begin{elpicode}
-link-eta L R :- var L, maybe-eta R HolesInR, !,
-  declare_constraint (link-eta L R) [L|HolesInR].
+link-eta L R :- var L, maybe-eta R, !,
+  declare_constraint (link-eta L R) [L,R].
 link-eta L R :- not (var L), !, eta-progress-lhs L R.
@@
 \end{elpicode}
-\todo{check progress-rhs is back}
-\noteD{first rule is wrong, since maybe-eta is 1-argument pred}
 

### 2024-04-25 paper/main.tex
typo
reformulate 2.2

@@
 
-\subsection{Equational theories an unification}
+\subsection{Equational theories and unification}
 
@@
 \newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
+\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
 \newcommand{\runFD}[2]{\ensuremath{%
@@
 \newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
+\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
 \newcommand{\runHD}[3]{\ensuremath{%
@@
 $$
-  \runF{\foUnifPb}{\mathcal{N}}{\rho}
-  \Leftrightarrow
-  \runH{\foUnifPb}{\mathcal{N}}{\rho}
+  \runFx{\foUnifPb}{\mathcal{N}}{\rho} \land
+  \runHx{\foUnifPb}{\mathcal{N}}{\rho'}
+  \Rightarrow
+  \forall s \in \foUnifPb, \rho s \Eo \rho' s
 $$
@@
 \noindent
-That is, the two executions give the same result if all terms in
+That is, the two executions give equivalent results if all terms in
 \foUnifPb are in the pattern fragment. Moreover:
@@
 To address this issue, we post-process the compiler's output
-to ensure the follwoing property:
+to ensure the following property:
 

### 2024-04-25 paper/main.tex
fix typos

@@
 higher-order unification but are reather leaning towards an interactive
-system where explicit
-control is given the user.
+system where the user is given
+control over the search procedure.
 

### 2024-04-25 paper/main.tex
massive rename F -> O bject, H -> M eta, T -> Q , N -> n

@@
 
-\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
-\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
-\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
-\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
+\newcommand{\Ue}{\ensuremath{\UnifRel_m}\xspace}
+\newcommand{\nUe}{\ensuremath{\nUnifRel_m}\xspace}
+\newcommand{\Ee}{\ensuremath{\EqualRel_m}\xspace}
+\newcommand{\nEe}{\ensuremath{\nEqualRel_m}\xspace}
 \newcommand{\llambda}{\ensuremath{\mathcal{L}}\xspace}
@@
 %\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
-\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}\xspace}}{F}} % space non va
-\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}}\xspace}{H}}
+\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{O}\xspace}}{O}} % space non va
+\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{M}}\xspace}{M}}
 
@@
 \newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
-\newcommand{\hoUnifPb}{\ensuremath{\mathbb{T}}\xspace}
+\newcommand{\hoUnifPb}{\ensuremath{\mathbb{Q}}\xspace}
 
@@
 a minimal setting to ease its study. In this setting, we have
-a \Fo{} language (for first order) with a rich equational
-theory and a \Ho{} meta language with a simpler one.
+a (first order) language \Fo{} with a rich equational
+theory and a \Ho{} language with a simpler one.
 
@@
 We represent a logic program \emph{run} in \Fo{} as
-a sequence of \emph{steps} of length $\mathcal{N}$.
+a sequence of \emph{steps} of length $n$.
 At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
@@
 empty substitution $\rho_0$ produces the final
-substitution $\rho_\mathcal{N}$, which is the result of the
+substitution $\rho_n$, which is the result of the
 logic program execution.
@@
 \newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
-\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
-\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
-\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
-\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
+\newcommand{\fstep}{\ensuremath{\mathrm{ostep}}\xspace}
+\newcommand{\hstep}{\ensuremath{\mathrm{mstep}}\xspace}
+\newcommand{\frun}{\ensuremath{\mathrm{orun}}\xspace}
+\newcommand{\hrun}{\ensuremath{\mathrm{mrun}}\xspace}
 \newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
@@
 \stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
-\runF{\foUnifPb}{\mathcal{N}}{\rho}
+\runF{\foUnifPb}{n}{\rho}
 \deff
-\runFD{\foUnifPb}{\mathcal{N}}
+\runFD{\foUnifPb}{n}
 \end{array}
@@
   \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
-  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
+  \runH{\foUnifPb}{n}{\rho} \deff \vspace{2pt}\\
   \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
-  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
-  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
+  \qquad \runHD{\hoUnifPb}{n}{\linkStore}\vspace{2pt}\\
+  \qquad \D{\sigma_{n}}{\mapStore}{\linkStore_{n}}{\rho_{n}}
 \end{array}
@@
 \begin{proposition}[Run equivalence]\label{prop:simulation}
-$\forall \foUnifPb, \forall \mathcal{N},$ if $~\foUnifPb \subseteq \llambda$
+$\forall \foUnifPb, \forall n,$ if $~\foUnifPb \subseteq \llambda$
 $$
-  \runFx{\foUnifPb}{\mathcal{N}}{\rho} \land
-  \runHx{\foUnifPb}{\mathcal{N}}{\rho'}
+  \runFx{\foUnifPb}{n}{\rho} \land
+  \runHx{\foUnifPb}{n}{\rho'}
   \Rightarrow
@@
 In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
-$\forall p \in 1 \ldots \mathcal{N},$
+$\forall p \in 1 \ldots n,$
 $$

### 2024-04-25 paper/main.tex
add comment to simulation and fidelity

@@
 
+In \cref{sec:llam,sec:eta,sec:beta,sec:invariant1}, we will ensure
+\Cref{prop:fidelity}, since we consider this property quite important. Our aim
+is to guarantee that the success of \hstep coincides with the success of \fstep
+at each step $p$, regardless of any newly introduced modifications in our
+unification process. We postpone the discussion of \Cref{prop:fidelity-recovery}
+to \cref{sec:beta}, where we delve into unifications involving terms outside
+\llambda. Moreover, we reserve the proof of \Cref{prop:simulation} only in
+\cref{sec:llam}, due to the absence of links. In the other sections, we omit the
+proof of this latter property due to space constraints. However, the underlying
+idea remain constant: link activation refines $\sigma$ so that if $\rho_i
+\foUnifPb_{i_l} \Uo \sigma_i \foUnifPb_{i_r}$ then $\sigma_i \hoUnifPb_{i_l} \Ue
+\sigma_i \hoUnifPb_{i_r}$. Finally, \D{\sigma_n}{\mapStore}{\linkStore_n}{\rho}
+and $\rho$ is equivalent to the substitution returned by $\frun(\foUnifPb,n)$.
+
 \subsection{The solution (in a nutshell)}

### 2024-04-26 paper/main.tex
move description of proofs to end of section 2 since we need the idea of links to get it. Also make it short.

--- 2024-04-25 paper/main.tex
@@
 
+We prove \fullref{prop:fidelity} in
+\cref{sec:eta,sec:beta,sec:invariant1}
+and \fullref{prop:simulation} only in \cref{sec:llam},
+where links are not present yet.
+In order to prove \cref{prop:simulation} in the presence of links
+one needs to refine \cref{prop:fidelity} by relating the two
+substitutions: they are equivalent only by using decompilation to 
+take into account terms in \linkStore.
+Then, by induction
+on the number of steps, \ref{prop:fidelity} implies \Cref{prop:simulation}.
+We discuss \fullref{prop:fidelity-recovery} in \cref{sec:beta}.
+
 % Note that proposition \ref{prop:w-preservation} does not hold for \hrun as a whole
@@
 
-When detailing examples, we represent links as equations between two
-terms under a context.
+In the examples we represent links as equations
+%between two terms
+under a context.
 The equality sign is subscripted with the

### 2024-04-29 paper/main.tex
proofread section 2

@@
 For both languages, we extend the equational theory
-over ground terms to the full language by adding the reflexivity for
-unification variables (a variable is equal to itself).
+over ground terms to the full language by adding reflexivity for
+unification variables, i.e. a unification variable is equal to itself.
 
@@
 \noindent
-Note that the symbol \elpiIn{|} separates the head of a list from the tail,
+Note that the symbol \elpiIn{|} separates the head of a list from the tail;
 that \elpiIn{forall2} applies a binary predicate to two lists (of same
-length). Finally \elpiIn{=>} is $\lambda$Prolog syntax to augment the
-current program with an extra rule, and both procedures use it to
+length); and that \elpiIn{=>} is $\lambda$Prolog syntax for
+implication an is used to augment the
+current program with an extra rule. Both procedures use implication to
 assume that the fresh nominal constants introduces by \elpiIn{pi} are
@@
 it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
-weak head normal form of <<\elpiIn{app [F|L]}>>.
+weak-head normal form of <<\elpiIn{fapp [F|L]}>>.
 
@@
 We write \llambda as the set of terms that are in the pattern-fragment, i.e.
-every higher-order variable is applied to a list of distinct names.
+every unification variable is applied to a list of distinct names.
 
@@
 Even if we provide an implementation of the object-language unification
-\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
-logic program.
+\Uo{} in \cref{sec:founif}, our real goal is to simulate the execution of
+an entire logic-program.
 
@@
 We represent a logic program \emph{run} in \Fo{} as
-a sequence of \emph{steps} of length $n$.
-At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
+a sequence of $n$ \emph{steps}. Each step consists in
+unifying two terms, $\foUnifPb_{p_l}$ and
 $\foUnifPb_{p_r}$, taken from the list of all unification
@@
 \noindent
-In order to simulate a \Fo{} logic program in \Ho{}, we compile
-each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
+In order to simulate a logic program we start by compiling
+\foUnifPb to \hoUnifPb, i.e.
+each \Fo{}-term $s \in \foUnifPb$ is translated to a \Ho{}-term $t \in \hoUnifPb$.
 We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
@@
 description is deferred to \cref{sec:nutshell}.
-We write $\foUnifPb_p = \{~ \foUnifPb_{p_l}, \foUnifPb_{p_r} ~\}$
+We write $\foUnifPb_p = \{~ \foUnifPb_{p_l},\, \foUnifPb_{p_r} ~\}$
 and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.
 
-We simulate each run in \Fo{} with a run in \Ho as follows:
+Then we simulate each run in \Fo{} with a run in \Ho as follows:
 %
@@
 for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
-for the set $\{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$.\\
-\hstep{} is made of two sub-steps: a call to the meta language
+for the set $\{~ \hoUnifPb_{p_l},\, \hoUnifPb_{p_r} ~\}$.\\
+The \hstep{} procedure is made of two sub-steps: a call to the meta-language
 unification and a check for \progress{} on the set of links, that intuitively
 will compensate for the weaker equational theory honored by \Ue.
-\hrun{} compiles all terms in \foUnifPb{}, then executes each step, and
+Procedure \hrun{} compiles all terms in \foUnifPb{}, then executes each step, and
 finally decompiles the solution.
@@
 
+\noindent
 In other words, if the two terms involved in a step
@@
 succeeds by backtracking, exploring multiple branches, could be
-modeled as a set of runs with (possibly non-empty) common prefixes.
+modeled as a set of runs with a (possibly non-empty) common prefix.
 
@@
 A term $s$ is compiled to a term $t$ where every
-``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
+``problematic'' subterm $e$ is replaced by a fresh unification variable $h$
 with an accessory \emph{link} that represents a suspended unification problem
-$h \Ue p$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
+$h \Ue e$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
 it does not contradict \Eo as it would otherwise do on the
@@
 makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
-which is the eta long form of $f$. This term is problematic since
+which is the eta-long form of $f$. This term is problematic since
 its leading $\lambda$ abstraction cannot justify a
@@
 but they cannot be handled directly by the unification of the meta language, which
-is only required to handle terms in \llambda. Still, there exists a
-substitution $\rho$ such that $\rho s \in \llambda$.
+is only required to handle terms in \llambda. Still, for a $s \in \notllambda$
+there may exist a substitution $\rho$ such that $\rho s \in \llambda$.
 
@@
 \newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
-We write $\subterm{t}$ the set of sub-terms of $t$, and
+We write $\subterm{t}$ for the set of sub-terms of $t$, and
 %\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
@@
 \begin{definition}[Well behaved set]
-Given a set of terms $X \subseteq \Ho{}$,
+Given a set of terms $X$, % \subseteq \Ho{}$,
 $$
@@
 input is not. And if two terms can unify in the source language
-then these terms also unify in the target language. 
+then their images also unify in the target language. 
 Note that the property holds for any substitution, $\rho$ could be given by an
 oracle and/or not necessarily be a most general one: when applied to terms in \wellb,
-\Ue{} does not contradict \Eo. Also note that a compiler that
+\Ue{} does not contradict \Eo. Note that a compiler that
 translate all terms to unification variables would satisfy this
-property, but wound fail to verify \cref{prop:fidelity}.
+property, but wound certainly fail to verify \fullref{prop:fidelity}.
 
@@
 one needs to refine \cref{prop:fidelity} by relating the two
-substitutions: they are equivalent only by using decompilation to 
+substitutions $\rho_p$ and $\sigma_p$: they are equivalent only by using decompilation to 
 take into account terms in \linkStore.
 Then, by induction
-on the number of steps, \ref{prop:fidelity} implies \Cref{prop:simulation}.
+on the number of steps, \cref{prop:fidelity} implies \cref{prop:simulation}.
 We discuss \fullref{prop:fidelity-recovery} in \cref{sec:beta}.

### 2024-04-29 paper/main.tex
proof read abstract

@@
   pattern fragment~\cite{miller92jsc}.
-  Elpi offers an encoding of Coq terms suitable
+  Elpi offers an encoding of Coq terms that is suitable
   for meta programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}

### 2024-04-29 paper/main.tex
update tex
remove 2 todos

@@
 A good unification $\UnifRel\;$ for equality $\;\EqualRel\;$ in the domain
-$\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
+$\mathcal{D}$ satisfies the following properties:%\noteD{use $\sigma$, not $\rho$}
   \begin{gather}
@@
 \begin{proof}[Proof sketch]
- In this setting, \Ee is as strong as \Eo on ground terms\noteD{module var reflexivity}. What we have to show is that whenever two different \Fo{}
+In this setting, \Ee is as strong as \Eo on ground terms and reflexivity for
+unification variables. What we have to show is that whenever two different \Fo{}
 terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
@@
 
-\begin{proof}[Proof sketch]\todo{to reformulate}
+\begin{proof}[Proof sketch] %\todo{to reformulate}
   % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$

### 2024-04-29 paper/main.tex
abstract no cite for copy paste

@@
 \end{abstract}
+
+\begin{comment}
+Specifying and implementing a proof system from scratch requires significant
+effort. Logical Frameworks and Higher Order Logic Programming Languages provide
+dedicated, high-level meta languages to facilitate this task in two ways: 1)
+variable binding and substitution are for free when meta language binders
+represent object logic ones; 2) proof construction, and
+proof search, are greatly simplified by leveraging the unification procedure
+provided by the meta language. Notable examples of meta languages are Elf,
+Twelf, $\lambda$Prolog, Beluga, Abella and Isabelle which have been used to
+implement or specify many formal systems such as First Order Logic, Set Theory,
+Higher Order Logic, and the Calculus of Constructions.
+
+The object logic we are interested in is Coq's type theory. We aim to develop a
+higher-order unification-based proof search procedure using the meta language
+Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
+$\eta\beta$-equivalence and features a higher-order unification procedure \Ue
+for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
+for meta programming but that restricts \Ue to first-order unification problems
+only. We refer to this basic encoding as \Fo.
+
+In this paper we propose a more well-behaved encoding called \Ho, and show how
+to translate unification problems from \Fo{} to corresponding ones in \Ho.
+Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
+that honours $\eta\beta$-equivalence (for Coq functions), addresses problems
+within the pattern fragment, and allows for the use of heuristics to deal with
+problems outside the pattern fragment. Moreover, as \Uo{} delegates most of the
+work to \Ue, it can be used to efficiently simulate a logic program in \Fo{} by
+taking advantage of unification-related optimizations of the meta language, such
+as clause indexing.
+\end{comment}
 

### 2024-04-29 paper/main.tex
minor

@@
 of unification problems that constitute the execution of a logic program.
+
 In fact
@@
 % after the instantiation of \elpiIn{P}:
-
 % \begin{elpicode}
@@
 % \end{elpicode}
-
 % \noindent
@@
 We detail an encoding of a logic program in \Fo{} to a strongly related
-logic program in \Ho (the language of the meta language) and we show that
+logic program in \Ho, the language of the meta language, and we show that
 the higher-order unification procedure of the meta language \Ue{} can be
-efficiently used to simulate a higher-order unification procedure \Uo for
+used to simulate a higher-order unification procedure \Uo for
 the object language that features $\eta\beta$-conversion. We show how \Uo

### 2024-04-30 paper/main.tex
minor
some typos in conclusion

@@
 but they cannot be handled directly by the unification of the meta language, which
-is only required to handle terms in \llambda. Still, for a $s \in \notllambda$
+is only required to handle terms in \llambda. Still, given $s \in \notllambda$
 there may exist a substitution $\rho$ such that $\rho s \in \llambda$.
@@
 mode analysis for clause filtering.
-It is worth mentioning that we only replace terms with variables only
+It is worth mentioning that we replace terms with variables only
 when it is strictly needed, leaving the rest of the term structure intact
@@
 overhead on small goals but is consistently faster than
-the Coq one starting from goals made of XXX nodes.\\
+the Coq one starting from goals made of XXX nodes.
 \todo{D: how many?}
-Our approach is flexible enough to accommodate different strategies
+Moreover our approach is flexible enough to accommodate different strategies
 and heuristics to handle terms outside the pattern fragment.
-Moreover, the unification process we obtain is ready to take
-advantage of potential improvements to the programming engine,
-such as tabled search. Finally, our shallow encoding can
-enable to apply forms of static analysis for
-the meta language, such as determinacy, also to the object language.
+Finally, our encoding is sufficiently shallow to: 1)
+allow to lift forms of static analysis for
+the meta language, such as determinacy, also to the object
+language; 2) take advantage of future improvements to the
+logic-programming engine of Elpi, such as tabled search.
 
@@
 The main reason is that the logic of Abella does not let one quantify over
-predicates nor user the cut operator. Eliminating cut or
-combinators like \elpiIn{map} is surely possible, but 
+predicates nor use the cut operator. Eliminating cut or
+higher order
+combinators like \elpiIn{map} or \elpiIn{forall2} is surely
+possible, but 
 has a high cost in verbosity given how pervasively they are used.
@@
 between the verified and the actual code, which we found to
-hinder our exploratory phase.
+hinder our exploration. We did, however, establish a
+test suite with about a hundred cases.
+It is available at \githubUrl.
 

### 2024-04-30 paper/main.tex
fix code snippet

@@
 
-In this paper we did study a compiler and simulation loop on a minimal language.
+In this paper we did study a compiler and a simulation loop on a minimal language.
 The actual implementation uses the Coq-Elpi meta language to both compile
@@
 the runtime of the programming language. In particular the runtime iterates
-\Ue, but \hstep also needs to check for \progress. Luckily Elpi extends~\cite{TASSI_2019}
-$\lambda$Prolog with syntactic constraints (suspended goals) and
-Constraint Handling Rules (CHR)~\cite{fruehwirth2017constraint} to operate on them.
+\Ue, but \hstep also needs to check links for \progress. Luckily Elpi extends
+$\lambda$Prolog with constraints (suspended goals) and
+Constraint Handling Rules (CHR) to operate on them~\cite{TASSI_2019,fruehwirth2017constraint}.
 An important design choice is that constraints are suspended on a user
-computed list of unification variables, and are resumed as soon as one of them
-is assigned, before any other existing goal is considered. In turn this
-property enables a timely activation of links that grants~\fullref{prop:fidelity}.
+computed list of unification variables, and are resumed \emph{as soon as one of them
+is assigned}, before any other existing goal is considered. In turn this
+property allows for a timely activation of links that grants~\fullref{prop:fidelity}.
 For brevity we only provide two pseaudo-code snippets. The first one depicts
@@
 link-eta L R :- not (var L), !, eta-progress-lhs L R.
-link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
+link-eta L R :- not (maybe-eta R), !, eta-progress-rhs L R.
 \end{elpicode}

### 2024-05-02 paper/main.tex
update constraint code snippet
minor
proofreading

@@
 analogy with Logic Programming we want to leverage the
-Elpi meta-programming language~\cite{tassi:hal-01637063},
+Elpi meta programming language~\cite{tassi:hal-01637063},
 a dialect of $\lambda$Prolog, already used to extend
@@
 a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
-The solver has to infer a value for \coqIn{P} by unifying the conclusion
+The solver infers a value for \coqIn{P} by unifying the conclusion
 of rule 3 with the goal, and in particular, it has to solve the unification
 problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
-pattern fragment and admits a unique
-solution $\customlabel{solution:intro}{\rho}$ that assigns
-the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
+pattern fragment and admits a unique solution:
+\begin{elpicode}
+P = ~$\lambda$~x.nfact x 3                                       ~\customlabel{solution:intro}{(\rho)}~
+\end{elpicode}
 In the rest of this paper, we refer to the pattern fragment as
@@
 language its type is \elpiIn{tm}. If we try to back-chain the rule
-\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below
+\ref{clause:r3} on the encoding of the goal \ref{goal:g} given here
 
@@
 the equational theory of Coq is much richer, for efficiency and predictability
-reasons, automatic proof search procedures typically employ a unification
+reasons automatic proof search procedures typically employ a unification
 procedure that only captures a $\eta\beta$-equivalence and only operates
@@
 
-To reason about unification, we provide a description of the
+In order to reason about unification we provide a description of the
 \Fo{} and \Ho languages where unification variables
@@
 in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
-the arguments of a higher-order variable are given via the \elpiIn{fapp}
+the ``arguments'' of a higher-order variable are given via the \elpiIn{fapp}
 constructor. For example the term <<\coqIn{P x}>> is represented as
 <<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
-of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
-In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
+of \elpiIn{P} and \elpiIn{x} is a bound variable.
+In \Ho{} the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
 since unification variables are higher-order and come equipped with an
@@
 over ground terms to the full language by adding reflexivity for
-unification variables, i.e. a unification variable is equal to itself.
+unification variables, i.e. a unification variable is equal to itself
+but different to any other term.
 
@@
 Note that the symbol \elpiIn{|} separates the head of a list from the tail;
-that \elpiIn{forall2} applies a binary predicate to two lists (of same
+that \elpiIn{forall2} applies a binary predicate to two lists (of equal
 length); and that \elpiIn{=>} is $\lambda$Prolog syntax for
-implication an is used to augment the
+implication and is used to augment the
 current program with an extra rule. Both procedures use implication to
@@
 $\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
-$x$ does not occur free in $t$, and conversely, $t$ is the $\eta$-contraction of
+$x$ does not occur free in $t$, and conversely we say that $t$ is the $\eta$-contraction of
 $\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
@@
 \begin{elpicode}
-link-eta L R :- var L, maybe-eta R, !,
-  declare_constraint (link-eta L R) [L,R].
 link-eta L R :- not (var L), !, eta-progress-lhs L R.
 link-eta L R :- not (maybe-eta R), !, eta-progress-rhs L R.
+link-eta L R :- declare_constraint (link-eta L R) [L,R].
 \end{elpicode}

### 2024-05-02 paper/main.tex
proofreading 1, 2 and 3

@@
 We represent a logic program \emph{run} in \Fo{} as
-a sequence of $n$ \emph{steps}. Each step consists in
+a sequence of $n$ \emph{steps}. Each step $p$ consists in
 unifying two terms, $\foUnifPb_{p_l}$ and
@@
 problems \foUnifPb.
+We write $\foUnifPb_p = \{~ \foUnifPb_{p_l},\, \foUnifPb_{p_r} ~\}$
+and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.
 % \footnote{If the same rule is used multiple times in a run we
@@
 \newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
-\newcommand{\fstep}{\ensuremath{\mathrm{ostep}}\xspace}
-\newcommand{\hstep}{\ensuremath{\mathrm{mstep}}\xspace}
-\newcommand{\frun}{\ensuremath{\mathrm{orun}}\xspace}
-\newcommand{\hrun}{\ensuremath{\mathrm{mrun}}\xspace}
+\newcommand{\fstep}{\ensuremath{{\mathrm{step}}_o}\xspace}
+\newcommand{\hstep}{\ensuremath{{\mathrm{step}}_m}\xspace}
+\newcommand{\frun}{\ensuremath{{\mathrm{run}}_o}\xspace}
+\newcommand{\hrun}{\ensuremath{{\mathrm{run}}_m}\xspace}
 \newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
@@
 The variable map connects unification variables in \Ho to variables
-in \Fo{} and is used to ``decompile'' the assignment,
+in \Fo{} and is used to ``decompile'' the substitution, that is written
 $\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
 description is deferred to \cref{sec:nutshell}.
-We write $\foUnifPb_p = \{~ \foUnifPb_{p_l},\, \foUnifPb_{p_r} ~\}$
-and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.
 
@@
 In the context of \hrun, if 
-$\rho_{p-1} \foUnifPb_{p} \in \llambda$ 
-(even if $\;\foUnifPb_{p} \not\in \llambda$)
+$\rho_{p-1} \foUnifPb_{p} \subseteq \llambda$ 
+(even if $\;\foUnifPb_{p} \not\subseteq \llambda$)
 then
@@
 
-\paragraph{Backtracking} We omit it from our model of a logic-program execution
+\paragraph{Backtracking} We omit it from our model of logic-program execution
 since it plays a very minor role, orthogonal to higher-order unification.
@@
 ``problematic'' sub-terms.\\
-We now define ``problematic'' and ``well behaved'' more formally
+We now define ``problematic'' and ``well behaved'' more formally.
 We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
-and all problematic terms are characterized by some ``uncertainty''.
+and all problematic terms are characterized by some uncertainty.
 
 \begin{definition}[\maybebeta]\label{def:maybebeta}
-  \maybebeta is the set of terms of the form $X \appsep x_1 \ldots x_n$
+  \maybebeta is the set of terms of the form $F \appsep x_1 \ldots x_n$
   such that $x_1 \ldots x_n$ are distinct names (of bound variables).
@@
 \begin{definition}[\notllambda]\label{def:notllambda}
-  \notllambda is the set of terms of the form $X \appsep t_1 \ldots t_n$
+  \notllambda is the set of terms of the form $F \appsep t_1 \ldots t_n$
   such that $t_1 \ldots t_n$ are not distinct names.
@@
 \noindent
-In other words the compiler outputs terms in \wellb, even if its
-input is not. And if two terms can unify in the source language
-then their images also unify in the target language. 
-Note that the property holds for any substitution, $\rho$ could be given by an
-oracle and/or not necessarily be a most general one: when applied to terms in \wellb,
-\Ue{} does not contradict \Eo. Note that a compiler that
+In other words the compiler outputs terms in \wellb even if its
+input is not. And if two terms can be unified in the source language
+then their images can also be unified in the target language. 
+Note that the property holds for any substitution, not necessarily a most
+general one: when applied to terms in \wellb
+the meta-language unification 
+\Ue{} agrees with the object-language equality \Eo. Note that a compiler that
 translate all terms to unification variables would satisfy this
-property, but wound certainly fail to verify \fullref{prop:fidelity}.
+property, but wound certainly fail to verify \cref{prop:fidelity}.
 
@@
 are not reintroduced by \hstep, hence \Ue{} can continue to operate properly.
-In \cref{sec:simulation,sec:eta,sec:beta}
+In \cref{sec:llam,sec:eta,sec:beta}
 we describe how to %the compiler
 recognize terms in \maybebeta, \maybeeta and
-\notllambda and how \progress takes care of them preserving \wellb
-and ensuring \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.
+\notllambda and how \progress takes care of links and how it preserves \wellb
+and ensures \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.
 
@@
 \subsubsection{Scope check}\label{sec:sc}
-The predicate \elpiIn{scope-check} walks through all links. 
-Each link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} is replaced by
+The predicate \elpiIn{scope-check} 
+replaces each
+link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} with
+a link
 \linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
 The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
-executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$. This unification
+executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$ using a fresh $Y$.
+This unification
 ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
-and therefore grants fidelity.
-In fact, a link represents a suspended unification problem between a unification variable $X$
-and a ``problematic'' term $t$. This unification is possible iff 
-the free variables in $t$ are also in the scope of $X$.
+and if it fails then \progress hence \hstep fails.
+In turn this grants fidelity in the case where the \lhs of an \linketa
+is pruned of a name that occurs (rigidly) in $t$:
+links represent suspended unification problems
+\emph{that may succeed}.
 
@@
 $\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
-a lambda in head position, but we surely know it contains $X$, hence,
-% $f~Y$ and 
-that fails the occur check.
+a lambda in head position, but we surely know it contains $X$.
 The procedure \elpiIn{occur-check-links} is in charge of
@@
 already happens to be in \wellb.
-Later, when the compiler will unsure \fullref{prop:w-enforcing}
+Later, when the compiler will ensure \fullref{prop:w-enforcing}
 and
@@
 
-\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
+\begin{theorem}[Properties of \Uo in \wellb]\label{prop:uniffow}
   The implementation of \Uo above is a good unification
@@
   in the domain \wellb.
-\end{lemma}
+\end{theorem}
 \begin{proof}[Proof sketch]
-In this setting, \Ee is as strong as \Eo on ground terms and reflexivity for
-unification variables. What we have to show is that whenever two different \Fo{}
-terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
-and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
+In this setting, \Ee is as strong as \Eo on ground terms.
+What we have to show is that whenever two different \Fo{}
+terms can be made equal by a substitution $\rho$,
+we can produce this $\rho$ by finding
 a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
-If we look at the syntax of \Fo{}, terms the only interesting case is
+If we look at the syntax of \Fo{}, the only interesting case is
 <<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
@@
   $\sigma$ be the substitution obtained solving
-  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of $\hstep_{i-1}$,
+  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of \hstep
+  for the $i-1$ time,
   each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 

### 2024-05-03 paper/main.tex
proofreading

@@
 \begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
-  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
-  \cref{prop:uniffowb}.
+  \Ue is as powerful as \Uo in $\wellb_\beta$, as well as in \wellb by
+  \cref{thm:fidfow}.
 \end{proof}
@@
 and is activated under some conditions.
-Activation is implemented by extending the \elpiIn{progress1}
+Activation is implemented
+in \cref{sec:etaprogress}
+by extending the \elpiIn{progress1}
 predicate defined in \cref{sec:execution}.
@@
 When compiling a term $t$, we need to determine if any
-subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
-where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
+subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$
+can be an $\eta$-expansion, i.e. if
 there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
@@
 $\eta$-expansion.\\
-We can now define more formally how \maybeeta terms are detected together with
-its auxiliary functions:
+We can now detail how \maybeeta terms are detected.
 
@@
 If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
-contraction. In that case, the term $t$ is under the spine of binders
-$x_1\ldots x_n$ can either be $x$ applied to terms that can
+contraction. In that case, the term $t$ under the spine of binders
+$x_1\ldots x_n$ can either be $x$ applied to terms that
 \reduceto{} these variables (case 2), or a unification variable
-that can be assigned to that application (case 3).
+that can be assigned to the application $x \appsep x_1 \ldots x_n$ (case 3).
 \end{proof}
@@
 \elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
-Invariant~\fullref{inv:linklhs} holds for this link. Moreover:
+\Fullref{inv:linklhs} holds for this link. Moreover:
 
@@
 \begin{lemma}
-  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb(t)$, enforcing \cref{inv:link-eta-right}.
+  Let $\lambda x.t$ be the \rhs of a \linketa, then $\wellb(t)$, enforcing \cref{inv:link-eta-right}.
   \label{lemma:unif-eta-aux}
@@
 \begin{proof}[Proof sketch]
-  By construction, every ``problematic'' term in \Fo{} is replaced with a variable
-  in the corresponding \Ho term. Therefore, $\wellb(t)$.
+  By \cref{lem:w-enforcement-maybeeta}.
 \end{proof}
@@
 \begin{proof}[Proof sketch]
-  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
+  Let $\sigma$ be a substitution such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
   By \cref{inv:link-eta-right}, we have $\wellb(t)$.
@@
   \linketaM{\Gamma_2}{Y}{\lambda x.t'} 
-  the unification is performed between $t$ and $t'$. Both terms are
-  \wellb and therefore, unification is done between \wellb terms.
+  the unification is performed between $t$ and $t'$.
+  By \cref{lem:w-enforcement-maybeeta} both terms are
+  \wellb.
 \end{proof}
@@
 \begin{proof}[Proof sketch]
-  Rules \cref{def:progressetaleft} and
-  \cref{def:progressetadedup} remove one link from \linkStore, hence they
+  Rules \ref{def:progressetaleft} and
+  \ref{def:progressetadedup} remove one link from \linkStore, hence they
   cannot be applied indefinitely.
@@
   % F x = z, F x y = z y
-  Given a list of unification problems \foUnifPb, such that
+  Given a list of unification problems \foUnifPb such that
   $\wellb_{\beta \eta}(\foUnifPb)$, if the 
-  memory map is bijective, then
+  memory map is bijective then
   the introduction 
@@
   Without loss of generality we consider a
-  \Fo unification problem of the form $s_1 \Uo s_2$
+  \Fo{} unification problem of the form $s_1 \Uo s_2$
   where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
@@
     \item $s_2 \notin \maybeeta$,
-          in this case $s_2$ is compile to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
+          in this case $s_2$ is compiled to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
           and triggers \progressetaleft.
@@
           progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
-          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \ref{clause:lam-lam}
+          $\lambda x.t_2\appsep x$, otherwise they unify thanks to rule \ref{clause:lam-lam}
           and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
           to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
-          since \ref{clause:lam-lam} or \ref{clause:eta1} try to unify under it, eventually finding two
+          since rule \ref{clause:lam-lam} or rule \ref{clause:eta1} try to unify under it, eventually finding two
           different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
@@
           with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
-          $X$ and $Y$, \progressetadedup is fired and, once the $\lambda$ are crossed,
+          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
           $t_1$ and $t_2$ are unified. 
@@
 resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
-By the definition of \linketa decompilation,
-decompilation generates $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$, since:
-$X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$.
+Since $X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$,
+by the definition of \linketa decompilation
+$\sigma$ is decompiled to $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$,
+
 % \todo{explain why decompile does so and which eta?}
@@
 with different arities.
-Then it is the case the
+When it is the case the
 memory map \mapStore generated by the compiler is not
@@
 However, there is no connection between these two variables
-and any incompatible assignements to them would not be detected, breaking,
-in turn \cref{prop:fidelity}. 
+and any incompatible assignements to them would not be detected,
+in turn breaking \cref{prop:fidelity}. 
 % This incompatibility will only be detected during the
-decompilation phase.
+% decompilation phase.
 To address this issue, we post-process the compiler's output
@@
 \begin{proposition}[\mapStore is a bijection]
-  After compilation, for each \Fo{} variable $X$ in \foUnifPb{} and for each \Ho{}
-  variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
-  (for any arity $n$).
+  After compilation, for each \Fo{}-variable $X$ in \foUnifPb{} and for each
+  \Ho{}-variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
+  (for an arity $n$).
 \end{proposition}
@@
 The intuition is that we $\eta$-expand the occurrence of the variable
-with higher arity to match the lower arity. Since each \linketa can
+with lower arity to match the higher arity. Since each \linketa can
 add exactly one lambda, we need as many links as the difference between the
@@
   so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
-  Note that in this case fidelity is guaranteed, since term unified to $C^n$
+  Note that in this case fidelity is guaranteed, since the term unified with $C^n$
   sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
-  links is shrinked by $q$. Note that if this length is bigger then $n-m$ again
-  the full chain collapse and $A^m$ and $C^n$ are agian tiny coupled. Otherwise,
+  links is shrinked by $q$. Note that if this length is bigger then $n-m$
+  the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
+  Otherwise,
   if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
@@
 \end{proof}
+\todo{this proof is still unreadable}
 % \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}
@@
 $X_1$ and $X_2$ are unified with different terms.
+\todo{what are these X1 and X2?}
 
@@
 \noindent
-Note that $\mapping{X}{B}{1}$ desappears from \mapStore
+Note that $\mapping{X}{C}{0}$ desappears from \mapStore
+\todo{update this description to match the example} 
 in favour of the auxiliary \linketa \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
@@
 
-Even if we do not requrie \Ue to handle problems outside of \llambda it is
-worth handling terms in \notllambda since, in practice, these terms often
-re-enter \llambda at runtime, as observed in~\cite{Nadathur2001} for example,
-or to accommodate for the introduction of heuristics.
-
-In the following example problem $\foUnifPb_2$ is $X \appsep a \Uo a$  that
-admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
-and $\rho_2 = \{X \mapsto \lambda x.a\}$. The unification algorithm alone
-cannot chose, since no solution is more general than the other,
-although the first problem solve this dilemma at run-time making the second
-problem well behaved:
+As observed in~\cite{Nadathur2001} it is worth 
+handling terms in \notllambda since, in practice, these terms often
+re-enter \llambda at runtime.
+%for example,
+%or to accommodate for the introduction of heuristics.
+
+In the following example problem $\foUnifPb_2$, namely $X \appsep a \Uo a$,
+admits two different solutions: $\rho_1 = \{~X \mapsto \lambda x.x~\}$
+and $\rho_2 = \{~X \mapsto \lambda x.a~\}$.
+The unification algorithm alone
+cannot chose the right one, since no solution is more general than the other,
+but for the first problem the choice is obvious, and in turn
+this choice makes $\foUnifPb_2 \subseteq \llambda$:
 %
@@
 \noindent
-We want to support this scenario but we have to improve a little our compiler
+In order to support this scenario we have to improve a little our compiler
 and \progress routines.
@@
 \begin{invariant}[\linkbeta \rhs]
-  The \rhs of any \linkbeta has the shape $X\appsep t_1\dots t_m$
+  The \rhs of any \linkbeta has the shape $X_{s_1\ldots s_n}\appsep t_1\dots t_m$
   where $X$ is a unification variable in \llambda (with scope $s_1\dots s_n$)
@@
 \noindent Note that the shape of such as \rhs is
-  <<\elpiIn{app [uva X S | L]}>>, where \elpiIn{S}~$= s_1\dots s_n$
-  and \elpiIn{L}~$=t_1\dots t_m$.
+  <<\elpiIn{app [uva X S | L]}>>, where
+  \elpiIn{S} is \elpiIn{[s~$_1$~, ~$\ldots$~, s~$_n$~]}
+  and \elpiIn{L} is  \elpiIn{[t~$_1$~, ~$\ldots$~, t~$_m$~]}.
 
@@
 
-By inserting this rule
-before rule~\ref{rule:compapp}.
-we identify terms \notllambda, i.e. flexible applications with arguments that
-are not distinct names:
+We insert this rule
+just before rule~\ref{rule:compapp}:
 
@@
 \noindent
-The list \elpiIn{Ag} is split into the two parts: 
+The list \elpiIn{Ag} is split into two parts: 
 \elpiIn{Pf} is in \llambda, and can be empty; 
 \elpiIn{Extra} cannot be empty and is such that
-\elpiIn{append Pf Extra Ag}.
+<<\elpiIn{append Pf Extra Ag}>>.
 The \rhs of the \linkbeta is
@@
 in scope all names in \elpiIn{Pf1} (the compilation of \elpiIn{Pf}).
-The variable \elpiIn{B}, returned has the compiled term, is a fresh variable
+The variable \elpiIn{B}, returned as the compiled term, is a fresh variable
 having in scope all the free variables occurring in \elpiIn{Pf1} and
@@
   Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
-  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
+  not occurring in $s_1\dots s_n$. If $m = 1$, then $l$ is removed and
+  the \lhs is
   unified with $X_{s_1\ldots s_n~s}$.
@@
 \begin{definition}[\progBetaFail]
-  \progress fails whenever
-  there exists another link $l' \in \linkStore$ with the same \lhs % as $l$;
-  or the \lhs % of $l$
+  \progress fails whenever:
+  \begin{enumerate}
+  \item
+  there exists another link $l' \in \linkStore$ with the same \lhs as $l$; or
+  \item the \lhs of $l$
   become rigid.
+  \end{enumerate}
   \label{def:progBetaFail}
 \end{definition}
-
+\noindent
 We relax this condition and accommodate for
@@
 
-Finally, in order to correctly correlate a variable appearing in the \lhs of a
-\linketa and in the \rhs of a \linkbeta, we should introduce the following
-\linketa progression rule.
+Finally
+%in order to correctly correlate a variable appearing in the \lhs of a
+%\linketa and in the \rhs of a \linkbeta, 
+we introduce the following
+\linketa progress rule.
 
@@
 \end{definition}
-
-The idea behind this is to instantiate the variable of a \linketa whenever its
-\rhs is for sure a \wellb term.
+\todo{reformulate, maybe with itemize, too messy}
+
+The idea behind this rule is to instantiate the variable of a \linketa
+whenever its \rhs is for sure a \wellb term.
 
@@
   %\elpiIn{progress}
-  it immediately fails and \progressetaright terminates since it executes 
-  terminating instructions and if its permises succeed, ther considered
-  \linketa is removed from \linkStore.
+  it immediately fails and \progressetaright terminates since it performs 
+  terminating operations and, if its permises succeed, the considered
+  \linketa is removed from \linkStore.\todo{unclear}
 %
@@
   The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
-  if \Uo{} falls outside \llambda.
+  if $\foUnifPb \not\subseteq \llambda$.
 \end{theorem}
@@
 Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification
-problems until the substitution makes them reenter \llambda. Other systems, for
-example of the unification algorithm of Coq used in its type-class
-solver~\cite{sozeau08}, apply heuristics like preferring projection over mimic,
-and commit to that solution.
+problems until the substitution makes them reenter \llambda.
+Other systems apply heuristics like preferring projection over mimic,
+and commit to that solution. This is the case for
+the unification algorithm Coq uses in its type-class
+solver~\cite{sozeau08},
 
@@
 \noindent
-If we want the object language unification to delay the first unification problem
+If we want the object-language unification to delay the first unification problem
 (waiting for $X$ to be be instantiated),

### 2024-05-03 paper/main.tex
solve todo in simulation-fidelity of mapdeduplication

@@
 %\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
-\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{O}\xspace}}{O}} % space non va
+\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{O}}\xspace}{O}}
 \newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{M}}\xspace}{M}}
@@
   $\wellb_{\beta \eta}(\foUnifPb)$,
-  if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
+  if $\ \foUnifPb$ contains the same \Fo{} variable used at different arities,
   then \mapdeduplication guarantees \fullref{prop:fidelity}
@@
 
+\def\chain{\ensuremath{\mathcal{C}}\xspace}
+
 \begin{proof}[Proof sketch]
-  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
-  variable used at different arities are compiled to two
-  variables $A^m$ and $C^n$ related by a chain of \linketa.
+  Let $X$ a \Fo variable appearing in two different subterms $s_1$ and $s_2$. in
+  $s_1$, $X$ is applied to the list of terms $x_1\dots x_m$ and, in $s_2$, it is
+  applied to $y_1\dots y_n$. Here, $m < n$. By \mapdeduplication, the
+  compilation genereates two distinct \Ho variables for $X$, say $A$ and $C$,
+  related by a chain \chain of \linketa{}s, with length $n - m$. We prove that
+  if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
+  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
+  these terms fails to unify. By \progressetaleft, the assignment of $A$
+  activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
+  chain collapses and unification is performed between $\lambda x_n\dots
+  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
+  we work with \wellb terms, this unification succeeds iff it does in \Fo. If,
+  otherwise, $p < n$, then two cases should be considered: $t_1$ is a rigid
+  term, then, $t_2$ must be an eta expansion of $t_1$, implying $q = 0$. If it
+  is not the case, then the successive progressions made by \progressetaleft
+  would eventually cause a failure. If, on the other hand, $t_1$ is a
+  unification variable, \elpiIn{scope-check} guarantees that $t_2$ does not use
+  $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
+  After this check, $m - n - p$ links of \chain remain suspended, as we expect:
+  decompilation is charged wake all links in \linkStore and eventually relate
+  $t_1$ with $t_2$.
+  %
   % Attenzione a questo caso:
@@
   % 
-  Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
-  $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
-  term, then the chain of links between $A^m$ and $C^n$ is completely triggered
-  so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
-  Note that in this case fidelity is guaranteed, since the term unified with $C^n$
-  sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
-  links is shrinked by $q$. Note that if this length is bigger then $n-m$
-  the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
-  Otherwise,
-  if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
-  with at least $q$ lambdas. In this second case fidelity is also guaranteed,
-  due to the constraint of the number of lambda by which $C^n$ should start.
+  % Nella nuova versione sottintendo questo caso: p = 0 e t_1 è flessibile
+  %
+  % Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
+  % $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
+  % term, then the chain of links between $A^m$ and $C^n$ is completely triggered
+  % so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
+  % Note that in this case fidelity is guaranteed, since the term unified with $C^n$
+  % sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
+  % links is shrinked by $q$. Note that if this length is bigger then $n-m$
+  % the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
+  % Otherwise,
+  % if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
+  % with at least $q$ lambdas. In this second case fidelity is also guaranteed,
+  % due to the constraint of the number of lambda by which $C^n$ should start.
   %
@@
 \end{proof}
-\todo{this proof is still unreadable}
+
+% \todo{this proof is still unreadable}
 % \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}
 
-\noindent
-Note that shortening the chain of links allows unification to fail if
-$X_1$ and $X_2$ are unified with different terms.
-\todo{what are these X1 and X2?}
+% Note that shortening the chain of links allows unification to fail if
+% $X_1$ and $X_2$ are unified with different terms.
+% \todo{what are these X1 and X2?}
 

### 2024-05-03 paper/main.tex
corrected example of mapdedup

@@
 \paragraph{Example of \mapdeduplication}
-If we take back the example give at the beginning of this section,
+If we take back the example given at the beginning of this section,
 \mapdeduplication produces the following result:
@@
 Note that $\mapping{X}{C}{0}$ desappears from \mapStore
-\todo{update this description to match the example} 
-in favour of the auxiliary \linketa \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
+in favour of the auxiliary \linketa \linketaM{}{C}{\lambda x.B_{x}}.
 The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up

### 2024-05-03 paper/main.tex
reformulate definition progress-eta-rhs

@@
 \begin{definition}[\progressetaright]\label{def:progressetaright} A link
-\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
-$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
-$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
-first case, $X$ is unified with $T$, and in the second one, $X$ is unified with
-$T'$ (under the context $\Gamma$).
+\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 
+\begin{itemize}
+  \item $\testmaybeeta~T$ does not hold (anymore). In this case $X$ is unified
+        with $T$
+  \item $T$ $\eta$-contracts to a term $T'$ not starting with the \elpiIn{lam}
+        constructor. In this case $X$ is unified with $T'$
+\end{itemize}
 \end{definition}
-\todo{reformulate, maybe with itemize, too messy}
-
-The idea behind this rule is to instantiate the variable of a \linketa
+
+The idea behind this rule is to instantiate the \lhs variable of a \linketa
 whenever its \rhs is for sure a \wellb term.

### 2024-05-03 paper/main.tex
rephrase progress beta (+prog-eta-rhs) termination

@@
 \end{definition}
-\todo{comment this rule}
 \end{comment}
@@
 \begin{proof}[Proof sketch]
-  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
-  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
-  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
-  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
-  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
-  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
-  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
-  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
-  terminates. \progBetaFail guarantees termination since 
-  %it makes 
-  %\elpiIn{progress}
-  it immediately fails and \progressetaright terminates since it performs 
+  Let $l = \linkbetaM{\Gamma}{T}{\rhsBeta}$ a \linkbeta in the store \linkStore.
+  The progression made by \progBetaRH{} terminates: $l$ is removed from
+  \linkStore after having performed terminating instructions. If $l$ is not
+  activated by \progBetaRH, then $X$ is a variables. The activation
+  of \progBetaLL replaces $l$ with the new \linkbeta 
+  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$.
+  This progression can be triggered at most $m$ times, since at each time
+  the number of terms applied to $X$ decreases. In the end $l_m$
+  will be removed from \linkStore after the unification between \lhs with \rhs.
+  We also note that each \progBetaLL generates a \linketa, yet, according to
+  \cref{lemma:prog-eta-terminates}, this does not affect termination. 
+  \progBetaFail terminates since it stop progression with failure.
+  Finally, \progressetaright performs 
   terminating operations and, if its permises succeed, the considered
-  \linketa is removed from \linkStore.\todo{unclear}
+  \linketa is removed from \linkStore, ensuring termination.
   %

### 2024-05-03 paper/main.tex
minor

@@
 % decompilation phase.
-To address this issue, we post-process the compiler's output
+To address this issue, we post-process \mapStore
 to ensure the following property:
@@
   \Ho{}-variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
-  (for an arity $n$).
+  in \mapStore.
 \end{proposition}
@@
 \noindent
-Note that the execution in \Ho{} may require allocating new variables as
-explained in~\cref{sec:decompilation}. The property above does not
-concern these variables since they are fresh and have no corresponding
+Note that running \Ue{} may require allocating new variables in \Ho{} as
+explained in~\cref{sec:decompilation}. The property above ignores these variables since they are fresh and have no corresponding
 variable in \Fo.
@@
   $\wellb_{\beta \eta}(\foUnifPb)$,
-  if $\ \foUnifPb$ contains the same \Fo{} variable used at different arities,
+  if $\ \foUnifPb$ contains the same \Fo{}-variable used at different arities,
   then \mapdeduplication guarantees \fullref{prop:fidelity}
@@
 
-\def\chain{\ensuremath{\mathcal{C}}\xspace}
+\def\chain{\ensuremath{\vec{c}}\xspace}
 
 \begin{proof}[Proof sketch]
-  Let $X$ a \Fo variable appearing in two different subterms $s_1$ and $s_2$. in
-  $s_1$, $X$ is applied to the list of terms $x_1\dots x_m$ and, in $s_2$, it is
-  applied to $y_1\dots y_n$. Here, $m < n$. By \mapdeduplication, the
-  compilation genereates two distinct \Ho variables for $X$, say $A$ and $C$,
-  related by a chain \chain of \linketa{}s, with length $n - m$. We prove that
+  Let $X$ be a \Fo{}-variable appearing in two different subterms $s_1$ and $s_2$.
+  In $s_1$ variable $X$ is applied to the list of terms $x_1\dots x_m$ while, in $s_2$,
+  it is applied to $y_1\dots y_n$. Without loss of generality we take $m < n$.
+  After \mapdeduplication we have two distinct \Ho{}-variables for $X$, say $A$ and $C$,
+  related by a chain \chain of \linketa{}s of length $n - m$. We prove that
   if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
   x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
-  these terms fails to unify. By \progressetaleft, the assignment of $A$
+  these terms fails to unify.\todo{a unification fails if they fail to unify is a tautology, to be clarified} By \progressetaleft, the assignment of $A$
   activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
@@
   x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
-  we work with \wellb terms, this unification succeeds iff it does in \Fo. If,
-  otherwise, $p < n$, then two cases should be considered: $t_1$ is a rigid
-  term, then, $t_2$ must be an eta expansion of $t_1$, implying $q = 0$. If it
-  is not the case, then the successive progressions made by \progressetaleft
-  would eventually cause a failure. If, on the other hand, $t_1$ is a
-  unification variable, \elpiIn{scope-check} guarantees that $t_2$ does not use
+  we work with \wellb terms, this unification succeeds iff it does in \Fo.
+  Otherwise if $p < n$ then we have to consider two cases. 1) If $t_1$ is a rigid
+  term then $t_2$ must be the $\eta$-expansion of $t_1$, implying $q = 0$. If it
+  is not the case, then \progressetaleft
+  would eventually cause a failure. 2) If $t_1$ is a
+  unification variable then \elpiIn{scope-check} guarantees that $t_2$ does not use
   $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
   After this check, $m - n - p$ links of \chain remain suspended, as we expect:
-  decompilation is charged wake all links in \linkStore and eventually relate
+  decompilation is charged to wake all links in \linkStore and eventually relate
   $t_1$ with $t_2$.
@@
 The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
-turn triggers $\linkStore_1$ by \progressetaleft. This link perfroms
-$\lambda x.a \Ue \lambda x.(a \appsep x)$, which, as expected, fails.
+turn triggers $\linkStore_1$ by \progressetaleft. In turn, this performs
+$\lambda x.a \Ue \lambda x.(a \appsep x)$ that fails as expected.
 
@@
 and \progress routines.
-In particular $\hoUnifPb_1$ generates $\sigma = \{ A
-\mapsto \lambda x. a\}$ making $\sigma\hoUnifPb_2$ equal to
+In particular $\hoUnifPb_1$ generates $\sigma = \{~ A
+\mapsto \lambda x. a ~\}$ making $\sigma\hoUnifPb_2$ equal to
 $(\lambda x. a) \appsep a \Ue a$ that \Ue{} cannot solve since it
@@
 
-To address this problem the compiler must recognize and replace
-\notllambda terms with fresh variables and generate a new kind of links
+To address this problem the compiler must recognize
+\notllambda terms
+and replace them
+with fresh variables and generate a new kind of links
 that we call \linkbeta.
@@
 \begin{definition}[\progBetaFail]
-  \progress fails whenever:
-  \begin{enumerate}
+  \progress fails when either
+  \begin{itemize}
   \item
@@
   become rigid.
-  \end{enumerate}
+  \end{itemize}
   \label{def:progBetaFail}
@@
 \begin{itemize}
-  \item $\testmaybeeta~T$ does not hold (anymore). In this case $X$ is unified
-        with $T$
+  \item $\testmaybeeta~T$ does not hold (anymore), in this case $X$ is unified
+        with $T$; or
   \item $T$ $\eta$-contracts to a term $T'$ not starting with the \elpiIn{lam}
@@
 \begin{proof}[Proof sketch]
-  Let $\foUnifPb_i$ be the first problem in \foUnifPb, such that it is not in \llambda and le
+  Let $\foUnifPb_i$ be the first problem in \foUnifPb{} such that it is not in \llambda and let
   $\sigma$ be the substitution obtained solving
@@
   each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 
-  unification problem, is instantiated if it is known for sure
-  that $X$ is no more a \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
+  unification problem, is instantiated iff it is known for sure
+  that $X$ is no more in \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
   % each variable that the

### 2024-05-03 paper/main.tex
minor
small update of proof: Fidelity with map-deduplication

@@
   if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
-  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
-  these terms fails to unify.\todo{a unification fails if they fail to unify is a tautology, to be clarified} By \progressetaleft, the assignment of $A$
+  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a failure occurs iff
+  these terms do not unify. By \progressetaleft, the assignment of $A$
   activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
@@
 between the verified and the actual code, which we found to
-hinder our exploration. We did, however, establish a
+hinder our exploration. However, we did establish a
 test suite with about a hundred cases.

### 2024-05-09 paper/main.tex
add node nb

@@
 overhead on small goals but is consistently faster than
-the Coq one starting from goals made of XXX nodes.
-\todo{D: how many?}
+the Coq one starting from goals made of $32$ nodes.
 Moreover our approach is flexible enough to accommodate different strategies

### 2024-05-13 paper/main.tex
[fh]variable renamed to [om]variable

@@
 \noindent
-Each \elpiIn{hvariable} is stored in the mapping together with
+Each \elpiIn{mvariable} is stored in the mapping together with
 its arity (a number) so that the code of \ref{clause:malloc} below can preserve:
@@
 \noindent
-When a single \elpiIn{fvariable} occurs multiple times with different numbers
+When a single \elpiIn{ovariable} occurs multiple times with different numbers
 of arguments, the compiler generates multiple mappings for it, on a first

### 2024-05-13 paper/appendix.tex
rename object constructors

@@
   type tm->fm map -> ho.tm -> fo.fm -> o.
-  tm->fm _ (ho.con C)  (fo.fcon C).
-  tm->fm L (ho.lam B1) (fo.flam B2) :- 
+  tm->fm _ (ho.con C)  (fo.o-con C).
+  tm->fm L (ho.lam B1) (fo.o-lam B2) :- 
     pi x y\ tm->fm _ x y => tm->fm L (B1 x) (B2 y).
@@
   tm->fm L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
-    map (tm->fm L) TL T1, fo.mk-app (fo.fuva VO) T1 T.
+    map (tm->fm L) TL T1, fo.mk-app (fo.o-uva VO) T1 T.
 

### 2024-05-13 paper/main.tex
rename object constructors

@@
 Unification variables
-in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
-the ``arguments'' of a higher-order variable are given via the \elpiIn{fapp}
+in \Fo{} (\elpiIn{o-uva} term constructor) have no explicit scope:
+the ``arguments'' of a higher-order variable are given via the \elpiIn{o-app}
 constructor. For example the term <<\coqIn{P x}>> is represented as
-<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
+<<\elpiIn{o-app [o-uva N, x]}>>, where \elpiIn{N} is the memory address
 of \elpiIn{P} and \elpiIn{x} is a bound variable.
@@
 it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
-weak-head normal form of <<\elpiIn{fapp [F|L]}>>.
+weak-head normal form of <<\elpiIn{o-app [F|L]}>>.
 
@@
 % has no corresponding operation in \Ho, and only ensures that
-% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
-% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
+% terms of the form <<\elpiIn{o-app[o-app L1|L2]}>> are replaced by
+% <<\elpiIn{o-app L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
 % and \elpiIn{L2}. The reasons for this asymmetry is
-% that an \elpiIn{fapp} node with a flexible head is always mapped
+% that an \elpiIn{o-app} node with a flexible head is always mapped
 % to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
@@
 node carries a term that is easy to decompile since \mapStore is a bijection.
-Since the \elpiIn{flam} node
+Since the \elpiIn{o-lam} node
 carries no additional information (other than the function body),
-each \elpiIn{abs} node can be decompiled to a \elpiIn{flam} one.
-
-However, if the \elpiIn{flam} node was carrying more information,
+each \elpiIn{abs} node can be decompiled to a \elpiIn{o-lam} one.
+
+However, if the \elpiIn{o-lam} node was carrying more information,
 as is the case for Coq where it holds the type of the bound variable,
@@
 If we look at the syntax of \Fo{}, the only interesting case is
-<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
+<<\elpiIn{o-uva X ~\Uo~s}>>. In this case, after compilation, we have
 <<\elpiIn{uva Y [] ~\Ue~t}>>
@@
   If we look at the \Fo{} terms, there is one more interesting case,
-  namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
+  namely \elpiIn{o-app[o-uva X|W] ~\Uo~s} when \elpiIn{W}
   are distinct names compiled to $\vec{w}$. In this case
@@
 %   problematic terms are mapped to uva by comp,
-%   the problematic fapp node is gone.
+%   the problematic o-app node is gone.
 %   if F x = t then  F = lam x.t[]
@@
 % \begin{elpicode}
-% Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
-%       , flam x\ fapp[fcon"g", fcon"a"] ]
+% Terms [ o-lam x\ o-app[o-con"g",o-app[o-uva z, x]]
+%       , o-lam x\ o-app[o-con"g", o-con"a"] ]
 % Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
@@
 % HS = [some (abs x\con"a")]
-% S = [some (flam x\fcon a)]
+% S = [some (o-lam x\o-con a)]
 % \end{elpicode}
@@
 % \begin{elpicode}
-%   Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
-%   , flam x\ fapp[fcon"g", fcon"a"] ]
+%   Terms [ o-lam x\ o-app[o-con"g",o-app[o-uva z, x]]
+%   , o-lam x\ o-app[o-con"g", o-con"a"] ]
 % Problems = [ pr 0 1   % $A = \lambda x.x$
@@
 % HS = [some (abs x\con"a")]
-% S = [some (flam x\fcon a)]
+% S = [some (o-lam x\o-con a)]
 % lam x\ app[f, app[X, x]] = Y,
@@
 \noindent
-Whenever \elpiIn{flam F} is detected to be in
+Whenever \elpiIn{o-lam F} is detected to be in
 \maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
@@
   % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
-  % and the one flam-flam we find different terms.
+  % and the one o-lam-o-lam we find different terms.
   % (5.9.1) only adds lambdas that are forcibly consumed by
@@
 %   Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
-%   \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
+%   \Fo{} term has the shape \elpiIn{o-app[o-uva M | Ag]} where \elpiIn{Ag} is a
 %   list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
@@
 %   By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
-%   the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
+%   the list \elpiIn{Ag} from the original term \elpiIn{o-app [o-uva A|Ag]}.
 %   $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is

### 2024-05-13 paper/appendix.tex
fm -> to, fapp -> o-app, flam -> o-app

--- 2024-04-11 paper/appendix.tex
@@
     ho.deref-assmt H T TTT,
-    abs->lam TTT T', tm->fm Map T' T1, 
+    abs->lam TTT T', tm->to Map T' T1, 
     fo.eta-contract T1 T2, mem.assign VO F T2 F1,
@@
 
-  type tm->fm map -> ho.tm -> fo.fm -> o.
-  tm->fm _ (ho.con C)  (fo.fcon C).
-  tm->fm L (ho.lam B1) (fo.flam B2) :- 
-    pi x y\ tm->fm _ x y => tm->fm L (B1 x) (B2 y).
-  tm->fm L (ho.app L1) T :- map (tm->fm L) L1 [Hd|Tl], 
+  type tm->to map -> ho.tm -> fo.to -> o.
+  tm->to _ (ho.con C)  (fo.o-con C).
+  tm->to L (ho.lam B1) (fo.o-lam B2) :- 
+    pi x y\ tm->to _ x y => tm->to L (B1 x) (B2 y).
+  tm->to L (ho.app L1) T :- map (tm->to L) L1 [Hd|Tl], 
     fo.mk-app Hd Tl T.
-  tm->fm L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
-    map (tm->fm L) TL T1, fo.mk-app (fo.fuva VO) T1 T.
+  tm->to L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
+    map (tm->to L) TL T1, fo.mk-app (fo.o-uva VO) T1 T.
 

### 2024-05-13 paper/appendix.tex
fv -> ov & hv -> mv

@@
   decompl-subst _ [] _ F F.
-  decompl-subst Map [mapping (fv VO) (hv VM _)|Tl] H F F2 :- 
+  decompl-subst Map [mapping (ov VO) (mv VM _)|Tl] H F F2 :- 
     mem.set? VM H T, !, 
@@
     decompl-subst Map Tl H F1 F2. 
-  decompl-subst Map [mapping _ (hv VM _)|Tl] H F F2 :- 
+  decompl-subst Map [mapping _ (mv VM _)|Tl] H F F2 :- 
     mem.unset? VM H, decompl-subst Map Tl H F F2.
@@
     fo.mk-app Hd Tl T.
-  tm->to L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
+  tm->to L (ho.uva VM TL) T :- mem L (mapping (ov VO) (mv VM _)), 
     map (tm->to L) TL T1, fo.mk-app (fo.o-uva VO) T1 T.
@@
   add-new-map _ (ho.uva N _) Map [] F1 F1 :- 
-    mem Map (mapping _ (hv N _)), !.
+    mem Map (mapping _ (mv N _)), !.
   add-new-map H (ho.uva N L) Map [Map1 | MapL] F1 F3 :-
     mem.new F1 M F2,
-    len L Arity, Map1 = mapping (fv M) (hv N (arity Arity)),
+    len L Arity, Map1 = mapping (ov M) (mv N (arity Arity)),
     add-new-map H (ho.app L) [Map1 | Map] MapL F2 F3.

### 2024-05-13 paper/main.tex
fm -> to, fapp -> o-app, flam -> o-app
typos

@@
 an Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
-terms, called \elpiIn{tm}, that includes (among others) the following
+Coq terms, called \elpiIn{tm}, that includes (among others) the following
 constructors:
@@
 \begin{proposition}[Simulation fidelity]\label{prop:fidelity}
-In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
+In the context of  \frun and \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
 $\forall p \in 1 \ldots n,$
@@
 \begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
-In the context of \hrun, if 
+In the context of \frun and \hrun, if 
 $\rho_{p-1} \foUnifPb_{p} \subseteq \llambda$ 
@@
 \begin{definition}[\maybeeta]\label{def:maybeeta}
-  \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho \lambda x.s$
+  \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho (\lambda x.s)$
   is an eta expansion. % without the lambda, 5.11 is not true, that
@@
 $\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
-makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
+makes $\rho s \Eo{} \lambda x.\lambda y.f \appsep x \appsep y$,
 which is the eta-long form of $f$. This term is problematic since
@@
 \noindent
-A solution to a \Fo{} variable is a plain term, that is, \elpiIn{fsubst}
-is an abbreviation for \elpiIn{mem fm}.
+A solution to a \Fo{} variable is, instead, a plain term, that is, \elpiIn{fsubst}
+is an abbreviation for \elpiIn{mem to}.
 %We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
@@
 % \input{code/fderef}
-It is worth examining the code of \elpiIn{deref}, which
+It is worth examining the code of \elpiIn{deref}, that
 applies the substitution to a \Ho{} term. Notice how assignments are moved
@@
 
-The simple compiler described in this section serves as a basis for the
+The simple compiler described in this section serves as a base for the
 extensions in \cref{sec:llam,sec:eta,sec:beta}.
@@
 In order to bring back the substitution from \Ho{} to \Fo{} the compiler
-builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
-using %the routine
-\ref{clause:malloc}.
+builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones.
+% using the routine \ref{clause:malloc}.
 
@@
 but plays a major role in \cref{sec:llam,sec:eta,sec:beta}.
+
+\input{code/comp_base}
+\input{code/compile.tex}
+
+\noindent
 With respect to \cref{sec:problem-statement}, the signature also allows
 for updates to the substitution.
-
-\input{code/comp_base}
-\input{code/compile.tex}
-
-\noindent
 The code above uses that possibility
@@
 
-A step in \Ho consists of unifying two terms and reconsidering all
+A step in \Ho consists in unifying two terms and reconsidering all
 links for progress. If either of these tasks fails, we consider the entire step to
@@
 Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
-and explain why the don't hinder termination.
+and explain why the do not hinder termination.
 % For brevity we omit the code
@@
 \linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
-The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
+The term $t' = \lambda x.Y_{{x_1\ldots x_n}~x}$ is obtained after
 executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$ using a fresh $Y$.
@@
 This operation is possible thanks to \fullref{inv:linklhs},
-which ensures that no link causes an occur-check~(\ref{sec:oc})
+and that no link causes an occur-check~(\ref{sec:oc})
 and the fact that \linkStore is duplicate-free (see \cref{sec:invariant1}).
@@
 already happens to be in \wellb.
-Later, when the compiler will ensure \fullref{prop:w-enforcing}
-and
+Later, when the compiler will ensure \fullref{prop:w-enforcing},
 the proof will be refined to cover for the new cases.
@@
   {{{t_1,t_2},{t_3,t_4}}}
-  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
+  {{{X_1,A_1,{{n}}},{X_2,A_2,{{m}}}}}
   {{{\eta,\Gamma,a,b}}}
@@
    is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
-   Thanks to \ref{clause:beta1}
+   Thanks to \ref{clause:beta1} we have
    $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
@@
 \noindent
-In other word the compiler eliminates all subterms that are in \maybebeta.
+In other words the compiler eliminates all subterms that are in \maybebeta.
 
@@
 \end{theorem}
-\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
+\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta},
   \Ue is as powerful as \Uo in $\wellb_\beta$, as well as in \wellb by
@@
 \noindent
-The reason is that \elpiIn{lam x\ uva A [x]} and
-\elpiIn{con"f"} start with different, rigid, term constructors.
+The reason is that <<\elpiIn{lam x\ uva A [x]}>> and
+<<\elpiIn{con"f"}>> start with different term constructors.
 
@@
   has the shape $\lambda x.t$
-  and $t$ is \wellb. 
+  and $t$ is in \wellb. 
   %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
@@
 \begin{proof}[Proof sketch]
-Since our terms are in $\beta$-normal form there is only
-one rule that can play a role (namely \ref{clause:eta1}),
-hence if the term $s$ is not exactly $x$ (case 1)
+Since our terms are in $\beta$-normal form the only
+rule that can play a role is \ref{clause:eta1}, and note that
+that rule uses \elpiIn{beta} to add an argument to an
+application or to the scope of a variable.
+If the term $s$ is not exactly $x$ (case 1)
 it can only be an $\eta$-expansion of $x$, or a unification
@@
 of $X$ can make $x$ disappears from $t$.
-Moreover, note that $\eta$-contracting $t$ cannot make $x$ disappear, since
+Note that $\eta$-contracting $t$ cannot make $x$ disappear, since
 $x$ is not a locally bound variable inside $t$.
@@
   \begin{enumerate}
-    \item $t$ is a constant or a name applied to the arguments
+    \item $t$ is a constant $c$ or a name $y$ applied to the arguments
       $l_1 \ldots l_m$ such that 
@@
       \reduceto{} $x_i$, and
-      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n}$;
+      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n} ~ y$;
     \item $t$ is a
@@
 \linketa{}s are meant to delay the unification of ``problematic'' terms until
-we know for sure if their head lambdas can be $\eta$-contracted or not.
+we know for sure if their head lambdas can be $\eta$-contracted.
 
@@
 A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
-$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
+$X$ becomes rigid. Let $y\in\Gamma$. There are two cases:
 \begin{enumerate}
@@
   Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaleft is between
-  terms in $\wellb$  
+  terms in \wellb.
   \label{lemma:unif-eta-1}
@@
   If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
-  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
+  Otherwise, $\lambda x. X\appsep x$ is unified with $\lambda x. t$, and $X\appsep x$ and $t$
   are both in \wellb.
@@
   the unification is performed between $t$ and $t'$.
-  By \cref{lem:w-enforcement-maybeeta} both terms are
+  By \cref{lem:w-enforcement-maybeeta} both terms are in
   \wellb.
@@
   By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
-  unification performed by the activation of a \linketa is done between
+  unification performed by the activation of a \linketa is between
   terms in \wellb.
@@
   of \linketa guarantees \fullref{prop:fidelity}.
-  \footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
+  %\footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
   \label{lemma:fidelity-maybeeta}
@@
   \begin{itemize}
-    \item $s_2 \notin \maybeeta$,
-          in this case $s_2$ is compiled to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
-          and triggers \progressetaleft.
-          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
-          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
-          $\lambda x.t_2\appsep x$, otherwise they unify thanks to rule \ref{clause:lam-lam}
-          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
-          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
-          since rule \ref{clause:lam-lam} or rule \ref{clause:eta1} try to unify under it, eventually finding two
-          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
-          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
-          $\eta$-expands $t_2$.
-
     \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
@@
           This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
+
+    \item $s_2 \notin \maybeeta$,
+          in this case $s_2$ is compiled to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
+          and triggers \progressetaleft.
+          If $s_1$ and $s_2$ are equal thanks to \ref{clause:eta1} then
+          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
+          $\lambda x.t_2\appsep x$, otherwise they are equal thanks to rule \ref{clause:lam-lam}
+          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.
+          Alternatively, if $s_1$ are different $s_2$
+          it cannot be because of the $\lambda$ constructor in the head of $t_1$,
+          since rule \ref{clause:lam-lam} or rule \ref{clause:eta1} move under it, eventually finding two
+          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
+          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
+          $\eta$-expands $t_2$.
+
   \end{itemize}
@@
 \begin{corollary}[Properties of \Uo in $\wellb_{\beta\eta}$]\label{prop:unifetawb}
-  Given the updated compilation scheme, the \Uo of~\cref{sec:founif}
+  Given the updated compilation scheme, procedure \Uo of~\cref{sec:founif}
   is a good unification (as per \cref{prop:good-unif})
@@
 resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
-Since $X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$,
+Since $X$ is mapped to $B$ and $B_x$ is decompiled into $\lambda x.f \appsep x$,
 by the definition of \linketa decompilation
-$\sigma$ is decompiled to $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$,
+$\sigma$ is decompiled to $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$.
 
@@
 \printAlll
-  {{{\lambda x.(X\appsep x),a},
+  {{{\lambda x.(X\appsep x),f},
     {X,\lambda x.a}}}
-  {{{A,a},
+  {{{A,f},
     {C,\lambda x.a}}}
@@
 The core procedure of this post-processing step is \emph{align-arity} that
-is iterated by \elpiIn{map-deduplication}:
+is iterated by \emph{map-deduplication}:
 
@@
 The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
-$\linkStore_2$ by \progressetaleft, assigning <<$a \appsep x$>> to $B_x$.
+$\linkStore_2$ by \progressetaleft, assigning <<$f \appsep x$>> to $B_x$.
 The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
 turn triggers $\linkStore_1$ by \progressetaleft. In turn, this performs
-$\lambda x.a \Ue \lambda x.(a \appsep x)$ that fails as expected.
+$\lambda x.a \Ue \lambda x.(f \appsep x)$ that fails as expected.
 
@@
 \elpiIn{Extra1} (the compilation of \elpiIn{Extra}).
-This construction enforces \cref{inv:beta-rhs}.
+This construction enforces \fullref{inv:beta-rhs}.
 
@@
   by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
-  is solved and removed. Note that, in this case, all
+  is solved and removed. In this case, all
   calls to \Ue{} are between terms in \llambda and by \cref{th:fidelity-beta}
@@
   then by \cref{def:progBetaFail} unification fails as the
-  corresponding unification in \Fo{} would.
+  corresponding unification in \Fo{} would (it is called outside its domain).
 \end{proof}
@@
 
-The main difference is that we cannot implement \hrun since it comes as part of
+The main difference is that we cannot implement \hrun since it is
 the runtime of the programming language. In particular the runtime iterates
@@
 by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
-operator) \elpiIn{N} heigenvariables. Sequents are presented to
+operator) heigenvariables in \elpiIn{N}. Sequents are presented to
 CHR with their higher order unification variables replaced by
@@
 a constraint on the same variable; the third
-relocates \elpiIn{T2} and the latter crafs the new goal that unifies
+relocates \elpiIn{T2} and the latter crafts the new goal that unifies
 \elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
@@
 implementing a unification procedure in the meta language is likely to be significantly
-slower compared to the built-in one.
+slower compared to the built-in one on the common domain of first order terms.
 
@@
 not always feasible to adopt it for Coq due to the fact that the type system
-of the meta language is too limited to accommodate for that one the object language,
+of the meta language is too limited to accommodate for the one the object language,
 e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\

### 2024-05-13 paper/main.tex
typo

@@
 not always feasible to adopt it for Coq due to the fact that the type system
-of the meta language is too limited to accommodate for the one the object language,
+of the meta language is too limited to accommodate for the one of the object language,
 e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\

### 2024-06-28 paper/rebuttal.md
rebuttal draft

Dear reviewers and PC members, we are happy for the accurate and positive
reviews.
We take the occasion to answer a few questions, hopefully to satisfy the
interest of the reviewers.

----------------------- REVIEW 1 ---------------------
> Line 1368: The authors mention that one reason for delaying the
> application of substitutions is to keep the term structure indexable.
> One can imagine another reason for making such a delay, especially in
> a logic programming setting, is that failure can be encountered, and
> the work done in computing complete normal forms would be wasted.  Is
> there any evidence of that justification?

According to our experience, that is specific to the use of LP in the context
of Coq, delaying substitution (esp beta reduction) was not a crucial advantage.
An early prototype of Elpi did implement an explicit substitution calculus
but we could not observe a considerable gain (enough to justify its complexity).
Our intuition is that, in Coq, long reductions are not made of beta steps,
but rather fixpoint unfolding and ML-like pattern matching (interleaved with
betas). So just delaying beta, and sharing beta-normal forms, is not enough to
increase the performance in this context.

We hope to have correctly understood the question.

> It might interest the reader to explain what features of lambda Prolog
> that are not in Prolog remain in your specification.  For example,
> your code does not reduce HOU to first-order logic/Prolog.  The
> universal quantifier in goals and the pattern fragment unification are
> critical to your encoding, right?

We absolutely need:

1. pattern unification (and pi to induce fresh nominals)
2. delaying of goals (our links)

CHR rules, or a reified main loop (scheduling of goals), is needed only to
guarantee fidelity, that amounts at failing as soon as possible.
This may or may not be needed, depending on the use case.
It surely is in ours.

----------------------- REVIEW 2 ---------------------
> "our real goal is to simulate the execution of an entire logic-program."
> I do not understand what you mean by a "real goal".
> Is this in contrast to some goals that are not "real". I do not understand what this means.

Apologies for our English, we are not native speakers.

What we meant is that our ultimate/true goal is not (only) to implement an HO
unification procedure, but rather to simulate the execution of a whole
logic-program, that is made of many calls to that unification procedure.

We will reformulate and avoid the word "real", it is maybe a false friend with
"true" (in Italian).

> "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) below can preserve:"
> what can it preserve?

Sorry again, sloppy wording.

We should have written "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) can preserve
Invariant 1 written below".

----------------------- REVIEW 3 ---------------------
> Usually, I think of the eta rule as a type-directed expansion (capturing
> a form of extensionality) rather than a contraction.  I wonder if the
> authors could expand on (a) on the significance of eta for the object
> language as it relates to type-class resolution, and (b) whether it might
> be possible to make encoding sufficiently aware of the object-language
> types as eta might be modeled via an expansion.

The main reason why we did focus on contraction is that it is what the user
wants to see and typically writes when declaring type class instances. Eg
"Decidable prime" and not "Decidable (fun x => prime x)".

DAVIDE CHECK THIS: ma dove è che contraiamo? si può fare (b) facilmente? Serve chiamare coq.typecheck?

> In terms of related work, I might point out the paper
> "Higher-Order Logic Programming as Constraint Logic Programming" by
> Michaylov and Pfenning from PPCP 1993 which makes similar points as
> lines 1287-1291.  The Twelf manual also describes how nonpattern constraints
> are postponed, and how other constraint domains are integrated (Chapters 5.5 and 6),
> although not at the level of generality as CHR.

We sincerely thank the reviewer for this pointer.

### 2024-06-28 paper/rebuttal.md
update rebuttal

@@
 
+On the other hand, we think it is possible to produce "eta-expanded" solutions
+by slightly modifying our decompilation procedure: note that the substitution we
+provide to the object language is produced at the very end of the unification
+procedure. Therefore, let `MV` and `OV` a mapping of a meta variable and an
+object variable. If in the meta-substitution sigma, we assign a term `T` to
+`MV`, then when we translate the `T` for `OV`, we can, at first, get the type of
+`OV`, say `a -> b -> c`. If `T` is like `lambda x. f x x` then we can decompile
+it into `lambda y z x.(f x x) z y`
+
 DAVIDE CHECK THIS: ma dove è che contraiamo? si può fare (b) facilmente? Serve chiamare coq.typecheck?

### 2024-06-28 paper/rebuttal.md
rebuttal

--- 2024-06-28 paper/rebuttal.md
@@
 
-According to our experience, that is specific to the use of LP in the context
-of Coq, delaying substitution (esp beta reduction) was not a crucial advantage.
-An early prototype of Elpi did implement an explicit substitution calculus
-but we could not observe a considerable gain (enough to justify its complexity).
-Our intuition is that, in Coq, long reductions are not made of beta steps,
-but rather fixpoint unfolding and ML-like pattern matching (interleaved with
-betas). So just delaying beta, and sharing beta-normal forms, is not enough to
-increase the performance in this context.
+We hope to have correctly understood the question.
 
-We hope to have correctly understood the question.
+In our experience type-class search does not exhibit much backtracking,
+especially if the rules are indexed properly, deep enough to find the
+discriminating term constructor. It is out of scope for this paper, but
+another aspect we are investigating is a functionality/determinacy analysis
+since many type classes happen to be functions.
 
@@
 This may or may not be needed, depending on the use case.
-It surely is in ours.
+It surely is needed for our use case.
+
+We will try to make these points clear in the conclusion.
 
@@
 
-DAVIDE CHECK THIS: ma dove è che contraiamo? si può fare (b) facilmente? Serve chiamare coq.typecheck?
+On the other hand, we think it is possible to produce "eta-expanded" solutions
+by slightly modifying our decompilation procedure that is run at the very end
+in order to produce the substitution for the object language.
+Since Coq-Elpi runs withing Coq we could invoke Coq's type checker and use
+the type to perform eta expansion.
 
@@
 
-We sincerely thank the reviewer for this pointer.
+We sincerely thank the reviewer for these pointers, we will integrate them in
+the last section.

### 2024-07-01 paper/TODO_rebuttal.md
add TODO rebuttal with checklist

# REVIEW 1

- [ ] Line 1368: The authors mention that one reason for delaying the
  application of substitutions is to keep the term structure indexable.
  One can imagine another reason for making such a delay, especially in
  a logic programming setting, is that failure can be encountered, and
  the work done in computing complete normal forms would be wasted.  Is
  there any evidence of that justification?
- [ ] It might interest the reader to explain what features of lambda Prolog
  that are not in Prolog remain in your specification.  For example,
  your code does not reduce HOU to first-order logic/Prolog.  The
  universal quantifier in goals and the pattern fragment unification are
  critical to your encoding, right?
- [ ] line 26: This is the first time I've seen the "\beta\eta"
   conversion written as "\eta\beta$.  Is there a good reason to
   switch this order from the normal?
- [ ] line 70: "for the Coq's system" - rewrite
- [ ] line 112: The reference for pattern unification is not [11] but
   Miller's 1991 paper in the J. of Logic and Computation.
- [ ] line 247: Using "subscript" as a verb is odd.  Revise that
   sentence.
- [ ] line 240: Definition 2.6 asks that a term is an eta expansion.
   Does this mean the term is an eta-redex, a term of the form 
   (lambda x. M x) where x is not free in M ?
- [ ] line 1117 "desappears"
- [ ] line 1301: "heigenvariables"
- [ ] line 1311: "mathes"
- [ ] line 1324; "since it removes all THE data from THE rule's heads"
- [ ] line 1335: "to accommodate for the one" drop "for"?
- [ ] line 1355: "reather"
- [ ] An early attempt to reduce HOU (at least, 2nd order unification) to
   the pattern fragment can be found in Miller 1991 paper "Unification
   of Simply Typed Lambda-Terms as Logic Programming", ICLP91. That
   early approach does not provide the flexibility apparent in the
   present paper.

# REVIEW 2

- [ ] The abstract is a little bit vague. It writes "more well-behaved" and 
      "problems" without any hint about what this might mean. A stronger 
      abstract would define the problem more clearly.
- [ ] "Instance forall_dec: ∀A P, Finite A → (* r3 *) ∀x:A, Decision (P x) →
      Decision (∀x:A, P x)." I think parentheses are missing around "∀x:A, 
      Decision (P x)"?
- [ ] "an Higher-Order Abstract Syntax" an => a
- [ ] "for the Coq’s system" => "for Coq's type system"
- [ ] I'm a bit surprised it is called "good unification" and not "most general 
      unification"
- [ ] "our real goal is to simulate the execution of an entire logic-program."
      I do not understand what you mean by a "real goal". Is this in contrast 
      to some goals that are not "real". I do not understand what this means.
- [ ] "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) below can preserve" what can it preserve?
- [ ] "ensures the mapping are bijective" are => is
- [ ] In the definition of comp: "(arity z)" What is "arity z"?

# REVIEW 3

- [ ] Usually, I think of the eta rule as a type-directed expansion (capturing
    a form of extensionality) rather than a contraction.  I wonder if the
    authors could expand on (a) on the significance of eta for the object
    language as it relates to type-class resolution, and (b) whether it might
    be possible to make encoding sufficiently aware of the object-language
    types as eta might be modeled via an expansion.
- [ ] In terms of related work, I might point out the paper
    "Higher-Order Logic Programming as Constraint Logic Programming" by
    Michaylov and Pfenning from PPCP 1993 which makes similar points as
    lines 1287-1291.  The Twelf manual also describes how nonpattern constraints
    are postponed, and how other constraint domains are integrated (Chapters 5.5 and 6),
    although not at the level of generality as CHR.

### 2024-07-01 paper/TODO_rebuttal.md
add comment TODO_rebuttal

@@
    switch this order from the normal?
+   > Comment: We use beta-eta since we treat beta before eta in the paper
 - [ ] line 70: "for the Coq's system" - rewrite

### 2024-07-01 paper/TODO_rebuttal.md
update TODO_rebuttal

--- 2024-07-01 paper/TODO_rebuttal.md
@@
    switch this order from the normal?
+   > Comment: We use beta-eta since we treat beta before eta in the paper
 - [ ] line 70: "for the Coq's system" - rewrite
-- [ ] line 112: The reference for pattern unification is not [11] but
-   Miller's 1991 paper in the J. of Logic and Computation.
+- [x] line 112: The reference for pattern unification is not [11] but
+   Miller's 1991 paper in the J. of Logic and Computation. (cf 827b11dd2df213b9f446b77434fef8cb6591f00b)
 - [ ] line 247: Using "subscript" as a verb is odd.  Revise that

### 2024-07-01 paper/main.tex
correct ref to miller
desappears -> disappears
mathes -> matches

@@
 In the rest of this paper, we refer to the pattern fragment as
-\llambda~\cite{miller92jsc}.
+\llambda~\cite{miller1991}.
 
@@
 \noindent
-Note that $\mapping{X}{C}{0}$ desappears from \mapStore
+Note that $\mapping{X}{C}{0}$ disappears from \mapStore
 in favour of the auxiliary \linketa \linketaM{}{C}{\lambda x.B_{x}}.
@@
 The first directive matches a constraint
-whose \lhs is a variable \elpiIn{X}; the second mathes and removes
+whose \lhs is a variable \elpiIn{X}; the second matches and removes
 a constraint on the same variable; the third

### 2024-07-01 paper/main.tex
line 1324; "since it removes all THE data from THE rule's heads"

@@
 programming engine provided by the meta language since
-it removes all data from rule's heads and makes
+it removes all the data from the rule's heads and makes
 indexing degrade. Additionally,

### 2024-07-01 paper/main.tex
line 1335: "to accommodate for the one" drop "for"?

@@
 not always feasible to adopt it for Coq due to the fact that the type system
-of the meta language is too limited to accommodate for the one of the object language,
+of the meta language is too limited to accommodate the one of the object language,
 e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\

### 2024-07-01 paper/main.tex
reather -> rather

@@
 the authors are not interested in exploiting a decidable fragment of
-higher-order unification but are reather leaning towards an interactive
+higher-order unification but are rather leaning towards an interactive
 system where the user is given

### 2024-07-01 paper/main.tex
forall x:A, Decision (P x) in parentheses
an Higher -> a Higher

@@
 Instance forall_dec: ~$\forall$~A P, Finite A ~$\to$~            (* r3 *)
-  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
+  ~($\forall$~x:A, Decision (P x)) ~$\to$~ Decision (~$\forall$~x:A, P x).
 \end{coqcode}
@@
 Elpi comes equipped with
-an Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
+a Higher-Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
 Coq terms, called \elpiIn{tm}, that includes (among others) the following

### 2024-07-01 paper/main.tex
"for the Coq’s system" => "for Coq's type system"

@@
 form of proof search known as type-class resolution~\cite{wadler89,sozeau08}
-for the Coq's system~\cite{Coq-refman}.
+for Coq's type system~\cite{Coq-refman}.
 Type-class solvers are unification based proof search procedures

### 2024-07-01 paper/main.tex
ensures the mapping are bijective" are => is

@@
 of arguments, the compiler generates multiple mappings for it, on a first
-approximation, and then ensures the mapping are bijective by introducing
+approximation, and then ensures the mapping is bijective by introducing
 \linketa; this detail is discussed in section \ref{sec:invariant1}.

### 2024-07-01 paper/main.tex
"so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) below can preserve" what can it preserve?
Plus I removed a line break to not exceed the 12 pages

@@
 %the one of \Ho.
-
+%
 The compiler establishes a mapping between variables of the two languages.
@@
 Each \elpiIn{mvariable} is stored in the mapping together with
-its arity (a number) so that the code of \ref{clause:malloc} below can preserve:
+its arity (a number) so that the code of \ref{clause:malloc} below can preserve
+the following property:
 

### 2024-07-01 paper/main.tex
heigenvariables -> eigenvariables

@@
 by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
-operator) heigenvariables in \elpiIn{N}. Sequents are presented to
+operator) eigenvariables in \elpiIn{N}. Sequents are presented to
 CHR with their higher order unification variables replaced by

### 2024-07-01 paper/main.tex
line 247: Using "subscript" as a verb is odd. Revise that sentence.

@@
 When it is clear from the context, we shall use the same syntax for \Fo{} terms
-(although we never subscripts unification variables).\\
+(although we never use subscripts for unification variables).\\
 We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$, \ldots for

### 2024-07-01 paper/main.tex
eta expansion -> $\eta$-expansion
real goal -> final goal

@@
 Even if we provide an implementation of the object-language unification
-\Uo{} in \cref{sec:founif}, our real goal is to simulate the execution of
+\Uo{} in \cref{sec:founif}, our final goal is to simulate the execution of
 an entire logic-program.
@@
   \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho (\lambda x.s)$
-  is an eta expansion. % without the lambda, 5.11 is not true, that
+  is an $\eta$-expansion. % without the lambda, 5.11 is not true, that
   % is we assume variables are in W

### 2024-07-01 paper/main.tex
rework last paragraph abstract

@@
 
-  In this paper we propose a more well-behaved encoding called \Ho,
-  and show how to translate unification problems from \Fo{}
-  to corresponding ones in \Ho.
-  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
-    that honours $\eta\beta$-equivalence (for Coq functions), addresses
-  problems within the pattern fragment, and allows
-  for the use of heuristics to deal with problems outside the
-  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
+  In this paper we propose to translate \Fo{} unification problems
+  into an alternative encoding called \Ho, from which we derive \Uo,
+  the higher-order unification procedure of \Fo{}. \Uo
+  honours $\eta\beta$-equivalence for terms
+  within the pattern fragment, and allows the use for 
+  heuristics when unification is performed with terms outside the pattern fragment.
+  Moreover, as \Uo{} delegates most of the work to \Ue,
   it can be used to efficiently simulate a logic program in \Fo{} by

### 2024-07-01 paper/main.tex
(sec 8) add ref Higher-Order Logic Programming as Constraint Logic Programming

@@
 Constraint Handling Rules (CHR) to operate on them~\cite{TASSI_2019,fruehwirth2017constraint}.
-An important design choice is that constraints are suspended on a user
-computed list of unification variables, and are resumed \emph{as soon as one of them
-is assigned}, before any other existing goal is considered. In turn this
-property allows for a timely activation of links that grants~\fullref{prop:fidelity}.
+Similarly to \cite{pfenning1993}, a constraint is a delayed unification problems
+suspended on a list of variables and is resumed \emph{as soon as one of these variable
+is assigned}, before any other existing goal is considered. We point out that
+we extend \cite{pfenning1993}, since we delay not only \notllambda unification problems, 
+but also those problems that are still in \maybeeta.  
+In turn link activation grants~\fullref{prop:fidelity}.
 For brevity we only provide two pseaudo-code snippets. The first one depicts

### 2024-07-02 paper/main.tex
eta-exp -> eta-redex

@@
   \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho (\lambda x.s)$
-  is an $\eta$-expansion. % without the lambda, 5.11 is not true, that
+  is an $\eta$-redex. % without the lambda, 5.11 is not true, that
   % is we assume variables are in W
@@
 A term 
-$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
+$\lambda x. t \appsep x$ is said to be the $\eta$-redex of $t$ if
 $x$ does not occur free in $t$, and conversely we say that $t$ is the $\eta$-contraction of
@@
 subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$
-can be an $\eta$-expansion, i.e. if
+can be an $\eta$-redex, i.e. if
 there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
@@
 The first two examples are easy, and show how a unification variable can expose
-or erase a variable in its scope, turning the resulting term into an $\eta$-expansion or not.\\
+or erase a variable in its scope, turning the resulting term into an $\eta$-redex or not.\\
 The third example shows that when a variable occurs outside the scope of a unification
-variable, it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
+variable, it cannot be erased and can hence prevent a term from being an $\eta$-redex.\\
 The last example shows the recursive nature of the check we need to implement.
@@
 $f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
-$\eta$-expansion.\\
+$\eta$-redex.\\
 We can now detail how \maybeeta terms are detected.
@@
 If the term $s$ is not exactly $x$ (case 1)
-it can only be an $\eta$-expansion of $x$, or a unification
+it can only be an $\eta$-redex of $x$, or a unification
 variable that can be assigned to $x$, or a combination of both.
@@
 there exists no substitution $\rho$ such that $\rho t$ is an
-$\eta$-expansion. A simple counter example is
+$\eta$-redex. A simple counter example is
 $\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
@@
   \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
-    we unify the $\eta$-expansion of $X$ with $t$, that is we run
+    we unify the $\eta$-redex of $X$ with $t$, that is we run
     $\lambda x.X \appsep x\Ue{} t$
@@
 %   $X$ is variable, and, by \cref{def:progressetaright}, $\lambda x.t$ is either no more
-%   a \maybeeta, i.e. it is not a $\eta$-expansion and, so, $\wellb(\lambda x.t)$, 
+%   a \maybeeta, i.e. it is not a $\eta$-redex and, so, $\wellb(\lambda x.t)$, 
 %   the unification is performed between $X$ and $\lambda x.t$.
@@
   Moreover each rule only relies on terminating operations such as \Ue,
-  $\eta$-contraction, $\eta$-expansion, relocation (a recursive copy of a
+  $\eta$-contraction, $\eta$-redex, relocation (a recursive copy of a
   finite term).
@@
           If $s_1$ and $s_2$ are equal thanks to \ref{clause:eta1} then
-          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
+          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-redex of $t_2$ namly
           $\lambda x.t_2\appsep x$, otherwise they are equal thanks to rule \ref{clause:lam-lam}
@@
   Otherwise if $p < n$ then we have to consider two cases. 1) If $t_1$ is a rigid
-  term then $t_2$ must be the $\eta$-expansion of $t_1$, implying $q = 0$. If it
+  term then $t_2$ must be the $\eta$-redex of $t_1$, implying $q = 0$. If it
   is not the case, then \progressetaleft
@@
 % the other hand, we also point out that the \maybeeta detection spot out
-% potential $\eta$-expansion for terms that are not in \llambda. For example,
+% potential $\eta$-redex for terms that are not in \llambda. For example,
 % $\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the

### 2024-07-02 paper/main.tex
eta-beta -> beta-eta

@@
   Elpi's equational theory includes
-  $\eta\beta$-equivalence and features a
+  $\beta\eta$-equivalence and features a
   higher-order unification procedure \Ue 
@@
   the higher-order unification procedure of \Fo{}. \Uo
-  honours $\eta\beta$-equivalence for terms
+  honours $\beta\eta$-equivalence for terms
   within the pattern fragment, and allows the use for 
@@
 Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
-$\eta\beta$-equivalence and features a higher-order unification procedure \Ue
+$\beta\eta$-equivalence and features a higher-order unification procedure \Ue
 for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
@@
 Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
-that honours $\eta\beta$-equivalence (for Coq functions), addresses problems
+that honours $\beta\eta$-equivalence (for Coq functions), addresses problems
 within the pattern fragment, and allows for the use of heuristics to deal with
@@
 The equational theory of the meta-language Elpi encompasses
-$\eta\beta$-equivalence and its unification procedure can solve higher-order
+$\beta\eta$-equivalence and its unification procedure can solve higher-order
 problems in the pattern fragment. Although 
@@
 reasons automatic proof search procedures typically employ a unification
-procedure that only captures a $\eta\beta$-equivalence and only operates
+procedure that only captures a $\beta\eta$-equivalence and only operates
 in \llambda. The similarity is striking, but one needs to exercise some caution in order to
@@
 used to simulate a higher-order unification procedure \Uo for
-the object language that features $\eta\beta$-conversion. We show how \Uo
+the object language that features $\beta\eta$-conversion. We show how \Uo
 can be extended with heuristics to deal with problems outside the pattern
@@
 % a \hstep that respects $\beta$-conversion for terms in \llambda.
-% The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
+% The extension to $\beta\eta$-conversion is described in \cref{sec:eta} and
 % the support for terms outside \llambda in \cref{sec:beta}.

### 2024-07-15 paper/main.tex
minor rewording

@@
 
-  In this paper we propose to translate \Fo{} unification problems
-  into an alternative encoding called \Ho, from which we derive \Uo,
+  In this paper we translate unification problems in \Fo{} 
+  to an alternative encoding called \Ho, from which we derive \Uo,
   the higher-order unification procedure of \Fo{}. \Uo
   honours $\beta\eta$-equivalence for terms
-  within the pattern fragment, and allows the use for 
-  heuristics when unification is performed with terms outside the pattern fragment.
+  within the pattern fragment, and allows for the use of
+  heuristics when the terms fall outside the pattern fragment.
   Moreover, as \Uo{} delegates most of the work to \Ue,
@@
 \end{abstract}
-
-\begin{comment}
-Specifying and implementing a proof system from scratch requires significant
-effort. Logical Frameworks and Higher Order Logic Programming Languages provide
-dedicated, high-level meta languages to facilitate this task in two ways: 1)
-variable binding and substitution are for free when meta language binders
-represent object logic ones; 2) proof construction, and
-proof search, are greatly simplified by leveraging the unification procedure
-provided by the meta language. Notable examples of meta languages are Elf,
-Twelf, $\lambda$Prolog, Beluga, Abella and Isabelle which have been used to
-implement or specify many formal systems such as First Order Logic, Set Theory,
-Higher Order Logic, and the Calculus of Constructions.
-
-The object logic we are interested in is Coq's type theory. We aim to develop a
-higher-order unification-based proof search procedure using the meta language
-Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
-$\beta\eta$-equivalence and features a higher-order unification procedure \Ue
-for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
-for meta programming but that restricts \Ue to first-order unification problems
-only. We refer to this basic encoding as \Fo.
-
-In this paper we propose a more well-behaved encoding called \Ho, and show how
-to translate unification problems from \Fo{} to corresponding ones in \Ho.
-Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
-that honours $\beta\eta$-equivalence (for Coq functions), addresses problems
-within the pattern fragment, and allows for the use of heuristics to deal with
-problems outside the pattern fragment. Moreover, as \Uo{} delegates most of the
-work to \Ue, it can be used to efficiently simulate a logic program in \Fo{} by
-taking advantage of unification-related optimizations of the meta language, such
-as clause indexing.
-\end{comment}
 
@@
 Constraint Handling Rules (CHR) to operate on them~\cite{TASSI_2019,fruehwirth2017constraint}.
-Similarly to \cite{pfenning1993}, a constraint is a delayed unification problems
-suspended on a list of variables and is resumed \emph{as soon as one of these variable
-is assigned}, before any other existing goal is considered. We point out that
-we extend \cite{pfenning1993}, since we delay not only \notllambda unification problems, 
-but also those problems that are still in \maybeeta.  
+Similarly to \cite{pfenning1993} a constraint is a goal % unification problems
+suspended on a list of variables that is resumed \emph{as soon as one of these variable
+is assigned}, before any other existing goal is considered. 
 In turn link activation grants~\fullref{prop:fidelity}.
+We point out that \cite{pfenning1993} delays \notllambda unification problems
+while we also delay problems that are in \maybeeta.
 For brevity we only provide two pseaudo-code snippets. The first one depicts

### 2024-07-24 paper/paper_for_taps/source/code/mem.tex
add paper_for_taps folder (to build the zip)

\begin{elpicode}
  kind addr type.
  type addr nat -> addr.
  typeabbrev (mem A) (list (option A)).
  type set? addr -> mem A -> A -> o.
  set? (addr A) Mem Val :- get A Mem Val.

  type unset? addr -> mem A -> o.
  unset? Addr Mem :- not (set? Addr Mem _).

  type assign-aux nat -> mem A -> A -> mem A -> o.
  assign-aux z (none :: L) Y (some Y :: L).
  assign-aux (s N) (X :: L) Y (X :: L1) :- assign-aux N L Y L1.

  type assign addr -> mem A -> A -> mem A -> o.
  assign (addr A) Mem1 Val Mem2 :- assign-aux A Mem1 Val Mem2.

  type get nat -> mem A -> A -> o.
  get z (some Y :: _) Y.
  get (s N) (_ :: L) X :- get N L X.

  type alloc-aux nat -> mem A -> mem A -> o.
  alloc-aux z [] [none] :- !.
  alloc-aux z L L.
  alloc-aux (s N) [] [none | M] :- alloc-aux N [] M.
  alloc-aux (s N) [X | L] [X | M] :- alloc-aux N L M.

  type alloc addr -> mem A -> mem A -> o.
  alloc (addr A as Ad) Mem1 Mem2 :- unset? Ad Mem1, 
    alloc-aux A Mem1 Mem2.

  type new-aux mem A -> nat -> mem A -> o.
  new-aux [] z [none].
  new-aux [A | As] (s N) [A | Bs] :- new-aux As N Bs.

  type new mem A -> addr -> mem A -> o.
  new Mem1 (addr Ad) Mem2 :- new-aux Mem1 Ad Mem2.

\end{elpicode}

### 2024-07-24 paper/paper_for_taps/source/code/meta_language.tex
add paper_for_taps folder (to build the zip)

\begin{elpicode}

  kind inctx type -> type.                              ~($\customlabel{data:inctx}{\cdot\vdash\cdot}$)~
  type abs (tm -> inctx A) -> inctx A.
  type val A -> inctx A.
  typeabbrev assignment (inctx tm).
  typeabbrev subst (mem assignment).

  kind tm  type.
  type app list tm -> tm.
  type lam (tm -> tm) -> tm.
  type con string -> tm.
  type uva addr -> list tm -> tm.

  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
  (con C ~$\Ue$~ con C) S S.
  (app L1 ~$\Ue$~ app L2) S S1 :- fold2 (~$\Ue$~) L1 L2 S S1.
  (lam F1 ~$\Ue$~ lam F2) S S1 :- 
    pi x\ (pi S\ (x ~$\Ue$~ x) S S) => (F1 x ~$\Ue$~ F2 x) S S1.
  (uva N Args ~$\Ue$~ T) S S1 :- 
    set? N S F,!, move F Args T1, (T1 ~$\Ue$~ T) S S1.
  (T ~$\Ue$~ uva N Args) S S1 :- 
    set? N S F,!, move F Args T1, (T ~$\Ue$~ T1) S S1.
  (uva M A1 ~$\Ue$~ uva N A2) S1 S2 :- !, 
    pattern-fragment A1, pattern-fragment A2, 
    prune! M A1 N A2 S1 S2.
  (uva N Args ~$\Ue$~ T) S S1 :- not_occ N S T, pattern-fragment Args, 
    bind T Args T1, assign N S T1 S1.
  (T ~$\Ue$~ uva N Args) S S1 :- not_occ N S T, pattern-fragment Args, 
    bind T Args T1, assign N S T1 S1.

  type prune! addr -> list tm -> addr -> 
              list tm -> subst -> subst -> o.
  /* no pruning needed */               
  prune! N A  N A  S S :- !.
  prune! M A  N A  S1 S2 :- !, bind (uva M A) A Ass, 
    assign N S1 Ass S2.
  /* prune different arguments */
  prune! N A1 N A2 S1 S3 :- !,
    new S1 W S2, prune-same-variable W A1 A2 [] Ass,
    assign N S2 Ass S3.
  /* prune to the intersection of scopes */
  prune! N A1 M A2 S1 S4 :- !,
    new S1 W S2, prune-diff-variables W A1 A2 Ass1 Ass2, 
    assign N S2 Ass1 S3,
    assign M S3 Ass2 S4.

  type prune-same-variable addr -> list tm -> list tm -> 
                            list tm -> assignment -> o.
  prune-same-variable N [] [] ACC (val (uva N Args)) :- 
    rev ACC Args.
  prune-same-variable N [X|XS] [X|YS] ACC (abs F) :- 
    pi x\ prune-same-variable N XS YS [x|ACC] (F x).
  prune-same-variable N [_|XS] [_|YS] ACC (abs F) :- 
    pi x\ prune-same-variable N XS YS ACC (F x).

  type permute list nat -> list tm -> list tm -> o.
  permute [] _ [].
  permute [P|PS] Args [T|TS] :-
    nth P Args T,
    permute PS Args TS.

  type build-perm-assign addr -> list tm -> list bool -> 
                      list nat -> assignment -> o.
  build-perm-assign N ArgsR [] Perm (val (uva N PermutedArgs)) :- 
    rev ArgsR Args, permute Perm Args PermutedArgs.
  build-perm-assign N Acc [tt|L] Perm (abs T) :- 
    pi x\ build-perm-assign N [x|Acc] L Perm (T x).
  build-perm-assign N Acc [ff|L] Perm (abs T) :- 
    pi x\ build-perm-assign N Acc L Perm (T x).

  type keep list A -> A -> bool -> o.
  keep L A tt :- mem L A, !.
  keep _ _ ff.

  type prune-diff-variables addr -> list tm -> list tm -> 
                            assignment -> assignment -> o.
  prune-diff-variables N Args1 Args2 Ass1 Ass2 :-
    map (keep Args2) Args1 Bits1, 
    map (keep Args1) Args2 Bits2,
    filter Args1 (mem Args2) ToKeep1,
    filter Args2 (mem Args1) ToKeep2,
    map (index ToKeep1) ToKeep1 IdPerm,
    map (index ToKeep1) ToKeep2 Perm21,
    build-perm-assign N [] Bits1 IdPerm Ass1,
    build-perm-assign N [] Bits2 Perm21 Ass2.

  type beta tm -> list tm -> tm -> o.
  beta A [] A :- !.
  beta (lam Bo) [H | L] R :- beta (Bo H) L R1, beta-aux R1 R.
  beta (app A) L (app X) :- append A L X.
  beta (con H) L (app [con H | L]).
  beta X L (app[X|L]) :- name X.

  type beta-aux tm -> tm -> o.
  beta-aux (app [HD|TL]) R :- !, beta HD TL R.
  beta-aux A A.

  /* occur check for N before crossing a functor */
  type not_occ  addr -> subst -> tm -> o.
  not_occ N S (uva M Args) :- set? M S F, 
    move F Args T, not_occ N S T.
  not_occ N S (uva M Args) :- unset? M S, not (M = N),
    forall1 (not_occ_aux N S) Args.
  not_occ _ _ (con _).
  not_occ N S (app L) :- not_occ_aux N S (app L).
  /* Note: lam is a functor for the meta language! */
  not_occ N S (lam L) :- pi x\ not_occ_aux N S (L x).
  not_occ _ _ X :- name X.
  /* finding N is ok */
  not_occ N _ (uva N _).

  /* occur check for X after crossing a functor */
  type not_occ_aux addr -> subst -> tm -> o.
  not_occ_aux N S (uva M _) :- unset? M S, not (N = M).
  not_occ_aux N S (uva M Args) :- set? M S F, 
    move F Args T, not_occ_aux N S T.
  not_occ_aux N S (app L) :- forall1 (not_occ_aux N S) L.
  not_occ_aux N S (lam F) :- pi x\ not_occ_aux N S (F x).
  not_occ_aux _ _ (con _).
  not_occ_aux _ _ X :- name X.
  /* finding N is ko, hence no rule */

  /* copy T T' vails if T contains a free variable, i.e. it
     performs scope checking for bind */
  type copy tm -> tm -> o.
  copy (con C)   (con C).
  copy (app L)   (app L') :- map copy L L'.
  copy (lam T)   (lam T') :- pi x\ copy x x => copy (T x) (T' x).
  copy (uva A L) (uva A L') :- map copy L L'.

  type bind tm -> list tm -> assignment -> o.
  bind T [] (val T') :- copy T T'.
  bind T [X | TL] (abs T') :- pi x\ copy X x => bind T TL (T' x).

  type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
  deref _ (con C) (con C).
  deref S (app A) (app B) :- map (deref S) A B.
  deref S (lam F) (lam G) :-
    pi x\ deref S x x => deref S (F x) (G x).
  deref S (uva N L) R :- set? N S A,
    move A L T, deref S T R.
  deref S (uva N A) (uva N B) :- unset? N S,
    map (deref S) A B.

  type move assignment -> list tm -> tm -> o.
  move (abs Bo) [H|L] R :- move (Bo H) L R.
  move (val A)  [] A :- !.

  type deref-assmt subst -> assignment -> assignment -> o.
  deref-assmt S (abs T) (abs R) :- pi x\ deref-assmt S (T x) (R x).
  deref-assmt S (val T) (val R) :- deref S T R.
\end{elpicode}

### 2024-07-24 paper/paper_for_taps/source/main.tex
add paper_for_taps folder (to build the zip)

--- 2024-07-15 paper/main.tex
@@
-\documentclass[sigconf,natbib=false,review]{acmart}
+\documentclass[sigconf,natbib=false]{acmart}
 \usepackage[]{biblatex}
@@
 \usepackage{enumitem}
-\usepackage{intcalc}
-\usepackage{nameref}
-
-% TODO: set this fields
-%\setcopyright{cc}
-%\setcctype{by}
-%\copyrightyear{2024}
-%\acmYear{XXXX 2024}
-%\acmBooktitle{YYY}
-%\acmDOI{ZZZZZZZZZZZZ}
+
+\setcopyright{acmlicensed}  %
+\copyrightyear{2024}
+\acmYear{2024}
+\acmDOI{10.1145/NNNNNNN.NNNNNNN}  % leave as is for now
+\acmConference[PPDP '24]{PPDP '24}{September 10-11, 2024}{Milano, Italy}
+\acmBooktitle{Proceedings of the 26th Symposium on Principles and Practice of Declarative Programming, PPDP 2024, Milano, Italy, September 10-11, 2024}
+\acmPrice{15.00}
+\acmISBN{9-8-4007-0969-297}
 
@@
 }
-%\todo{use faddr for Fo}
-%
-%faddr and addr are memory addresses, the details are given in bla,
-%here they are just unique identifiers for unif variables and their types
-%do now allow to mistake one for the other.
-%\todo{cleanup}
 
@@
 
-% An example of $t$ in \notllambda{} is $F \appsep a$ for a constant $a$.
-% Note however tha
-% an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
-% that makes the resulting term fall back in \llambda.
-% \todo{said before}
-
 \newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
 We write $\subterm{t}$ for the set of sub-terms of $t$, and
-%\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
-%  largest set $\subterm{t}$ that can be obtained by the following rules.
-%$$
-%\begin{array}{l}
-%t \in \subterm{t}\\
-%t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
-%t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
-%\end{array}
-%$$
-%\end{definition}
-%\noindent
 we write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.
@@
 % \noindent
-% \todo{remove pruning optimization from code above}
 
@@
 \end{lemma}
-% \begin{proof}[Proof sketch]
-%   problematic terms are mapped to uva by comp,
-%   the problematic o-app node is gone.
-%   if F x = t then  F = lam x.t[]
-%   \todo{finish/maybe W enforcement is not needed here}
-% \end{proof}
 \noindent
@@
   of \linketa guarantees \fullref{prop:fidelity}.
-  %\footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
   \label{lemma:fidelity-maybeeta}
@@
 
-\begin{proof}[Proof sketch] %\todo{to reformulate}
-  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
-  % succeeds. A $\hstep_i$ is made by a unification
-  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
-  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
+\begin{proof}[Proof sketch]
   We want to prove that \fstep succeeds iff \hstep
@@
 
-% \todo{explain why decompile does so and which eta?}
-
-\begin{comment}
-\paragraph{Example of \progressetadedup}
-
-A very basic example of \linketa deduplication, is given below:
-% make test ONLY=7002 TEX=tex
-\printAlll
-  {{{\lambda x.(X\appsep x),\lambda x.(Y\appsep x)}}}
-  {{{A,C}}}
-  {{{X,B,1},
-    {Y,D,1}}}
-  {{{\eta,,A,\lambda x.B_{x}},
-    {\eta,,C,\lambda x.D_{x}}}}
-
-\noindent
-The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
-By unifying the two \rhs we get
-$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
-In turn, given the map \mapStore, this second assignment is decompiled to
-$\rho = \{~ X \mapsto Y ~\}$ as expected.
-\end{comment}
-
-%\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
 \section{Making \mapStore a bijection}
@@
 a bijection. For example:
-\begin{comment}
-%
-% make test ONLY=7001 TEX=tex
-\printAlll
-  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
-    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
-  {{{A,\lambda x.\lambda y.x},
-    {D,F}}}
-  {{{X,E,1},
-    {Y,F,0},
-    {X,C,2}}}
-  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
-    {\eta,,A,\lambda x.B_{x}},
-    {\eta,x,B_{x},\lambda y.C_{y x}}}}
-\end{comment}
 % make test ONLY=7007 TEX=tex
@@
   {{{\eta,,A,\lambda x.B_{x}}}}
-% \todo{simplify example since no prog right}
 in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
@@
 
-% \todo{this proof is still unreadable}
-% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}
-
 % Note that shortening the chain of links allows unification to fail if
 % $X_1$ and $X_2$ are unified with different terms.
-% \todo{what are these X1 and X2?}
 
@@
 
-\begin{comment}
-\paragraph{Example of \progBetaLL}
-Consider the \linkbeta below:
-%
-% make test ONLY=7004 TEX=tex
-\printAlll
-  {{{X,\lambda x.x},
-    {\lambda x.(Y\appsep (X\appsep x)),f}}}
-  {{{A,\lambda x.x},
-    {B,f}}}
-  {{{Y,D,0},
-    {X,A,0}}}
-  {{{\eta,,A,\lambda x.E_{x}},
-    {\eta,,B,\lambda x.C_{x}},
-    {\llambda,x,C_{x},(D\appsep E_{x})}}}
-
-\noindent
-Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
-The first unification problem results in $\sigma =
-\{A \mapsto \lambda x.x\}$. In turn this instantiation
-triggers $\linkStore_1$ by \progressetaleft 
-and $E_x$ is assigned to $x$.
-Under this substitution the \linkbeta becomes
-\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
-it is replaced with the link:
-\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
-problem assigns $f$ to $B$, that in turn activates
-the second \linketa ($f$ is assigned to $C$), and then all the remaining links
-are solved. The final \Ho substitution is $\sigma = \{~A _{} \mapsto \lambda x.x ~;~
-B _{} \mapsto f ~;~ C _{x} \mapsto (f\appsep x) ~;~
-D _{} \mapsto f ~;~ E _{x} \mapsto x ~;~
-F _{x} \mapsto C_{x} ~\}$ and is decompiled into $\rho = \{~X \mapsto \lambda x.x ~;~
-Y \mapsto f~\}$.
-
-\paragraph{Example of \progBetaRH}
-We give a second look at the opening example in \cref{sec:beta}.
-% \todo{this is a little more complex that the opening problem}
-The problem is compiled into:
-%
-% make test ONLY=7003 TEX=tex
-\printAlll
-  {{{X,\lambda x.a},
-    {(X\appsep a),a}}}
-  {{{A,\lambda x.a},
-    {B,a}}}
-  {{{X,A,0}}}
-  {{{\llambda,,B,(A\appsep a)}}}
-
-% \todo{Rinominare i print dei problemi}
-
-The first unification problem is solved with the substitution
-$\sigma = \{~A \mapsto \lambda x.a~\}$. The \linkbeta becomes
-\linkbetaM{}{B}{((\lambda x.a) \appsep a)} whose \rhs can be $\beta$-reduced to
-$a$. $a$ is in \llambda and is unified with $B$. The resolution of the second
-unification problem
-confirm that $B$ equal $a$. The final substitution $\sigma = \{~A \mapsto \lambda x.a ~;~ B \mapsto a ~\}$ is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
-a ~\}$.
-\end{comment} 
 \subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}

### 2024-07-24 paper/paper_for_taps/source/code/generate_tex.py

@@ -1,0 +1,74 @@
+import os, sys, re
+
+def get_file_cnt(lines):
+    res = []
+    try:
+        indexBegin = lines.index("%BEGIN\n")
+        indexEnd = lines.index("%END\n")
+        res = lines[indexBegin+1:indexEnd]
+    finally:
+        return res
+
+
+def print_tex(lines, fout, raw = False):
+    with open(fout, "w") as f:
+        if not raw:
+            f.write("\\begin{elpicode}\n")
+        for l in lines:
+            l = re.sub("^ *% +.*\n","",l)   
+            l = re.sub("%~(.*)",r"~\g<1>",l)   
+            l = re.sub("^ *%SNIP.*\n","",l)   
+            l = re.sub("^ *%ENDSNIP.*\n","",l)   
+            l = re.sub("^ *%%%.*\n","",l)   
+            l = re.sub("==l",r"~$\\Ue$~",l) 
+            l = re.sub("===o",r"~$\\Uo$~",l)
+            l = re.sub("==o",r"~$\\Eo$~",l)
+            l = re.sub(".*% *HIDE.*\n","",l)
+            l = re.sub("% label: (.*).* cnt: (.*)",r"~\\customlabel{\g<1>}{(\g<2>)}~",l)
+            l = re.sub("type \(~\$([^ ]+)\$~\) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{(\g<1>)} \\PYG{k+kt}{\g<2>}~",l)
+            l = re.sub("type (\([^ ]+\)) ([^\.]+)",r"~\\PYG{k+kd}{type} \\PYG{n+nf}{\g<1>} \\PYG{k+kt}{\g<2>}~",l)
+            f.write(l)
+        if not raw:
+            f.write("\\end{elpicode}\n")
+
+def mk_fname(fname):
+    return fname.split("/")[-1][:-4] + "tex"
+
+def get_snippets(lines):
+    snips = {}
+    ingrp = False
+    name = ""
+    curgrp = []
+    for l in lines:
+        m = re.match(r"^%ENDSNIP",l)
+        if not (m is None):
+            snips[name] = curgrp
+            ingrp = False
+            curgrp = []
+        if ingrp is True:
+            curgrp = curgrp + [l]
+        m = re.match(r"^%SNIP: *(.*) *$",l)
+        if not (m is None):
+            ingrp = True
+            name = m.group(1)
+            if name in snips:
+                curgrp = snips[name]
+            else:
+                curgrp = []
[…]

### 2024-07-24 paper/paper_for_taps/source/elpi.py

@@ -1,0 +1,178 @@
+"""
+    pygments.lexers.elpi
+    ~~~~~~~~~~~~~~~~~~~~
+
+    Lexer for the `Elpi <http://github.com/LPCIC/elpi>`_ programming language.
+
+    :copyright: Copyright 2006-2024 by the Pygments team, see AUTHORS.
+    :license: BSD, see LICENSE for details.
+"""
+
+from pygments.lexer import RegexLexer, bygroups, include
+from pygments.token import Text, Comment, Operator, Keyword, Name, String, \
+    Number, Punctuation
+
+__all__ = ['ElpiLexer']
+
+
+class ElpiLexer(RegexLexer):
+    """
+    Lexer for the Elpi programming language.
+    """
+
+    name = 'Elpi'
+    url = 'http://github.com/LPCIC/elpi'
+    aliases = ['elpi']
+    filenames = ['*.elpi']
+    mimetypes = ['text/x-elpi']
+    version_added = '2.11'
+
+    lcase_re = r"[a-z]"
+    ucase_re = r"[A-Z]"
+    digit_re = r"[0-9]"
+    schar2_re = r"([+*^?/<>`'@#~=&!])"
+    schar_re = r"({}|-|\$|_)".format(schar2_re)
+    idchar_re = r"({}|{}|{}|{})".format(lcase_re,ucase_re,digit_re,schar_re)
+    idcharstarns_re = r"({}*(\.({}|{}){}*)*)".format(idchar_re, lcase_re, ucase_re, idchar_re)
+    symbchar_re = r"({}|{}|{}|{}|:)".format(lcase_re, ucase_re, digit_re, schar_re)
+    constant_re = r"({}{}*|{}{}|{}{}*|_{}+)".format(ucase_re, idchar_re, lcase_re, idcharstarns_re, schar2_re, symbchar_re, idchar_re)
+    symbol_re = r"(,|<=>|->|:-|;|\?-|->|&|=>|\bas\b|\buvar\b|<|=<|=|==|>=|>|\bi<|\bi=<|\bi>=|\bi>|\bis\b|\br<|\br=<|\br>=|\br>|\bs<|\bs=<|\bs>=|\bs>|@|::|\[\]|`->|`:|`:=|\^|-|\+|\bi-|\bi\+|r-|r\+|/|\*|\bdiv\b|\bi\*|\bmod\b|\br\*|~|\bi~|\br~)"
+    escape_re = r"\(({}|{})\)".format(constant_re,symbol_re)
[…]

### 2024-07-24 paper/paper_for_taps/source/main.tex
update metadata paper

@@
 \acmYear{2024}
-\acmDOI{10.1145/NNNNNNN.NNNNNNN}  % leave as is for now
-\acmConference[PPDP '24]{PPDP '24}{September 10-11, 2024}{Milano, Italy}
-\acmBooktitle{Proceedings of the 26th Symposium on Principles and Practice of Declarative Programming, PPDP 2024, Milano, Italy, September 10-11, 2024}
-\acmPrice{15.00}
-\acmISBN{9-8-4007-0969-297}
+\setcopyright{rightsretained}
+\acmConference[PPDP 2024]{26th International Symposium on Principles and Practice of Declarative Programming}{September 09--11, 2024}{Milano, Italy}
+\acmBooktitle{26th International Symposium on Principles and Practice of Declarative Programming (PPDP 2024), September 09--11, 2024, Milano, Italy}
+\acmPrice{}
+\acmDOI{10.1145/3678232.3678233}
+\acmISBN{979-8-4007-0969-2/24/09}
 

### 2024-07-25 paper/paper_for_taps/source/main.tex
mv pb_printer.tex to pb_printer.sty
rm sty files : content moved into main.tex

--- 2024-07-15 paper/main.tex
@@
 
-\usepackage{myTools}
-\usepackage{macros}
+\usepackage[newfloat]{minted}
+\usepackage{xspace}
+\usepackage{cleveref}
+\usepackage{calc}
+\usepackage{tikz}
 \usepackage{enumitem}
@@
 \usepackage{nameref}
+
+\newcommand{\hoareTriple}[3]{
+  \{#1\}\ #2\ \{#3\}
+}
+
+\newcommand\sepimp{\mathrel{-\mkern-6mu*}}
+
+\newcommand*{\eqdef}{\ensuremath{\mathrel{\overset{\mathrm{def}}{=}}}}
+
+\def\prop{\texttt{Prop}}
+
+% Guillemots for maths
+\def\gl{\text{``}}
+\def\gr{\text{''}}
+\newcommand{\guillemot}[1]{\gl #1 \gr}
+
+% after usepackage{minted}
+\newcommand{\elpiIn}[1]{\mintinline[fontsize=\small,escapeinside=~~]{elpi.py:ElpiLexer -x}{#1}}
+\newcommand{\elpiInFN}[1]{\mintinline[fontsize=\footnotesize,escapeinside=~~]{elpi.py:ElpiLexer -x}{#1}}
+\newcommand{\coqIn}[1]{\mintinline[fontsize=\small,escapeinside=~~]{coq}{#1}}
+\newcommand{\textIn}[1]{\mintinline[escapeinside=~~]{text}{#1}}
+
+\newcommand{\elpiInEq}[2][]{
+  \begin{equation}
+    \texttt{\elpiIn{#2}}
+    \label{#1}
+  \end{equation}
+}
+
+\newcommand{\elpiInEqq}[2][]{
+  \begin{equation*}
+    \texttt{\elpiIn{#2}}
+    \label{#1}
+  \end{equation*}
+}
+
+\newcommand{\coqInEq}[2][]{
+  \begin{equation}
+    \texttt{\coqIn{#2}}
+    \label{#1}
+  \end{equation}
+}
+
+% \newminted{coq}{escapeinside=||}
+% \newminted{elpi}{escapeinside=||}
+
+\newcommand{\inputrawelpicode}[1]{%
+\inputminted[fontsize=\small,autogobble,escapeinside=~~,mathescape=true,frame=leftline,framerule=0pt,framesep=0em]{elpi.py:ElpiLexer -x}{#1_raw.tex}%
+}
+
+\newenvironment{elpicode}{%
+  \vspace{0.2em}%
+  \VerbatimEnvironment
+    \begin{minted}[fontsize=\small,autogobble,escapeinside=~~,mathescape=true,frame=leftline,framerule=0pt,framesep=1em]{elpi.py:ElpiLexer -x}%
+}{%
+  \end{minted}%
+  \vspace{0.2em}%
+}
+\newenvironment{elpicodetab}{%
+  \VerbatimEnvironment
+    \begin{minted}[fontsize=\small,autogobble,escapeinside=~~,mathescape=true,frame=leftline,framerule=0pt,framesep=0em]{elpi.py:ElpiLexer -x}%
+}{%
+  \end{minted}%
+  \vspace{0.2em}%
+}
+
+\newenvironment{textcode}{%
+  \vspace{0.2em}%
+  \VerbatimEnvironment
+  \begin{minted}[breaklines,fontsize=\small,autogobble,escapeinside=~~,mathescape=true,frame=leftline,framerule=0pt,framesep=1em]{text}%
+}{%
+  \end{minted}%
+  \vspace{0.2em}%
+}
+
+\newenvironment{coqcode}{%
+  \vspace{0.2em}%
+  \VerbatimEnvironment
+    \begin{minted}[fontsize=\small,autogobble,escapeinside=~~,mathescape=true,frame=leftline,framerule=0pt,framesep=1em]{coq}%
+}{%
+  \end{minted}%
+  \vspace{0.2em}%
+}
+
+% from amsmath and pifont
+\newcommand{\cmark}{\ding{51}}%
+\newcommand{\xmark}{\ding{55}}%
+
+\newcommand{\proglang}[1]{\texttt{#1}\xspace}
+\newcommand{\pl}[1]{\proglang{#1}}
+
+\def\old{\noindent\hrulefill OLD \noindent\hrulefill}
+
+\makeatletter
+\newcommand{\customlabel}[2]{%
+   \protected@write \@auxout {}{\string \newlabel {#1}{{\ensuremath{#2}}{\thepage}{#2}{#1}{}} }%
+   \hypertarget{#1}{\ensuremath{#2}}
+}
+\makeatother
+
+\newcounter{llength}
+\newcommand{\listlength}[1]{
+  \setcounter{llength}{0}%
+  \foreach \i in #1 {\stepcounter{llength}}%
+}
+
+\newcommand*{\fullref}[1]{\cref{#1} (\textsc{\nameref{#1}})}
+\newcommand*{\Fullref}[1]{\Cref{#1} (\textsc{\nameref{#1}})}
+
+\makeatletter
+\newcommand{\addXdef}[2]{\protected@xdef#1{#1 #2}}
+
+\newcommand{\nth}[3]{
+  \foreach \x [count=\k] in #2 {\ifnum\k=#3 \addXdef{#1}{\x} \fi}
+}
+
+% takes a unif symbol (\Ue or \Uo) and a list of pairs: (lhs, rhs) of unif pb
+\newcommand{\printPb}[2]{
+  \listlength{#2}
+  \def\tableData{}% empty table
+  \foreach \x [count=\i] in #2
+    {
+      \foreach \y [count=\j] in \x {
+        \ifthenelse{\isodd{\j}}
+          {\addXdef{\tableData}{\y & #1 &}}
+          {\addXdef{\tableData}{\y}}
+      } 
+      \ifthenelse{\equal{\i}{2}}{}{
+        \ifthenelse{\equal{\i}{\thellength}}{}{\addXdef{\tableData}{&\quad}}
+      }
+    }
+  \tableData
+}
+
+% Takes a list of triple: coq-var, elpi-var, arity
+\newcommand{\printMap}[1]{
+  \def\tableData{}% empty table
+
+  \newcommand{\printMapAux}[1]{
+    \nth{\tableData}{##1}{1} \addXdef{\tableData}{\mapsto} 
+    \nth{\tableData}{##1}{2} \addXdef{\tableData}{^} \nth{\tableData}{##1}{3}}
+
+  \listlength{#1}
+  \foreach \x [count=\i] in #1
+    {
+      \printMapAux{\x}
+      \ifthenelse{\equal{\thellength}{\i}}{}{
+        \ifthenelse{\equal{\intcalcMod{\i}{3}}{0}}
+        {\addXdef{\tableData}{\\}}
+        {\addXdef{\tableData}{&\quad}}
+        }
+    }
+  \ensuremath{
+    \begin{array}{ccc}
+      \tableData
+    \end{array}
+  }
+}
+
+% Take a list of triple for eta-links: (link-type, ctx, lhs, rhs), link-type is \eta or \beta
+\newcommand{\printLink}[1]{
+  \listlength{#1}
+  \def\tableData{}% empty table
+
+  \newcommand{\printLinkAux}[1]{
+    \nth{\tableData}{##1}{2} \addXdef{\tableData}{\vdash&} 
+    \nth{\tableData}{##1}{3} \addXdef{\tableData}{&=_}
+    \nth{\tableData}{##1}{1} \addXdef{\tableData}{&} 
+    \nth{\tableData}{##1}{4}
+  }
+
+  \foreach \x [count=\i] in #1
+    {
+      \printLinkAux{\x}
+      \ifthenelse{\equal{\thellength}{\i}}{}{
+        \ifthenelse{\isodd{\i}}
+          {\addXdef{\tableData}{&\quad}}
+          {\addXdef{\tableData}{\\}}
+      }
+    }
+  \ensuremath{
+    \begin{array}{rcclrccl}
+      \tableData
+    \end{array}
+  }
+}
+
+\newcommand{\putBigPar}[2]{
+  \listlength{#1}
+  \ifthenelse{\thellength>#2}{\Big}{}
+}
+
+\newcommand{\hideEmpty}[2]{
+  \ifthenelse{\equal{#1}{{}}}{}{#2}
+}
+
+% Input is: 
+%   P = a list of pairs for FoUnifPb    (leftPb, rightPb)
+%   T = a list of pairs for HoUnifPb    (leftPb, rightPb)
+%   M = a list of triples for mappings  (FoVar, HoVar, Arity)
+%   L = a list of 4-uplet for links     (link-type, ctx, lhs, rhs)
+% T, M, L can be empty
+% Note: this macro is used when the length of P > 1
+\newcommand{\printAlll}[4]{
+  \arraycolsep=2pt
+  $$
+  \begin{array}{rrclrcll}
+    \foUnifPb = \{ & \printPb{\Uo}{#1} &\}
+    \hideEmpty{#2}{\\
+      \hoUnifPb = \{ & \printPb{\Ue}{#2} &\}
+    }
+    \hideEmpty{#3}{\\
+      \mapStore = \putBigPar{#3}{3}\{ & \multicolumn{7}{l}{
+        \printMap{#3} ~\putBigPar{#3}{3}\}
+    }}
+    \hideEmpty{#4}{\\
+      \linkStore = \putBigPar{#4}{2}\{ & \multicolumn{7}{l}{
+        \printLink{#4} ~\putBigPar{#4}{2}\}
+    }}
+  \end{array}
+  $$
+}
+
+% Same as printAlll, but the length of P is 1, the if then else
+% seems not to work on the fst parameter of multicolumn
+\newcommand{\printAlllSingle}[4]{
+  \arraycolsep=2pt
+  $$
+  \begin{array}{rrcll}
+    \foUnifPb = \{ & \printPb{\Uo}{#1} &\}
+    \hideEmpty{#2}{\\
+      \hoUnifPb = \{ & \printPb{\Ue}{#2} &\}
+    }
+    \hideEmpty{#3}{\\
+      \mapStore = \putBigPar{#3}{3}\{ & \multicolumn{4}{l}{
+        \printMap{#3} ~\putBigPar{#3}{3}\}
+    }}
+    \hideEmpty{#4}{\\
+      \linkStore = \putBigPar{#4}{2}\{ & \multicolumn{4}{l}{
+        \printLink{#4} ~\putBigPar{#4}{2}\}
+    }}
+  \end{array}
+  $$
+}
+\makeatother
 
@@
 
+\setcopyright{acmlicensed}  %
+\copyrightyear{2024}
+\acmYear{2024}
+\setcopyright{rightsretained}
+\acmConference[PPDP 2024]{26th International Symposium on Principles and Practice of Declarative Programming}{September 09--11, 2024}{Milano, Italy}
+\acmBooktitle{26th International Symposium on Principles and Practice of Declarative Programming (PPDP 2024), September 09--11, 2024, Milano, Italy}
+\acmPrice{}
+\acmDOI{10.1145/3678232.3678233}
+\acmISBN{979-8-4007-0969-2/24/09}
+
 \def\githubUrl{\url{https://github.com/FissoreD/ho-unif-for-free}}
@@
 \newcommand{\hoUnifPb}{\ensuremath{\mathbb{Q}}\xspace}
-
-\include{pb_printer}
 

### 2024-07-26 paper/paper_for_taps/source_src/code/common.tex
new version (seems accepted by taps)

\begin{elpicode}
kind nat type.
type z nat.
type s nat -> nat.

kind bool type.
type tt bool.
type ff bool.

kind option type -> type.
type some A -> option A.
type none option A.

type int->nat int -> nat -> o.
int->nat 0 z.
int->nat R (s N) :- R > 0, R' is R - 1, int->nat R' N.

type nth nat -> list A -> A -> o.
nth z (X :: _) Y  :- !, X = Y.
nth (s N) (_ :: L) X :- nth N L X.

type list-init nat -> nat -> (nat -> A -> o) -> list A -> o.
list-init Len Len _ [].
list-init Len Pos F [A | TL] :- F Pos A, list-init Len (s Pos) F TL.

type iota nat -> list nat -> o.
iota N L :- list-init N z (=) L.

type partition (A -> o) -> list A -> list A -> list A -> o.
partition _ [] [] [].
partition F [A|As] [A|Bs] Cs :- F A, !, partition F As Bs Cs.
partition F [A|As] Bs [A|Cs] :- partition F As Bs Cs.

type index list A -> A -> nat -> o.
index [A|_] A z :- !.
index [_|TL] A (s P) :- index TL A P.

type take-list list A -> A -> list A -> o.
take-list [L|XS] L XS.
take-list [L|XS] L' [L|YS] :- not(L = L'), take-list XS L' YS.

type not-in list A -> A -> o.
not-in [] _.
not-in [X | TL] T :- not (X = T), not-in TL T.

type exists (A -> o) -> list A -> o.
exists F [X|XS] :- F X ; exists F XS.

type forall2 (A -> B -> o) -> list A -> list B -> o.
forall2 _ [] [].
forall2 F [X|XS] [Y|YS] :- F X Y, forall2 F XS YS.

type forall3 (A -> B -> C -> o) -> list A -> list B -> list C -> o.
forall3 _ [] [] [].
forall3 F [X|XS] [Y|YS] [Z|ZS] :- F X Y Z, forall3 F XS YS ZS.

/* same as forall2, sometimes more intuitive name */
type map (A -> B -> o) -> list A -> list B -> o.
map _ [] [].
map F [X|XS] [Y|YS] :- F X Y, map F XS YS.

type fold2 (A -> A1 -> B -> B -> o) -> list A -> list A1 -> B -> B -> o.
fold2 _ [] [] A A.
fold2 F [X|XS] [Y|YS] A A1 :- F X Y A A0, fold2 F XS YS A0 A1.

type fold (A -> B -> B -> o) -> list A -> B -> B -> o.
fold _ [] A A.
fold F [X|XS] A A1 :- F X A A0, fold F XS A0 A1.

~\PYG{k+kd}{type} \PYG{n+nf}{(>n)} \PYG{k+kt}{nat -> nat -> o}~.
s _ >n z.
s N >n s M :- N >n M.

~\PYG{k+kd}{type} \PYG{n+nf}{(>=n)} \PYG{k+kt}{nat -> nat -> o}~.
_ >=n z.
s N >=n s M :- N >=n M.

type diff nat -> nat -> o.
diff N M :- N >n M ; M >n N.

type map-fold2 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> B -> C -> list A1 -> B -> C -> o.
map-fold2 _F [] B C [] B C.
map-fold2 F [X | XS] B C [Y | YS] B2 C2 :-
  F X Y B B1 C C1,
  map-fold2 F XS B1 C1 YS B2 C2.

type mem (list A) -> A -> o.
mem [A|_] A :- !.
mem [_|L] A :- mem L A.

type append list A -> list A -> list A -> o.
append [] A A.
append [A | TL] L [A | L'] :- append TL L L'.

type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
fold4 _ [] [] A A B B.
fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.

type fold6 (A -> A1 -> B -> B -> C -> C -> D -> D -> o) -> list A -> list A1 -> B -> B -> C -> C -> D -> D -> o.
fold6 _ [] [] A A B B C C.
fold6 F [X|XS] [Y|YS] A A1 B B1 C C1 :- F X Y A A0 B B0 C C0, fold6 F XS YS A0 A1 B0 B1 C0 C1.

type flatten-aux list A -> list (list A) -> list A -> o.
flatten-aux [] L L1 :- flatten L L1.
flatten-aux [A|As] L [A|L1] :- flatten-aux As L L1.

type flatten list (list A) -> list A -> o.
flatten [] [].
flatten [H|L] R :- flatten-aux H L R.

type all list o -> o.
all [].
all [P|PS] :- P, all PS.

type forall1 (X -> o) -> list X -> o.
forall1 _ L :- var L, !.
forall1 P [X|XS] :- P X, forall1 P XS.
forall1 _ [].

type none (X -> o) -> list X -> o.
none _ [].
none P [X|Xs] :- not (P X), none P Xs.

type undup list A -> list A -> o.
undup [] [].
undup [X|Xs] [X|R] :- not-in Xs X, !, undup Xs R.
undup [_|Xs] R :- undup Xs R.

type len list A -> nat -> o.
len [] z.
len [_|L] (s X) :- len L X.

type rev list A -> list A -> o.
rev L RL  :- rev.aux L []  RL.
rev.aux [X|XS] ACC R :- rev.aux XS [X|ACC] R.
rev.aux [] L L.

type filter list A -> (A -> prop) -> list A -> o.
filter []    _ [].
filter [X|L] P R :- if (P X) (R = X :: L1) (R = L1), filter L P L1.

type filter2 list A -> list B -> (A -> B -> prop) -> list A -> o.
filter2 [] []   _ [].
filter2 [X|L] [Y|YS] P R :- if (P X Y) (R = X :: L1) (R = L1), filter2 L YS P L1.

type pattern-fragment-prefix-aux list A -> list A -> list A -> list A -> o.
pattern-fragment-prefix-aux [] _ [] [].
pattern-fragment-prefix-aux [X | L] ACC [X | PF] NPF :- name X, not-in ACC X, !, 
  pattern-fragment-prefix-aux L [X | ACC] PF NPF.
pattern-fragment-prefix-aux L _ [] L.

type pattern-fragment-prefix list A -> list A -> list A -> o.
pattern-fragment-prefix L PF NPF :- pattern-fragment-prefix-aux L [] PF NPF.

type pattern-fragment list A -> o.
pattern-fragment L :- distinct_names L.
\end{elpicode}

### 2024-07-26 paper/paper_for_taps/source_src/code/comp.tex
new version (seems accepted by taps)

\begin{elpicode}
  kind arity type.
  type arity nat -> arity.
  kind ovariable type.
  type ov addr -> ovariable.
  kind mvariable type.
  type mv addr -> arity -> mvariable.
  kind mapping type.
  ~\PYG{k+kd}{type} \PYG{n+nf}{(<->)} \PYG{k+kt}{ovariable -> mvariable -> mapping}~.
  typeabbrev mmap (list mapping).

  typeabbrev scope (list tm).
  typeabbrev inctx ho.inctx.
  kind baselink type.
  type link-eta  tm -> tm -> baselink.
  type link-llam tm -> tm -> baselink.
  typeabbrev links (list (inctx baselink)).
  typeabbrev link (inctx baselink).

  macro @val-link-eta T1 T2 :- ho.val (link-eta T1 T2).
  macro @val-link-llam T1 T2 :- ho.val (link-llam T1 T2).

  type get-lhs link -> tm -> o.
  get-lhs (val (link-llam A _)) A.
  get-lhs (val (link-eta A _)) A.

  type get-rhs link -> tm -> o.
  get-rhs (val (link-llam _ A)) A.
  get-rhs (val (link-eta _ A)) A.

  type occurs-rigidly to -> to -> o.
  occurs-rigidly N N.
  occurs-rigidly _ (o-app [o-uva _|_]) :- !, fail.
  occurs-rigidly N (o-app L) :- exists (occurs-rigidly N) L.
  occurs-rigidly N (o-lam B) :- pi x\ occurs-rigidly N (B x).

  type reducible-to list to -> to -> to -> o.
  reducible-to _ N N :- !.
  reducible-to L N (o-app[o-uva _|Args]) :- !, 
    forall1 (x\ exists (reducible-to [] x) Args) [N|L]. 
  reducible-to L N (o-lam B) :- !, 
    pi x\ reducible-to [x | L] N (B x).
  reducible-to L N (o-app [N|Args]) :-
    last-n {len L} Args R,
    forall2 (reducible-to []) R {rev L}.

  type maybe-eta to -> list to -> o.                  ~\customlabel{rule:maybeeta}{(\maybeeta)}~
  maybe-eta (o-app[o-uva _|Args]) L :- !,
    forall1 (x\ exists (reducible-to [] x) Args) L, !. 
  maybe-eta (o-lam B) L :- !, pi x\ maybe-eta (B x) [x | L].
  maybe-eta (o-app [T|Args]) L :- (name T; T = o-con _),
    split-last-n {len L} Args First Last,
    none (x\ exists (y\ occurs-rigidly x y) First) L,
    forall2 (reducible-to []) {rev L} Last.

  type locally-bound tm -> o.
  type get-scope-aux tm -> list tm -> o.
  get-scope-aux (con _) [].
  get-scope-aux (uva _ L) L1 :- 
    forall2 get-scope-aux L R,
    flatten R L1.
  get-scope-aux (lam B) L1 :- 
    pi x\ locally-bound x => get-scope-aux (B x) L1.
  get-scope-aux (app L) L1 :- 
    forall2 get-scope-aux L R,
    flatten R L1.
  get-scope-aux X [X] :- name X, not (locally-bound X).
  get-scope-aux X [] :- name X, (locally-bound X).

  type names1 list tm -> o.
  names1 L :-
    names L1, 
    new_int N,
    if (1 is N mod 2) (L1 = L) (rev L1 L).

  type get-scope tm -> list tm -> o.
  get-scope T Scope :-
    get-scope-aux T ScopeDuplicata,
    undup ScopeDuplicata Scope.
  type rigid to -> o.
  rigid X :- not (X = o-uva _).

  type comp-lam (to -> to) -> (tm -> tm) -> 
    mmap -> mmap -> links -> links -> subst -> subst -> o.
  comp-lam F G M1 M2 L1 L3 S1 S2 :-
    pi x y\ (pi M L S\ comp x y M M L L S S) =>         ~\customlabel{rule:complamh}{(\ensuremath{H_\lambda})}~
      comp (F x) (G y) M1 M2 L1 (L2 y) S1 S2,
    close-links L2 L3.

  type close-links (tm -> links) -> links -> o.
  close-links (v\[X v|L v]) [abs X|R] :- close-links L R.
  close-links (_\[]) [].
  type comp to -> tm -> mmap -> mmap -> links -> links ->
    subst -> subst -> o.
  comp (o-con C) (con C) M M L L S S.
  comp (o-lam F) (uva A Scope) M1 M2 L1 L3 S1 S3 :-
    maybe-eta (o-lam F) [], !,
      alloc S1 A S2,
      comp-lam F F1 M1 M2 L1 L2 S2 S3,
      get-scope (lam F1) Scope, 
      L3 = [val (link-eta (uva A Scope) (lam F1)) | L2].
  comp (o-lam F) (lam F1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:complam}{(\ensuremath{c_\lambda})}~
    comp-lam F F1 M1 M2 L1 L2 S1 S2.
  comp (o-uva A) (uva B []) M1 M2 L L S1 S2 :- 
    m-alloc (ov A) (mv B (arity z)) M1 M2 S1 S2.
  comp (o-app [o-uva A|Ag]) (uva B Ag1) M1 M2 L L S1 S2 :-
    pattern-fragment Ag, !,
      fold6 comp Ag Ag1 M1 M1 L L S1 S1,
      len Ag Arity, 
      m-alloc (ov A) (mv B (arity Arity)) M1 M2 S1 S2.
  comp (o-app [o-uva A|Ag]) (uva B Sc) M1 M3 L1 L3 S1 S4 :- !,
    pattern-fragment-prefix Ag Pf Extra, alloc S1 B S2,
    len Pf Ar, m-alloc (ov A) (mv C (arity Ar)) M1 M2 S2 S3,
    fold6 comp Pf    Pf1    M2 M2 L1 L1 S3 S3,
    fold6 comp Extra Extra1 M2 M3 L1 L2 S3 S4,
    Beta = app [uva C Pf1 | Extra1],
    get-scope Beta Sc, 
    L3 = [val (link-llam (uva B Sc) Beta) | L2].
  comp (o-app A) (app A1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:compapp}{(\ensuremath{c_@})}~
    fold6 comp A A1 M1 M2 L1 L2 S1 S2.

  type alloc mem A -> addr -> mem A -> o.
  alloc S N S1 :- mem.new S N S1.

  type compile-terms-diagnostic 
    triple diagnostic to to -> 
    triple diagnostic tm tm -> 
    mmap -> mmap -> 
    links -> links -> 
    subst -> subst -> o.
  compile-terms-diagnostic (triple D FO1 FO2) (triple D HO1 HO2) M1 M3 L1 L3 S1 S3 :-
    beta-normal FO1 FO1',
    beta-normal FO2 FO2',
    comp FO1' HO1 M1 M2 L1 L2 S1 S2,
    comp FO2' HO2 M2 M3 L2 L3 S2 S3.

  type compile-terms 
    list (triple diagnostic to to) -> 
    list (triple diagnostic tm tm) -> 
    mmap -> links -> subst -> o.
  compile-terms T H M L S :- 
    fold6 compile-terms-diagnostic T H [] M_ [] L_ [] S_,
    print-compil-result T H L_ M_, 
    deduplicate-map M_ M S_ S L_ L.

  type make-eta-link-aux nat -> addr -> addr -> 
    list tm -> links -> subst -> subst -> o.
  make-eta-link-aux z Ad1 Ad2 Scope1 L H1 H1  :-
    rev Scope1 Scope, eta-expand (uva Ad2 Scope) T1,
    L = [val (link-eta (uva Ad1 Scope) T1)].
  make-eta-link-aux (s N) Ad1 Ad2 Scope1 L H1 H3 :-
    rev Scope1 Scope, alloc H1 Ad H2, 
    eta-expand (uva Ad Scope) T2,
    (pi x\ make-eta-link-aux N Ad Ad2 [x|Scope1] (L1 x) H2 H3),
    close-links L1 L2,
    L = [val (link-eta (uva Ad1 Scope) T2) | L2].

  type make-eta-link nat -> nat -> addr -> addr -> 
          list tm -> links -> subst -> subst -> o.
  make-eta-link (s N) z Ad1 Ad2 Vars L H H1 :- 
    make-eta-link-aux N Ad2 Ad1 Vars L H H1.
  make-eta-link z (s N) Ad1 Ad2 Vars L H H1 :- 
    make-eta-link-aux N Ad1 Ad2 Vars L H H1.
  make-eta-link (s N) (s M) Ad1 Ad2 Vars Links H H1 :-
    (pi x\ make-eta-link N M Ad1 Ad2 [x|Vars] (L x) H H1),
    close-links L Links.

  type deduplicate-map mmap -> mmap -> 
      subst -> subst -> links -> links -> o.
  deduplicate-map [] [] H H L L.
  deduplicate-map [((ov O <-> mv M (arity LenM)) as X1) | Map1] Map2 H1 H3 L1 L3 :-
    take-list Map1 ((ov O <-> mv M' (arity LenM'))) _, !,
    std.assert! (not (LenM = LenM')) "Deduplicate map, there is a bug",
    print "arity-fix links:" {ppmapping X1} "~!~" {ppmapping ((ov O <-> mv M' (arity LenM')))}, 
    make-eta-link LenM LenM' M M' [] New H1 H2,
    print "new eta link" {pplinks New},
    append New L1 L2,
    deduplicate-map Map1 Map2 H2 H3 L2 L3.
  deduplicate-map [A|As] [A|Bs] H1 H2 L1 L2 :- 
    deduplicate-map As Bs H1 H2 L1 L2, !.
  deduplicate-map [A|_] _ H _ _ _ :- 
    halt "deduplicating mapping error" {ppmapping A} {ho.ppsubst H}.
\end{elpicode}

### 2024-07-26 paper/paper_for_taps/source_src/code/maybe_eta.tex
new version (seems accepted by taps)

--- 2024-07-26 paper/paper_for_taps/source_src/code/comp.tex
@@
 \begin{elpicode}
-  kind arity type.
-  type arity nat -> arity.
-  kind ovariable type.
-  type ov addr -> ovariable.
-  kind mvariable type.
-  type mv addr -> arity -> mvariable.
-  kind mapping type.
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(<->)} \PYG{k+kt}{ovariable -> mvariable -> mapping}~.
-  typeabbrev mmap (list mapping).
-
-  typeabbrev scope (list tm).
-  typeabbrev inctx ho.inctx.
-  kind baselink type.
-  type link-eta  tm -> tm -> baselink.
-  type link-llam tm -> tm -> baselink.
-  typeabbrev links (list (inctx baselink)).
-  typeabbrev link (inctx baselink).
-
-  macro @val-link-eta T1 T2 :- ho.val (link-eta T1 T2).
-  macro @val-link-llam T1 T2 :- ho.val (link-llam T1 T2).
-
-  type get-lhs link -> tm -> o.
-  get-lhs (val (link-llam A _)) A.
-  get-lhs (val (link-eta A _)) A.
-
-  type get-rhs link -> tm -> o.
-  get-rhs (val (link-llam _ A)) A.
-  get-rhs (val (link-eta _ A)) A.
-
   type occurs-rigidly to -> to -> o.
@@
     forall2 (reducible-to []) {rev L} Last.
-
-  type locally-bound tm -> o.
-  type get-scope-aux tm -> list tm -> o.
-  get-scope-aux (con _) [].
-  get-scope-aux (uva _ L) L1 :- 
-    forall2 get-scope-aux L R,
-    flatten R L1.
-  get-scope-aux (lam B) L1 :- 
-    pi x\ locally-bound x => get-scope-aux (B x) L1.
-  get-scope-aux (app L) L1 :- 
-    forall2 get-scope-aux L R,
-    flatten R L1.
-  get-scope-aux X [X] :- name X, not (locally-bound X).
-  get-scope-aux X [] :- name X, (locally-bound X).
-
-  type names1 list tm -> o.
-  names1 L :-
-    names L1, 
-    new_int N,
-    if (1 is N mod 2) (L1 = L) (rev L1 L).
-
-  type get-scope tm -> list tm -> o.
-  get-scope T Scope :-
-    get-scope-aux T ScopeDuplicata,
-    undup ScopeDuplicata Scope.
-  type rigid to -> o.
-  rigid X :- not (X = o-uva _).
-
-  type comp-lam (to -> to) -> (tm -> tm) -> 
-    mmap -> mmap -> links -> links -> subst -> subst -> o.
-  comp-lam F G M1 M2 L1 L3 S1 S2 :-
-    pi x y\ (pi M L S\ comp x y M M L L S S) =>         ~\customlabel{rule:complamh}{(\ensuremath{H_\lambda})}~
-      comp (F x) (G y) M1 M2 L1 (L2 y) S1 S2,
-    close-links L2 L3.
-
-  type close-links (tm -> links) -> links -> o.
-  close-links (v\[X v|L v]) [abs X|R] :- close-links L R.
-  close-links (_\[]) [].
-  type comp to -> tm -> mmap -> mmap -> links -> links ->
-    subst -> subst -> o.
-  comp (o-con C) (con C) M M L L S S.
-  comp (o-lam F) (uva A Scope) M1 M2 L1 L3 S1 S3 :-
-    maybe-eta (o-lam F) [], !,
-      alloc S1 A S2,
-      comp-lam F F1 M1 M2 L1 L2 S2 S3,
-      get-scope (lam F1) Scope, 
-      L3 = [val (link-eta (uva A Scope) (lam F1)) | L2].
-  comp (o-lam F) (lam F1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:complam}{(\ensuremath{c_\lambda})}~
-    comp-lam F F1 M1 M2 L1 L2 S1 S2.
-  comp (o-uva A) (uva B []) M1 M2 L L S1 S2 :- 
-    m-alloc (ov A) (mv B (arity z)) M1 M2 S1 S2.
-  comp (o-app [o-uva A|Ag]) (uva B Ag1) M1 M2 L L S1 S2 :-
-    pattern-fragment Ag, !,
-      fold6 comp Ag Ag1 M1 M1 L L S1 S1,
-      len Ag Arity, 
-      m-alloc (ov A) (mv B (arity Arity)) M1 M2 S1 S2.
-  comp (o-app [o-uva A|Ag]) (uva B Sc) M1 M3 L1 L3 S1 S4 :- !,
-    pattern-fragment-prefix Ag Pf Extra, alloc S1 B S2,
-    len Pf Ar, m-alloc (ov A) (mv C (arity Ar)) M1 M2 S2 S3,
-    fold6 comp Pf    Pf1    M2 M2 L1 L1 S3 S3,
-    fold6 comp Extra Extra1 M2 M3 L1 L2 S3 S4,
-    Beta = app [uva C Pf1 | Extra1],
-    get-scope Beta Sc, 
-    L3 = [val (link-llam (uva B Sc) Beta) | L2].
-  comp (o-app A) (app A1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:compapp}{(\ensuremath{c_@})}~
-    fold6 comp A A1 M1 M2 L1 L2 S1 S2.
-
-  type alloc mem A -> addr -> mem A -> o.
-  alloc S N S1 :- mem.new S N S1.
-
-  type compile-terms-diagnostic 
-    triple diagnostic to to -> 
-    triple diagnostic tm tm -> 
-    mmap -> mmap -> 
-    links -> links -> 
-    subst -> subst -> o.
-  compile-terms-diagnostic (triple D FO1 FO2) (triple D HO1 HO2) M1 M3 L1 L3 S1 S3 :-
-    beta-normal FO1 FO1',
-    beta-normal FO2 FO2',
-    comp FO1' HO1 M1 M2 L1 L2 S1 S2,
-    comp FO2' HO2 M2 M3 L2 L3 S2 S3.
-
-  type compile-terms 
-    list (triple diagnostic to to) -> 
-    list (triple diagnostic tm tm) -> 
-    mmap -> links -> subst -> o.
-  compile-terms T H M L S :- 
-    fold6 compile-terms-diagnostic T H [] M_ [] L_ [] S_,
-    print-compil-result T H L_ M_, 
-    deduplicate-map M_ M S_ S L_ L.
-
-  type make-eta-link-aux nat -> addr -> addr -> 
-    list tm -> links -> subst -> subst -> o.
-  make-eta-link-aux z Ad1 Ad2 Scope1 L H1 H1  :-
-    rev Scope1 Scope, eta-expand (uva Ad2 Scope) T1,
-    L = [val (link-eta (uva Ad1 Scope) T1)].
-  make-eta-link-aux (s N) Ad1 Ad2 Scope1 L H1 H3 :-
-    rev Scope1 Scope, alloc H1 Ad H2, 
-    eta-expand (uva Ad Scope) T2,
-    (pi x\ make-eta-link-aux N Ad Ad2 [x|Scope1] (L1 x) H2 H3),
-    close-links L1 L2,
-    L = [val (link-eta (uva Ad1 Scope) T2) | L2].
-
-  type make-eta-link nat -> nat -> addr -> addr -> 
-          list tm -> links -> subst -> subst -> o.
-  make-eta-link (s N) z Ad1 Ad2 Vars L H H1 :- 
-    make-eta-link-aux N Ad2 Ad1 Vars L H H1.
-  make-eta-link z (s N) Ad1 Ad2 Vars L H H1 :- 
-    make-eta-link-aux N Ad1 Ad2 Vars L H H1.
-  make-eta-link (s N) (s M) Ad1 Ad2 Vars Links H H1 :-
-    (pi x\ make-eta-link N M Ad1 Ad2 [x|Vars] (L x) H H1),
-    close-links L Links.
-
-  type deduplicate-map mmap -> mmap -> 
-      subst -> subst -> links -> links -> o.
-  deduplicate-map [] [] H H L L.
-  deduplicate-map [((ov O <-> mv M (arity LenM)) as X1) | Map1] Map2 H1 H3 L1 L3 :-
-    take-list Map1 ((ov O <-> mv M' (arity LenM'))) _, !,
-    std.assert! (not (LenM = LenM')) "Deduplicate map, there is a bug",
-    print "arity-fix links:" {ppmapping X1} "~!~" {ppmapping ((ov O <-> mv M' (arity LenM')))}, 
-    make-eta-link LenM LenM' M M' [] New H1 H2,
-    print "new eta link" {pplinks New},
-    append New L1 L2,
-    deduplicate-map Map1 Map2 H2 H3 L2 L3.
-  deduplicate-map [A|As] [A|Bs] H1 H2 L1 L2 :- 
-    deduplicate-map As Bs H1 H2 L1 L2, !.
-  deduplicate-map [A|_] _ H _ _ _ :- 
-    halt "deduplicating mapping error" {ppmapping A} {ho.ppsubst H}.
 \end{elpicode}

### 2024-07-26 paper/paper_for_taps/source_src/code/maybe_eta_raw.tex
new version (seems accepted by taps)

--- 2024-07-26 paper/paper_for_taps/source_src/code/comp.tex
@@
-\begin{elpicode}
-  kind arity type.
-  type arity nat -> arity.
-  kind ovariable type.
-  type ov addr -> ovariable.
-  kind mvariable type.
-  type mv addr -> arity -> mvariable.
-  kind mapping type.
-  ~\PYG{k+kd}{type} \PYG{n+nf}{(<->)} \PYG{k+kt}{ovariable -> mvariable -> mapping}~.
-  typeabbrev mmap (list mapping).
-
-  typeabbrev scope (list tm).
-  typeabbrev inctx ho.inctx.
-  kind baselink type.
-  type link-eta  tm -> tm -> baselink.
-  type link-llam tm -> tm -> baselink.
-  typeabbrev links (list (inctx baselink)).
-  typeabbrev link (inctx baselink).
-
-  macro @val-link-eta T1 T2 :- ho.val (link-eta T1 T2).
-  macro @val-link-llam T1 T2 :- ho.val (link-llam T1 T2).
-
-  type get-lhs link -> tm -> o.
-  get-lhs (val (link-llam A _)) A.
-  get-lhs (val (link-eta A _)) A.
-
-  type get-rhs link -> tm -> o.
-  get-rhs (val (link-llam _ A)) A.
-  get-rhs (val (link-eta _ A)) A.
-
   type occurs-rigidly to -> to -> o.
@@
     forall2 (reducible-to []) {rev L} Last.
-
-  type locally-bound tm -> o.
-  type get-scope-aux tm -> list tm -> o.
-  get-scope-aux (con _) [].
-  get-scope-aux (uva _ L) L1 :- 
-    forall2 get-scope-aux L R,
-    flatten R L1.
-  get-scope-aux (lam B) L1 :- 
-    pi x\ locally-bound x => get-scope-aux (B x) L1.
-  get-scope-aux (app L) L1 :- 
-    forall2 get-scope-aux L R,
-    flatten R L1.
-  get-scope-aux X [X] :- name X, not (locally-bound X).
-  get-scope-aux X [] :- name X, (locally-bound X).
-
-  type names1 list tm -> o.
-  names1 L :-
-    names L1, 
-    new_int N,
-    if (1 is N mod 2) (L1 = L) (rev L1 L).
-
-  type get-scope tm -> list tm -> o.
-  get-scope T Scope :-
-    get-scope-aux T ScopeDuplicata,
-    undup ScopeDuplicata Scope.
-  type rigid to -> o.
-  rigid X :- not (X = o-uva _).
-
-  type comp-lam (to -> to) -> (tm -> tm) -> 
-    mmap -> mmap -> links -> links -> subst -> subst -> o.
-  comp-lam F G M1 M2 L1 L3 S1 S2 :-
-    pi x y\ (pi M L S\ comp x y M M L L S S) =>         ~\customlabel{rule:complamh}{(\ensuremath{H_\lambda})}~
-      comp (F x) (G y) M1 M2 L1 (L2 y) S1 S2,
-    close-links L2 L3.
-
-  type close-links (tm -> links) -> links -> o.
-  close-links (v\[X v|L v]) [abs X|R] :- close-links L R.
-  close-links (_\[]) [].
-  type comp to -> tm -> mmap -> mmap -> links -> links ->
-    subst -> subst -> o.
-  comp (o-con C) (con C) M M L L S S.
-  comp (o-lam F) (uva A Scope) M1 M2 L1 L3 S1 S3 :-
-    maybe-eta (o-lam F) [], !,
-      alloc S1 A S2,
-      comp-lam F F1 M1 M2 L1 L2 S2 S3,
-      get-scope (lam F1) Scope, 
-      L3 = [val (link-eta (uva A Scope) (lam F1)) | L2].
-  comp (o-lam F) (lam F1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:complam}{(\ensuremath{c_\lambda})}~
-    comp-lam F F1 M1 M2 L1 L2 S1 S2.
-  comp (o-uva A) (uva B []) M1 M2 L L S1 S2 :- 
-    m-alloc (ov A) (mv B (arity z)) M1 M2 S1 S2.
-  comp (o-app [o-uva A|Ag]) (uva B Ag1) M1 M2 L L S1 S2 :-
-    pattern-fragment Ag, !,
-      fold6 comp Ag Ag1 M1 M1 L L S1 S1,
-      len Ag Arity, 
-      m-alloc (ov A) (mv B (arity Arity)) M1 M2 S1 S2.
-  comp (o-app [o-uva A|Ag]) (uva B Sc) M1 M3 L1 L3 S1 S4 :- !,
-    pattern-fragment-prefix Ag Pf Extra, alloc S1 B S2,
-    len Pf Ar, m-alloc (ov A) (mv C (arity Ar)) M1 M2 S2 S3,
-    fold6 comp Pf    Pf1    M2 M2 L1 L1 S3 S3,
-    fold6 comp Extra Extra1 M2 M3 L1 L2 S3 S4,
-    Beta = app [uva C Pf1 | Extra1],
-    get-scope Beta Sc, 
-    L3 = [val (link-llam (uva B Sc) Beta) | L2].
-  comp (o-app A) (app A1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:compapp}{(\ensuremath{c_@})}~
-    fold6 comp A A1 M1 M2 L1 L2 S1 S2.
-
-  type alloc mem A -> addr -> mem A -> o.
-  alloc S N S1 :- mem.new S N S1.
-
-  type compile-terms-diagnostic 
-    triple diagnostic to to -> 
-    triple diagnostic tm tm -> 
-    mmap -> mmap -> 
-    links -> links -> 
-    subst -> subst -> o.
-  compile-terms-diagnostic (triple D FO1 FO2) (triple D HO1 HO2) M1 M3 L1 L3 S1 S3 :-
-    beta-normal FO1 FO1',
-    beta-normal FO2 FO2',
-    comp FO1' HO1 M1 M2 L1 L2 S1 S2,
-    comp FO2' HO2 M2 M3 L2 L3 S2 S3.
-
-  type compile-terms 
-    list (triple diagnostic to to) -> 
-    list (triple diagnostic tm tm) -> 
-    mmap -> links -> subst -> o.
-  compile-terms T H M L S :- 
-    fold6 compile-terms-diagnostic T H [] M_ [] L_ [] S_,
-    print-compil-result T H L_ M_, 
-    deduplicate-map M_ M S_ S L_ L.
-
-  type make-eta-link-aux nat -> addr -> addr -> 
-    list tm -> links -> subst -> subst -> o.
-  make-eta-link-aux z Ad1 Ad2 Scope1 L H1 H1  :-
-    rev Scope1 Scope, eta-expand (uva Ad2 Scope) T1,
-    L = [val (link-eta (uva Ad1 Scope) T1)].
-  make-eta-link-aux (s N) Ad1 Ad2 Scope1 L H1 H3 :-
-    rev Scope1 Scope, alloc H1 Ad H2, 
-    eta-expand (uva Ad Scope) T2,
-    (pi x\ make-eta-link-aux N Ad Ad2 [x|Scope1] (L1 x) H2 H3),
-    close-links L1 L2,
-    L = [val (link-eta (uva Ad1 Scope) T2) | L2].
-
-  type make-eta-link nat -> nat -> addr -> addr -> 
-          list tm -> links -> subst -> subst -> o.
-  make-eta-link (s N) z Ad1 Ad2 Vars L H H1 :- 
-    make-eta-link-aux N Ad2 Ad1 Vars L H H1.
-  make-eta-link z (s N) Ad1 Ad2 Vars L H H1 :- 
-    make-eta-link-aux N Ad1 Ad2 Vars L H H1.
-  make-eta-link (s N) (s M) Ad1 Ad2 Vars Links H H1 :-
-    (pi x\ make-eta-link N M Ad1 Ad2 [x|Vars] (L x) H H1),
-    close-links L Links.
-
-  type deduplicate-map mmap -> mmap -> 
-      subst -> subst -> links -> links -> o.
-  deduplicate-map [] [] H H L L.
-  deduplicate-map [((ov O <-> mv M (arity LenM)) as X1) | Map1] Map2 H1 H3 L1 L3 :-
-    take-list Map1 ((ov O <-> mv M' (arity LenM'))) _, !,
-    std.assert! (not (LenM = LenM')) "Deduplicate map, there is a bug",
-    print "arity-fix links:" {ppmapping X1} "~!~" {ppmapping ((ov O <-> mv M' (arity LenM')))}, 
-    make-eta-link LenM LenM' M M' [] New H1 H2,
-    print "new eta link" {pplinks New},
-    append New L1 L2,
-    deduplicate-map Map1 Map2 H2 H3 L2 L3.
-  deduplicate-map [A|As] [A|Bs] H1 H2 L1 L2 :- 
-    deduplicate-map As Bs H1 H2 L1 L2, !.
-  deduplicate-map [A|_] _ H _ _ _ :- 
-    halt "deduplicating mapping error" {ppmapping A} {ho.ppsubst H}.
-\end{elpicode}

### 2024-07-26 paper/paper_for_taps/source_src/code/test.tex
new version (seems accepted by taps)

\begin{elpicode}

type hstep tm -> tm -> links -> links -> subst -> subst -> o.
hstep T1 T2 L1 L2 S1 S3 :-
  (T1 ~$\Ue$~ T2) S1 S2,
  progress L1 L2 S2 S3.

type compile to -> tm -> mmap -> mmap ->  links -> links -> 
  subst -> subst -> o.
compile F G M1 M2 L1 L2 S1 S2 :-
  beta-normal F F', comp F' G M1 M2 L1 L2 S1 S2.

~\PYG{k+kd}{type} \PYG{n+nf}{(\Uo)} \PYG{k+kt}{to -> to -> fsubst -> o}~.
(A ~$\Uo$~ B) F :-
  compile A A' [] M1 [] L1 [] S1,
  compile B B' M1 M2 L1 L2 S1 S2,
  hstep A' B' L2 L3 S2 S3,
  decompile M2 L3 S3 [] F.

\end{elpicode}

### 2024-07-29 paper/paper_for_taps/source/PPDP2024-1.tex
new version (seems accepted by taps)
change dates 09-11 -> 10-11

--- 2024-07-25 paper/paper_for_taps/source/main.tex
@@
 
-\usepackage[newfloat]{minted}
+\usepackage[newfloat,frozencache]{minted}
 \usepackage{xspace}
@@
 \setcopyright{rightsretained}
-\acmConference[PPDP 2024]{26th International Symposium on Principles and Practice of Declarative Programming}{September 09--11, 2024}{Milano, Italy}
-\acmBooktitle{26th International Symposium on Principles and Practice of Declarative Programming (PPDP 2024), September 09--11, 2024, Milano, Italy}
+\acmConference[PPDP 2024]{26th International Symposium on Principles and Practice of Declarative Programming}{September 10--11, 2024}{Milano, Italy}
+\acmBooktitle{26th International Symposium on Principles and Practice of Declarative Programming (PPDP 2024), September 10--11, 2024, Milano, Italy}
 \acmPrice{}
@@
 
+\begin{acks}
+
+This work has been supported by the French government, through the France 2030
+investment plan managed by the Agence Nationale de la Recherche, as part of the
+"UCA DS4H" project, reference ANR-17-EURE-0004.
+
+\end{acks}
 % ---

### 2024-07-29 paper/paper_for_taps/source_src/PPDP2024-1.tex
new version (seems accepted by taps)
change dates 09-11 -> 10-11

--- 2024-07-25 paper/paper_for_taps/source/main.tex
@@
 
-\usepackage[newfloat]{minted}
+\usepackage[newfloat,finalizecache]{minted}
 \usepackage{xspace}
